---
title: "Step 0 — Check the R environment and Cell Ranger ARC inputs"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# What this step teaches

Before creating a Seurat object, we verify three things:

1. R can load every required package from the same library that will be used for the analysis.
2. Each row of the sample sheet represents a real biological library and points to the expected Cell Ranger ARC outputs.
3. The genome annotation uses the same mouse genome naming convention as the ATAC data.

This is a light validation step. A reasonable SOL RStudio request is **2 CPU cores, 16 GiB memory, 2 hours, R 4.4.2, no GPU**.

Do not run analysis on a SOL login node. Open this notebook from an allocated Open OnDemand RStudio Server session.

# 1. Open the RStudio Project

Open `multiome_scRNA-ATAC.Rproj` before running this notebook. The setup chunk finds the repository root whether the notebook working directory is the project root or the `analysis/` directory.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 8,
  fig.height = 5
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
```

# 2. Confirm R and package namespaces

`installed.packages()` can report that a top-level package exists even when one of its imported dependencies is missing. For that reason, this check calls `loadNamespace()` on each required package. It catches the same class of dependency failures previously seen for `RANN`, `spam`, `cluster`, `igraph`, `fastmatch`, and `biovizBase`.

If this chunk stops, run `setup/install_packages.R` in a clean R 4.4.2 session and then restart RStudio.

```{r package-preflight}
check_r_version()
package_status <- package_preflight()
package_status
```

```{r current-libraries}
cat("R version:\n", R.version.string, "\n\n")
cat("R executable:\n", Sys.which("R"), "\n\n")
cat("R_LIBS_USER:\n", Sys.getenv("R_LIBS_USER"), "\n\n")
cat("Library search paths:\n")
print(.libPaths())
```

# 3. Read the sample sheet

The sample sheet is the biological map for the analysis. `sample_id` names a library; `replicate_id` names the independent animal/specimen that could eventually support statistical replication.

```{r read-sample-sheet}
sample_sheet <- read_sample_sheet()
sample_sheet[, c(
  "sample_id",
  "replicate_id",
  "genotype",
  "diet",
  "outs_dir"
)]
```

## Important design limitation

There is one `CON` sample and one `HFD` sample. Therefore, diet and sample are completely confounded. Thousands of nuclei do not turn one mouse into thousands of biological replicates. Downstream HFD-versus-CON summaries will be descriptive until additional independent animals are added.

```{r replication-table}
replication_table <- as.data.frame(with(
  sample_sheet,
  table(diet, replicate_id)
))
replication_table[replication_table$Freq > 0, ]
```

# 4. Validate required Cell Ranger ARC files

For each library we require:

- `filtered_feature_bc_matrix.h5`: paired Gene Expression and Peaks matrices;
- `atac_fragments.tsv.gz` and `.tbi`: ATAC fragments plus tabix index;
- `atac_peaks.bed`: sample-specific Cell Ranger peak calls;
- `per_barcode_metrics.csv`: paired RNA/ATAC barcode metrics.

The optional `web_summary.html` should also be reviewed before cell-level QC because a severely failed library cannot be rescued by filtering individual barcodes.

```{r validate-inputs}
input_report <- validate_cellranger_inputs(sample_sheet)
input_report
```

```{r locate-web-summaries}
data.frame(
  sample_id = sample_sheet$sample_id,
  web_summary = sample_sheet$web_summary,
  exists = file.exists(sample_sheet$web_summary)
)
```

Open each existing `web_summary.html` from the RStudio Files pane. At the library level, discuss:

- how many nuclei Cell Ranger called;
- sequencing saturation and read depth;
- fraction of reads in peaks;
- TSS enrichment;
- whether RNA and ATAC metrics both look plausible;
- whether one library differs dramatically from the other.

# 5. Build the mm10 gene annotation

Signac needs gene coordinates to calculate TSS enrichment and later to connect peaks to genes. The annotation is derived from `EnsDb.Mmusculus.v79`, converted to UCSC chromosome names such as `chr1`, and labeled `mm10` to match the current Cell Ranger ARC reference.

```{r build-annotation}
initialize_output_dirs()

annotation_path <- file.path(
  OUTPUT_DIR,
  "00_run_info",
  "mm10_gene_annotation.rds"
)

gene_annotation <- build_mm10_annotation()
saveRDS(gene_annotation, annotation_path, compress = FALSE)

gene_annotation
```

# 6. Save the validation record

These small files document exactly what R environment and input paths were used. They are written to the project results directory on SOL, not to GitHub.

```{r save-validation}
readr::write_csv(
  input_report,
  file.path(OUTPUT_DIR, "00_run_info", "cellranger_input_report.csv")
)
readr::write_csv(
  sample_sheet,
  file.path(OUTPUT_DIR, "00_run_info", "sample_sheet_used.csv")
)
save_run_information(package_status)

writeLines(
  paste("Step 0 completed:", format(Sys.time())),
  file.path(OUTPUT_DIR, "00_run_info", "STEP_00_COMPLETE.txt")
)
```

# Interpretation checkpoint

Before moving to Step 1, you should be able to answer:

1. What is the difference between a sequencing library and a biological replicate?
2. Why are `atac_fragments.tsv.gz` and its `.tbi` index both required?
3. Why do we inspect Cell Ranger's library-level report before filtering cells?
4. Why is HFD-versus-CON inference not valid with one mouse per diet?

Expected completion marker:

```{r expected-output, eval=FALSE}
file.path(OUTPUT_DIR, "00_run_info", "STEP_00_COMPLETE.txt")
```

# References

See `docs/REFERENCES.md`, especially the official 10x Genomics output documentation and the Signac Multiome vignette.

