---
title: "Step 1 — Import each library and calculate Multiome QC metrics"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# What this step does—and does not do

This step imports one Cell Ranger ARC library at a time and calculates QC information. It deliberately **does not make the final keep/remove decision**.

For each paired barcode it records:

- RNA counts, detected genes, mitochondrial fraction, and RNA complexity;
- ATAC counts, TSS enrichment, nucleosome signal, and Cell Ranger FRiP;
- a likely singlet/doublet call from `scDblFinder` using RNA counts.

The output is an unfiltered QC checkpoint for each sample, a metadata CSV that is easy to plot, and a sample-aware starting suggestion for thresholds.

This is a substantial computation. Begin with **8 CPU cores, 96 GiB memory, 12 hours, R 4.4.2, no GPU**. If SOL reports an out-of-memory termination, rerun this same step with 128 GiB; completed sample checkpoints are reused.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 8,
  fig.height = 5
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()
configure_parallel_plan()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
})

# TRUE resumes from an existing per-sample QC-metric checkpoint.
# Set FALSE only when the import/QC calculation itself must be rebuilt.
REUSE_QC_METRIC_CHECKPOINTS <- TRUE
```

# 2. Confirm Step 0 and load project metadata

```{r confirm-step-zero}
step_zero_marker <- file.path(
  OUTPUT_DIR,
  "00_run_info",
  "STEP_00_COMPLETE.txt"
)
assert_file(step_zero_marker, "Step 0 completion marker")

sample_sheet <- read_sample_sheet()
validate_cellranger_inputs(sample_sheet)

annotation_path <- file.path(
  OUTPUT_DIR,
  "00_run_info",
  "GRCm39_GENCODE_vM33_gene_annotation.rds"
)

assert_file(
  annotation_path,
  "saved GRCm39 GENCODE vM33 annotation"
)

gene_annotation <- readRDS(annotation_path)

if (!identical(unique(GenomeInfoDb::genome(gene_annotation)), "GRCm39")) {
  stop(
    "The saved gene annotation is not labeled as GRCm39.",
    call. = FALSE
  )
}
```

# 3. Why QC begins separately by sample

Two libraries can differ in capture efficiency, sequencing depth, or ATAC complexity even when they contain similar biology. A global count cutoff applied after merging can preferentially remove cells from the shallower library. We therefore calculate and inspect distributions within each library.

RNA and ATAC metrics are also not interchangeable:

| Metric | Modality | What an extreme value can suggest |
|---|---|---|
| `nCount_RNA` | RNA | very low RNA content or unusually high counts/doublet |
| `nFeature_RNA` | RNA | low complexity or unusually broad expression |
| `percent.mt` | RNA | stressed/damaged cell or ambient mitochondrial RNA; context dependent in nuclei |
| `nCount_ATAC` | ATAC | too few fragments/cut sites or unusually high accessibility counts |
| `TSS.enrichment` | ATAC | enrichment of insertions near promoters; low values suggest poor signal-to-background |
| `nucleosome_signal` | ATAC | mono-nucleosomal relative to nucleosome-free fragments; lower is generally cleaner |
| `frip_cr` | ATAC | fraction of ATAC fragments overlapping called peak regions |

# 4. Import and calculate metrics one sample at a time

The loop below is intentionally sample-by-sample. It saves a checkpoint and releases memory before proceeding to the next library.

Only near-empty barcodes are removed before `scDblFinder`. Cell Ranger's filtered matrix has already performed cell calling, and ordinary low-quality cells should remain present while doublets are modeled. Final QC happens in Step 2.

```{r calculate-qc, cache=FALSE}
threshold_suggestions <- vector("list", nrow(sample_sheet))
names(threshold_suggestions) <- sample_sheet$sample_id

for (row_index in seq_len(nrow(sample_sheet))) {
  sample_id <- sample_sheet$sample_id[row_index]
  message_step("Step 1 sample: ", sample_id)

  checkpoint_path <- file.path(
    OUTPUT_DIR,
    "01_qc",
    "checkpoints",
    paste0(sample_id, "_qc_metrics_unfiltered.rds")
  )
  metadata_path <- file.path(OUTPUT_DIR,
                             "01_qc",
                             "metadata",
                             paste0(sample_id, "_qc_metrics.csv"))

  if (REUSE_QC_METRIC_CHECKPOINTS && file.exists(checkpoint_path)) {
    message("Loading completed checkpoint: ", checkpoint_path)
    object <- readRDS(checkpoint_path)

    checkpoint_genome <- unique(GenomeInfoDb::genome(Signac::Annotation(object[["ATAC"]])))

    if (!identical(checkpoint_genome, GRCM39_GENOME_LABEL)) {
      stop(
        "Checkpoint for ",
        sample_id,
        " was created using genome annotation ",
        paste(checkpoint_genome, collapse = ", "),
        " rather than ",
        GRCM39_GENOME_LABEL,
        ". Rebuild this checkpoint.",
        call. = FALSE
      )
    }

  } else {
    counts <- extract_10x_modalities(sample_sheet$h5[row_index])
    rna_counts <- counts[["Gene Expression"]]
    atac_counts <- counts[["Peaks"]]

    if (!identical(colnames(rna_counts), colnames(atac_counts))) {
      stop("RNA and ATAC barcodes differ in ", sample_id)
    }

    object <- CreateSeuratObject(
      counts = rna_counts,
      assay = "RNA",
      project = sample_id,
      min.cells = 0,
      min.features = 0
    )
    object[["ATAC"]] <- CreateChromatinAssay(
      counts = atac_counts,
      sep = c(":", "-"),
      fragments = sample_sheet$fragments[row_index],
      annotation = gene_annotation,
      min.cells = 0,
      min.features = 0
    )

    object$sample_id <- sample_id
    object$replicate_id <- sample_sheet$replicate_id[row_index]
    object$genotype <- sample_sheet$genotype[row_index]
    object$diet <- sample_sheet$diet[row_index]

    object$percent.mt <- PercentageFeatureSet(object, assay = "RNA", pattern = "^(mt-|MT-)")
    object$percent.ribo <- PercentageFeatureSet(object, assay = "RNA", pattern = "^(Rpl|Rps)")
    object$log10_genes_per_umi <- log10(object$nFeature_RNA + 1) /
      log10(object$nCount_RNA + 1)
    object <- add_cellranger_metrics(object, sample_sheet$barcode_metrics[row_index])

    initial_keep <- (
      object$nCount_RNA >= ULTRA_LOW_RNA_UMI &
        object$nFeature_RNA >= ULTRA_LOW_RNA_FEATURES
    )
    object <- subset(object, cells = colnames(object)[initial_keep])

    DefaultAssay(object) <- "ATAC"
    object <- NucleosomeSignal(object)
    object <- TSSEnrichment(object, fast = TRUE)

    DefaultAssay(object) <- "RNA"
    object <- run_scdblfinder(object)

    saveRDS(object, checkpoint_path, compress = FALSE)
    rm(counts, rna_counts, atac_counts)
  }

  required_metadata <- c(
    "nCount_RNA",
    "nFeature_RNA",
    "percent.mt",
    "nCount_ATAC",
    "TSS.enrichment",
    "nucleosome_signal",
    "frip_cr",
    "scDblFinder.class"
  )
  absent_metadata <- setdiff(required_metadata, colnames(object[[]]))
  if (length(absent_metadata) > 0L) {
    stop(
      "Checkpoint for ",
      sample_id,
      " lacks: ",
      paste(absent_metadata, collapse = ", "),
      ". Set REUSE_QC_METRIC_CHECKPOINTS <- FALSE to rebuild it."
    )
  }

  metadata <- object[[]]
  metadata$barcode <- rownames(metadata)
  readr::write_csv(metadata, metadata_path)

  plot_prefix <- file.path(OUTPUT_DIR,
                           "01_qc",
                           "plots",
                           paste0(sample_id, "_before_filtering"))
  save_sample_qc_plots(metadata, sample_id, plot_prefix)

  threshold_suggestions[[sample_id]] <- make_qc_threshold_suggestion(object, sample_id)

  rm(object, metadata)
  invisible(gc())
}
```

# 5. Save starting suggestions—without approving them

Count cutoffs are suggested from broad hard limits plus four median absolute deviations on a log scale within each sample. This is a robust way to identify extreme tails, but it cannot decide whether a tail represents a rare, valid colon population. That requires plots and marker-aware biological judgment.

This chunk fills previously blank numeric cells in `config/qc_thresholds.csv`. It does **not** change `approved` to `TRUE` and does not overwrite numeric decisions that are already present.

```{r save-suggestions}
threshold_suggestions <- dplyr::bind_rows(threshold_suggestions)

suggestion_path <- file.path(
  OUTPUT_DIR,
  "01_qc",
  "qc_thresholds_suggested.csv"
)
readr::write_csv(threshold_suggestions, suggestion_path)

qc_decision_draft <- populate_qc_decision_draft(threshold_suggestions)
qc_decision_draft
```

```{r complete-step-one}
writeLines(
  paste("Step 1 completed:", format(Sys.time())),
  file.path(OUTPUT_DIR, "01_qc", "STEP_01_COMPLETE.txt")
)
```

# What to inspect before Step 2

For each sample, inspect the three PDFs under `01_qc/plots/`:

1. QC distributions: are there shoulders, long tails, or clear low-quality modes?
2. RNA complexity: is the count–feature relationship smooth, and do high-mitochondrial cells occupy a distinct region?
3. ATAC quality: do low-count barcodes also show low TSS enrichment or high nucleosome signal?

Do not select a threshold because one isolated cell looks unusual. Look for distributions and joint patterns.

# Interpretation checkpoint

1. Why were samples processed separately at this stage?
2. Why were only near-empty barcodes removed before doublet calling?
3. What biological or technical property does each QC metric measure?
4. Why are the automatically suggested thresholds still marked unapproved?

# References

See `docs/REFERENCES.md` for the official Signac QC workflow, 10x per-barcode metrics, and `scDblFinder` vignette.

