---
title: "Step 2 — Review Multiome QC and create post-QC objects"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The decision made in this notebook

This is the first formal manual checkpoint. We review RNA and ATAC distributions separately within each sample, record thresholds, and then make one joint keep/remove decision for every paired barcode.

The logical rule is:

```text
qc_pass = qc_pass_rna AND qc_pass_atac AND qc_pass_doublet
```

A barcode with excellent RNA but unusable ATAC is not suitable for a joint WNN analysis. The reverse is also true. The separate flags remain in the metadata so the reason for failure is auditable.

Use **4 CPU cores, 64 GiB memory, 8 hours, R 4.4.2, no GPU**. Only one full Seurat object is loaded at a time.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 6
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
})
```

# 2. Load the small QC tables first

Reviewing CSV metadata does not require loading the large RDS checkpoints. This lets us inspect distributions before spending memory on filtering.

```{r load-qc-tables}
step_one_marker <- file.path(OUTPUT_DIR, "01_qc", "STEP_01_COMPLETE.txt")
assert_file(step_one_marker, "Step 1 completion marker")

sample_sheet <- read_sample_sheet()

metadata_tables <- lapply(sample_sheet$sample_id, function(sample_id) {
  path <- file.path(
    OUTPUT_DIR,
    "01_qc",
    "metadata",
    paste0(sample_id, "_qc_metrics.csv")
  )
  assert_file(path, paste(sample_id, "QC metadata"))
  suppressMessages(readr::read_csv(path, show_col_types = FALSE))
})
names(metadata_tables) <- sample_sheet$sample_id
qc_metadata <- dplyr::bind_rows(metadata_tables)

suggestions <- suppressMessages(readr::read_csv(
  file.path(OUTPUT_DIR, "01_qc", "qc_thresholds_suggested.csv"),
  show_col_types = FALSE
))
decisions <- suppressMessages(readr::read_csv(
  QC_DECISION_PATH,
  show_col_types = FALSE
))

suggestions
decisions
```

# 3. Compare samples without pooling their thresholds

The following plots put samples side by side for diagnosis. This does **not** imply one threshold must be shared. Look for whether an entire library shifts, whether low RNA and low ATAC coincide, and whether a high-count tail is continuous or separated.

```{r compare-distributions}
metric_order <- c(
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "nCount_ATAC",
  "TSS.enrichment",
  "nucleosome_signal",
  "frip_cr"
)

qc_long <- tidyr::pivot_longer(
  qc_metadata,
  cols = dplyr::all_of(metric_order),
  names_to = "metric",
  values_to = "value"
)
qc_long$metric <- factor(qc_long$metric, levels = metric_order)

ggplot2::ggplot(qc_long, ggplot2::aes(sample_id, value, fill = sample_id)) +
  ggplot2::geom_violin(scale = "width", color = "grey25") +
  ggplot2::facet_wrap(~ metric, scales = "free_y", ncol = 4) +
  ggplot2::labs(
    title = "Pre-filter QC distributions by Cell Ranger library",
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_classic(base_size = 11) +
  ggplot2::theme(legend.position = "none")
```

```{r compare-rna}
ggplot2::ggplot(
  qc_metadata,
  ggplot2::aes(nCount_RNA, nFeature_RNA, color = percent.mt)
) +
  ggplot2::geom_point(alpha = 0.25, size = 0.35) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::scale_color_viridis_c() +
  ggplot2::facet_wrap(~ sample_id) +
  ggplot2::labs(title = "RNA complexity before filtering") +
  ggplot2::theme_classic(base_size = 11)
```

```{r compare-atac}
ggplot2::ggplot(
  qc_metadata,
  ggplot2::aes(nCount_ATAC, TSS.enrichment, color = nucleosome_signal)
) +
  ggplot2::geom_point(alpha = 0.25, size = 0.35) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_color_viridis_c() +
  ggplot2::facet_wrap(~ sample_id) +
  ggplot2::labs(title = "ATAC quality before filtering") +
  ggplot2::theme_classic(base_size = 11)
```

# 3A. Inspect doublet calls relative to library size
```{r inspect doublets}
doublet_summary <- qc_metadata |>
  dplyr::count(
    sample_id,
    scDblFinder.class,
    name = "n"
  ) |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    percent = 100 * n / sum(n)
  ) |>
  dplyr::ungroup()

doublet_summary

ggplot2::ggplot(
  qc_metadata,
  ggplot2::aes(
    x = nCount_RNA,
    y = nCount_ATAC,
    color = scDblFinder.class
  )
) +
  ggplot2::geom_point(
    alpha = 0.35,
    size = 0.45
  ) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::facet_wrap(~ sample_id) +
  ggplot2::labs(
    title = "RNA and ATAC library size by scDblFinder call",
    x = "RNA UMIs",
    y = "ATAC counts",
    color = "scDblFinder"
  ) +
  ggplot2::theme_classic(base_size = 11)
```
# 3B. Lower limit summary for RNA and ATAC
```{r}
lower_tail_summary <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    metadata |>
      dplyr::mutate(
        RNA_bin = cut(
          nCount_RNA,
          breaks = c(-Inf, 200, 300, 400, 500, 750, 1000, Inf),
          labels = c(
            "<200",
            "200-299",
            "300-399",
            "400-499",
            "500-749",
            "750-999",
            ">=1000"
          ),
          right = FALSE
        ),
        ATAC_bin = cut(
          nCount_ATAC,
          breaks = c(-Inf, 500, 750, 1000, 1500, 2000, Inf),
          labels = c(
            "<500",
            "500-749",
            "750-999",
            "1000-1499",
            "1500-1999",
            ">=2000"
          ),
          right = FALSE
        )
      ) |>
      dplyr::group_by(RNA_bin) |>
      dplyr::summarise(
        sample_id = .env$sample_id,
        modality = "RNA",
        n = dplyr::n(),
        median_features = median(nFeature_RNA),
        median_mt = median(percent.mt),
        singlet_pct = 100 * mean(scDblFinder.class == "singlet"),
        .groups = "drop"
      )
  }
) |>
  dplyr::bind_rows()

lower_tail_summary

lower_atac_summary <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    metadata |>
      dplyr::mutate(
        ATAC_bin = cut(
          nCount_ATAC,
          breaks = c(-Inf, 500, 750, 1000, 1500, 2000, Inf),
          labels = c(
            "<500",
            "500-749",
            "750-999",
            "1000-1499",
            "1500-1999",
            ">=2000"
          ),
          right = FALSE
        )
      ) |>
      dplyr::group_by(ATAC_bin) |>
      dplyr::summarise(
        sample_id = .env$sample_id,
        n = dplyr::n(),
        median_TSS = median(TSS.enrichment, na.rm = TRUE),
        median_FRiP = median(frip_cr, na.rm = TRUE),
        median_nucleosome = median(nucleosome_signal, na.rm = TRUE),
        singlet_pct = 100 * mean(scDblFinder.class == "singlet"),
        .groups = "drop"
      )
  }
) |>
  dplyr::bind_rows()

lower_atac_summary

# Percentage passing independent QC metrics within RNA-count bins

rna_bin_pass_summary <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    thresholds <- decisions_preview[
      decisions_preview$sample_id == sample_id,
      ,
      drop = FALSE
    ]

    metadata |>
      dplyr::mutate(
        RNA_bin = cut(
          nCount_RNA,
          breaks = c(-Inf, 200, 300, 400, 500, 750, 1000, Inf),
          labels = c(
            "<200",
            "200-299",
            "300-399",
            "400-499",
            "500-749",
            "750-999",
            ">=1000"
          ),
          right = FALSE
        )
      ) |>
      dplyr::group_by(RNA_bin) |>
      dplyr::summarise(
        sample_id = .env$sample_id,
        n = dplyr::n(),

        feature_pass_pct =
          100 * mean(
            nFeature_RNA >= thresholds$nFeature_RNA_min,
            na.rm = TRUE
          ),

        mt_pass_pct =
          100 * mean(
            percent.mt <= thresholds$percent_mt_max,
            na.rm = TRUE
          ),

        TSS_pass_pct =
          100 * mean(
            TSS.enrichment >= thresholds$TSS_enrichment_min,
            na.rm = TRUE
          ),

        FRiP_pass_pct =
          100 * mean(
            frip_cr >= thresholds$frip_min,
            na.rm = TRUE
          ),

        singlet_pct =
          100 * mean(
            scDblFinder.class == "singlet",
            na.rm = TRUE
          ),

        .groups = "drop"
      )
  }
) |>
  dplyr::bind_rows()

rna_bin_pass_summary

# Percentage passing independent QC metrics within ATAC-count bins

atac_bin_pass_summary <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    thresholds <- decisions_preview[
      decisions_preview$sample_id == sample_id,
      ,
      drop = FALSE
    ]

    metadata |>
      dplyr::mutate(
        ATAC_bin = cut(
          nCount_ATAC,
          breaks = c(-Inf, 500, 750, 1000, 1500, 2000, Inf),
          labels = c(
            "<500",
            "500-749",
            "750-999",
            "1000-1499",
            "1500-1999",
            ">=2000"
          ),
          right = FALSE
        )
      ) |>
      dplyr::group_by(ATAC_bin) |>
      dplyr::summarise(
        sample_id = .env$sample_id,
        n = dplyr::n(),

        RNA_feature_pass_pct =
          100 * mean(
            nFeature_RNA >= thresholds$nFeature_RNA_min,
            na.rm = TRUE
          ),

        mt_pass_pct =
          100 * mean(
            percent.mt <= thresholds$percent_mt_max,
            na.rm = TRUE
          ),

        TSS_pass_pct =
          100 * mean(
            TSS.enrichment >= thresholds$TSS_enrichment_min,
            na.rm = TRUE
          ),

        FRiP_pass_pct =
          100 * mean(
            frip_cr >= thresholds$frip_min,
            na.rm = TRUE
          ),

        singlet_pct =
          100 * mean(
            scDblFinder.class == "singlet",
            na.rm = TRUE
          ),

        .groups = "drop"
      )
  }
) |>
  dplyr::bind_rows()

atac_bin_pass_summary
```
# 4. Record the decision

Open `config/qc_thresholds.csv` in RStudio. Step 1 populated blank numeric cells with starting suggestions. For every sample:

1. compare the threshold to its full distribution;
2. change values when biology or a clear quality mode supports the change;
3. enter `reviewer`, `review_date`, and a short `decision_notes` explanation;
4. set `approved` to `TRUE` only after review.

The file is deliberately version-controlled. A future reader should be able to see not only the final cells, but the decision that produced them.

```{r open-decision-file, eval=FALSE}
file.edit(QC_DECISION_PATH)
```

Questions to discuss while reviewing:

- Would the lower RNA cutoff remove a coherent low-RNA cell population or mostly a diffuse poor-quality tail?
- Do high RNA counts and high ATAC counts occur together, consistent with possible doublets?
- Does low TSS enrichment coincide with low ATAC counts?
- Are thresholds broadly comparable between samples, and if not, is the difference technically plausible?
- Is the retained fraction unexpectedly different between CON and HFD?

# 5. Preview the recorded thresholds on each sample

Rerun this chunk after saving the CSV. It produces threshold-overlay PDFs, but it does not yet filter cells.

```{r preview-thresholds}
decisions_preview <- suppressMessages(readr::read_csv(
  QC_DECISION_PATH,
  show_col_types = FALSE
))

for (sample_id in sample_sheet$sample_id) {
  sample_metadata <- metadata_tables[[sample_id]]
  threshold_row <- decisions_preview[
    decisions_preview$sample_id == sample_id,
    ,
    drop = FALSE
  ]

  save_sample_qc_plots(
    sample_metadata,
    sample_id,
    file.path(
      OUTPUT_DIR,
      "01_qc",
      "plots",
      paste0(sample_id, "_threshold_preview")
    ),
    thresholds = threshold_row
  )
}

candidate_retention <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    thresholds <- decisions_preview[
      decisions_preview$sample_id == sample_id,
      ,
      drop = FALSE
    ]

    pass_rna <- with(
      metadata,
      nCount_RNA >= thresholds$nCount_RNA_min &
        nCount_RNA <= thresholds$nCount_RNA_max &
        nFeature_RNA >= thresholds$nFeature_RNA_min &
        nFeature_RNA <= thresholds$nFeature_RNA_max &
        percent.mt <= thresholds$percent_mt_max
    )

    pass_atac <- with(
      metadata,
      nCount_ATAC >= thresholds$nCount_ATAC_min &
        nCount_ATAC <= thresholds$nCount_ATAC_max &
        TSS.enrichment >= thresholds$TSS_enrichment_min &
        nucleosome_signal <= thresholds$nucleosome_signal_max &
        frip_cr >= thresholds$frip_min
    )

    pass_doublet <- metadata$scDblFinder.class == "singlet"

    pass_joint <- pass_rna & pass_atac & pass_doublet

    data.frame(
      sample_id = sample_id,
      total_barcodes = nrow(metadata),
      pass_rna = sum(pass_rna, na.rm = TRUE),
      pass_atac = sum(pass_atac, na.rm = TRUE),
      pass_doublet = sum(pass_doublet, na.rm = TRUE),
      pass_joint = sum(pass_joint, na.rm = TRUE),
      percent_joint_retained =
        100 * mean(pass_joint, na.rm = TRUE),
      percent_doublets =
        100 * mean(!pass_doublet, na.rm = TRUE)
    )
  }
) |>
  dplyr::bind_rows()

candidate_retention

criterion_failure_summary <- lapply(
  sample_sheet$sample_id,
  function(sample_id) {

    metadata <- metadata_tables[[sample_id]]

    thresholds <- decisions_preview[
      decisions_preview$sample_id == sample_id,
      ,
      drop = FALSE
    ]

    criterion_pass <- data.frame(
      RNA_count_min =
        metadata$nCount_RNA >= thresholds$nCount_RNA_min,

      RNA_count_max =
        metadata$nCount_RNA <= thresholds$nCount_RNA_max,

      RNA_features_min =
        metadata$nFeature_RNA >= thresholds$nFeature_RNA_min,

      RNA_features_max =
        metadata$nFeature_RNA <= thresholds$nFeature_RNA_max,

      mitochondrial =
        metadata$percent.mt <= thresholds$percent_mt_max,

      ATAC_count_min =
        metadata$nCount_ATAC >= thresholds$nCount_ATAC_min,

      ATAC_count_max =
        metadata$nCount_ATAC <= thresholds$nCount_ATAC_max,

      TSS =
        metadata$TSS.enrichment >= thresholds$TSS_enrichment_min,

      nucleosome =
        metadata$nucleosome_signal <= thresholds$nucleosome_signal_max,

      FRiP =
        metadata$frip_cr >= thresholds$frip_min,

      singlet =
        metadata$scDblFinder.class == "singlet"
    )

    data.frame(
      sample_id = sample_id,
      criterion = names(criterion_pass),
      n_failed = vapply(
        criterion_pass,
        function(x) sum(!x, na.rm = TRUE),
        integer(1)
      ),
      percent_failed = vapply(
        criterion_pass,
        function(x) 100 * mean(!x, na.rm = TRUE),
        numeric(1)
      )
    )
  }
) |>
  dplyr::bind_rows()

criterion_failure_summary
```

# 6. Approval gate

This chunk stops on purpose if a threshold is missing or a sample is not approved. That is not a software failure; it prevents an unreviewed filtering decision from propagating downstream.

```{r approval-gate}
qc_decisions <- read_and_validate_qc_decisions(sample_sheet)
qc_decisions
```

# 7. Apply separate modality flags and one joint decision

This loop never overwrites the Step 1 checkpoints. It writes new post-QC RDS objects to `02_objects/`.

```{r filter-and-save, cache=FALSE}
retention_summaries <- vector("list", nrow(sample_sheet))
names(retention_summaries) <- sample_sheet$sample_id

for (row_index in seq_len(nrow(sample_sheet))) {
  sample_id <- sample_sheet$sample_id[row_index]
  message_step("Step 2 sample: ", sample_id)

  input_checkpoint <- file.path(
    OUTPUT_DIR,
    "01_qc",
    "checkpoints",
    paste0(sample_id, "_qc_metrics_unfiltered.rds")
  )
  output_checkpoint <- file.path(
    OUTPUT_DIR,
    "02_objects",
    paste0(sample_id, "_post_QC.rds")
  )
  assert_file(input_checkpoint, paste(sample_id, "unfiltered QC checkpoint"))

  object <- readRDS(input_checkpoint)
  checkpoint_genome <- unique(
  GenomeInfoDb::genome(
    Signac::Annotation(object[["ATAC"]])
  )
)

if (!identical(
  checkpoint_genome,
  GRCM39_GENOME_LABEL
)) {
  stop(
    "Checkpoint for ", sample_id,
    " uses genome annotation ",
    paste(checkpoint_genome, collapse = ", "),
    " rather than ",
    GRCM39_GENOME_LABEL,
    ".",
    call. = FALSE
  )
}
  threshold_row <- qc_decisions[
    qc_decisions$sample_id == sample_id,
    ,
    drop = FALSE
  ]

  object <- add_joint_qc_flags(object, threshold_row)
  retention_summaries[[sample_id]] <- summarize_qc_decision(object, sample_id)

  metadata_with_decision <- object[[]]
  metadata_with_decision$barcode <- rownames(metadata_with_decision)
  readr::write_csv(
    metadata_with_decision,
    file.path(
      OUTPUT_DIR,
      "01_qc",
      "metadata",
      paste0(sample_id, "_qc_decision_all_barcodes.csv")
    )
  )

  passing_cells <- rownames(metadata_with_decision)[metadata_with_decision$qc_pass]
  filtered_object <- subset(object, cells = passing_cells)
  DefaultAssay(filtered_object) <- "RNA"
  saveRDS(filtered_object, output_checkpoint, compress = FALSE)

  filtered_metadata <- filtered_object[[]]
  filtered_metadata$barcode <- rownames(filtered_metadata)
  readr::write_csv(
    filtered_metadata,
    file.path(
      OUTPUT_DIR,
      "01_qc",
      "metadata",
      paste0(sample_id, "_post_QC_metadata.csv")
    )
  )

  save_sample_qc_plots(
    filtered_metadata,
    sample_id,
    file.path(
      OUTPUT_DIR,
      "01_qc",
      "plots",
      paste0(sample_id, "_after_filtering")
    )
  )

  rm(object, filtered_object, metadata_with_decision, filtered_metadata)
  invisible(gc())
}

retention_summary <- dplyr::bind_rows(retention_summaries)
retention_summary
```

# 8. Save the audit trail and completion marker

```{r save-step-two}
readr::write_csv(
  retention_summary,
  file.path(OUTPUT_DIR, "01_qc", "qc_cell_retention_summary.csv")
)
readr::write_csv(
  qc_decisions,
  file.path(OUTPUT_DIR, "01_qc", "qc_thresholds_used.csv")
)

writeLines(
  paste("Step 2 completed:", format(Sys.time())),
  file.path(OUTPUT_DIR, "02_objects", "STEP_02_COMPLETE.txt")
)
```

# 9. Final checks before common peaks

```{r verify-post-qc-objects}
post_qc_paths <- file.path(
  OUTPUT_DIR,
  "02_objects",
  paste0(sample_sheet$sample_id, "_post_QC.rds")
)

data.frame(
  sample_id = sample_sheet$sample_id,
  checkpoint = post_qc_paths,
  exists = file.exists(post_qc_paths),
  size_gb = unname(file.info(post_qc_paths)$size / 1024^3)
)
```

Before proceeding, inspect `qc_cell_retention_summary.csv` and the after-filtering plots. A very different retained fraction between CON and HFD is not automatically wrong, but it must be understood before biological comparisons.

# Interpretation checkpoint

1. Which metrics were evaluated separately for RNA and ATAC?
2. At what level was the final decision made?
3. How would you find cells that failed only ATAC QC?
4. Why are the unfiltered QC checkpoints preserved?
5. Which parts of the threshold decision should be committed to GitHub?

# References

See `docs/REFERENCES.md` for the official Signac QC and Multiome analysis documentation and the `scDblFinder` vignette.

