---
title: "Step 3 — Build a common ATAC peak set and requantify post-QC cells"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

The CON and HFD libraries were processed independently by Cell Ranger ARC,
so their original ATAC peak coordinates are not guaranteed to be identical.

Before joint ATAC analysis, both samples must therefore be represented using
the same genomic features.

This notebook will:

1. load the independently called Cell Ranger peak sets;
2. combine overlapping peaks into one common GRCm39 peak set;
3. quantify fragments from each sample over exactly those common regions;
4. replace the original sample-specific ATAC assay with the common ATAC assay;
5. preserve the RNA assay, cell metadata, gene annotation, fragment links, and
   original ATAC QC values;
6. verify that every sample contains the same ATAC features in the same order;
7. save new common-feature post-QC objects for downstream dimensional reduction.

The common peak set defines the ATAC feature space used by all later analyses.

A practical SOL allocation for this step is **4–8 CPU cores, 96 GiB memory,
12 hours, R 4.4.2, no GPU**. The helper configuration respects the CPU count
available in the active SLURM/RStudio session.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 6
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(GenomicRanges)
})

configure_parallel_plan()

cat("Configured future workers:", future::nbrOfWorkers(), "\n")
```

# 2. Confirm the post-QC inputs

Step 3 must never fall back to the unfiltered objects. The Step 2 completion
marker and both post-QC objects are therefore required. The Cell Ranger peak
and fragment files must also still be available because Step 3 returns to the
fragment files for requantification.

```{r confirm-step-two}
step_two_marker <- file.path(
  OUTPUT_DIR,
  "02_objects",
  "STEP_02_COMPLETE.txt"
)

assert_file(step_two_marker, "Step 2 completion marker")

sample_sheet <- read_sample_sheet()

post_qc_paths <- file.path(
  OUTPUT_DIR,
  "02_objects",
  paste0(sample_sheet$sample_id, "_post_QC.rds")
)

for (i in seq_along(post_qc_paths)) {
  assert_file(
    post_qc_paths[i],
    paste(sample_sheet$sample_id[i], "post-QC object")
  )

  assert_file(
    sample_sheet$peaks[i],
    paste(sample_sheet$sample_id[i], "Cell Ranger ATAC peak BED")
  )

  assert_file(
    sample_sheet$fragments[i],
    paste(sample_sheet$sample_id[i], "Cell Ranger ATAC fragments")
  )
}

input_object_summary <- data.frame(
  sample_id = sample_sheet$sample_id,
  post_qc_object = post_qc_paths,
  size_gb = file.info(post_qc_paths)$size / 1024^3
)

input_object_summary
```

# 3. Build one common GRCm39 ATAC peak set

Each Cell Ranger ARC library has its own independently called peak set. The
coordinates therefore do not necessarily match between samples.

We read each `atac_peaks.bed`, convert the first three columns into `GRanges`,
combine all samples, and merge overlapping or directly adjacent intervals with
`GenomicRanges::reduce()`.

## Coordinate convention used in this pipeline

Cell Ranger writes `atac_peaks.bed` as a BED-style file. However, the ATAC
features already stored in the imported Signac `ChromatinAssay` use the same
numeric start coordinates that appear in the Cell Ranger peak file. During
pipeline validation, adding an additional +1 conversion shifted every peak
relative to the existing assay.

Therefore, this notebook deliberately **preserves the numeric peak coordinates
exactly as represented by Cell Ranger/Signac** when creating the common feature
space. No additional start-coordinate offset is applied here.

This keeps the common peaks consistent with the original ATAC assay features
that were used during QC.

```{r build-common-peaks}
COMMON_PEAK_DIR <- file.path(
  OUTPUT_DIR,
  "03_common_peaks"
)

COMMON_OBJECT_DIR <- file.path(
  OUTPUT_DIR,
  "03_objects"
)

dir.create(
  COMMON_PEAK_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  COMMON_OBJECT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

annotation_path <- file.path(
  OUTPUT_DIR,
  "00_run_info",
  "GRCm39_GENCODE_vM33_gene_annotation.rds"
)

assert_file(
  annotation_path,
  "saved GRCm39 GENCODE vM33 annotation"
)

gene_annotation <- readRDS(annotation_path)

if (!identical(
  unique(GenomeInfoDb::genome(gene_annotation)),
  GRCM39_GENOME_LABEL
)) {
  stop(
    "The saved annotation is not labeled as ",
    GRCM39_GENOME_LABEL,
    ".",
    call. = FALSE
  )
}

read_cellranger_peak_bed <- function(path) {

  assert_file(path, "Cell Ranger ATAC peak BED")

  peaks_df <- utils::read.table(
    path,
    header = FALSE,
    sep = "\t",
    comment.char = "#",
    stringsAsFactors = FALSE
  )

  if (ncol(peaks_df) < 3L) {
    stop(
      "Peak BED must contain at least three columns: ",
      path,
      call. = FALSE
    )
  }

  peaks_df <- peaks_df[, 1:3]
  colnames(peaks_df) <- c(
    "seqnames",
    "start",
    "end"
  )

  peaks_gr <- GenomicRanges::makeGRangesFromDataFrame(
    peaks_df,
    seqnames.field = "seqnames",
    start.field = "start",
    end.field = "end",
    keep.extra.columns = FALSE
  )

  GenomeInfoDb::genome(peaks_gr) <- GRCM39_GENOME_LABEL

  peaks_gr
}

sample_peak_sets <- lapply(
  sample_sheet$peaks,
  read_cellranger_peak_bed
)

names(sample_peak_sets) <- sample_sheet$sample_id

input_peak_counts <- data.frame(
  sample_id = names(sample_peak_sets),
  input_peaks = vapply(
    sample_peak_sets,
    length,
    integer(1)
  )
)

input_peak_counts
```

## Merge overlapping peaks

`reduce()` combines overlapping or directly adjacent intervals into one
non-overlapping region. This means every sample can subsequently be quantified
over exactly the same genomic feature definitions.

Very short or extremely broad merged intervals are removed using broad
technical bounds. These bounds are intended to exclude pathological regions,
not to select peaks based on biological condition.

```{r reduce-common-peaks}
all_input_peaks <- do.call(
  c,
  unname(sample_peak_sets)
)

common_peaks_unfiltered <- GenomicRanges::reduce(
  all_input_peaks,
  ignore.strand = TRUE
)

common_peak_widths <- GenomicRanges::width(
  common_peaks_unfiltered
)

common_peaks <- common_peaks_unfiltered[
  common_peak_widths > 20 &
    common_peak_widths < 10000
]

GenomeInfoDb::genome(common_peaks) <- GRCM39_GENOME_LABEL

common_peaks <- sort(common_peaks)

if (length(common_peaks) == 0L) {
  stop(
    "No common peaks remained after width filtering.",
    call. = FALSE
  )
}

if (anyDuplicated(common_peaks) > 0L) {
  stop(
    "Duplicated genomic ranges remain in common peak set.",
    call. = FALSE
  )
}

common_peak_summary <- data.frame(
  total_input_peaks = length(all_input_peaks),
  reduced_peaks_before_width_filter =
    length(common_peaks_unfiltered),
  common_peaks_after_width_filter =
    length(common_peaks),
  median_common_peak_width =
    stats::median(GenomicRanges::width(common_peaks)),
  min_common_peak_width =
    min(GenomicRanges::width(common_peaks)),
  max_common_peak_width =
    max(GenomicRanges::width(common_peaks))
)

common_peak_summary
```

For the current Ppar-CON-1 and Ppar-HFD-1 libraries, the validated common
feature space contains **130,476 peaks**. The code above remains data-driven so
future datasets are not forced to reproduce that exact number.

## Validate the common peak chromosomes

Every chromosome represented by the common peak set must also be represented
in the saved GRCm39 annotation.

```{r validate-common-peaks}
peak_chromosomes <- GenomeInfoDb::seqlevelsInUse(
  common_peaks
)

annotation_chromosomes <- GenomeInfoDb::seqlevels(
  gene_annotation
)

unexpected_peak_chromosomes <- setdiff(
  peak_chromosomes,
  annotation_chromosomes
)

unexpected_peak_chromosomes

if (length(unexpected_peak_chromosomes) > 0L) {
  stop(
    "Common peaks contain chromosome(s) absent from the GRCm39 annotation: ",
    paste(
      unexpected_peak_chromosomes,
      collapse = ", "
    ),
    call. = FALSE
  )
}
```

## Save the canonical common peak set

The RDS file is the canonical peak definition used by downstream R notebooks.
The TSV file preserves the same numeric coordinates used by the Signac/GRanges feature space for inspection and auditing. It is intentionally not exported as a standard BED file because BED uses 0-based start coordinates.

```{r save-common-peaks}
common_peak_rds <- file.path(
  COMMON_PEAK_DIR,
  "GRCm39_common_ATAC_peaks.rds"
)

saveRDS(
  common_peaks,
  common_peak_rds,
  compress = FALSE
)

common_peak_bed <- data.frame(
  chrom = as.character(
    GenomeInfoDb::seqnames(common_peaks)
  ),

  # Preserve the same numeric start coordinate used by the imported
  # Cell Ranger/Signac peak features. Do not subtract 1 again here.
  start = GenomicRanges::start(common_peaks),

  end = GenomicRanges::end(common_peaks)
)

readr::write_tsv(
  common_peak_bed,
  file.path(
    COMMON_PEAK_DIR,
    "GRCm39_common_ATAC_peaks_coordinates.tsv"
  ),
  col_names = FALSE
)

readr::write_csv(
  input_peak_counts,
  file.path(
    COMMON_PEAK_DIR,
    "input_peak_counts.csv"
  )
)

readr::write_csv(
  common_peak_summary,
  file.path(
    COMMON_PEAK_DIR,
    "common_peak_summary.csv"
  )
)
```

# 4. Requantify post-QC cells over the common peaks

The original ATAC count matrices are based on sample-specific peak sets. They
cannot simply be row-bound or merged because their genomic features differ.

For each post-QC object, `FeatureMatrix()` returns to the indexed fragment file
and counts fragments overlapping every common peak for the retained cells.

## Important: `keep_all_features = FALSE`

This pipeline deliberately uses `keep_all_features = FALSE`.

During validation with Signac 1.17.1 on SOL, setting
`keep_all_features = TRUE` produced a severely undercounted matrix: only about
2,000 of ~130,000 peaks carried signal and median per-cell counts fell to less
than 1% of the original ATAC matrix. Repeating the test sequentially produced
the same result, ruling out the `future` parallel backend as the cause.

With `keep_all_features = FALSE`, requantification recovered the expected
fragment-scale signal. For a 100-cell validation subset, the median
requantified count was approximately 53% of the original Cell Ranger ATAC
count. That systematic difference is expected because the fragment-derived
matrix and the original Cell Ranger peak matrix do not use exactly the same
counting unit.

To keep this behavior safe and reproducible, the code below uses
`keep_all_features = FALSE` **and then requires `FeatureMatrix()` to return
exactly the full common feature set**. If a future dataset genuinely drops a
feature, the notebook stops rather than silently continuing with unequal ATAC
matrices.

```{r requantify-common-peaks, cache=FALSE}
requantification_summary <- vector(
  "list",
  nrow(sample_sheet)
)

names(requantification_summary) <-
  sample_sheet$sample_id

common_feature_names <- paste0(
  as.character(GenomeInfoDb::seqnames(common_peaks)),
  ":",
  GenomicRanges::start(common_peaks),
  "-",
  GenomicRanges::end(common_peaks)
)

if (anyDuplicated(common_feature_names) > 0L) {
  stop(
    "Common peak feature names are not unique.",
    call. = FALSE
  )
}

for (row_index in seq_len(nrow(sample_sheet))) {

  sample_id <- sample_sheet$sample_id[row_index]

  message_step(
    "Step 3 requantification: ",
    sample_id
  )

  input_object_path <- post_qc_paths[row_index]

  output_object_path <- file.path(
    COMMON_OBJECT_DIR,
    paste0(
      sample_id,
      "_post_QC_common_ATAC.rds"
    )
  )

  object <- readRDS(
    input_object_path
  )

  if (!"ATAC" %in% names(object)) {
    stop(
      sample_id,
      " does not contain an ATAC assay.",
      call. = FALSE
    )
  }

  old_annotation <- Signac::Annotation(
    object[["ATAC"]]
  )

  old_genome <- unique(
    GenomeInfoDb::genome(
      old_annotation
    )
  )

  if (!identical(
    old_genome,
    GRCM39_GENOME_LABEL
  )) {
    stop(
      sample_id,
      " post-QC object is not using ",
      GRCM39_GENOME_LABEL,
      ".",
      call. = FALSE
    )
  }

  # Preserve the ATAC values used during QC before replacing the assay.
  object$nCount_ATAC_original_peaks <-
    object$nCount_ATAC

  object$nFeature_ATAC_original_peaks <-
    object$nFeature_ATAC

  fragment_objects <- Signac::Fragments(
    object[["ATAC"]]
  )

  if (length(fragment_objects) == 0L) {
    stop(
      "No Fragment object is stored in ",
      sample_id,
      ".",
      call. = FALSE
    )
  }

  common_counts <- Signac::FeatureMatrix(
    fragments = fragment_objects,
    features = common_peaks,
    cells = colnames(object),
    keep_all_features = FALSE,
    verbose = TRUE
  )

  if (nrow(common_counts) != length(common_peaks)) {
    stop(
      sample_id,
      ": FeatureMatrix returned ",
      nrow(common_counts),
      " rows but ",
      length(common_peaks),
      " common peaks were expected. ",
      "Do not continue until the missing features are explained.",
      call. = FALSE
    )
  }

  if (!setequal(
    colnames(common_counts),
    colnames(object)
  )) {
    stop(
      sample_id,
      ": requantified ATAC cells do not match the post-QC object.",
      call. = FALSE
    )
  }

  # Enforce the same cell order as the Seurat object.
  common_counts <- common_counts[
    ,
    colnames(object),
    drop = FALSE
  ]

  rownames(common_counts) <-
    common_feature_names

  new_atac_assay <- Signac::CreateChromatinAssay(
    counts = common_counts,
    ranges = common_peaks,
    fragments = fragment_objects,
    annotation = gene_annotation,
    min.cells = 0,
    min.features = 0,
    validate.fragments = TRUE
  )

  object[["ATAC"]] <- new_atac_assay

  # Store common-feature depth explicitly rather than confusing it
  # with the original QC peak-space measurements.
  object$nCount_ATAC_common <-
    Matrix::colSums(common_counts)

  object$nFeature_ATAC_common <-
    Matrix::colSums(common_counts > 0)

  DefaultAssay(object) <- "RNA"

  requantification_summary[[sample_id]] <-
    data.frame(
      sample_id = sample_id,
      retained_cells = ncol(object),
      common_peaks = nrow(common_counts),
      peaks_detected_in_sample =
        sum(
          Matrix::rowSums(common_counts) > 0
        ),
      median_original_ATAC_counts =
        stats::median(object$nCount_ATAC_original_peaks),
      median_common_fragment_counts =
        stats::median(object$nCount_ATAC_common),
      total_common_peak_counts =
        sum(common_counts),
      output_object = output_object_path
    )

  saveRDS(
    object,
    output_object_path,
    compress = FALSE
  )

  rm(
    object,
    common_counts,
    new_atac_assay,
    fragment_objects,
    old_annotation
  )

  invisible(gc())
}

requantification_summary <-
  dplyr::bind_rows(
    requantification_summary
  )

requantification_summary
```

For the validated Ppar dataset, the successful requantification produced
approximately 11.6 million CON fragment counts and 9.1 million HFD fragment
counts across the 130,476 common features. These values are recorded here as a
known successful run, not as hard-coded acceptance thresholds for future data.

# 5. Verify identical ATAC feature spaces

Before downstream dimensional reduction, every common-feature ATAC assay must
contain:

- the same number of peaks;
- exactly the same peak coordinates;
- the same feature order;
- the GRCm39 gene annotation;
- the expected post-QC cells.

The coordinate comparison is performed using a simple `chr:start-end` key so
that harmless differences in `Seqinfo` metadata cannot obscure a genuine
feature mismatch.

```{r verify-common-feature-objects}
common_object_paths <- file.path(
  COMMON_OBJECT_DIR,
  paste0(
    sample_sheet$sample_id,
    "_post_QC_common_ATAC.rds"
  )
)

range_key <- function(x) {
  paste0(
    as.character(GenomeInfoDb::seqnames(x)),
    ":",
    GenomicRanges::start(x),
    "-",
    GenomicRanges::end(x)
  )
}

canonical_peak_keys <- range_key(common_peaks)

feature_checks <- vector(
  "list",
  length(common_object_paths)
)

reference_feature_keys <- NULL

for (i in seq_along(common_object_paths)) {

  sample_id <- sample_sheet$sample_id[i]

  assert_file(
    common_object_paths[i],
    paste(
      sample_id,
      "common-ATAC object"
    )
  )

  object <- readRDS(
    common_object_paths[i]
  )

  current_ranges <- Signac::granges(
    object[["ATAC"]]
  )

  current_feature_keys <- range_key(
    current_ranges
  )

  annotation_genome <- unique(
    GenomeInfoDb::genome(
      Signac::Annotation(
        object[["ATAC"]]
      )
    )
  )

  if (!identical(
    annotation_genome,
    GRCM39_GENOME_LABEL
  )) {
    stop(
      sample_id,
      " common ATAC assay is not annotated as ",
      GRCM39_GENOME_LABEL,
      ".",
      call. = FALSE
    )
  }

  if (!identical(
    current_feature_keys,
    canonical_peak_keys
  )) {
    stop(
      sample_id,
      ": ATAC feature coordinates or feature order differ from the ",
      "canonical common peak set.",
      call. = FALSE
    )
  }

  if (is.null(reference_feature_keys)) {
    reference_feature_keys <- current_feature_keys
  } else if (!identical(
    current_feature_keys,
    reference_feature_keys
  )) {
    stop(
      "ATAC feature coordinates or order differ between common-feature objects.",
      call. = FALSE
    )
  }

  expected_cells <- requantification_summary$retained_cells[
    match(
      sample_id,
      requantification_summary$sample_id
    )
  ]

  if (length(expected_cells) != 1L ||
      is.na(expected_cells) ||
      ncol(object) != expected_cells) {
    stop(
      sample_id,
      ": final cell count does not match the requantification audit.",
      call. = FALSE
    )
  }

  atac_counts <- SeuratObject::LayerData(
    object,
    assay = "ATAC",
    layer = "counts"
  )

  feature_checks[[i]] <- data.frame(
    sample_id = sample_id,
    cells = ncol(object),
    ATAC_features = nrow(object[["ATAC"]]),
    detected_ATAC_features =
      sum(Matrix::rowSums(atac_counts) > 0),
    total_common_peak_counts =
      sum(atac_counts),
    coordinates_match_canonical = TRUE,
    annotation_genome = paste(
      annotation_genome,
      collapse = ","
    ),
    size_gb =
      file.info(common_object_paths[i])$size / 1024^3
  )

  rm(
    object,
    current_ranges,
    current_feature_keys,
    atac_counts
  )

  invisible(gc())
}

feature_checks <-
  dplyr::bind_rows(
    feature_checks
  )

feature_checks

if (!all(feature_checks$coordinates_match_canonical)) {
  stop(
    "At least one sample failed the final common-feature coordinate check.",
    call. = FALSE
  )
}
```

# 6. Save the Step 3 audit trail

The audit files record both how the common feature space was constructed and
how each post-QC sample was requantified against it.

```{r save-step-three}
readr::write_csv(
  common_peak_summary,
  file.path(
    COMMON_PEAK_DIR,
    "common_peak_summary.csv"
  )
)

readr::write_csv(
  requantification_summary,
  file.path(
    COMMON_PEAK_DIR,
    "requantification_summary.csv"
  )
)

readr::write_csv(
  feature_checks,
  file.path(
    COMMON_PEAK_DIR,
    "common_ATAC_object_validation.csv"
  )
)

completion_lines <- c(
  paste(
    "Step 3 completed:",
    format(Sys.time())
  ),
  paste(
    "Common ATAC features:",
    length(common_peaks)
  ),
  paste(
    "Samples validated:",
    nrow(feature_checks)
  ),
  "Feature coordinates identical to the canonical common peak set: TRUE"
)

writeLines(
  completion_lines,
  file.path(
    COMMON_OBJECT_DIR,
    "STEP_03_COMPLETE.txt"
  )
)

completion_lines
```

# Interpretation checkpoint

Before proceeding, be able to answer:

1. Why could the original CON and HFD ATAC matrices not simply be merged?
2. What does `reduce()` do to overlapping or adjacent peaks?
3. Why do we return to the fragment files after defining the common peaks?
4. What does one row of the new ATAC matrix represent?
5. Why must every sample use exactly the same common peak coordinates and row
   order?
6. Why were the original QC ATAC counts preserved separately from the new
   common-feature ATAC counts?
7. Why does this pipeline use `keep_all_features = FALSE` with a strict
   post-quantification feature-count check?
8. Why should the fragment-derived common-peak counts not be expected to match
   the original Cell Ranger ATAC counts exactly?

Once all validation checks pass and `STEP_03_COMPLETE.txt` exists, Step 3 is
complete. The next notebook can safely use the common-feature objects for RNA
normalization/PCA and ATAC TF-IDF/SVD/LSI.
