---
title: "Step 5 - Build the WNN graph and evaluate multimodal clusters"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 4 described every retained nucleus in two independent low-dimensional
coordinate systems:

- **RNA PCA** describes coordinated transcriptional variation;
- **ATAC LSI** describes coordinated chromatin-accessibility variation.

Step 5 combines those two descriptions using **weighted nearest neighbors
(WNN)**. WNN does not simply average RNA and ATAC. Instead, it estimates how
informative each modality is for defining the neighborhood of each individual
nucleus and builds a multimodal neighbor graph from those cell-specific
weights.

This notebook will:

1. load and validate the completed Step 4 object;
2. construct the WNN neighbor graph using RNA PCs 1–30 and ATAC LSI dimensions 2–30;
3. visualize the WNN embedding by sample identity;
4. inspect the RNA-versus-ATAC modality weights learned for each nucleus;
5. explore several graph-clustering resolutions without silently choosing one;
6. compare cluster sizes and descriptive sample composition across resolutions;
7. pause for a **manual clustering-resolution decision**;
8. after a resolution is chosen, inspect mouse-colon epithelial marker programs;
9. save a clustered WNN object for downstream annotation and biological analysis.

A practical SOL allocation for this step is **4 CPU cores, 64–96 GiB memory, 8–12 hours, R 4.4.2, no GPU**.

## One statistical limitation remains important

There is one pooled CON Cell Ranger library and one pooled HFD Cell Ranger library. Each library contains colon epithelial cells pooled from multiple mice, so individual mouse identities are not recoverable after pooling.
Therefore, pooled-library identity and diet condition remain perfectly confounded.

Cluster composition can therefore be described by pooled library, but a difference in CON-versus-HFD abundance must not be presented as a replicated treatment effect.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
})

ggplot2::theme_set(theme_multiome())

sample_sheet <- read_sample_sheet()

unknown_diets <- setdiff(
  unique(sample_sheet$diet),
  names(DIET_COLORS)
)

if (length(unknown_diets) > 0L) {
  stop(
    "No plotting colors defined for: ",
    paste(unknown_diets, collapse = ", "),
    call. = FALSE
  )
}

sample_colors <- stats::setNames(
  unname(DIET_COLORS[sample_sheet$diet]),
  sample_sheet$sample_id
)

configure_parallel_plan()

cat("Configured future workers:", future::nbrOfWorkers(), "\n")
```

# 2. Define Step 5 parameters and outputs

The dimensional choices below are carried forward from Step 4 rather than
selected again after seeing the WNN result.

RNA uses PCs 1–30. ATAC uses LSI dimensions 2–30 because Step 4 showed that
LSI 1 was almost perfectly associated with ATAC sequencing depth
(Spearman rho approximately -0.992).

We begin clustering with a small **resolution grid**. Resolution is not a
measure of statistical significance. It controls how finely the shared-neighbor
graph is partitioned: larger values generally produce more clusters.

```{r step-five-parameters}
STEP5_DIR <- file.path(OUTPUT_DIR, "05_wnn")

STEP5_PLOT_DIR <- file.path(STEP5_DIR, "plots")

step5_review_object_path <- file.path(
  STEP5_DIR,
  "multiome_WNN_resolution_grid.rds"
)

legacy_step5_review_object_path <- file.path(
  STEP5_DIR,
  "Ppar_iKO_multiome_WNN_resolution_grid.rds"
)

step5_final_object_path <- file.path(
  STEP5_DIR,
  "multiome_WNN_clustered.rds"
)

dir.create(STEP5_DIR, recursive = TRUE, showWarnings = FALSE)

dir.create(STEP5_PLOT_DIR,
           recursive = TRUE,
           showWarnings = FALSE)

RNA_DIMS_WNN <- 1:30
ATAC_DIMS_WNN <- 2:30
WNN_K <- 20L

CLUSTER_RESOLUTIONS <- c(0.2, 0.4, 0.6, 0.8, 1.0)

resolution_column <- wnn_resolution_column

step5_parameters <- data.frame(
  parameter = c(
    "RNA WNN dimensions",
    "ATAC WNN dimensions",
    "WNN neighbors (k.nn)",
    "Clustering algorithm",
    "Resolution grid"
  ),
  value = c(
    paste(RNA_DIMS_WNN, collapse = ","),
    paste(ATAC_DIMS_WNN, collapse = ","),
    as.character(WNN_K),
    "3 (SLM)",
    paste(CLUSTER_RESOLUTIONS, collapse = ",")
  )
)

step5_parameters
```

# 3. Load and validate the Step 4 checkpoint

Step 5 should never silently rebuild PCA or LSI. Those decisions were made and
audited in Step 4, so this notebook requires the Step 4 completion marker and
saved reduced object.

```{r load-step-four}
step4_marker <- file.path(OUTPUT_DIR, "04_reduction", "STEP_04_COMPLETE.txt")

step4_object_path <- file.path(OUTPUT_DIR,
                               "04_reduction",
                               "Ppar_iKO_multiome_merged_reduced.rds")

assert_file(step4_marker, "Step 4 completion marker")

assert_file(step4_object_path, "Step 4 reduced merged object")

merged <- readRDS(step4_object_path)

required_reductions <- c("pca", "lsi", "umap.rna", "umap.atac")

missing_reductions <- setdiff(required_reductions, names(merged@reductions))

if (length(missing_reductions) > 0L) {
  stop(
    "Step 4 object is missing reduction(s): ",
    paste(missing_reductions, collapse = ", "),
    call. = FALSE
  )
}

if (max(RNA_DIMS_WNN) > ncol(Embeddings(merged, "pca"))) {
  stop("Requested RNA WNN dimensions exceed the available PCA dimensions.",
       call. = FALSE)
}

if (max(ATAC_DIMS_WNN) > ncol(Embeddings(merged, "lsi"))) {
  stop("Requested ATAC WNN dimensions exceed the available LSI dimensions.",
       call. = FALSE)
}

step4_validation <- data.frame(
  cells = ncol(merged),
  RNA_features = nrow(merged[["RNA"]]),
  ATAC_features = nrow(merged[["ATAC"]]),
  PCA_dimensions = ncol(Embeddings(merged, "pca")),
  LSI_dimensions = ncol(Embeddings(merged, "lsi")),
  fragment_objects = length(Signac::Fragments(merged[["ATAC"]])),
  samples = length(unique(merged$sample_id))
)

step4_validation

table(merged$sample_id)
```

# 4. Construct the weighted nearest-neighbor graph

A nearest-neighbor graph asks which nuclei are most similar to each other.
WNN extends this idea by asking how useful RNA and ATAC are **for each nucleus**
when defining that neighborhood.

Conceptually:

```text
RNA PCA similarity ----\
                        > cell-specific modality weighting -> WNN neighbors
ATAC LSI similarity ---/
```

The resulting objects have different jobs:

- `weighted.nn` stores the multimodal nearest-neighbor structure;
- `wknn` is the weighted k-nearest-neighbor graph;
- `wsnn` is the weighted shared-nearest-neighbor graph used for clustering;
- `RNA.weight` records the learned RNA contribution for each nucleus.

```{r build-wnn, cache=FALSE}
step5_complete_marker <- file.path(STEP5_DIR, "STEP_05_COMPLETE.txt")

if (file.exists(step5_complete_marker)) {
  message("Invalidating the previous Step 5 completion marker before rebuilding WNN.")

  unlink(step5_complete_marker)
}

set.seed(20260727)

merged <- FindMultiModalNeighbors(
  merged,
  reduction.list = list("pca", "lsi"),
  dims.list = list(RNA_DIMS_WNN, ATAC_DIMS_WNN),
  k.nn = WNN_K,
  knn.graph.name = "wknn",
  snn.graph.name = "wsnn",
  weighted.nn.name = "weighted.nn",
  modality.weight.name = c("RNA.weight", "ATAC.weight"),
  verbose = TRUE
)
```

# 5. Validate and inspect the WNN modality weights

The WNN graph should contain every nucleus from Step 4. The RNA modality weight
should also be finite and between zero and one.

For this two-modality WNN analysis, Seurat stores both `RNA.weight` and `ATAC.weight` directly in the metadata. These weights describe how informative each modality is for local neighborhood prediction. They are **not** RNA or ATAC expression/accessibility percentages.

```{r validate-wnn}
if (!"weighted.nn" %in% names(merged@neighbors)) {
  stop(
    "The weighted.nn neighbor object was not created.",
    call. = FALSE
  )
}

required_graphs <- c(
  "wknn",
  "wsnn"
)

missing_graphs <- setdiff(
  required_graphs,
  names(merged@graphs)
)

if (length(missing_graphs) > 0L) {
  stop(
    "WNN graph(s) were not created: ",
    paste(missing_graphs, collapse = ", "),
    call. = FALSE
  )
}

required_weights <- c(
  "RNA.weight",
  "ATAC.weight"
)

missing_weights <- setdiff(
  required_weights,
  colnames(merged[[]])
)

if (length(missing_weights) > 0L) {
  stop(
    "WNN modality weight(s) were not stored in the merged metadata: ",
    paste(missing_weights, collapse = ", "),
    call. = FALSE
  )
}

rna_weight <- as.numeric(
  merged$RNA.weight
)

atac_weight <- as.numeric(
  merged$ATAC.weight
)

if (any(!is.finite(rna_weight))) {
  stop(
    "Non-finite RNA modality weights were detected.",
    call. = FALSE
  )
}

if (any(!is.finite(atac_weight))) {
  stop(
    "Non-finite ATAC modality weights were detected.",
    call. = FALSE
  )
}

if (any(rna_weight < 0 | rna_weight > 1)) {
  stop(
    "RNA modality weights fall outside the expected 0-1 range.",
    call. = FALSE
  )
}

if (any(atac_weight < 0 | atac_weight > 1)) {
  stop(
    "ATAC modality weights fall outside the expected 0-1 range.",
    call. = FALSE
  )
}

weight_sum_error <- abs(
  rna_weight + atac_weight - 1
)

if (any(weight_sum_error > 1e-6)) {
  stop(
    "RNA and ATAC modality weights do not sum to one.",
    call. = FALSE
  )
}

modality_weight_summary <- merged[[]] |>
  dplyr::group_by(sample_id) |>
  dplyr::summarise(
    cells = dplyr::n(),
    RNA_weight_q25 = stats::quantile(
      RNA.weight,
      0.25
    ),
    RNA_weight_median = stats::median(
      RNA.weight
    ),
    RNA_weight_q75 = stats::quantile(
      RNA.weight,
      0.75
    ),
    ATAC_weight_median = stats::median(
      ATAC.weight
    ),
    .groups = "drop"
  )

modality_weight_summary

summary(merged$RNA.weight)
summary(merged$ATAC.weight)
```

## Plot modality weights by sample

A systematic shift in modality weight between libraries is worth noticing, but
it is not automatically evidence of a technical problem. Different biological
cell-state composition can also change which modality is locally informative.

```{r modality-weight-by-sample, fig.width=9, fig.height=5}
p_weight_violin <- ggplot2::ggplot(
  merged[[]],
  ggplot2::aes(
    x = sample_id,
    y = RNA.weight
  )
) +
  ggplot2::geom_violin(
    scale = "width",
    trim = TRUE
  ) +
  ggplot2::geom_boxplot(
    width = 0.12,
    outlier.shape = NA
  ) +
  ggplot2::labs(
    title = "RNA modality weight by sample",
    x = NULL,
    y = "RNA modality weight"
  ) +
  theme_multiome()

print(p_weight_violin)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    "01_RNA_modality_weight_by_sample.pdf"
  ),
  plot = p_weight_violin,
  width = 8,
  height = 5
)
```
RNA and ATAC modality-weight distributions were similar between the two pooled libraries, supporting use of the joint WNN representation.

# 6. Build the WNN UMAP

UMAP is now constructed from the multimodal `weighted.nn` neighborhood rather
than from RNA PCA or ATAC LSI alone.

The WNN UMAP is a visualization of the multimodal neighborhood graph. The graph,
not the two-dimensional UMAP coordinates, is what will be used for clustering.

```{r run-wnn-umap, cache=FALSE}
set.seed(20260727)

merged <- RunUMAP(
  merged,
  nn.name = "weighted.nn",
  reduction.name = "wnn.umap",
  reduction.key = "wnnUMAP_",
  seed.use = 20260727,
  verbose = TRUE
)
```

## WNN UMAP by sample

Because there is one pooled Cell Ranger library per condition, this plot is descriptive. We want to know whether both pooled libraries occupy the major multimodal states, not to claim that local sample-density differences are replicated diet effects.

```{r plot-wnn-by-sample, fig.width=9, fig.height=7}
p_wnn_sample <- DimPlot(
  merged,
  reduction = "wnn.umap",
  group.by = "sample_id",
  cols = sample_colors,
  pt.size = 0.25,
  seed = 20260727,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Multimodal structure across pooled libraries",
    subtitle = paste(
      "WNN combines RNA and ATAC neighborhood information;",
      "library and diet remain confounded."
    ),
    color = NULL
  ) +
  theme_multiome()

print(p_wnn_sample)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    "02_WNN_UMAP_by_sample.pdf"
  ),
  plot = p_wnn_sample,
  width = 9,
  height = 7
)
```

## Map modality weights onto the WNN embedding

```{r plot-modality-weights-on-wnn, fig.width=12, fig.height=6}
wnn_embedding <- as.data.frame(
  Embeddings(
    merged,
    "wnn.umap"
  )
)

colnames(wnn_embedding)[1:2] <- c(
  "WNN_1",
  "WNN_2"
)

wnn_embedding$RNA.weight <- merged$RNA.weight
wnn_embedding$ATAC.weight <- merged$ATAC.weight
wnn_embedding$sample_id <- merged$sample_id

p_rna_weight <- ggplot2::ggplot(
  wnn_embedding,
  ggplot2::aes(
    x = WNN_1,
    y = WNN_2,
    color = RNA.weight
  )
) +
  ggplot2::geom_point(
    size = 0.25,
    alpha = 0.8
  ) +
  ggplot2::scale_color_viridis_c() +
  ggplot2::labs(
    title = "RNA contribution to WNN neighborhoods",
    x = "WNN UMAP 1",
    y = "WNN UMAP 2",
    color = "RNA weight"
  ) +
  theme_multiome()

p_atac_weight <- ggplot2::ggplot(
  wnn_embedding,
  ggplot2::aes(
    x = WNN_1,
    y = WNN_2,
    color = ATAC.weight
  )
) +
  ggplot2::geom_point(
    size = 0.25,
    alpha = 0.8
  ) +
  ggplot2::scale_color_viridis_c() +
  ggplot2::labs(
    title = "ATAC contribution to WNN neighborhoods",
    x = "WNN UMAP 1",
    y = "WNN UMAP 2",
    color = "ATAC weight"
  ) +
  theme_multiome()

print(p_rna_weight)
print(p_atac_weight)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    "03_WNN_UMAP_RNA_weight.pdf"
  ),
  plot = p_rna_weight,
  width = 8,
  height = 6
)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    "04_WNN_UMAP_ATAC_weight.pdf"
  ),
  plot = p_atac_weight,
  width = 8,
  height = 6
)
```

# 7. Explore a clustering-resolution grid

Clusters are communities in the WNN shared-nearest-neighbor graph. They are not
created from the visual shape of the UMAP.

We deliberately calculate several plausible resolutions before choosing one.
A useful resolution should separate coherent biological programs without
splitting continuous populations into many arbitrary microclusters.

**Do not choose a resolution because it produces the number of cell types you
expected in advance.**

```{r cluster-resolution-grid, cache=FALSE}
# Clustering uses random starts.
# Seurat parallelizes FindClusters() through future_lapply() when
# multiple future workers are active, which can trigger an
# undeclared parallel-RNG warning.
#
# For reproducibility, run this small resolution grid sequentially.

future::plan(future::sequential)

cat("Workers used for clustering:", future::nbrOfWorkers(), "\n")

for (resolution in CLUSTER_RESOLUTIONS) {
  cluster_column <- resolution_column(resolution)

  merged <- FindClusters(
    merged,
    graph.name = "wsnn",
    algorithm = 3,
    resolution = resolution,
    cluster.name = cluster_column,
    random.seed = 20260727,
    verbose = FALSE
  )
}

# Restore the project-wide parallel configuration afterward.
configure_parallel_plan()

cat("Workers restored after clustering:",
    future::nbrOfWorkers(),
    "\n")
```

## Summarize cluster number and size

A very small cluster is not automatically invalid: rare enteroendocrine or tuft
populations can be biologically real. This table is meant to expose the
consequences of resolution, not impose a minimum cluster size by itself.

```{r summarize-resolution-grid}
cluster_count_summary <- dplyr::bind_rows(
  lapply(
    CLUSTER_RESOLUTIONS,
    function(resolution) {

      cluster_column <- resolution_column(
        resolution
      )

      cluster_values <- merged[[]][[
        cluster_column
      ]]

      cluster_sizes <- table(
        cluster_values
      )

      data.frame(
        resolution = resolution,
        cluster_column = cluster_column,
        n_clusters = length(cluster_sizes),
        smallest_cluster = min(cluster_sizes),
        median_cluster = stats::median(as.numeric(cluster_sizes)),
        largest_cluster = max(cluster_sizes)
      )
    }
  )
)

cluster_count_summary
```

## Descriptive sample composition of each resolution

These are **descriptive counts**, not treatment-abundance statistics. Because
sample and diet are confounded, do not use this table to perform a CON-versus-HFD
hypothesis test.

```{r cluster-sample-composition}
cluster_sample_composition <- dplyr::bind_rows(
  lapply(
    CLUSTER_RESOLUTIONS,
    function(resolution) {

      cluster_column <- resolution_column(
        resolution
      )

      temp <- data.frame(
        resolution = resolution,
        cluster = as.character(
          merged[[]][[
            cluster_column
          ]]
        ),
        sample_id = merged$sample_id
      )

      temp |>
        dplyr::count(
          resolution,
          cluster,
          sample_id,
          name = "cells"
        ) |>
        dplyr::group_by(
          resolution,
          cluster
        ) |>
        dplyr::mutate(
          percent_of_cluster = 100 * cells / sum(cells)
        ) |>
        dplyr::ungroup()
    }
  )
)

cluster_sample_composition
```

## Plot every candidate resolution on the same WNN embedding

The coordinates remain fixed. Only the graph partition changes.

```{r plot-resolution-grid, fig.width=9, fig.height=7}
for (resolution in CLUSTER_RESOLUTIONS) {
  cluster_column <- resolution_column(resolution)

  p_resolution <- DimPlot(
    merged,
    reduction = "wnn.umap",
    group.by = cluster_column,
    label = TRUE,
    repel = TRUE,
    label.size = 3,
    pt.size = 0.2
  ) +
    NoLegend() +
    ggplot2::ggtitle(paste("WNN clusters - resolution", resolution))

  print(p_resolution)

  ggplot2::ggsave(
    filename = file.path(
      STEP5_PLOT_DIR,
      paste0("WNN_clusters_resolution_", resolution, ".pdf")
    ),
    plot = p_resolution,
    width = 9,
    height = 7
  )
}
```

# 8. Save the WNN resolution-grid checkpoint

At this point the WNN analysis is complete, but the clustering resolution has
**not** been biologically approved. We save this state so that changing the
chosen resolution later does not require rebuilding WNN.

```{r save-resolution-grid-checkpoint}
readr::write_csv(
  step5_parameters,
  file.path(
    STEP5_DIR,
    "WNN_parameters.csv"
  )
)

readr::write_csv(
  modality_weight_summary,
  file.path(
    STEP5_DIR,
    "modality_weight_summary.csv"
  )
)

readr::write_csv(
  cluster_count_summary,
  file.path(
    STEP5_DIR,
    "cluster_resolution_summary.csv"
  )
)

readr::write_csv(
  cluster_sample_composition,
  file.path(
    STEP5_DIR,
    "cluster_sample_composition_descriptive.csv"
  )
)

saveRDS(
  merged,
  step5_review_object_path,
  compress = FALSE
)

step5_review_object_info <- data.frame(
  path = step5_review_object_path,
  exists = file.exists(step5_review_object_path),
  size_gb = file.info(step5_review_object_path)$size / 1024^3
)

step5_review_object_info

writeLines(
  c(
    paste(
      "Step 5 WNN review checkpoint created:",
      format(Sys.time())
    ),
    paste(
      "Cells:",
      ncol(merged)
    ),
    paste(
      "RNA WNN dimensions:",
      paste(RNA_DIMS_WNN, collapse = ",")
    ),
    paste(
      "ATAC WNN dimensions:",
      paste(ATAC_DIMS_WNN, collapse = ",")
    ),
    paste(
      "Resolution grid:",
      paste(CLUSTER_RESOLUTIONS, collapse = ",")
    ),
    paste(
      "Clustering decision table:",
      CLUSTERING_DECISION_PATH
    ),
    "Final clustering resolution approved: FALSE"
  ),
  file.path(
    STEP5_DIR,
    "STEP_05_READY_FOR_CLUSTER_REVIEW.txt"
  )
)

clustering_template_created <- write_clustering_decision_template()

if (isTRUE(clustering_template_created)) {
  message(
    "Created an unapproved clustering decision template: ",
    CLUSTERING_DECISION_PATH
  )
}
```

# STOP HERE FOR THE MANUAL RESOLUTION REVIEW

Before continuing, inspect:

1. `02_WNN_UMAP_by_sample.pdf`;
2. the RNA- and ATAC-weight WNN plots;
3. `cluster_resolution_summary.csv`;
4. each candidate-resolution UMAP;
5. whether any clusters are populated almost exclusively by one library;
6. whether increasing resolution is separating visibly coherent states or only
   fragmenting continuous structures.

Then record the reviewed resolution in the active cohort's
`clustering_decision.csv`. The decision table is the auditable checkpoint;
the notebook does not contain a cohort-independent default resolution.

---

# 9. Choose one clustering resolution after review

This chunk validates the versioned cohort decision before applying it. In a
fresh R session, it reloads the saved resolution-grid object automatically;
there is no need to rebuild WNN simply to finalize a reviewed resolution.

```{r choose-resolution}
if (!exists("merged", inherits = FALSE)) {
  review_checkpoint_candidates <- c(
    step5_review_object_path,
    legacy_step5_review_object_path
  )

  available_review_checkpoints <- review_checkpoint_candidates[
    file.exists(review_checkpoint_candidates)
  ]

  if (length(available_review_checkpoints) == 0L) {
    stop(
      "No Step 5 resolution-grid checkpoint was found. Run through Step 8 ",
      "before validating the clustering decision.",
      call. = FALSE
    )
  }

  checkpoint_to_load <- available_review_checkpoints[[1]]

  if (identical(checkpoint_to_load, legacy_step5_review_object_path)) {
    message(
      "Loading legacy-named resolution-grid checkpoint: ",
      checkpoint_to_load
    )
  }

  merged <- readRDS(checkpoint_to_load)
}

clustering_decision <- read_and_validate_clustering_decision(
  available_resolutions = CLUSTER_RESOLUTIONS,
  metadata_columns = colnames(merged[[]])
)

CHOSEN_RESOLUTION <- clustering_decision$resolution[[1]]
CHOSEN_CLUSTER_COLUMN <- clustering_decision$cluster_column[[1]]

if (
  length(CHOSEN_RESOLUTION) != 1L ||
    is.na(CHOSEN_RESOLUTION) ||
    !CHOSEN_RESOLUTION %in% CLUSTER_RESOLUTIONS
) {
  stop(
    paste0(
      "The approved clustering resolution must match one value in ",
      "CLUSTER_RESOLUTIONS."
    ),
    call. = FALSE
  )
}

Idents(merged) <- merged[[]][[
  CHOSEN_CLUSTER_COLUMN
]]

cat(
  "Chosen WNN clustering resolution:",
  CHOSEN_RESOLUTION,
  "\n"
)

cat(
  "Number of clusters:",
  length(levels(Idents(merged))),
  "\n"
)
```

# 10. Audit a mouse-colon epithelial marker panel

This dataset contains **mouse colonic epithelium**, not small-intestinal
epithelium. Annotation therefore uses a colon-focused marker framework.

A marker should not silently disappear from a plot. Before visualization, this
chunk records whether each requested symbol exists in the RNA feature set and
whether any RNA molecules were actually detected.

This is a symbol/detection audit, not a comprehensive synonym database. A gene
that is absent from the matrix can be investigated for aliases later if needed.

```{r colon-marker-audit}
colon_marker_panel <- list(
  Stem_crypt = c(
    "Lgr5",
    "Ascl2",
    "Smoc2",
    "Axin2",
    "Sox9"
  ),
  Proliferative_TA = c(
    "Mki67",
    "Top2a",
    "Ccnb1",
    "Ccna2"
  ),
  Absorptive_colonocyte = c(
    "Slc26a3",
    "Car1",
    "Aqp8",
    "Krt20"
  ),
  Goblet = c(
    "Muc2",
    "Tff3",
    "Clca1",
    "Fcgbp",
    "Agr2"
  ),
  Deep_crypt_secretory = c(
    "Reg4",
    "Mptx1",
    "Kit",
    "Spink4"
  ),
  Enteroendocrine = c(
    "Chga",
    "Chgb",
    "Neurod1",
    "Pyy",
    "Gcg",
    "Tph1"
  ),
  Tuft = c(
    "Dclk1",
    "Pou2f3",
    "Trpm5"
  )
)

marker_lookup <- data.frame(
  marker_program = rep(
    names(colon_marker_panel),
    lengths(colon_marker_panel)
  ),
  gene = unname(
    unlist(
      colon_marker_panel
    )
  ),
  stringsAsFactors = FALSE
)

DefaultAssay(merged) <- "RNA"

rna_counts <- SeuratObject::LayerData(
  merged,
  assay = "RNA",
  layer = "counts"
)

marker_audit <- lapply(
  seq_len(nrow(marker_lookup)),
  function(i) {

    gene <- marker_lookup$gene[i]
    program <- marker_lookup$marker_program[i]

    exact_present <- gene %in% rownames(rna_counts)

    case_insensitive_matches <- grep(
      paste0("^", gene, "$"),
      rownames(rna_counts),
      value = TRUE,
      ignore.case = TRUE
    )

    if (exact_present) {
      x <- rna_counts[
        gene,
        ,
        drop = TRUE
      ]

      total_umis <- sum(x)
      positive_cells <- sum(x > 0)
      percent_positive <- 100 * mean(x > 0)
    } else {
      total_umis <- NA_real_
      positive_cells <- NA_integer_
      percent_positive <- NA_real_
    }

    data.frame(
      marker_program = program,
      gene = gene,
      exact_symbol_present = exact_present,
      case_insensitive_match = paste(
        case_insensitive_matches,
        collapse = ";"
      ),
      total_UMIs = total_umis,
      positive_cells = positive_cells,
      percent_positive = percent_positive,
      stringsAsFactors = FALSE
    )
  }
)

marker_audit <- dplyr::bind_rows(
  marker_audit
)

marker_audit$status <- ifelse(
  !marker_audit$exact_symbol_present,
  "not_in_RNA_feature_set",
  ifelse(
    marker_audit$total_UMIs == 0,
    "present_but_zero_counts",
    "detected"
  )
)

marker_audit
```

# 11. Visualize colon marker programs across the chosen WNN clusters

A cluster should be annotated from coordinated marker programs, not from one
marker in isolation. The dot plot summarizes both the prevalence and average
expression of the detected marker genes across the chosen WNN clusters.

```{r colon-marker-dotplot, fig.width=15, fig.height=8}
detected_marker_genes <- marker_audit$gene[
  marker_audit$status == "detected"
]

detected_marker_genes <- unique(
  detected_marker_genes
)

if (length(detected_marker_genes) == 0L) {
  stop(
    "None of the requested mouse-colon marker genes were detected.",
    call. = FALSE
  )
}

p_colon_markers <- suppressMessages(
  Seurat::DotPlot(
    merged,
    features = detected_marker_genes,
    group.by = CHOSEN_CLUSTER_COLUMN,
    assay = "RNA",
    scale = TRUE,
    col.min = -2,
    col.max = 2
  ) +
    ggplot2::scale_color_gradient2(
      low = "#2166AC",
      mid = "#D9D9D9",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-2, 2),
      oob = scales::squish,
      name = "Scaled average\nexpression"
    ) +
    Seurat::RotatedAxis() +
    ggplot2::labs(
      title = paste0(
        "Mouse-colon marker programs across WNN clusters (resolution ",
        CHOSEN_RESOLUTION,
        ")"
      ),
      x = NULL,
      y = "WNN cluster"
    ) +
    theme_multiome(base_size = 10)
)

print(
  p_colon_markers
)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    paste0(
      "05_colon_marker_DotPlot_resolution_",
      CHOSEN_RESOLUTION,
      ".pdf"
    )
  ),
  plot = p_colon_markers,
  width = 20,
  height = 8
)
```

## Chosen clusters on the WNN embedding

```{r chosen-resolution-plot, fig.width=9, fig.height=7}
p_chosen_clusters <- DimPlot(
  merged,
  reduction = "wnn.umap",
  group.by = CHOSEN_CLUSTER_COLUMN,
  label = TRUE,
  repel = TRUE,
  label.size = 3,
  pt.size = 0.2
) +
  NoLegend() +
  ggplot2::ggtitle(
    paste0(
      "Chosen WNN clusters - resolution ",
      CHOSEN_RESOLUTION
    )
  )

print(
  p_chosen_clusters
)

ggplot2::ggsave(
  filename = file.path(
    STEP5_PLOT_DIR,
    paste0(
      "06_chosen_WNN_clusters_resolution_",
      CHOSEN_RESOLUTION,
      ".pdf"
    )
  ),
  plot = p_chosen_clusters,
  width = 9,
  height = 7
)
```

# 12. Save the finalized Step 5 checkpoint

This checkpoint preserves the WNN graph, WNN UMAP, all tested clustering
resolutions, the manually selected resolution, and the marker-audit metadata.
It does **not** assign final biological cell-type labels yet.

```{r save-step-five}
chosen_cluster_sizes <- as.data.frame(
  table(
    cluster = as.character(
      merged[[]][[
        CHOSEN_CLUSTER_COLUMN
      ]]
    )
  ),
  stringsAsFactors = FALSE
)

colnames(chosen_cluster_sizes) <- c(
  "cluster",
  "cells"
)

readr::write_csv(
  marker_audit,
  file.path(
    STEP5_DIR,
    "mouse_colon_marker_audit.csv"
  )
)

readr::write_csv(
  chosen_cluster_sizes,
  file.path(
    STEP5_DIR,
    paste0(
      "chosen_cluster_sizes_resolution_",
      CHOSEN_RESOLUTION,
      ".csv"
    )
  )
)

DefaultAssay(merged) <- "RNA"

saveRDS(
  merged,
  step5_final_object_path,
  compress = FALSE
)

step5_final_object_info <- data.frame(
  path = step5_final_object_path,
  exists = file.exists(step5_final_object_path),
  size_gb = file.info(step5_final_object_path)$size / 1024^3,
  chosen_resolution = CHOSEN_RESOLUTION,
  chosen_cluster_column = CHOSEN_CLUSTER_COLUMN,
  clusters = length(levels(Idents(merged)))
)

step5_final_object_info

if (!isTRUE(step5_final_object_info$exists)) {
  stop(
    "The finalized Step 5 WNN object was not saved successfully.",
    call. = FALSE
  )
}

writeLines(
  c(
    paste(
      "Step 5 completed:",
      format(Sys.time())
    ),
    paste(
      "Cells:",
      ncol(merged)
    ),
    paste(
      "RNA WNN dimensions:",
      paste(RNA_DIMS_WNN, collapse = ",")
    ),
    paste(
      "ATAC WNN dimensions:",
      paste(ATAC_DIMS_WNN, collapse = ",")
    ),
    paste(
      "Chosen clustering resolution:",
      CHOSEN_RESOLUTION
    ),
    paste(
      "Chosen cluster column:",
      CHOSEN_CLUSTER_COLUMN
    ),
    paste(
      "Clustering decision table:",
      CLUSTERING_DECISION_PATH
    ),
    paste(
      "Clustering decision reviewer:",
      clustering_decision$reviewer[[1]]
    ),
    paste(
      "Clustering decision review date:",
      clustering_decision$review_date[[1]]
    ),
    paste(
      "Clusters:",
      length(levels(Idents(merged)))
    ),
    "Final biological cell-type labels assigned: FALSE",
    "CON-versus-HFD treatment inference performed: FALSE"
  ),
  file.path(
    STEP5_DIR,
    "STEP_05_COMPLETE.txt"
  )
)
```

# Interpretation checkpoint

Before proceeding to formal cell-state annotation, be able to answer:

1. What problem does WNN solve that RNA-only PCA or ATAC-only LSI cannot solve alone?
2. What does `RNA.weight` mean for an individual nucleus?
3. Why is a WNN UMAP not itself the object used to calculate clusters?
4. What is the difference between the `weighted.nn`, `wknn`, and `wsnn` objects?
5. What does the clustering-resolution parameter change?
6. Why should the resolution not be chosen simply because it produces the expected number of cell types?
7. Why can a small WNN cluster still be biologically plausible?
8. Why do we inspect sample composition within clusters but avoid a formal CON-versus-HFD abundance test here?
9. Why should a final cell identity be supported by a multi-gene marker program rather than a single marker?
10. What additional evidence would you want before declaring a cluster a specific mouse-colon epithelial state?

# What comes next?

Step 6 will focus on **cell-state annotation and biological interpretation**.
That step can combine:

- cluster-level RNA marker discovery;
- mouse-colon epithelial marker programs;
- accessibility around lineage-defining genes;
- RNA/ATAC concordance;
- careful identification of rare populations;
- descriptive comparison of cluster representation between the two libraries.

Formal CON-versus-HFD treatment inference remains limited by the absence of
independent biological replicates.

# References used for this step

- Seurat weighted nearest-neighbor analysis:
  https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis
- Seurat `FindMultiModalNeighbors()`:
  https://satijalab.org/seurat/reference/findmultimodalneighbors
- Seurat `FindClusters()`:
  https://satijalab.org/seurat/reference/findclusters
