---
title: "Step 5B - Review the IL17 C11/C17 biological offshoot"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 5 identified an IL17 WNN branch in which global cluster 11 was strongly
enriched for the CON library. Focused review showed that this global cluster
contains two different components:

1. a small, coherent immune-like tuft population supported by RNA, ATAC, and
   WNN structure;
2. a larger low-complexity epithelial population that should not be assigned a
   tuft identity.

Global cluster 17 is included because clusters 11 and 17 descend from the same
resolution-0.2 parent and therefore provide the nearest enteroendocrine
comparison. This notebook recomputes RNA PCA, ATAC LSI, and WNN after subsetting
those two global clusters. It then evaluates a focused resolution grid, applies
the reviewed cohort-specific decision, and exports barcode-level assignments
for Step 6.

The focused analysis does not replace the global WNN solution. It creates a
reviewed annotation override for one stable child population while preserving
the original global cluster identities.

A practical SOL allocation is **4 CPU cores, 32-64 GiB memory, 4 hours,
R 4.4.2, no GPU**.

## Statistical limitation

There is one pooled CON library and one pooled HFD library. Sample identity and
diet are therefore perfectly confounded. Sample composition is descriptive and
must not be interpreted as a replicated diet effect.

# 1. Setup

Set the active dataset and cohort before running this notebook:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "il17"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tibble)
  library(patchwork)
})

ggplot2::theme_set(theme_multiome())

sample_sheet <- read_sample_sheet()

unknown_diets <- setdiff(
  unique(sample_sheet$diet),
  names(DIET_COLORS)
)

if (length(unknown_diets) > 0L) {
  stop(
    "No plotting colors defined for: ",
    paste(unknown_diets, collapse = ", "),
    call. = FALSE
  )
}

sample_colors <- stats::setNames(
  unname(DIET_COLORS[sample_sheet$diet]),
  sample_sheet$sample_id
)

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "il17")
) {
  stop(
    "Step 5B is the reviewed IL17 C11/C17 case study. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=il17 before running it.",
    call. = FALSE
  )
}

set.seed(12345)
future::plan(future::sequential)
```

# 2. Define inputs and outputs

The resolution grid includes 0.5 because it locates the transition between the
stable four-cluster solution and the additional enteroendocrine subdivision
that first appears at resolution 0.6.

```{r define-inputs-and-outputs}
STEP5_DIR <- file.path(OUTPUT_DIR, "05_wnn")

FOCUSED_DIR <- file.path(
  STEP5_DIR,
  "focused_offshoot_C11_C17"
)

STEP5_REVIEW_OBJECT_PATH <- file.path(
  STEP5_DIR,
  "multiome_WNN_resolution_grid.rds"
)

SUBCLUSTERING_DECISION_PATH <- file.path(
  COHORT_CONFIG_DIR,
  "subclustering_decision.csv"
)

FOCUSED_OBJECT_PATH <- file.path(
  FOCUSED_DIR,
  "IL17_C11_C17_focused_WNN.rds"
)

FOCUSED_ASSIGNMENT_PATH <- file.path(
  FOCUSED_DIR,
  "focused_subcluster_assignments.csv"
)

FOCUSED_RESOLUTIONS <- c(
  0.1,
  0.2,
  0.3,
  0.4,
  0.5,
  0.6,
  0.8
)

RNA_DIMS_FOCUSED <- 1:20
ATAC_DIMS_FOCUSED <- 2:20
WNN_K_FOCUSED <- 20L

dir.create(
  FOCUSED_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

assert_file(
  STEP5_REVIEW_OBJECT_PATH,
  "Step 5 WNN resolution-grid object"
)

assert_file(
  SUBCLUSTERING_DECISION_PATH,
  "focused subclustering decision table"
)
```

# 3. Read and validate the reviewed decision

The decision table is kept under the active cohort configuration directory. It
must contain exactly one approved row and complete audit fields.

```{r validate-focused-decision}
subclustering_decision <- suppressMessages(
  readr::read_csv(
    SUBCLUSTERING_DECISION_PATH,
    show_col_types = FALSE,
    progress = FALSE
  )
)

subclustering_decision <- as.data.frame(
  subclustering_decision,
  stringsAsFactors = FALSE
)

required_decision_columns <- c(
  "cohort_id",
  "parent_cluster_column",
  "parent_clusters",
  "focused_resolution",
  "focused_cluster_column",
  "approved_child_cluster",
  "annotation_unit",
  "approved",
  "reviewer",
  "review_date",
  "decision_notes"
)

missing_decision_columns <- setdiff(
  required_decision_columns,
  colnames(subclustering_decision)
)

if (length(missing_decision_columns) > 0L) {
  stop(
    "Subclustering decision table is missing: ",
    paste(missing_decision_columns, collapse = ", "),
    call. = FALSE
  )
}

if (nrow(subclustering_decision) != 1L) {
  stop(
    "Subclustering decision table must contain exactly one row.",
    call. = FALSE
  )
}

approved <- tolower(
  trimws(as.character(subclustering_decision$approved))
) %in% c("true", "t", "1", "yes", "y")

audit_columns <- c(
  "reviewer",
  "review_date",
  "decision_notes"
)

missing_audit <- vapply(
  audit_columns,
  function(column_name) {
    value <- trimws(
      as.character(subclustering_decision[[column_name]])
    )
    is.na(value) || value == ""
  },
  logical(1)
)

if (!isTRUE(approved) || any(missing_audit)) {
  stop(
    "Focused subclustering is paused intentionally. Complete the ",
    "decision table, its audit fields, and set approved=TRUE.",
    call. = FALSE
  )
}

if (
  !identical(
    tolower(trimws(subclustering_decision$cohort_id)),
    COHORT_ID
  )
) {
  stop(
    "Subclustering decision cohort does not match the active cohort.",
    call. = FALSE
  )
}

PARENT_CLUSTER_COLUMN <- trimws(
  subclustering_decision$parent_cluster_column
)

PARENT_CLUSTERS <- trimws(
  strsplit(
    subclustering_decision$parent_clusters,
    ";",
    fixed = TRUE
  )[[1]]
)

CHOSEN_FOCUSED_RESOLUTION <- suppressWarnings(
  as.numeric(subclustering_decision$focused_resolution)
)

CHOSEN_FOCUSED_CLUSTER_COLUMN <- trimws(
  subclustering_decision$focused_cluster_column
)

APPROVED_CHILD_CLUSTER <- trimws(
  as.character(subclustering_decision$approved_child_cluster)
)

APPROVED_ANNOTATION_UNIT <- trimws(
  subclustering_decision$annotation_unit
)

if (
  !is.finite(CHOSEN_FOCUSED_RESOLUTION) ||
    !CHOSEN_FOCUSED_RESOLUTION %in% FOCUSED_RESOLUTIONS
) {
  stop(
    "The approved focused resolution is not in FOCUSED_RESOLUTIONS.",
    call. = FALSE
  )
}

expected_focused_column <- paste0(
  "offshoot_res_",
  format(
    CHOSEN_FOCUSED_RESOLUTION,
    trim = TRUE,
    scientific = FALSE,
    nsmall = 1
  )
)

if (!identical(CHOSEN_FOCUSED_CLUSTER_COLUMN, expected_focused_column)) {
  stop(
    "The decision table's focused cluster column does not match its ",
    "resolution. Expected: ",
    expected_focused_column,
    call. = FALSE
  )
}

subclustering_decision
```

# 4. Load the global Step 5 checkpoint and isolate the branch

```{r load-and-subset-global-wnn}
merged <- readRDS(STEP5_REVIEW_OBJECT_PATH)

required_metadata <- c(
  PARENT_CLUSTER_COLUMN,
  "sample_id",
  "diet"
)

missing_metadata <- setdiff(
  required_metadata,
  colnames(merged[[]])
)

if (length(missing_metadata) > 0L) {
  stop(
    "Step 5 object is missing metadata: ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

if (!all(c("RNA", "ATAC") %in% names(merged@assays))) {
  stop(
    "Step 5 object must contain RNA and ATAC assays.",
    call. = FALSE
  )
}

parent_values <- as.character(
  merged[[]][[PARENT_CLUSTER_COLUMN]]
)

focused_cells <- rownames(merged[[]])[
  parent_values %in% PARENT_CLUSTERS
]

if (length(focused_cells) == 0L) {
  stop(
    "No cells matched the reviewed parent clusters.",
    call. = FALSE
  )
}

focused <- subset(
  merged,
  cells = focused_cells
)

focused$parent_wnn_cluster <- as.character(
  focused[[]][[PARENT_CLUSTER_COLUMN]]
)

parent_sample_crosswalk <- focused[[]] |>
  tibble::rownames_to_column("barcode") |>
  dplyr::count(
    parent_wnn_cluster,
    sample_id,
    diet,
    name = "nuclei"
  )

readr::write_csv(
  parent_sample_crosswalk,
  file.path(
    FOCUSED_DIR,
    "parent_cluster_sample_crosswalk.csv"
  )
)

parent_sample_crosswalk
```

# 5. Create fresh focused RNA and ATAC assays

Fresh assay copies prevent the focused analysis from reusing scaled RNA values
or ATAC inverse-document-frequency values calculated on the full cohort. ATAC
features with zero counts in the focused cells are removed before TF-IDF.

```{r create-focused-assays}
rna_counts <- GetAssayData(
  focused,
  assay = "RNA",
  layer = "counts"
)

focused[["RNA.focused"]] <- CreateAssayObject(
  counts = rna_counts
)

focused[["ATAC.focused"]] <- focused[["ATAC"]]
Key(focused[["ATAC.focused"]]) <- "ATACFOCUSED_"

atac_counts <- GetAssayData(
  focused,
  assay = "ATAC.focused",
  layer = "counts"
)

nonzero_atac_features <- Matrix::rowSums(atac_counts) > 0

atac_feature_filter_summary <- data.frame(
  ATAC_features_before = length(nonzero_atac_features),
  zero_count_features_removed = sum(!nonzero_atac_features),
  ATAC_features_retained = sum(nonzero_atac_features)
)

focused[["ATAC.focused"]] <- focused[["ATAC.focused"]][
  nonzero_atac_features,
]

readr::write_csv(
  atac_feature_filter_summary,
  file.path(
    FOCUSED_DIR,
    "focused_ATAC_feature_filter_summary.csv"
  )
)

rm(rna_counts, atac_counts)
invisible(gc())

atac_feature_filter_summary
```

# 6. Recompute focused RNA PCA and ATAC LSI

```{r focused-rna-reduction}
DefaultAssay(focused) <- "RNA.focused"

focused <- NormalizeData(
  focused,
  assay = "RNA.focused",
  verbose = FALSE
)

focused <- FindVariableFeatures(
  focused,
  assay = "RNA.focused",
  selection.method = "vst",
  nfeatures = 3000,
  verbose = FALSE
)

focused <- ScaleData(
  focused,
  assay = "RNA.focused",
  features = VariableFeatures(focused[["RNA.focused"]]),
  verbose = FALSE
)

focused <- RunPCA(
  focused,
  assay = "RNA.focused",
  features = VariableFeatures(focused[["RNA.focused"]]),
  npcs = 30,
  reduction.name = "offshoot.pca",
  reduction.key = "OFFRNA_",
  verbose = FALSE
)

focused <- RunUMAP(
  focused,
  reduction = "offshoot.pca",
  dims = RNA_DIMS_FOCUSED,
  reduction.name = "offshoot.rna.umap",
  reduction.key = "OFFRNAUMAP_",
  seed.use = 12345,
  verbose = FALSE
)
```

```{r focused-atac-reduction}
DefaultAssay(focused) <- "ATAC.focused"

focused <- RunTFIDF(
  focused,
  assay = "ATAC.focused",
  verbose = FALSE
)

focused <- FindTopFeatures(
  focused,
  assay = "ATAC.focused",
  min.cutoff = "q5",
  verbose = FALSE
)

focused <- RunSVD(
  focused,
  assay = "ATAC.focused",
  n = 30,
  reduction.name = "offshoot.lsi",
  reduction.key = "OFFLSI_",
  verbose = FALSE
)

focused <- RunUMAP(
  focused,
  reduction = "offshoot.lsi",
  dims = ATAC_DIMS_FOCUSED,
  reduction.name = "offshoot.atac.umap",
  reduction.key = "OFFATACUMAP_",
  seed.use = 12345,
  verbose = FALSE
)
```

# 7. Build the focused WNN representation

One metadata name is supplied for each modality so the focused RNA and ATAC
weights remain explicit and cannot be confused with the global WNN weights.

```{r focused-wnn}
focused <- FindMultiModalNeighbors(
  focused,
  reduction.list = list(
    "offshoot.pca",
    "offshoot.lsi"
  ),
  dims.list = list(
    RNA_DIMS_FOCUSED,
    ATAC_DIMS_FOCUSED
  ),
  k.nn = WNN_K_FOCUSED,
  knn.graph.name = "offshoot.wknn",
  snn.graph.name = "offshoot.wsnn",
  weighted.nn.name = "offshoot.weighted.nn",
  modality.weight.name = c(
    "focused.RNA.weight",
    "focused.ATAC.weight"
  ),
  verbose = FALSE
)

focused <- RunUMAP(
  focused,
  nn.name = "offshoot.weighted.nn",
  reduction.name = "offshoot.wnn.umap",
  reduction.key = "OFFWNNUMAP_",
  seed.use = 12345,
  verbose = FALSE
)

reduction_validation <- vapply(
  c(
    "offshoot.pca",
    "offshoot.lsi",
    "offshoot.rna.umap",
    "offshoot.atac.umap",
    "offshoot.wnn.umap"
  ),
  function(reduction_name) {
    all(
      is.finite(
        Embeddings(
          focused,
          reduction = reduction_name
        )
      )
    )
  },
  logical(1)
)

if (!all(reduction_validation)) {
  stop(
    "One or more focused reductions contain non-finite values.",
    call. = FALSE
  )
}

reduction_validation
```

# 8. Record strict and inclusive tuft-program support

The strict definition excludes `St18` because global cluster 17 expresses it
without a complete tuft program. The inclusive definition retains `St18` to
reproduce the original candidate review. Neither definition is used to build
the focused clusters; they are independent post-clustering validation labels.

```{r tuft-program-support}
strict_tuft_genes <- c(
  "Dclk1",
  "Pou2f3",
  "Trpm5",
  "Gnat3",
  "Plcb2",
  "Rgs13",
  "Sh2d6",
  "Fyb",
  "Ltc4s",
  "Avil",
  "Hck"
)

inclusive_tuft_genes <- union(
  strict_tuft_genes,
  "St18"
)

rna_focused_data <- GetAssayData(
  focused,
  assay = "RNA.focused",
  layer = "data"
)

count_detected_genes <- function(features) {
  available_features <- intersect(
    features,
    rownames(rna_focused_data)
  )

  Matrix::colSums(
    rna_focused_data[
      available_features,
      colnames(focused),
      drop = FALSE
    ] > 0
  )
}

strict_detected <- count_detected_genes(
  strict_tuft_genes
)

inclusive_detected <- count_detected_genes(
  inclusive_tuft_genes
)

focused$tuft_program_genes_detected_strict <-
  strict_detected[colnames(focused)]

focused$tuft_program_genes_detected_inclusive <-
  inclusive_detected[colnames(focused)]

focused$tuft_program_status_strict <- factor(
  ifelse(
    focused$tuft_program_genes_detected_strict >= 2,
    "program_positive",
    "program_negative"
  ),
  levels = c("program_negative", "program_positive")
)

focused$tuft_program_status_inclusive <- factor(
  ifelse(
    focused$tuft_program_genes_detected_inclusive >= 2,
    "program_positive",
    "program_negative"
  ),
  levels = c("program_negative", "program_positive")
)

strict_crosswalk <- focused[[]] |>
  tibble::rownames_to_column("barcode") |>
  dplyr::count(
    parent_wnn_cluster,
    sample_id,
    diet,
    tuft_program_status_strict,
    name = "nuclei"
  )

inclusive_crosswalk <- focused[[]] |>
  tibble::rownames_to_column("barcode") |>
  dplyr::count(
    parent_wnn_cluster,
    sample_id,
    diet,
    tuft_program_status_inclusive,
    name = "nuclei"
  )

readr::write_csv(
  strict_crosswalk,
  file.path(
    FOCUSED_DIR,
    "strict_tuft_program_crosswalk.csv"
  )
)

readr::write_csv(
  inclusive_crosswalk,
  file.path(
    FOCUSED_DIR,
    "inclusive_tuft_program_crosswalk.csv"
  )
)

strict_crosswalk
inclusive_crosswalk
```

# 9. Construct and summarize the focused resolution grid

```{r focused-resolution-grid}
focused_resolution_columns <- character(
  length(FOCUSED_RESOLUTIONS)
)

for (resolution_index in seq_along(FOCUSED_RESOLUTIONS)) {
  resolution <- FOCUSED_RESOLUTIONS[[resolution_index]]

  focused <- FindClusters(
    focused,
    graph.name = "offshoot.wsnn",
    algorithm = 3,
    resolution = resolution,
    random.seed = 12345,
    verbose = FALSE
  )

  resolution_column <- paste0(
    "offshoot_res_",
    format(
      resolution,
      trim = TRUE,
      scientific = FALSE,
      nsmall = 1
    )
  )

  focused[[resolution_column]] <- as.character(
    Idents(focused)
  )

  focused_resolution_columns[[resolution_index]] <-
    resolution_column
}

focused_metadata <- focused[[]] |>
  tibble::as_tibble()

total_strict_positive <- sum(
  focused_metadata$tuft_program_status_strict ==
    "program_positive"
)

total_inclusive_positive <- sum(
  focused_metadata$tuft_program_status_inclusive ==
    "program_positive"
)

focused_resolution_summary <- dplyr::bind_rows(
  lapply(
    seq_along(FOCUSED_RESOLUTIONS),
    function(resolution_index) {
      resolution <- FOCUSED_RESOLUTIONS[[resolution_index]]

      resolution_column <-
        focused_resolution_columns[[resolution_index]]

      focused_metadata |>
        dplyr::group_by(
          focused_subcluster = .data[[resolution_column]]
        ) |>
        dplyr::summarise(
          nuclei = dplyr::n(),
          parent_C11_percent = 100 * mean(
            parent_wnn_cluster == "11"
          ),
          parent_C17_percent = 100 * mean(
            parent_wnn_cluster == "17"
          ),
          strict_positive_n = sum(
            tuft_program_status_strict ==
              "program_positive"
          ),
          strict_positive_percent = 100 * mean(
            tuft_program_status_strict ==
              "program_positive"
          ),
          strict_positive_capture_percent = 100 *
            sum(
              tuft_program_status_strict ==
                "program_positive"
            ) /
            total_strict_positive,
          inclusive_positive_n = sum(
            tuft_program_status_inclusive ==
              "program_positive"
          ),
          inclusive_positive_percent = 100 * mean(
            tuft_program_status_inclusive ==
              "program_positive"
          ),
          inclusive_positive_capture_percent = 100 *
            sum(
              tuft_program_status_inclusive ==
                "program_positive"
            ) /
            total_inclusive_positive,
          CON_percent = 100 * mean(
            as.character(diet) == "CON"
          ),
          HFD_percent = 100 * mean(
            as.character(diet) == "HFD"
          ),
          nCount_RNA_median = median(nCount_RNA),
          nCount_ATAC_median = median(nCount_ATAC),
          RNA_weight_median = median(
            .data[["focused.RNA.weight"]],
            na.rm = TRUE
          ),
          .groups = "drop"
        ) |>
        dplyr::mutate(
          resolution = resolution,
          .before = 1
        )
    }
  )
) |>
  tibble::as_tibble()

readr::write_csv(
  focused_resolution_summary,
  file.path(
    FOCUSED_DIR,
    "focused_resolution_summary.csv"
  )
)

focused_resolution_summary
```

# 10. Save the focused resolution and modality-support plots

```{r save-focused-resolution-plots}
for (resolution_index in seq_along(FOCUSED_RESOLUTIONS)) {
  resolution <- FOCUSED_RESOLUTIONS[[resolution_index]]

  resolution_column <-
    focused_resolution_columns[[resolution_index]]

  resolution_plot <- DimPlot(
    focused,
    reduction = "offshoot.wnn.umap",
    group.by = resolution_column,
    label = TRUE,
    repel = TRUE
  ) +
    NoLegend() +
    ggtitle(
      paste(
        "Focused C11/C17 WNN - resolution",
        resolution
      )
    )

  ggsave(
    filename = file.path(
      FOCUSED_DIR,
      paste0(
        "focused_WNN_clusters_resolution_",
        resolution,
        ".pdf"
      )
    ),
    plot = resolution_plot,
    width = 8,
    height = 6
  )
}
```

```{r save-focused-support-plots}
program_colors <- c(
  program_negative = "#B8B8B8",
  program_positive = "#D55E00"
)

rna_program_plot <- DimPlot(
  focused,
  reduction = "offshoot.rna.umap",
  group.by = "tuft_program_status_strict",
  cols = program_colors
) +
  ggtitle("Focused RNA support")

atac_program_plot <- DimPlot(
  focused,
  reduction = "offshoot.atac.umap",
  group.by = "tuft_program_status_strict",
  cols = program_colors
) +
  ggtitle("Focused ATAC support")

wnn_program_plot <- DimPlot(
  focused,
  reduction = "offshoot.wnn.umap",
  group.by = "tuft_program_status_strict",
  cols = program_colors
) +
  ggtitle("Focused WNN support")

ggsave(
  filename = file.path(
    FOCUSED_DIR,
    "focused_RNA_ATAC_WNN_tuft_program_support.pdf"
  ),
  plot = patchwork::wrap_plots(
    rna_program_plot,
    atac_program_plot,
    wnn_program_plot,
    nrow = 1
  ),
  width = 18,
  height = 6
)

sample_plot <- DimPlot(
  focused,
  reduction = "offshoot.wnn.umap",
  group.by = "sample_id",
  cols = sample_colors
) +
  ggtitle("Focused C11/C17 WNN by sample")

ggsave(
  filename = file.path(
    FOCUSED_DIR,
    "focused_WNN_by_sample.pdf"
  ),
  plot = sample_plot,
  width = 8,
  height = 6
)
```

# 11. Preserve the exploratory enteroendocrine subdivision

Resolution 0.6 divides the global-cluster-17 enteroendocrine branch into two
mixed-library children. This comparison is retained for cross-cohort follow-up,
but the subdivision is not promoted to the primary IL17 annotation.

```{r exploratory-eec-audit}
EEC_REVIEW_COLUMN <- "offshoot_res_0.6"
EEC_REVIEW_CLUSTERS <- c("1", "4")

if (
  EEC_REVIEW_COLUMN %in% colnames(focused[[]]) &&
    all(
      EEC_REVIEW_CLUSTERS %in%
        unique(focused[[]][[EEC_REVIEW_COLUMN]])
    )
) {
  DefaultAssay(focused) <- "RNA.focused"
  Idents(focused) <- EEC_REVIEW_COLUMN

  eec_split_markers <- FindMarkers(
    focused,
    ident.1 = "4",
    ident.2 = "1",
    assay = "RNA.focused",
    logfc.threshold = 0.25,
    min.pct = 0.10,
    only.pos = FALSE
  ) |>
    tibble::rownames_to_column("gene") |>
    dplyr::arrange(dplyr::desc(avg_log2FC))

  readr::write_csv(
    eec_split_markers,
    file.path(
      FOCUSED_DIR,
      "RNA_EEC_subclusters_4_vs_1_resolution_0.6.csv"
    )
  )

  eec_review <- subset(
    focused,
    idents = EEC_REVIEW_CLUSTERS
  )

  eec_panel <- intersect(
    c(
      "Chga",
      "Chgb",
      "Insm1",
      "Neurod1",
      "Rfx6",
      "Pax6",
      "Isl1",
      "Tph1",
      "Ddc",
      "Lmx1a",
      "Tac1",
      "Gcg",
      "Pyy",
      "Nts",
      "Sst",
      "Cck",
      "Ghrl",
      "Reg4"
    ),
    rownames(eec_review[["RNA.focused"]])
  )

  eec_marker_plot <- DotPlot(
    eec_review,
    features = eec_panel,
    group.by = EEC_REVIEW_COLUMN,
    assay = "RNA.focused"
  ) +
    RotatedAxis() +
    ggtitle(
      "Focused enteroendocrine split at resolution 0.6"
    )

  ggsave(
    filename = file.path(
      FOCUSED_DIR,
      "focused_EEC_marker_audit_resolution_0.6.pdf"
    ),
    plot = eec_marker_plot,
    width = 11,
    height = 5
  )
}
```

# 12. Apply the reviewed focused decision

The selected child is identified from unsupervised focused WNN membership. The
tuft-program flags remain validation evidence and do not directly determine
which nuclei receive the override.

```{r apply-focused-decision}
if (
  !CHOSEN_FOCUSED_CLUSTER_COLUMN %in%
    colnames(focused[[]])
) {
  stop(
    "Chosen focused cluster column is absent from the object: ",
    CHOSEN_FOCUSED_CLUSTER_COLUMN,
    call. = FALSE
  )
}

focused$focused_subcluster <- as.character(
  focused[[]][[CHOSEN_FOCUSED_CLUSTER_COLUMN]]
)

if (
  !APPROVED_CHILD_CLUSTER %in%
    unique(focused$focused_subcluster)
) {
  stop(
    "Approved focused child cluster is absent from the chosen solution.",
    call. = FALSE
  )
}

approved_child_cells <- colnames(focused)[
  focused$focused_subcluster == APPROVED_CHILD_CLUSTER
]

approved_child_parents <- unique(
  focused$parent_wnn_cluster[
    focused$focused_subcluster == APPROVED_CHILD_CLUSTER
  ]
)

if (length(approved_child_parents) != 1L) {
  stop(
    "Approved focused child spans more than one global parent cluster.",
    call. = FALSE
  )
}

focused$annotation_unit_override <- ifelse(
  focused$focused_subcluster == APPROVED_CHILD_CLUSTER,
  APPROVED_ANNOTATION_UNIT,
  NA_character_
)

focused_assignments <- focused[[]] |>
  tibble::rownames_to_column("barcode") |>
  dplyr::transmute(
    barcode,
    sample_id,
    diet,
    parent_cluster_column = PARENT_CLUSTER_COLUMN,
    parent_wnn_cluster,
    focused_resolution = CHOSEN_FOCUSED_RESOLUTION,
    focused_cluster_column = CHOSEN_FOCUSED_CLUSTER_COLUMN,
    focused_subcluster,
    tuft_program_status_strict,
    tuft_program_status_inclusive,
    focused_RNA_weight = .data[["focused.RNA.weight"]],
    focused_ATAC_weight = .data[["focused.ATAC.weight"]],
    annotation_unit_override
  )

approved_assignment_count <- sum(
  focused_assignments$annotation_unit_override ==
    APPROVED_ANNOTATION_UNIT,
  na.rm = TRUE
)

if (approved_assignment_count != length(approved_child_cells)) {
  stop(
    "Focused assignment count does not match the approved child size.",
    call. = FALSE
  )
}

readr::write_csv(
  focused_assignments,
  FOCUSED_ASSIGNMENT_PATH
)

focused_decision_audit <- data.frame(
  cohort_id = COHORT_ID,
  parent_cluster_column = PARENT_CLUSTER_COLUMN,
  parent_cluster = approved_child_parents,
  parent_clusters_reviewed = paste(
    PARENT_CLUSTERS,
    collapse = ";"
  ),
  focused_resolution = CHOSEN_FOCUSED_RESOLUTION,
  focused_cluster_column = CHOSEN_FOCUSED_CLUSTER_COLUMN,
  approved_child_cluster = APPROVED_CHILD_CLUSTER,
  annotation_unit = APPROVED_ANNOTATION_UNIT,
  approved_nuclei = approved_assignment_count,
  reviewer = subclustering_decision$reviewer,
  review_date = subclustering_decision$review_date,
  stringsAsFactors = FALSE
)

readr::write_csv(
  focused_decision_audit,
  file.path(
    FOCUSED_DIR,
    "focused_subclustering_decision_applied.csv"
  )
)

focused_decision_audit
```

# 13. Save the focused checkpoint

```{r save-focused-checkpoint}
DefaultAssay(focused) <- "RNA"

saveRDS(
  focused,
  FOCUSED_OBJECT_PATH,
  compress = FALSE
)

writeLines(
  c(
    paste(
      "Focused subclustering completed:",
      format(Sys.time())
    ),
    paste(
      "Parent global clusters:",
      paste(PARENT_CLUSTERS, collapse = ",")
    ),
    paste(
      "Focused nuclei:",
      ncol(focused)
    ),
    paste(
      "Chosen focused resolution:",
      CHOSEN_FOCUSED_RESOLUTION
    ),
    paste(
      "Chosen focused cluster column:",
      CHOSEN_FOCUSED_CLUSTER_COLUMN
    ),
    paste(
      "Approved child cluster:",
      APPROVED_CHILD_CLUSTER
    ),
    paste(
      "Annotation unit override:",
      APPROVED_ANNOTATION_UNIT
    ),
    paste(
      "Approved nuclei:",
      approved_assignment_count
    ),
    "Exploratory EEC subdivision promoted to primary annotation: FALSE"
  ),
  file.path(
    FOCUSED_DIR,
    "STEP_05B_COMPLETE.txt"
  )
)

focused_object_info <- data.frame(
  path = FOCUSED_OBJECT_PATH,
  exists = file.exists(FOCUSED_OBJECT_PATH),
  size_gb = file.info(FOCUSED_OBJECT_PATH)$size / 1024^3,
  assignments_path = FOCUSED_ASSIGNMENT_PATH,
  assignments_exist = file.exists(FOCUSED_ASSIGNMENT_PATH),
  approved_nuclei = approved_assignment_count,
  stringsAsFactors = FALSE
)

if (
  !isTRUE(focused_object_info$exists) ||
    !isTRUE(focused_object_info$assignments_exist)
) {
  stop(
    "Focused Step 5B outputs were not saved successfully.",
    call. = FALSE
  )
}

focused_object_info
```

# 14. Interpretation checkpoint

The approved 42-nucleus child is a stable, unsupervised WNN population derived
entirely from global cluster 11. It is supported by both diets and by a
coordinated immune-like tuft program. Step 6 will use
`focused_subcluster_assignments.csv` to create the annotation unit `11_tuft`.

The remaining global-cluster-11 nuclei retain their global identity and should
receive a conservative low-complexity epithelial annotation. The exploratory
resolution-0.6 enteroendocrine split is preserved in this checkpoint for later
comparison with WT and PPAR but does not alter the primary IL17 labels.
