---
title: "Step 5C - Bidirectional mapping of Gourab d21 scRNA and WT multiome RNA"
subtitle: "Version 2 - presto marker testing and reciprocal WT-to-source mapping"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

This auxiliary review maps the independently generated Gourab d21 whole-cell
scRNA-seq reference onto the RNA component of the reviewed WT snMultiome
object. The source reference contains matched CON and HFD libraries and 13
named cell-state annotations recovered from the processed d21 loom file.

The analysis has five purposes:

1. reconstruct the source reference from the full Cell Ranger count matrices
   while preserving the loom-derived biological labels;
2. test how the source Stem, TAC, DSC, Goblet, and other identities correspond
   to the current WT WNN clusters and cell-state annotations;
3. provide external evidence for the focused review of WT clusters 0, 6, and
   8 and the related secretory populations;
4. reverse the mapping direction by projecting the reviewed WT states and WNN
   clusters onto the Gourab d21 cells, then quantify bidirectional agreement;
5. visualize cross-platform alignment with Harmony as a secondary sensitivity
   analysis.

The source labels and transferred labels are evidence, not automatic
annotation overrides. Harmony is used only for visualization and neighborhood
alignment. Original uncorrected RNA counts remain the basis for markers and
all subsequent expression analyses.

Although this notebook is numbered Step 5C to keep it beside the focused
clustering reviews, it reads the current reviewed Step 6 WT object so that the
external mapping can be compared directly with the existing annotations. It
does not modify or replace that object.

A practical SOL allocation is **8 CPU cores, 64-96 GiB memory, 8 hours,
R 4.4.2, no GPU**.

## Statistical limitation

The recovered source metadata identifies cells only as CON or HFD. It does not
retain mouse-level assignments. The WT snMultiome likewise contains one pooled
library per diet. Consequently, this notebook performs no diet-effect
inference and no abundance testing. Diet-specific summaries are descriptive.

# 1. Setup

Set the active dataset and cohort before running this notebook:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(tibble)
  library(Matrix)
  library(patchwork)
  library(scales)
})

if (!requireNamespace("hdf5r", quietly = TRUE)) {
  stop("Package 'hdf5r' is required to read the d21 loom metadata.")
}

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 5C requires MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}

ggplot2::theme_set(theme_multiome())
set.seed(12345)
future::plan(future::sequential)
```

# 2. Define inputs, outputs, and review parameters

```{r define-inputs-and-outputs}
ANALYSIS_VERSION <- "05c_v2_bidirectional_presto"

REFERENCE_ROOT <- file.path(
  PROJECT_DIR,
  "reference_data",
  "gourab_vilb_d21_scrna"
)

CON_H5_PATH <- file.path(
  REFERENCE_ROOT,
  "cellranger",
  "ConCon21",
  "outs",
  "filtered_feature_bc_matrix.h5"
)

HFD_H5_PATH <- file.path(
  REFERENCE_ROOT,
  "cellranger",
  "HFHF21",
  "outs",
  "filtered_feature_bc_matrix.h5"
)

LOOM_PATH <- file.path(
  REFERENCE_ROOT,
  "processed",
  "day21",
  "Day21_data_POST.loom"
)

STEP6_OBJECT_PATH <- file.path(
  OUTPUT_DIR,
  "06_annotation",
  "multiome_annotated.rds"
)

STEP5C_DIR <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "gourab_d21_rna_reference_mapping"
)

STEP5C_PLOT_DIR <- file.path(STEP5C_DIR, "plots")
STEP5C_TABLE_DIR <- file.path(STEP5C_DIR, "tables")

SOURCE_REFERENCE_PATH <- file.path(
  STEP5C_DIR,
  "Gourab_d21_annotated_scRNA_reference.rds"
)

WT_TRANSFER_ASSIGNMENT_PATH <- file.path(
  STEP5C_DIR,
  "WT_Gourab_d21_label_transfer_assignments.csv"
)

for (output_directory in c(
  STEP5C_DIR,
  STEP5C_PLOT_DIR,
  STEP5C_TABLE_DIR
)) {
  dir.create(
    output_directory,
    recursive = TRUE,
    showWarnings = FALSE
  )
}

for (input_path in c(
  CON_H5_PATH,
  HFD_H5_PATH,
  LOOM_PATH,
  STEP6_OBJECT_PATH
)) {
  assert_file(input_path)
}

SOURCE_VARIABLE_FEATURES <- 3000L
SOURCE_PCS <- 50L
TRANSFER_DIMS <- 1:30
HARMONY_DIMS <- 1:30

# These score bands are descriptive review aids, not biological probabilities
# or substitutes for the manually assigned annotation-confidence field.
HIGH_MAPPING_SCORE <- 0.75
MODERATE_MAPPING_SCORE <- 0.50
WT_INVERSE_REFERENCE_CONFIDENCE <- c("high", "moderate")

CRYPT_FOCUS_CLUSTERS <- c("0", "6", "8")
SECRETORY_FOCUS_CLUSTERS <- c("4", "5", "7", "9", "13", "18")
FOCUS_CLUSTERS <- unique(c(
  CRYPT_FOCUS_CLUSTERS,
  SECRETORY_FOCUS_CLUSTERS
))

SOURCE_MARKER_PANEL <- list(
  Stem = c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Sox9"),
  Proliferation = c("Mki67", "Top2a", "Cenpf", "Ube2c"),
  DSC = c("Reg4", "Spink4", "Kit", "Agr2"),
  Goblet = c("Muc2", "Spdef", "Fcgbp", "Clca1", "Zg16"),
  Enterocyte = c("Car4", "Krt20", "Aqp8", "Slc26a3"),
  Enteroendocrine = c("Chga", "Chgb", "Neurod1", "Pyy"),
  Tuft = c("Pou2f3", "Dclk1", "Trpm5", "Gnat3"),
  Immune = c("Ptprc", "Cd3d", "Cd79a", "Nkg7", "Lyz2")
)

SOURCE_MARKER_GENES <- unique(unlist(SOURCE_MARKER_PANEL))

save_pdf_plot <- function(path, plot_object, width, height) {
  grDevices::pdf(
    file = path,
    width = width,
    height = height,
    useDingbats = FALSE
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot_object)
  invisible(path)
}
```

# 3. Recover the loom-derived source annotations

The loom contains the named cell identities and QC metadata but only 2,000
features. Its expression matrix is therefore not used. The `_1` and `_2`
suffixes were appended during integration; the original 10x barcode is
recovered by removing only the final numeric suffix.

```{r recover-source-loom-metadata}
read_loom_metadata <- function(path) {
  loom_file <- hdf5r::H5File$new(path, mode = "r")
  on.exit(loom_file$close_all(), add = TRUE)

  read_attribute <- function(attribute_name) {
    loom_file[[file.path("col_attrs", attribute_name)]][]
  }

  tibble::tibble(
    cell_id = as.character(read_attribute("CellID")),
    source_diet = as.character(read_attribute("orig.ident")),
    source_cell_state = as.character(
      read_attribute("clusters_named")
    ),
    source_integrated_cluster = as.character(
      read_attribute("integrated_snn_res.0.5")
    ),
    source_seurat_cluster = as.character(
      read_attribute("seurat_clusters")
    ),
    loom_nCount_RNA = as.numeric(read_attribute("nCount_RNA")),
    loom_nFeature_RNA = as.numeric(
      read_attribute("nFeature_RNA")
    ),
    loom_percent_mt = as.numeric(read_attribute("percent.mt"))
  )
}

source_metadata <- read_loom_metadata(LOOM_PATH) |>
  dplyr::mutate(
    raw_barcode = sub("_[0-9]+$", "", cell_id),
    integration_suffix = sub(
      "^.*_([0-9]+)$",
      "\\1",
      cell_id
    )
  )

if (anyDuplicated(source_metadata$cell_id) > 0L) {
  stop("The d21 loom contains duplicated CellID values.", call. = FALSE)
}

if (anyDuplicated(
  paste(source_metadata$source_diet, source_metadata$raw_barcode)
) > 0L) {
  stop(
    "The d21 loom contains duplicated diet-specific raw barcodes.",
    call. = FALSE
  )
}

expected_source_diets <- c("CON", "HFD")

if (!setequal(unique(source_metadata$source_diet), expected_source_diets)) {
  stop(
    "Expected exactly CON and HFD in the d21 loom metadata.",
    call. = FALSE
  )
}

suffix_audit <- source_metadata |>
  dplyr::count(
    source_diet,
    integration_suffix,
    name = "cells"
  ) |>
  dplyr::arrange(source_diet, integration_suffix)

suffix_counts_by_diet <- suffix_audit |>
  dplyr::count(source_diet, name = "suffixes")

if (any(suffix_counts_by_diet$suffixes != 1L)) {
  stop(
    "Each source diet must map to exactly one integration suffix.",
    call. = FALSE
  )
}

if (dplyr::n_distinct(suffix_audit$integration_suffix) != 2L) {
  stop(
    "CON and HFD must have distinct integration suffixes.",
    call. = FALSE
  )
}

readr::write_csv(
  suffix_audit,
  file.path(STEP5C_TABLE_DIR, "source_integration_suffix_audit.csv")
)

source_label_counts <- source_metadata |>
  dplyr::count(
    source_diet,
    source_cell_state,
    name = "cells"
  ) |>
  dplyr::group_by(source_diet) |>
  dplyr::mutate(
    percent_within_diet = 100 * cells / sum(cells)
  ) |>
  dplyr::ungroup()

readr::write_csv(
  source_label_counts,
  file.path(STEP5C_TABLE_DIR, "source_label_counts_by_diet.csv")
)

suffix_audit
source_label_counts
```

# 4. Reconnect the named cells to the full source count matrices

Barcode matching is performed separately within CON and HFD. This avoids
ambiguity from the 10x barcode sequences shared between independent runs.

```{r reconstruct-full-source-counts}
read_gene_expression_h5 <- function(path) {
  count_object <- Seurat::Read10X_h5(
    filename = path,
    use.names = TRUE,
    unique.features = TRUE
  )

  if (is.list(count_object)) {
    if ("Gene Expression" %in% names(count_object)) {
      count_object <- count_object[["Gene Expression"]]
    } else if (length(count_object) == 1L) {
      count_object <- count_object[[1]]
    } else {
      stop(
        "The H5 file contains multiple modalities but no Gene Expression ",
        "matrix.",
        call. = FALSE
      )
    }
  }

  count_object
}

con_counts <- read_gene_expression_h5(CON_H5_PATH)
hfd_counts <- read_gene_expression_h5(HFD_H5_PATH)

shared_source_genes <- intersect(
  rownames(con_counts),
  rownames(hfd_counts)
)

if (length(shared_source_genes) == 0L) {
  stop("The source H5 matrices have no shared genes.", call. = FALSE)
}

con_metadata <- source_metadata |>
  dplyr::filter(source_diet == "CON")

hfd_metadata <- source_metadata |>
  dplyr::filter(source_diet == "HFD")

source_cell_recovery_audit <- tibble::tibble(
  source_diet = c("CON", "HFD"),
  filtered_matrix_cells = c(ncol(con_counts), ncol(hfd_counts)),
  annotated_loom_cells = c(nrow(con_metadata), nrow(hfd_metadata)),
  annotated_barcodes_found = c(
    sum(con_metadata$raw_barcode %in% colnames(con_counts)),
    sum(hfd_metadata$raw_barcode %in% colnames(hfd_counts))
  ),
  annotated_cell_recovery_percent = 100 *
    annotated_loom_cells / filtered_matrix_cells
)

if (any(
  source_cell_recovery_audit$annotated_barcodes_found !=
    source_cell_recovery_audit$annotated_loom_cells
)) {
  stop(
    "Not every loom-annotated barcode was recovered from its diet-specific ",
    "H5 matrix.",
    call. = FALSE
  )
}

con_annotated_counts <- con_counts[
  shared_source_genes,
  con_metadata$raw_barcode,
  drop = FALSE
]

hfd_annotated_counts <- hfd_counts[
  shared_source_genes,
  hfd_metadata$raw_barcode,
  drop = FALSE
]

colnames(con_annotated_counts) <- con_metadata$cell_id
colnames(hfd_annotated_counts) <- hfd_metadata$cell_id

source_counts <- cbind(
  con_annotated_counts,
  hfd_annotated_counts
)

source_counts <- source_counts[
  ,
  source_metadata$cell_id,
  drop = FALSE
]

if (!identical(colnames(source_counts), source_metadata$cell_id)) {
  stop(
    "Reconstructed source counts are not aligned to the loom metadata.",
    call. = FALSE
  )
}

raw_nCount_RNA <- Matrix::colSums(source_counts)
raw_nFeature_RNA <- Matrix::colSums(source_counts > 0)

source_count_concordance <- source_metadata |>
  dplyr::mutate(
    raw_nCount_RNA = as.numeric(.env$raw_nCount_RNA[cell_id]),
    raw_nFeature_RNA = as.numeric(.env$raw_nFeature_RNA[cell_id]),
    nCount_difference = raw_nCount_RNA - loom_nCount_RNA,
    nFeature_difference = raw_nFeature_RNA - loom_nFeature_RNA
  ) |>
  dplyr::group_by(source_diet) |>
  dplyr::summarise(
    cells = dplyr::n(),
    exact_nCount_matches = sum(nCount_difference == 0),
    exact_nFeature_matches = sum(nFeature_difference == 0),
    median_absolute_nCount_difference = stats::median(
      abs(nCount_difference)
    ),
    median_absolute_nFeature_difference = stats::median(
      abs(nFeature_difference)
    ),
    .groups = "drop"
  )

readr::write_csv(
  source_cell_recovery_audit,
  file.path(STEP5C_TABLE_DIR, "source_cell_recovery_audit.csv")
)

readr::write_csv(
  source_count_concordance,
  file.path(STEP5C_TABLE_DIR, "source_count_metadata_concordance.csv")
)

source_cell_recovery_audit
source_count_concordance
```

# 5. Build and independently audit the annotated source reference

```{r build-source-reference}
source_metadata_frame <- source_metadata |>
  dplyr::select(
    cell_id,
    source_diet,
    source_cell_state,
    source_integrated_cluster,
    source_seurat_cluster,
    integration_suffix,
    raw_barcode,
    loom_nCount_RNA,
    loom_nFeature_RNA,
    loom_percent_mt
  ) |>
  as.data.frame(stringsAsFactors = FALSE)

rownames(source_metadata_frame) <- source_metadata_frame$cell_id
source_metadata_frame$cell_id <- NULL

source_reference <- Seurat::CreateSeuratObject(
  counts = source_counts,
  assay = "RNA",
  project = "Gourab_d21_scRNA",
  meta.data = source_metadata_frame,
  min.cells = 0,
  min.features = 0
)

source_reference$dataset_platform <- "Gourab d21 scRNA"
source_reference$source_or_query <- "Annotated reference"
source_reference$mapping_state <- source_reference$source_cell_state

Seurat::Idents(source_reference) <- "source_cell_state"

source_reference <- Seurat::NormalizeData(
  source_reference,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

source_reference <- Seurat::FindVariableFeatures(
  source_reference,
  selection.method = "vst",
  nfeatures = SOURCE_VARIABLE_FEATURES,
  verbose = FALSE
)

source_reference <- Seurat::ScaleData(
  source_reference,
  features = Seurat::VariableFeatures(source_reference),
  verbose = FALSE
)

source_reference <- Seurat::RunPCA(
  source_reference,
  features = Seurat::VariableFeatures(source_reference),
  npcs = SOURCE_PCS,
  verbose = FALSE
)

source_reference <- Seurat::RunUMAP(
  source_reference,
  reduction = "pca",
  dims = TRANSFER_DIMS,
  reduction.name = "umap.source",
  reduction.key = "sourceUMAP_",
  seed.use = 12345,
  verbose = FALSE
)

source_state_levels <- sort(unique(
  as.character(source_reference$source_cell_state)
))

source_state_colors <- stats::setNames(
  scales::hue_pal()(length(source_state_levels)),
  source_state_levels
)

p_source_state <- Seurat::DimPlot(
  source_reference,
  reduction = "umap.source",
  group.by = "source_cell_state",
  cols = source_state_colors,
  label = TRUE,
  repel = TRUE,
  raster = TRUE
) +
  ggplot2::labs(
    title = "Reconstructed Gourab d21 scRNA reference",
    subtitle = "Named identities recovered from the processed d21 loom"
  ) +
  ggplot2::theme(legend.position = "right")

p_source_diet <- Seurat::DimPlot(
  source_reference,
  reduction = "umap.source",
  group.by = "source_diet",
  cols = DIET_COLORS,
  raster = TRUE
) +
  ggplot2::labs(
    title = "Source reference by diet",
    subtitle = "Composition is descriptive; mouse-level identity is absent"
  )

p_source_overview <- p_source_state + p_source_diet +
  patchwork::plot_layout(widths = c(1.3, 1))

print(p_source_overview)

save_pdf_plot(
  file.path(STEP5C_PLOT_DIR, "01_source_reference_UMAP.pdf"),
  p_source_overview,
  width = 16,
  height = 7
)

available_source_markers <- intersect(
  SOURCE_MARKER_GENES,
  rownames(source_reference)
)

missing_source_markers <- setdiff(
  SOURCE_MARKER_GENES,
  available_source_markers
)

source_marker_availability <- tibble::tibble(
  gene = SOURCE_MARKER_GENES,
  present = SOURCE_MARKER_GENES %in% available_source_markers
)

readr::write_csv(
  source_marker_availability,
  file.path(STEP5C_TABLE_DIR, "source_marker_panel_availability.csv")
)

if (length(available_source_markers) > 0L) {
  p_source_markers <- Seurat::DotPlot(
    source_reference,
    features = available_source_markers,
    group.by = "source_cell_state",
    assay = "RNA",
    cols = c("grey90", "#2166AC"),
    dot.scale = 5
  ) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      title = "Source-reference marker audit",
      subtitle = paste0(
        "Full H5 expression; ",
        length(missing_source_markers),
        " panel gene(s) absent"
      ),
      x = NULL,
      y = NULL
    ) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(
        angle = 45,
        hjust = 1
      )
    )

  print(p_source_markers)

  save_pdf_plot(
    file.path(STEP5C_PLOT_DIR, "02_source_marker_DotPlot.pdf"),
    p_source_markers,
    width = 14,
    height = 9
  )
}

saveRDS(
  source_reference,
  SOURCE_REFERENCE_PATH,
  compress = FALSE
)
```

# 6. Recalculate source-reference marker evidence

These markers are calculated from the reconstructed full H5 counts rather
than the 2,000-gene loom representation. They provide an auditable check of
the inherited labels; they are not used to redefine those labels here.

```{r discover-source-reference-markers}
if (!requireNamespace("presto", quietly = TRUE)) {
  stop(
    "Source marker discovery requires 'presto' for a tractable Wilcoxon ",
    "implementation. Run source('setup/install_packages.R') in a fresh ",
    "R 4.4.2 session, restart R, and rerun this chunk.",
    call. = FALSE
  )
}

message(
  "presto ",
  as.character(utils::packageVersion("presto")),
  " is available; Seurat will use its accelerated Wilcoxon implementation."
)

Seurat::Idents(source_reference) <- "source_cell_state"

source_all_positive_markers <- Seurat::FindAllMarkers(
  source_reference,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.10,
  logfc.threshold = 0.25,
  test.use = "wilcox",
  verbose = FALSE
)

source_top_markers <- source_all_positive_markers |>
  dplyr::group_by(cluster) |>
  dplyr::arrange(
    p_val_adj,
    dplyr::desc(avg_log2FC),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = 25L) |>
  dplyr::ungroup()

readr::write_csv(
  source_all_positive_markers,
  file.path(
    STEP5C_TABLE_DIR,
    "source_RNA_all_positive_markers.csv"
  )
)

readr::write_csv(
  source_top_markers,
  file.path(
    STEP5C_TABLE_DIR,
    "source_RNA_top_annotation_candidates.csv"
  )
)

source_top_markers |>
  dplyr::select(
    cluster,
    gene,
    avg_log2FC,
    pct.1,
    pct.2,
    p_val_adj
  ) |>
  print(n = 100)
```

# 7. Build a clean WT RNA query from the reviewed Step 6 object

The query is rebuilt from the original RNA counts so that reference and query
receive the same log-normalization workflow. Existing WT annotations and WNN
coordinates are retained only as metadata and independent comparison fields.

```{r build-wt-rna-query}
multiome <- readRDS(STEP6_OBJECT_PATH)
wt_metadata <- multiome[[]]

required_wt_metadata <- c(
  "sample_id",
  "diet",
  "wnn_cluster",
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence"
)

missing_wt_metadata <- setdiff(
  required_wt_metadata,
  colnames(wt_metadata)
)

if (length(missing_wt_metadata) > 0L) {
  stop(
    "The WT Step 6 object is missing metadata column(s): ",
    paste(missing_wt_metadata, collapse = ", "),
    call. = FALSE
  )
}

if (!"RNA" %in% names(multiome)) {
  stop("The WT Step 6 object does not contain an RNA assay.")
}

rna_layers <- SeuratObject::Layers(multiome[["RNA"]])

if (!"counts" %in% rna_layers) {
  stop(
    "The WT RNA assay must contain a joined counts layer.",
    call. = FALSE
  )
}

wt_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)

if (!setequal(colnames(wt_counts), rownames(wt_metadata))) {
  stop(
    "WT RNA counts and Step 6 metadata do not contain the same cells.",
    call. = FALSE
  )
}

wt_counts <- wt_counts[
  ,
  rownames(wt_metadata),
  drop = FALSE
]

wt_metadata$wnn_cluster <- as.character(wt_metadata$wnn_cluster)
wt_metadata$cell_state <- as.character(wt_metadata$cell_state)
wt_metadata$cell_state_full <- as.character(
  wt_metadata$cell_state_full
)
wt_metadata$annotation_confidence <- tolower(as.character(
  wt_metadata$annotation_confidence
))

wt_query <- Seurat::CreateSeuratObject(
  counts = wt_counts,
  assay = "RNA",
  project = "WT_snMultiome_RNA",
  meta.data = wt_metadata,
  min.cells = 0,
  min.features = 0
)

wt_query$dataset_platform <- "WT snMultiome RNA"
wt_query$source_or_query <- "WT query"

wt_query <- Seurat::NormalizeData(
  wt_query,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

wt_query <- Seurat::FindVariableFeatures(
  wt_query,
  selection.method = "vst",
  nfeatures = SOURCE_VARIABLE_FEATURES,
  verbose = FALSE
)

wt_query <- Seurat::ScaleData(
  wt_query,
  features = Seurat::VariableFeatures(wt_query),
  verbose = FALSE
)

wt_query <- Seurat::RunPCA(
  wt_query,
  features = Seurat::VariableFeatures(wt_query),
  npcs = SOURCE_PCS,
  verbose = FALSE
)

shared_reference_query_genes <- intersect(
  rownames(source_reference),
  rownames(wt_query)
)

reference_query_feature_audit <- tibble::tibble(
  source_genes = nrow(source_reference),
  query_genes = nrow(wt_query),
  shared_genes = length(shared_reference_query_genes),
  source_variable_features = length(
    Seurat::VariableFeatures(source_reference)
  ),
  transferable_source_variable_features = sum(
    Seurat::VariableFeatures(source_reference) %in%
      shared_reference_query_genes
  )
)

readr::write_csv(
  reference_query_feature_audit,
  file.path(
    STEP5C_TABLE_DIR,
    "reference_query_feature_overlap_audit.csv"
  )
)

reference_query_feature_audit
```

# 8. Transfer Gourab d21 identities onto the WT RNA query

Transfer scores quantify mapping support within this specific reference. They
are not percentages of biological identity and are not equivalent to the
manually assigned annotation-confidence field.

```{r transfer-source-identities-to-wt}
transfer_features <- intersect(
  Seurat::VariableFeatures(source_reference),
  shared_reference_query_genes
)

if (length(transfer_features) < 1000L) {
  warning(
    "Fewer than 1,000 source variable features are available for transfer."
  )
}

transfer_anchors <- Seurat::FindTransferAnchors(
  reference = source_reference,
  query = wt_query,
  normalization.method = "LogNormalize",
  reference.reduction = "pca",
  reduction = "pcaproject",
  features = transfer_features,
  dims = TRANSFER_DIMS,
  verbose = TRUE
)

transfer_predictions <- Seurat::TransferData(
  anchorset = transfer_anchors,
  refdata = source_reference$source_cell_state,
  dims = TRANSFER_DIMS,
  verbose = TRUE
)

transfer_predictions <- as.data.frame(
  transfer_predictions,
  stringsAsFactors = FALSE
)

transfer_predictions$gourab_d21_predicted_state <- as.character(
  transfer_predictions$predicted.id
)

transfer_predictions$gourab_d21_prediction_score <- as.numeric(
  transfer_predictions$prediction.score.max
)

class_score_columns <- grep(
  "^prediction[.]score[.]",
  colnames(transfer_predictions),
  value = TRUE
)

class_score_columns <- setdiff(
  class_score_columns,
  "prediction.score.max"
)

if (length(class_score_columns) < 2L) {
  stop(
    "TransferData did not return at least two class-score columns.",
    call. = FALSE
  )
}

class_score_matrix <- as.matrix(
  transfer_predictions[, class_score_columns, drop = FALSE]
)

storage.mode(class_score_matrix) <- "numeric"
class_score_matrix[!is.finite(class_score_matrix)] <- 0

class_score_sums <- rowSums(class_score_matrix)

normalized_class_scores <- class_score_matrix
valid_score_rows <- class_score_sums > 0

normalized_class_scores[valid_score_rows, ] <-
  normalized_class_scores[valid_score_rows, , drop = FALSE] /
  class_score_sums[valid_score_rows]

normalized_class_scores[!valid_score_rows, ] <- 0

transfer_predictions$gourab_d21_prediction_entropy <- apply(
  normalized_class_scores,
  1,
  function(score_values) {
    positive_scores <- score_values[score_values > 0]

    if (length(positive_scores) <= 1L) {
      return(0)
    }

    -sum(positive_scores * log(positive_scores)) /
      log(ncol(normalized_class_scores))
  }
)

top_two_scores <- t(apply(
  class_score_matrix,
  1,
  function(score_values) {
    sort(score_values, decreasing = TRUE)[1:2]
  }
))

transfer_predictions$gourab_d21_prediction_margin <-
  top_two_scores[, 1] - top_two_scores[, 2]

transfer_predictions$gourab_d21_mapping_score_band <- dplyr::case_when(
  transfer_predictions$gourab_d21_prediction_score >=
    HIGH_MAPPING_SCORE ~ "high_score",
  transfer_predictions$gourab_d21_prediction_score >=
    MODERATE_MAPPING_SCORE ~ "moderate_score",
  TRUE ~ "low_score"
)

transfer_predictions$predicted.id <- NULL
transfer_predictions$prediction.score.max <- NULL

wt_query <- Seurat::AddMetaData(
  wt_query,
  metadata = transfer_predictions
)

wt_query$mapping_state <- wt_query$gourab_d21_predicted_state

wt_transfer_assignments <- wt_query[[]] |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::select(
    cell_barcode,
    sample_id,
    diet,
    wnn_cluster,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence,
    gourab_d21_predicted_state,
    gourab_d21_prediction_score,
    gourab_d21_prediction_entropy,
    gourab_d21_prediction_margin,
    gourab_d21_mapping_score_band,
    dplyr::all_of(class_score_columns)
  )

readr::write_csv(
  wt_transfer_assignments,
  WT_TRANSFER_ASSIGNMENT_PATH
)

mapping_score_summary <- wt_transfer_assignments |>
  dplyr::summarise(
    WT_nuclei = dplyr::n(),
    median_prediction_score = stats::median(
      gourab_d21_prediction_score
    ),
    prediction_score_q25 = stats::quantile(
      gourab_d21_prediction_score,
      0.25
    ),
    prediction_score_q75 = stats::quantile(
      gourab_d21_prediction_score,
      0.75
    ),
    high_score_percent = 100 * mean(
      gourab_d21_mapping_score_band == "high_score"
    ),
    low_score_percent = 100 * mean(
      gourab_d21_mapping_score_band == "low_score"
    ),
    median_prediction_entropy = stats::median(
      gourab_d21_prediction_entropy
    ),
    median_prediction_margin = stats::median(
      gourab_d21_prediction_margin
    )
  )

readr::write_csv(
  mapping_score_summary,
  file.path(STEP5C_TABLE_DIR, "WT_label_transfer_score_summary.csv")
)

mapping_score_summary
```

# 9. Compare transferred identities with the current WT annotations

The primary crosswalk is calculated within each current WNN cluster. No label
is changed automatically, regardless of mapping score.

```{r create-reference-query-crosswalks}
wt_cluster_crosswalk <- wt_transfer_assignments |>
  dplyr::count(
    wnn_cluster,
    cell_state,
    cell_state_full,
    gourab_d21_predicted_state,
    name = "nuclei"
  ) |>
  dplyr::group_by(wnn_cluster) |>
  dplyr::mutate(
    percent_within_WT_cluster = 100 * nuclei / sum(nuclei)
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    suppressWarnings(as.integer(wnn_cluster)),
    dplyr::desc(nuclei)
  )

wt_cluster_mapping_confidence <- wt_transfer_assignments |>
  dplyr::group_by(
    wnn_cluster,
    cell_state,
    cell_state_full,
    gourab_d21_predicted_state
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    median_prediction_score = stats::median(
      gourab_d21_prediction_score
    ),
    prediction_score_q25 = stats::quantile(
      gourab_d21_prediction_score,
      0.25
    ),
    prediction_score_q75 = stats::quantile(
      gourab_d21_prediction_score,
      0.75
    ),
    median_prediction_entropy = stats::median(
      gourab_d21_prediction_entropy
    ),
    median_prediction_margin = stats::median(
      gourab_d21_prediction_margin
    ),
    high_score_percent = 100 * mean(
      gourab_d21_mapping_score_band == "high_score"
    ),
    CON_percent = 100 * mean(diet == "CON"),
    HFD_percent = 100 * mean(diet == "HFD"),
    .groups = "drop"
  )

wt_cell_state_crosswalk <- wt_transfer_assignments |>
  dplyr::count(
    cell_state,
    gourab_d21_predicted_state,
    name = "nuclei"
  ) |>
  dplyr::group_by(cell_state) |>
  dplyr::mutate(
    percent_within_current_cell_state = 100 * nuclei / sum(nuclei)
  ) |>
  dplyr::ungroup()

wt_focus_crosswalk <- wt_cluster_mapping_confidence |>
  dplyr::filter(wnn_cluster %in% FOCUS_CLUSTERS) |>
  dplyr::group_by(wnn_cluster) |>
  dplyr::mutate(
    percent_within_WT_cluster = 100 * nuclei / sum(nuclei)
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    match(wnn_cluster, FOCUS_CLUSTERS),
    dplyr::desc(nuclei)
  )

readr::write_csv(
  wt_cluster_crosswalk,
  file.path(
    STEP5C_TABLE_DIR,
    "WT_WNN_cluster_by_Gourab_d21_state_crosswalk.csv"
  )
)

readr::write_csv(
  wt_cluster_mapping_confidence,
  file.path(
    STEP5C_TABLE_DIR,
    "WT_WNN_cluster_mapping_confidence.csv"
  )
)

readr::write_csv(
  wt_cell_state_crosswalk,
  file.path(
    STEP5C_TABLE_DIR,
    "WT_current_cell_state_by_Gourab_d21_state_crosswalk.csv"
  )
)

readr::write_csv(
  wt_focus_crosswalk,
  file.path(
    STEP5C_TABLE_DIR,
    "WT_focus_clusters_0_6_8_and_secretory_crosswalk.csv"
  )
)

wt_focus_crosswalk |>
  print(n = Inf, width = Inf)
```

# 10. Display transferred identities in the original WT WNN space

The WNN coordinates are not recomputed. This plot asks where the independently
transferred RNA identities fall within the existing multimodal structure.

```{r plot-transferred-identities-in-wt-wnn-space}
if (!"wnn.umap" %in% names(multiome@reductions)) {
  stop("The WT Step 6 object does not contain the wnn.umap reduction.")
}

wnn_embeddings <- Seurat::Embeddings(
  multiome,
  reduction = "wnn.umap"
)

if (!setequal(rownames(wnn_embeddings), rownames(wt_query[[]]))) {
  stop("WT WNN coordinates and transferred query cells do not match.")
}

wnn_embeddings <- wnn_embeddings[
  rownames(wt_query[[]]),
  ,
  drop = FALSE
]

wt_wnn_mapping_data <- wt_query[[]] |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::mutate(
    wnnUMAP_1 = wnn_embeddings[cell_barcode, 1],
    wnnUMAP_2 = wnn_embeddings[cell_barcode, 2]
  )

p_wt_current <- ggplot2::ggplot(
  wt_wnn_mapping_data,
  ggplot2::aes(x = wnnUMAP_1, y = wnnUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = cell_state),
    size = 0.25,
    alpha = 0.75
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Current WT annotations",
    x = "wnnUMAP_1",
    y = "wnnUMAP_2",
    color = "WT cell state"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_wt_transferred <- ggplot2::ggplot(
  wt_wnn_mapping_data,
  ggplot2::aes(x = wnnUMAP_1, y = wnnUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = gourab_d21_predicted_state),
    size = 0.25,
    alpha = 0.75
  ) +
  ggplot2::scale_color_manual(values = source_state_colors) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Gourab d21 transferred identities",
    subtitle = "Reference mapping of RNA; WNN coordinates are unchanged",
    x = "wnnUMAP_1",
    y = "wnnUMAP_2",
    color = "Transferred state"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_wt_mapping_score <- ggplot2::ggplot(
  wt_wnn_mapping_data,
  ggplot2::aes(x = wnnUMAP_1, y = wnnUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = gourab_d21_prediction_score),
    size = 0.25,
    alpha = 0.80
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    limits = c(0, 1)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Maximum reference-mapping score",
    subtitle = "Score is algorithmic support, not annotation confidence",
    x = "wnnUMAP_1",
    y = "wnnUMAP_2",
    color = "Maximum score"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_wt_mapping_overview <- p_wt_current +
  p_wt_transferred +
  p_wt_mapping_score +
  patchwork::plot_layout(ncol = 3)

print(p_wt_mapping_overview)

save_pdf_plot(
  file.path(STEP5C_PLOT_DIR, "03_WT_WNN_reference_mapping_overview.pdf"),
  p_wt_mapping_overview,
  width = 22,
  height = 7
)
```

# 11. Plot the focused crypt and secretory crosswalk

```{r plot-focused-crosswalk}
focus_plot_data <- wt_focus_crosswalk |>
  dplyr::mutate(
    wnn_cluster = factor(
      wnn_cluster,
      levels = rev(FOCUS_CLUSTERS)
    ),
    gourab_d21_predicted_state = factor(
      gourab_d21_predicted_state,
      levels = source_state_levels
    )
  )

p_focus_crosswalk <- ggplot2::ggplot(
  focus_plot_data,
  ggplot2::aes(
    x = gourab_d21_predicted_state,
    y = wnn_cluster,
    fill = percent_within_WT_cluster
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = ifelse(
        percent_within_WT_cluster >= 3,
        paste0(nuclei, "\n", round(percent_within_WT_cluster, 1), "%"),
        ""
      )
    ),
    size = 3
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "C",
    limits = c(0, 100)
  ) +
  ggplot2::labs(
    title = "External-reference mapping of focused WT populations",
    subtitle = paste0(
      "Crypt review: 0, 6, 8; secretory context: ",
      paste(SECRETORY_FOCUS_CLUSTERS, collapse = ", ")
    ),
    x = "Gourab d21 transferred identity",
    y = "Current WT WNN cluster",
    fill = "% of WT cluster"
  ) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_focus_crosswalk)

save_pdf_plot(
  file.path(STEP5C_PLOT_DIR, "04_focused_cluster_mapping_crosswalk.pdf"),
  p_focus_crosswalk,
  width = 14,
  height = 7
)
```

# 12. Map the reviewed WT annotations back onto Gourab d21 cells

The reciprocal transfer reverses the reference and query roles. Only WT cells
with high- or moderate-confidence reviewed annotations are allowed into the WT
reference. Gourab's observed labels are retained for comparison and are never
used to train this inverse mapping.

Reciprocal agreement is intentionally conservative. A broad source label may
capture most cells from one WT state in the forward direction while mapping
back to several WT states in the reverse direction. Both directions must be
reviewed before claiming that two annotations represent equivalent
populations.

```{r transfer-wt-identities-to-source-reference}
prepare_transfer_predictions <- function(
  prediction_object,
  prefix,
  high_score = HIGH_MAPPING_SCORE,
  moderate_score = MODERATE_MAPPING_SCORE
) {
  prediction_table <- as.data.frame(
    prediction_object,
    stringsAsFactors = FALSE
  )

  score_columns <- grep(
    "^prediction[.]score[.]",
    colnames(prediction_table),
    value = TRUE
  )

  score_columns <- setdiff(
    score_columns,
    "prediction.score.max"
  )

  if (length(score_columns) < 2L) {
    stop(
      "TransferData did not return at least two class-score columns for ",
      prefix,
      ".",
      call. = FALSE
    )
  }

  score_matrix <- as.matrix(
    prediction_table[, score_columns, drop = FALSE]
  )

  storage.mode(score_matrix) <- "numeric"
  score_matrix[!is.finite(score_matrix)] <- 0

  score_sums <- rowSums(score_matrix)
  normalized_scores <- score_matrix
  valid_rows <- score_sums > 0

  normalized_scores[valid_rows, ] <-
    normalized_scores[valid_rows, , drop = FALSE] /
    score_sums[valid_rows]

  normalized_scores[!valid_rows, ] <- 0

  entropy_values <- apply(
    normalized_scores,
    1,
    function(score_values) {
      positive_scores <- score_values[score_values > 0]

      if (length(positive_scores) <= 1L) {
        return(0)
      }

      -sum(positive_scores * log(positive_scores)) /
        log(ncol(normalized_scores))
    }
  )

  top_two_scores <- t(apply(
    score_matrix,
    1,
    function(score_values) {
      sort(score_values, decreasing = TRUE)[1:2]
    }
  ))

  maximum_scores <- as.numeric(
    prediction_table$prediction.score.max
  )

  output_table <- as.data.frame(
    matrix(
      nrow = nrow(prediction_table),
      ncol = 0L,
      dimnames = list(rownames(prediction_table), character())
    )
  )

  output_table[[paste0(prefix, "_predicted_label")]] <- as.character(
    prediction_table$predicted.id
  )

  output_table[[paste0(prefix, "_prediction_score")]] <- maximum_scores
  output_table[[paste0(prefix, "_prediction_entropy")]] <- entropy_values
  output_table[[paste0(prefix, "_prediction_margin")]] <-
    top_two_scores[, 1] - top_two_scores[, 2]

  output_table[[paste0(prefix, "_mapping_score_band")]] <-
    dplyr::case_when(
      maximum_scores >= high_score ~ "high_score",
      maximum_scores >= moderate_score ~ "moderate_score",
      TRUE ~ "low_score"
    )

  renamed_score_table <- prediction_table[
    ,
    score_columns,
    drop = FALSE
  ]

  colnames(renamed_score_table) <- paste0(
    prefix,
    "_class_score_",
    make.names(
      sub("^prediction[.]score[.]", "", score_columns),
      unique = TRUE
    )
  )

  cbind(output_table, renamed_score_table)
}

wt_reference_metadata <- wt_query[[]]

wt_inverse_reference_cells <- rownames(wt_reference_metadata)[
  wt_reference_metadata$annotation_confidence %in%
    WT_INVERSE_REFERENCE_CONFIDENCE &
    !is.na(wt_reference_metadata$cell_state) &
    nzchar(as.character(wt_reference_metadata$cell_state))
]

if (length(wt_inverse_reference_cells) == 0L) {
  stop(
    "No high- or moderate-confidence WT cells are available for inverse ",
    "reference mapping.",
    call. = FALSE
  )
}

wt_reference_counts <- SeuratObject::LayerData(
  wt_query[["RNA"]],
  layer = "counts"
)[
  ,
  wt_inverse_reference_cells,
  drop = FALSE
]

wt_reference_metadata <- wt_reference_metadata[
  wt_inverse_reference_cells,
  ,
  drop = FALSE
]

wt_reference <- Seurat::CreateSeuratObject(
  counts = wt_reference_counts,
  assay = "RNA",
  project = "WT_reviewed_reference",
  meta.data = wt_reference_metadata,
  min.cells = 0,
  min.features = 0
)

wt_reference$wt_reference_cell_state <- as.character(
  wt_reference$cell_state
)
wt_reference$wt_reference_wnn_cluster <- as.character(
  wt_reference$wnn_cluster
)
wt_reference$wt_reference_broad_compartment <- as.character(
  wt_reference$broad_compartment
)

wt_inverse_reference_composition <- wt_reference[[]] |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::count(
    wt_reference_wnn_cluster,
    wt_reference_cell_state,
    wt_reference_broad_compartment,
    annotation_confidence,
    name = "nuclei"
  ) |>
  dplyr::arrange(
    suppressWarnings(as.integer(wt_reference_wnn_cluster))
  )

readr::write_csv(
  wt_inverse_reference_composition,
  file.path(
    STEP5C_TABLE_DIR,
    "WT_inverse_reference_composition.csv"
  )
)

wt_reference <- Seurat::NormalizeData(
  wt_reference,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

wt_reference <- Seurat::FindVariableFeatures(
  wt_reference,
  selection.method = "vst",
  nfeatures = SOURCE_VARIABLE_FEATURES,
  verbose = FALSE
)

wt_reference <- Seurat::ScaleData(
  wt_reference,
  features = Seurat::VariableFeatures(wt_reference),
  verbose = FALSE
)

wt_reference <- Seurat::RunPCA(
  wt_reference,
  features = Seurat::VariableFeatures(wt_reference),
  npcs = SOURCE_PCS,
  verbose = FALSE
)

source_query <- Seurat::CreateSeuratObject(
  counts = source_counts,
  assay = "RNA",
  project = "Gourab_d21_query",
  meta.data = source_metadata_frame,
  min.cells = 0,
  min.features = 0
)

source_query <- Seurat::NormalizeData(
  source_query,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = FALSE
)

source_query <- Seurat::FindVariableFeatures(
  source_query,
  selection.method = "vst",
  nfeatures = SOURCE_VARIABLE_FEATURES,
  verbose = FALSE
)

source_query <- Seurat::ScaleData(
  source_query,
  features = Seurat::VariableFeatures(source_query),
  verbose = FALSE
)

source_query <- Seurat::RunPCA(
  source_query,
  features = Seurat::VariableFeatures(source_query),
  npcs = SOURCE_PCS,
  verbose = FALSE
)

inverse_shared_genes <- intersect(
  rownames(wt_reference),
  rownames(source_query)
)

inverse_transfer_features <- intersect(
  Seurat::VariableFeatures(wt_reference),
  inverse_shared_genes
)

if (length(inverse_transfer_features) < 1000L) {
  warning(
    "Fewer than 1,000 WT variable features are available for inverse ",
    "reference mapping."
  )
}

inverse_transfer_anchors <- Seurat::FindTransferAnchors(
  reference = wt_reference,
  query = source_query,
  normalization.method = "LogNormalize",
  reference.reduction = "pca",
  reduction = "pcaproject",
  features = inverse_transfer_features,
  dims = TRANSFER_DIMS,
  verbose = TRUE
)

inverse_cell_state_raw <- Seurat::TransferData(
  anchorset = inverse_transfer_anchors,
  refdata = wt_reference$wt_reference_cell_state,
  dims = TRANSFER_DIMS,
  verbose = TRUE
)

inverse_wnn_cluster_raw <- Seurat::TransferData(
  anchorset = inverse_transfer_anchors,
  refdata = wt_reference$wt_reference_wnn_cluster,
  dims = TRANSFER_DIMS,
  verbose = TRUE
)

inverse_cell_state_predictions <- prepare_transfer_predictions(
  inverse_cell_state_raw,
  prefix = "WT_cell_state"
)

inverse_wnn_cluster_predictions <- prepare_transfer_predictions(
  inverse_wnn_cluster_raw,
  prefix = "WT_wnn_cluster"
)

source_query <- Seurat::AddMetaData(
  source_query,
  metadata = inverse_cell_state_predictions
)

source_query <- Seurat::AddMetaData(
  source_query,
  metadata = inverse_wnn_cluster_predictions
)

wt_state_to_broad_lookup <- wt_reference[[]] |>
  dplyr::distinct(
    wt_reference_cell_state,
    wt_reference_broad_compartment
  )

state_broad_multiplicity <- wt_state_to_broad_lookup |>
  dplyr::count(wt_reference_cell_state, name = "broad_labels")

if (any(state_broad_multiplicity$broad_labels != 1L)) {
  stop(
    "At least one reviewed WT cell-state label maps to multiple broad ",
    "compartments.",
    call. = FALSE
  )
}

wt_state_to_broad_vector <- stats::setNames(
  wt_state_to_broad_lookup$wt_reference_broad_compartment,
  wt_state_to_broad_lookup$wt_reference_cell_state
)

source_query$WT_predicted_broad_compartment <- unname(
  wt_state_to_broad_vector[
    source_query$WT_cell_state_predicted_label
  ]
)

source_inverse_assignments <- source_query[[]] |>
  tibble::rownames_to_column("source_cell_id") |>
  dplyr::select(
    source_cell_id,
    source_diet,
    source_cell_state,
    source_integrated_cluster,
    source_seurat_cluster,
    WT_cell_state_predicted_label,
    WT_predicted_broad_compartment,
    WT_cell_state_prediction_score,
    WT_cell_state_prediction_entropy,
    WT_cell_state_prediction_margin,
    WT_cell_state_mapping_score_band,
    WT_wnn_cluster_predicted_label,
    WT_wnn_cluster_prediction_score,
    WT_wnn_cluster_prediction_entropy,
    WT_wnn_cluster_prediction_margin,
    WT_wnn_cluster_mapping_score_band,
    dplyr::starts_with("WT_cell_state_class_score_"),
    dplyr::starts_with("WT_wnn_cluster_class_score_")
  )

SOURCE_INVERSE_ASSIGNMENT_PATH <- file.path(
  STEP5C_DIR,
  "Gourab_d21_WT_inverse_label_transfer_assignments.csv"
)

readr::write_csv(
  source_inverse_assignments,
  SOURCE_INVERSE_ASSIGNMENT_PATH
)

inverse_mapping_summary <- source_inverse_assignments |>
  dplyr::summarise(
    source_cells = dplyr::n(),
    WT_reference_nuclei = ncol(wt_reference),
    WT_reference_cell_states = dplyr::n_distinct(
      wt_reference$wt_reference_cell_state
    ),
    median_cell_state_prediction_score = stats::median(
      WT_cell_state_prediction_score
    ),
    median_cluster_prediction_score = stats::median(
      WT_wnn_cluster_prediction_score
    ),
    high_cell_state_score_percent = 100 * mean(
      WT_cell_state_mapping_score_band == "high_score"
    ),
    low_cell_state_score_percent = 100 * mean(
      WT_cell_state_mapping_score_band == "low_score"
    )
  )

readr::write_csv(
  inverse_mapping_summary,
  file.path(STEP5C_TABLE_DIR, "inverse_mapping_score_summary.csv")
)

inverse_mapping_summary
```

# 13. Calculate bidirectional annotation agreement

The forward percentage answers: what fraction of a current WT population maps
to a Gourab label? The inverse percentage answers: what fraction of that
Gourab label maps back to the same WT population? The conservative minimum is
the smaller percentage, while the geometric mean summarizes two-way strength
without allowing one direction to completely dominate.

```{r calculate-reciprocal-mapping-agreement}
forward_state_pairs <- wt_transfer_assignments |>
  dplyr::group_by(
    WT_cell_state = cell_state,
    Gourab_cell_state = gourab_d21_predicted_state
  ) |>
  dplyr::summarise(
    forward_nuclei = dplyr::n(),
    forward_median_score = stats::median(
      gourab_d21_prediction_score
    ),
    .groups = "drop_last"
  ) |>
  dplyr::mutate(
    forward_percent_within_WT_state = 100 *
      forward_nuclei / sum(forward_nuclei)
  ) |>
  dplyr::ungroup()

inverse_state_pairs <- source_inverse_assignments |>
  dplyr::group_by(
    Gourab_cell_state = source_cell_state,
    WT_cell_state = WT_cell_state_predicted_label
  ) |>
  dplyr::summarise(
    inverse_cells = dplyr::n(),
    inverse_median_score = stats::median(
      WT_cell_state_prediction_score
    ),
    .groups = "drop_last"
  ) |>
  dplyr::mutate(
    inverse_percent_within_Gourab_state = 100 *
      inverse_cells / sum(inverse_cells)
  ) |>
  dplyr::ungroup()

reciprocal_state_agreement <- dplyr::full_join(
  forward_state_pairs,
  inverse_state_pairs,
  by = c("WT_cell_state", "Gourab_cell_state")
) |>
  dplyr::mutate(
    dplyr::across(
      c(
        forward_nuclei,
        inverse_cells,
        forward_percent_within_WT_state,
        inverse_percent_within_Gourab_state
      ),
      ~ tidyr::replace_na(.x, 0)
    ),
    reciprocal_geometric_mean_percent = sqrt(
      forward_percent_within_WT_state *
        inverse_percent_within_Gourab_state
    ),
    reciprocal_conservative_min_percent = pmin(
      forward_percent_within_WT_state,
      inverse_percent_within_Gourab_state
    )
  ) |>
  dplyr::arrange(
    dplyr::desc(reciprocal_conservative_min_percent),
    dplyr::desc(reciprocal_geometric_mean_percent)
  )

forward_cluster_pairs <- wt_transfer_assignments |>
  dplyr::group_by(
    WT_wnn_cluster = wnn_cluster,
    Gourab_cell_state = gourab_d21_predicted_state
  ) |>
  dplyr::summarise(
    forward_nuclei = dplyr::n(),
    forward_median_score = stats::median(
      gourab_d21_prediction_score
    ),
    .groups = "drop_last"
  ) |>
  dplyr::mutate(
    forward_percent_within_WT_cluster = 100 *
      forward_nuclei / sum(forward_nuclei)
  ) |>
  dplyr::ungroup()

inverse_cluster_pairs <- source_inverse_assignments |>
  dplyr::group_by(
    Gourab_cell_state = source_cell_state,
    WT_wnn_cluster = WT_wnn_cluster_predicted_label
  ) |>
  dplyr::summarise(
    inverse_cells = dplyr::n(),
    inverse_median_score = stats::median(
      WT_wnn_cluster_prediction_score
    ),
    .groups = "drop_last"
  ) |>
  dplyr::mutate(
    inverse_percent_within_Gourab_state = 100 *
      inverse_cells / sum(inverse_cells)
  ) |>
  dplyr::ungroup()

reciprocal_cluster_agreement <- dplyr::full_join(
  forward_cluster_pairs,
  inverse_cluster_pairs,
  by = c("WT_wnn_cluster", "Gourab_cell_state")
) |>
  dplyr::mutate(
    dplyr::across(
      c(
        forward_nuclei,
        inverse_cells,
        forward_percent_within_WT_cluster,
        inverse_percent_within_Gourab_state
      ),
      ~ tidyr::replace_na(.x, 0)
    ),
    reciprocal_geometric_mean_percent = sqrt(
      forward_percent_within_WT_cluster *
        inverse_percent_within_Gourab_state
    ),
    reciprocal_conservative_min_percent = pmin(
      forward_percent_within_WT_cluster,
      inverse_percent_within_Gourab_state
    )
  ) |>
  dplyr::arrange(
    suppressWarnings(as.integer(WT_wnn_cluster)),
    dplyr::desc(reciprocal_conservative_min_percent)
  )

reciprocal_focus_cluster_agreement <- reciprocal_cluster_agreement |>
  dplyr::filter(WT_wnn_cluster %in% FOCUS_CLUSTERS)

readr::write_csv(
  reciprocal_state_agreement,
  file.path(
    STEP5C_TABLE_DIR,
    "reciprocal_WT_state_Gourab_state_agreement.csv"
  )
)

readr::write_csv(
  reciprocal_cluster_agreement,
  file.path(
    STEP5C_TABLE_DIR,
    "reciprocal_WT_cluster_Gourab_state_agreement.csv"
  )
)

readr::write_csv(
  reciprocal_focus_cluster_agreement,
  file.path(
    STEP5C_TABLE_DIR,
    "reciprocal_focus_cluster_Gourab_state_agreement.csv"
  )
)

reciprocal_focus_cluster_agreement |>
  print(n = Inf, width = Inf)
```

# 14. Plot inverse and reciprocal mapping evidence

```{r plot-inverse-and-reciprocal-mapping}
inverse_state_crosswalk <- source_inverse_assignments |>
  dplyr::count(
    source_cell_state,
    WT_cell_state_predicted_label,
    name = "cells"
  ) |>
  dplyr::group_by(source_cell_state) |>
  dplyr::mutate(
    percent_within_Gourab_state = 100 * cells / sum(cells)
  ) |>
  dplyr::ungroup()

inverse_cluster_crosswalk <- source_inverse_assignments |>
  dplyr::count(
    source_cell_state,
    WT_wnn_cluster_predicted_label,
    name = "cells"
  ) |>
  dplyr::group_by(source_cell_state) |>
  dplyr::mutate(
    percent_within_Gourab_state = 100 * cells / sum(cells)
  ) |>
  dplyr::ungroup()

inverse_source_label_score_summary <- source_inverse_assignments |>
  dplyr::group_by(source_cell_state) |>
  dplyr::summarise(
    median_cell_prediction_score = stats::median(
      WT_cell_state_prediction_score
    ),
    median_cell_prediction_entropy = stats::median(
      WT_cell_state_prediction_entropy
    ),
    median_cell_prediction_margin = stats::median(
      WT_cell_state_prediction_margin
    ),
    high_score_percent = 100 * mean(
      WT_cell_state_mapping_score_band == "high_score"
    ),
    low_score_percent = 100 * mean(
      WT_cell_state_mapping_score_band == "low_score"
    ),
    .groups = "drop"
  )

inverse_reference_state_count <- dplyr::n_distinct(
  wt_reference$wt_reference_cell_state
)

inverse_source_label_specificity <- inverse_state_crosswalk |>
  dplyr::group_by(source_cell_state) |>
  dplyr::summarise(
    source_cells = sum(cells),
    dominant_predicted_WT_state = WT_cell_state_predicted_label[
      which.max(percent_within_Gourab_state)
    ],
    dominant_predicted_WT_state_percent = max(
      percent_within_Gourab_state
    ),
    WT_states_at_least_5_percent = sum(
      percent_within_Gourab_state >= 5
    ),
    WT_states_at_least_10_percent = sum(
      percent_within_Gourab_state >= 10
    ),
    normalized_label_distribution_entropy = {
      label_probabilities <- percent_within_Gourab_state / 100
      label_probabilities <- label_probabilities[
        label_probabilities > 0
      ]

      if (inverse_reference_state_count <= 1L) {
        0
      } else {
        -sum(label_probabilities * log(label_probabilities)) /
          log(inverse_reference_state_count)
      }
    },
    .groups = "drop"
  ) |>
  dplyr::left_join(
    inverse_source_label_score_summary,
    by = "source_cell_state"
  ) |>
  dplyr::arrange(
    dplyr::desc(normalized_label_distribution_entropy),
    dominant_predicted_WT_state_percent
  )

readr::write_csv(
  inverse_state_crosswalk,
  file.path(
    STEP5C_TABLE_DIR,
    "Gourab_state_by_predicted_WT_state_crosswalk.csv"
  )
)

readr::write_csv(
  inverse_cluster_crosswalk,
  file.path(
    STEP5C_TABLE_DIR,
    "Gourab_state_by_predicted_WT_cluster_crosswalk.csv"
  )
)

readr::write_csv(
  inverse_source_label_specificity,
  file.path(
    STEP5C_TABLE_DIR,
    "inverse_source_label_specificity_summary.csv"
  )
)

source_umap_embeddings <- Seurat::Embeddings(
  source_reference,
  reduction = "umap.source"
)

if (!setequal(
  rownames(source_umap_embeddings),
  source_inverse_assignments$source_cell_id
)) {
  stop(
    "Source UMAP coordinates and inverse-mapping assignments do not match.",
    call. = FALSE
  )
}

source_umap_embeddings <- source_umap_embeddings[
  source_inverse_assignments$source_cell_id,
  ,
  drop = FALSE
]

inverse_umap_data <- source_inverse_assignments |>
  dplyr::mutate(
    sourceUMAP_1 = source_umap_embeddings[source_cell_id, 1],
    sourceUMAP_2 = source_umap_embeddings[source_cell_id, 2]
  )

p_inverse_observed <- ggplot2::ggplot(
  inverse_umap_data,
  ggplot2::aes(x = sourceUMAP_1, y = sourceUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = source_cell_state),
    size = 0.25,
    alpha = 0.75
  ) +
  ggplot2::scale_color_manual(values = source_state_colors) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Observed Gourab d21 labels",
    x = "sourceUMAP_1",
    y = "sourceUMAP_2",
    color = "Gourab state"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_inverse_predicted <- ggplot2::ggplot(
  inverse_umap_data,
  ggplot2::aes(x = sourceUMAP_1, y = sourceUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = WT_cell_state_predicted_label),
    size = 0.25,
    alpha = 0.75
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Reviewed WT states transferred to d21",
    subtitle = "Same source RNA embedding",
    x = "sourceUMAP_1",
    y = "sourceUMAP_2",
    color = "Predicted WT state"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_inverse_score <- ggplot2::ggplot(
  inverse_umap_data,
  ggplot2::aes(x = sourceUMAP_1, y = sourceUMAP_2)
) +
  ggplot2::geom_point(
    ggplot2::aes(color = WT_cell_state_prediction_score),
    size = 0.25,
    alpha = 0.80
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    limits = c(0, 1)
  ) +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Inverse mapping score",
    subtitle = "Algorithmic support, not annotation confidence",
    x = "sourceUMAP_1",
    y = "sourceUMAP_2",
    color = "Maximum score"
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank()
  )

p_inverse_overview <- p_inverse_observed +
  p_inverse_predicted +
  p_inverse_score +
  patchwork::plot_layout(ncol = 3)

print(p_inverse_overview)

save_pdf_plot(
  file.path(
    STEP5C_PLOT_DIR,
    "05_inverse_Gourab_to_WT_mapping_overview.pdf"
  ),
  p_inverse_overview,
  width = 22,
  height = 7
)

p_inverse_state <- ggplot2::ggplot(
  inverse_state_crosswalk,
  ggplot2::aes(
    x = WT_cell_state_predicted_label,
    y = source_cell_state,
    fill = percent_within_Gourab_state
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.35) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = ifelse(
        percent_within_Gourab_state >= 3,
        paste0(cells, "\n", round(percent_within_Gourab_state, 1), "%"),
        ""
      )
    ),
    size = 2.6
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "C",
    limits = c(0, 100)
  ) +
  ggplot2::labs(
    title = "WT annotations transferred onto Gourab d21 cells",
    subtitle = paste0(
      "Rows sum to 100%; WT reference includes only high/moderate-confidence ",
      "annotations"
    ),
    x = "Predicted reviewed WT cell state",
    y = "Observed Gourab d21 label",
    fill = "% of Gourab label"
  ) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_inverse_state)

save_pdf_plot(
  file.path(
    STEP5C_PLOT_DIR,
    "06_inverse_Gourab_to_WT_state_crosswalk.pdf"
  ),
  p_inverse_state,
  width = 18,
  height = 8
)

reciprocal_focus_plot_data <- reciprocal_focus_cluster_agreement |>
  dplyr::mutate(
    WT_wnn_cluster = factor(
      WT_wnn_cluster,
      levels = rev(FOCUS_CLUSTERS)
    ),
    Gourab_cell_state = factor(
      Gourab_cell_state,
      levels = source_state_levels
    )
  )

p_reciprocal_focus <- ggplot2::ggplot(
  reciprocal_focus_plot_data,
  ggplot2::aes(
    x = Gourab_cell_state,
    y = WT_wnn_cluster,
    fill = reciprocal_conservative_min_percent
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.35) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = ifelse(
        forward_percent_within_WT_cluster >= 3 |
          inverse_percent_within_Gourab_state >= 3,
        paste0(
          "F ", round(forward_percent_within_WT_cluster),
          " / R ", round(inverse_percent_within_Gourab_state)
        ),
        ""
      )
    ),
    size = 2.7
  ) +
  ggplot2::scale_fill_viridis_c(
    option = "C",
    limits = c(0, 100)
  ) +
  ggplot2::labs(
    title = "Bidirectional agreement for focused WT clusters",
    subtitle = paste0(
      "F = % of WT cluster mapping to source label; ",
      "R = % of source label mapping back"
    ),
    x = "Gourab d21 cell state",
    y = "WT WNN cluster",
    fill = "Conservative\nminimum (%)"
  ) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_reciprocal_focus)

save_pdf_plot(
  file.path(
    STEP5C_PLOT_DIR,
    "07_reciprocal_focused_cluster_agreement.pdf"
  ),
  p_reciprocal_focus,
  width = 15,
  height = 7
)
```

# 15. Harmony co-embedding sensitivity analysis

This section is intentionally secondary. Harmony adjusts the combined PCA
embedding by platform/study label. It does not replace expression values,
restore cytoplasmic transcripts missing from nuclei, or justify direct
cross-platform abundance comparisons.

If the `harmony` package is unavailable, the primary reference mapping remains
complete and this section is skipped with an explicit audit record.

```{r harmony-coembedding-sensitivity}
harmony_available <- requireNamespace("harmony", quietly = TRUE)
harmony_completed <- FALSE
harmony_error <- NA_character_

if (harmony_available) {
  harmony_result <- tryCatch(
    {
      source_for_harmony <- Seurat::CreateSeuratObject(
        counts = SeuratObject::LayerData(
          source_reference[["RNA"]],
          layer = "counts"
        ),
        project = "Gourab_d21_scRNA",
        meta.data = source_reference[[]],
        min.cells = 0,
        min.features = 0
      )

      query_for_harmony <- Seurat::CreateSeuratObject(
        counts = SeuratObject::LayerData(
          wt_query[["RNA"]],
          layer = "counts"
        ),
        project = "WT_snMultiome_RNA",
        meta.data = wt_query[[]],
        min.cells = 0,
        min.features = 0
      )

      source_for_harmony <- Seurat::RenameCells(
        source_for_harmony,
        add.cell.id = "GourabD21"
      )

      query_for_harmony <- Seurat::RenameCells(
        query_for_harmony,
        add.cell.id = "WTMultiome"
      )

      combined_rna <- merge(
        source_for_harmony,
        y = query_for_harmony,
        merge.data = FALSE
      )

      combined_rna <- SeuratObject::JoinLayers(
        combined_rna,
        assay = "RNA"
      )

      combined_rna <- Seurat::NormalizeData(
        combined_rna,
        normalization.method = "LogNormalize",
        scale.factor = 10000,
        verbose = FALSE
      )

      combined_rna <- Seurat::FindVariableFeatures(
        combined_rna,
        selection.method = "vst",
        nfeatures = SOURCE_VARIABLE_FEATURES,
        verbose = FALSE
      )

      combined_rna <- Seurat::ScaleData(
        combined_rna,
        features = Seurat::VariableFeatures(combined_rna),
        verbose = FALSE
      )

      combined_rna <- Seurat::RunPCA(
        combined_rna,
        features = Seurat::VariableFeatures(combined_rna),
        npcs = SOURCE_PCS,
        verbose = FALSE
      )

      combined_rna <- Seurat::RunUMAP(
        combined_rna,
        reduction = "pca",
        dims = HARMONY_DIMS,
        reduction.name = "umap.unintegrated",
        reduction.key = "unintegratedUMAP_",
        seed.use = 12345,
        verbose = FALSE
      )

      combined_rna <- harmony::RunHarmony(
        object = combined_rna,
        group.by.vars = "dataset_platform",
        reduction.use = "pca",
        dims.use = HARMONY_DIMS,
        reduction.save = "harmony",
        plot_convergence = FALSE,
        verbose = FALSE
      )

      combined_rna <- Seurat::RunUMAP(
        combined_rna,
        reduction = "harmony",
        dims = HARMONY_DIMS,
        reduction.name = "umap.harmony",
        reduction.key = "harmonyUMAP_",
        seed.use = 12345,
        verbose = FALSE
      )

      platform_colors <- c(
        "Gourab d21 scRNA" = "#0072B2",
        "WT snMultiome RNA" = "#D55E00"
      )

      p_unintegrated_platform <- Seurat::DimPlot(
        combined_rna,
        reduction = "umap.unintegrated",
        group.by = "dataset_platform",
        cols = platform_colors,
        raster = TRUE
      ) +
        ggplot2::labs(
          title = "Uncorrected shared RNA embedding",
          subtitle = "Platform separation is expected"
        )

      p_harmony_platform <- Seurat::DimPlot(
        combined_rna,
        reduction = "umap.harmony",
        group.by = "dataset_platform",
        cols = platform_colors,
        raster = TRUE
      ) +
        ggplot2::labs(
          title = "Harmony sensitivity embedding",
          subtitle = "Corrected by dataset/platform only"
        )

      p_harmony_state <- Seurat::DimPlot(
        combined_rna,
        reduction = "umap.harmony",
        group.by = "mapping_state",
        cols = source_state_colors,
        label = TRUE,
        repel = TRUE,
        raster = TRUE
      ) +
        ggplot2::labs(
          title = "Harmony embedding by reference/mapped identity",
          subtitle = paste0(
            "Reference cells use observed labels; WT cells use ",
            "transferred labels"
          )
        )

      p_harmony_overview <- p_unintegrated_platform +
        p_harmony_platform +
        p_harmony_state +
        patchwork::plot_layout(ncol = 3)

      print(p_harmony_overview)

      save_pdf_plot(
        file.path(
          STEP5C_PLOT_DIR,
          "08_Harmony_cross_platform_sensitivity.pdf"
        ),
        p_harmony_overview,
        width = 23,
        height = 7
      )

      harmony_embeddings <- Seurat::Embeddings(
        combined_rna,
        reduction = "umap.harmony"
      )

      harmony_embedding_table <- combined_rna[[]] |>
        tibble::rownames_to_column("cell_id") |>
        dplyr::mutate(
          harmonyUMAP_1 = harmony_embeddings[cell_id, 1],
          harmonyUMAP_2 = harmony_embeddings[cell_id, 2]
        ) |>
        dplyr::select(
          cell_id,
          dataset_platform,
          source_or_query,
          mapping_state,
          source_diet,
          sample_id,
          diet,
          wnn_cluster,
          cell_state,
          gourab_d21_prediction_score,
          harmonyUMAP_1,
          harmonyUMAP_2
        )

      readr::write_csv(
        harmony_embedding_table,
        file.path(
          STEP5C_TABLE_DIR,
          "Harmony_cross_platform_embedding.csv"
        )
      )

      list(
        completed = TRUE,
        error = NA_character_
      )
    },
    error = function(error_condition) {
      current_harmony_error <- conditionMessage(error_condition)
      warning(
        "Harmony sensitivity analysis failed: ",
        current_harmony_error,
        call. = FALSE
      )
      list(
        completed = FALSE,
        error = current_harmony_error
      )
    }
  )

  harmony_completed <- isTRUE(harmony_result$completed)
  harmony_error <- harmony_result$error
}

harmony_audit <- tibble::tibble(
  harmony_package_available = harmony_available,
  harmony_completed = harmony_completed,
  harmony_error = harmony_error
)

readr::write_csv(
  harmony_audit,
  file.path(STEP5C_TABLE_DIR, "Harmony_sensitivity_audit.csv")
)

harmony_audit
```

# 16. Save completion records

No current WT annotation is changed in this notebook. The transferred-label
table and focused crosswalk are inputs for the subsequent manual crypt-axis
review.

```{r save-completion-records}
step5c_summary <- tibble::tibble(
  analysis_version = ANALYSIS_VERSION,
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  source_CON_filtered_cells = ncol(con_counts),
  source_HFD_filtered_cells = ncol(hfd_counts),
  source_annotated_cells = ncol(source_reference),
  source_named_states = dplyr::n_distinct(
    source_reference$source_cell_state
  ),
  WT_query_nuclei = ncol(wt_query),
  shared_reference_query_genes = length(
    shared_reference_query_genes
  ),
  median_transfer_score = stats::median(
    wt_query$gourab_d21_prediction_score
  ),
  high_score_WT_percent = 100 * mean(
    wt_query$gourab_d21_mapping_score_band == "high_score"
  ),
  WT_inverse_reference_nuclei = ncol(wt_reference),
  inverse_source_cells = nrow(source_inverse_assignments),
  inverse_median_cell_state_score = stats::median(
    source_inverse_assignments$WT_cell_state_prediction_score
  ),
  inverse_high_score_source_percent = 100 * mean(
    source_inverse_assignments$WT_cell_state_mapping_score_band ==
      "high_score"
  ),
  presto_version = as.character(utils::packageVersion("presto")),
  harmony_completed = harmony_completed,
  annotation_overrides_applied = FALSE,
  biological_diet_inference_performed = FALSE
)

readr::write_csv(
  step5c_summary,
  file.path(STEP5C_DIR, "step5c_summary.csv")
)

writeLines(
  c(
    paste("Step 5C completed:", Sys.time()),
    paste("Analysis version:", ANALYSIS_VERSION),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste("Source annotated cells:", ncol(source_reference)),
    paste(
      "Source named states:",
      dplyr::n_distinct(source_reference$source_cell_state)
    ),
    paste("WT query nuclei:", ncol(wt_query)),
    paste(
      "Shared reference-query genes:",
      length(shared_reference_query_genes)
    ),
    paste(
      "Median transfer score:",
      signif(stats::median(
        wt_query$gourab_d21_prediction_score
      ), 4)
    ),
    paste("presto version:", utils::packageVersion("presto")),
    paste("WT inverse-reference nuclei:", ncol(wt_reference)),
    paste(
      "Inverse median cell-state score:",
      signif(stats::median(
        source_inverse_assignments$WT_cell_state_prediction_score
      ), 4)
    ),
    paste("Harmony sensitivity completed:", harmony_completed),
    "WT annotation overrides applied: FALSE",
    "Biological diet inference performed: FALSE",
    paste("Source reference object:", SOURCE_REFERENCE_PATH),
    paste("WT transfer assignments:", WT_TRANSFER_ASSIGNMENT_PATH),
    paste("Source inverse assignments:", SOURCE_INVERSE_ASSIGNMENT_PATH)
  ),
  con = file.path(STEP5C_DIR, "STEP_05C_COMPLETE.txt")
)

writeLines(
  c(
    "Step 5C external-reference mapping is ready for manual review.",
    paste(
      "Review focused crosswalk:",
      file.path(
        STEP5C_TABLE_DIR,
        "WT_focus_clusters_0_6_8_and_secretory_crosswalk.csv"
      )
    ),
    paste(
      "Review mapping overview:",
      file.path(
        STEP5C_PLOT_DIR,
        "03_WT_WNN_reference_mapping_overview.pdf"
      )
    ),
    paste(
      "Review inverse mapping overview:",
      file.path(
        STEP5C_PLOT_DIR,
        "05_inverse_Gourab_to_WT_mapping_overview.pdf"
      )
    ),
    paste(
      "Review reciprocal focus table:",
      file.path(
        STEP5C_TABLE_DIR,
        "reciprocal_focus_cluster_Gourab_state_agreement.csv"
      )
    ),
    "Do not overwrite WT annotations without a separate reviewed decision."
  ),
  con = file.path(STEP5C_DIR, "STEP_05C_READY_FOR_REVIEW.txt")
)

step5c_summary
```

# 17. Session information

```{r session-information}
sessionInfo()
```
