---
title: "Step 5D - Symmetric pseudobulk validation of Gourab d21 and WT epithelial annotations"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# Purpose

Step 5C established that one-way transfer of Gourab's labels onto WT nuclei is
useful as descriptive evidence, but the inverse WT-to-Gourab transfer collapsed
onto the WT Reg4-positive DCS reference. That collapse makes the inverse labels
unsuitable for biological interpretation.

This notebook therefore replaces reciprocal label transfer with a symmetric
comparison that does not train either dataset on the other. It:

1. records a formal QC disposition for the Step 5C inverse transfer;
2. retains Gourab's original integrated clusters inside each named state;
3. constructs cluster-by-diet RNA pseudobulks independently in each dataset;
4. compares relative marker programs after gene-wise standardization within
   each dataset and diet;
5. evaluates similarity in both directions, including mutual-best matches;
6. bootstraps genes to measure match stability; and
7. reports focused evidence for WT crypt and secretory clusters.

This is annotation validation, not a diet-effect test. Both datasets contain
one pooled library per diet, so diet is used as a sensitivity stratum and is
balanced in the primary identity comparison. No WT labels are overwritten.

A practical SOL allocation is **8 CPU cores, 64 GiB memory, 6 hours, R 4.4.2,
no GPU**.

# 1. Setup

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 12,
  fig.height = 8
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(readr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(scales)
})

if (!requireNamespace("presto", quietly = TRUE)) {
  stop(
    "Package 'presto' is required so the integrated-cluster marker audit is ",
    "tractable. Run setup/install_packages.R in a clean R session first.",
    call. = FALSE
  )
}

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 5D requires MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}

ggplot2::theme_set(theme_multiome())
future::plan(future::sequential)
set.seed(12345)
```

# 2. Inputs, outputs, and review parameters

```{r define-inputs-and-outputs}
ANALYSIS_VERSION <- "05d_symmetric_pseudobulk_v1"

STEP5C_DIR <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "gourab_d21_rna_reference_mapping"
)

STEP5C_TABLE_DIR <- file.path(STEP5C_DIR, "tables")

STEP5D_DIR <- file.path(
  STEP5C_DIR,
  "05d_symmetric_pseudobulk_validation"
)

STEP5D_TABLE_DIR <- file.path(STEP5D_DIR, "tables")
STEP5D_PLOT_DIR <- file.path(STEP5D_DIR, "plots")

STEP6_OBJECT_PATH <- file.path(
  OUTPUT_DIR,
  "06_annotation",
  "multiome_annotated.rds"
)

WT_MARKER_PATH <- file.path(
  OUTPUT_DIR,
  "06_annotation",
  "markers",
  "RNA_all_positive_markers.csv"
)

SOURCE_REFERENCE_PATH <- file.path(
  STEP5C_DIR,
  "Gourab_d21_annotated_scRNA_reference.rds"
)

INVERSE_ASSIGNMENT_PATH <- file.path(
  STEP5C_DIR,
  "Gourab_d21_WT_inverse_label_transfer_assignments.csv"
)

for (output_directory in c(
  STEP5D_DIR,
  STEP5D_TABLE_DIR,
  STEP5D_PLOT_DIR
)) {
  dir.create(output_directory, recursive = TRUE, showWarnings = FALSE)
}

for (input_path in c(
  STEP6_OBJECT_PATH,
  WT_MARKER_PATH,
  SOURCE_REFERENCE_PATH,
  INVERSE_ASSIGNMENT_PATH
)) {
  assert_file(input_path)
}

MIN_CELLS_PER_CLUSTER_DIET <- 10L
TOP_MARKERS_PER_CLUSTER <- 100L
MIN_PRIMARY_FEATURES <- 100L
GENE_BOOTSTRAP_ITERATIONS <- 200L
BOOTSTRAP_SEED <- 20260921L

PRIMARY_WT_CONFIDENCE <- c("high", "moderate")
SOURCE_EPITHELIAL_STATES <- c(
  "Stem", "TAC", "DSC", "Goblet", "EC", "EEC", "Tuft"
)

CRYPT_FOCUS_CLUSTERS <- c("0", "6", "8")
SECRETORY_FOCUS_CLUSTERS <- c("4", "5", "7", "9", "13", "18")
FOCUS_WT_CLUSTERS <- c(CRYPT_FOCUS_CLUSTERS, SECRETORY_FOCUS_CLUSTERS)

CELL_CYCLE_GENES <- unique(c(
  Seurat::cc.genes.updated.2019$s.genes,
  Seurat::cc.genes.updated.2019$g2m.genes
))

TECHNICAL_GENE_PATTERN <- paste0(
  "^(mt-|Rpl|Rps|Gm[0-9]+|Mir[0-9]+|Snor|Hist[0-9]|Hsp[0-9])",
  "|Rik$"
)

CURATED_PROGRAMS <- list(
  ISC_crypt = c(
    "Lgr5", "Ascl2", "Smoc2", "Axin2", "Olfm4", "Sox9",
    "Prom1", "Hopx", "Tert", "Lrig1"
  ),
  Cycling = c(
    "Mki67", "Top2a", "Cenpf", "Ube2c", "Cdk1", "Ccnb1",
    "Tpx2", "Nusap1", "Kif11", "Birc5"
  ),
  Crypt_progenitor = c(
    "Dmbt1", "Cftr", "Htr4", "Kcnq1", "Slc12a2", "Sema5a",
    "Hmga2", "Clca3a2", "Kitl", "Hmgcs2"
  ),
  Reg4_DSC = c(
    "Reg4", "Spink4", "Kit", "Agr2", "Klk1", "Hepacam2",
    "Slc12a8", "Sytl2", "Car8", "Plcb4"
  ),
  Goblet = c(
    "Muc2", "Spdef", "Fcgbp", "Clca1", "Zg16", "Agr2",
    "Tff3", "Scin", "Bcas1", "Clec2h"
  ),
  Absorptive = c(
    "Krt20", "Car4", "Aqp8", "Slc26a3", "Mxd1", "Lama3",
    "Tgm3", "Trpm6", "Slc9a2", "Muc3"
  ),
  Enteroendocrine = c(
    "Chga", "Chgb", "Neurod1", "Rfx6", "Cpe", "Pcsk1",
    "Pyy", "Gcg", "Insl5", "Tph1"
  ),
  Enterochromaffin = c(
    "Tph1", "Tac1", "Ddc", "Slc18a1", "Lmx1a", "Chga", "Chgb"
  ),
  Distal_L = c(
    "Gcg", "Pyy", "Insl5", "Pcsk1", "Neurod1", "Chga", "Chgb"
  ),
  Tuft = c(
    "Pou2f3", "Dclk1", "Trpm5", "Gnat3", "Sh2d6", "Rgs13",
    "Avil", "Ltc4s"
  )
)

EXPECTED_WT_FAMILIES <- list(
  Stem = c("0", "6", "8"),
  TAC = c("0", "18"),
  DSC = c("4", "7", "9", "13", "18"),
  Goblet = c("4", "5", "7", "13"),
  EC = c("1", "3", "19"),
  EEC = c("14", "16"),
  Tuft = c("20")
)

save_pdf_plot <- function(path, plot_object, width, height) {
  grDevices::pdf(
    file = path,
    width = width,
    height = height,
    useDingbats = FALSE
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot_object)
  invisible(path)
}
```

# 3. Formal disposition of the Step 5C inverse transfer

This chunk does not retry the transfer. It quantifies whether the previous
inverse result is sufficiently diverse and lineage-consistent to be used as
annotation evidence.

```{r audit-step5c-inverse-transfer}
inverse_assignments <- suppressMessages(
  readr::read_csv(INVERSE_ASSIGNMENT_PATH, show_col_types = FALSE)
)

required_inverse_columns <- c(
  "source_cell_state",
  "WT_cell_state_predicted_label",
  "WT_wnn_cluster_predicted_label"
)

missing_inverse_columns <- setdiff(
  required_inverse_columns,
  colnames(inverse_assignments)
)

if (length(missing_inverse_columns) > 0L) {
  stop(
    "Step 5C inverse assignments are missing column(s): ",
    paste(missing_inverse_columns, collapse = ", "),
    call. = FALSE
  )
}

inverse_cluster_distribution <- inverse_assignments |>
  dplyr::count(
    WT_wnn_cluster_predicted_label,
    name = "cells",
    sort = TRUE
  ) |>
  dplyr::mutate(percent = 100 * cells / sum(cells))

inverse_state_distribution <- inverse_assignments |>
  dplyr::count(
    WT_cell_state_predicted_label,
    name = "cells",
    sort = TRUE
  ) |>
  dplyr::mutate(percent = 100 * cells / sum(cells))

inverse_expected_family_recovery <- purrr::imap_dfr(
  EXPECTED_WT_FAMILIES,
  function(expected_clusters, source_state) {
    state_data <- inverse_assignments |>
      dplyr::filter(source_cell_state == source_state)

    tibble::tibble(
      source_cell_state = source_state,
      source_cells = nrow(state_data),
      expected_WT_clusters = paste(expected_clusters, collapse = ";"),
      expected_family_cells = sum(
        as.character(state_data$WT_wnn_cluster_predicted_label) %in%
          expected_clusters
      ),
      expected_family_percent = if (nrow(state_data) == 0L) {
        NA_real_
      } else {
        100 * expected_family_cells / nrow(state_data)
      },
      dominant_predicted_cluster = if (nrow(state_data) == 0L) {
        NA_character_
      } else {
        names(sort(table(
          as.character(state_data$WT_wnn_cluster_predicted_label)
        ), decreasing = TRUE))[[1]]
      }
    )
  }
)

dominant_cluster_percent <- inverse_cluster_distribution$percent[[1]]
predicted_cluster_coverage <- dplyr::n_distinct(
  inverse_assignments$WT_wnn_cluster_predicted_label
)
predicted_state_coverage <- dplyr::n_distinct(
  inverse_assignments$WT_cell_state_predicted_label
)
median_expected_family_recovery <- stats::median(
  inverse_expected_family_recovery$expected_family_percent,
  na.rm = TRUE
)

inverse_failed <- (
  dominant_cluster_percent >= 50 ||
    predicted_cluster_coverage < 8 ||
    predicted_state_coverage < 8 ||
    median_expected_family_recovery < 25
)

inverse_qc_disposition <- tibble::tibble(
  metric = c(
    "dominant_predicted_WT_cluster_percent",
    "predicted_WT_cluster_coverage",
    "predicted_WT_state_coverage",
    "median_expected_family_recovery_percent"
  ),
  observed = c(
    dominant_cluster_percent,
    predicted_cluster_coverage,
    predicted_state_coverage,
    median_expected_family_recovery
  ),
  failure_rule = c(">= 50", "< 8", "< 8", "< 25"),
  failed = c(
    dominant_cluster_percent >= 50,
    predicted_cluster_coverage < 8,
    predicted_state_coverage < 8,
    median_expected_family_recovery < 25
  ),
  overall_disposition = ifelse(
    inverse_failed,
    "FAILED_FOR_BIOLOGICAL_INTERPRETATION",
    "PASSED_DESCRIPTIVE_QC"
  )
)

readr::write_csv(
  inverse_cluster_distribution,
  file.path(STEP5D_TABLE_DIR, "step5c_inverse_cluster_distribution.csv")
)

readr::write_csv(
  inverse_state_distribution,
  file.path(STEP5D_TABLE_DIR, "step5c_inverse_state_distribution.csv")
)

readr::write_csv(
  inverse_expected_family_recovery,
  file.path(
    STEP5D_TABLE_DIR,
    "step5c_inverse_expected_family_recovery.csv"
  )
)

readr::write_csv(
  inverse_qc_disposition,
  file.path(STEP5D_TABLE_DIR, "step5c_inverse_qc_disposition.csv")
)

inverse_qc_disposition
```

# 4. Load and validate both independently annotated datasets

```{r load-and-validate-objects}
source_reference <- readRDS(SOURCE_REFERENCE_PATH)
multiome <- readRDS(STEP6_OBJECT_PATH)

source_metadata <- source_reference[[]]
wt_metadata <- multiome[[]]

required_source_metadata <- c(
  "source_diet",
  "source_cell_state",
  "source_integrated_cluster"
)

required_wt_metadata <- c(
  "diet",
  "wnn_cluster",
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence"
)

missing_source_metadata <- setdiff(
  required_source_metadata,
  colnames(source_metadata)
)

missing_wt_metadata <- setdiff(
  required_wt_metadata,
  colnames(wt_metadata)
)

if (length(missing_source_metadata) > 0L) {
  stop(
    "The source reference is missing metadata column(s): ",
    paste(missing_source_metadata, collapse = ", "),
    call. = FALSE
  )
}

if (length(missing_wt_metadata) > 0L) {
  stop(
    "The WT Step 6 object is missing metadata column(s): ",
    paste(missing_wt_metadata, collapse = ", "),
    call. = FALSE
  )
}

if (!"RNA" %in% names(source_reference) || !"RNA" %in% names(multiome)) {
  stop("Both objects must contain an RNA assay.", call. = FALSE)
}

if (!"counts" %in% SeuratObject::Layers(source_reference[["RNA"]])) {
  stop("The source RNA assay lacks a joined counts layer.", call. = FALSE)
}

if (!"counts" %in% SeuratObject::Layers(multiome[["RNA"]])) {
  stop("The WT RNA assay lacks a joined counts layer.", call. = FALSE)
}

source_counts <- SeuratObject::LayerData(
  source_reference[["RNA"]],
  layer = "counts"
)

wt_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)

if (!setequal(colnames(source_counts), rownames(source_metadata))) {
  stop("Source counts and metadata do not contain the same cells.")
}

if (!setequal(colnames(wt_counts), rownames(wt_metadata))) {
  stop("WT counts and metadata do not contain the same cells.")
}

source_counts <- source_counts[, rownames(source_metadata), drop = FALSE]
wt_counts <- wt_counts[, rownames(wt_metadata), drop = FALSE]

source_metadata <- source_metadata |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::mutate(
    diet = as.character(source_diet),
    source_cell_state = as.character(source_cell_state),
    source_integrated_cluster = as.character(source_integrated_cluster),
    source_cluster_id = paste0("G", source_integrated_cluster),
    source_cluster_label = paste0(
      source_cluster_id,
      " (", source_cell_state, ")"
    )
  )

wt_metadata <- wt_metadata |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::mutate(
    diet = as.character(diet),
    wnn_cluster = as.character(wnn_cluster),
    broad_compartment = as.character(broad_compartment),
    cell_state = as.character(cell_state),
    cell_state_full = as.character(cell_state_full),
    annotation_confidence = tolower(as.character(annotation_confidence)),
    wt_cluster_id = paste0("W", wnn_cluster),
    wt_cluster_label = paste0("W", wnn_cluster, " (", cell_state, ")")
  )

if (!setequal(unique(source_metadata$diet), c("CON", "HFD"))) {
  stop("Expected exactly CON and HFD in the source reference.")
}

if (!setequal(unique(wt_metadata$diet), c("CON", "HFD"))) {
  stop("Expected exactly CON and HFD in the WT object.")
}

source_keep <- source_metadata$source_cell_state %in%
  SOURCE_EPITHELIAL_STATES

wt_keep <- (
  wt_metadata$broad_compartment == "epithelial" &
    wt_metadata$annotation_confidence %in% PRIMARY_WT_CONFIDENCE
)

source_metadata_epithelial <- source_metadata[source_keep, , drop = FALSE]
wt_metadata_epithelial <- wt_metadata[wt_keep, , drop = FALSE]

source_counts_epithelial <- source_counts[
  , source_keep, drop = FALSE
]

wt_counts_epithelial <- wt_counts[, wt_keep, drop = FALSE]

dataset_cell_summary <- dplyr::bind_rows(
  source_metadata_epithelial |>
    dplyr::count(
      dataset = "Gourab_d21_scRNA",
      diet,
      cluster = source_cluster_label,
      state = source_cell_state,
      name = "cells"
    ),
  wt_metadata_epithelial |>
    dplyr::count(
      dataset = "WT_snMultiome_RNA",
      diet,
      cluster = wt_cluster_label,
      state = cell_state,
      name = "cells"
    )
)

readr::write_csv(
  dataset_cell_summary,
  file.path(STEP5D_TABLE_DIR, "epithelial_cluster_cell_counts_by_diet.csv")
)

dataset_cell_summary
```

# 5. Audit Gourab's original integrated clusters

The named Gourab labels aggregate multiple integrated clusters. Marker testing
at the integrated-cluster level reveals whether a named state is internally
coherent or spans several biological subprograms.

```{r source-integrated-cluster-audit}
source_epithelial <- subset(
  source_reference,
  cells = source_metadata_epithelial$cell_barcode
)

Seurat::Idents(source_epithelial) <- "source_integrated_cluster"

source_integrated_markers <- Seurat::FindAllMarkers(
  source_epithelial,
  assay = "RNA",
  only.pos = TRUE,
  test.use = "wilcox",
  min.pct = 0.10,
  logfc.threshold = 0.25,
  return.thresh = 0.05,
  verbose = FALSE
)

if (nrow(source_integrated_markers) == 0L) {
  stop("No Gourab integrated-cluster markers passed the thresholds.")
}

source_fc_column <- intersect(
  c("avg_log2FC", "avg_logFC"),
  colnames(source_integrated_markers)
)

if (length(source_fc_column) != 1L) {
  stop("Could not identify the source marker fold-change column.")
}

source_integrated_markers$effect_size <-
  source_integrated_markers[[source_fc_column]]

source_integrated_markers$pct_delta <- with(
  source_integrated_markers,
  pct.1 - pct.2
)

source_cluster_annotation <- source_metadata_epithelial |>
  dplyr::distinct(
    source_integrated_cluster,
    source_cluster_id,
    source_cluster_label,
    source_cell_state
  )

source_integrated_markers <- source_integrated_markers |>
  dplyr::mutate(
    cluster = as.character(cluster),
    gene = as.character(gene)
  ) |>
  dplyr::left_join(
    source_cluster_annotation,
    by = c("cluster" = "source_integrated_cluster")
  )

source_integrated_top_markers <- source_integrated_markers |>
  dplyr::filter(!grepl(TECHNICAL_GENE_PATTERN, gene)) |>
  dplyr::group_by(cluster) |>
  dplyr::arrange(
    p_val_adj,
    dplyr::desc(pct_delta),
    dplyr::desc(effect_size),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = TOP_MARKERS_PER_CLUSTER) |>
  dplyr::ungroup()

readr::write_csv(
  source_integrated_markers,
  file.path(
    STEP5D_TABLE_DIR,
    "source_integrated_cluster_all_positive_markers.csv"
  )
)

readr::write_csv(
  source_integrated_top_markers,
  file.path(
    STEP5D_TABLE_DIR,
    "source_integrated_cluster_top_markers.csv"
  )
)

source_integrated_top_markers |>
  dplyr::select(
    source_cluster_label,
    gene,
    effect_size,
    pct.1,
    pct.2,
    p_val_adj
  )
```

# 6. Build cluster-by-diet pseudobulks

```{r build-cluster-diet-pseudobulks}
aggregate_sparse_counts <- function(counts, groups) {
  groups <- as.character(groups)
  group_levels <- sort(unique(groups))
  group_factor <- factor(groups, levels = group_levels)
  design <- Matrix::sparse.model.matrix(~ 0 + group_factor)
  colnames(design) <- group_levels
  counts %*% design
}

log2_cpm <- function(pseudobulk_counts) {
  library_sizes <- Matrix::colSums(pseudobulk_counts)

  if (any(library_sizes <= 0)) {
    stop("A pseudobulk has a non-positive library size.")
  }

  cpm <- t(t(pseudobulk_counts) / library_sizes) * 1e6
  log2(cpm + 1)
}

source_metadata_epithelial <- source_metadata_epithelial |>
  dplyr::mutate(
    pseudobulk_id = paste(source_cluster_id, diet, sep = "__")
  )

wt_metadata_epithelial <- wt_metadata_epithelial |>
  dplyr::mutate(
    pseudobulk_id = paste(wt_cluster_id, diet, sep = "__")
  )

source_pseudobulk_counts <- aggregate_sparse_counts(
  source_counts_epithelial,
  source_metadata_epithelial$pseudobulk_id
)

wt_pseudobulk_counts <- aggregate_sparse_counts(
  wt_counts_epithelial,
  wt_metadata_epithelial$pseudobulk_id
)

source_pb_metadata <- source_metadata_epithelial |>
  dplyr::count(
    pseudobulk_id,
    source_cluster_id,
    source_cluster_label,
    source_integrated_cluster,
    source_cell_state,
    diet,
    name = "cells"
  ) |>
  dplyr::arrange(match(pseudobulk_id, colnames(source_pseudobulk_counts)))

wt_pb_metadata <- wt_metadata_epithelial |>
  dplyr::count(
    pseudobulk_id,
    wt_cluster_id,
    wt_cluster_label,
    wnn_cluster,
    cell_state,
    cell_state_full,
    annotation_confidence,
    diet,
    name = "cells"
  ) |>
  dplyr::arrange(match(pseudobulk_id, colnames(wt_pseudobulk_counts)))

if (!identical(
  source_pb_metadata$pseudobulk_id,
  colnames(source_pseudobulk_counts)
)) {
  stop("Source pseudobulk metadata are not aligned with counts.")
}

if (!identical(
  wt_pb_metadata$pseudobulk_id,
  colnames(wt_pseudobulk_counts)
)) {
  stop("WT pseudobulk metadata are not aligned with counts.")
}

source_pb_keep <- source_pb_metadata$cells >= MIN_CELLS_PER_CLUSTER_DIET
wt_pb_keep <- wt_pb_metadata$cells >= MIN_CELLS_PER_CLUSTER_DIET

source_pseudobulk_counts <- source_pseudobulk_counts[
  , source_pb_keep, drop = FALSE
]
source_pb_metadata <- source_pb_metadata[source_pb_keep, , drop = FALSE]

wt_pseudobulk_counts <- wt_pseudobulk_counts[, wt_pb_keep, drop = FALSE]
wt_pb_metadata <- wt_pb_metadata[wt_pb_keep, , drop = FALSE]

source_pseudobulk_logcpm <- log2_cpm(source_pseudobulk_counts)
wt_pseudobulk_logcpm <- log2_cpm(wt_pseudobulk_counts)

readr::write_csv(
  source_pb_metadata,
  file.path(STEP5D_TABLE_DIR, "source_pseudobulk_metadata.csv")
)

readr::write_csv(
  wt_pb_metadata,
  file.path(STEP5D_TABLE_DIR, "WT_pseudobulk_metadata.csv")
)

saveRDS(
  list(
    source_counts = source_pseudobulk_counts,
    source_log2_cpm = source_pseudobulk_logcpm,
    source_metadata = source_pb_metadata,
    wt_counts = wt_pseudobulk_counts,
    wt_log2_cpm = wt_pseudobulk_logcpm,
    wt_metadata = wt_pb_metadata
  ),
  file.path(STEP5D_DIR, "symmetric_pseudobulk_matrices.rds")
)
```

# 7. Define independent marker and curated feature panels

```{r define-feature-panels}
wt_markers <- suppressMessages(
  readr::read_csv(WT_MARKER_PATH, show_col_types = FALSE)
)

required_wt_marker_columns <- c("cluster", "gene", "p_val_adj")
missing_wt_marker_columns <- setdiff(
  required_wt_marker_columns,
  colnames(wt_markers)
)

if (length(missing_wt_marker_columns) > 0L) {
  stop(
    "WT marker table is missing column(s): ",
    paste(missing_wt_marker_columns, collapse = ", ")
  )
}

wt_fc_column <- intersect(
  c("effect_size", "avg_log2FC", "avg_logFC"),
  colnames(wt_markers)
)

if (length(wt_fc_column) == 0L) {
  stop("Could not identify a WT marker fold-change column.")
}

wt_fc_column <- wt_fc_column[[1]]

eligible_wt_clusters <- sort(unique(wt_pb_metadata$wnn_cluster))

wt_top_markers <- wt_markers |>
  dplyr::mutate(
    cluster = as.character(cluster),
    gene = as.character(gene),
    effect_size_for_ranking = .data[[wt_fc_column]]
  ) |>
  dplyr::filter(
    cluster %in% eligible_wt_clusters,
    p_val_adj < 0.05,
    !grepl(TECHNICAL_GENE_PATTERN, gene)
  ) |>
  dplyr::group_by(cluster) |>
  dplyr::arrange(
    p_val_adj,
    dplyr::desc(effect_size_for_ranking),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = TOP_MARKERS_PER_CLUSTER) |>
  dplyr::ungroup()

shared_expression_genes <- intersect(
  rownames(source_pseudobulk_logcpm),
  rownames(wt_pseudobulk_logcpm)
)

source_marker_genes <- unique(source_integrated_top_markers$gene)
wt_marker_genes <- unique(wt_top_markers$gene)
curated_lineage_genes <- unique(unlist(CURATED_PROGRAMS))

feature_panels <- list(
  shared_identity_no_cell_cycle = setdiff(
    intersect(source_marker_genes, wt_marker_genes),
    CELL_CYCLE_GENES
  ),
  union_identity_no_cell_cycle = setdiff(
    union(source_marker_genes, wt_marker_genes),
    CELL_CYCLE_GENES
  ),
  union_identity_with_cell_cycle = union(
    source_marker_genes,
    wt_marker_genes
  ),
  curated_lineage = curated_lineage_genes
)

feature_panels <- lapply(
  feature_panels,
  function(features) {
    sort(intersect(unique(features), shared_expression_genes))
  }
)

feature_panel_summary <- purrr::imap_dfr(
  feature_panels,
  function(features, panel_name) {
    tibble::tibble(
      feature_panel = panel_name,
      genes = length(features),
      cell_cycle_genes = sum(features %in% CELL_CYCLE_GENES),
      primary_panel = panel_name == "shared_identity_no_cell_cycle"
    )
  }
)

if (
  length(feature_panels$shared_identity_no_cell_cycle) <
    MIN_PRIMARY_FEATURES
) {
  stop(
    "The primary shared marker panel contains only ",
    length(feature_panels$shared_identity_no_cell_cycle),
    " genes; expected at least ", MIN_PRIMARY_FEATURES,
    ". Review the marker inputs before lowering this threshold.",
    call. = FALSE
  )
}

readr::write_csv(
  feature_panel_summary,
  file.path(STEP5D_TABLE_DIR, "feature_panel_summary.csv")
)

readr::write_csv(
  purrr::imap_dfr(
    feature_panels,
    ~ tibble::tibble(feature_panel = .y, gene = .x)
  ),
  file.path(STEP5D_TABLE_DIR, "feature_panel_genes.csv")
)

feature_panel_summary
```

# 8. Standardize expression within each dataset and diet

Gene-wise z-scores compare the relative enrichment pattern among clusters
within a platform. This reduces global whole-cell-versus-nucleus detection
differences. It does not make the two technologies identical.

```{r standardize-expression-and-balance-diets}
zscore_by_diet <- function(expression_matrix, pseudobulk_metadata) {
  output <- expression_matrix

  for (diet_value in unique(pseudobulk_metadata$diet)) {
    diet_columns <- which(pseudobulk_metadata$diet == diet_value)
    diet_matrix <- expression_matrix[, diet_columns, drop = FALSE]

    gene_means <- Matrix::rowMeans(diet_matrix)
    gene_sds <- apply(diet_matrix, 1, stats::sd)
    gene_sds[!is.finite(gene_sds) | gene_sds == 0] <- 1

    output[, diet_columns] <- (
      diet_matrix - gene_means
    ) / gene_sds
  }

  output
}

average_cluster_profiles <- function(
  standardized_matrix,
  pseudobulk_metadata,
  cluster_column,
  diet_filter = c("CON", "HFD")
) {
  metadata_subset <- pseudobulk_metadata |>
    dplyr::filter(diet %in% diet_filter)

  matrix_subset <- standardized_matrix[
    , match(metadata_subset$pseudobulk_id, colnames(standardized_matrix)),
    drop = FALSE
  ]

  cluster_values <- metadata_subset[[cluster_column]]
  cluster_levels <- sort(unique(cluster_values))

  profile_matrix <- vapply(
    cluster_levels,
    function(cluster_value) {
      Matrix::rowMeans(
        matrix_subset[, cluster_values == cluster_value, drop = FALSE]
      )
    },
    numeric(nrow(matrix_subset))
  )

  rownames(profile_matrix) <- rownames(matrix_subset)
  colnames(profile_matrix) <- cluster_levels
  profile_matrix
}

source_z <- zscore_by_diet(
  source_pseudobulk_logcpm,
  source_pb_metadata
)

wt_z <- zscore_by_diet(
  wt_pseudobulk_logcpm,
  wt_pb_metadata
)

source_profiles_by_scope <- list(
  diet_balanced = average_cluster_profiles(
    source_z,
    source_pb_metadata,
    "source_cluster_id",
    c("CON", "HFD")
  ),
  CON = average_cluster_profiles(
    source_z,
    source_pb_metadata,
    "source_cluster_id",
    "CON"
  ),
  HFD = average_cluster_profiles(
    source_z,
    source_pb_metadata,
    "source_cluster_id",
    "HFD"
  )
)

wt_profiles_by_scope <- list(
  diet_balanced = average_cluster_profiles(
    wt_z,
    wt_pb_metadata,
    "wt_cluster_id",
    c("CON", "HFD")
  ),
  CON = average_cluster_profiles(
    wt_z,
    wt_pb_metadata,
    "wt_cluster_id",
    "CON"
  ),
  HFD = average_cluster_profiles(
    wt_z,
    wt_pb_metadata,
    "wt_cluster_id",
    "HFD"
  )
)
```

# 9. Compute symmetric similarities and mutual-best matches

```{r compute-symmetric-similarities}
compute_similarity <- function(
  source_profiles,
  wt_profiles,
  features,
  feature_panel,
  diet_scope
) {
  features <- Reduce(
    intersect,
    list(features, rownames(source_profiles), rownames(wt_profiles))
  )

  if (length(features) < 10L) {
    stop("Fewer than 10 genes are available for ", feature_panel, ".")
  }

  correlation_matrix <- stats::cor(
    source_profiles[features, , drop = FALSE],
    wt_profiles[features, , drop = FALSE],
    method = "spearman",
    use = "pairwise.complete.obs"
  )

  as.data.frame(as.table(correlation_matrix)) |>
    tibble::as_tibble() |>
    dplyr::rename(
      source_cluster_id = Var1,
      wt_cluster_id = Var2,
      spearman_rho = Freq
    ) |>
    dplyr::mutate(
      source_cluster_id = as.character(source_cluster_id),
      wt_cluster_id = as.character(wt_cluster_id),
      feature_panel = feature_panel,
      diet_scope = diet_scope,
      genes = length(features)
    )
}

similarity_all <- purrr::imap_dfr(
  feature_panels,
  function(features, panel_name) {
    purrr::map_dfr(
      names(source_profiles_by_scope),
      function(scope_name) {
        compute_similarity(
          source_profiles_by_scope[[scope_name]],
          wt_profiles_by_scope[[scope_name]],
          features,
          panel_name,
          scope_name
        )
      }
    )
  }
)

source_lookup <- source_cluster_annotation |>
  dplyr::select(
    source_cluster_id,
    source_cluster_label,
    source_integrated_cluster,
    source_cell_state
  )

wt_lookup <- wt_metadata_epithelial |>
  dplyr::distinct(
    wt_cluster_id,
    wt_cluster_label,
    wnn_cluster,
    cell_state,
    cell_state_full,
    annotation_confidence
  )

similarity_all <- similarity_all |>
  dplyr::left_join(source_lookup, by = "source_cluster_id") |>
  dplyr::left_join(wt_lookup, by = "wt_cluster_id")

rank_similarity_pairs <- function(similarity_table) {
  source_ranked <- similarity_table |>
    dplyr::group_by(feature_panel, diet_scope, source_cluster_id) |>
    dplyr::arrange(dplyr::desc(spearman_rho), .by_group = TRUE) |>
    dplyr::mutate(
      source_to_WT_rank = dplyr::row_number(),
      source_to_WT_margin = dplyr::first(spearman_rho) -
        dplyr::nth(spearman_rho, 2, default = dplyr::first(spearman_rho))
    ) |>
    dplyr::ungroup() |>
    dplyr::select(
      feature_panel,
      diet_scope,
      source_cluster_id,
      wt_cluster_id,
      source_to_WT_rank,
      source_to_WT_margin
    )

  wt_ranked <- similarity_table |>
    dplyr::group_by(feature_panel, diet_scope, wt_cluster_id) |>
    dplyr::arrange(dplyr::desc(spearman_rho), .by_group = TRUE) |>
    dplyr::mutate(
      WT_to_source_rank = dplyr::row_number(),
      WT_to_source_margin = dplyr::first(spearman_rho) -
        dplyr::nth(spearman_rho, 2, default = dplyr::first(spearman_rho))
    ) |>
    dplyr::ungroup() |>
    dplyr::select(
      feature_panel,
      diet_scope,
      source_cluster_id,
      wt_cluster_id,
      WT_to_source_rank,
      WT_to_source_margin
    )

  similarity_table |>
    dplyr::left_join(
      source_ranked,
      by = c(
        "feature_panel", "diet_scope",
        "source_cluster_id", "wt_cluster_id"
      )
    ) |>
    dplyr::left_join(
      wt_ranked,
      by = c(
        "feature_panel", "diet_scope",
        "source_cluster_id", "wt_cluster_id"
      )
    ) |>
    dplyr::mutate(
      mutual_top1 = source_to_WT_rank == 1L & WT_to_source_rank == 1L
    )
}

similarity_ranked <- rank_similarity_pairs(similarity_all)

readr::write_csv(
  similarity_ranked,
  file.path(STEP5D_TABLE_DIR, "symmetric_cluster_similarity_all.csv")
)

primary_similarity <- similarity_ranked |>
  dplyr::filter(
    feature_panel == "shared_identity_no_cell_cycle",
    diet_scope == "diet_balanced"
  )

readr::write_csv(
  primary_similarity,
  file.path(
    STEP5D_TABLE_DIR,
    "primary_diet_balanced_symmetric_similarity.csv"
  )
)

primary_similarity |>
  dplyr::filter(source_to_WT_rank <= 3L | WT_to_source_rank <= 3L) |>
  dplyr::arrange(source_cluster_id, source_to_WT_rank)
```

# 10. Gene-bootstrap stability of the primary comparison

```{r bootstrap-primary-similarity}
primary_features <- feature_panels$shared_identity_no_cell_cycle
source_primary_profiles <- source_profiles_by_scope$diet_balanced
wt_primary_profiles <- wt_profiles_by_scope$diet_balanced

set.seed(BOOTSTRAP_SEED)

bootstrap_similarity <- purrr::map_dfr(
  seq_len(GENE_BOOTSTRAP_ITERATIONS),
  function(iteration) {
    sampled_features <- sample(
      primary_features,
      size = length(primary_features),
      replace = TRUE
    )

    correlation_matrix <- stats::cor(
      source_primary_profiles[sampled_features, , drop = FALSE],
      wt_primary_profiles[sampled_features, , drop = FALSE],
      method = "spearman",
      use = "pairwise.complete.obs"
    )

    as.data.frame(as.table(correlation_matrix)) |>
      tibble::as_tibble() |>
      dplyr::rename(
        source_cluster_id = Var1,
        wt_cluster_id = Var2,
        spearman_rho = Freq
      ) |>
      dplyr::mutate(
        iteration = iteration,
        source_cluster_id = as.character(source_cluster_id),
        wt_cluster_id = as.character(wt_cluster_id)
      )
  }
)

bootstrap_source_top1 <- bootstrap_similarity |>
  dplyr::group_by(iteration, source_cluster_id) |>
  dplyr::slice_max(
    order_by = spearman_rho,
    n = 1L,
    with_ties = FALSE
  ) |>
  dplyr::ungroup() |>
  dplyr::count(
    source_cluster_id,
    wt_cluster_id,
    name = "source_to_WT_top1_iterations"
  ) |>
  dplyr::mutate(
    source_to_WT_top1_frequency =
      source_to_WT_top1_iterations / GENE_BOOTSTRAP_ITERATIONS
  )

bootstrap_wt_top1 <- bootstrap_similarity |>
  dplyr::group_by(iteration, wt_cluster_id) |>
  dplyr::slice_max(
    order_by = spearman_rho,
    n = 1L,
    with_ties = FALSE
  ) |>
  dplyr::ungroup() |>
  dplyr::count(
    wt_cluster_id,
    source_cluster_id,
    name = "WT_to_source_top1_iterations"
  ) |>
  dplyr::mutate(
    WT_to_source_top1_frequency =
      WT_to_source_top1_iterations / GENE_BOOTSTRAP_ITERATIONS
  )

bootstrap_pair_summary <- bootstrap_similarity |>
  dplyr::group_by(source_cluster_id, wt_cluster_id) |>
  dplyr::summarise(
    bootstrap_median_rho = stats::median(spearman_rho),
    bootstrap_q025_rho = stats::quantile(spearman_rho, 0.025),
    bootstrap_q975_rho = stats::quantile(spearman_rho, 0.975),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    bootstrap_source_top1,
    by = c("source_cluster_id", "wt_cluster_id")
  ) |>
  dplyr::left_join(
    bootstrap_wt_top1,
    by = c("source_cluster_id", "wt_cluster_id")
  ) |>
  dplyr::mutate(
    source_to_WT_top1_frequency = dplyr::coalesce(
      source_to_WT_top1_frequency,
      0
    ),
    WT_to_source_top1_frequency = dplyr::coalesce(
      WT_to_source_top1_frequency,
      0
    )
  ) |>
  dplyr::left_join(source_lookup, by = "source_cluster_id") |>
  dplyr::left_join(wt_lookup, by = "wt_cluster_id")

primary_similarity_bootstrapped <- primary_similarity |>
  dplyr::left_join(
    bootstrap_pair_summary |>
      dplyr::select(
        source_cluster_id,
        wt_cluster_id,
        bootstrap_median_rho,
        bootstrap_q025_rho,
        bootstrap_q975_rho,
        source_to_WT_top1_frequency,
        WT_to_source_top1_frequency
      ),
    by = c("source_cluster_id", "wt_cluster_id")
  )

readr::write_csv(
  bootstrap_pair_summary,
  file.path(STEP5D_TABLE_DIR, "primary_similarity_gene_bootstrap.csv")
)

readr::write_csv(
  primary_similarity_bootstrapped,
  file.path(
    STEP5D_TABLE_DIR,
    "primary_similarity_with_bootstrap_support.csv"
  )
)
```

# 11. Curated lineage-program specificity

```{r curated-lineage-program-specificity}
score_programs <- function(profile_matrix, programs, dataset_name) {
  purrr::imap_dfr(
    programs,
    function(program_genes, program_name) {
      detected_genes <- intersect(program_genes, rownames(profile_matrix))

      if (length(detected_genes) == 0L) {
        return(tibble::tibble())
      }

      tibble::tibble(
        dataset = dataset_name,
        cluster_id = colnames(profile_matrix),
        program = program_name,
        mean_relative_expression = Matrix::colMeans(
          profile_matrix[detected_genes, , drop = FALSE]
        ),
        detected_program_genes = length(detected_genes),
        requested_program_genes = length(program_genes)
      )
    }
  )
}

curated_program_scores <- dplyr::bind_rows(
  score_programs(
    source_profiles_by_scope$diet_balanced,
    CURATED_PROGRAMS,
    "Gourab d21 scRNA"
  ) |>
    dplyr::left_join(
      source_lookup,
      by = c("cluster_id" = "source_cluster_id")
    ) |>
    dplyr::mutate(cluster_label = source_cluster_label),
  score_programs(
    wt_profiles_by_scope$diet_balanced,
    CURATED_PROGRAMS,
    "WT snMultiome RNA"
  ) |>
    dplyr::left_join(
      wt_lookup,
      by = c("cluster_id" = "wt_cluster_id")
    ) |>
    dplyr::mutate(cluster_label = wt_cluster_label)
)

readr::write_csv(
  curated_program_scores,
  file.path(STEP5D_TABLE_DIR, "curated_lineage_program_scores.csv")
)
```

# 12. Focused crypt and secretory review tables

```{r focused-crypt-and-secretory-review}
focus_similarity <- primary_similarity_bootstrapped |>
  dplyr::filter(wnn_cluster %in% FOCUS_WT_CLUSTERS) |>
  dplyr::group_by(wt_cluster_id) |>
  dplyr::arrange(dplyr::desc(spearman_rho), .by_group = TRUE) |>
  dplyr::mutate(WT_match_rank_within_focus_table = dplyr::row_number()) |>
  dplyr::filter(
    WT_match_rank_within_focus_table <= 5L |
      mutual_top1 |
      WT_to_source_top1_frequency >= 0.10
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    match(wnn_cluster, FOCUS_WT_CLUSTERS),
    WT_match_rank_within_focus_table
  )

focus_panel_sensitivity <- similarity_ranked |>
  dplyr::filter(wnn_cluster %in% FOCUS_WT_CLUSTERS) |>
  dplyr::group_by(feature_panel, diet_scope, wt_cluster_id) |>
  dplyr::slice_max(
    order_by = spearman_rho,
    n = 3L,
    with_ties = FALSE
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    match(wnn_cluster, FOCUS_WT_CLUSTERS),
    feature_panel,
    diet_scope,
    dplyr::desc(spearman_rho)
  )

readr::write_csv(
  focus_similarity,
  file.path(
    STEP5D_TABLE_DIR,
    "focused_WT_crypt_secretory_primary_matches.csv"
  )
)

readr::write_csv(
  focus_panel_sensitivity,
  file.path(
    STEP5D_TABLE_DIR,
    "focused_WT_crypt_secretory_panel_and_diet_sensitivity.csv"
  )
)

focus_similarity
```

# 13. Visual review

```{r source-integrated-cluster-umap, fig.width=15, fig.height=7}
if ("umap.source" %in% names(source_reference@reductions)) {
  source_umap <- Seurat::Embeddings(source_reference, "umap.source") |>
    as.data.frame() |>
    tibble::rownames_to_column("cell_barcode") |>
    dplyr::left_join(
      source_metadata |>
        dplyr::select(
          cell_barcode,
          diet,
          source_cell_state,
          source_integrated_cluster,
          source_cluster_label
        ),
      by = "cell_barcode"
    )

  p_source_cluster <- ggplot2::ggplot(
    source_umap,
    ggplot2::aes(
      x = .data[[colnames(source_umap)[[2]]]],
      y = .data[[colnames(source_umap)[[3]]]],
      color = source_integrated_cluster
    )
  ) +
    ggplot2::geom_point(size = 0.35, alpha = 0.75) +
    ggplot2::facet_wrap(~ diet, nrow = 1) +
    ggplot2::coord_equal() +
    ggplot2::labs(
      title = "Gourab d21 original integrated clusters",
      subtitle = "Named states may contain more than one integrated cluster",
      x = "UMAP 1",
      y = "UMAP 2",
      color = "Integrated\ncluster"
    ) +
    theme_multiome() +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.text = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank()
    )

  print(p_source_cluster)
  save_pdf_plot(
    file.path(STEP5D_PLOT_DIR, "01_source_integrated_clusters_by_diet.pdf"),
    p_source_cluster,
    15,
    7
  )
}
```

```{r source-integrated-marker-dotplot, fig.width=18, fig.height=11}
source_marker_display_genes <- source_integrated_top_markers |>
  dplyr::group_by(cluster) |>
  dplyr::slice_head(n = 5L) |>
  dplyr::ungroup() |>
  dplyr::pull(gene) |>
  unique()

source_marker_display_genes <- intersect(
  source_marker_display_genes,
  rownames(source_epithelial)
)

p_source_markers <- Seurat::DotPlot(
  source_epithelial,
  features = source_marker_display_genes,
  group.by = "source_integrated_cluster",
  assay = "RNA"
) +
  ggplot2::coord_flip() +
  ggplot2::labs(
    title = "Gourab d21 integrated-cluster marker audit",
    x = "Integrated cluster",
    y = NULL
  ) +
  theme_multiome() +
  ggplot2::theme(
    axis.text.y = ggplot2::element_text(size = 7)
  )

print(p_source_markers)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "02_source_integrated_cluster_markers.pdf"),
  p_source_markers,
  18,
  11
)
```

```{r primary-similarity-heatmap, fig.width=15, fig.height=11}
p_primary_heatmap <- ggplot2::ggplot(
  primary_similarity,
  ggplot2::aes(
    x = wt_cluster_label,
    y = source_cluster_label,
    fill = spearman_rho
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.15) +
  ggplot2::geom_point(
    data = primary_similarity |>
      dplyr::filter(mutual_top1),
    shape = 21,
    size = 3,
    stroke = 0.9,
    fill = NA,
    color = "black"
  ) +
  ggplot2::scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    limits = c(-1, 1),
    oob = scales::squish
  ) +
  ggplot2::labs(
    title = "Diet-balanced symmetric epithelial similarity",
    subtitle = paste0(
      "Spearman correlation of ",
      length(primary_features),
      " shared identity markers; circles mark mutual top-1 matches"
    ),
    x = "WT snMultiome WNN cluster",
    y = "Gourab d21 integrated cluster",
    fill = "Spearman\nrho"
  ) +
  theme_multiome() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 55, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 8)
  )

print(p_primary_heatmap)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "03_primary_diet_balanced_similarity.pdf"),
  p_primary_heatmap,
  15,
  11
)
```

```{r sensitivity-heatmaps, fig.width=18, fig.height=14}
panel_sensitivity_plot_data <- similarity_ranked |>
  dplyr::filter(diet_scope == "diet_balanced")

p_panel_sensitivity <- ggplot2::ggplot(
  panel_sensitivity_plot_data,
  ggplot2::aes(
    x = wt_cluster_label,
    y = source_cluster_label,
    fill = spearman_rho
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.1) +
  ggplot2::facet_wrap(~ feature_panel, ncol = 2) +
  ggplot2::scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    limits = c(-1, 1),
    oob = scales::squish
  ) +
  ggplot2::labs(
    title = "Similarity sensitivity to feature-panel definition",
    x = "WT cluster",
    y = "Gourab integrated cluster",
    fill = "Spearman\nrho"
  ) +
  theme_multiome() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 70, hjust = 1, size = 6),
    axis.text.y = ggplot2::element_text(size = 6)
  )

print(p_panel_sensitivity)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "04_feature_panel_sensitivity.pdf"),
  p_panel_sensitivity,
  18,
  14
)

diet_sensitivity_plot_data <- similarity_ranked |>
  dplyr::filter(feature_panel == "shared_identity_no_cell_cycle")

p_diet_sensitivity <- ggplot2::ggplot(
  diet_sensitivity_plot_data,
  ggplot2::aes(
    x = wt_cluster_label,
    y = source_cluster_label,
    fill = spearman_rho
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.1) +
  ggplot2::facet_wrap(~ diet_scope, ncol = 1) +
  ggplot2::scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    limits = c(-1, 1),
    oob = scales::squish
  ) +
  ggplot2::labs(
    title = "Similarity sensitivity to diet stratum",
    subtitle = "CON and HFD are descriptive pooled-library strata",
    x = "WT cluster",
    y = "Gourab integrated cluster",
    fill = "Spearman\nrho"
  ) +
  theme_multiome() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 70, hjust = 1, size = 6),
    axis.text.y = ggplot2::element_text(size = 6)
  )

print(p_diet_sensitivity)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "05_diet_stratum_sensitivity.pdf"),
  p_diet_sensitivity,
  17,
  18
)
```

```{r curated-program-heatmap, fig.width=17, fig.height=14}
p_curated_programs <- ggplot2::ggplot(
  curated_program_scores,
  ggplot2::aes(
    x = program,
    y = cluster_label,
    fill = mean_relative_expression
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.15) +
  ggplot2::facet_wrap(~ dataset, scales = "free_y", ncol = 1) +
  ggplot2::scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0
  ) +
  ggplot2::labs(
    title = "Curated lineage-program specificity",
    subtitle = "Mean relative expression in diet-balanced profiles",
    x = NULL,
    y = NULL,
    fill = "Mean relative\nexpression"
  ) +
  theme_multiome() +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 7)
  )

print(p_curated_programs)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "06_curated_lineage_program_specificity.pdf"),
  p_curated_programs,
  17,
  14
)
```

```{r bootstrap-focus-plot, fig.width=13, fig.height=11}
focus_best_matches <- primary_similarity_bootstrapped |>
  dplyr::filter(wnn_cluster %in% FOCUS_WT_CLUSTERS) |>
  dplyr::group_by(wt_cluster_id) |>
  dplyr::slice_max(
    order_by = spearman_rho,
    n = 3L,
    with_ties = FALSE
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    comparison_label = paste0(
      wt_cluster_label,
      "  <-  ",
      source_cluster_label
    ),
    comparison_label = factor(
      comparison_label,
      levels = rev(unique(comparison_label))
    )
  )

p_bootstrap_focus <- ggplot2::ggplot(
  focus_best_matches,
  ggplot2::aes(
    x = bootstrap_median_rho,
    y = comparison_label,
    color = WT_to_source_top1_frequency
  )
) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      xmin = bootstrap_q025_rho,
      xmax = bootstrap_q975_rho
    ),
    orientation = "y",
    width = 0.20,
    linewidth = 0.6
  ) +
  ggplot2::geom_point(size = 3) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  ggplot2::scale_color_viridis_c(
    limits = c(0, 1),
    labels = scales::label_percent()
  ) +
  ggplot2::labs(
    title = "Stability of focused WT crypt and secretory matches",
    subtitle = paste0(
      GENE_BOOTSTRAP_ITERATIONS,
      " gene bootstraps; intervals are 2.5th-97.5th percentiles"
    ),
    x = "Bootstrap Spearman correlation",
    y = NULL,
    color = "WT-to-source\ntop-1 frequency"
  ) +
  theme_multiome()

print(p_bootstrap_focus)
save_pdf_plot(
  file.path(STEP5D_PLOT_DIR, "07_focused_match_bootstrap_stability.pdf"),
  p_bootstrap_focus,
  13,
  11
)
```

# 14. Save review summary and completion records

```{r save-completion-records}
primary_mutual_matches <- primary_similarity_bootstrapped |>
  dplyr::filter(mutual_top1) |>
  dplyr::arrange(dplyr::desc(spearman_rho))

readr::write_csv(
  primary_mutual_matches,
  file.path(STEP5D_TABLE_DIR, "primary_mutual_top1_matches.csv")
)

step5d_summary <- tibble::tibble(
  analysis_version = ANALYSIS_VERSION,
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  inverse_transfer_disposition = ifelse(
    inverse_failed,
    "FAILED_FOR_BIOLOGICAL_INTERPRETATION",
    "PASSED_DESCRIPTIVE_QC"
  ),
  source_epithelial_cells = nrow(source_metadata_epithelial),
  source_integrated_clusters = dplyr::n_distinct(
    source_metadata_epithelial$source_integrated_cluster
  ),
  WT_epithelial_nuclei = nrow(wt_metadata_epithelial),
  WT_clusters = dplyr::n_distinct(wt_metadata_epithelial$wnn_cluster),
  primary_feature_genes = length(primary_features),
  primary_mutual_top1_matches = nrow(primary_mutual_matches),
  bootstrap_iterations = GENE_BOOTSTRAP_ITERATIONS,
  annotation_overrides_applied = FALSE,
  biological_diet_inference_performed = FALSE
)

readr::write_csv(
  step5d_summary,
  file.path(STEP5D_DIR, "step5d_summary.csv")
)

writeLines(
  c(
    paste("Step 5D completed:", Sys.time()),
    paste("Analysis version:", ANALYSIS_VERSION),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste(
      "Step 5C inverse-transfer disposition:",
      step5d_summary$inverse_transfer_disposition
    ),
    paste("Source epithelial cells:", nrow(source_metadata_epithelial)),
    paste(
      "Source integrated clusters:",
      dplyr::n_distinct(source_metadata_epithelial$source_integrated_cluster)
    ),
    paste("WT epithelial nuclei:", nrow(wt_metadata_epithelial)),
    paste(
      "WT clusters:",
      dplyr::n_distinct(wt_metadata_epithelial$wnn_cluster)
    ),
    paste("Primary identity genes:", length(primary_features)),
    paste("Primary mutual top-1 matches:", nrow(primary_mutual_matches)),
    paste("Gene bootstrap iterations:", GENE_BOOTSTRAP_ITERATIONS),
    "WT annotation overrides applied: FALSE",
    "Biological diet inference performed: FALSE"
  ),
  con = file.path(STEP5D_DIR, "STEP_05D_COMPLETE.txt")
)

writeLines(
  c(
    "Step 5D symmetric pseudobulk validation is ready for manual review.",
    paste(
      "Review inverse-transfer QC:",
      file.path(STEP5D_TABLE_DIR, "step5c_inverse_qc_disposition.csv")
    ),
    paste(
      "Review primary similarity:",
      file.path(
        STEP5D_TABLE_DIR,
        "primary_similarity_with_bootstrap_support.csv"
      )
    ),
    paste(
      "Review focused crypt/secretory matches:",
      file.path(
        STEP5D_TABLE_DIR,
        "focused_WT_crypt_secretory_primary_matches.csv"
      )
    ),
    paste(
      "Review panel/diet sensitivity:",
      file.path(
        STEP5D_TABLE_DIR,
        "focused_WT_crypt_secretory_panel_and_diet_sensitivity.csv"
      )
    ),
    "Do not revise WT annotations until the symmetric matches, curated ",
    "programs, and bootstrap stability have been reviewed together."
  ),
  con = file.path(STEP5D_DIR, "STEP_05D_READY_FOR_REVIEW.txt")
)

step5d_summary
```

# 15. Session information

```{r session-information}
sessionInfo()
```
