---
title: "Step 5E - Focused WT crypt stem and progenitor refinement"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# Purpose

Step 5D showed that the existing WT crypt annotations are too coarse for a
defensible stem-versus-progenitor interpretation. In particular, global WNN
cluster 0 contains a strong cycling program, cluster 6 resembles a
`Dmbt1`/`Cftr`-rich crypt progenitor state, and cluster 8 does not behave as a
clean canonical ISC population in the symmetric source comparison.

This notebook performs a **review-only** focused analysis. It does not replace
the approved Step 6 labels. It:

1. refits RNA PCA, ATAC LSI, and WNN neighborhoods within clusters 0, 6, and 8;
2. builds an RNA-only clustering sensitivity on those same nuclei;
3. repeats the WNN fit after adding clusters 1 and 18 as absorptive and
   secretory boundary sensitivities;
4. scores cell cycle, curated crypt programs, and marker programs derived from
   Gourab's original d21 integrated clusters;
5. audits cluster stability, parent composition, diet representation, QC, and
   modality weights across a resolution grid;
6. performs formal marker review at the primary resolution 0.3 and sensitivity
   resolutions 0.4 and 0.6;
7. pairs targeted RNA evidence with exact promoter accessibility within 2 kb
   of the annotated TSS; and
8. exports marker and crosswalk tables for a manual refinement decision.

The focused clusters are discovered independently of all marker-program
scores. Those scores are used only after clustering to interpret the result.
No diet-effect inference is performed because each diet is represented by one
pooled library.

A practical SOL allocation is **8 CPU cores, 96 GiB memory, 8 hours, R 4.4.2,
no GPU**.

# 1. Setup

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 12,
  fig.height = 8
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Signac)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(readr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(scales)
})

if (!requireNamespace("presto", quietly = TRUE)) {
  stop(
    "Package 'presto' is required for the focused marker audit. Run ",
    "setup/install_packages.R in a clean R session first.",
    call. = FALSE
  )
}

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 5E requires MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}

ggplot2::theme_set(theme_multiome())
future::plan(future::sequential)
set.seed(12345)
```

# 2. Inputs, outputs, and audit parameters

```{r define-inputs-and-outputs}
ANALYSIS_VERSION <- "05e_focused_wt_crypt_v5"

STEP6_OBJECT_PATH <- file.path(
  OUTPUT_DIR,
  "06_annotation",
  "multiome_annotated.rds"
)

STEP5D_DIR <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "gourab_d21_rna_reference_mapping",
  "05d_symmetric_pseudobulk_validation"
)

SOURCE_MARKER_PATH <- file.path(
  STEP5D_DIR,
  "tables",
  "source_integrated_cluster_all_positive_markers.csv"
)

STEP5E_DIR <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "focused_wt_crypt_stem_progenitor"
)

STEP5E_TABLE_DIR <- file.path(STEP5E_DIR, "tables")
STEP5E_PLOT_DIR <- file.path(STEP5E_DIR, "plots")
STEP5E_OBJECT_DIR <- file.path(STEP5E_DIR, "objects")

for (output_directory in c(
  STEP5E_DIR,
  STEP5E_TABLE_DIR,
  STEP5E_PLOT_DIR,
  STEP5E_OBJECT_DIR
)) {
  dir.create(output_directory, recursive = TRUE, showWarnings = FALSE)
}

for (input_path in c(STEP6_OBJECT_PATH, SOURCE_MARKER_PATH)) {
  assert_file(input_path)
}

CORE_PARENT_CLUSTERS <- c("0", "6", "8")
BOUNDARY_PARENT_CLUSTERS <- c("0", "1", "6", "8", "18")

FOCUSED_RESOLUTIONS <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.8)
PRIMARY_REVIEW_RESOLUTION <- 0.3
MARKER_REVIEW_RESOLUTIONS <- c(0.3, 0.4, 0.6)

RNA_NFEATURES <- 3000L
RNA_NPCS <- 30L
RNA_DIMS <- 1:20
ATAC_NLSI <- 30L
ATAC_DIMS <- 2:20
WNN_K <- 20L

MIN_PROGRAM_GENES <- 3L
SOURCE_SIGNATURE_GENES_PER_CLUSTER <- 50L
MARKER_MIN_PCT <- 0.10
MARKER_LOGFC_THRESHOLD <- 0.25
MARKER_ADJUSTED_P_THRESHOLD <- 0.05

TARGETED_EVIDENCE_PROGRAMS <- list(
  canonical_ISC = c(
    "Lgr5", "Ascl2", "Smoc2", "Axin2", "Olfm4", "Lrig1", "Rnf43"
  ),
  cell_cycle = c(
    "Mki67", "Top2a", "Cenpf", "Ube2c", "Cdk1", "Ccnb1"
  ),
  crypt_progenitor = c(
    "Dmbt1", "Cftr", "Htr4", "Kcnq1", "Slc12a2", "Hmgcs2", "Hsd3b3"
  ),
  secretory_DSC = c(
    "Atoh1", "Dll1", "Spdef", "Muc2", "Reg4", "Spink4",
    "Slc12a8", "Kcnma1", "Sytl2"
  )
)

TARGETED_EVIDENCE_GENES <- unique(unlist(TARGETED_EVIDENCE_PROGRAMS))
PROMOTER_WINDOW_BP <- 2000L

TECHNICAL_GENE_PATTERN <- paste0(
  "^(mt-|Rpl|Rps|Gm[0-9]+|Mir[0-9]+|Snor|Hist[0-9])",
  "|Rik$"
)

CURATED_PROGRAMS <- list(
  curated_canonical_ISC = c(
    "Lgr5", "Ascl2", "Smoc2", "Axin2", "Olfm4", "Lrig1",
    "Prom1", "Sox9", "Hopx", "Tert"
  ),
  curated_cell_cycle = c(
    "Mki67", "Top2a", "Cenpf", "Ube2c", "Cdk1", "Ccnb1",
    "Ccnb2", "Tpx2", "Nusap1", "Kif11", "Birc5", "Prc1"
  ),
  curated_crypt_progenitor = c(
    "Dmbt1", "Cftr", "Htr4", "Kcnq1", "Slc12a2", "Sema5a",
    "Hmga2", "Clca3a2", "Kitl", "Hmgcs2"
  ),
  curated_early_absorptive = c(
    "Slc26a3", "Slc9a2", "Lama3", "Car4", "Aqp8", "Krt20",
    "Mxd1", "Tgm3", "Muc3", "Cyp2c55"
  ),
  curated_secretory_commitment = c(
    "Atoh1", "Dll1", "Spdef", "Muc2", "Agr2", "Gfi1",
    "Neurog3", "Fcgbp", "Clca1", "Zg16"
  ),
  curated_Reg4_DSC = c(
    "Reg4", "Spink4", "Kit", "Klk1", "Car8", "Hepacam2",
    "Slc12a8", "Sytl2", "Plcb4", "Agr2"
  )
)

SOURCE_PROGRAM_CLUSTER_MAP <- list(
  source_G6_canonical_ISC = "6",
  source_G7_mixed_crypt_secretory = "7",
  source_G8_absorptive_like_legacy_stem = "8",
  source_G9_crypt_progenitor = "9",
  source_G5_G11_cycling_TA = c("5", "11"),
  source_G3_G12_cycling_secretory = c("3", "12")
)

resolution_tag <- function(resolution) {
  format(resolution, nsmall = 1, trim = TRUE)
}

resolution_column <- function(prefix, resolution) {
  paste0(prefix, "_res_", resolution_tag(resolution))
}

CORE_WNN_PREFIX <- "crypt_core_wnn"
CORE_RNA_PREFIX <- "crypt_core_rna"
BOUNDARY_WNN_PREFIX <- "crypt_boundary_wnn"
```

# 3. Helper functions

```{r define-helper-functions}
get_aligned_layer <- function(object, assay, layer, cells = colnames(object)) {
  matrix_data <- SeuratObject::LayerData(
    object[[assay]],
    layer = layer
  )

  if (!setequal(colnames(matrix_data), colnames(object))) {
    stop(
      "Assay ", assay, " layer ", layer,
      " does not contain the same cells as the Seurat object.",
      call. = FALSE
    )
  }

  missing_cells <- setdiff(cells, colnames(matrix_data))

  if (length(missing_cells) > 0L) {
    stop(
      "Requested cells are absent from assay ", assay, ": ",
      paste(head(missing_cells, 10L), collapse = ", "),
      call. = FALSE
    )
  }

  matrix_data[, cells, drop = FALSE]
}

derive_source_programs <- function(marker_table) {
  required_columns <- c(
    "cluster", "gene", "effect_size", "pct.1", "pct.2", "p_val_adj"
  )

  missing_columns <- setdiff(required_columns, colnames(marker_table))

  if (length(missing_columns) > 0L) {
    stop(
      "The Step 5D source marker table is missing column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }

  marker_table <- marker_table |>
    dplyr::mutate(
      cluster = as.character(cluster),
      gene = as.character(gene),
      pct_delta = pct.1 - pct.2
    ) |>
    dplyr::filter(
      is.finite(effect_size),
      effect_size > 0,
      is.finite(p_val_adj),
      p_val_adj < 0.05,
      !grepl(TECHNICAL_GENE_PATTERN, gene)
    )

  programs <- lapply(
    SOURCE_PROGRAM_CLUSTER_MAP,
    function(source_clusters) {
      marker_table |>
        dplyr::filter(cluster %in% source_clusters) |>
        dplyr::group_by(cluster) |>
        dplyr::arrange(
          p_val_adj,
          dplyr::desc(pct_delta),
          dplyr::desc(effect_size),
          .by_group = TRUE
        ) |>
        dplyr::slice_head(n = SOURCE_SIGNATURE_GENES_PER_CLUSTER) |>
        dplyr::ungroup() |>
        dplyr::distinct(gene) |>
        dplyr::pull(gene)
    }
  )

  too_small <- names(programs)[lengths(programs) < MIN_PROGRAM_GENES]

  if (length(too_small) > 0L) {
    stop(
      "Too few genes were recovered for source program(s): ",
      paste(too_small, collapse = ", "),
      call. = FALSE
    )
  }

  programs
}

add_program_scores <- function(object, assay, programs) {
  normalized_data <- SeuratObject::LayerData(
    object[[assay]],
    layer = "data"
  )

  available_programs <- lapply(
    programs,
    function(genes) intersect(unique(genes), rownames(normalized_data))
  )

  program_gene_summary <- tibble::tibble(
    program = names(programs),
    genes_requested = lengths(programs),
    genes_available = lengths(available_programs),
    available_genes = vapply(
      available_programs,
      paste,
      character(1),
      collapse = ";"
    )
  )

  invalid_programs <- program_gene_summary$program[
    program_gene_summary$genes_available < MIN_PROGRAM_GENES
  ]

  if (length(invalid_programs) > 0L) {
    stop(
      "Fewer than ", MIN_PROGRAM_GENES,
      " genes were available for program(s): ",
      paste(invalid_programs, collapse = ", "),
      call. = FALSE
    )
  }

  for (program_name in names(available_programs)) {
    program_features <- available_programs[[program_name]]
    temporary_prefix <- paste0("temporary_", program_name, "_")
    metadata_before <- colnames(object[[]])

    object <- Seurat::AddModuleScore(
      object,
      features = list(program_features),
      assay = assay,
      name = temporary_prefix,
      seed = 12345,
      search = FALSE
    )

    added_columns <- setdiff(colnames(object[[]]), metadata_before)

    if (length(added_columns) != 1L) {
      stop(
        "Could not identify the AddModuleScore output for ",
        program_name,
        call. = FALSE
      )
    }

    score_column <- paste0(program_name, "_score")
    detection_column <- paste0(program_name, "_genes_detected")

    object[[score_column]] <- object[[]][[added_columns]]
    object[[added_columns]] <- NULL
    object[[detection_column]] <- Matrix::colSums(
      normalized_data[program_features, , drop = FALSE] > 0
    )
  }

  list(
    object = object,
    gene_summary = program_gene_summary,
    programs = available_programs
  )
}

build_focused_reductions <- function(
  parent_object,
  parent_clusters,
  reduction_prefix,
  build_rna_graph = FALSE
) {
  parent_values <- as.character(parent_object$wnn_cluster)
  focused_cells <- colnames(parent_object)[
    parent_values %in% parent_clusters
  ]

  if (length(focused_cells) == 0L) {
    stop(
      "No nuclei matched parent clusters: ",
      paste(parent_clusters, collapse = ", "),
      call. = FALSE
    )
  }

  focused <- subset(parent_object, cells = focused_cells)
  focused$parent_wnn_cluster <- as.character(focused$wnn_cluster)

  rna_counts <- get_aligned_layer(
    parent_object,
    assay = "RNA",
    layer = "counts",
    cells = focused_cells
  )

  focused[["RNA.crypt"]] <- SeuratObject::CreateAssayObject(
    counts = rna_counts
  )
  SeuratObject::Key(focused[["RNA.crypt"]]) <- "CRYPT_"

  # The Seurat object is already restricted to focused_cells. Copying the
  # complete ChromatinAssay first preserves its ranges, fragments, and genome
  # annotations. Features are then filtered with subset(), avoiding the
  # layer-indexing failure produced by assay[logical_vector, ].
  focused[["ATAC.crypt"]] <- focused[["ATAC"]]
  SeuratObject::Key(focused[["ATAC.crypt"]]) <- "ATACCRYPT_"

  atac_counts <- SeuratObject::LayerData(
    focused[["ATAC.crypt"]],
    layer = "counts"
  )

  if (!identical(colnames(atac_counts), focused_cells)) {
    if (!setequal(colnames(atac_counts), focused_cells)) {
      stop("Focused ATAC counts do not contain the expected nuclei.")
    }
    atac_counts <- atac_counts[, focused_cells, drop = FALSE]
  }

  nonzero_atac_features <- rownames(atac_counts)[
    Matrix::rowSums(atac_counts) > 0
  ]

  if (length(nonzero_atac_features) == 0L) {
    stop("No nonzero ATAC features remain in the focused object.")
  }

  focused[["ATAC.crypt"]] <- subset(
    x = focused[["ATAC.crypt"]],
    features = nonzero_atac_features,
    cells = focused_cells
  )

  feature_filter_summary <- tibble::tibble(
    analysis_scope = reduction_prefix,
    nuclei = length(focused_cells),
    parent_clusters = paste(parent_clusters, collapse = ";"),
    ATAC_features_before = nrow(atac_counts),
    zero_count_features_removed = nrow(atac_counts) -
      length(nonzero_atac_features),
    ATAC_features_retained = length(nonzero_atac_features)
  )

  rm(rna_counts, atac_counts)
  invisible(gc())

  rna_pca_name <- paste0(reduction_prefix, ".pca")
  rna_umap_name <- paste0(reduction_prefix, ".rna.umap")
  atac_lsi_name <- paste0(reduction_prefix, ".lsi")
  atac_umap_name <- paste0(reduction_prefix, ".atac.umap")
  wnn_umap_name <- paste0(reduction_prefix, ".wnn.umap")
  wknn_name <- paste0(reduction_prefix, ".wknn")
  wsnn_name <- paste0(reduction_prefix, ".wsnn")
  weighted_nn_name <- paste0(reduction_prefix, ".weighted.nn")
  rna_nn_name <- paste0(reduction_prefix, ".rna.nn")
  rna_snn_name <- paste0(reduction_prefix, ".rna.snn")

  Seurat::DefaultAssay(focused) <- "RNA.crypt"

  focused <- Seurat::NormalizeData(
    focused,
    assay = "RNA.crypt",
    verbose = FALSE
  )
  focused <- Seurat::FindVariableFeatures(
    focused,
    assay = "RNA.crypt",
    selection.method = "vst",
    nfeatures = RNA_NFEATURES,
    verbose = FALSE
  )
  focused <- Seurat::ScaleData(
    focused,
    assay = "RNA.crypt",
    features = Seurat::VariableFeatures(focused[["RNA.crypt"]]),
    verbose = FALSE
  )
  focused <- Seurat::RunPCA(
    focused,
    assay = "RNA.crypt",
    features = Seurat::VariableFeatures(focused[["RNA.crypt"]]),
    npcs = RNA_NPCS,
    reduction.name = rna_pca_name,
    reduction.key = paste0(toupper(gsub("[^A-Za-z0-9]", "", reduction_prefix)), "PCA_"),
    verbose = FALSE
  )
  focused <- Seurat::RunUMAP(
    focused,
    reduction = rna_pca_name,
    dims = RNA_DIMS,
    reduction.name = rna_umap_name,
    reduction.key = paste0(toupper(gsub("[^A-Za-z0-9]", "", reduction_prefix)), "RNAUMAP_"),
    seed.use = 12345,
    verbose = FALSE
  )

  Seurat::DefaultAssay(focused) <- "ATAC.crypt"
  focused <- Signac::RunTFIDF(
    focused,
    assay = "ATAC.crypt",
    verbose = FALSE
  )
  focused <- Signac::FindTopFeatures(
    focused,
    assay = "ATAC.crypt",
    min.cutoff = "q5",
    verbose = FALSE
  )
  focused <- Signac::RunSVD(
    focused,
    assay = "ATAC.crypt",
    n = ATAC_NLSI,
    reduction.name = atac_lsi_name,
    reduction.key = paste0(toupper(gsub("[^A-Za-z0-9]", "", reduction_prefix)), "LSI_"),
    verbose = FALSE
  )
  focused <- Seurat::RunUMAP(
    focused,
    reduction = atac_lsi_name,
    dims = ATAC_DIMS,
    reduction.name = atac_umap_name,
    reduction.key = paste0(toupper(gsub("[^A-Za-z0-9]", "", reduction_prefix)), "ATACUMAP_"),
    seed.use = 12345,
    verbose = FALSE
  )

  focused <- Seurat::FindMultiModalNeighbors(
    focused,
    reduction.list = list(rna_pca_name, atac_lsi_name),
    dims.list = list(RNA_DIMS, ATAC_DIMS),
    k.nn = WNN_K,
    knn.graph.name = wknn_name,
    snn.graph.name = wsnn_name,
    weighted.nn.name = weighted_nn_name,
    modality.weight.name = c(
      paste0(reduction_prefix, ".RNA.weight"),
      paste0(reduction_prefix, ".ATAC.weight")
    ),
    verbose = FALSE
  )
  focused <- Seurat::RunUMAP(
    focused,
    nn.name = weighted_nn_name,
    reduction.name = wnn_umap_name,
    reduction.key = paste0(toupper(gsub("[^A-Za-z0-9]", "", reduction_prefix)), "WNNUMAP_"),
    seed.use = 12345,
    verbose = FALSE
  )

  if (build_rna_graph) {
    focused <- Seurat::FindNeighbors(
      focused,
      reduction = rna_pca_name,
      dims = RNA_DIMS,
      graph.name = c(rna_nn_name, rna_snn_name),
      verbose = FALSE
    )
  }

  expected_reductions <- c(
    rna_pca_name,
    rna_umap_name,
    atac_lsi_name,
    atac_umap_name,
    wnn_umap_name
  )

  valid_reductions <- vapply(
    expected_reductions,
    function(reduction_name) {
      all(is.finite(Seurat::Embeddings(focused, reduction_name)))
    },
    logical(1)
  )

  if (!all(valid_reductions)) {
    stop(
      "Non-finite values were found in focused reduction(s): ",
      paste(names(valid_reductions)[!valid_reductions], collapse = ", "),
      call. = FALSE
    )
  }

  list(
    object = focused,
    feature_filter_summary = feature_filter_summary,
    rna_pca = rna_pca_name,
    rna_umap = rna_umap_name,
    atac_lsi = atac_lsi_name,
    atac_umap = atac_umap_name,
    wnn_umap = wnn_umap_name,
    wsnn = wsnn_name,
    rna_snn = rna_snn_name
  )
}

add_resolution_grid <- function(
  object,
  graph_name,
  column_prefix,
  resolutions = FOCUSED_RESOLUTIONS
) {
  resolution_columns <- character(length(resolutions))

  for (resolution_index in seq_along(resolutions)) {
    resolution <- resolutions[[resolution_index]]
    output_column <- resolution_column(column_prefix, resolution)

    object <- Seurat::FindClusters(
      object,
      graph.name = graph_name,
      algorithm = 3,
      resolution = resolution,
      random.seed = 12345,
      verbose = FALSE
    )

    object[[output_column]] <- as.character(Seurat::Idents(object))
    resolution_columns[[resolution_index]] <- output_column
  }

  list(object = object, columns = resolution_columns)
}

summarise_resolution_grid <- function(
  object,
  resolution_columns,
  analysis_scope,
  rna_weight_column,
  atac_weight_column,
  score_columns
) {
  metadata <- object[[]] |>
    tibble::rownames_to_column("cell_barcode")

  available_qc <- intersect(
    c(
      "nCount_RNA", "nFeature_RNA", "nCount_ATAC", "nFeature_ATAC",
      "percent.mt", "TSS.enrichment", "nucleosome_signal", "frip_cr",
      "scDblFinder.score", rna_weight_column, atac_weight_column,
      "S.Score", "G2M.Score"
    ),
    colnames(metadata)
  )

  cluster_summary <- purrr::map2_dfr(
    FOCUSED_RESOLUTIONS,
    resolution_columns,
    function(resolution, cluster_column) {
      metadata |>
        dplyr::mutate(focused_subcluster = .data[[cluster_column]]) |>
        dplyr::group_by(focused_subcluster) |>
        dplyr::summarise(
          nuclei = dplyr::n(),
          CON_nuclei = sum(as.character(diet) == "CON"),
          HFD_nuclei = sum(as.character(diet) == "HFD"),
          CON_percent = 100 * mean(as.character(diet) == "CON"),
          HFD_percent = 100 * mean(as.character(diet) == "HFD"),
          S_phase_percent = 100 * mean(as.character(Phase) == "S"),
          G2M_phase_percent = 100 * mean(as.character(Phase) == "G2M"),
          dplyr::across(
            dplyr::all_of(c(available_qc, score_columns)),
            ~ stats::median(.x, na.rm = TRUE),
            .names = "{.col}_median"
          ),
          .groups = "drop"
        ) |>
        dplyr::mutate(
          analysis_scope = analysis_scope,
          resolution = resolution,
          cluster_column = cluster_column,
          .before = 1
        )
    }
  )

  parent_composition <- purrr::map2_dfr(
    FOCUSED_RESOLUTIONS,
    resolution_columns,
    function(resolution, cluster_column) {
      metadata |>
        dplyr::transmute(
          focused_subcluster = .data[[cluster_column]],
          parent_wnn_cluster = as.character(parent_wnn_cluster)
        ) |>
        dplyr::count(
          focused_subcluster,
          parent_wnn_cluster,
          name = "nuclei"
        ) |>
        dplyr::group_by(focused_subcluster) |>
        dplyr::mutate(
          percent_of_focused_subcluster = 100 * nuclei / sum(nuclei)
        ) |>
        dplyr::ungroup() |>
        dplyr::mutate(
          analysis_scope = analysis_scope,
          resolution = resolution,
          cluster_column = cluster_column,
          .before = 1
        )
    }
  )

  list(
    cluster_summary = cluster_summary,
    parent_composition = parent_composition
  )
}

make_adjacent_resolution_crosswalk <- function(
  object,
  resolution_columns,
  analysis_scope
) {
  metadata <- object[[]]

  purrr::map_dfr(
    seq_len(length(resolution_columns) - 1L),
    function(index) {
      from_column <- resolution_columns[[index]]
      to_column <- resolution_columns[[index + 1L]]

      tibble::tibble(
        from_cluster = as.character(metadata[[from_column]]),
        to_cluster = as.character(metadata[[to_column]])
      ) |>
        dplyr::count(from_cluster, to_cluster, name = "nuclei") |>
        dplyr::group_by(from_cluster) |>
        dplyr::mutate(
          percent_of_from_cluster = 100 * nuclei / sum(nuclei)
        ) |>
        dplyr::ungroup() |>
        dplyr::mutate(
          analysis_scope = analysis_scope,
          from_resolution = FOCUSED_RESOLUTIONS[[index]],
          to_resolution = FOCUSED_RESOLUTIONS[[index + 1L]],
          from_column = from_column,
          to_column = to_column,
          .before = 1
        )
    }
  )
}

make_same_resolution_crosswalk <- function(
  left_object,
  left_columns,
  right_object,
  right_columns,
  comparison
) {
  shared_cells <- intersect(colnames(left_object), colnames(right_object))

  if (length(shared_cells) == 0L) {
    stop("No shared nuclei were available for ", comparison, ".")
  }

  left_metadata <- left_object[[]][shared_cells, , drop = FALSE]
  right_metadata <- right_object[[]][shared_cells, , drop = FALSE]

  purrr::pmap_dfr(
    list(FOCUSED_RESOLUTIONS, left_columns, right_columns),
    function(resolution, left_column, right_column) {
      tibble::tibble(
        left_cluster = as.character(left_metadata[[left_column]]),
        right_cluster = as.character(right_metadata[[right_column]])
      ) |>
        dplyr::count(left_cluster, right_cluster, name = "nuclei") |>
        dplyr::group_by(left_cluster) |>
        dplyr::mutate(
          percent_of_left_cluster = 100 * nuclei / sum(nuclei)
        ) |>
        dplyr::ungroup() |>
        dplyr::mutate(
          comparison = comparison,
          resolution = resolution,
          left_column = left_column,
          right_column = right_column,
          .before = 1
        )
    }
  )
}

run_marker_audit <- function(object, cluster_column, analysis_scope) {
  Seurat::DefaultAssay(object) <- "RNA.crypt"
  Seurat::Idents(object) <- cluster_column

  markers <- Seurat::FindAllMarkers(
    object,
    assay = "RNA.crypt",
    only.pos = TRUE,
    test.use = "wilcox",
    min.pct = MARKER_MIN_PCT,
    logfc.threshold = MARKER_LOGFC_THRESHOLD,
    return.thresh = 0.05,
    verbose = FALSE
  )

  if (nrow(markers) == 0L) {
    warning("No markers passed thresholds for ", cluster_column)
    return(tibble::tibble())
  }

  fold_change_column <- intersect(
    c("avg_log2FC", "avg_logFC"),
    colnames(markers)
  )

  if (length(fold_change_column) != 1L) {
    stop("Could not identify the marker fold-change column.")
  }

  markers |>
    dplyr::mutate(
      analysis_scope = analysis_scope,
      cluster_column = cluster_column,
      cluster = as.character(cluster),
      gene = as.character(gene),
      effect_size = .data[[fold_change_column]],
      pct_delta = pct.1 - pct.2,
      .before = 1
    ) |>
    dplyr::filter(
      is.finite(effect_size),
      effect_size > 0,
      is.finite(p_val_adj),
      p_val_adj < MARKER_ADJUSTED_P_THRESHOLD
    )
}

top_marker_table <- function(markers, n = 30L) {
  if (nrow(markers) == 0L) {
    return(markers)
  }

  markers |>
    dplyr::filter(!grepl(TECHNICAL_GENE_PATTERN, gene)) |>
    dplyr::group_by(analysis_scope, cluster_column, cluster) |>
    dplyr::arrange(
      p_val_adj,
      dplyr::desc(pct_delta),
      dplyr::desc(effect_size),
      .by_group = TRUE
    ) |>
    dplyr::slice_head(n = n) |>
    dplyr::ungroup()
}

safe_z_score <- function(values) {
  finite_values <- is.finite(values)

  if (!any(finite_values)) {
    return(rep(NA_real_, length(values)))
  }

  value_sd <- stats::sd(values[finite_values])

  if (!is.finite(value_sd) || value_sd <= 0) {
    output <- rep(0, length(values))
    output[!finite_values] <- NA_real_
    return(output)
  }

  output <- (values - mean(values[finite_values])) / value_sd
  output[!finite_values] <- NA_real_
  output
}

make_targeted_rna_promoter_audit <- function(
  object,
  cluster_column,
  targeted_programs,
  promoter_window_bp = 2000L,
  rna_assay = "RNA.crypt",
  atac_assay = "ATAC.crypt"
) {
  message(
    "Running targeted RNA/promoter audit helper v5 ",
    "(direct-vector TSS mapping)."
  )

  cells <- colnames(object)
  target_genes <- unique(unlist(targeted_programs))

  rna_counts <- get_aligned_layer(
    object,
    assay = rna_assay,
    layer = "counts",
    cells = cells
  )
  rna_data <- get_aligned_layer(
    object,
    assay = rna_assay,
    layer = "data",
    cells = cells
  )
  atac_counts <- get_aligned_layer(
    object,
    assay = atac_assay,
    layer = "counts",
    cells = cells
  )

  annotation_ranges <- Signac::Annotation(object[[atac_assay]])

  if (is.null(annotation_ranges) || length(annotation_ranges) == 0L) {
    stop("The focused ATAC assay does not contain gene annotations.")
  }

  annotation_table <- as.data.frame(annotation_ranges)

  if (!"gene_name" %in% colnames(annotation_table)) {
    stop("The focused ATAC annotation lacks a gene_name column.")
  }

  gene_annotation <- annotation_table |>
    dplyr::filter(gene_name %in% target_genes)

  if (
    "type" %in% colnames(gene_annotation) &&
      any(gene_annotation$type == "gene", na.rm = TRUE)
  ) {
    gene_annotation <- gene_annotation |>
      dplyr::filter(type == "gene")
  }

  gene_annotation <- gene_annotation |>
    dplyr::mutate(
      strand = as.character(strand),
      tss = dplyr::if_else(strand == "-", end, start),
      promoter_start = pmax(1L, tss - promoter_window_bp),
      promoter_end = tss + promoter_window_bp
    ) |>
    dplyr::distinct(
      seqnames,
      tss,
      promoter_start,
      promoter_end,
      strand,
      gene_name
    )

  if (nrow(gene_annotation) == 0L) {
    stop("No targeted genes were found in the focused ATAC annotation.")
  }

  promoter_ranges <- GenomicRanges::GRanges(
    seqnames = gene_annotation$seqnames,
    ranges = IRanges::IRanges(
      start = gene_annotation$promoter_start,
      end = gene_annotation$promoter_end
    ),
    strand = gene_annotation$strand
  )

  peak_ranges <- GenomicRanges::granges(object[[atac_assay]])
  promoter_overlaps <- GenomicRanges::findOverlaps(
    peak_ranges,
    promoter_ranges,
    ignore.strand = TRUE
  )

  overlap_peak_indices <- S4Vectors::queryHits(promoter_overlaps)
  overlap_promoter_indices <- S4Vectors::subjectHits(promoter_overlaps)

  if (!all(c("gene_name", "tss") %in% colnames(gene_annotation))) {
    stop(
      "The promoter annotation table lacks gene_name or tss after ",
      "promoter construction.",
      call. = FALSE
    )
  }

  overlap_genes <- as.character(
    gene_annotation[["gene_name"]][overlap_promoter_indices]
  )
  overlap_tss <- as.numeric(
    gene_annotation[["tss"]][overlap_promoter_indices]
  )
  overlap_peak_starts <- BiocGenerics::start(peak_ranges)[
    overlap_peak_indices
  ]
  overlap_peak_ends <- BiocGenerics::end(peak_ranges)[
    overlap_peak_indices
  ]
  overlap_peak_centers <- floor(
    (overlap_peak_starts + overlap_peak_ends) / 2
  )

  expected_overlap_length <- length(overlap_peak_indices)
  overlap_vector_lengths <- c(
    genes = length(overlap_genes),
    tss = length(overlap_tss),
    peak_starts = length(overlap_peak_starts),
    peak_ends = length(overlap_peak_ends),
    peak_centers = length(overlap_peak_centers)
  )

  if (any(overlap_vector_lengths != expected_overlap_length)) {
    stop(
      "Promoter-overlap vectors are not aligned: ",
      paste(
        names(overlap_vector_lengths),
        overlap_vector_lengths,
        sep = "=",
        collapse = ", "
      ),
      call. = FALSE
    )
  }

  promoter_peak_map <- tibble::tibble(
    peak_index = overlap_peak_indices,
    promoter_index = overlap_promoter_indices,
    gene = overlap_genes,
    peak = rownames(atac_counts)[overlap_peak_indices],
    chromosome = as.character(
      GenomeInfoDb::seqnames(peak_ranges)[overlap_peak_indices]
    ),
    peak_start = overlap_peak_starts,
    peak_end = overlap_peak_ends,
    peak_center = overlap_peak_centers,
    tss_coordinate = overlap_tss,
    distance_to_tss = overlap_peak_centers - overlap_tss
  ) |>
    dplyr::distinct(gene, peak_index, .keep_all = TRUE)

  program_map <- purrr::imap_dfr(
    targeted_programs,
    ~ tibble::tibble(program = .y, gene = .x)
  ) |>
    dplyr::distinct(gene, .keep_all = TRUE)

  cell_metadata <- object[[]] |>
    tibble::rownames_to_column("cell_barcode") |>
    dplyr::transmute(
      cell_barcode,
      focused_subcluster = as.character(.data[[cluster_column]]),
      diet = as.character(diet)
    )

  total_atac_counts <- Matrix::colSums(atac_counts)

  cell_level <- purrr::map_dfr(
    target_genes,
    function(gene) {
      rna_is_available <- gene %in% rownames(rna_counts)

      if (rna_is_available) {
        rna_gene_counts <- as.numeric(rna_counts[gene, ])
        rna_gene_data <- as.numeric(rna_data[gene, ])
      } else {
        rna_gene_counts <- rep(0, length(cells))
        rna_gene_data <- rep(NA_real_, length(cells))
      }

      peak_indices <- promoter_peak_map$peak_index[
        promoter_peak_map$gene == gene
      ]

      if (length(peak_indices) > 0L) {
        promoter_counts <- Matrix::colSums(
          atac_counts[unique(peak_indices), , drop = FALSE]
        )
      } else {
        promoter_counts <- rep(0, length(cells))
      }

      tibble::tibble(
        cell_barcode = cells,
        gene = gene,
        RNA_available = rna_is_available,
        RNA_detected = rna_gene_counts > 0,
        RNA_log_normalized = rna_gene_data,
        promoter_peak_count = length(unique(peak_indices)),
        promoter_detected = promoter_counts > 0,
        promoter_log_normalized = log1p(
          10000 * promoter_counts / pmax(total_atac_counts, 1)
        )
      )
    }
  ) |>
    dplyr::left_join(program_map, by = "gene") |>
    dplyr::left_join(cell_metadata, by = "cell_barcode")

  group_columns <- c(
    "focused_subcluster",
    "diet",
    "program",
    "gene"
  )

  cluster_diet_summary <- cell_level |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_columns))) |>
    dplyr::summarise(
      nuclei = dplyr::n(),
      RNA_available = all(RNA_available),
      RNA_percent_detected = 100 * mean(RNA_detected),
      RNA_mean_log_normalized = mean(
        RNA_log_normalized,
        na.rm = TRUE
      ),
      promoter_peak_count = max(promoter_peak_count),
      promoter_percent_detected = 100 * mean(promoter_detected),
      promoter_mean_log_normalized = mean(promoter_log_normalized),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      RNA_mean_log_normalized = dplyr::if_else(
        is.nan(RNA_mean_log_normalized),
        NA_real_,
        RNA_mean_log_normalized
      )
    )

  cluster_summary <- cell_level |>
    dplyr::group_by(focused_subcluster, program, gene) |>
    dplyr::summarise(
      nuclei = dplyr::n(),
      RNA_available = all(RNA_available),
      RNA_percent_detected = 100 * mean(RNA_detected),
      RNA_mean_log_normalized = mean(
        RNA_log_normalized,
        na.rm = TRUE
      ),
      promoter_peak_count = max(promoter_peak_count),
      promoter_percent_detected = 100 * mean(promoter_detected),
      promoter_mean_log_normalized = mean(promoter_log_normalized),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      RNA_mean_log_normalized = dplyr::if_else(
        is.nan(RNA_mean_log_normalized),
        NA_real_,
        RNA_mean_log_normalized
      )
    )

  list(
    cluster_summary = cluster_summary,
    cluster_diet_summary = cluster_diet_summary,
    promoter_peak_map = promoter_peak_map
  )
}

make_targeted_evidence_dotplot <- function(
  summary_table,
  value_column,
  detection_column,
  title,
  color_label
) {
  plot_data <- summary_table |>
    dplyr::group_by(gene) |>
    dplyr::mutate(within_gene_z = safe_z_score(.data[[value_column]])) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      gene = factor(gene, levels = TARGETED_EVIDENCE_GENES),
      program = factor(
        program,
        levels = names(TARGETED_EVIDENCE_PROGRAMS)
      ),
      cluster_diet = paste0("C", focused_subcluster, "_", diet)
    )

  cluster_levels <- plot_data |>
    dplyr::distinct(focused_subcluster, diet, cluster_diet) |>
    dplyr::mutate(
      cluster_number = suppressWarnings(as.numeric(focused_subcluster)),
      diet_order = match(diet, c("CON", "HFD"))
    ) |>
    dplyr::arrange(cluster_number, focused_subcluster, diet_order) |>
    dplyr::pull(cluster_diet)

  plot_data$cluster_diet <- factor(
    plot_data$cluster_diet,
    levels = rev(unique(cluster_levels))
  )

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = gene,
      y = cluster_diet,
      size = .data[[detection_column]],
      fill = within_gene_z
    )
  ) +
    ggplot2::geom_point(shape = 21, color = "grey25", stroke = 0.25) +
    ggplot2::facet_grid(
      cols = ggplot2::vars(program),
      scales = "free_x",
      space = "free_x"
    ) +
    ggplot2::scale_size_continuous(
      range = c(0, 7),
      limits = c(0, 100),
      name = "Percent detected"
    ) +
    ggplot2::scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      na.value = "grey90",
      name = color_label
    ) +
    ggplot2::labs(
      title = title,
      subtitle = paste0(
        "Resolution ", resolution_tag(PRIMARY_REVIEW_RESOLUTION),
        "; color is standardized within each gene"
      ),
      x = NULL,
      y = "Focused cluster and diet"
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      panel.grid = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold")
    )
}

save_pdf <- function(filename, plot, width = 12, height = 8) {
  output_path <- file.path(STEP5E_PLOT_DIR, filename)
  grDevices::pdf(output_path, width = width, height = height, useDingbats = FALSE)
  print(plot)
  grDevices::dev.off()
  invisible(output_path)
}
```

# 4. Load and validate the reviewed WT object

```{r load-and-validate-wt-object}
multiome <- readRDS(STEP6_OBJECT_PATH)
metadata <- multiome[[]]

required_metadata <- c(
  "sample_id",
  "diet",
  "wnn_cluster",
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence",
  "nCount_RNA",
  "nCount_ATAC"
)

missing_metadata <- setdiff(required_metadata, colnames(metadata))

if (length(missing_metadata) > 0L) {
  stop(
    "The Step 6 object is missing metadata column(s): ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

if (!all(c("RNA", "ATAC") %in% names(multiome))) {
  stop("The Step 6 object must contain RNA and ATAC assays.")
}

if (!"counts" %in% SeuratObject::Layers(multiome[["RNA"]])) {
  stop("The Step 6 RNA assay lacks a joined counts layer.")
}

if (!"counts" %in% SeuratObject::Layers(multiome[["ATAC"]])) {
  stop("The Step 6 ATAC assay lacks a joined counts layer.")
}

metadata$wnn_cluster <- as.character(metadata$wnn_cluster)
metadata$diet <- as.character(metadata$diet)

if (!all(BOUNDARY_PARENT_CLUSTERS %in% unique(metadata$wnn_cluster))) {
  stop(
    "One or more requested focused parent clusters are absent: ",
    paste(
      setdiff(BOUNDARY_PARENT_CLUSTERS, unique(metadata$wnn_cluster)),
      collapse = ", "
    )
  )
}

if (!setequal(unique(metadata$diet), c("CON", "HFD"))) {
  stop("Expected exactly CON and HFD in the WT object.")
}

parent_annotation_audit <- metadata |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::filter(wnn_cluster %in% BOUNDARY_PARENT_CLUSTERS) |>
  dplyr::mutate(
    wnn_cluster = as.character(wnn_cluster),
    sample_id = as.character(sample_id),
    diet = as.character(diet),
    broad_compartment = as.character(broad_compartment),
    cell_state = as.character(cell_state),
    cell_state_full = as.character(cell_state_full),
    annotation_confidence = as.character(annotation_confidence)
  ) |>
  dplyr::count(
    wnn_cluster,
    sample_id,
    diet,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence,
    name = "nuclei"
  )

readr::write_csv(
  parent_annotation_audit,
  file.path(STEP5E_TABLE_DIR, "parent_cluster_annotation_audit.csv")
)

parent_annotation_audit
```

# 5. Derive independent source and curated program panels

```{r derive-source-programs}
source_marker_table <- readr::read_csv(
  SOURCE_MARKER_PATH,
  show_col_types = FALSE
)

SOURCE_PROGRAMS <- derive_source_programs(source_marker_table)
ALL_PROGRAMS <- c(CURATED_PROGRAMS, SOURCE_PROGRAMS)

program_definition_table <- purrr::imap_dfr(
  ALL_PROGRAMS,
  function(genes, program) {
    tibble::tibble(
      program = program,
      source = ifelse(
        grepl("^source_", program),
        "Gourab d21 integrated-cluster marker program",
        "curated biological program"
      ),
      gene = genes
    )
  }
)

readr::write_csv(
  program_definition_table,
  file.path(STEP5E_TABLE_DIR, "crypt_program_definitions.csv")
)

program_definition_table |>
  dplyr::count(program, source, name = "genes")
```

# 6. Build the core focused RNA, ATAC, and WNN representations

```{r build-core-focused-reductions}
core_fit <- build_focused_reductions(
  parent_object = multiome,
  parent_clusters = CORE_PARENT_CLUSTERS,
  reduction_prefix = "crypt.core",
  build_rna_graph = TRUE
)

core <- core_fit$object

core_grid <- add_resolution_grid(
  object = core,
  graph_name = core_fit$wsnn,
  column_prefix = CORE_WNN_PREFIX
)
core <- core_grid$object
core_wnn_columns <- core_grid$columns

core_rna_grid <- add_resolution_grid(
  object = core,
  graph_name = core_fit$rna_snn,
  column_prefix = CORE_RNA_PREFIX
)
core <- core_rna_grid$object
core_rna_columns <- core_rna_grid$columns

core_fit$object <- NULL
invisible(gc())
```

# 7. Add cell-cycle and biological support after clustering

```{r score-core-programs}
Seurat::DefaultAssay(core) <- "RNA.crypt"

s_genes <- Seurat::CaseMatch(
  search = Seurat::cc.genes.updated.2019$s.genes,
  match = rownames(core[["RNA.crypt"]])
)
g2m_genes <- Seurat::CaseMatch(
  search = Seurat::cc.genes.updated.2019$g2m.genes,
  match = rownames(core[["RNA.crypt"]])
)

s_genes <- unique(stats::na.omit(s_genes))
g2m_genes <- unique(stats::na.omit(g2m_genes))

if (length(s_genes) < 20L || length(g2m_genes) < 20L) {
  stop("Too few cell-cycle genes were available for CellCycleScoring.")
}

core <- Seurat::CellCycleScoring(
  core,
  s.features = s_genes,
  g2m.features = g2m_genes,
  set.ident = FALSE
)

core_program_result <- add_program_scores(
  object = core,
  assay = "RNA.crypt",
  programs = ALL_PROGRAMS
)

core <- core_program_result$object
program_score_columns <- paste0(names(ALL_PROGRAMS), "_score")
program_detection_columns <- paste0(
  names(ALL_PROGRAMS),
  "_genes_detected"
)

readr::write_csv(
  core_program_result$gene_summary,
  file.path(STEP5E_TABLE_DIR, "crypt_program_gene_availability.csv")
)
```

# 8. Build the expanded-boundary WNN sensitivity

```{r build-boundary-sensitivity}
boundary_fit <- build_focused_reductions(
  parent_object = multiome,
  parent_clusters = BOUNDARY_PARENT_CLUSTERS,
  reduction_prefix = "crypt.boundary",
  build_rna_graph = FALSE
)

boundary <- boundary_fit$object

boundary_grid <- add_resolution_grid(
  object = boundary,
  graph_name = boundary_fit$wsnn,
  column_prefix = BOUNDARY_WNN_PREFIX
)
boundary <- boundary_grid$object
boundary_wnn_columns <- boundary_grid$columns

Seurat::DefaultAssay(boundary) <- "RNA.crypt"
boundary <- Seurat::CellCycleScoring(
  boundary,
  s.features = intersect(s_genes, rownames(boundary[["RNA.crypt"]])),
  g2m.features = intersect(g2m_genes, rownames(boundary[["RNA.crypt"]])),
  set.ident = FALSE
)

boundary_program_result <- add_program_scores(
  object = boundary,
  assay = "RNA.crypt",
  programs = ALL_PROGRAMS
)

boundary <- boundary_program_result$object
boundary_fit$object <- NULL
invisible(gc())
```

# 9. Summarize resolutions, parent origins, and sensitivity agreement

```{r summarize-resolution-grid}
core_wnn_summary <- summarise_resolution_grid(
  object = core,
  resolution_columns = core_wnn_columns,
  analysis_scope = "core_WNN_0_6_8",
  rna_weight_column = "crypt.core.RNA.weight",
  atac_weight_column = "crypt.core.ATAC.weight",
  score_columns = c(program_score_columns, program_detection_columns)
)

core_rna_summary <- summarise_resolution_grid(
  object = core,
  resolution_columns = core_rna_columns,
  analysis_scope = "core_RNA_only_0_6_8",
  rna_weight_column = "crypt.core.RNA.weight",
  atac_weight_column = "crypt.core.ATAC.weight",
  score_columns = c(program_score_columns, program_detection_columns)
)

boundary_wnn_summary <- summarise_resolution_grid(
  object = boundary,
  resolution_columns = boundary_wnn_columns,
  analysis_scope = "boundary_WNN_0_1_6_8_18",
  rna_weight_column = "crypt.boundary.RNA.weight",
  atac_weight_column = "crypt.boundary.ATAC.weight",
  score_columns = c(program_score_columns, program_detection_columns)
)

resolution_cluster_summary <- dplyr::bind_rows(
  core_wnn_summary$cluster_summary,
  core_rna_summary$cluster_summary,
  boundary_wnn_summary$cluster_summary
)

resolution_parent_composition <- dplyr::bind_rows(
  core_wnn_summary$parent_composition,
  core_rna_summary$parent_composition,
  boundary_wnn_summary$parent_composition
)

adjacent_resolution_crosswalk <- dplyr::bind_rows(
  make_adjacent_resolution_crosswalk(
    core,
    core_wnn_columns,
    "core_WNN_0_6_8"
  ),
  make_adjacent_resolution_crosswalk(
    core,
    core_rna_columns,
    "core_RNA_only_0_6_8"
  ),
  make_adjacent_resolution_crosswalk(
    boundary,
    boundary_wnn_columns,
    "boundary_WNN_0_1_6_8_18"
  )
)

core_wnn_vs_rna_crosswalk <- make_same_resolution_crosswalk(
  left_object = core,
  left_columns = core_wnn_columns,
  right_object = core,
  right_columns = core_rna_columns,
  comparison = "core_WNN_vs_RNA"
)

core_vs_boundary_crosswalk <- make_same_resolution_crosswalk(
  left_object = core,
  left_columns = core_wnn_columns,
  right_object = boundary,
  right_columns = boundary_wnn_columns,
  comparison = "core_WNN_vs_boundary_WNN"
)

readr::write_csv(
  resolution_cluster_summary,
  file.path(STEP5E_TABLE_DIR, "focused_resolution_cluster_summary.csv")
)
readr::write_csv(
  resolution_parent_composition,
  file.path(STEP5E_TABLE_DIR, "focused_resolution_parent_composition.csv")
)
readr::write_csv(
  adjacent_resolution_crosswalk,
  file.path(STEP5E_TABLE_DIR, "adjacent_resolution_crosswalk.csv")
)
readr::write_csv(
  core_wnn_vs_rna_crosswalk,
  file.path(STEP5E_TABLE_DIR, "core_WNN_vs_RNA_crosswalk.csv")
)
readr::write_csv(
  core_vs_boundary_crosswalk,
  file.path(STEP5E_TABLE_DIR, "core_vs_boundary_WNN_crosswalk.csv")
)
readr::write_csv(
  resolution_cluster_summary |>
    dplyr::filter(
      analysis_scope == "core_WNN_0_6_8",
      resolution == PRIMARY_REVIEW_RESOLUTION
    ),
  file.path(STEP5E_TABLE_DIR, "primary_r0.3_cluster_summary.csv")
)
readr::write_csv(
  resolution_parent_composition |>
    dplyr::filter(
      analysis_scope == "core_WNN_0_6_8",
      resolution == PRIMARY_REVIEW_RESOLUTION
    ),
  file.path(STEP5E_TABLE_DIR, "primary_r0.3_parent_composition.csv")
)

resolution_cluster_summary |>
  dplyr::select(
    analysis_scope,
    resolution,
    focused_subcluster,
    nuclei,
    CON_nuclei,
    HFD_nuclei,
    S_phase_percent,
    G2M_phase_percent,
    dplyr::contains("canonical_ISC_score_median"),
    dplyr::contains("crypt_progenitor_score_median")
  )
```

# 10. Focused RNA marker audit at the primary and sensitivity resolutions

Resolution 0.3 is the primary review candidate because it preserves the major
stable branches without the unsupported splits that appear at higher
resolutions. Resolutions 0.4 and 0.6 remain marker-level sensitivities. Marker
exports are explicitly restricted to positive genes with adjusted P values
below 0.05. No annotation override is applied in this notebook.

```{r focused-marker-audit}
marker_tables <- purrr::map_dfr(
  MARKER_REVIEW_RESOLUTIONS,
  function(resolution) {
    cluster_column <- resolution_column(CORE_WNN_PREFIX, resolution)
    run_marker_audit(
      object = core,
      cluster_column = cluster_column,
      analysis_scope = paste0("core_WNN_resolution_", resolution_tag(resolution))
    )
  }
)

top_marker_tables <- top_marker_table(marker_tables, n = 30L)

marker_count_summary <- purrr::map_dfr(
  MARKER_REVIEW_RESOLUTIONS,
  function(resolution) {
    cluster_column <- resolution_column(CORE_WNN_PREFIX, resolution)

    tibble::tibble(
      resolution = resolution,
      cluster_column = cluster_column,
      cluster = sort(unique(as.character(core[[]][[cluster_column]])))
    )
  }
) |>
  dplyr::left_join(
    marker_tables |>
      dplyr::count(cluster_column, cluster, name = "significant_positive_markers"),
    by = c("cluster_column", "cluster")
  ) |>
  dplyr::mutate(
    significant_positive_markers = tidyr::replace_na(
      significant_positive_markers,
      0L
    )
  )

readr::write_csv(
  marker_tables,
  file.path(STEP5E_TABLE_DIR, "core_WNN_positive_markers_review_resolutions.csv.gz")
)
readr::write_csv(
  top_marker_tables,
  file.path(STEP5E_TABLE_DIR, "core_WNN_top_markers_review_resolutions.csv")
)
readr::write_csv(
  marker_count_summary,
  file.path(STEP5E_TABLE_DIR, "core_WNN_marker_count_review_resolutions.csv")
)

top_marker_tables |>
  dplyr::select(
    cluster_column,
    cluster,
    gene,
    effect_size,
    pct.1,
    pct.2,
    p_val_adj
  )
```

# 11. Export cell-level audit assignments

```{r export-cell-level-assignments}
core_assignment_columns <- c(
  "sample_id",
  "diet",
  "parent_wnn_cluster",
  "cell_state",
  "cell_state_full",
  "annotation_confidence",
  "nCount_RNA",
  "nFeature_RNA",
  "nCount_ATAC",
  "nFeature_ATAC",
  "percent.mt",
  "TSS.enrichment",
  "nucleosome_signal",
  "frip_cr",
  "scDblFinder.score",
  "crypt.core.RNA.weight",
  "crypt.core.ATAC.weight",
  "Phase",
  "S.Score",
  "G2M.Score",
  core_wnn_columns,
  core_rna_columns,
  program_score_columns,
  program_detection_columns
)

core_assignments <- core[[]] |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::select(
    cell_barcode,
    dplyr::any_of(core_assignment_columns)
  )

boundary_assignment_columns <- c(
  "sample_id",
  "diet",
  "parent_wnn_cluster",
  "cell_state",
  "cell_state_full",
  "annotation_confidence",
  "nCount_RNA",
  "nCount_ATAC",
  "crypt.boundary.RNA.weight",
  "crypt.boundary.ATAC.weight",
  "Phase",
  "S.Score",
  "G2M.Score",
  boundary_wnn_columns,
  program_score_columns,
  program_detection_columns
)

boundary_assignments <- boundary[[]] |>
  tibble::rownames_to_column("cell_barcode") |>
  dplyr::select(
    cell_barcode,
    dplyr::any_of(boundary_assignment_columns)
  )

readr::write_csv(
  core_assignments,
  file.path(STEP5E_TABLE_DIR, "core_cell_assignments.csv.gz")
)
readr::write_csv(
  boundary_assignments,
  file.path(STEP5E_TABLE_DIR, "boundary_cell_assignments.csv.gz")
)
```

# 12. Parent, modality, and diet overview plots

```{r plot-focused-overview}
p_core_parent <- Seurat::DimPlot(
  core,
  reduction = core_fit$wnn_umap,
  group.by = "parent_wnn_cluster",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) +
  ggplot2::labs(
    title = "Core focused WNN: original parent clusters",
    subtitle = "WT global clusters 0, 6, and 8"
  ) +
  ggplot2::theme_void(base_size = 14)

p_core_diet <- Seurat::DimPlot(
  core,
  reduction = core_fit$wnn_umap,
  group.by = "diet",
  split.by = "diet",
  raster = FALSE
) +
  ggplot2::labs(
    title = "Core focused WNN by diet",
    subtitle = "Diet is descriptive because library and diet are confounded"
  ) +
  ggplot2::theme_void(base_size = 14)

p_core_modality <- Seurat::FeaturePlot(
  core,
  reduction = core_fit$wnn_umap,
  features = c(
    "crypt.core.RNA.weight",
    "crypt.core.ATAC.weight"
  ),
  min.cutoff = "q02",
  max.cutoff = "q98",
  raster = FALSE,
  ncol = 2
) +
  patchwork::plot_annotation(
    title = "Focused WNN modality weights"
  )

save_pdf(
  "01_core_WNN_parent_clusters.pdf",
  p_core_parent,
  width = 10,
  height = 8
)
save_pdf(
  "02_core_WNN_by_diet.pdf",
  p_core_diet,
  width = 14,
  height = 7
)
save_pdf(
  "03_core_WNN_modality_weights.pdf",
  p_core_modality,
  width = 14,
  height = 7
)
```

# 13. Resolution grids and RNA-only sensitivity

```{r plot-resolution-grids}
for (resolution_index in seq_along(FOCUSED_RESOLUTIONS)) {
  resolution <- FOCUSED_RESOLUTIONS[[resolution_index]]
  wnn_column <- core_wnn_columns[[resolution_index]]
  rna_column <- core_rna_columns[[resolution_index]]
  boundary_column <- boundary_wnn_columns[[resolution_index]]

  p_core_wnn <- Seurat::DimPlot(
    core,
    reduction = core_fit$wnn_umap,
    group.by = wnn_column,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  ) +
    ggplot2::labs(
      title = paste0("Core WNN resolution ", resolution_tag(resolution))
    ) +
    ggplot2::theme_void(base_size = 14)

  p_core_rna <- Seurat::DimPlot(
    core,
    reduction = core_fit$rna_umap,
    group.by = rna_column,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  ) +
    ggplot2::labs(
      title = paste0("Core RNA-only resolution ", resolution_tag(resolution))
    ) +
    ggplot2::theme_void(base_size = 14)

  p_comparison <- p_core_wnn + p_core_rna +
    patchwork::plot_annotation(
      title = paste0(
        "Core graph sensitivity at resolution ",
        resolution_tag(resolution)
      ),
      subtitle = "The two panels contain exactly the same nuclei"
    )

  save_pdf(
    paste0(
      "04_core_WNN_vs_RNA_resolution_",
      resolution_tag(resolution),
      ".pdf"
    ),
    p_comparison,
    width = 15,
    height = 7
  )

  p_boundary <- Seurat::DimPlot(
    boundary,
    reduction = boundary_fit$wnn_umap,
    group.by = boundary_column,
    label = TRUE,
    repel = TRUE,
    raster = FALSE
  ) +
    ggplot2::labs(
      title = paste0(
        "Expanded-boundary WNN resolution ",
        resolution_tag(resolution)
      ),
      subtitle = "Global clusters 0, 1, 6, 8, and 18"
    ) +
    ggplot2::theme_void(base_size = 14)

  save_pdf(
    paste0(
      "05_boundary_WNN_resolution_",
      resolution_tag(resolution),
      ".pdf"
    ),
    p_boundary,
    width = 10,
    height = 8
  )
}
```

# 14. Program-score and cell-cycle support

```{r plot-program-support}
program_plot_features <- c(
  "curated_canonical_ISC_score",
  "curated_cell_cycle_score",
  "curated_crypt_progenitor_score",
  "curated_early_absorptive_score",
  "curated_secretory_commitment_score",
  "curated_Reg4_DSC_score"
)

p_curated_programs <- Seurat::FeaturePlot(
  core,
  reduction = core_fit$wnn_umap,
  features = program_plot_features,
  min.cutoff = "q02",
  max.cutoff = "q98",
  raster = FALSE,
  ncol = 3,
  keep.scale = "feature"
) +
  patchwork::plot_annotation(
    title = "Independent curated programs on the core focused WNN"
  )

source_plot_features <- paste0(
  names(SOURCE_PROGRAMS),
  "_score"
)

p_source_programs <- Seurat::FeaturePlot(
  core,
  reduction = core_fit$wnn_umap,
  features = source_plot_features,
  min.cutoff = "q02",
  max.cutoff = "q98",
  raster = FALSE,
  ncol = 3,
  keep.scale = "feature"
) +
  patchwork::plot_annotation(
    title = "Gourab d21 integrated-cluster programs on WT crypt nuclei"
  )

p_cell_cycle <- Seurat::DimPlot(
  core,
  reduction = core_fit$wnn_umap,
  group.by = "Phase",
  raster = FALSE
) +
  ggplot2::labs(title = "Cell-cycle phase on the core focused WNN") +
  ggplot2::theme_void(base_size = 14)

save_pdf(
  "06_core_curated_program_support.pdf",
  p_curated_programs,
  width = 16,
  height = 11
)
save_pdf(
  "07_core_source_program_support.pdf",
  p_source_programs,
  width = 16,
  height = 11
)
save_pdf(
  "08_core_cell_cycle_phase.pdf",
  p_cell_cycle,
  width = 10,
  height = 8
)
```

# 15. Marker dot plots for manual review

```{r plot-marker-audits}
marker_audit_genes <- unique(c(
  "Lgr5", "Ascl2", "Smoc2", "Axin2", "Olfm4", "Lrig1", "Rnf43",
  "Mki67", "Top2a", "Cenpf", "Ube2c", "Cdk1", "Ccnb1",
  "Dmbt1", "Cftr", "Htr4", "Kcnq1", "Slc12a2", "Hmga2",
  "Hmgcs2", "Hsd3b3",
  "Slc26a3", "Slc9a2", "Car4", "Aqp8", "Krt20", "Mxd1",
  "Atoh1", "Dll1", "Spdef", "Muc2", "Reg4", "Spink4",
  "Slc12a8", "Kcnma1", "Sytl2"
))

marker_audit_genes <- intersect(
  marker_audit_genes,
  rownames(core[["RNA.crypt"]])
)

for (resolution in MARKER_REVIEW_RESOLUTIONS) {
  cluster_column <- resolution_column(CORE_WNN_PREFIX, resolution)
  plot_group_column <- paste0(cluster_column, "_diet")

  core[[plot_group_column]] <- paste0(
    "C",
    core[[]][[cluster_column]],
    "_",
    as.character(core$diet)
  )

  p_markers <- Seurat::DotPlot(
    core,
    assay = "RNA.crypt",
    features = marker_audit_genes,
    group.by = plot_group_column,
    dot.scale = 7
  ) +
    Seurat::RotatedAxis() +
    ggplot2::labs(
      title = paste0(
        "Core WNN crypt marker audit: resolution ",
        resolution_tag(resolution)
      ),
      subtitle = "Each focused subcluster is displayed separately by diet",
      x = NULL,
      y = "Focused cluster and diet"
    ) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(size = 8),
      axis.text.y = ggplot2::element_text(size = 10)
    )

  save_pdf(
    paste0(
      "09_core_marker_audit_resolution_",
      resolution_tag(resolution),
      ".pdf"
    ),
    p_markers,
    width = 17,
    height = 9
  )
}
```

# 16. Targeted RNA and exact-promoter evidence at resolution 0.3

This audit directly pairs RNA detection with accessibility of consensus peaks
overlapping the annotated TSS plus or minus 2 kb. Promoter accessibility is
supporting regulatory evidence rather than proof of identity. Color is
standardized within each gene; dot size reports the percentage of nuclei with
detectable RNA or at least one promoter-overlapping ATAC fragment.

```{r targeted-rna-promoter-evidence}
primary_cluster_column <- resolution_column(
  CORE_WNN_PREFIX,
  PRIMARY_REVIEW_RESOLUTION
)

targeted_evidence <- make_targeted_rna_promoter_audit(
  object = core,
  cluster_column = primary_cluster_column,
  targeted_programs = TARGETED_EVIDENCE_PROGRAMS,
  promoter_window_bp = PROMOTER_WINDOW_BP
)

readr::write_csv(
  targeted_evidence$cluster_summary,
  file.path(
    STEP5E_TABLE_DIR,
    "primary_r0.3_targeted_RNA_promoter_summary_by_cluster.csv"
  )
)
readr::write_csv(
  targeted_evidence$cluster_diet_summary,
  file.path(
    STEP5E_TABLE_DIR,
    "primary_r0.3_targeted_RNA_promoter_summary_by_cluster_diet.csv"
  )
)
readr::write_csv(
  targeted_evidence$promoter_peak_map,
  file.path(
    STEP5E_TABLE_DIR,
    "primary_r0.3_targeted_promoter_peak_map.csv"
  )
)

p_targeted_rna <- make_targeted_evidence_dotplot(
  summary_table = targeted_evidence$cluster_diet_summary,
  value_column = "RNA_mean_log_normalized",
  detection_column = "RNA_percent_detected",
  title = "Targeted RNA evidence for the primary focused solution",
  color_label = "Within-gene RNA\nZ score"
)

p_targeted_promoter <- make_targeted_evidence_dotplot(
  summary_table = targeted_evidence$cluster_diet_summary,
  value_column = "promoter_mean_log_normalized",
  detection_column = "promoter_percent_detected",
  title = "Exact-promoter ATAC evidence for the primary focused solution",
  color_label = "Within-gene ATAC\nZ score"
)

save_pdf(
  "10_primary_r0.3_targeted_RNA_evidence.pdf",
  p_targeted_rna,
  width = 18,
  height = 9
)
save_pdf(
  "11_primary_r0.3_targeted_promoter_ATAC_evidence.pdf",
  p_targeted_promoter,
  width = 18,
  height = 9
)
```

# 17. Cluster-standardized program summaries

Module-score magnitudes are not compared between gene sets. Within each
resolution, every program is standardized across the focused clusters. The
primary resolution receives a dedicated summary, while the full grid remains
available as a stability visualization.

```{r plot-program-summary-heatmaps}
program_score_summary_long <- resolution_cluster_summary |>
  dplyr::select(
    analysis_scope,
    resolution,
    focused_subcluster,
    nuclei,
    dplyr::ends_with("_score_median")
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::ends_with("_score_median"),
    names_to = "program",
    values_to = "median_score"
  ) |>
  dplyr::mutate(
    program = sub("_score_median$", "", program),
    program_source = dplyr::if_else(
      grepl("^curated_", program),
      "Curated programs",
      "Gourab d21 source programs"
    ),
    cluster_label = paste0(
      "r", resolution_tag(resolution),
      ":C", focused_subcluster,
      " (n=", nuclei, ")"
    ),
    program = factor(
      program,
      levels = c(names(CURATED_PROGRAMS), names(SOURCE_PROGRAMS))
    )
  ) |>
  dplyr::group_by(analysis_scope, resolution, program) |>
  dplyr::mutate(program_z = safe_z_score(median_score)) |>
  dplyr::ungroup()

readr::write_csv(
  program_score_summary_long,
  file.path(STEP5E_TABLE_DIR, "focused_program_score_summary_long.csv")
)

for (scope in unique(program_score_summary_long$analysis_scope)) {
  plot_data <- program_score_summary_long |>
    dplyr::filter(analysis_scope == scope) |>
    dplyr::arrange(resolution, suppressWarnings(as.numeric(focused_subcluster)))

  plot_data$cluster_label <- factor(
    plot_data$cluster_label,
    levels = rev(unique(plot_data$cluster_label))
  )

  p_heatmap <- ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = program,
      y = cluster_label,
      fill = program_z
    )
  ) +
    ggplot2::geom_tile(color = "white", linewidth = 0.2) +
    ggplot2::scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      name = "Within-program\nZ score"
    ) +
    ggplot2::facet_grid(
      cols = ggplot2::vars(program_source),
      scales = "free_x",
      space = "free_x"
    ) +
    ggplot2::labs(
      title = paste0("Program support across resolutions: ", scope),
      subtitle = "Each program is standardized across clusters within a resolution",
      x = NULL,
      y = "Resolution and focused cluster"
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(
        angle = 45,
        hjust = 1,
        vjust = 1
      ),
      panel.grid = ggplot2::element_blank()
    )

  save_pdf(
    paste0(
      "10_program_heatmap_",
      gsub("[^A-Za-z0-9]+", "_", scope),
      ".pdf"
    ),
    p_heatmap,
    width = 15,
    height = 12
  )
}

primary_program_plot_data <- program_score_summary_long |>
  dplyr::filter(
    analysis_scope == "core_WNN_0_6_8",
    resolution == PRIMARY_REVIEW_RESOLUTION
  ) |>
  dplyr::arrange(suppressWarnings(as.numeric(focused_subcluster)))

primary_program_plot_data$cluster_label <- factor(
  primary_program_plot_data$cluster_label,
  levels = rev(unique(primary_program_plot_data$cluster_label))
)

p_primary_program_heatmap <- ggplot2::ggplot(
  primary_program_plot_data,
  ggplot2::aes(x = program, y = cluster_label, fill = program_z)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.3) +
  ggplot2::scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Within-program\nZ score"
  ) +
  ggplot2::facet_grid(
    cols = ggplot2::vars(program_source),
    scales = "free_x",
    space = "free_x"
  ) +
  ggplot2::labs(
    title = "Program support in the primary focused WNN solution",
    subtitle = paste0(
      "Resolution ", resolution_tag(PRIMARY_REVIEW_RESOLUTION),
      "; compare clusters within a program, not magnitudes between programs"
    ),
    x = NULL,
    y = "Focused cluster"
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    panel.grid = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )

save_pdf(
  "12_primary_r0.3_program_heatmap.pdf",
  p_primary_program_heatmap,
  width = 16,
  height = 7
)
```

# 18. Save review objects and completion records

```{r save-review-checkpoints}
feature_filter_summary <- dplyr::bind_rows(
  core_fit$feature_filter_summary,
  boundary_fit$feature_filter_summary
)

readr::write_csv(
  feature_filter_summary,
  file.path(STEP5E_TABLE_DIR, "focused_ATAC_feature_filter_summary.csv")
)

saveRDS(
  core,
  file.path(STEP5E_OBJECT_DIR, "WT_crypt_core_0_6_8_focused.rds"),
  compress = FALSE
)
saveRDS(
  boundary,
  file.path(
    STEP5E_OBJECT_DIR,
    "WT_crypt_boundary_0_1_6_8_18_focused.rds"
  ),
  compress = FALSE
)

step5e_summary <- tibble::tibble(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  cohort_group = COHORT_GROUP,
  analysis_version = ANALYSIS_VERSION,
  core_parent_clusters = paste(CORE_PARENT_CLUSTERS, collapse = ";"),
  core_nuclei = ncol(core),
  boundary_parent_clusters = paste(
    BOUNDARY_PARENT_CLUSTERS,
    collapse = ";"
  ),
  boundary_nuclei = ncol(boundary),
  resolutions_tested = paste(
    vapply(FOCUSED_RESOLUTIONS, resolution_tag, character(1)),
    collapse = ";"
  ),
  marker_review_resolutions = paste(
    vapply(MARKER_REVIEW_RESOLUTIONS, resolution_tag, character(1)),
    collapse = ";"
  ),
  primary_review_resolution = PRIMARY_REVIEW_RESOLUTION,
  source_programs = length(SOURCE_PROGRAMS),
  curated_programs = length(CURATED_PROGRAMS),
  targeted_RNA_promoter_audit_completed = TRUE,
  annotation_override_applied = FALSE,
  biological_diet_inference_performed = FALSE
)

readr::write_csv(
  step5e_summary,
  file.path(STEP5E_DIR, "step5e_summary.csv")
)

writeLines(
  c(
    paste("Step 5E completed:", Sys.time()),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste("Core parent clusters:", paste(CORE_PARENT_CLUSTERS, collapse = ",")),
    paste("Core nuclei:", ncol(core)),
    paste(
      "Boundary parent clusters:",
      paste(BOUNDARY_PARENT_CLUSTERS, collapse = ",")
    ),
    paste("Boundary nuclei:", ncol(boundary)),
    paste(
      "Resolutions tested:",
      paste(FOCUSED_RESOLUTIONS, collapse = ",")
    ),
    paste("Primary review resolution:", PRIMARY_REVIEW_RESOLUTION),
    paste(
      "Marker review resolutions:",
      paste(MARKER_REVIEW_RESOLUTIONS, collapse = ",")
    ),
    "Core WNN analysis completed: TRUE",
    "Core RNA-only sensitivity completed: TRUE",
    "Expanded boundary sensitivity completed: TRUE",
    "Targeted RNA and exact-promoter ATAC audit completed: TRUE",
    "Annotation override applied: FALSE",
    "Biological diet inference performed: FALSE"
  ),
  con = file.path(STEP5E_DIR, "STEP_05E_COMPLETE.txt")
)

writeLines(
  c(
    "Step 5E is ready for focused manual review.",
    "No global or Step 6 annotation has been changed.",
    "Review cluster stability across adjacent resolutions, WNN-versus-RNA ",
    "agreement, boundary sensitivity, parent origins, marker coherence, ",
    "cell-cycle composition, modality weights, QC, and diet representation.",
    "A proposed refined label should not be promoted solely because it has ",
    "the highest relative module score in a sparse comparison."
  ),
  con = file.path(STEP5E_DIR, "STEP_05E_READY_FOR_REVIEW.txt")
)

step5e_summary
```

# 19. Manual review rules

Do not select a focused resolution or child label from UMAP geometry alone.
A candidate canonical ISC, cycling stem/early-TA, or crypt-progenitor child
should satisfy all of the following:

1. it is stable across adjacent resolutions rather than appearing only once;
2. it has a coherent detection-supported RNA marker program, not merely a high
   relative module score;
3. its broad biological program remains interpretable in the RNA-only
   sensitivity, while exact one-to-one RNA/WNN cluster correspondence is not
   required;
4. the same core nuclei remain coherent when clusters 1 and 18 are added;
5. its QC and modality weights do not indicate a low-complexity or
   single-modality artifact;
6. its cell-cycle composition is explicitly reported, without treating
   proliferation as either proof against stemness or proof of stemness; and
7. RNA expression and exact-promoter accessibility are evaluated separately;
   promoter accessibility is supporting evidence and is not treated as proof
   of expression or identity; and
8. both diet libraries contribute enough nuclei for later descriptive
   diet-effect comparisons. A one-diet child may be biologically interesting,
   but it cannot anchor a within-state CON-versus-HFD comparison here.

If a stable refinement is supported, record that decision in a new reviewed
configuration table and apply it in a later workflow step. Step 5E itself is
provenance-preserving and never mutates the approved Step 6 object.
