---
title: "Step 6 - Annotate mouse-colon cell states"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 5 converted the RNA and ATAC representations into a weighted
nearest-neighbor graph and selected resolution 0.8 after manual review. That
resolution produced 27 computational clusters (0-26). 

Step 6 asks a different question: 
**what biological compartment or cell state is supported for each cluster?**

This notebook deliberately separates evidence generation from label approval.
It will:

1. load and validate the finalized Step 5 object and its recorded clustering
   decision;
2. inventory all 27 WNN clusters and their descriptive library composition;
3. discover positive RNA markers without treating a single gene as sufficient
   evidence;
4. audit broad epithelial, immune, vascular, stromal, and neural programs;
5. resolve epithelial states, including absorptive, goblet, Reg4-positive deep
   crypt secretory, enteroendocrine, and tuft programs;
6. adjudicate the candidate crypt clusters using per-cell co-detection,
   cell-cycle state, pairwise RNA contrasts, cross-library consistency, and
   targeted ATAC coverage;
7. write a proposed annotation table for manual review;
8. stop until an approved annotation decision table is present; and
9. apply approved labels, create final annotation plots, and save an annotated
   multimodal object.

A practical SOL allocation is **4 CPU cores, 96 GiB memory, 12–24 hours,
R 4.4.2, no GPU**. Marker discovery and ATAC coverage are the most expensive
parts of this notebook.

## Statistical limitation

There is one pooled CON library and one pooled HFD library. Individual mouse
identities are not recoverable after pooling, so diet and library remain
perfectly confounded. Cross-library consistency is useful annotation evidence,
but this notebook does **not** perform replicated CON-versus-HFD inference.

## Annotation philosophy

- WNN clusters are computational partitions, not biological truths.
- A label should be supported by a coherent positive marker program and the
  absence of a convincing competing lineage program.
- Cycling status is recorded as a state layered onto lineage identity; it is
  not regressed away or automatically treated as a separate lineage.
- Rare populations remain plausible when supported by multiple concordant
  markers and appropriate multimodal structure.
- Ambiguous clusters remain explicitly ambiguous until the evidence improves.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
})

ggplot2::theme_set(theme_multiome())

sample_sheet <- read_sample_sheet()

unknown_diets <- setdiff(unique(sample_sheet$diet), names(DIET_COLORS))

if (length(unknown_diets) > 0L) {
  stop("No plotting colors defined for: ",
       paste(unknown_diets, collapse = ", "),
       call. = FALSE)
}

sample_colors <- stats::setNames(unname(DIET_COLORS[sample_sheet$diet]), sample_sheet$sample_id)

configure_parallel_plan()

cat("Configured future workers:", future::nbrOfWorkers(), "\n")
```

# 2. Define Step 6 outputs and review parameters

The annotation decision table is dataset-version specific. The notebook writes
a proposal into the results directory, but the manually reviewed table belongs
under `config/datasets/<dataset version>/` so that the decision is versioned
with the analysis code.

The focused cluster sets below are **review scopes, not final labels**. They
encode the hypotheses raised during Step 5 review and make those hypotheses
easy to challenge with additional evidence.

```{r step-six-parameters}
STEP5_DIR <- file.path(OUTPUT_DIR, "05_wnn")
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP6_PLOT_DIR <- file.path(STEP6_DIR, "plots")
STEP6_MARKER_DIR <- file.path(STEP6_DIR, "markers")

dir.create(STEP6_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP6_PLOT_DIR,
           recursive = TRUE,
           showWarnings = FALSE)
dir.create(STEP6_MARKER_DIR,
           recursive = TRUE,
           showWarnings = FALSE)

ANNOTATION_DECISION_PATH <- file.path(DATASET_CONFIG_DIR, "cell_state_annotations.csv")

ANNOTATION_PROPOSAL_PATH <- file.path(STEP6_DIR, "cell_state_annotations_proposed.csv")

MIN_MARKER_PCT <- 0.10
MIN_MARKER_LOG2FC <- 0.25
TOP_MARKERS_PER_CLUSTER <- 15L

CRYPT_REVIEW_CLUSTERS <- c("3", "6", "19")

EPITHELIAL_REVIEW_CLUSTERS <- c("3", "5", "6", "7", "8", "9", "13", "14", "17", "19", "26")

AMBIGUOUS_REVIEW_CLUSTERS <- c("10", "11", "12", "22", "23", "24", "25")

RUN_TARGETED_ATAC_COVERAGE <- TRUE

step6_parameters <- data.frame(
  parameter = c(
    "minimum marker detection fraction",
    "minimum marker log2 fold change",
    "top markers retained per cluster",
    "targeted ATAC coverage enabled"
  ),
  value = c(
    MIN_MARKER_PCT,
    MIN_MARKER_LOG2FC,
    TOP_MARKERS_PER_CLUSTER,
    RUN_TARGETED_ATAC_COVERAGE
  )
)

step6_parameters
```

# 3. Load and validate the Step 5 checkpoint

Step 6 reads the clustering decision from the Step 5 completion marker rather
than silently redefining it. This prevents a future edit from annotating one
resolution while believing it is using another.

```{r load-step-five}
step5_complete_marker <- file.path(STEP5_DIR, "STEP_05_COMPLETE.txt")

step5_object_path <- file.path(STEP5_DIR, "Ppar_iKO_multiome_WNN_clustered.rds")

assert_file(step5_complete_marker, "Step 5 completion marker")
assert_file(step5_object_path, "Step 5 finalized WNN object")

step5_marker_lines <- readLines(step5_complete_marker)

extract_step5_value <- function(prefix) {
  matched_lines <- step5_marker_lines[startsWith(step5_marker_lines, prefix)]
  
  if (length(matched_lines) != 1L) {
    stop(
      "Expected exactly one Step 5 marker line beginning with '",
      prefix,
      "'; found ",
      length(matched_lines),
      ".",
      call. = FALSE
    )
  }
  
  trimws(sub(prefix, "", matched_lines, fixed = TRUE))
}

CHOSEN_RESOLUTION <- as.numeric(extract_step5_value("Chosen clustering resolution:"))

CHOSEN_CLUSTER_COLUMN <- extract_step5_value("Chosen cluster column:")

EXPECTED_CLUSTER_COUNT <- as.integer(extract_step5_value("Clusters:"))

if (!is.finite(CHOSEN_RESOLUTION) ||
    !nzchar(CHOSEN_CLUSTER_COLUMN) ||
    !is.finite(EXPECTED_CLUSTER_COUNT)) {
  stop("The Step 5 marker contains an invalid clustering decision.",
       call. = FALSE)
}

multiome <- readRDS(step5_object_path)

required_assays <- c("RNA", "ATAC")
missing_assays <- setdiff(required_assays, names(multiome))

if (length(missing_assays) > 0L) {
  stop("Step 5 object is missing assay(s): ",
       paste(missing_assays, collapse = ", "),
       call. = FALSE)
}

required_reductions <- c("pca", "lsi", "wnn.umap")
missing_reductions <- setdiff(required_reductions, names(multiome@reductions))

if (length(missing_reductions) > 0L) {
  stop(
    "Step 5 object is missing reduction(s): ",
    paste(missing_reductions, collapse = ", "),
    call. = FALSE
  )
}

required_graphs <- c("wknn", "wsnn")
missing_graphs <- setdiff(required_graphs, names(multiome@graphs))

if (length(missing_graphs) > 0L) {
  stop("Step 5 object is missing graph(s): ",
       paste(missing_graphs, collapse = ", "),
       call. = FALSE)
}

required_metadata_columns <- c("sample_id", CHOSEN_CLUSTER_COLUMN)
missing_metadata_columns <- setdiff(required_metadata_columns, colnames(multiome[[]]))

if (length(missing_metadata_columns) > 0L) {
  stop(
    "Step 5 object is missing metadata column(s): ",
    paste(missing_metadata_columns, collapse = ", "),
    call. = FALSE
  )
}

sample_match <- match(as.character(multiome$sample_id), sample_sheet$sample_id)

if (anyNA(sample_match)) {
  stop("One or more Step 5 nuclei could not be matched to samples.csv.",
       call. = FALSE)
}

multiome$diet <- factor(sample_sheet$diet[sample_match], levels = names(DIET_COLORS))

cluster_values <- as.character(multiome[[]][[CHOSEN_CLUSTER_COLUMN]])

if (anyNA(cluster_values) || any(!nzchar(cluster_values))) {
  stop("Chosen Step 5 cluster column contains missing or blank values.",
       call. = FALSE)
}

cluster_levels <- unique(cluster_values)

if (all(grepl("^[0-9]+$", cluster_levels))) {
  cluster_levels <- cluster_levels[order(as.integer(cluster_levels))]
} else {
  cluster_levels <- sort(cluster_levels)
}

if (length(cluster_levels) != EXPECTED_CLUSTER_COUNT) {
  stop(
    "Step 5 marker expected ",
    EXPECTED_CLUSTER_COUNT,
    " clusters but the object contains ",
    length(cluster_levels),
    ".",
    call. = FALSE
  )
}

for (review_clusters in list(CRYPT_REVIEW_CLUSTERS,
                             EPITHELIAL_REVIEW_CLUSTERS,
                             AMBIGUOUS_REVIEW_CLUSTERS)) {
  missing_review_clusters <- setdiff(review_clusters, cluster_levels)
  
  if (length(missing_review_clusters) > 0L) {
    stop(
      "Review cluster(s) absent from the Step 5 object: ",
      paste(missing_review_clusters, collapse = ", "),
      call. = FALSE
    )
  }
}

multiome$wnn_cluster <- factor(cluster_values, levels = cluster_levels)

Idents(multiome) <- "wnn_cluster"
DefaultAssay(multiome) <- "RNA"

cluster_colors <- stats::setNames(grDevices::hcl.colors(n = length(cluster_levels), palette = "Dark 3"),
                                  cluster_levels)

step5_validation <- data.frame(
  cells = ncol(multiome),
  RNA_features = nrow(multiome[["RNA"]]),
  ATAC_features = nrow(multiome[["ATAC"]]),
  WNN_clusters = length(cluster_levels),
  chosen_resolution = CHOSEN_RESOLUTION,
  chosen_cluster_column = CHOSEN_CLUSTER_COLUMN,
  WNN_graph_present = "wknn" %in% names(multiome@graphs),
  WSNN_graph_present = "wsnn" %in% names(multiome@graphs)
)

step5_validation

readr::write_csv(step5_validation,
                 file.path(STEP6_DIR, "step5_annotation_input_validation.csv"))

step6_complete_marker <- file.path(STEP6_DIR, "STEP_06_COMPLETE.txt")

if (file.exists(step6_complete_marker)) {
  message(
    "Invalidating the previous Step 6 completion marker before rebuilding annotation evidence."
  )
  
  unlink(step6_complete_marker)
}
```

# 4. Inventory the WNN clusters

The library-composition summaries below are descriptive. A cluster enriched in
one pooled library may represent biology, library-specific sampling, or another
unreplicated difference; this table cannot distinguish those explanations.

```{r cluster-inventory}
cluster_sample_inventory <- multiome[[]] |>
  dplyr::mutate(cluster = as.character(wnn_cluster),
                sample_id = as.character(sample_id)) |>
  dplyr::count(cluster, sample_id, name = "nuclei") |>
  dplyr::group_by(cluster) |>
  dplyr::mutate(
    cluster_total = sum(nuclei),
    percent_within_cluster = 100 * nuclei / cluster_total
  ) |>
  dplyr::ungroup() |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    sample_total = sum(nuclei),
    percent_of_sample = 100 * nuclei / sample_total
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    cluster = factor(cluster, levels = cluster_levels),
    sample_id = factor(sample_id, levels = sample_sheet$sample_id)
  ) |>
  dplyr::arrange(cluster, sample_id)

cluster_inventory <- cluster_sample_inventory |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    nuclei = sum(nuclei),
    percent_of_all_nuclei = 100 * nuclei / ncol(multiome),
    smallest_library_fraction = min(percent_within_cluster),
    largest_library_fraction = max(percent_within_cluster),
    .groups = "drop"
  )

cluster_inventory

readr::write_csv(cluster_inventory,
                 file.path(STEP6_DIR, "cluster_inventory.csv"))

readr::write_csv(
  cluster_sample_inventory,
  file.path(STEP6_DIR, "cluster_sample_composition_descriptive.csv")
)

p_cluster_umap <- DimPlot(
  multiome,
  reduction = "wnn.umap",
  group.by = "wnn_cluster",
  cols = cluster_colors,
  label = TRUE,
  repel = TRUE,
  pt.size = 0.25,
  seed = 20260727,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = paste0(
      "WNN clusters entering annotation (resolution ",
      CHOSEN_RESOLUTION,
      ")"
    ),
    subtitle = "Numbers are computational clusters, not final biological labels.",
    color = "WNN cluster"
  ) +
  theme_multiome()

print(p_cluster_umap)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "01_WNN_clusters_for_annotation.pdf"),
  plot = p_cluster_umap,
  width = 10,
  height = 8
)

p_cluster_composition <- ggplot2::ggplot(
  cluster_sample_inventory,
  ggplot2::aes(x = cluster, y = percent_within_cluster, fill = sample_id)
) +
  ggplot2::geom_col(width = 0.85) +
  ggplot2::scale_fill_manual(values = sample_colors) +
  ggplot2::scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    expand = ggplot2::expansion(mult = c(0, 0.03))
  ) +
  ggplot2::labs(
    title = "Descriptive pooled-library composition of WNN clusters",
    subtitle = "Library and diet are confounded; no treatment test is performed.",
    x = "WNN cluster",
    y = "Nuclei within cluster",
    fill = NULL
  ) +
  theme_multiome()

print(p_cluster_composition)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "02_cluster_sample_composition_descriptive.pdf"
  ),
  plot = p_cluster_composition,
  width = 12,
  height = 6
)
```

# 5. Discover positive RNA markers

`FindAllMarkers()` identifies genes enriched in each cluster relative to all
other nuclei. These are annotation candidates, not automatic labels. The full
table is retained so that later decisions can be audited.

```{r discover-rna-markers, cache=FALSE}
DefaultAssay(multiome) <- "RNA"
Idents(multiome) <- "wnn_cluster"

future::plan(future::sequential)

set.seed(20260824)

rna_all_positive_markers <- FindAllMarkers(
  object = multiome,
  assay = "RNA",
  only.pos = TRUE,
  test.use = "wilcox",
  min.pct = MIN_MARKER_PCT,
  logfc.threshold = MIN_MARKER_LOG2FC,
  return.thresh = 0.05,
  verbose = TRUE
)

configure_parallel_plan()

if (nrow(rna_all_positive_markers) == 0L) {
  stop("FindAllMarkers returned no positive RNA markers.", call. = FALSE)
}

marker_fc_column <- intersect(c("avg_log2FC", "avg_logFC"),
                              colnames(rna_all_positive_markers))

if (length(marker_fc_column) != 1L) {
  stop("Could not identify exactly one marker fold-change column.",
       call. = FALSE)
}

rna_all_positive_markers$effect_size <-
  rna_all_positive_markers[[marker_fc_column]]

rna_all_positive_markers$pct_delta <- with(rna_all_positive_markers, pct.1 - pct.2)

top_cluster_markers <- rna_all_positive_markers |>
  dplyr::filter(p_val_adj < 0.05) |>
  dplyr::group_by(cluster) |>
  dplyr::arrange(dplyr::desc(pct_delta), dplyr::desc(effect_size), .by_group = TRUE) |>
  dplyr::slice_head(n = TOP_MARKERS_PER_CLUSTER) |>
  dplyr::ungroup()

marker_cluster_coverage <- data.frame(
  expected_cluster = cluster_levels,
  marker_cluster_present = cluster_levels %in%
    as.character(unique(rna_all_positive_markers$cluster))
)

if (any(!marker_cluster_coverage$marker_cluster_present)) {
  warning(
    "No positive markers passed the current thresholds for cluster(s): ",
    paste(marker_cluster_coverage$expected_cluster[!marker_cluster_coverage$marker_cluster_present], collapse = ", "),
    call. = FALSE
  )
}

readr::write_csv(
  rna_all_positive_markers,
  file.path(STEP6_MARKER_DIR, "RNA_all_positive_markers.csv")
)

readr::write_csv(
  top_cluster_markers,
  file.path(STEP6_MARKER_DIR, "RNA_top_annotation_candidates.csv")
)

readr::write_csv(marker_cluster_coverage,
                 file.path(STEP6_DIR, "RNA_marker_cluster_coverage.csv"))

top_cluster_markers |>
  dplyr::select(cluster, gene, effect_size, pct.1, pct.2, p_val_adj)
```

## Visualize the strongest cluster discriminants

This heatmap is intentionally a presentation view rather than a second marker
test. Ribosomal, mitochondrial, predicted, and common technical genes are
excluded so that coherent biological programs are easier to inspect. The full
unfiltered marker table remains the authoritative result.

```{r top-marker-heatmap, fig.width=20, fig.height=20}
presentation_exclusion_pattern <- paste0("^(mt-|Rpl|Rps|Gm[0-9]+|Mir[0-9]+|Snor)", "|Rik$")

heatmap_marker_table <- rna_all_positive_markers |>
  dplyr::filter(
    p_val_adj < 0.05,
    effect_size >= 0.50,
    pct_delta >= 0.10,!grepl(presentation_exclusion_pattern, gene),!gene %in% c("Malat1", "Xist", "Tsix")
  ) |>
  dplyr::group_by(cluster) |>
  dplyr::arrange(dplyr::desc(pct_delta), dplyr::desc(effect_size), .by_group = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::ungroup()

heatmap_marker_coverage <- data.frame(cluster = cluster_levels, presentation_markers = as.integer(table(
  factor(as.character(heatmap_marker_table$cluster), levels = cluster_levels)
)))

if (any(heatmap_marker_coverage$presentation_markers == 0L)) {
  warning(
    "No presentation-quality heatmap markers passed for cluster(s): ",
    paste(heatmap_marker_coverage$cluster[heatmap_marker_coverage$presentation_markers == 0L], collapse = ", "),
    call. = FALSE
  )
}

heatmap_genes <- unique(heatmap_marker_table$gene)

if (length(heatmap_genes) == 0L) {
  stop("No genes were available for the cluster marker heatmap.", call. = FALSE)
}

set.seed(20260824)

heatmap_cells <- unlist(lapply(cluster_levels, function(cluster_id) {
  cluster_cells <- WhichCells(multiome, idents = cluster_id)
  
  sample(cluster_cells,
         size = min(60L, length(cluster_cells)),
         replace = FALSE)
}), use.names = FALSE)

readr::write_csv(
  heatmap_marker_table,
  file.path(STEP6_MARKER_DIR, "RNA_cluster_heatmap_markers.csv")
)

readr::write_csv(
  heatmap_marker_coverage,
  file.path(STEP6_DIR, "RNA_cluster_heatmap_marker_coverage.csv")
)

heatmap_object <- ScaleData(multiome,
                            assay = "RNA",
                            features = heatmap_genes,
                            verbose = FALSE)

p_cluster_heatmap <- suppressMessages(
  suppressWarnings(
    Seurat::DoHeatmap(
      heatmap_object,
      features = heatmap_genes,
      cells = heatmap_cells,
      assay = "RNA",
      slot = "scale.data",
      group.by = "wnn_cluster",
      group.colors = unname(cluster_colors[cluster_levels]),
      disp.min = -2,
      disp.max = 2,
      raster = TRUE,
      draw.lines = TRUE,
      angle = 0,
      hjust = 0.5,
      size = 4
    ) +
      ggplot2::scale_fill_gradientn(
        colors = c("#2166AC", "#D9D9D9", "#B2182B"),
        limits = c(-2, 2),
        oob = scales::squish
      ) +
      ggplot2::labs(
        title = "Strongest RNA discriminants across WNN clusters",
        subtitle = paste(
          "Up to five genes and 60 nuclei per cluster;",
          "the full unfiltered marker table is retained separately."
        )
      ) +
      ggplot2::theme(axis.text.y = ggplot2::element_text(face = "italic", size = 7))
  )
)

print(p_cluster_heatmap)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "03_RNA_cluster_marker_heatmap.pdf"),
  plot = p_cluster_heatmap,
  width = 20,
  height = 20,
  limitsize = FALSE
)

rm(heatmap_object)
invisible(gc())
```

# 6. Define and audit marker programs

The marker panels are intentionally redundant at the biological-program level,
but every plotted gene appears only once. The audit table records markers that
are absent from the RNA feature space instead of silently dropping them.

```{r define-marker-programs}
broad_lineage_programs <- list(
  Epithelial = c("Epcam", "Krt8", "Krt18", "Krt19", "Cdh1"),
  T_NK = c("Ptprc", "Cd3d", "Cd3e", "Trbc1", "Nkg7", "Klrd1"),
  B_cell = c("Cd79a", "Cd79b", "Ms4a1", "Cd74", "Cd37"),
  Plasma_cell = c("Jchain", "Mzb1", "Sdc1", "Igkc"),
  Myeloid = c("Lyz2", "Adgre1", "Csf1r", "C1qa", "C1qb", "C1qc"),
  Dendritic = c("Itgax", "Flt3", "H2-Ab1", "Ccr7"),
  Plasmacytoid_DC = c("Siglech", "Flt3", "Ccr9", "Pld4", "Gzmb"),
  Neutrophil = c("S100a8", "S100a9", "Ly6g", "Csf3r"),
  Mast = c("Kit", "Ms4a2", "Cpa3", "Mcpt1", "Mcpt2"),
  Vascular_endothelial = c("Cdh5", "Pecam1", "Kdr", "Esam", "Emcn"),
  Lymphatic_endothelial = c("Flt4", "Prox1", "Lyve1", "Ccl21a", "Pdpn"),
  Fibroblast = c("Col1a1", "Col1a2", "Col3a1", "Dcn", "Pdgfra"),
  Smooth_muscle = c("Carmn", "Acta2", "Tagln", "Myh11", "Mylk"),
  Mesothelial = c("Lrrn4", "Upk3b", "Msln", "Wt1", "Rspo1"),
  Enteric_glia = c("S100b", "Sox10", "Plp1", "Sox2", "Fabp7"),
  Enteric_neuron = c("Snap25", "Tubb3", "Elavl4", "Syp", "Uchl1")
)

epithelial_state_programs <- list(
  Stem_Wnt = c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Ephb2"),
  Cell_cycle = c("Mki67", "Pcna", "Top2a", "Cdk1", "Ccnb1", "Ube2c", "Birc5"),
  Secretory_commitment = c("Atoh1", "Dll1", "Gfi1", "Spdef"),
  Absorptive_progenitor = c("Selenbp1", "Cyp2c55", "Car1"),
  Mature_colonocyte = c("Krt20", "Slc26a3", "Aqp8", "Car1", "Car2", "Otop2"),
  Goblet = c("Muc2", "Tff3", "Clca1", "Fcgbp", "Agr2", "Spdef"),
  Deep_crypt_secretory = c("Reg4", "Mptx1", "Spink4", "Kit"),
  EEC_general = c("Chga", "Chgb", "Neurod1", "Pax6", "Isl1"),
  Enterochromaffin = c("Tph1", "Lmx1a", "Fev", "Tac1"),
  L_cell = c("Gcg", "Pyy", "Insl5"),
  Tuft = c("Pou2f3", "Dclk1", "Trpm5", "Avil", "Il17rb")
)

sex_marker_programs <- list(
  X_linked = c("Xist", "Tsix"),
  Y_linked = c("Ddx3y", "Eif2s3y", "Kdm5d", "Uty")
)

audit_marker_programs <- function(program_list, panel_name, RNA_features) {
  marker_table <- do.call(rbind, lapply(names(program_list), function(program_name) {
    data.frame(
      panel = panel_name,
      program = program_name,
      gene = program_list[[program_name]],
      stringsAsFactors = FALSE
    )
  }))
  
  marker_table$present_in_RNA <- marker_table$gene %in% RNA_features
  marker_table
}

marker_program_audit <- dplyr::bind_rows(
  audit_marker_programs(broad_lineage_programs, "broad_lineage", rownames(multiome[["RNA"]])),
  audit_marker_programs(
    epithelial_state_programs,
    "epithelial_state",
    rownames(multiome[["RNA"]])
  ),
  audit_marker_programs(sex_marker_programs, "sex_audit", rownames(multiome[["RNA"]]))
)

marker_program_audit

readr::write_csv(
  marker_program_audit,
  file.path(STEP6_DIR, "annotation_marker_program_audit.csv")
)

make_scaled_dotplot <- function(object,
                                features,
                                group_by,
                                title,
                                subtitle,
                                base_size = 9) {
  detected_features <- features[features %in% rownames(object[["RNA"]])]
  
  if (length(detected_features) == 0L) {
    stop("No requested DotPlot features are present in RNA.", call. = FALSE)
  }
  
  suppressMessages(
    suppressWarnings(
      Seurat::DotPlot(
        object,
        features = detected_features,
        group.by = group_by,
        assay = "RNA",
        scale = TRUE,
        col.min = -2,
        col.max = 2
      ) +
        ggplot2::scale_color_gradient2(
          low = "#2166AC",
          mid = "#D9D9D9",
          high = "#B2182B",
          midpoint = 0,
          limits = c(-2, 2),
          oob = scales::squish,
          name = "Scaled average\nexpression"
        ) +
        Seurat::RotatedAxis() +
        ggplot2::labs(
          title = title,
          subtitle = subtitle,
          x = NULL,
          y = "WNN cluster"
        ) +
        theme_multiome(base_size = base_size)
    )
  )
}
```

# 7. Broad lineage adjudication

This panel tests whether the computational clusters contain coherent epithelial
or non-epithelial lineage programs. `Kit` is intentionally interpreted in
context: `Kit` with `Reg4`/`Spink4` supports a deep crypt secretory program,
whereas `Kit` with `Ms4a2`/`Cpa3` supports mast-cell identity.

```{r broad-lineage-dotplot, fig.width=22, fig.height=9}
broad_lineage_features <- unique(
  unlist(broad_lineage_programs, use.names = FALSE)
)

p_broad_lineage <- make_scaled_dotplot(
  object = multiome,
  features = broad_lineage_features,
  group_by = "wnn_cluster",
  title = "Broad lineage programs across WNN clusters",
  subtitle = paste(
    "Color is scaled average expression within each gene;",
    "dot size is the fraction of nuclei expressing the gene."
  ),
  base_size = 9
)

print(p_broad_lineage)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "04_broad_lineage_DotPlot.pdf"),
  plot = p_broad_lineage,
  width = 30,
  height = 9
)
```

# 8. Focused epithelial-state adjudication

This panel keeps all clusters visible. Restricting the plot to clusters already
believed to be epithelial would make it too easy to miss a rare epithelial
population or a mixed cluster.

```{r epithelial-state-dotplot, fig.width=23, fig.height=9}
epithelial_state_features <- unique(
  unlist(epithelial_state_programs, use.names = FALSE)
)

p_epithelial_states <- make_scaled_dotplot(
  object = multiome,
  features = epithelial_state_features,
  group_by = "wnn_cluster",
  title = "Mouse-colon epithelial programs across WNN clusters",
  subtitle = paste(
    "Candidate states require multi-gene support;",
    "negative scaled values mean below-gene-average expression, not negative counts."
  ),
  base_size = 9
)

print(p_epithelial_states)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "05_epithelial_state_DotPlot.pdf"),
  plot = p_epithelial_states,
  width = 23,
  height = 9
)
```

# 9. Sex-marker audit

Each library pools female and male mice. Sex-linked signal is therefore an
expected within-library source of heterogeneity. This audit helps prevent a
sex-associated cluster from being mislabeled as a novel colon state.

```{r sex-marker-audit, fig.width=10, fig.height=5}
sex_marker_features <- unique(
  unlist(sex_marker_programs, use.names = FALSE)
)

detected_sex_markers <- sex_marker_features[
  sex_marker_features %in% rownames(multiome[["RNA"]])
]

if (length(detected_sex_markers) > 0L) {
  p_sex_audit <- suppressMessages(
    suppressWarnings(
      Seurat::DotPlot(
        multiome,
        features = detected_sex_markers,
        group.by = "wnn_cluster",
        assay = "RNA",
        scale = TRUE,
        col.min = -2,
        col.max = 2
      ) +
        ggplot2::scale_color_gradient2(
          low = "#2166AC",
          mid = "#D9D9D9",
          high = "#B2182B",
          midpoint = 0,
          limits = c(-2, 2),
          oob = scales::squish,
          name = "Scaled average\nexpression"
        ) +
        Seurat::RotatedAxis() +
        ggplot2::labs(
          title = "Sex-linked RNA signal across WNN clusters",
          subtitle = "Both libraries contain pooled female and male mice.",
          x = NULL,
          y = "WNN cluster"
        ) +
        theme_multiome()
    )
  )

  print(p_sex_audit)

  ggplot2::ggsave(
    filename = file.path(STEP6_PLOT_DIR, "06_sex_marker_audit.pdf"),
    plot = p_sex_audit,
    width = 10,
    height = 7
  )
}
```

# 10. Focused crypt-cluster adjudication

Step 5 suggested three related but distinct hypotheses:

- cluster 6: noncycling ISC or early progenitor;
- cluster 3: cycling ISC or early transit-amplifying state;
- cluster 19: cycling secretory progenitor with `Muc2`-associated signal.

Those names remain hypotheses. The following analyses test whether the three
clusters differ in stem/Wnt, proliferation, secretory-commitment, and mature
secretory programs.

## Focused RNA marker programs

```{r focused-crypt-dotplot, fig.width=15, fig.height=6}
crypt_programs <- list(
  Stem_Wnt = c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Ephb2"),
  Proliferative = c("Mki67", "Pcna", "Top2a", "Cdk1", "Ccnb1", "Ube2c", "Birc5"),
  Secretory_commitment = c("Atoh1", "Dll1", "Gfi1", "Spdef"),
  Goblet_DCS = c("Muc2", "Agr2", "Tff3", "Fcgbp", "Spink4", "Reg4", "Kit")
)

crypt_review_cells <- WhichCells(
  multiome,
  idents = CRYPT_REVIEW_CLUSTERS
)

crypt_review_object <- subset(
  multiome,
  cells = crypt_review_cells
)

crypt_review_object$wnn_cluster <- factor(
  as.character(crypt_review_object$wnn_cluster),
  levels = CRYPT_REVIEW_CLUSTERS
)

Idents(crypt_review_object) <- "wnn_cluster"
DefaultAssay(crypt_review_object) <- "RNA"

crypt_features <- unique(unlist(crypt_programs, use.names = FALSE))

p_crypt_programs <- make_scaled_dotplot(
  object = crypt_review_object,
  features = crypt_features,
  group_by = "wnn_cluster",
  title = "Focused crypt-state adjudication",
  subtitle = paste(
    "Clusters 3, 6, and 19 were selected from Step 5 as competing",
    "stem, cycling, and secretory-progenitor hypotheses."
  ),
  base_size = 10
)

print(p_crypt_programs)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "07_focused_crypt_DotPlot.pdf"),
  plot = p_crypt_programs,
  width = 15,
  height = 6
)
```

## Per-nucleus marker co-detection

Unlike scaled cluster averages, this table asks whether marker programs are
co-detected in the same nucleus. These thresholds are transparent descriptive
rules; they are not trained classifiers.

```{r crypt-codetection}
crypt_counts <- SeuratObject::GetAssayData(
  crypt_review_object,
  assay = "RNA",
  layer = "counts"
)

detected_crypt_programs <- lapply(
  crypt_programs,
  function(program_genes) {
    program_genes[program_genes %in% rownames(crypt_counts)]
  }
)

if (any(lengths(detected_crypt_programs) == 0L)) {
  stop(
    "At least one focused crypt program has no detected RNA features.",
    call. = FALSE
  )
}

crypt_program_detection <- do.call(
  cbind,
  lapply(
    detected_crypt_programs,
    function(program_genes) {
      Matrix::colSums(
        crypt_counts[program_genes, , drop = FALSE] > 0
      )
    }
  )
)

colnames(crypt_program_detection) <- names(detected_crypt_programs)

crypt_review_metadata <- crypt_review_object[[]][
  colnames(crypt_counts),
  ,
  drop = FALSE
]

if (!identical(rownames(crypt_review_metadata), colnames(crypt_counts))) {
  stop(
    "Focused crypt metadata could not be aligned to the RNA count matrix.",
    call. = FALSE
  )
}

per_nucleus_crypt_detection <- data.frame(
  nucleus = colnames(crypt_counts),
  cluster = as.character(crypt_review_metadata$wnn_cluster),
  sample_id = as.character(crypt_review_metadata$sample_id),
  crypt_program_detection,
  stringsAsFactors = FALSE
)

per_nucleus_crypt_detection$Stem_Wnt_positive <-
  per_nucleus_crypt_detection$Stem_Wnt >= 2L

per_nucleus_crypt_detection$Proliferative_positive <-
  per_nucleus_crypt_detection$Proliferative >= 2L

per_nucleus_crypt_detection$Secretory_commitment_positive <-
  per_nucleus_crypt_detection$Secretory_commitment >= 1L

per_nucleus_crypt_detection$Goblet_DCS_positive <-
  per_nucleus_crypt_detection$Goblet_DCS >= 2L

crypt_codetection_summary <- per_nucleus_crypt_detection |>
  dplyr::group_by(cluster, sample_id) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    stem_Wnt_positive_pct = 100 * mean(Stem_Wnt_positive),
    proliferative_positive_pct = 100 * mean(Proliferative_positive),
    secretory_commitment_positive_pct =
      100 * mean(Secretory_commitment_positive),
    goblet_DCS_positive_pct = 100 * mean(Goblet_DCS_positive),
    stem_and_proliferative_pct = 100 * mean(
      Stem_Wnt_positive & Proliferative_positive
    ),
    proliferative_and_secretory_pct = 100 * mean(
      Proliferative_positive & Secretory_commitment_positive
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    cluster = factor(cluster, levels = CRYPT_REVIEW_CLUSTERS),
    sample_id = factor(sample_id, levels = sample_sheet$sample_id)
  ) |>
  dplyr::arrange(cluster, sample_id)

crypt_codetection_summary

readr::write_csv(
  per_nucleus_crypt_detection,
  file.path(STEP6_DIR, "focused_crypt_per_nucleus_codetection.csv")
)

readr::write_csv(
  crypt_codetection_summary,
  file.path(STEP6_DIR, "focused_crypt_codetection_summary.csv")
)

crypt_codetection_plot_data <- crypt_codetection_summary |>
  tidyr::pivot_longer(
    cols = dplyr::ends_with("_pct"),
    names_to = "program",
    values_to = "percent_positive"
  ) |>
  dplyr::mutate(
    program = factor(
      program,
      levels = c(
        "stem_Wnt_positive_pct",
        "proliferative_positive_pct",
        "secretory_commitment_positive_pct",
        "goblet_DCS_positive_pct",
        "stem_and_proliferative_pct",
        "proliferative_and_secretory_pct"
      ),
      labels = c(
        "Stem/Wnt",
        "Proliferative",
        "Secretory commitment",
        "Goblet/DCS",
        "Stem + proliferative",
        "Proliferative + secretory"
      )
    )
  )

p_crypt_codetection <- ggplot2::ggplot(
  crypt_codetection_plot_data,
  ggplot2::aes(
    x = cluster,
    y = percent_positive,
    fill = sample_id
  )
) +
  ggplot2::geom_col(
    position = ggplot2::position_dodge(width = 0.8),
    width = 0.7
  ) +
  ggplot2::facet_wrap(ggplot2::vars(program), ncol = 3) +
  ggplot2::scale_fill_manual(values = sample_colors) +
  ggplot2::scale_y_continuous(
    limits = c(0, 100),
    labels = scales::label_percent(scale = 1),
    expand = ggplot2::expansion(mult = c(0, 0.03))
  ) +
  ggplot2::labs(
    title = "Per-nucleus co-detection in candidate crypt clusters",
    subtitle = "Bars are shown separately for each pooled library as an annotation check.",
    x = "WNN cluster",
    y = "Nuclei meeting rule",
    fill = NULL
  ) +
  theme_multiome()

print(p_crypt_codetection)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "08_focused_crypt_codetection.pdf"),
  plot = p_crypt_codetection,
  width = 13,
  height = 8
)
```

## Pairwise RNA contrasts

```{r pairwise-crypt-markers, cache=FALSE}
future::plan(future::sequential)

crypt_cluster_pairs <- combn(
  CRYPT_REVIEW_CLUSTERS,
  2,
  simplify = FALSE
)

pairwise_crypt_markers <- lapply(
  crypt_cluster_pairs,
  function(cluster_pair) {
    pair_markers <- FindMarkers(
      crypt_review_object,
      ident.1 = cluster_pair[[1]],
      ident.2 = cluster_pair[[2]],
      assay = "RNA",
      test.use = "wilcox",
      min.pct = 0.05,
      logfc.threshold = 0,
      verbose = FALSE
    )

    pair_markers$gene <- rownames(pair_markers)
    pair_markers$ident_1 <- cluster_pair[[1]]
    pair_markers$ident_2 <- cluster_pair[[2]]
    pair_markers$comparison <- paste0(
      "cluster_",
      cluster_pair[[1]],
      "_vs_",
      cluster_pair[[2]]
    )

    rownames(pair_markers) <- NULL
    pair_markers
  }
)

pairwise_crypt_markers <- dplyr::bind_rows(pairwise_crypt_markers)

configure_parallel_plan()

pairwise_fc_column <- intersect(
  c("avg_log2FC", "avg_logFC"),
  colnames(pairwise_crypt_markers)
)

if (length(pairwise_fc_column) != 1L) {
  stop(
    "Could not identify exactly one pairwise fold-change column.",
    call. = FALSE
  )
}

readr::write_csv(
  pairwise_crypt_markers,
  file.path(STEP6_MARKER_DIR, "RNA_focused_crypt_pairwise_markers.csv")
)

pairwise_crypt_markers |>
  dplyr::filter(p_val_adj < 0.05) |>
  dplyr::group_by(comparison) |>
  dplyr::arrange(
    dplyr::desc(abs(.data[[pairwise_fc_column]])),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = 15) |>
  dplyr::ungroup()
```

## Cell-cycle context without regression

```{r crypt-cell-cycle}
mouse_s_features <- unique(
  stats::na.omit(
    unname(
      Seurat::CaseMatch(
        search = cc.genes.updated.2019$s.genes,
        match = rownames(multiome[["RNA"]])
      )
    )
  )
)

mouse_g2m_features <- unique(
  stats::na.omit(
    unname(
      Seurat::CaseMatch(
        search = cc.genes.updated.2019$g2m.genes,
        match = rownames(multiome[["RNA"]])
      )
    )
  )
)

if (length(mouse_s_features) < 10L || length(mouse_g2m_features) < 10L) {
  stop(
    "Too few cell-cycle genes matched mouse RNA symbols: ",
    length(mouse_s_features),
    " S and ",
    length(mouse_g2m_features),
    " G2M genes.",
    call. = FALSE
  )
}

multiome <- Seurat::CellCycleScoring(
  object = multiome,
  s.features = mouse_s_features,
  g2m.features = mouse_g2m_features,
  set.ident = FALSE,
  assay = "RNA"
)

cell_cycle_feature_audit <- dplyr::bind_rows(
  data.frame(phase_program = "S", gene = mouse_s_features),
  data.frame(phase_program = "G2M", gene = mouse_g2m_features)
)

readr::write_csv(
  cell_cycle_feature_audit,
  file.path(STEP6_DIR, "cell_cycle_features_matched_to_mouse_RNA.csv")
)

crypt_cycle_summary <- multiome[[]] |>
  dplyr::mutate(cluster = as.character(wnn_cluster)) |>
  dplyr::filter(cluster %in% CRYPT_REVIEW_CLUSTERS) |>
  dplyr::count(cluster, sample_id, Phase, name = "nuclei") |>
  dplyr::group_by(cluster, sample_id) |>
  dplyr::mutate(
    total_nuclei = sum(nuclei),
    percent_within_cluster_sample = 100 * nuclei / total_nuclei
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    cluster = factor(cluster, levels = CRYPT_REVIEW_CLUSTERS),
    Phase = factor(Phase, levels = c("G1", "S", "G2M"))
  )

crypt_cycle_summary

readr::write_csv(
  crypt_cycle_summary,
  file.path(STEP6_DIR, "focused_crypt_cell_cycle_summary.csv")
)

phase_colors <- c(
  G1 = "#B8B8B8",
  S = "yellow2",
  G2M = "green4"
)

p_crypt_cycle <- ggplot2::ggplot(
  crypt_cycle_summary,
  ggplot2::aes(
    x = cluster,
    y = percent_within_cluster_sample,
    fill = Phase
  )
) +
  ggplot2::geom_col(width = 0.8) +
  ggplot2::facet_wrap(ggplot2::vars(sample_id)) +
  ggplot2::scale_fill_manual(values = phase_colors) +
  ggplot2::scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    expand = ggplot2::expansion(mult = c(0, 0.03))
  ) +
  ggplot2::labs(
    title = "Cell-cycle state within candidate crypt clusters",
    subtitle = "Cell-cycle scores are retained as biological context and are not regressed.",
    x = "WNN cluster",
    y = "Nuclei within cluster and library",
    fill = NULL
  ) +
  theme_multiome()

print(p_crypt_cycle)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "09_focused_crypt_cell_cycle.pdf"),
  plot = p_crypt_cycle,
  width = 11,
  height = 6
)
```

## Cross-library RNA consistency

The following summary uses log-normalized RNA values. It asks whether the
direction of a marker program is visible in both pooled libraries; it does not
test a diet effect.

```{r crypt-cross-library-expression}
crypt_review_genes <- unique(
  c(
    "Lgr5", "Ascl2", "Smoc2", "Mki67", "Top2a",
    "Atoh1", "Muc2", "Reg4", "Kit"
  )
)

crypt_review_genes <- crypt_review_genes[
  crypt_review_genes %in% rownames(multiome[["RNA"]])
]

rna_data <- SeuratObject::GetAssayData(
  multiome,
  assay = "RNA",
  layer = "data"
)

crypt_metadata <- multiome[[]] |>
  dplyr::mutate(
    nucleus = rownames(multiome[[]]),
    cluster = as.character(wnn_cluster),
    sample_id = as.character(sample_id)
  ) |>
  dplyr::filter(cluster %in% CRYPT_REVIEW_CLUSTERS)

crypt_group_cells <- split(
  crypt_metadata$nucleus,
  interaction(
    crypt_metadata$cluster,
    crypt_metadata$sample_id,
    drop = TRUE,
    sep = "__"
  )
)

crypt_cross_library_expression <- dplyr::bind_rows(
  lapply(
    names(crypt_group_cells),
    function(group_name) {
      group_cells <- crypt_group_cells[[group_name]]
      group_parts <- strsplit(group_name, "__", fixed = TRUE)[[1]]

      data.frame(
        cluster = group_parts[[1]],
        sample_id = group_parts[[2]],
        gene = crypt_review_genes,
        average_log_normalized_expression = as.numeric(
          Matrix::rowMeans(
            rna_data[
              crypt_review_genes,
              group_cells,
              drop = FALSE
            ]
          )
        ),
        percent_detected = 100 * as.numeric(
          Matrix::rowMeans(
            rna_data[
              crypt_review_genes,
              group_cells,
              drop = FALSE
            ] > 0
          )
        ),
        nuclei = length(group_cells),
        stringsAsFactors = FALSE
      )
    }
  )
) |>
  dplyr::mutate(
    cluster = factor(cluster, levels = CRYPT_REVIEW_CLUSTERS),
    sample_id = factor(sample_id, levels = sample_sheet$sample_id),
    cluster_sample = interaction(
      cluster,
      sample_id,
      sep = " | ",
      drop = TRUE
    ),
    gene = factor(gene, levels = rev(crypt_review_genes))
  )

readr::write_csv(
  crypt_cross_library_expression,
  file.path(STEP6_DIR, "focused_crypt_cross_library_expression.csv")
)

p_crypt_cross_library <- ggplot2::ggplot(
  crypt_cross_library_expression,
  ggplot2::aes(
    x = cluster_sample,
    y = gene,
    size = percent_detected,
    color = average_log_normalized_expression
  )
) +
  ggplot2::geom_point() +
  ggplot2::scale_size_continuous(
    range = c(0.5, 8),
    limits = c(0, 100),
    name = "Nuclei detected (%)"
  ) +
  ggplot2::labs(
    title = "Cross-library consistency of focused crypt markers",
    subtitle = "Library-separated values are descriptive annotation evidence.",
    x = NULL,
    y = NULL
  ) +
  theme_multiome()

print(p_crypt_cross_library)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "10_focused_crypt_cross_library_expression.pdf"
  ),
  plot = p_crypt_cross_library,
  width = 12,
  height = 7
)

rm(rna_data)
invisible(gc())
```

## Focused RNA expression on the WNN embedding

```{r focused-crypt-feature-plots, fig.width=15, fig.height=10}
feature_plot_genes <- c(
  "Lgr5", "Smoc2", "Mki67", "Top2a", "Atoh1", "Muc2", "Reg4", "Kit"
)

feature_plot_genes <- feature_plot_genes[
  feature_plot_genes %in% rownames(multiome[["RNA"]])
]

p_crypt_features <- Seurat::FeaturePlot(
  multiome,
  features = feature_plot_genes,
  reduction = "wnn.umap",
  order = TRUE,
  min.cutoff = "q05",
  max.cutoff = "q95",
  pt.size = 0.2,
  ncol = 4
) &
  theme_multiome(base_size = 9)

print(p_crypt_features)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "11_crypt_marker_FeaturePlots.pdf"),
  plot = p_crypt_features,
  width = 15,
  height = 10
)
```

# 11. Targeted ATAC support for the crypt hypotheses

Coverage plots are not used as an automatic classifier. They provide a
multimodal check that RNA-supported differences are accompanied by plausible
accessibility near lineage-defining loci. Tracks are split by WNN cluster and
pooled library.

```{r targeted-crypt-atac-coverage, cache=FALSE, eval=RUN_TARGETED_ATAC_COVERAGE}
targeted_crypt_loci <- c("Lgr5", "Ascl2", "Atoh1", "Muc2", "Reg4", "Kit")

fragment_objects <- Signac::Fragments(multiome[["ATAC"]])

if (length(fragment_objects) == 0L) {
  stop("The ATAC assay has no Fragment objects for CoveragePlot().", call. = FALSE)
}

atac_annotation <- Signac::Annotation(multiome[["ATAC"]])

if (length(atac_annotation) == 0L) {
  stop("The ATAC assay has no gene annotation.", call. = FALSE)
}

annotation_columns <- colnames(S4Vectors::mcols(atac_annotation))

if (!"gene_name" %in% annotation_columns) {
  stop(
    "The ATAC annotation lacks a gene_name metadata column.",
    call. = FALSE
  )
}

annotated_gene_names <- unique(
  as.character(S4Vectors::mcols(atac_annotation)$gene_name)
)

coverage_locus_audit <- data.frame(
  locus = targeted_crypt_loci,
  present_in_ATAC_annotation = targeted_crypt_loci %in% annotated_gene_names
)

readr::write_csv(
  coverage_locus_audit,
  file.path(STEP6_DIR, "targeted_crypt_ATAC_locus_audit.csv")
)

available_crypt_loci <- coverage_locus_audit$locus[
  coverage_locus_audit$present_in_ATAC_annotation
]

if (length(available_crypt_loci) == 0L) {
  stop(
    "None of the targeted crypt loci are present in the ATAC annotation.",
    call. = FALSE
  )
}

coverage_object <- subset(multiome, cells = crypt_review_cells)

coverage_group_sizes <- coverage_object[[]] |>
  dplyr::mutate(
    cluster = as.character(wnn_cluster),
    sample_id = as.character(sample_id)
  ) |>
  dplyr::count(cluster, sample_id, name = "nuclei") |>
  dplyr::mutate(
    cluster = factor(cluster, levels = CRYPT_REVIEW_CLUSTERS),
    sample_id = factor(sample_id, levels = sample_sheet$sample_id),
    coverage_group = paste0(
      "Cluster ",
      cluster,
      " | ",
      sample_id,
      " (n=",
      nuclei,
      ")"
    )
  ) |>
  dplyr::arrange(cluster, sample_id)

coverage_group_lookup <- stats::setNames(
  coverage_group_sizes$coverage_group,
  paste(
    coverage_group_sizes$cluster,
    coverage_group_sizes$sample_id,
    sep = "__"
  )
)

coverage_group_key <- paste(
  as.character(coverage_object$wnn_cluster),
  as.character(coverage_object$sample_id),
  sep = "__"
)

coverage_object$coverage_group <- factor(
  unname(coverage_group_lookup[coverage_group_key]),
  levels = coverage_group_sizes$coverage_group
)

if (anyNA(coverage_object$coverage_group)) {
  stop(
    "One or more focused crypt nuclei lack an ATAC coverage group.",
    call. = FALSE
  )
}

readr::write_csv(
  coverage_group_sizes,
  file.path(STEP6_DIR, "targeted_crypt_ATAC_coverage_group_sizes.csv")
)

DefaultAssay(coverage_object) <- "ATAC"

coverage_pdf_path <- file.path(
  STEP6_PLOT_DIR,
  "12_targeted_crypt_ATAC_coverage.pdf"
)

grDevices::pdf(
  file = coverage_pdf_path,
  width = 13,
  height = 9,
  onefile = TRUE
)

for (locus in available_crypt_loci) {
  p_coverage <- Signac::CoveragePlot(
    object = coverage_object,
    region = locus,
    features = locus,
    assay = "ATAC",
    expression.assay = "RNA",
    expression.slot = "data",
    annotation = "gene",
    peaks = TRUE,
    links = FALSE,
    tile = FALSE,
    group.by = "coverage_group",
    extend.upstream = 5000,
    extend.downstream = 5000,
    ymax = "q95"
  ) +
    patchwork::plot_annotation(
      title = paste0(locus, " locus: ATAC coverage with RNA expression"),
      subtitle = paste(
        "Candidate crypt clusters split by pooled library;",
        "tracks are normalized by group depth and cell number."
      )
    )

  print(p_coverage)
}

grDevices::dev.off()

rm(coverage_object, fragment_objects, atac_annotation)
invisible(gc())
```

# 12. Write the manual annotation proposal

The `review_hint` column preserves the hypotheses developed during Step 5. It
is deliberately separate from the blank decision fields. Copy the proposed CSV
to the printed config path, then complete one row per WNN cluster.

Required decision fields are:

- `broad_compartment`: epithelial, immune, stromal, vascular, neural, or an
  explicit unresolved category;
- `cell_state`: the final biological label to store per nucleus;
- `annotation_confidence`: `high`, `moderate`, or `low`;
- `evidence_summary`: concise multi-gene and/or ATAC justification; and
- `approved`: `TRUE` only after manual review.

```{r write-annotation-proposal}
top_marker_strings <- top_cluster_markers |>
  dplyr::mutate(cluster = as.character(cluster)) |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    top_RNA_markers = paste(gene, collapse = ";"),
    .groups = "drop"
  )

review_hint_lookup <- c(
  `3` = "cycling stem/early progenitor candidate",
  `5` = "goblet candidate",
  `6` = "noncycling stem/early progenitor candidate",
  `7` = "absorptive/colonocyte candidate",
  `8` = "absorptive/colonocyte candidate",
  `9` = "absorptive/colonocyte candidate",
  `13` = "Reg4-positive deep crypt secretory candidate",
  `14` = "enterochromaffin-like EEC candidate",
  `17` = "Pyy/Gcg-positive L-cell candidate",
  `19` = "cycling secretory progenitor candidate",
  `10` = "broad non-epithelial lineage review",
  `11` = "broad non-epithelial lineage review",
  `12` = "broad non-epithelial lineage review",
  `22` = "broad non-epithelial lineage review",
  `23` = "Kit context review: mast versus deep crypt secretory",
  `24` = "broad non-epithelial lineage review",
  `25` = "broad non-epithelial lineage review",
  `26` = "tuft candidate"
)

annotation_proposal <- cluster_inventory |>
  dplyr::mutate(cluster = as.character(cluster)) |>
  dplyr::left_join(top_marker_strings, by = "cluster") |>
  dplyr::mutate(
    review_scope = dplyr::case_when(
      cluster %in% CRYPT_REVIEW_CLUSTERS ~ "focused_crypt_review",
      cluster %in% EPITHELIAL_REVIEW_CLUSTERS ~ "epithelial_review",
      cluster %in% AMBIGUOUS_REVIEW_CLUSTERS ~ "broad_lineage_review",
      TRUE ~ "standard_review"
    ),
    review_hint = unname(review_hint_lookup[cluster]),
    review_hint = dplyr::if_else(
      is.na(review_hint),
      "no preassigned hypothesis",
      review_hint
    ),
    broad_compartment = NA_character_,
    cell_state = NA_character_,
    annotation_confidence = NA_character_,
    evidence_summary = NA_character_,
    approved = FALSE
  ) |>
  dplyr::arrange(factor(cluster, levels = cluster_levels))

readr::write_csv(annotation_proposal, ANNOTATION_PROPOSAL_PATH)

annotation_proposal

cat(
  "\nManual annotation proposal written to:\n",
  ANNOTATION_PROPOSAL_PATH,
  "\n\nReviewed decision table expected at:\n",
  ANNOTATION_DECISION_PATH,
  "\n"
)
```

# 13. Manual annotation approval gate

The first run is expected to stop here after producing the evidence files and
proposal. Do not bypass this gate by inserting labels directly into the Seurat
object. The decision CSV is the auditable record of the biological judgment.

```{r validate-annotation-decisions}
if (!file.exists(ANNOTATION_DECISION_PATH)) {
  stop(
    "Manual annotation checkpoint reached.\n",
    "Review the Step 6 plots and tables, copy:\n",
    ANNOTATION_PROPOSAL_PATH,
    "\nto:\n",
    ANNOTATION_DECISION_PATH,
    "\nComplete every decision field and set approved=TRUE only after review.",
    call. = FALSE
  )
}

annotation_decisions <- suppressMessages(
  readr::read_csv(
    ANNOTATION_DECISION_PATH,
    show_col_types = FALSE
  )
)

required_decision_columns <- c(
  "cluster",
  "broad_compartment",
  "cell_state",
  "annotation_confidence",
  "evidence_summary",
  "approved"
)

missing_decision_columns <- setdiff(
  required_decision_columns,
  colnames(annotation_decisions)
)

if (length(missing_decision_columns) > 0L) {
  stop(
    "Annotation decision table is missing required column(s): ",
    paste(missing_decision_columns, collapse = ", "),
    call. = FALSE
  )
}

annotation_decisions$cluster <- as.character(annotation_decisions$cluster)

if (anyDuplicated(annotation_decisions$cluster) > 0L) {
  stop("Annotation decision table contains duplicate clusters.", call. = FALSE)
}

if (!setequal(annotation_decisions$cluster, cluster_levels)) {
  stop(
    "Annotation decision clusters do not exactly match the Step 5 clusters.",
    call. = FALSE
  )
}

annotation_decisions <- annotation_decisions[
  match(cluster_levels, annotation_decisions$cluster),
  ,
  drop = FALSE
]

decision_text_columns <- c(
  "broad_compartment",
  "cell_state",
  "annotation_confidence",
  "evidence_summary"
)

for (column_name in decision_text_columns) {
  annotation_decisions[[column_name]] <- trimws(
    as.character(annotation_decisions[[column_name]])
  )

  if (
    anyNA(annotation_decisions[[column_name]]) ||
      any(!nzchar(annotation_decisions[[column_name]]))
  ) {
    stop(
      "Annotation decision column '",
      column_name,
      "' contains missing or blank values.",
      call. = FALSE
    )
  }
}

valid_confidence <- c("high", "moderate", "low")

annotation_decisions$annotation_confidence <- tolower(
  annotation_decisions$annotation_confidence
)

invalid_confidence <- setdiff(
  unique(annotation_decisions$annotation_confidence),
  valid_confidence
)

if (length(invalid_confidence) > 0L) {
  stop(
    "Invalid annotation confidence value(s): ",
    paste(invalid_confidence, collapse = ", "),
    call. = FALSE
  )
}

approved_values <- tolower(as.character(annotation_decisions$approved))

if (anyNA(approved_values) || any(!approved_values %in% c("true", "t", "1"))) {
  stop(
    "Every annotation row must have approved=TRUE before labels are applied.",
    call. = FALSE
  )
}

annotation_decisions
```

# 14. Apply approved labels

Numeric WNN cluster identities are preserved in `wnn_cluster`. The final
biological label is stored separately in `cell_state`, and
`cell_state_cluster` keeps one-to-one traceability when multiple clusters share
the same state label.

```{r apply-approved-labels}
cluster_key <- as.character(multiome$wnn_cluster)

broad_compartment_lookup <- stats::setNames(
  annotation_decisions$broad_compartment,
  annotation_decisions$cluster
)

cell_state_lookup <- stats::setNames(
  annotation_decisions$cell_state,
  annotation_decisions$cluster
)

confidence_lookup <- stats::setNames(
  annotation_decisions$annotation_confidence,
  annotation_decisions$cluster
)

evidence_lookup <- stats::setNames(
  annotation_decisions$evidence_summary,
  annotation_decisions$cluster
)

multiome$broad_compartment <- unname(
  broad_compartment_lookup[cluster_key]
)

multiome$cell_state <- unname(cell_state_lookup[cluster_key])

multiome$annotation_confidence <- unname(
  confidence_lookup[cluster_key]
)

multiome$annotation_evidence <- unname(
  evidence_lookup[cluster_key]
)

multiome$cell_state_cluster <- paste0(
  multiome$cell_state,
  " | C",
  multiome$wnn_cluster
)

applied_annotation_columns <- c(
  "broad_compartment",
  "cell_state",
  "annotation_confidence",
  "annotation_evidence",
  "cell_state_cluster"
)

if (
  any(
    vapply(
      multiome[[]][, applied_annotation_columns, drop = FALSE],
      anyNA,
      logical(1)
    )
  )
) {
  stop("Missing values remain after applying annotation labels.", call. = FALSE)
}

multiome$broad_compartment <- factor(
  multiome$broad_compartment,
  levels = unique(annotation_decisions$broad_compartment)
)

multiome$cell_state <- factor(
  multiome$cell_state,
  levels = unique(annotation_decisions$cell_state)
)

multiome$annotation_confidence <- factor(
  multiome$annotation_confidence,
  levels = c("high", "moderate", "low")
)

multiome$cell_state_cluster <- factor(
  multiome$cell_state_cluster,
  levels = paste0(
    annotation_decisions$cell_state,
    " | C",
    annotation_decisions$cluster
  )
)

readr::write_csv(
  annotation_decisions,
  file.path(STEP6_DIR, "cell_state_annotations_applied.csv")
)

annotation_application_summary <- multiome[[]] |>
  dplyr::mutate(
    wnn_cluster = as.character(wnn_cluster),
    broad_compartment = as.character(broad_compartment),
    cell_state = as.character(cell_state),
    annotation_confidence = as.character(annotation_confidence)
  ) |>
  dplyr::count(
    wnn_cluster,
    broad_compartment,
    cell_state,
    annotation_confidence,
    name = "nuclei"
  ) |>
  dplyr::arrange(factor(wnn_cluster, levels = cluster_levels))

annotation_application_summary

readr::write_csv(
  annotation_application_summary,
  file.path(STEP6_DIR, "cell_state_annotation_summary.csv")
)
```

# 15. Visualize the approved annotation

```{r approved-annotation-plots, fig.width=13, fig.height=9}
compartment_levels <- levels(multiome$broad_compartment)
cell_state_levels <- levels(multiome$cell_state)

compartment_colors <- stats::setNames(
  grDevices::hcl.colors(
    length(compartment_levels),
    palette = "Dark 3"
  ),
  compartment_levels
)

cell_state_colors <- stats::setNames(
  grDevices::hcl.colors(
    length(cell_state_levels),
    palette = "Dynamic"
  ),
  cell_state_levels
)

p_broad_annotation <- DimPlot(
  multiome,
  reduction = "wnn.umap",
  group.by = "broad_compartment",
  cols = compartment_colors,
  label = TRUE,
  repel = TRUE,
  pt.size = 0.25,
  seed = 20260727,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Approved broad compartments",
    subtitle = "Labels were applied from the versioned manual decision table.",
    color = NULL
  ) +
  theme_multiome()

p_cell_state_annotation <- DimPlot(
  multiome,
  reduction = "wnn.umap",
  group.by = "cell_state",
  cols = cell_state_colors,
  label = TRUE,
  repel = TRUE,
  pt.size = 0.25,
  seed = 20260727,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Approved mouse-colon cell states",
    subtitle = "WNN clusters remain available separately as wnn_cluster.",
    color = NULL
  ) +
  theme_multiome()

print(p_broad_annotation)
print(p_cell_state_annotation)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "13_WNN_UMAP_broad_compartment.pdf"),
  plot = p_broad_annotation,
  width = 11,
  height = 8
)

ggplot2::ggsave(
  filename = file.path(STEP6_PLOT_DIR, "14_WNN_UMAP_cell_state.pdf"),
  plot = p_cell_state_annotation,
  width = 13,
  height = 9
)

cell_state_sample_summary <- multiome[[]] |>
  dplyr::mutate(
    cell_state = as.character(cell_state),
    sample_id = as.character(sample_id)
  ) |>
  dplyr::count(sample_id, cell_state, name = "nuclei") |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    sample_total = sum(nuclei),
    percent_of_sample = 100 * nuclei / sample_total
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    sample_id = factor(sample_id, levels = sample_sheet$sample_id),
    cell_state = factor(cell_state, levels = cell_state_levels)
  )

readr::write_csv(
  cell_state_sample_summary,
  file.path(STEP6_DIR, "cell_state_sample_composition_descriptive.csv")
)

p_cell_state_composition <- ggplot2::ggplot(
  cell_state_sample_summary,
  ggplot2::aes(
    x = sample_id,
    y = percent_of_sample,
    fill = cell_state
  )
) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::scale_fill_manual(values = cell_state_colors) +
  ggplot2::scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    expand = ggplot2::expansion(mult = c(0, 0.03))
  ) +
  ggplot2::labs(
    title = "Descriptive cell-state composition by pooled library",
    subtitle = "This visualization is not a replicated diet-composition test.",
    x = NULL,
    y = "Nuclei within pooled library",
    fill = NULL
  ) +
  theme_multiome()

print(p_cell_state_composition)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "15_cell_state_sample_composition_descriptive.pdf"
  ),
  plot = p_cell_state_composition,
  width = 12,
  height = 8
)
```

# 16. Save the annotated multimodal object

```{r save-step-six}
DefaultAssay(multiome) <- "RNA"
Idents(multiome) <- "cell_state"

step6_object_path <- file.path(
  STEP6_DIR,
  "Ppar_iKO_multiome_annotated.rds"
)

saveRDS(multiome, step6_object_path, compress = FALSE)

step6_object_info <- data.frame(
  path = step6_object_path,
  exists = file.exists(step6_object_path),
  size_gb = file.info(step6_object_path)$size / 1024^3,
  cells = ncol(multiome),
  WNN_clusters = length(cluster_levels),
  broad_compartments = nlevels(multiome$broad_compartment),
  cell_states = nlevels(multiome$cell_state)
)

step6_object_info

if (!isTRUE(step6_object_info$exists)) {
  stop("The Step 6 annotated object was not saved successfully.", call. = FALSE)
}

readr::write_csv(
  step6_parameters,
  file.path(STEP6_DIR, "annotation_parameters.csv")
)

writeLines(
  c(
    paste("Step 6 completed:", format(Sys.time())),
    paste("Cells:", ncol(multiome)),
    paste("Chosen clustering resolution:", CHOSEN_RESOLUTION),
    paste("Chosen cluster column:", CHOSEN_CLUSTER_COLUMN),
    paste("WNN clusters:", length(cluster_levels)),
    paste("Broad compartments:", nlevels(multiome$broad_compartment)),
    paste("Cell states:", nlevels(multiome$cell_state)),
    paste("Annotation decision table:", ANNOTATION_DECISION_PATH),
    "Final biological cell-state labels assigned: TRUE",
    "CON-versus-HFD treatment inference performed: FALSE"
  ),
  step6_complete_marker
)
```

# Interpretation checkpoint

Before moving to downstream biological analysis, be able to answer:

1. Which marker programs distinguish epithelial from immune, stromal,
   vascular, and neural clusters?
2. Why does `Kit` require contextual markers before distinguishing DCS from
   mast cells?
3. Which evidence separates noncycling stem/early progenitor, cycling
   stem/progenitor, and cycling secretory-progenitor hypotheses?
4. Do crypt-state conclusions remain directionally consistent in both pooled
   libraries?
5. Which clusters remain low confidence, and what additional experiment or
   analysis would resolve them?
6. Why are scaled negative DotPlot values not negative transcript counts?
7. Which labels are supported by both RNA and local ATAC accessibility?
8. Why must descriptive differences between CON and HFD libraries not be
   presented as replicated diet effects?

# What comes next?

With an approved annotated object, subsequent notebooks can perform:

- cell-state-specific RNA and ATAC characterization;
- regulatory-element and motif analyses within biologically coherent states;
- descriptive library composition and pseudobulk summaries; and
- carefully limited exploratory CON-versus-HFD contrasts that retain the
  pooled-library confounding caveat.

# References used for this step

- Sasaki et al. Reg4-positive deep crypt secretory cells support colonic stem
  cells: https://pmc.ncbi.nlm.nih.gov/articles/PMC5027439/
- Billing et al. Molecular and cellular diversity of large-intestinal
  enteroendocrine cells: https://pmc.ncbi.nlm.nih.gov/articles/PMC6812004/
- Seurat differential-expression testing:
  https://satijalab.org/seurat/reference/findallmarkers
- Signac coverage visualization:
  https://stuartlab.org/signac/reference/coverageplot
