---
title: "Step 6 - Annotate mouse-colon epithelial cell states"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 5 established the multimodal WNN graph and selected a working clustering resolution of 0.8. Those clusters are computational communities, not biological cell-type labels.

Step 6 converts those clusters into reviewed mouse-colon epithelial cell-state annotations.

The annotation strategy is deliberately evidence-based:

1. discover RNA markers for every WNN cluster;
2. examine marker effect size and prevalence rather than relying on P values;
3. evaluate coordinated mouse-colon marker programs;
4. determine whether cluster markers are consistent across the two pooled libraries;
5. audit sex-associated transcription because the two pools differ in sex composition;
6. investigate ambiguous or closely related populations directly;
7. use ATAC accessibility or gene activity as targeted supporting evidence;
8. record the final annotation decision separately from the Seurat object.

Final biological labels will **not** be assigned automatically from `FindAllMarkers()`.

## Statistical limitation

Marker discovery compares cells belonging to different biological states and is useful for annotation. However, cell-level P values must not be interpreted as replicated CON-versus-HFD treatment inference.

There is one pooled Cell Ranger library per diet, so diet and pooled-library identity remain confounded.

# 1. Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 10,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(readr)
})

configure_parallel_plan()

cat(
  "Configured future workers:",
  future::nbrOfWorkers(),
  "\n"
)
```

# 2. Define Step 6 parameters and output directories

The clustering resolution was reviewed in Step 5. Step 6 therefore treats it as
an input rather than reopening the clustering-resolution decision.

Marker discovery is intentionally permissive at first. We retain positive
markers expressed in at least 10% of cells in a cluster and with at least
0.25 log2-fold enrichment. Stronger prevalence and specificity filters are
applied later when ranking annotation candidates.

```{r step-six-parameters}
STEP6_DIR <- file.path(
  OUTPUT_DIR,
  "06_annotation"
)

STEP6_PLOT_DIR <- file.path(
  STEP6_DIR,
  "plots"
)

STEP6_MARKER_DIR <- file.path(
  STEP6_DIR,
  "markers"
)

dir.create(
  STEP6_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  STEP6_PLOT_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  STEP6_MARKER_DIR,
  recursive = TRUE,
  showWarnings = FALSE
)

CHOSEN_RESOLUTION <- 0.8
CHOSEN_CLUSTER_COLUMN <- "wnn_res_0.8"

MARKER_MIN_PCT <- 0.10
MARKER_LOGFC_THRESHOLD <- 0.25

step6_parameters <- data.frame(
  parameter = c(
    "WNN clustering resolution",
    "WNN cluster column",
    "RNA marker minimum pct",
    "RNA marker log2FC threshold"
  ),
  value = c(
    CHOSEN_RESOLUTION,
    CHOSEN_CLUSTER_COLUMN,
    MARKER_MIN_PCT,
    MARKER_LOGFC_THRESHOLD
  )
)

step6_parameters
```

# 3. Load and validate the finalized Step 5 object

Step 6 must begin from the finalized Step 5 checkpoint. It should never silently
recalculate WNN or choose a new clustering resolution.

```{r load-step-five}
step5_marker <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "STEP_05_COMPLETE.txt"
)

step5_object_path <- file.path(
  OUTPUT_DIR,
  "05_wnn",
  "Ppar_iKO_multiome_WNN_clustered.rds"
)

assert_file(
  step5_marker,
  "Step 5 completion marker"
)

assert_file(
  step5_object_path,
  "finalized Step 5 WNN object"
)

multiome <- readRDS(
  step5_object_path
)

required_assays <- c(
  "RNA",
  "ATAC"
)

missing_assays <- setdiff(
  required_assays,
  names(multiome)
)

if (length(missing_assays) > 0L) {
  stop(
    "Step 5 object is missing assay(s): ",
    paste(missing_assays, collapse = ", "),
    call. = FALSE
  )
}

required_reductions <- c(
  "pca",
  "lsi",
  "wnn.umap"
)

missing_reductions <- setdiff(
  required_reductions,
  names(multiome@reductions)
)

if (length(missing_reductions) > 0L) {
  stop(
    "Step 5 object is missing reduction(s): ",
    paste(missing_reductions, collapse = ", "),
    call. = FALSE
  )
}

if (!CHOSEN_CLUSTER_COLUMN %in% colnames(multiome[[]])) {
  stop(
    "The reviewed WNN cluster column was not found: ",
    CHOSEN_CLUSTER_COLUMN,
    call. = FALSE
  )
}

# Synchronize sample-level metadata from the current sample sheet.
#
# Older checkpoints may predate metadata fields that were later added to the
# workflow, such as pool_id. Rather than rebuilding expensive upstream
# checkpoints solely to add sample-level metadata, reconstruct those fields
# from sample_id and verify that any existing values agree with samples.csv.

sample_sheet <- read_sample_sheet()

sample_metadata_columns <- intersect(
  c(
    "sample_id",
    "pool_id",
    "genotype",
    "diet",
    "mice_pooled",
    "female_mice",
    "male_mice",
    "pooling_design"
  ),
  colnames(sample_sheet)
)

sample_index <- match(
  as.character(multiome$sample_id),
  sample_sheet$sample_id
)

if (anyNA(sample_index)) {
  missing_sample_ids <- unique(
    as.character(multiome$sample_id)[
      is.na(sample_index)
    ]
  )

  stop(
    "The Step 5 object contains sample_id values absent from samples.csv: ",
    paste(missing_sample_ids, collapse = ", "),
    call. = FALSE
  )
}

expected_sample_metadata <- sample_sheet[
  sample_index,
  sample_metadata_columns,
  drop = FALSE
]

rownames(expected_sample_metadata) <- colnames(
  multiome
)

# If a field already exists in the checkpoint, confirm that samples.csv does
# not disagree with it before replacing/refreshing the metadata.
existing_metadata_columns <- intersect(
  sample_metadata_columns,
  colnames(multiome[[]])
)

for (column_name in existing_metadata_columns) {

  existing_values <- multiome[[]][[
    column_name
  ]]

  expected_values <- expected_sample_metadata[[
    column_name
  ]]

  disagreement <- (
    !is.na(existing_values) &
      !is.na(expected_values) &
      as.character(existing_values) !=
        as.character(expected_values)
  )

  if (any(disagreement)) {
    stop(
      "Checkpoint metadata disagrees with samples.csv for column: ",
      column_name,
      call. = FALSE
    )
  }
}

multiome <- AddMetaData(
  object = multiome,
  metadata = expected_sample_metadata
)

required_metadata <- c(
  "sample_id",
  "pool_id",
  "diet",
  CHOSEN_CLUSTER_COLUMN
)

missing_metadata <- setdiff(
  required_metadata,
  colnames(multiome[[]])
)

if (length(missing_metadata) > 0L) {
  stop(
    "Step 5 object is missing required metadata after synchronization: ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

DefaultAssay(multiome) <- "RNA"

rna_layers <- SeuratObject::Layers(
  multiome[["RNA"]]
)

if (!"data" %in% rna_layers) {
  stop(
    "RNA normalized data layer is absent. ",
    "Step 6 should use the normalized RNA assay created in Step 4.",
    call. = FALSE
  )
}

Idents(multiome) <- multiome[[]][[
  CHOSEN_CLUSTER_COLUMN
]]

step5_validation <- data.frame(
  cells = ncol(multiome),
  RNA_features = nrow(multiome[["RNA"]]),
  ATAC_features = nrow(multiome[["ATAC"]]),
  WNN_clusters = length(unique(Idents(multiome))),
  samples = length(unique(multiome$sample_id)),
  RNA_layers = paste(rna_layers, collapse = ",")
)

step5_validation

table(
  Idents(multiome)
)
```

# 4. Inventory the WNN clusters before annotating them

Before looking at marker genes, record how many nuclei from each pooled library
occupy every WNN cluster.

These counts are useful for identifying sample-restricted clusters or rare
populations. They are **not** differential-abundance statistics.

```{r cluster-inventory}
# Total retained nuclei contributed by each pooled library.
sample_totals <- multiome[[]] |>
  dplyr::count(
    sample_id,
    diet,
    name = "sample_total"
  ) |>
  dplyr::mutate(
    global_sample_fraction =
      sample_total / sum(sample_total)
  )

sample_totals

cluster_inventory <- multiome[[]] |>
  dplyr::mutate(
    cluster = as.character(
      .data[[CHOSEN_CLUSTER_COLUMN]]
    )
  ) |>
  dplyr::count(
    cluster,
    sample_id,
    diet,
    name = "cells"
  ) |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::mutate(
    cluster_total = sum(cells),
    percent_of_cluster =
      100 * cells / cluster_total
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    sample_totals,
    by = c(
      "sample_id",
      "diet"
    )
  ) |>
  dplyr::mutate(
    # Fraction of that library's entire retained-cell population
    # represented by this cluster.
    percent_of_sample =
      100 * cells / sample_total,

    # Number of cells expected in this cluster if cluster membership
    # simply followed the overall library contribution.
    expected_cells_from_global_fraction =
      cluster_total * global_sample_fraction,

    # Values >1 indicate over-representation relative to the
    # library's overall contribution to the merged dataset.
    representation_ratio =
      cells / expected_cells_from_global_fraction,

    log2_representation_ratio =
      log2(representation_ratio)
  )

cluster_inventory

cluster_composition_wide <- cluster_inventory |>
  dplyr::select(
    cluster,
    diet,
    cells,
    percent_of_sample,
    representation_ratio
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = c(
      cells,
      percent_of_sample,
      representation_ratio
    ),
    values_fill = 0
  ) |>
  dplyr::mutate(
    CON_to_HFD_fraction_ratio =
      dplyr::if_else(
        percent_of_sample_HFD > 0,
        percent_of_sample_CON /
          percent_of_sample_HFD,
        NA_real_
      )
  ) |>
  dplyr::arrange(
    as.integer(cluster)
  )

cluster_composition_wide

readr::write_csv(
  cluster_inventory,
  file.path(
    STEP6_DIR,
    "cluster_inventory_by_library.csv"
  )
)

readr::write_csv(
  cluster_composition_wide,
  file.path(
    STEP6_DIR,
    "cluster_composition_normalized_by_library.csv"
  )
)
```

## Plot the reviewed WNN clusters

```{r plot-reviewed-clusters, fig.width=10, fig.height=8}
p_step6_clusters <- DimPlot(
  multiome,
  reduction = "wnn.umap",
  group.by = CHOSEN_CLUSTER_COLUMN,
  label = TRUE,
  repel = TRUE,
  label.size = 3,
  pt.size = 0.2
) +
  NoLegend() +
  ggplot2::ggtitle(
    "WNN clusters entering cell-state annotation"
  )

print(
  p_step6_clusters
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "01_WNN_clusters_entering_annotation.pdf"
  ),
  plot = p_step6_clusters,
  width = 10,
  height = 8
)
```

# 5. Discover positive RNA markers for every WNN cluster

`FindAllMarkers()` compares each cluster with the remaining nuclei.

The P values produced by this calculation are useful as part of marker
discovery, but they are not the primary criterion for biological annotation.
With thousands of nuclei, even very small effects can have extremely small
P values.

For annotation, we care especially about:

- effect size;
- fraction of cells expressing the marker in the cluster;
- expression outside the cluster;
- whether several genes support the same biological program;
- whether the pattern is reproducible across the two pooled libraries.

Marker discovery is run sequentially here so that the result is deterministic
and does not multiply the large Seurat object across several R workers.

```{r discover-rna-markers, cache=FALSE}
DefaultAssay(multiome) <- "RNA"

future::plan(
  future::sequential
)

cat(
  "Workers used for RNA marker discovery:",
  future::nbrOfWorkers(),
  "\n"
)

set.seed(20260727)

rna_markers <- FindAllMarkers(
  object = multiome,
  assay = "RNA",
  only.pos = TRUE,
  test.use = "wilcox",
  min.pct = MARKER_MIN_PCT,
  logfc.threshold = MARKER_LOGFC_THRESHOLD,
  return.thresh = 1,
  verbose = TRUE
)

configure_parallel_plan()

cat(
  "Workers restored after marker discovery:",
  future::nbrOfWorkers(),
  "\n"
)

if (nrow(rna_markers) == 0L) {
  stop(
    "FindAllMarkers returned no RNA markers.",
    call. = FALSE
  )
}

fc_column <- intersect(
  c(
    "avg_log2FC",
    "avg_logFC"
  ),
  colnames(rna_markers)
)

if (length(fc_column) != 1L) {
  stop(
    "Could not identify the fold-change column returned by FindAllMarkers().",
    call. = FALSE
  )
}

rna_markers$effect_size <- rna_markers[[
  fc_column
]]

rna_markers$pct_delta <- (
  rna_markers$pct.1 -
    rna_markers$pct.2
)

readr::write_csv(
  rna_markers,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_positive_markers_all_clusters.csv"
  )
)

cat(
  "RNA marker rows:",
  nrow(rna_markers),
  "\n"
)

cat(
  "Clusters with RNA markers:",
  length(unique(rna_markers$cluster)),
  "\n"
)
```

# 6. Rank marker candidates by effect size and prevalence

The table below is a review tool, not an automated annotation algorithm.

For the first-pass shortlist we require:

- at least 20% of cells in the cluster expressing the gene;
- at least a 10-percentage-point prevalence advantage over cells outside the
  cluster;
- positive effect size of at least 0.25 log2 fold-change.

We then retain the 15 strongest genes by effect size for each cluster.

```{r rank-rna-markers}
rna_marker_candidates <- rna_markers |>
  dplyr::filter(
    effect_size >= 0.25,
    pct.1 >= 0.20,
    pct_delta >= 0.10
  ) |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::arrange(
    dplyr::desc(effect_size),
    dplyr::desc(pct_delta),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(
    n = 15
  ) |>
  dplyr::ungroup()

readr::write_csv(
  rna_marker_candidates,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_top15_marker_candidates_by_cluster.csv"
  )
)

marker_review_summary <- rna_marker_candidates |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::summarise(
    top_genes = paste(
      gene,
      collapse = ", "
    ),
    .groups = "drop"
  )

marker_review_summary
```

# 7. Review broad mouse-colon epithelial marker programs

The broad marker panel is intentionally redundant. A biological identity should
be supported by several genes from the same program rather than one canonical
marker.

These programs are starting hypotheses. They will be refined after inspecting
the unbiased cluster-marker tables.

```{r define-colon-marker-programs}
colon_marker_programs <- list(
  Stem_crypt = c(
    "Lgr5",
    "Ascl2",
    "Smoc2",
    "Axin2",
    "Sox9",
    "Olfm4"
  ),
  Proliferative = c(
    "Mki67",
    "Top2a",
    "Ccna2",
    "Ccnb1",
    "Cdk1"
  ),
  Absorptive_colonocyte = c(
    "Slc26a3",
    "Aqp8",
    "Car1",
    "Car2",
    "Krt20"
  ),
  Goblet = c(
    "Muc2",
    "Tff3",
    "Clca1",
    "Fcgbp",
    "Agr2",
    "Spdef"
  ),
  Deep_crypt_secretory = c(
    "Reg4",
    "Mptx1",
    "Spink4",
    "Kit"
  ),
  Enteroendocrine = c(
    "Chga",
    "Chgb",
    "Neurod1",
    "Pyy",
    "Gcg",
    "Tph1"
  ),
  Tuft = c(
    "Dclk1",
    "Pou2f3",
    "Trpm5",
    "Avil"
  )
)

colon_marker_table <- data.frame(
  marker_program = rep(
    names(colon_marker_programs),
    lengths(colon_marker_programs)
  ),
  gene = unname(
    unlist(
      colon_marker_programs
    )
  ),
  stringsAsFactors = FALSE
)

colon_marker_table$detected <- (
  colon_marker_table$gene %in%
    rownames(multiome[["RNA"]])
)

colon_marker_table
```

```{r broad-marker-dotplot, fig.width=16, fig.height=9}
detected_colon_markers <- unique(
  colon_marker_table$gene[
    colon_marker_table$detected
  ]
)

p_broad_markers <- DotPlot(
  multiome,
  features = detected_colon_markers,
  group.by = CHOSEN_CLUSTER_COLUMN,
  assay = "RNA"
) +
  RotatedAxis() +
  ggplot2::ggtitle(
    "Broad mouse-colon epithelial programs across WNN clusters"
  )

print(
  p_broad_markers
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "02_broad_colon_marker_programs.pdf"
  ),
  plot = p_broad_markers,
  width = 16,
  height = 9
)
```

# 8. Audit sex-associated transcription before assigning labels

The two pooled libraries differ in sex composition. This means that
sex-associated transcription can contribute to apparent cluster or
library-specific expression patterns.

We do not attempt to infer the originating mouse for an individual nucleus.
Instead, this audit asks whether canonical X- or Y-linked transcripts appear
among the signals distinguishing clusters.

```{r sex-associated-marker-audit}
sex_associated_genes <- c(
  "Xist",
  "Tsix",
  "Ddx3y",
  "Eif2s3y",
  "Kdm5d",
  "Uty"
)

sex_marker_audit <- data.frame(
  gene = sex_associated_genes,
  present = sex_associated_genes %in%
    rownames(multiome[["RNA"]])
)

sex_marker_audit
```

```{r sex-associated-dotplot, fig.width=10, fig.height=7}
detected_sex_markers <- sex_marker_audit$gene[
  sex_marker_audit$present
]

if (length(detected_sex_markers) > 0L) {

  p_sex_markers <- DotPlot(
    multiome,
    features = detected_sex_markers,
    group.by = CHOSEN_CLUSTER_COLUMN,
    assay = "RNA"
  ) +
    RotatedAxis() +
    ggplot2::ggtitle(
      "Sex-associated transcription across WNN clusters"
    )

  print(
    p_sex_markers
  )

  ggplot2::ggsave(
    filename = file.path(
      STEP6_PLOT_DIR,
      "03_sex_associated_marker_audit.pdf"
    ),
    plot = p_sex_markers,
    width = 10,
    height = 7
  )
}
```

# 9. Audit broad epithelial, immune, vascular, and stromal lineages

Although the input material was enriched for colonic epithelium, unbiased RNA
marker discovery identified several clusters with strong immune, vascular, or
stromal transcriptional programs.

These populations should not be forced into epithelial cell-state labels.

Before proceeding to detailed epithelial annotation, we therefore perform a
broad-lineage audit using redundant marker programs. The purpose is to
distinguish major cellular compartments, not to assign fine subtypes from a
single marker.

```{r broad-lineage-marker-audit}
broad_lineage_programs <- list(
  Epithelial = c(
    "Epcam",
    "Krt8",
    "Krt18",
    "Krt19"
  ),
  T_NK = c(
    "Ptprc",
    "Cd3d",
    "Cd3e",
    "Trbc1",
    "Trbc2",
    "Nkg7",
    "Klrd1"
  ),
  B_cell = c(
    "Ptprc",
    "Cd79a",
    "Cd79b",
    "Ms4a1",
    "Cd74"
  ),
  Macrophage_myeloid = c(
    "Ptprc",
    "Lyz2",
    "Adgre1",
    "Csf1r",
    "C1qa",
    "C1qb",
    "C1qc"
  ),
  Mast = c(
    "Ptprc",
    "Kit",
    "Ms4a2",
    "Cpa3",
    "Mcpt1",
    "Mcpt2"
  ),
  Plasmacytoid_DC = c(
    "Ptprc",
    "Siglech",
    "Flt3",
    "Ccr9",
    "Pld4"
  ),
  Vascular_endothelial = c(
    "Cdh5",
    "Pecam1",
    "Kdr",
    "Esam"
  ),
  Lymphatic_endothelial = c(
    "Cdh5",
    "Flt4",
    "Prox1",
    "Lyve1",
    "Ccl21a",
    "Pdpn"
  ),
  Smooth_muscle = c(
    "Carmn",
    "Acta2",
    "Tagln",
    "Myh11",
    "Mylk"
  ),
  Fibroblast_stromal = c(
    "Col1a1",
    "Col1a2",
    "Col3a1",
    "Dcn",
    "Pdgfra"
  ),
  Mesothelial = c(
    "Lrrn4",
    "Upk3b",
    "Msln",
    "Wt1",
    "Rspo1"
  )
)

broad_lineage_table <- data.frame(
  marker_program = rep(
    names(broad_lineage_programs),
    lengths(broad_lineage_programs)
  ),
  gene = unname(
    unlist(
      broad_lineage_programs
    )
  ),
  stringsAsFactors = FALSE
)

broad_lineage_table$detected <- (
  broad_lineage_table$gene %in%
    rownames(multiome[["RNA"]])
)

broad_lineage_table
```

```{r broad-lineage-dotplot, fig.width=18, fig.height=9}
detected_broad_lineage_markers <- unique(
  broad_lineage_table$gene[
    broad_lineage_table$detected
  ]
)

p_broad_lineages <- DotPlot(
  multiome,
  features = detected_broad_lineage_markers,
  group.by = CHOSEN_CLUSTER_COLUMN,
  assay = "RNA"
) +
  RotatedAxis() +
  ggplot2::ggtitle(
    "Broad lineage programs across WNN clusters"
  )

print(
  p_broad_lineages
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "04_broad_lineage_marker_audit.pdf"
  ),
  plot = p_broad_lineages,
  width = 18,
  height = 9
)

readr::write_csv(
  broad_lineage_table,
  file.path(
    STEP6_DIR,
    "broad_lineage_marker_audit.csv"
  )
)
```

# 10. Focused epithelial-state audit with literature-supported markers

The broad-lineage audit separates epithelial cells from immune, vascular, stromal, and mesothelial populations. We can now examine epithelial states at higher resolution.

Marker programs used here are not treated as undocumented canonical knowledge.
Each program is linked to primary mouse-colon or large-intestinal literature, with curated marker resources used only as secondary support.

The marker panel is intentionally conservative. Dataset-specific genes emerging from `FindAllMarkers()` are considered additional evidence but are not silently promoted to literature-defined markers.

```{r epithelial-marker-sources}
epithelial_marker_sources <- data.frame(
  source_key = c(
    "AGING_MOUSE_COLON_2022",
    "CELLTAG_LARGE_INTESTINE_2019",
    "ROTHENBERG_COLON_CRYPT",
    "SASAKI_DCS_2016",
    "LARGE_INTESTINE_EEC_2019",
    "INTESTINAL_MARKER_REVIEW_2022",
    "PANGLOADB_2019"
  ),
  source = c(
    "Single-cell atlas of the aging mouse colon",
    "CellTag indexing: large-intestinal single-cell dataset",
    "Identification of a cKit+ colonic crypt base secretory cell",
    "Reg4+ deep crypt secretory cells function as epithelial niche for Lgr5+ stem cells",
    "Single-cell transcriptomic profiling of large intestinal enteroendocrine cells in mice",
    "Intestinal cellular heterogeneity and disease development revealed by single-cell technology",
    "PanglaoDB"
  ),
  evidence_scope = c(
    "mouse colon",
    "mouse large intestine",
    "mouse colon",
    "mouse colon",
    "mouse colon and rectum",
    "mouse and human intestine review",
    "curated mouse/human marker database"
  ),
  url = c(
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC9035718/",
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC6509836/",
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC3911891/",
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC5027439/",
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC6812004/",
    "https://pmc.ncbi.nlm.nih.gov/articles/PMC9433512/",
    "https://doi.org/10.1093/database/baz046"
  ),
  stringsAsFactors = FALSE
)

epithelial_marker_sources

readr::write_csv(
  epithelial_marker_sources,
  file.path(
    STEP6_DIR,
    "epithelial_marker_sources.csv"
  )
)
```

```{r define-evidence-backed-epithelial-programs}
epithelial_cluster_ids <- c(
  "0",
  "1",
  "2",
  "3",
  "4",
  "5",
  "6",
  "7",
  "8",
  "9",
  "11",
  "12",
  "14",
  "16",
  "18"
)

missing_epithelial_clusters <- setdiff(
  epithelial_cluster_ids,
  levels(Idents(multiome))
)

if (length(missing_epithelial_clusters) > 0L) {
  stop(
    "Requested epithelial cluster(s) are absent: ",
    paste(missing_epithelial_clusters, collapse = ", "),
    call. = FALSE
  )
}

epithelial_marker_programs <- list(
  Stem_ISC = c(
    "Lgr5",
    "Ascl2",
    "Smoc2",
    "Slc12a2"
  ),
  Proliferative = c(
    "Mki67",
    "Pcna",
    "Top2a"
  ),
  Colonocyte_precursor = c(
    "Selenbp1",
    "Cyp2c55",
    "Car1"
  ),
  Mature_colonocyte = c(
    "Krt20",
    "Slc26a3"
  ),
  Goblet = c(
    "Muc2",
    "Tff3",
    "Spdef",
    "Agr2",
    "Fcgbp",
    "Clca1"
  ),
  Deep_crypt_secretory = c(
    "Reg4",
    "Kit"
  ),
  Enteroendocrine_general = c(
    "Chga",
    "Chgb",
    "Neurod1"
  ),
  Enterochromaffin = c(
    "Tph1"
  ),
  L_cell = c(
    "Gcg",
    "Pyy",
    "Insl5"
  ),
  Tuft = c(
    "Dclk1",
    "Pou2f3",
    "Trpm5"
  )
)

marker_program_sources <- c(
  Stem_ISC =
    "AGING_MOUSE_COLON_2022;CELLTAG_LARGE_INTESTINE_2019",
  Proliferative =
    "AGING_MOUSE_COLON_2022;INTESTINAL_MARKER_REVIEW_2022",
  Colonocyte_precursor =
    "AGING_MOUSE_COLON_2022",
  Mature_colonocyte =
    "ROTHENBERG_COLON_CRYPT;CELLTAG_LARGE_INTESTINE_2019",
  Goblet =
    "ROTHENBERG_COLON_CRYPT;CELLTAG_LARGE_INTESTINE_2019",
  Deep_crypt_secretory =
    "SASAKI_DCS_2016;ROTHENBERG_COLON_CRYPT",
  Enteroendocrine_general =
    "LARGE_INTESTINE_EEC_2019;INTESTINAL_MARKER_REVIEW_2022",
  Enterochromaffin =
    "LARGE_INTESTINE_EEC_2019",
  L_cell =
    "LARGE_INTESTINE_EEC_2019",
  Tuft =
    "INTESTINAL_MARKER_REVIEW_2022;PANGLOADB_2019"
)

epithelial_marker_table <- data.frame(
  marker_program = rep(
    names(epithelial_marker_programs),
    lengths(epithelial_marker_programs)
  ),
  gene = unname(
    unlist(
      epithelial_marker_programs
    )
  ),
  stringsAsFactors = FALSE
)

epithelial_marker_table$source_keys <- unname(
  marker_program_sources[
    epithelial_marker_table$marker_program
  ]
)

# A gene being present as a row in the 10x feature matrix does not prove that
# any RNA molecules were detected. Audit both feature-set presence and actual
# nonzero counts.

rna_counts <- SeuratObject::LayerData(
  object = multiome,
  assay = "RNA",
  layer = "counts"
)

feature_index <- match(
  epithelial_marker_table$gene,
  rownames(rna_counts)
)

epithelial_marker_table$present_in_feature_set <- !is.na(
  feature_index
)

epithelial_marker_table$total_UMIs <- NA_real_
epithelial_marker_table$positive_nuclei <- NA_integer_
epithelial_marker_table$percent_positive <- NA_real_

present_rows <- which(
  epithelial_marker_table$present_in_feature_set
)

if (length(present_rows) > 0L) {

  marker_counts <- rna_counts[
    feature_index[present_rows],
    ,
    drop = FALSE
  ]

  epithelial_marker_table$total_UMIs[present_rows] <-
    as.numeric(
      Matrix::rowSums(marker_counts)
    )

  epithelial_marker_table$positive_nuclei[present_rows] <-
    as.integer(
      Matrix::rowSums(marker_counts > 0)
    )

  epithelial_marker_table$percent_positive[present_rows] <-
    100 *
      epithelial_marker_table$positive_nuclei[present_rows] /
      ncol(rna_counts)
}

epithelial_marker_table$detected <- (
  epithelial_marker_table$present_in_feature_set &
    !is.na(epithelial_marker_table$total_UMIs) &
    epithelial_marker_table$total_UMIs > 0
)

epithelial_marker_table$status <- dplyr::case_when(
  !epithelial_marker_table$present_in_feature_set ~
    "not_in_RNA_feature_set",
  epithelial_marker_table$total_UMIs == 0 ~
    "present_but_zero_counts",
  TRUE ~
    "detected"
)

epithelial_marker_table

readr::write_csv(
  epithelial_marker_table,
  file.path(
    STEP6_DIR,
    "epithelial_marker_evidence.csv"
  )
)
```

```{r focused-epithelial-dotplot, fig.width=19, fig.height=9.5}
# Keep machine-readable program names in the evidence table, but use shorter
# display labels so narrow DotPlot feature groups remain legible.

epithelial_plot_labels <- c(
  Stem_ISC = "Stem / ISC",
  Proliferative = "Proliferative",
  Colonocyte_precursor = "Colonocyte precursor",
  Mature_colonocyte = "Mature colonocyte",
  Goblet = "Goblet",
  Deep_crypt_secretory = "DCS",
  Enteroendocrine_general = "EEC",
  Enterochromaffin = "EC",
  L_cell = "L cell",
  Tuft = "Tuft"
)

missing_plot_labels <- setdiff(
  names(epithelial_marker_programs),
  names(epithelial_plot_labels)
)

if (length(missing_plot_labels) > 0L) {
  stop(
    "Plot labels are missing for marker program(s): ",
    paste(missing_plot_labels, collapse = ", "),
    call. = FALSE
  )
}

detected_epithelial_genes <- unique(
  epithelial_marker_table$gene[
    epithelial_marker_table$detected
  ]
)

epithelial_plot_programs <- lapply(
  epithelial_marker_programs,
  function(genes) {
    genes[
      genes %in% detected_epithelial_genes
    ]
  }
)

empty_programs <- names(
  epithelial_plot_programs
)[
  lengths(epithelial_plot_programs) == 0L
]

if (length(empty_programs) > 0L) {
  warning(
    "No detected genes were available for program(s): ",
    paste(empty_programs, collapse = ", ")
  )
}

epithelial_plot_programs <- epithelial_plot_programs[
  lengths(epithelial_plot_programs) > 0L
]

names(epithelial_plot_programs) <- unname(
  epithelial_plot_labels[
    names(epithelial_plot_programs)
  ]
)

p_epithelial_states <- DotPlot(
  multiome,
  features = epithelial_plot_programs,
  assay = "RNA",
  idents = epithelial_cluster_ids,
  group.by = CHOSEN_CLUSTER_COLUMN
) +
  RotatedAxis() +
  ggplot2::labs(
    title =
      "Literature-supported epithelial-state programs across WNN clusters",
    subtitle =
      "DCS = deep-crypt secretory; EEC = enteroendocrine; EC = enterochromaffin"
  ) +
  ggplot2::theme(
    strip.text.x = ggplot2::element_text(
      size = 9,
      face = "bold"
    ),
    axis.text.x = ggplot2::element_text(
      size = 8
    )
  )

print(
  p_epithelial_states
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "05_focused_epithelial_state_audit.pdf"
  ),
  plot = p_epithelial_states,
  width = 19,
  height = 9.5
)

readr::write_csv(
  p_epithelial_states$data,
  file.path(
    STEP6_DIR,
    "focused_epithelial_state_DotPlot_data.csv"
  )
)
```

# 11. Focused adjudication of crypt, stem, and proliferative clusters

The broad epithelial-state audit identified clusters 4, 7, and 18 as closely
related crypt-associated populations requiring direct comparison.

These clusters are not assigned final identities here. In particular,
proliferation is treated as a cell state rather than as proof of a
transit-amplifying lineage. A cycling ISC, absorptive progenitor, or
secretory-committed progenitor can all express cell-cycle genes.

This section asks:

1. which genes distinguish clusters 4, 7, and 18 directly;
2. whether ISC, proliferation, and secretory programs are detected in the same
   nuclei;
3. whether the distinguishing patterns are visible in both pooled libraries;
4. whether cluster 18 shows evidence of abnormal complexity or residual
   doublets.

```{r define-focused-crypt-programs}
crypt_cluster_ids <- c(
  "4",
  "7",
  "18"
)

missing_crypt_clusters <- setdiff(
  crypt_cluster_ids,
  levels(Idents(multiome))
)

if (length(missing_crypt_clusters) > 0L) {
  stop(
    "Focused crypt cluster(s) are absent: ",
    paste(missing_crypt_clusters, collapse = ", "),
    call. = FALSE
  )
}

crypt_marker_programs <- list(
  ISC_Wnt = c(
    "Lgr5",
    "Ascl2",
    "Smoc2",
    "Axin2",
    "Olfm4",
    "Lrig1",
    "Ephb2",
    "Tnfrsf19"
  ),
  Proliferative = c(
    "Mki67",
    "Pcna",
    "Top2a",
    "Cdk1",
    "Ccnb1",
    "Ube2c",
    "Birc5"
  ),
  Secretory_commitment = c(
    "Atoh1",
    "Dll1",
    "Gfi1",
    "Spdef"
  ),
  Goblet_DCS = c(
    "Muc2",
    "Agr2",
    "Tff3",
    "Fcgbp",
    "Clca1",
    "Spink4",
    "Mptx1",
    "Reg4",
    "Kit"
  )
)

crypt_marker_table <- data.frame(
  marker_program = rep(
    names(crypt_marker_programs),
    lengths(crypt_marker_programs)
  ),
  gene = unname(
    unlist(crypt_marker_programs)
  ),
  stringsAsFactors = FALSE
)

rna_counts <- SeuratObject::LayerData(
  multiome,
  assay = "RNA",
  layer = "counts"
)

crypt_feature_index <- match(
  crypt_marker_table$gene,
  rownames(rna_counts)
)

crypt_marker_table$present_in_feature_set <- !is.na(
  crypt_feature_index
)

crypt_marker_table$total_UMIs <- NA_real_
crypt_marker_table$positive_nuclei <- NA_integer_
crypt_marker_table$percent_positive <- NA_real_

present_crypt_rows <- which(
  crypt_marker_table$present_in_feature_set
)

if (length(present_crypt_rows) > 0L) {

  crypt_marker_counts <- rna_counts[
    crypt_feature_index[present_crypt_rows],
    ,
    drop = FALSE
  ]

  crypt_marker_table$total_UMIs[present_crypt_rows] <-
    as.numeric(
      Matrix::rowSums(crypt_marker_counts)
    )

  crypt_marker_table$positive_nuclei[present_crypt_rows] <-
    as.integer(
      Matrix::rowSums(crypt_marker_counts > 0)
    )

  crypt_marker_table$percent_positive[present_crypt_rows] <-
    100 *
      crypt_marker_table$positive_nuclei[present_crypt_rows] /
      ncol(rna_counts)
}

crypt_marker_table$detected <- (
  crypt_marker_table$present_in_feature_set &
    !is.na(crypt_marker_table$total_UMIs) &
    crypt_marker_table$total_UMIs > 0
)

crypt_marker_table

readr::write_csv(
  crypt_marker_table,
  file.path(
    STEP6_DIR,
    "focused_crypt_marker_evidence.csv"
  )
)
```

```{r focused-crypt-program-dotplot, fig.width=13, fig.height=5.5}
detected_crypt_genes <- unique(
  crypt_marker_table$gene[
    crypt_marker_table$detected
  ]
)

crypt_plot_programs <- lapply(
  crypt_marker_programs,
  function(genes) {
    genes[
      genes %in% detected_crypt_genes
    ]
  }
)

crypt_plot_programs <- crypt_plot_programs[
  lengths(crypt_plot_programs) > 0L
]

names(crypt_plot_programs) <- c(
  "ISC / Wnt",
  "Proliferative",
  "Secretory commitment",
  "Goblet / DCS"
)[
  match(
    names(crypt_plot_programs),
    names(crypt_marker_programs)
  )
]

p_crypt_programs <- DotPlot(
  multiome,
  features = crypt_plot_programs,
  assay = "RNA",
  idents = crypt_cluster_ids,
  group.by = CHOSEN_CLUSTER_COLUMN
) +
  RotatedAxis() +
  ggplot2::labs(
    title =
      "Focused crypt-state programs across clusters 4, 7, and 18",
    subtitle =
      "Proliferation is evaluated separately from lineage identity"
  ) +
  ggplot2::theme(
    strip.text.x = ggplot2::element_text(
      size = 10,
      face = "bold"
    )
  )

print(
  p_crypt_programs
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "06_focused_crypt_program_audit.pdf"
  ),
  plot = p_crypt_programs,
  width = 13,
  height = 5.5
)

readr::write_csv(
  p_crypt_programs$data,
  file.path(
    STEP6_DIR,
    "focused_crypt_program_DotPlot_data.csv"
  )
)
```

## Direct pairwise RNA comparisons

Each comparison uses the normalized RNA assay and asks which genes distinguish
one cluster from the other. The adjusted P values are cell-level marker-ranking
statistics, not replicated biological inference. Annotation decisions emphasize
effect size, prevalence difference, coordinated programs, and consistency
between the pooled libraries.

```{r pairwise-rna-clusters-4-7-18, cache=FALSE}
pairwise_comparisons <- data.frame(
  ident_1 = c(
    "4",
    "4",
    "7"
  ),
  ident_2 = c(
    "7",
    "18",
    "18"
  ),
  stringsAsFactors = FALSE
)

get_effect_column <- function(marker_table) {

  candidates <- intersect(
    c(
      "avg_log2FC",
      "avg_log2_FC",
      "avg_logFC"
    ),
    colnames(marker_table)
  )

  if (length(candidates) != 1L) {
    stop(
      "Could not identify one fold-change column.",
      call. = FALSE
    )
  }

  candidates
}

future::plan(
  future::sequential
)

pairwise_rna_results <- vector(
  "list",
  nrow(pairwise_comparisons)
)

for (i in seq_len(nrow(pairwise_comparisons))) {

  ident_1 <- pairwise_comparisons$ident_1[i]
  ident_2 <- pairwise_comparisons$ident_2[i]

  marker_result <- FindMarkers(
    object = multiome,
    ident.1 = ident_1,
    ident.2 = ident_2,
    assay = "RNA",
    test.use = "wilcox",
    only.pos = FALSE,
    min.pct = 0.10,
    logfc.threshold = 0,
    return.thresh = 1,
    verbose = TRUE
  )

  effect_column <- get_effect_column(
    marker_result
  )

  marker_result$gene <- rownames(
    marker_result
  )

  marker_result$effect_size <- marker_result[[
    effect_column
  ]]

  marker_result$pct_delta <- (
    marker_result$pct.1 -
      marker_result$pct.2
  )

  marker_result$comparison <- paste0(
    ident_1,
    "_vs_",
    ident_2
  )

  marker_result$ident_1 <- ident_1
  marker_result$ident_2 <- ident_2

  marker_result$higher_in <- dplyr::case_when(
    marker_result$effect_size > 0 ~ ident_1,
    marker_result$effect_size < 0 ~ ident_2,
    TRUE ~ "equal"
  )

  rownames(marker_result) <- NULL

  pairwise_rna_results[[i]] <- marker_result
}

configure_parallel_plan()

pairwise_rna_results <- dplyr::bind_rows(
  pairwise_rna_results
)

pairwise_review_candidates <- pairwise_rna_results |>
  dplyr::filter(
    abs(effect_size) >= 0.25,
    abs(pct_delta) >= 0.10
  ) |>
  dplyr::group_by(
    comparison,
    higher_in
  ) |>
  dplyr::slice_max(
    order_by = abs(effect_size),
    n = 25,
    with_ties = FALSE
  ) |>
  dplyr::ungroup()

readr::write_csv(
  pairwise_rna_results,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_pairwise_clusters_4_7_18_all.csv"
  )
)

readr::write_csv(
  pairwise_review_candidates,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_pairwise_clusters_4_7_18_review_candidates.csv"
  )
)

pairwise_review_candidates
```

## Per-cell RNA co-detection

A cluster-level DotPlot can show two genes in the same cluster even when they
occur in different nuclei. This audit therefore measures whether lineage and
proliferation programs are detected within the same individual nuclei.

Because single-nucleus RNA measurements are sparse, failure to detect a pair is
not proof that the transcripts are biologically absent.

```{r per-cell-crypt-program-codetection}
cluster_vector <- as.character(
  multiome[[]][[
    CHOSEN_CLUSTER_COLUMN
  ]]
)

names(cluster_vector) <- rownames(
  multiome[[]]
)

crypt_cells <- names(
  cluster_vector
)[
  cluster_vector %in% crypt_cluster_ids
]

detected_crypt_programs <- lapply(
  crypt_marker_programs,
  function(genes) {
    genes[
      genes %in% detected_crypt_genes
    ]
  }
)

crypt_counts <- rna_counts[
  unique(
    unlist(
      detected_crypt_programs
    )
  ),
  crypt_cells,
  drop = FALSE
]

crypt_cell_evidence <- data.frame(
  cell = crypt_cells,
  cluster = unname(
    cluster_vector[
      crypt_cells
    ]
  ),
  stringsAsFactors = FALSE
)

for (program_name in names(detected_crypt_programs)) {

  program_genes <- detected_crypt_programs[[
    program_name
  ]]

  output_column <- paste0(
    program_name,
    "_genes_detected"
  )

  if (length(program_genes) == 0L) {
    crypt_cell_evidence[[
      output_column
    ]] <- 0L
  } else {
    crypt_cell_evidence[[
      output_column
    ]] <- as.integer(
      Matrix::colSums(
        crypt_counts[
          program_genes,
          ,
          drop = FALSE
        ] > 0
      )
    )
  }
}

crypt_codetection_summary <- crypt_cell_evidence |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),

    ISC_Wnt_any_pct =
      100 * mean(ISC_Wnt_genes_detected > 0),

    Proliferative_any_pct =
      100 * mean(Proliferative_genes_detected > 0),

    Secretory_commitment_any_pct =
      100 * mean(Secretory_commitment_genes_detected > 0),

    Goblet_DCS_any_pct =
      100 * mean(Goblet_DCS_genes_detected > 0),

    ISC_Wnt_and_Proliferative_pct =
      100 * mean(
        ISC_Wnt_genes_detected > 0 &
          Proliferative_genes_detected > 0
      ),

    Secretory_commitment_and_Proliferative_pct =
      100 * mean(
        Secretory_commitment_genes_detected > 0 &
          Proliferative_genes_detected > 0
      ),

    Goblet_DCS_and_Proliferative_pct =
      100 * mean(
        Goblet_DCS_genes_detected > 0 &
          Proliferative_genes_detected > 0
      ),

    .groups = "drop"
  )

gene_pair_specification <- data.frame(
  pair = c(
    "Lgr5_and_Mki67",
    "Smoc2_and_Mki67",
    "Muc2_and_Mki67",
    "Kit_and_Mki67",
    "Reg4_and_Mki67"
  ),
  gene_1 = c(
    "Lgr5",
    "Smoc2",
    "Muc2",
    "Kit",
    "Reg4"
  ),
  gene_2 = rep(
    "Mki67",
    5
  ),
  stringsAsFactors = FALSE
)

gene_pair_specification <- gene_pair_specification |>
  dplyr::filter(
    gene_1 %in% rownames(crypt_counts),
    gene_2 %in% rownames(crypt_counts)
  )

gene_pair_codetection <- dplyr::bind_rows(
  lapply(
    crypt_cluster_ids,
    function(cluster_id) {

      cluster_cells <- crypt_cells[
        cluster_vector[
          crypt_cells
        ] == cluster_id
      ]

      dplyr::bind_rows(
        lapply(
          seq_len(nrow(gene_pair_specification)),
          function(i) {

            gene_1 <- gene_pair_specification$gene_1[i]
            gene_2 <- gene_pair_specification$gene_2[i]

            gene_1_detected <- as.numeric(
              crypt_counts[
                gene_1,
                cluster_cells,
                drop = TRUE
              ] > 0
            )

            gene_2_detected <- as.numeric(
              crypt_counts[
                gene_2,
                cluster_cells,
                drop = TRUE
              ] > 0
            )

            data.frame(
              cluster = cluster_id,
              nuclei = length(cluster_cells),
              pair = gene_pair_specification$pair[i],
              gene_1 = gene_1,
              gene_2 = gene_2,
              percent_gene_1_detected =
                100 * mean(gene_1_detected),
              percent_gene_2_detected =
                100 * mean(gene_2_detected),
              percent_codetected =
                100 * mean(
                  gene_1_detected &
                    gene_2_detected
                )
            )
          }
        )
      )
    }
  )
)

readr::write_csv(
  crypt_cell_evidence,
  file.path(
    STEP6_DIR,
    "per_cell_crypt_program_detection.csv"
  )
)

readr::write_csv(
  crypt_codetection_summary,
  file.path(
    STEP6_DIR,
    "crypt_program_codetection_summary.csv"
  )
)

readr::write_csv(
  gene_pair_codetection,
  file.path(
    STEP6_DIR,
    "crypt_gene_pair_codetection_summary.csv"
  )
)

crypt_codetection_summary
gene_pair_codetection
```

## Refine per-nucleus crypt-program definitions

The broad `any gene in program` rule is too permissive for per-nucleus classification. In particular, broadly expressed crypt genes can inflate an ISC/Wnt flag without showing that the same nucleus expresses a bona fide ISC marker.

The following rules are deliberately explicit and conservative:

- ISC core: `Lgr5` plus `Ascl2` or `Smoc2`;
- proliferative: `Mki67` or `Top2a`;
- secretory commitment: `Atoh1` and at least one of `Dll1`, `Gfi1`, or `Spdef`;
- goblet-associated: `Muc2` detected;
- DCS-associated: both `Reg4` and `Kit` detected.

These are annotation-support flags, not final cell labels. `Olfm4` remains useful in descriptive marker plots but is excluded from the binary ISC rule because of its low detectability in this colon dataset.

```{r refined-crypt-program-codetection, fig.width=14, fig.height=7}
CRYPT_REVIEW_CLUSTERS <- c(
  "4",
  "7",
  "18"
)

CRYPT_CLUSTER_COLORS <- c(
  "4" = "green4",
  "7" = "greenyellow",
  "18" = "yellow"
)

missing_crypt_clusters <- setdiff(
  CRYPT_REVIEW_CLUSTERS,
  levels(Idents(multiome))
)

if (length(missing_crypt_clusters) > 0L) {
  stop(
    "Focused crypt cluster(s) are absent: ",
    paste(missing_crypt_clusters, collapse = ", "),
    call. = FALSE
  )
}

crypt_program_definitions <- data.frame(
  program = c(
    "ISC_core",
    "Proliferative",
    "Secretory_commitment",
    "Goblet_Muc2",
    "DCS_Reg4_Kit"
  ),
  display_name = c(
    "ISC core",
    "Proliferative",
    "Secretory commitment",
    "Goblet (Muc2+)",
    "DCS (Reg4+ Kit+)"
  ),
  genes = c(
    "Lgr5 AND (Ascl2 OR Smoc2)",
    "Mki67 OR Top2a",
    "Atoh1 OR Dll1 OR Gfi1 OR Spdef",
    "Muc2",
    "Reg4 AND Kit"
  ),
  rule = c(
    "Lgr5 > 0 and at least one of Ascl2 or Smoc2 > 0",
    "at least one of Mki67 or Top2a > 0",
    "at least one of Atoh1, Dll1, Gfi1, or Spdef > 0",
    "Muc2 > 0",
    "Reg4 > 0 and Kit > 0"
  ),
  source_keys = c(
    "AGING_MOUSE_COLON_2022;CELLTAG_LARGE_INTESTINE_2019",
    "AGING_MOUSE_COLON_2022;INTESTINAL_MARKER_REVIEW_2022",
    "ROTHENBERG_COLON_CRYPT;CELLTAG_LARGE_INTESTINE_2019",
    "ROTHENBERG_COLON_CRYPT;CELLTAG_LARGE_INTESTINE_2019",
    "SASAKI_DCS_2016;ROTHENBERG_COLON_CRYPT"
  ),
  interpretation = rep(
    "Per-nucleus RNA co-detection flag for annotation support; not a final label",
    5
  ),
  stringsAsFactors = FALSE
)

crypt_program_definitions

readr::write_csv(
  crypt_program_definitions,
  file.path(
    STEP6_DIR,
    "focused_crypt_program_definitions.csv"
  )
)

codetection_genes <- c(
  "Lgr5",
  "Ascl2",
  "Smoc2",
  "Mki67",
  "Top2a",
  "Atoh1",
  "Dll1",
  "Gfi1",
  "Spdef",
  "Muc2",
  "Reg4",
  "Kit"
)

DefaultAssay(multiome) <- "RNA"

rna_counts <- SeuratObject::LayerData(
  multiome,
  assay = "RNA",
  layer = "counts"
)

missing_codetection_genes <- setdiff(
  codetection_genes,
  rownames(rna_counts)
)

if (length(missing_codetection_genes) > 0L) {
  stop(
    "Required co-detection gene(s) are absent from the RNA feature set: ",
    paste(missing_codetection_genes, collapse = ", "),
    call. = FALSE
  )
}

crypt_review_cells <- unlist(
  lapply(
    CRYPT_REVIEW_CLUSTERS,
    function(cluster_id) {
      WhichCells(
        multiome,
        idents = cluster_id
      )
    }
  ),
  use.names = FALSE
)

crypt_detection_matrix <- t(
  as.matrix(
    rna_counts[
      codetection_genes,
      crypt_review_cells,
      drop = FALSE
    ] > 0
  )
)

crypt_detection_matrix <- as.data.frame(
  crypt_detection_matrix,
  check.names = FALSE
)

per_cell_crypt_program_detection <- data.frame(
  nucleus = rownames(crypt_detection_matrix),
  cluster = factor(
    as.character(
      multiome[[]][
        rownames(crypt_detection_matrix),
        CHOSEN_CLUSTER_COLUMN
      ]
    ),
    levels = CRYPT_REVIEW_CLUSTERS
  ),
  sample_id = as.character(
    multiome[[]][
      rownames(crypt_detection_matrix),
      "sample_id"
    ]
  ),
  crypt_detection_matrix,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

per_cell_crypt_program_detection$ISC_core <- with(
  per_cell_crypt_program_detection,
  Lgr5 & (Ascl2 | Smoc2)
)

per_cell_crypt_program_detection$Proliferative <- with(
  per_cell_crypt_program_detection,
  Mki67 | Top2a
)

per_cell_crypt_program_detection$Secretory_commitment <- with(
  per_cell_crypt_program_detection,
  Atoh1 & (Dll1 | Gfi1 | Spdef)
)

per_cell_crypt_program_detection$Goblet_Muc2 <- (
  per_cell_crypt_program_detection$Muc2
)

per_cell_crypt_program_detection$DCS_Reg4_Kit <- with(
  per_cell_crypt_program_detection,
  Reg4 & Kit
)

crypt_program_columns <- crypt_program_definitions$program

crypt_program_detection_summary <-
  per_cell_crypt_program_detection |>
  dplyr::select(
    nucleus,
    cluster,
    sample_id,
    dplyr::all_of(crypt_program_columns)
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(crypt_program_columns),
    names_to = "program",
    values_to = "detected"
  ) |>
  dplyr::group_by(
    cluster,
    program
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    detected_nuclei = sum(detected),
    percent_detected = 100 * mean(detected),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    crypt_program_definitions[
      ,
      c(
        "program",
        "display_name"
      )
    ],
    by = "program"
  ) |>
  dplyr::mutate(
    display_name = factor(
      display_name,
      levels = crypt_program_definitions$display_name
    )
  )

crypt_program_pair_values <- data.frame(
  nucleus = per_cell_crypt_program_detection$nucleus,
  cluster = per_cell_crypt_program_detection$cluster,
  sample_id = per_cell_crypt_program_detection$sample_id,
  ISC_core_and_proliferative = with(
    per_cell_crypt_program_detection,
    ISC_core & Proliferative
  ),
  Secretory_commitment_and_proliferative = with(
    per_cell_crypt_program_detection,
    Secretory_commitment & Proliferative
  ),
  Goblet_Muc2_and_proliferative = with(
    per_cell_crypt_program_detection,
    Goblet_Muc2 & Proliferative
  ),
  DCS_Reg4_Kit_and_proliferative = with(
    per_cell_crypt_program_detection,
    DCS_Reg4_Kit & Proliferative
  ),
  stringsAsFactors = FALSE
)

crypt_pair_labels <- c(
  ISC_core_and_proliferative = "ISC core + proliferative",
  Secretory_commitment_and_proliferative =
    "Secretory commitment + proliferative",
  Goblet_Muc2_and_proliferative = "Goblet (Muc2+) + proliferative",
  DCS_Reg4_Kit_and_proliferative =
    "DCS (Reg4+ Kit+) + proliferative"
)

crypt_program_codetection_summary <-
  crypt_program_pair_values |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(names(crypt_pair_labels)),
    names_to = "pair",
    values_to = "codetected"
  ) |>
  dplyr::group_by(
    cluster,
    pair
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    codetected_nuclei = sum(codetected),
    percent_codetected = 100 * mean(codetected),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    pair_label = factor(
      unname(crypt_pair_labels[pair]),
      levels = unname(crypt_pair_labels)
    )
  )

crypt_program_detection_summary
crypt_program_codetection_summary

readr::write_csv(
  per_cell_crypt_program_detection,
  file.path(
    STEP6_DIR,
    "per_cell_crypt_program_detection.csv"
  )
)

readr::write_csv(
  crypt_program_detection_summary,
  file.path(
    STEP6_DIR,
    "focused_crypt_program_detection_summary.csv"
  )
)

readr::write_csv(
  crypt_program_codetection_summary,
  file.path(
    STEP6_DIR,
    "focused_crypt_program_codetection_summary.csv"
  )
)

p_crypt_program_detection <- ggplot2::ggplot(
  crypt_program_detection_summary,
  ggplot2::aes(
    x = display_name,
    y = cluster,
    fill = percent_detected
  )
) +
  ggplot2::geom_tile(
    color = "white",
    linewidth = 0.8
  ) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = sprintf("%.1f%%", percent_detected)
    ),
    size = 3.2
  ) +
  ggplot2::scale_fill_gradientn(
    colors = c(
      "#F7FBFF",
      "#6BAED6",
      "#08306B"
    ),
    limits = c(
      0,
      100
    )
  ) +
  ggplot2::labs(
    title = "Per-nucleus program detection",
    x = NULL,
    y = "WNN cluster",
    fill = "Nuclei (%)"
  ) +
  ggplot2::theme_minimal(
    base_size = 11
  ) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    axis.text.x = ggplot2::element_text(
      angle = 35,
      hjust = 1
    )
  )

p_crypt_program_pairs <- ggplot2::ggplot(
  crypt_program_codetection_summary,
  ggplot2::aes(
    x = cluster,
    y = percent_codetected,
    fill = cluster
  )
) +
  ggplot2::geom_col(
    width = 0.68
  ) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = sprintf("%.1f%%", percent_codetected)
    ),
    vjust = -0.35,
    size = 3
  ) +
  ggplot2::facet_wrap(
    ~pair_label,
    ncol = 2,
    scales = "free_y"
  ) +
  ggplot2::scale_fill_manual(
    values = CRYPT_CLUSTER_COLORS,
    guide = "none"
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(
      mult = c(
        0,
        0.18
      )
    )
  ) +
  ggplot2::labs(
    title = "Per-nucleus co-detection with proliferation",
    x = "WNN cluster",
    y = "Nuclei with co-detection (%)"
  ) +
  ggplot2::theme_classic(
    base_size = 11
  ) +
  ggplot2::theme(
    strip.background = ggplot2::element_rect(
      fill = "grey95",
      color = "grey75"
    ),
    strip.text = ggplot2::element_text(
      face = "bold"
    )
  )

p_refined_codetection <- patchwork::wrap_plots(
  p_crypt_program_detection,
  p_crypt_program_pairs,
  widths = c(
    1,
    1.25
  )
) +
  patchwork::plot_annotation(
    title = "Refined crypt-state detection and proliferation co-detection",
    subtitle = paste(
      "Counts-layer definitions; descriptive annotation support;",
      "nuclei are not biological replicates"
    )
  )

print(
  p_refined_codetection
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "08_refined_crypt_program_codetection.pdf"
  ),
  plot = p_refined_codetection,
  width = 14,
  height = 7
)
```

## Cross-library consistency of focused marker effects

The two pooled libraries are not biological replicates. Nevertheless, a
cluster-defining marker effect observed in both libraries is less likely to be
caused entirely by one library's technical characteristics.

This audit calculates within-library fold changes for the focused marker panel.
No cross-library treatment P value is calculated.

```{r focused-cross-library-marker-consistency}
sample_ids <- sample_sheet$sample_id[
  sample_sheet$sample_id %in%
    unique(
      as.character(
        multiome$sample_id
      )
    )
]

combined_identity <- paste(
  cluster_vector,
  as.character(multiome$sample_id),
  sep = "__"
)

crypt_identity_order <- unlist(
  lapply(
    crypt_cluster_ids,
    function(cluster_id) {
      paste(
        cluster_id,
        sample_ids,
        sep = "__"
      )
    }
  ),
  use.names = FALSE
)

all_combined_identities <- unique(
  combined_identity
)

multiome$crypt_cluster_library <- factor(
  combined_identity,
  levels = c(
    crypt_identity_order,
    setdiff(
      all_combined_identities,
      crypt_identity_order
    )
  )
)

original_identities <- Idents(
  multiome
)

cross_library_fc_list <- list()
result_index <- 1L

tryCatch(
  {
    Idents(multiome) <- "crypt_cluster_library"

    identity_counts <- table(
      Idents(multiome)
    )

    for (sample_id in sample_ids) {

      for (comparison_index in seq_len(
        nrow(pairwise_comparisons)
      )) {

        ident_1 <- pairwise_comparisons$ident_1[
          comparison_index
        ]

        ident_2 <- pairwise_comparisons$ident_2[
          comparison_index
        ]

        group_1 <- paste(
          ident_1,
          sample_id,
          sep = "__"
        )

        group_2 <- paste(
          ident_2,
          sample_id,
          sep = "__"
        )

        groups_present <- (
          group_1 %in% names(identity_counts) &&
            group_2 %in% names(identity_counts)
        )

        groups_large_enough <- (
          groups_present &&
            identity_counts[[group_1]] >= 3L &&
            identity_counts[[group_2]] >= 3L
        )

        if (!groups_large_enough) {
          warning(
            "Skipping ",
            group_1,
            " versus ",
            group_2,
            ": fewer than three nuclei in at least one group."
          )

          next
        }

        fold_change_result <- FoldChange(
          object = multiome,
          ident.1 = group_1,
          ident.2 = group_2,
          assay = "RNA",
          features = detected_crypt_genes
        )

        effect_column <- get_effect_column(
          fold_change_result
        )

        fold_change_result$gene <- rownames(
          fold_change_result
        )

        fold_change_result$effect_size <-
          fold_change_result[[
            effect_column
          ]]

        fold_change_result$pct_delta <- (
          fold_change_result$pct.1 -
            fold_change_result$pct.2
        )

        fold_change_result$sample_id <- sample_id
        fold_change_result$comparison <- paste0(
          ident_1,
          "_vs_",
          ident_2
        )
        fold_change_result$ident_1 <- ident_1
        fold_change_result$ident_2 <- ident_2

        rownames(fold_change_result) <- NULL

        cross_library_fc_list[[
          result_index
        ]] <- fold_change_result

        result_index <- result_index + 1L
      }
    }
  },
  finally = {
    Idents(multiome) <- original_identities
  }
)

cross_library_fold_changes <- dplyr::bind_rows(
  cross_library_fc_list
)

cross_library_concordance <- cross_library_fold_changes |>
  dplyr::group_by(
    comparison,
    gene
  ) |>
  dplyr::summarise(
    libraries_compared =
      dplyr::n_distinct(sample_id),

    direction_concordant = (
      dplyr::n_distinct(sample_id) ==
        length(sample_ids) &&
        (
          all(effect_size > 0) ||
            all(effect_size < 0)
        )
    ),

    minimum_absolute_effect =
      min(
        abs(effect_size),
        na.rm = TRUE
      ),

    sample_effects = paste(
      paste0(
        sample_id,
        "=",
        round(effect_size, 3)
      ),
      collapse = ";"
    ),

    .groups = "drop"
  )

readr::write_csv(
  cross_library_fold_changes,
  file.path(
    STEP6_MARKER_DIR,
    "focused_crypt_marker_effects_by_library.csv"
  )
)

readr::write_csv(
  cross_library_concordance,
  file.path(
    STEP6_MARKER_DIR,
    "focused_crypt_marker_cross_library_concordance.csv"
  )
)

cross_library_concordance |>
  dplyr::filter(
    direction_concordant
  )
```

```{r focused-crypt-programs-by-library, fig.width=14, fig.height=7}
p_crypt_by_library <- DotPlot(
  multiome,
  features = crypt_plot_programs,
  assay = "RNA",
  idents = crypt_cluster_ids,
  group.by = "crypt_cluster_library"
) +
  RotatedAxis() +
  ggplot2::labs(
    title =
      "Focused crypt-state programs by WNN cluster and pooled library",
    subtitle =
      "Descriptive annotation support; libraries are not biological replicates",
    y = "Cluster__library"
  ) +
  ggplot2::theme(
    strip.text.x = ggplot2::element_text(
      size = 9,
      face = "bold"
    )
  )

print(
  p_crypt_by_library
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "08_focused_crypt_programs_by_library.pdf"
  ),
  plot = p_crypt_by_library,
  width = 14,
  height = 7
)

readr::write_csv(
  p_crypt_by_library$data,
  file.path(
    STEP6_DIR,
    "focused_crypt_programs_by_library_DotPlot_data.csv"
  )
)
```

## QC and residual-doublet audit for the focused clusters

All nuclei entering Step 6 passed the reviewed joint QC and scDblFinder
decisions in Step 2. This is not a second round of filtering.

The purpose is to determine whether cluster 18's mixed proliferative and
secretory program coincides with unusually high RNA/ATAC complexity or residual
doublet evidence relative to clusters 4 and 7.

```{r focused-crypt-qc-doublet-audit}
focused_qc_metadata <- multiome[[]] |>
  dplyr::mutate(
    cluster = as.character(
      .data[[CHOSEN_CLUSTER_COLUMN]]
    )
  ) |>
  dplyr::filter(
    cluster %in% crypt_cluster_ids
  )

qc_numeric_candidates <- c(
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "nCount_ATAC",
  "nCount_ATAC_common",
  "TSS.enrichment",
  "nucleosome_signal",
  "frip_cr",
  "scDblFinder.score",
  "RNA.weight",
  "ATAC.weight"
)

qc_numeric_metrics <- intersect(
  qc_numeric_candidates,
  colnames(focused_qc_metadata)
)

qc_numeric_metrics <- qc_numeric_metrics[
  vapply(
    focused_qc_metadata[
      qc_numeric_metrics
    ],
    is.numeric,
    logical(1)
  )
]

focused_qc_summary <- focused_qc_metadata |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    dplyr::across(
      dplyr::all_of(
        qc_numeric_metrics
      ),
      list(
        q25 = ~ stats::quantile(
          .x,
          0.25,
          na.rm = TRUE
        ),
        median = ~ stats::median(
          .x,
          na.rm = TRUE
        ),
        q75 = ~ stats::quantile(
          .x,
          0.75,
          na.rm = TRUE
        )
      ),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"
  )

if (
  "scDblFinder.class" %in%
    colnames(focused_qc_metadata)
) {
  focused_doublet_summary <- focused_qc_metadata |>
    dplyr::count(
      cluster,
      scDblFinder.class,
      name = "nuclei"
    ) |>
    dplyr::group_by(
      cluster
    ) |>
    dplyr::mutate(
      percent_of_cluster =
        100 * nuclei / sum(nuclei)
    ) |>
    dplyr::ungroup()
} else {
  focused_doublet_summary <- data.frame(
    note =
      "scDblFinder.class was absent from the Step 5 object."
  )
}

readr::write_csv(
  focused_qc_summary,
  file.path(
    STEP6_DIR,
    "focused_crypt_cluster_QC_summary.csv"
  )
)

readr::write_csv(
  focused_doublet_summary,
  file.path(
    STEP6_DIR,
    "focused_crypt_cluster_doublet_summary.csv"
  )
)

focused_qc_summary
focused_doublet_summary
```

```{r plot-focused-crypt-qc, fig.width=12, fig.height=9}
focused_qc_long <- focused_qc_metadata |>
  dplyr::select(
    cluster,
    dplyr::all_of(
      qc_numeric_metrics
    )
  ) |>
  tidyr::pivot_longer(
    cols = -cluster,
    names_to = "metric",
    values_to = "value"
  )

p_focused_qc <- ggplot2::ggplot(
  focused_qc_long,
  ggplot2::aes(
    x = cluster,
    y = value,
    fill = cluster
  )
) +
  ggplot2::geom_violin(
    scale = "width",
    trim = TRUE
  ) +
  ggplot2::geom_boxplot(
    width = 0.15,
    outlier.shape = NA
  ) +
  ggplot2::facet_wrap(
    ~ metric,
    scales = "free_y",
    ncol = 3
  ) +
  ggplot2::labs(
    title =
      "QC and modality characteristics of clusters 4, 7, and 18",
    x = "WNN cluster",
    y = NULL
  ) +
  ggplot2::theme_classic(
    base_size = 10
  ) +
  ggplot2::theme(
    legend.position = "none"
  )

print(
  p_focused_qc
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "09_focused_crypt_cluster_QC_audit.pdf"
  ),
  plot = p_focused_qc,
  width = 12,
  height = 9
)
```


# 12. Multimodal review of focused crypt states

RNA marker programs define the initial biological hypotheses, but the paired
ATAC assay should contribute before final labels are assigned. The following
plots compare the same nuclei across the saved RNA-only, ATAC-only, and WNN
embeddings; map selected RNA markers onto the WNN geometry; score cell-cycle
phase without regressing it out; display the strongest focused RNA
discriminants; and inspect accessibility at selected ISC and secretory loci.

These analyses remain descriptive. The two pooled libraries are not biological
replicates, and neither nuclei nor peaks are treated as replicated treatment
units.

## Compare focused clusters across RNA, ATAC, and WNN embeddings

Step 4 saved modality-specific UMAPs as `umap.rna` and `umap.atac`; Step 5
saved the reviewed multimodal embedding as `wnn.umap`. Reusing those coordinates
avoids recomputing neighborhoods or changing the reviewed cluster solution.

```{r focused-crypt-multimodal-embeddings, fig.width=16, fig.height=5.5}
required_review_reductions <- c(
  "umap.rna",
  "umap.atac",
  "wnn.umap"
)

missing_review_reductions <- setdiff(
  required_review_reductions,
  names(multiome@reductions)
)

if (length(missing_review_reductions) > 0L) {
  stop(
    "The finalized Step 5 object is missing review reduction(s): ",
    paste(missing_review_reductions, collapse = ", "),
    ". Revisit the Step 4 checkpoint rather than silently recomputing them here.",
    call. = FALSE
  )
}

crypt_cells_by_cluster <- lapply(
  CRYPT_REVIEW_CLUSTERS,
  function(cluster_id) {
    WhichCells(
      multiome,
      idents = cluster_id
    )
  }
)

names(crypt_cells_by_cluster) <- paste(
  "Cluster",
  CRYPT_REVIEW_CLUSTERS
)

plot_focused_reduction <- function(
  reduction_name,
  plot_title
) {
  DimPlot(
    multiome,
    reduction = reduction_name,
    cells.highlight = crypt_cells_by_cluster,
    cols.highlight = unname(
      CRYPT_CLUSTER_COLORS[
        CRYPT_REVIEW_CLUSTERS
      ]
    ),
    sizes.highlight = 0.55,
    pt.size = 0.12,
    raster = TRUE,
    shuffle = FALSE,
    na.value = "grey88"
  ) +
    ggplot2::ggtitle(
      plot_title
    ) +
    Seurat::NoAxes() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold",
        hjust = 0.5
      )
    )
}

p_crypt_rna_umap <- plot_focused_reduction(
  "umap.rna",
  "RNA-only UMAP"
)

p_crypt_atac_umap <- plot_focused_reduction(
  "umap.atac",
  "ATAC-only UMAP"
)

p_crypt_wnn_umap <- plot_focused_reduction(
  "wnn.umap",
  "WNN UMAP"
)

p_crypt_multimodal <- patchwork::wrap_plots(
  p_crypt_rna_umap,
  p_crypt_atac_umap,
  p_crypt_wnn_umap,
  ncol = 3,
  guides = "collect"
) +
  patchwork::plot_annotation(
    title = "Clusters 4, 7, and 18 across modality-specific and joint embeddings",
    subtitle = "All other nuclei are shown in grey; coordinates are inherited from Steps 4 and 5"
  ) &
  ggplot2::theme(
    legend.position = "bottom"
  )

print(
  p_crypt_multimodal
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "10_focused_crypt_RNA_ATAC_WNN_embeddings.pdf"
  ),
  plot = p_crypt_multimodal,
  width = 16,
  height = 5.5
)

focused_embedding_data <- dplyr::bind_rows(
  lapply(
    required_review_reductions,
    function(reduction_name) {
      embedding <- Embeddings(
        multiome,
        reduction = reduction_name
      )[
        crypt_review_cells,
        1:2,
        drop = FALSE
      ]

      data.frame(
        nucleus = rownames(embedding),
        reduction = reduction_name,
        dimension_1 = embedding[
          ,
          1
        ],
        dimension_2 = embedding[
          ,
          2
        ],
        cluster = as.character(
          multiome[[]][
            rownames(embedding),
            CHOSEN_CLUSTER_COLUMN
          ]
        ),
        sample_id = as.character(
          multiome[[]][
            rownames(embedding),
            "sample_id"
          ]
        ),
        stringsAsFactors = FALSE
      )
    }
  )
)

readr::write_csv(
  focused_embedding_data,
  file.path(
    STEP6_DIR,
    "focused_crypt_multimodal_embedding_coordinates.csv"
  )
)
```

## Map focused crypt markers onto the WNN embedding

The WNN feature maps ask whether the marker signals occupy coherent portions of
the multimodal geometry. Each gene uses its own color scale so that sparse but
spatially coherent markers are not hidden by highly expressed genes.

```{r focused-crypt-wnn-featureplots, fig.width=14, fig.height=8}
crypt_featureplot_genes <- c(
  "Lgr5",
  "Smoc2",
  "Mki67",
  "Top2a",
  "Atoh1",
  "Muc2",
  "Reg4",
  "Kit"
)

missing_featureplot_genes <- setdiff(
  crypt_featureplot_genes,
  rownames(multiome[["RNA"]])
)

if (length(missing_featureplot_genes) > 0L) {
  stop(
    "WNN FeaturePlot gene(s) are absent from the RNA feature set: ",
    paste(missing_featureplot_genes, collapse = ", "),
    call. = FALSE
  )
}

DefaultAssay(multiome) <- "RNA"

p_crypt_features_wnn <- FeaturePlot(
  multiome,
  features = crypt_featureplot_genes,
  reduction = "wnn.umap",
  slot = "data",
  order = TRUE,
  min.cutoff = "q05",
  max.cutoff = "q95",
  keep.scale = "feature",
  cols = c(
    "grey94",
    "#2C7FB8",
    "#D7301F"
  ),
  pt.size = 0.18,
  raster = FALSE,
  ncol = 4
) +
  patchwork::plot_annotation(
    title = "Focused crypt-state RNA markers on the WNN embedding",
    subtitle = paste(
      "Feature-specific scales;",
      "spatial pattern is emphasized over cross-gene magnitude"
    )
  )

print(
  p_crypt_features_wnn
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "11_focused_crypt_markers_on_WNN_UMAP.pdf"
  ),
  plot = p_crypt_features_wnn,
  width = 14,
  height = 8
)

rna_data <- SeuratObject::LayerData(
  multiome,
  assay = "RNA",
  layer = "data"
)

focused_wnn_embedding <- Embeddings(
  multiome,
  reduction = "wnn.umap"
)[
  crypt_review_cells,
  1:2,
  drop = FALSE
]

focused_featureplot_data <- data.frame(
  nucleus = crypt_review_cells,
  WNN_1 = focused_wnn_embedding[
    crypt_review_cells,
    1
  ],
  WNN_2 = focused_wnn_embedding[
    crypt_review_cells,
    2
  ],
  cluster = as.character(
    multiome[[]][
      crypt_review_cells,
      CHOSEN_CLUSTER_COLUMN
    ]
  ),
  sample_id = as.character(
    multiome[[]][
      crypt_review_cells,
      "sample_id"
    ]
  ),
  t(
    as.matrix(
      rna_data[
        crypt_featureplot_genes,
        crypt_review_cells,
        drop = FALSE
      ]
    )
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

readr::write_csv(
  focused_featureplot_data,
  file.path(
    STEP6_DIR,
    "focused_crypt_WNN_FeaturePlot_data.csv"
  )
)
```

## Score cell-cycle phase without regressing it out

Cell-cycle scoring is used here to describe whether proliferative nuclei are
more S-like or G2M-like. The scores are not regressed from the data because
proliferation is a biological state that helps distinguish these crypt
populations. A phase assignment for a non-proliferative nucleus should not be
interpreted as proof that the nucleus was actively cycling.

```{r focused-crypt-cell-cycle-scoring, fig.width=12, fig.height=9}
DefaultAssay(multiome) <- "RNA"

mouse_s_features <- unique(
  unname(
    Seurat::CaseMatch(
      search = cc.genes.updated.2019$s.genes,
      match = rownames(multiome[["RNA"]])
    )
  )
)

mouse_g2m_features <- unique(
  unname(
    Seurat::CaseMatch(
      search = cc.genes.updated.2019$g2m.genes,
      match = rownames(multiome[["RNA"]])
    )
  )
)

if (
  length(mouse_s_features) < 10L ||
    length(mouse_g2m_features) < 10L
) {
  stop(
    "Too few cell-cycle genes matched the mouse RNA feature symbols: ",
    length(mouse_s_features),
    " S-phase and ",
    length(mouse_g2m_features),
    " G2M-phase genes.",
    call. = FALSE
  )
}

cell_cycle_feature_audit <- dplyr::bind_rows(
  data.frame(
    phase_program = "S",
    gene = mouse_s_features,
    stringsAsFactors = FALSE
  ),
  data.frame(
    phase_program = "G2M",
    gene = mouse_g2m_features,
    stringsAsFactors = FALSE
  )
)

readr::write_csv(
  cell_cycle_feature_audit,
  file.path(
    STEP6_DIR,
    "cell_cycle_features_matched_to_mouse_RNA.csv"
  )
)

cycle_scored <- Seurat::CellCycleScoring(
  object = multiome,
  s.features = mouse_s_features,
  g2m.features = mouse_g2m_features,
  set.ident = FALSE,
  assay = "RNA"
)

multiome <- AddMetaData(
  object = multiome,
  metadata = cycle_scored[[]][
    ,
    c(
      "S.Score",
      "G2M.Score",
      "Phase"
    ),
    drop = FALSE
  ]
)

rm(cycle_scored)
invisible(gc())

focused_cycle_data <- data.frame(
  nucleus = crypt_review_cells,
  cluster = factor(
    as.character(
      multiome[[]][
        crypt_review_cells,
        CHOSEN_CLUSTER_COLUMN
      ]
    ),
    levels = CRYPT_REVIEW_CLUSTERS
  ),
  sample_id = as.character(
    multiome[[]][
      crypt_review_cells,
      "sample_id"
    ]
  ),
  S.Score = multiome[[]][
    crypt_review_cells,
    "S.Score"
  ],
  G2M.Score = multiome[[]][
    crypt_review_cells,
    "G2M.Score"
  ],
  Phase = factor(
    multiome[[]][
      crypt_review_cells,
      "Phase"
    ],
    levels = c(
      "G1",
      "S",
      "G2M"
    )
  ),
  stringsAsFactors = FALSE
)

proliferative_index <- match(
  focused_cycle_data$nucleus,
  per_cell_crypt_program_detection$nucleus
)

if (anyNA(proliferative_index)) {
  stop(
    "Cell-cycle nuclei could not all be matched to the refined co-detection table.",
    call. = FALSE
  )
}

focused_cycle_data$Proliferative <-
  per_cell_crypt_program_detection$Proliferative[
    proliferative_index
  ]

focused_cycle_scope <- dplyr::bind_rows(
  focused_cycle_data |>
    dplyr::mutate(
      review_scope = "All focused crypt nuclei"
    ),
  focused_cycle_data |>
    dplyr::filter(
      Proliferative
    ) |>
    dplyr::mutate(
      review_scope = "Mki67 or Top2a detected"
    )
)

focused_cycle_phase_summary <-
  focused_cycle_scope |>
  dplyr::count(
    review_scope,
    cluster,
    Phase,
    name = "nuclei"
  ) |>
  dplyr::group_by(
    review_scope,
    cluster
  ) |>
  dplyr::mutate(
    total_nuclei = sum(nuclei),
    percent_of_scope = 100 * nuclei / total_nuclei
  ) |>
  dplyr::ungroup()

focused_cycle_phase_summary

readr::write_csv(
  focused_cycle_data,
  file.path(
    STEP6_DIR,
    "focused_crypt_cell_cycle_scores.csv"
  )
)

readr::write_csv(
  focused_cycle_phase_summary,
  file.path(
    STEP6_DIR,
    "focused_crypt_cell_cycle_phase_summary.csv"
  )
)

cell_cycle_colors <- c(
  "G1" = "grey70",
  "S" = "#56B4E9",
  "G2M" = "#D55E00"
)

p_cycle_scores <- ggplot2::ggplot(
  focused_cycle_data,
  ggplot2::aes(
    x = S.Score,
    y = G2M.Score
  )
) +
  ggplot2::geom_hline(
    yintercept = 0,
    color = "grey85",
    linewidth = 0.35
  ) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey85",
    linewidth = 0.35
  ) +
  ggplot2::geom_point(
    color = "grey88",
    size = 0.65,
    alpha = 0.45
  ) +
  ggplot2::geom_point(
    data = dplyr::filter(
      focused_cycle_data,
      Proliferative
    ),
    ggplot2::aes(
      color = Phase
    ),
    size = 0.9,
    alpha = 0.75
  ) +
  ggplot2::facet_wrap(
    ~cluster,
    nrow = 1
  ) +
  ggplot2::scale_color_manual(
    values = cell_cycle_colors,
    drop = FALSE
  ) +
  ggplot2::labs(
    title = "Cell-cycle scores",
    subtitle = "Colored nuclei have Mki67 or Top2a counts; grey nuclei provide cluster context",
    x = "S-phase score",
    y = "G2M-phase score",
    color = "Assigned phase"
  ) +
  ggplot2::theme_classic(
    base_size = 11
  )

p_cycle_phase <- ggplot2::ggplot(
  focused_cycle_phase_summary,
  ggplot2::aes(
    x = cluster,
    y = percent_of_scope,
    fill = Phase
  )
) +
  ggplot2::geom_col(
    width = 0.72
  ) +
  ggplot2::facet_wrap(
    ~review_scope,
    nrow = 1
  ) +
  ggplot2::scale_fill_manual(
    values = cell_cycle_colors,
    drop = FALSE
  ) +
  ggplot2::scale_y_continuous(
    limits = c(
      0,
      100
    ),
    expand = ggplot2::expansion(
      mult = c(
        0,
        0.02
      )
    )
  ) +
  ggplot2::labs(
    title = "Phase composition",
    x = "WNN cluster",
    y = "Nuclei in review scope (%)",
    fill = "Assigned phase"
  ) +
  ggplot2::theme_classic(
    base_size = 11
  )

p_cell_cycle_audit <- patchwork::wrap_plots(
  p_cycle_scores,
  p_cycle_phase,
  ncol = 1,
  heights = c(
    1.15,
    1
  ),
  guides = "collect"
) +
  patchwork::plot_annotation(
    title = "Cell-cycle phase audit of clusters 4, 7, and 18",
    subtitle = "Scores are descriptive and are not regressed from the annotation object"
  ) &
  ggplot2::theme(
    legend.position = "bottom"
  )

print(
  p_cell_cycle_audit
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "12_focused_crypt_cell_cycle_phase_audit.pdf"
  ),
  plot = p_cell_cycle_audit,
  width = 12,
  height = 9
)
```

## Display the strongest focused RNA discriminants

The heatmap uses a new one-versus-rest marker calculation restricted to
clusters 4, 7, and 18. Marker discovery uses every focused nucleus; only the
display is downsampled. The cell selection is deterministic and exported so
the visual can be reproduced exactly.

This is an annotation visualization, not a treatment differential-expression
analysis.

```{r focused-crypt-discriminant-heatmap, fig.width=11, fig.height=8, cache=FALSE}
focused_crypt_object <- subset(
  multiome,
  idents = CRYPT_REVIEW_CLUSTERS
)

DefaultAssay(focused_crypt_object) <- "RNA"

Idents(focused_crypt_object) <- factor(
  as.character(
    focused_crypt_object[[]][[
      CHOSEN_CLUSTER_COLUMN
    ]]
  ),
  levels = CRYPT_REVIEW_CLUSTERS
)

future::plan(
  future::sequential
)

set.seed(20260820)

focused_crypt_markers <- FindAllMarkers(
  object = focused_crypt_object,
  assay = "RNA",
  only.pos = TRUE,
  min.pct = 0.10,
  logfc.threshold = 0.50,
  test.use = "wilcox",
  return.thresh = 1,
  verbose = TRUE
)

configure_parallel_plan()

if (nrow(focused_crypt_markers) == 0L) {
  stop(
    "Focused FindAllMarkers returned no RNA markers.",
    call. = FALSE
  )
}

focused_fc_column <- intersect(
  c(
    "avg_log2FC",
    "avg_logFC"
  ),
  colnames(focused_crypt_markers)
)

if (length(focused_fc_column) != 1L) {
  stop(
    "Could not identify exactly one average log-fold-change column in focused markers.",
    call. = FALSE
  )
}

focused_crypt_markers$effect_size <-
  focused_crypt_markers[[
    focused_fc_column
  ]]

focused_crypt_markers$pct_delta <- with(
  focused_crypt_markers,
  pct.1 - pct.2
)

presentation_exclusion_pattern <- paste0(
  "^(mt-|Rpl|Rps|Gm[0-9]+|Mir[0-9]+|Snor)",
  "|Rik$"
)

focused_heatmap_gene_table <-
  focused_crypt_markers |>
  dplyr::filter(
    p_val_adj < 0.05,
    effect_size >= 0.50,
    pct_delta >= 0.10,
    !grepl(
      presentation_exclusion_pattern,
      gene
    ),
    !gene %in% c(
      "Malat1",
      "Xist",
      "Tsix"
    )
  ) |>
  dplyr::group_by(
    cluster
  ) |>
  dplyr::arrange(
    dplyr::desc(pct_delta),
    dplyr::desc(effect_size),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(
    n = 8
  ) |>
  dplyr::ungroup()

selected_genes_per_cluster <- table(
  focused_heatmap_gene_table$cluster
)

if (
  length(selected_genes_per_cluster) != length(CRYPT_REVIEW_CLUSTERS) ||
    any(selected_genes_per_cluster < 4L)
) {
  stop(
    "Fewer than four presentation-quality discriminants were selected for one or more focused clusters.",
    call. = FALSE
  )
}

heatmap_genes <- unique(
  focused_heatmap_gene_table$gene
)

readr::write_csv(
  focused_crypt_markers,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_focused_crypt_clusters_4_7_18_one_vs_rest.csv"
  )
)

readr::write_csv(
  focused_heatmap_gene_table,
  file.path(
    STEP6_MARKER_DIR,
    "RNA_focused_crypt_heatmap_genes.csv"
  )
)

set.seed(20260820)

heatmap_cells <- unlist(
  lapply(
    CRYPT_REVIEW_CLUSTERS,
    function(cluster_id) {
      cluster_cells <- WhichCells(
        focused_crypt_object,
        idents = cluster_id
      )

      sample(
        cluster_cells,
        size = min(
          100L,
          length(cluster_cells)
        ),
        replace = FALSE
      )
    }
  ),
  use.names = FALSE
)

heatmap_cell_table <- data.frame(
  display_order = seq_along(heatmap_cells),
  nucleus = heatmap_cells,
  cluster = as.character(
    Idents(focused_crypt_object)[
      heatmap_cells
    ]
  ),
  sample_id = as.character(
    focused_crypt_object[[]][
      heatmap_cells,
      "sample_id"
    ]
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  heatmap_cell_table,
  file.path(
    STEP6_DIR,
    "focused_crypt_heatmap_displayed_nuclei.csv"
  )
)

focused_crypt_object <- ScaleData(
  object = focused_crypt_object,
  assay = "RNA",
  features = heatmap_genes,
  verbose = FALSE
)

p_focused_crypt_heatmap <- suppressMessages(
  DoHeatmap(
    object = focused_crypt_object,
    features = heatmap_genes,
    cells = heatmap_cells,
    assay = "RNA",
    slot = "scale.data",
    group.by = CHOSEN_CLUSTER_COLUMN,
    group.colors = unname(
      CRYPT_CLUSTER_COLORS[
        CRYPT_REVIEW_CLUSTERS
      ]
    ),
    disp.min = -2,
    disp.max = 2,
    raster = FALSE,
    draw.lines = TRUE,
    angle = 0,
    hjust = 0.5,
    size = 4.5
  ) +
    ggplot2::scale_fill_gradientn(
      colors = c(
        "#2166AC",
        "#F7F7F7",
        "#B2182B"
      ),
      limits = c(
        -2,
        2
      ),
      oob = scales::squish
    ) +
    ggplot2::labs(
      title = "Strongest focused RNA discriminants",
      subtitle = paste(
        "Top eight genes per cluster; scaled across all focused nuclei;",
        "up to 100 nuclei per cluster displayed"
      )
    ) +
    ggplot2::theme(
      axis.text.y = ggplot2::element_text(
        face = "italic",
        size = 8
      )
    )
)

print(
  p_focused_crypt_heatmap
)

ggplot2::ggsave(
  filename = file.path(
    STEP6_PLOT_DIR,
    "13_focused_crypt_discriminant_heatmap.pdf"
  ),
  plot = p_focused_crypt_heatmap,
  width = 11,
  height = 8
)
```

## Add targeted ATAC support at ISC and secretory loci

Coverage tracks aggregate Tn5 insertions within each cluster-library group.
They are used to ask whether accessibility at selected loci is compatible with
the RNA-supported state assignments. Accessibility is not treated as a proxy
for expression, and a locus track alone is not sufficient for a final label.

The library split is retained because the two pooled libraries are descriptive
annotation-support strata, not biological replicates. Cluster 18 has only 39
HFD nuclei, so sparse differences in that track should be interpreted
cautiously.

```{r targeted-crypt-atac-coverage, fig.width=13, fig.height=9, cache=FALSE}
targeted_crypt_loci <- c(
  "Lgr5",
  "Ascl2",
  "Atoh1",
  "Muc2",
  "Reg4",
  "Kit"
)

if (
  length(
    Signac::Fragments(
      multiome[["ATAC"]]
    )
  ) == 0L
) {
  stop(
    "The ATAC assay has no Fragment objects for CoveragePlot().",
    call. = FALSE
  )
}

atac_annotation <- Signac::Annotation(
  multiome[["ATAC"]]
)

if (length(atac_annotation) == 0L) {
  stop(
    "The ATAC assay has no gene annotation for gene-name coverage plots.",
    call. = FALSE
  )
}

annotation_columns <- colnames(
  S4Vectors::mcols(
    atac_annotation
  )
)

if (!"gene_name" %in% annotation_columns) {
  stop(
    "The ATAC annotation does not contain a gene_name metadata column.",
    call. = FALSE
  )
}

missing_annotation_loci <- setdiff(
  targeted_crypt_loci,
  unique(
    as.character(
      S4Vectors::mcols(
        atac_annotation
      )$gene_name
    )
  )
)

if (length(missing_annotation_loci) > 0L) {
  stop(
    "Targeted locus/loci are absent from the ATAC gene annotation: ",
    paste(missing_annotation_loci, collapse = ", "),
    call. = FALSE
  )
}

coverage_object <- subset(
  multiome,
  idents = CRYPT_REVIEW_CLUSTERS
)

coverage_object$focused_crypt_cluster <- factor(
  as.character(
    coverage_object[[]][[
      CHOSEN_CLUSTER_COLUMN
    ]]
  ),
  levels = CRYPT_REVIEW_CLUSTERS
)

coverage_group_sizes <-
  coverage_object[[]] |>
  dplyr::mutate(
    nucleus = rownames(coverage_object[[]]),
    cluster = as.character(focused_crypt_cluster),
    sample_id = as.character(sample_id)
  ) |>
  dplyr::count(
    cluster,
    sample_id,
    name = "nuclei"
  ) |>
  dplyr::arrange(
    factor(
      cluster,
      levels = CRYPT_REVIEW_CLUSTERS
    ),
    sample_id
  ) |>
  dplyr::mutate(
    coverage_group = paste0(
      "Cluster ",
      cluster,
      " | ",
      sample_id,
      " (n=",
      nuclei,
      ")"
    )
  )

coverage_group_key <- paste(
  as.character(
    coverage_object$focused_crypt_cluster
  ),
  as.character(
    coverage_object$sample_id
  ),
  sep = "__"
)

coverage_group_lookup <- stats::setNames(
  coverage_group_sizes$coverage_group,
  paste(
    coverage_group_sizes$cluster,
    coverage_group_sizes$sample_id,
    sep = "__"
  )
)

coverage_object$coverage_group <- factor(
  unname(
    coverage_group_lookup[
      coverage_group_key
    ]
  ),
  levels = coverage_group_sizes$coverage_group
)

if (anyNA(coverage_object$coverage_group)) {
  stop(
    "One or more nuclei could not be assigned to an ATAC coverage group.",
    call. = FALSE
  )
}

readr::write_csv(
  coverage_group_sizes,
  file.path(
    STEP6_DIR,
    "targeted_crypt_ATAC_coverage_group_sizes.csv"
  )
)

DefaultAssay(coverage_object) <- "ATAC"

targeted_crypt_coverage_plots <- lapply(
  targeted_crypt_loci,
  function(locus) {
    Signac::CoveragePlot(
      object = coverage_object,
      region = locus,
      features = locus,
      assay = "ATAC",
      expression.assay = "RNA",
      expression.slot = "data",
      annotation = "gene",
      peaks = TRUE,
      links = FALSE,
      tile = FALSE,
      group.by = "coverage_group",
      extend.upstream = 5000,
      extend.downstream = 5000,
      ymax = "q95"
    ) +
      patchwork::plot_annotation(
        title = paste0(
          locus,
          " locus: ATAC coverage with RNA expression"
        ),
        subtitle = paste(
          "Clusters 4, 7, and 18 split by pooled library;",
          "coverage tracks are normalized by group depth and cell number"
        )
      )
  }
)

names(targeted_crypt_coverage_plots) <- targeted_crypt_loci

coverage_pdf_path <- file.path(
  STEP6_PLOT_DIR,
  "14_targeted_crypt_ATAC_coverage.pdf"
)

grDevices::pdf(
  file = coverage_pdf_path,
  width = 13,
  height = 9,
  onefile = TRUE
)

for (locus in targeted_crypt_loci) {
  print(
    targeted_crypt_coverage_plots[[
      locus
    ]]
  )
}

grDevices::dev.off()

for (locus in targeted_crypt_loci) {
  print(
    targeted_crypt_coverage_plots[[
      locus
    ]]
  )
}

DefaultAssay(multiome) <- "RNA"
```

## Focused crypt-state interpretation checkpoint

Before assigning final labels to clusters 4, 7, and 18, review whether:

1. the clusters remain coherent or separable in RNA-only, ATAC-only, and WNN
   embeddings;
2. selected RNA markers occupy coherent portions of the WNN geometry;
3. cluster 4 and cluster 18 differ in cell-cycle phase composition rather than
   merely in overall proliferation prevalence;
4. the downsampled heatmap separates each cluster with interpretable
   discriminants;
5. ISC-associated loci support clusters 4/7 and secretory-associated loci
   support cluster 18 in both pooled-library strata;
6. any conclusion remains descriptive rather than a CON-versus-HFD treatment
   claim.
