---
title: "Step 7 - Characterize within-cohort diet response"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 6 assigned reviewed biological labels while preserving the global WNN
clusters and any approved focused annotation units. Step 7 asks where CON and
HFD differ within one genotype cohort.

This notebook deliberately separates three questions:

1. Does the representation of an annotated population differ between the two
   pooled libraries?
2. Within a matched population, which RNA-expression and chromatin-accessibility
   effect sizes are most different between HFD and CON?
3. Are the strongest RNA and ATAC changes directionally concordant near the
   same genes?

The notebook is cohort-parameterized and can later be run without changing its
logic for WT, PPAR, and IL17. The first benchmark analysis should use WT.

## Statistical limitation

Each cohort currently contains one pooled CON library and one pooled HFD
library. Mouse identities are not recoverable after pooling, so diet and
library are perfectly confounded. Nuclei are observations, not independent
biological replicates.

Accordingly, this notebook reports:

- descriptive composition;
- pseudobulk log-fold changes between the two pooled libraries;
- within-cluster detection differences; and
- repeated balanced downsampling as a cell-sampling stability analysis.

It does **not** report barcode-level P values as evidence of a replicated diet
effect. Downsampling cannot create biological replication.

A practical first-pass allocation is **4 CPU cores, 96--128 GiB memory, and
12--24 hours**. The full ATAC effect table and repeated downsampling are the
most memory- and time-intensive sections.

# 1. Setup

Set the active dataset and cohort before running the notebook. For the WT
benchmark:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
  library(Matrix)
  library(patchwork)
})

ggplot2::theme_set(theme_multiome())

sample_sheet <- read_sample_sheet()

set.seed(20260904)
future::plan(future::sequential)
```

# 2. Define inputs, outputs, and analysis parameters

The preferred analysis unit is `annotation_unit`. Older Step 6 checkpoints
created before focused annotation units were introduced may not contain that
column. When no approved focused-subclustering decision exists, the notebook
safely falls back to the equivalent global `wnn_cluster`. For IL17 and any
future cohort with an approved focused refinement, `annotation_unit` remains
mandatory so that a child such as `11_tuft` cannot be silently collapsed.

Tier A requires at least 200 nuclei from each diet. Tier B requires at least
100. Tier C remains available for composition and exploratory visualization,
but is not promoted into the primary RNA/ATAC effect-size tables.

```{r define-inputs-and-parameters}
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP7_DIR <- file.path(OUTPUT_DIR, "07_diet_response")
STEP7_TABLE_DIR <- file.path(STEP7_DIR, "tables")
STEP7_PLOT_DIR <- file.path(STEP7_DIR, "plots")

STEP6_OBJECT_PATH <- file.path(
  STEP6_DIR,
  "multiome_annotated.rds"
)

STEP6_COMPLETE_MARKER <- file.path(
  STEP6_DIR,
  "STEP_06_COMPLETE.txt"
)

STEP7_COMPLETE_MARKER <- file.path(
  STEP7_DIR,
  "STEP_07_COMPLETE.txt"
)

dir.create(STEP7_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP7_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP7_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

assert_file(STEP6_OBJECT_PATH, "Step 6 annotated multimodal object")
assert_file(STEP6_COMPLETE_MARKER, "Step 6 completion marker")

PREFERRED_ANALYSIS_UNIT_COLUMN <- "annotation_unit"
LEGACY_ANALYSIS_UNIT_COLUMN <- "wnn_cluster"
REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"

TIER_A_MIN_NUCLEI_PER_DIET <- 200L
TIER_B_MIN_NUCLEI_PER_DIET <- 100L

RNA_MIN_TOTAL_COUNTS <- 20L
RNA_MIN_DETECTION_PERCENT <- 5
RNA_EFFECT_LOG2FC_THRESHOLD <- 0.50
RNA_EFFECT_DETECTION_DELTA_THRESHOLD <- 5
RNA_TOP_FEATURES_PER_DIRECTION <- 250L

ATAC_MIN_TOTAL_COUNTS <- 30L
ATAC_MIN_DETECTION_PERCENT <- 1
ATAC_EFFECT_LOG2FC_THRESHOLD <- 0.50
ATAC_TOP_FEATURES_PER_DIRECTION <- 500L

CPM_PSEUDOCOUNT <- 0.10

RUN_BALANCED_DOWNSAMPLING <- TRUE
DOWNSAMPLING_ITERATIONS <- 25L
DOWNSAMPLING_MAX_CELLS_PER_DIET <- 500L
DOWNSAMPLING_MAX_FEATURES_PER_UNIT <- 1000L

REQUEST_MOTIF_ENRICHMENT <- FALSE

regional_audit_programs <- list(
  candidate_proximal_crypt = c(
    "Cftr", "Htr4", "Slc12a2", "Kcnq1", "Dmbt1"
  ),
  candidate_mid_distal_secretory = c(
    "Fcgbp", "Clca1", "Zg16", "Bcas1", "Tff3"
  ),
  mature_surface_absorptive = c(
    "Car4", "Aqp8", "Slc26a3", "Krt20", "Tgm3"
  )
)

analysis_parameters <- data.frame(
  parameter = c(
    "preferred analysis unit column",
    "reference diet",
    "comparison diet",
    "Tier A minimum nuclei per diet",
    "Tier B minimum nuclei per diet",
    "RNA minimum total counts",
    "RNA minimum detection percent",
    "RNA candidate absolute log2 fold change",
    "ATAC minimum total counts",
    "ATAC minimum detection percent",
    "ATAC candidate absolute log2 fold change",
    "balanced downsampling performed",
    "balanced downsampling iterations",
    "maximum sampled cells per diet",
    "motif enrichment requested"
  ),
  value = c(
    PREFERRED_ANALYSIS_UNIT_COLUMN,
    REFERENCE_DIET,
    COMPARISON_DIET,
    TIER_A_MIN_NUCLEI_PER_DIET,
    TIER_B_MIN_NUCLEI_PER_DIET,
    RNA_MIN_TOTAL_COUNTS,
    RNA_MIN_DETECTION_PERCENT,
    RNA_EFFECT_LOG2FC_THRESHOLD,
    ATAC_MIN_TOTAL_COUNTS,
    ATAC_MIN_DETECTION_PERCENT,
    ATAC_EFFECT_LOG2FC_THRESHOLD,
    RUN_BALANCED_DOWNSAMPLING,
    DOWNSAMPLING_ITERATIONS,
    DOWNSAMPLING_MAX_CELLS_PER_DIET,
    REQUEST_MOTIF_ENRICHMENT
  ),
  stringsAsFactors = FALSE
)

```

# 3. Load and validate the annotated object

```{r load-and-validate-object}
multiome <- readRDS(STEP6_OBJECT_PATH)
metadata <- multiome[[]]

resolve_analysis_unit_column <- function(metadata) {
  if (PREFERRED_ANALYSIS_UNIT_COLUMN %in% colnames(metadata)) {
    return(PREFERRED_ANALYSIS_UNIT_COLUMN)
  }

  if (!LEGACY_ANALYSIS_UNIT_COLUMN %in% colnames(metadata)) {
    stop(
      "The Step 6 object contains neither '",
      PREFERRED_ANALYSIS_UNIT_COLUMN,
      "' nor '",
      LEGACY_ANALYSIS_UNIT_COLUMN,
      "'.",
      call. = FALSE
    )
  }

  focused_decision_path <- file.path(
    COHORT_CONFIG_DIR,
    "subclustering_decision.csv"
  )

  if (file.exists(focused_decision_path)) {
    focused_decision <- suppressMessages(
      readr::read_csv(
        focused_decision_path,
        show_col_types = FALSE,
        progress = FALSE
      )
    )

    if (!"approved" %in% colnames(focused_decision)) {
      stop(
        "The focused-subclustering decision table has no approved column: ",
        focused_decision_path,
        call. = FALSE
      )
    }

    approved_focused_decision <- tolower(
      trimws(as.character(focused_decision$approved))
    ) %in% c("true", "t", "1", "yes", "y")

    if (any(approved_focused_decision)) {
      stop(
        "This cohort has an approved focused-subclustering decision, but ",
        "the Step 6 object lacks annotation_unit. Rerun Step 6 so the ",
        "approved focused annotation units are retained.",
        call. = FALSE
      )
    }
  }

  message(
    "Legacy Step 6 object detected without annotation_unit; using ",
    "wnn_cluster because this cohort has no approved focused refinement."
  )

  LEGACY_ANALYSIS_UNIT_COLUMN
}

ANALYSIS_UNIT_COLUMN <- resolve_analysis_unit_column(metadata)

analysis_unit_resolution <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  preferred_column = PREFERRED_ANALYSIS_UNIT_COLUMN,
  resolved_column = ANALYSIS_UNIT_COLUMN,
  legacy_fallback_used = !identical(
    ANALYSIS_UNIT_COLUMN,
    PREFERRED_ANALYSIS_UNIT_COLUMN
  ),
  stringsAsFactors = FALSE
)

analysis_parameters <- dplyr::bind_rows(
  analysis_parameters,
  data.frame(
    parameter = "resolved analysis unit column",
    value = ANALYSIS_UNIT_COLUMN,
    stringsAsFactors = FALSE
  )
)

readr::write_csv(
  analysis_parameters,
  file.path(STEP7_DIR, "diet_response_parameters.csv")
)

readr::write_csv(
  analysis_unit_resolution,
  file.path(STEP7_DIR, "analysis_unit_resolution.csv")
)

required_assays <- c("RNA", "ATAC")
missing_assays <- setdiff(required_assays, names(multiome))

if (length(missing_assays) > 0L) {
  stop(
    "The Step 6 object is missing assay(s): ",
    paste(missing_assays, collapse = ", "),
    call. = FALSE
  )
}

required_metadata <- c(
  "sample_id",
  "diet",
  "wnn_cluster",
  ANALYSIS_UNIT_COLUMN,
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence",
  "nCount_RNA",
  "percent.mt",
  "TSS.enrichment",
  "frip_cr",
  "scDblFinder.score"
)

missing_metadata <- setdiff(required_metadata, colnames(metadata))

if (length(missing_metadata) > 0L) {
  stop(
    "The Step 6 object is missing metadata column(s): ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

atac_depth_column <- if (
  "nCount_ATAC_common" %in% colnames(metadata)
) {
  "nCount_ATAC_common"
} else if (
  "nCount_ATAC" %in% colnames(metadata)
) {
  "nCount_ATAC"
} else {
  stop(
    "No ATAC depth column was found in the Step 6 metadata.",
    call. = FALSE
  )
}

metadata$cell_barcode <- rownames(metadata)
metadata$sample_id <- as.character(metadata$sample_id)
metadata$diet <- as.character(metadata$diet)
metadata$analysis_unit <- as.character(
  metadata[[ANALYSIS_UNIT_COLUMN]]
)
metadata$wnn_cluster <- as.character(metadata$wnn_cluster)
metadata$broad_compartment <- as.character(
  metadata$broad_compartment
)
metadata$cell_state <- as.character(metadata$cell_state)
metadata$cell_state_full <- as.character(metadata$cell_state_full)
metadata$annotation_confidence <- tolower(
  as.character(metadata$annotation_confidence)
)

expected_diets <- c(REFERENCE_DIET, COMPARISON_DIET)
observed_diets <- sort(unique(metadata$diet))

if (!setequal(observed_diets, expected_diets)) {
  stop(
    "Expected exactly CON and HFD. Observed diet values: ",
    paste(observed_diets, collapse = ", "),
    call. = FALSE
  )
}

sample_diet_map <- metadata |>
  dplyr::distinct(sample_id, diet)

if (anyDuplicated(sample_diet_map$sample_id) > 0L) {
  stop(
    "At least one sample_id maps to more than one diet.",
    call. = FALSE
  )
}

libraries_per_diet <- sample_diet_map |>
  dplyr::count(diet, name = "libraries")

replicated_design <- all(libraries_per_diet$libraries >= 2L)

if (!replicated_design) {
  message(
    "No replicated diet inference will be performed: one or both diets ",
    "have fewer than two independent libraries."
  )
}

rna_layers <- SeuratObject::Layers(multiome[["RNA"]])
atac_layers <- SeuratObject::Layers(multiome[["ATAC"]])

if (!"counts" %in% rna_layers || !"data" %in% rna_layers) {
  stop(
    "The RNA assay must contain joined counts and data layers.",
    call. = FALSE
  )
}

if (!"counts" %in% atac_layers) {
  stop(
    "The ATAC assay must contain a counts layer.",
    call. = FALSE
  )
}

rna_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)

rna_data <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "data"
)

atac_counts <- SeuratObject::LayerData(
  multiome[["ATAC"]],
  layer = "counts"
)

metadata_cells <- rownames(metadata)

align_matrix_to_metadata <- function(
    feature_matrix,
    metadata_cells,
    matrix_label
) {
  matrix_cells <- colnames(feature_matrix)

  if (is.null(matrix_cells)) {
    stop(matrix_label, " has no cell names.", call. = FALSE)
  }

  if (anyDuplicated(matrix_cells) > 0L) {
    stop(matrix_label, " contains duplicate cell names.", call. = FALSE)
  }

  if (anyDuplicated(metadata_cells) > 0L) {
    stop("The Step 6 metadata contains duplicate cell names.", call. = FALSE)
  }

  missing_from_matrix <- setdiff(metadata_cells, matrix_cells)
  extra_in_matrix <- setdiff(matrix_cells, metadata_cells)

  if (
    length(missing_from_matrix) > 0L ||
      length(extra_in_matrix) > 0L
  ) {
    stop(
      matrix_label,
      " and the Step 6 metadata contain different cell sets. ",
      "Missing from matrix: ",
      length(missing_from_matrix),
      "; extra in matrix: ",
      length(extra_in_matrix),
      ". Example missing cells: ",
      paste(utils::head(missing_from_matrix, 5L), collapse = ", "),
      ". Example extra cells: ",
      paste(utils::head(extra_in_matrix, 5L), collapse = ", "),
      ".",
      call. = FALSE
    )
  }

  if (!identical(matrix_cells, metadata_cells)) {
    message(
      matrix_label,
      " contains the correct cells in a different order; ",
      "reordering its columns to match the Step 6 metadata."
    )

    feature_matrix <- feature_matrix[
      ,
      metadata_cells,
      drop = FALSE
    ]
  }

  if (!identical(colnames(feature_matrix), metadata_cells)) {
    stop(
      "Failed to align ",
      matrix_label,
      " to the Step 6 metadata.",
      call. = FALSE
    )
  }

  feature_matrix
}

matrix_alignment_audit <- data.frame(
  matrix = c("RNA counts", "RNA data", "ATAC counts"),
  matrix_cells = c(
    ncol(rna_counts),
    ncol(rna_data),
    ncol(atac_counts)
  ),
  metadata_cells = length(metadata_cells),
  same_cell_set = c(
    setequal(colnames(rna_counts), metadata_cells),
    setequal(colnames(rna_data), metadata_cells),
    setequal(colnames(atac_counts), metadata_cells)
  ),
  identical_order_before_alignment = c(
    identical(colnames(rna_counts), metadata_cells),
    identical(colnames(rna_data), metadata_cells),
    identical(colnames(atac_counts), metadata_cells)
  ),
  stringsAsFactors = FALSE
)

rna_counts <- align_matrix_to_metadata(
  rna_counts,
  metadata_cells,
  "RNA counts"
)

rna_data <- align_matrix_to_metadata(
  rna_data,
  metadata_cells,
  "RNA data"
)

atac_counts <- align_matrix_to_metadata(
  atac_counts,
  metadata_cells,
  "ATAC counts"
)

readr::write_csv(
  matrix_alignment_audit,
  file.path(STEP7_DIR, "matrix_cell_alignment_audit.csv")
)

run_inventory <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  cohort_group = COHORT_GROUP,
  nuclei = nrow(metadata),
  RNA_features = nrow(rna_counts),
  ATAC_features = nrow(atac_counts),
  independent_libraries = nrow(sample_diet_map),
  replicated_diet_design = replicated_design,
  stringsAsFactors = FALSE
)

run_inventory

readr::write_csv(
  run_inventory,
  file.path(STEP7_DIR, "diet_response_run_inventory.csv")
)
```

# 4. Define cell-state lineages and analysis eligibility

Eligibility is determined before looking at diet-associated RNA or ATAC effect
sizes. Low-confidence annotations are retained in the object and composition
tables but excluded from the primary molecular analysis.

```{r define-analysis-eligibility}
annotate_lineage <- function(cell_state, broad_compartment) {
  dplyr::case_when(
    broad_compartment != "epithelial" ~ broad_compartment,
    grepl(
      "ISC|crypt|transit",
      cell_state,
      ignore.case = TRUE
    ) ~ "crypt",
    grepl(
      "absorptive|colonocyte",
      cell_state,
      ignore.case = TRUE
    ) ~ "absorptive",
    grepl(
      "goblet|Reg4|secretory",
      cell_state,
      ignore.case = TRUE
    ) ~ "secretory",
    grepl(
      "Enterochromaffin|L cell",
      cell_state,
      ignore.case = TRUE
    ) ~ "enteroendocrine",
    grepl("Tuft", cell_state, ignore.case = TRUE) ~ "tuft",
    TRUE ~ "other_epithelial"
  )
}

metadata$analysis_lineage <- annotate_lineage(
  metadata$cell_state,
  metadata$broad_compartment
)

unit_annotation <- metadata |>
  dplyr::distinct(
    analysis_unit,
    wnn_cluster,
    broad_compartment,
    analysis_lineage,
    cell_state,
    cell_state_full,
    annotation_confidence
  )

duplicate_unit_annotations <- unit_annotation |>
  dplyr::count(analysis_unit, name = "annotation_rows") |>
  dplyr::filter(annotation_rows != 1L)

if (nrow(duplicate_unit_annotations) > 0L) {
  stop(
    "At least one annotation unit maps to multiple biological labels.",
    call. = FALSE
  )
}

unit_counts <- metadata |>
  dplyr::count(analysis_unit, diet, name = "nuclei") |>
  tidyr::complete(
    analysis_unit,
    diet = expected_diets,
    fill = list(nuclei = 0L)
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = nuclei,
    names_prefix = "nuclei_"
  )

eligibility <- unit_annotation |>
  dplyr::left_join(unit_counts, by = "analysis_unit") |>
  dplyr::mutate(
    minimum_nuclei_per_diet = pmin(
      .data[[paste0("nuclei_", REFERENCE_DIET)]],
      .data[[paste0("nuclei_", COMPARISON_DIET)]]
    ),
    eligibility_tier = dplyr::case_when(
      annotation_confidence == "low" ~ "exclude_primary",
      broad_compartment != "epithelial" ~ "secondary_compartment",
      minimum_nuclei_per_diet >= TIER_A_MIN_NUCLEI_PER_DIET ~
        "Tier_A",
      minimum_nuclei_per_diet >= TIER_B_MIN_NUCLEI_PER_DIET ~
        "Tier_B",
      TRUE ~ "Tier_C"
    ),
    primary_RNA_ATAC = eligibility_tier %in% c("Tier_A", "Tier_B"),
    interpretation = dplyr::case_when(
      eligibility_tier == "Tier_A" ~
        "Primary descriptive RNA and ATAC comparison",
      eligibility_tier == "Tier_B" ~
        "Biologically prioritized comparison with reduced cell support",
      eligibility_tier == "Tier_C" ~
        "Composition and exploratory marker review only",
      eligibility_tier == "secondary_compartment" ~
        "Analyze separately from the epithelial benchmark",
      TRUE ~
        "Retain for sensitivity analysis; exclude from primary inference"
    )
  ) |>
  dplyr::arrange(
    factor(
      eligibility_tier,
      levels = c(
        "Tier_A",
        "Tier_B",
        "Tier_C",
        "secondary_compartment",
        "exclude_primary"
      )
    ),
    analysis_unit
  )

eligibility

readr::write_csv(
  eligibility,
  file.path(STEP7_TABLE_DIR, "analysis_population_eligibility.csv")
)

primary_units <- eligibility |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::pull(analysis_unit)

if (length(primary_units) == 0L) {
  stop(
    "No annotation unit meets the primary Tier A or Tier B criteria.",
    call. = FALSE
  )
}

cat(
  "Primary RNA/ATAC analysis units:",
  paste(primary_units, collapse = ", "),
  "\n"
)
```

## Cluster-level QC sensitivity

```{r cluster-qc-sensitivity}
qc_metrics <- c(
  "nCount_RNA",
  atac_depth_column,
  "percent.mt",
  "TSS.enrichment",
  "frip_cr",
  "scDblFinder.score"
)

if ("RNA.weight" %in% colnames(metadata)) {
  qc_metrics <- c(qc_metrics, "RNA.weight")
}

cluster_qc_summary <- metadata |>
  dplyr::group_by(analysis_unit, diet) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    dplyr::across(
      dplyr::all_of(qc_metrics),
      list(
        median = ~ stats::median(.x, na.rm = TRUE),
        q25 = ~ stats::quantile(.x, 0.25, na.rm = TRUE),
        q75 = ~ stats::quantile(.x, 0.75, na.rm = TRUE)
      )
    ),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    eligibility |>
      dplyr::select(
        analysis_unit,
        cell_state,
        annotation_confidence,
        eligibility_tier
      ),
    by = "analysis_unit"
  )

readr::write_csv(
  cluster_qc_summary,
  file.path(STEP7_TABLE_DIR, "analysis_unit_QC_by_diet.csv")
)
```

# 5. Describe composition using explicit denominators

Composition is reported three ways because a change in captured stromal,
immune, or low-confidence nuclei can alter the percentage of every epithelial
population even when the epithelial biology is unchanged.

```{r composition-summaries}
composition_all_nuclei <- metadata |>
  dplyr::count(sample_id, diet, analysis_unit, name = "nuclei") |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    denominator = sum(nuclei),
    percent = 100 * nuclei / denominator,
    denominator_definition = "all_retained_nuclei"
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    unit_annotation |>
      dplyr::select(
        analysis_unit,
        broad_compartment,
        analysis_lineage,
        cell_state,
        annotation_confidence
      ),
    by = "analysis_unit"
  )

curated_epithelial_metadata <- metadata |>
  dplyr::filter(
    broad_compartment == "epithelial",
    annotation_confidence != "low"
  )

composition_curated_epithelium <- curated_epithelial_metadata |>
  dplyr::count(sample_id, diet, analysis_unit, name = "nuclei") |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    denominator = sum(nuclei),
    percent = 100 * nuclei / denominator,
    denominator_definition = "high_or_moderate_confidence_epithelium"
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    unit_annotation |>
      dplyr::select(
        analysis_unit,
        broad_compartment,
        analysis_lineage,
        cell_state,
        annotation_confidence
      ),
    by = "analysis_unit"
  )

composition_by_lineage <- curated_epithelial_metadata |>
  dplyr::count(
    sample_id,
    diet,
    analysis_lineage,
    analysis_unit,
    name = "nuclei"
  ) |>
  dplyr::group_by(sample_id, analysis_lineage) |>
  dplyr::mutate(
    denominator = sum(nuclei),
    percent = 100 * nuclei / denominator,
    denominator_definition = paste0(
      "within_",
      analysis_lineage
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    unit_annotation |>
      dplyr::select(
        analysis_unit,
        cell_state,
        annotation_confidence
      ),
    by = "analysis_unit"
  )

lineage_composition <- curated_epithelial_metadata |>
  dplyr::count(
    sample_id,
    diet,
    analysis_lineage,
    name = "nuclei"
  ) |>
  dplyr::group_by(sample_id) |>
  dplyr::mutate(
    epithelial_denominator = sum(nuclei),
    percent_of_curated_epithelium =
      100 * nuclei / epithelial_denominator
  ) |>
  dplyr::ungroup()

readr::write_csv(
  composition_all_nuclei,
  file.path(STEP7_TABLE_DIR, "composition_all_retained_nuclei.csv")
)

readr::write_csv(
  composition_curated_epithelium,
  file.path(
    STEP7_TABLE_DIR,
    "composition_curated_epithelial_nuclei.csv"
  )
)

readr::write_csv(
  composition_by_lineage,
  file.path(STEP7_TABLE_DIR, "composition_within_lineage.csv")
)

readr::write_csv(
  lineage_composition,
  file.path(STEP7_TABLE_DIR, "lineage_composition.csv")
)
```

## Composition effect-size plot

```{r plot-composition-effect-sizes}
composition_effects <- composition_all_nuclei |>
  dplyr::select(
    analysis_unit,
    cell_state,
    diet,
    percent
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = percent,
    values_fill = 0
  ) |>
  dplyr::mutate(
    difference_percentage_points =
      .data[[COMPARISON_DIET]] - .data[[REFERENCE_DIET]],
    abundance_ratio =
      (.data[[COMPARISON_DIET]] + 0.01) /
      (.data[[REFERENCE_DIET]] + 0.01),
    display_label = paste0(analysis_unit, ": ", cell_state)
  ) |>
  dplyr::left_join(
    eligibility |>
      dplyr::select(analysis_unit, eligibility_tier),
    by = "analysis_unit"
  ) |>
  dplyr::arrange(difference_percentage_points)

readr::write_csv(
  composition_effects,
  file.path(STEP7_TABLE_DIR, "composition_effect_sizes.csv")
)

composition_effects$display_label <- factor(
  composition_effects$display_label,
  levels = composition_effects$display_label
)

p_composition_effects <- ggplot2::ggplot(
  composition_effects,
  ggplot2::aes(
    x = difference_percentage_points,
    y = display_label,
    color = eligibility_tier
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey60",
    linewidth = 0.4
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = difference_percentage_points,
      yend = display_label
    ),
    linewidth = 0.7
  ) +
  ggplot2::geom_point(size = 2.4) +
  ggplot2::labs(
    title = paste0(
      COHORT_GROUP,
      ": descriptive cell-state composition difference"
    ),
    subtitle = paste0(
      COMPARISON_DIET,
      " minus ",
      REFERENCE_DIET,
      "; one pooled library per diet is not a replicated test"
    ),
    x = "Difference in percentage points",
    y = NULL,
    color = "Analysis tier"
  ) +
  theme_multiome(base_size = 9)

print(p_composition_effects)

ggplot2::ggsave(
  file.path(
    STEP7_PLOT_DIR,
    "01_composition_effect_sizes_all_nuclei.pdf"
  ),
  p_composition_effects,
  width = 10,
  height = 8
)
```

# 6. Audit candidate regional programs within matched states

These feature groups are an exploratory audit, not approved regional labels.
They intentionally remain separate from `cell_state`. Because many regional
markers also change during differentiation, comparisons should be interpreted
within the same annotation unit rather than across unrelated clusters.

```{r regional-program-audit}
regional_feature_key <- dplyr::bind_rows(
  lapply(
    names(regional_audit_programs),
    function(program_name) {
      data.frame(
        program = program_name,
        gene = regional_audit_programs[[program_name]],
        stringsAsFactors = FALSE
      )
    }
  )
)

regional_feature_key$present_in_RNA <-
  regional_feature_key$gene %in% rownames(rna_counts)

readr::write_csv(
  regional_feature_key,
  file.path(STEP7_TABLE_DIR, "regional_feature_audit.csv")
)

regional_feature_summary <- list()
regional_summary_index <- 1L

for (feature_index in seq_len(nrow(regional_feature_key))) {
  gene <- regional_feature_key$gene[[feature_index]]

  if (!gene %in% rownames(rna_counts)) {
    next
  }

  for (analysis_unit_value in unique(metadata$analysis_unit)) {
    for (diet_value in expected_diets) {
      selected_cells <- metadata$cell_barcode[
        metadata$analysis_unit == analysis_unit_value &
          metadata$diet == diet_value
      ]

      if (length(selected_cells) == 0L) {
        next
      }

      count_values <- rna_counts[gene, selected_cells, drop = TRUE]
      normalized_values <- rna_data[
        gene,
        selected_cells,
        drop = TRUE
      ]

      regional_feature_summary[[regional_summary_index]] <- data.frame(
        analysis_unit = analysis_unit_value,
        diet = diet_value,
        program = regional_feature_key$program[[feature_index]],
        gene = gene,
        nuclei = length(selected_cells),
        percent_detected = 100 * mean(count_values > 0),
        average_log_normalized_expression = mean(normalized_values),
        stringsAsFactors = FALSE
      )

      regional_summary_index <- regional_summary_index + 1L
    }
  }
}

regional_feature_summary <- dplyr::bind_rows(
  regional_feature_summary
) |>
  dplyr::left_join(
    unit_annotation |>
      dplyr::select(analysis_unit, cell_state),
    by = "analysis_unit"
  )

readr::write_csv(
  regional_feature_summary,
  file.path(
    STEP7_TABLE_DIR,
    "regional_feature_expression_by_unit_and_diet.csv"
  )
)

p_regional_audit <- regional_feature_summary |>
  dplyr::filter(analysis_unit %in% primary_units) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = analysis_unit,
      y = gene,
      size = percent_detected,
      color = average_log_normalized_expression
    )
  ) +
  ggplot2::geom_point() +
  ggplot2::facet_grid(
    program ~ diet,
    scales = "free_y",
    space = "free_y"
  ) +
  ggplot2::scale_color_gradientn(
    colors = c("#2C7BB6", "#F7F7F7", "#D7191C")
  ) +
  ggplot2::labs(
    title = "Candidate regional-axis audit within annotation units",
    subtitle = "Use within-state differences; these are not regional labels.",
    x = "Annotation unit",
    y = NULL,
    size = "Percent detected",
    color = "Average log-normalized expression"
  ) +
  theme_multiome(base_size = 9)

print(p_regional_audit)

ggplot2::ggsave(
  file.path(STEP7_PLOT_DIR, "02_regional_feature_audit.pdf"),
  p_regional_audit,
  width = 12,
  height = 8
)
```

# 7. Calculate descriptive RNA pseudobulk effects

Each annotation unit and diet is collapsed into one pooled count vector. The
reported log2 fold change is therefore a contrast between pooled libraries,
not a DESeq2-style estimate with biological variance.

```{r descriptive-effect-functions}
summarize_pooled_effect <- function(
  count_matrix,
  feature_label,
  metadata,
  analysis_unit_value,
  reference_diet,
  comparison_diet,
  minimum_total_counts,
  minimum_detection_percent,
  cpm_pseudocount
) {
  reference_cells <- metadata$cell_barcode[
    metadata$analysis_unit == analysis_unit_value &
      metadata$diet == reference_diet
  ]

  comparison_cells <- metadata$cell_barcode[
    metadata$analysis_unit == analysis_unit_value &
      metadata$diet == comparison_diet
  ]

  if (
    length(reference_cells) == 0L ||
      length(comparison_cells) == 0L
  ) {
    return(NULL)
  }

  reference_counts <- Matrix::rowSums(
    count_matrix[, reference_cells, drop = FALSE]
  )

  comparison_counts <- Matrix::rowSums(
    count_matrix[, comparison_cells, drop = FALSE]
  )

  reference_detected <- Matrix::rowSums(
    count_matrix[, reference_cells, drop = FALSE] > 0
  )

  comparison_detected <- Matrix::rowSums(
    count_matrix[, comparison_cells, drop = FALSE] > 0
  )

  reference_library_size <- sum(reference_counts)
  comparison_library_size <- sum(comparison_counts)

  if (
    reference_library_size <= 0 ||
      comparison_library_size <= 0
  ) {
    warning(
      "Skipping analysis unit ",
      analysis_unit_value,
      " because at least one pooled library has zero total counts."
    )
    return(NULL)
  }

  reference_cpm <- 1e6 * reference_counts / reference_library_size
  comparison_cpm <- 1e6 * comparison_counts / comparison_library_size

  reference_percent <-
    100 * reference_detected / length(reference_cells)
  comparison_percent <-
    100 * comparison_detected / length(comparison_cells)

  result <- data.frame(
    analysis_unit = analysis_unit_value,
    feature = rownames(count_matrix),
    feature_type = feature_label,
    reference_diet = reference_diet,
    comparison_diet = comparison_diet,
    reference_nuclei = length(reference_cells),
    comparison_nuclei = length(comparison_cells),
    reference_counts = reference_counts,
    comparison_counts = comparison_counts,
    total_counts = reference_counts + comparison_counts,
    reference_CPM = reference_cpm,
    comparison_CPM = comparison_cpm,
    log2FC_comparison_vs_reference = log2(
      (comparison_cpm + cpm_pseudocount) /
        (reference_cpm + cpm_pseudocount)
    ),
    reference_percent_detected = reference_percent,
    comparison_percent_detected = comparison_percent,
    detection_difference_percentage_points =
      comparison_percent - reference_percent,
    stringsAsFactors = FALSE
  )

  result |>
    dplyr::filter(
      total_counts >= minimum_total_counts,
      pmax(
        reference_percent_detected,
        comparison_percent_detected
      ) >= minimum_detection_percent
    )
}

select_top_effects <- function(
  effects,
  absolute_log2fc_threshold,
  top_features_per_direction,
  detection_delta_threshold = NULL
) {
  selected <- effects |>
    dplyr::filter(
      abs(log2FC_comparison_vs_reference) >=
        absolute_log2fc_threshold
    )

  if (!is.null(detection_delta_threshold)) {
    selected <- selected |>
      dplyr::filter(
        abs(detection_difference_percentage_points) >=
          detection_delta_threshold
      )
  }

  selected |>
    dplyr::mutate(
      direction = dplyr::if_else(
        log2FC_comparison_vs_reference > 0,
        paste0(COMPARISON_DIET, "_higher"),
        paste0(REFERENCE_DIET, "_higher")
      )
    ) |>
    dplyr::group_by(analysis_unit, direction) |>
    dplyr::slice_max(
      order_by = abs(log2FC_comparison_vs_reference),
      n = top_features_per_direction,
      with_ties = FALSE
    ) |>
    dplyr::ungroup()
}
```

```{r calculate-rna-effects}
rna_effects <- dplyr::bind_rows(
  lapply(
    primary_units,
    function(analysis_unit_value) {
      summarize_pooled_effect(
        count_matrix = rna_counts,
        feature_label = "gene",
        metadata = metadata,
        analysis_unit_value = analysis_unit_value,
        reference_diet = REFERENCE_DIET,
        comparison_diet = COMPARISON_DIET,
        minimum_total_counts = RNA_MIN_TOTAL_COUNTS,
        minimum_detection_percent =
          RNA_MIN_DETECTION_PERCENT,
        cpm_pseudocount = CPM_PSEUDOCOUNT
      )
    }
  )
) |>
  dplyr::left_join(
    eligibility |>
      dplyr::select(
        analysis_unit,
        cell_state,
        cell_state_full,
        eligibility_tier
      ),
    by = "analysis_unit"
  )

rna_top_candidates <- select_top_effects(
  effects = rna_effects,
  absolute_log2fc_threshold = RNA_EFFECT_LOG2FC_THRESHOLD,
  top_features_per_direction = RNA_TOP_FEATURES_PER_DIRECTION,
  detection_delta_threshold =
    RNA_EFFECT_DETECTION_DELTA_THRESHOLD
)

readr::write_csv(
  rna_effects,
  file.path(
    STEP7_TABLE_DIR,
    "RNA_pooled_library_effects_by_analysis_unit.csv.gz"
  )
)

readr::write_csv(
  rna_top_candidates,
  file.path(
    STEP7_TABLE_DIR,
    "RNA_top_diet_associated_candidates.csv"
  )
)

rna_top_candidates |>
  dplyr::count(analysis_unit, cell_state, direction)
```

# 8. Calculate descriptive ATAC pseudobulk effects

```{r calculate-atac-effects}
atac_effects <- dplyr::bind_rows(
  lapply(
    primary_units,
    function(analysis_unit_value) {
      summarize_pooled_effect(
        count_matrix = atac_counts,
        feature_label = "peak",
        metadata = metadata,
        analysis_unit_value = analysis_unit_value,
        reference_diet = REFERENCE_DIET,
        comparison_diet = COMPARISON_DIET,
        minimum_total_counts = ATAC_MIN_TOTAL_COUNTS,
        minimum_detection_percent =
          ATAC_MIN_DETECTION_PERCENT,
        cpm_pseudocount = CPM_PSEUDOCOUNT
      )
    }
  )
) |>
  dplyr::left_join(
    eligibility |>
      dplyr::select(
        analysis_unit,
        cell_state,
        cell_state_full,
        eligibility_tier
      ),
    by = "analysis_unit"
  )

atac_top_candidates <- select_top_effects(
  effects = atac_effects,
  absolute_log2fc_threshold = ATAC_EFFECT_LOG2FC_THRESHOLD,
  top_features_per_direction = ATAC_TOP_FEATURES_PER_DIRECTION
)

readr::write_csv(
  atac_effects,
  file.path(
    STEP7_TABLE_DIR,
    "ATAC_pooled_library_effects_by_analysis_unit.csv.gz"
  )
)

readr::write_csv(
  atac_top_candidates,
  file.path(
    STEP7_TABLE_DIR,
    "ATAC_top_diet_associated_candidates.csv"
  )
)

atac_top_candidates |>
  dplyr::count(analysis_unit, cell_state, direction)
```

# 9. Evaluate balanced cell-sampling stability

This analysis repeatedly uses the same number of nuclei from each diet. It
tests whether an effect is dominated by unequal cell numbers or a small set of
nuclei. It does not measure mouse-to-mouse reproducibility.

```{r balanced-downsampling-function}
downsample_effect_stability <- function(
  count_matrix,
  metadata,
  full_effects,
  analysis_units,
  iterations,
  maximum_cells_per_diet,
  maximum_features_per_unit,
  cpm_pseudocount,
  random_seed
) {
  set.seed(random_seed)
  iteration_results <- list()
  result_index <- 1L
  cell_library_sizes <- Matrix::colSums(count_matrix)

  for (analysis_unit_value in analysis_units) {
    unit_effects <- full_effects |>
      dplyr::filter(analysis_unit == analysis_unit_value) |>
      dplyr::slice_max(
        order_by = abs(log2FC_comparison_vs_reference),
        n = maximum_features_per_unit,
        with_ties = FALSE
      )

    if (nrow(unit_effects) == 0L) {
      warning(
        "Skipping downsampling for analysis unit ",
        analysis_unit_value,
        " because no features passed the pooled-effect filters."
      )
      next
    }

    feature_index <- match(
      unit_effects$feature,
      rownames(count_matrix)
    )

    valid_features <- !is.na(feature_index)
    unit_effects <- unit_effects[valid_features, , drop = FALSE]
    feature_index <- feature_index[valid_features]

    reference_cells <- metadata$cell_barcode[
      metadata$analysis_unit == analysis_unit_value &
        metadata$diet == REFERENCE_DIET
    ]

    comparison_cells <- metadata$cell_barcode[
      metadata$analysis_unit == analysis_unit_value &
        metadata$diet == COMPARISON_DIET
    ]

    sampled_nuclei <- min(
      length(reference_cells),
      length(comparison_cells),
      maximum_cells_per_diet
    )

    if (sampled_nuclei < 1L) {
      warning(
        "Skipping downsampling for analysis unit ",
        analysis_unit_value,
        " because one diet has no nuclei."
      )
      next
    }

    for (iteration in seq_len(iterations)) {
      sampled_reference <- sample(
        reference_cells,
        sampled_nuclei,
        replace = FALSE
      )

      sampled_comparison <- sample(
        comparison_cells,
        sampled_nuclei,
        replace = FALSE
      )

      sampled_reference_counts <- Matrix::rowSums(
        count_matrix[
          feature_index,
          sampled_reference,
          drop = FALSE
        ]
      )

      sampled_comparison_counts <- Matrix::rowSums(
        count_matrix[
          feature_index,
          sampled_comparison,
          drop = FALSE
        ]
      )

      sampled_reference_cpm <-
        1e6 * sampled_reference_counts /
        sum(cell_library_sizes[sampled_reference])

      sampled_comparison_cpm <-
        1e6 * sampled_comparison_counts /
        sum(cell_library_sizes[sampled_comparison])

      iteration_results[[result_index]] <- data.frame(
        analysis_unit = analysis_unit_value,
        feature = unit_effects$feature,
        iteration = iteration,
        sampled_nuclei_per_diet = sampled_nuclei,
        downsampled_log2FC = log2(
          (sampled_comparison_cpm + cpm_pseudocount) /
            (sampled_reference_cpm + cpm_pseudocount)
        ),
        full_log2FC =
          unit_effects$log2FC_comparison_vs_reference,
        stringsAsFactors = FALSE
      )

      result_index <- result_index + 1L
    }
  }

  stability_iterations <- dplyr::bind_rows(iteration_results)

  if (nrow(stability_iterations) == 0L) {
    stop(
      "Balanced downsampling produced no feature-level results.",
      call. = FALSE
    )
  }

  stability_iterations |>
    dplyr::group_by(analysis_unit, feature) |>
    dplyr::summarise(
      sampled_nuclei_per_diet = dplyr::first(
        sampled_nuclei_per_diet
      ),
      full_log2FC = dplyr::first(full_log2FC),
      downsampled_log2FC_median = stats::median(
        downsampled_log2FC
      ),
      downsampled_log2FC_q25 = stats::quantile(
        downsampled_log2FC,
        0.25
      ),
      downsampled_log2FC_q75 = stats::quantile(
        downsampled_log2FC,
        0.75
      ),
      sign_concordance_percent = 100 * mean(
        sign(downsampled_log2FC) == sign(full_log2FC)
      ),
      .groups = "drop"
    )
}
```

```{r run-balanced-downsampling}
if (RUN_BALANCED_DOWNSAMPLING) {
  rna_downsampling_stability <- downsample_effect_stability(
    count_matrix = rna_counts,
    metadata = metadata,
    full_effects = rna_effects,
    analysis_units = primary_units,
    iterations = DOWNSAMPLING_ITERATIONS,
    maximum_cells_per_diet = DOWNSAMPLING_MAX_CELLS_PER_DIET,
    maximum_features_per_unit =
      DOWNSAMPLING_MAX_FEATURES_PER_UNIT,
    cpm_pseudocount = CPM_PSEUDOCOUNT,
    random_seed = 20260904
  )

  atac_downsampling_stability <- downsample_effect_stability(
    count_matrix = atac_counts,
    metadata = metadata,
    full_effects = atac_effects,
    analysis_units = primary_units,
    iterations = DOWNSAMPLING_ITERATIONS,
    maximum_cells_per_diet = DOWNSAMPLING_MAX_CELLS_PER_DIET,
    maximum_features_per_unit =
      DOWNSAMPLING_MAX_FEATURES_PER_UNIT,
    cpm_pseudocount = CPM_PSEUDOCOUNT,
    random_seed = 20260905
  )

  readr::write_csv(
    rna_downsampling_stability,
    file.path(
      STEP7_TABLE_DIR,
      "RNA_balanced_downsampling_stability.csv"
    )
  )

  readr::write_csv(
    atac_downsampling_stability,
    file.path(
      STEP7_TABLE_DIR,
      "ATAC_balanced_downsampling_stability.csv"
    )
  )
}
```

# 10. Add nearest genes to the top ATAC candidates

Nearest-gene assignments are genomic context, not proof that a peak regulates
that gene. They are used here to nominate RNA/ATAC pairs for later coverage,
motif, and peak-to-gene review.

```{r annotate-top-atac-peaks}
annotate_nearest_gene <- function(object, peak_table) {
  if (nrow(peak_table) == 0L) {
    return(peak_table)
  }

  unique_peaks <- unique(peak_table$feature)

  nearest_result <- tryCatch(
    {
      peak_index <- match(
        unique_peaks,
        rownames(object[["ATAC"]])
      )

      if (anyNA(peak_index)) {
        stop("At least one candidate peak is absent from the ATAC assay.")
      }

      query_ranges <- GenomicRanges::granges(
        object[["ATAC"]]
      )[peak_index]

      names(query_ranges) <- unique_peaks

      gene_annotation <- Signac::Annotation(
        object[["ATAC"]]
      )

      if (is.null(gene_annotation) || length(gene_annotation) == 0L) {
        stop("The ATAC assay does not contain gene annotations.")
      }

      query_seqlevels <- as.character(
        GenomeInfoDb::seqnames(query_ranges)
      )

      compatible_query <- query_seqlevels %in%
        GenomeInfoDb::seqlevels(gene_annotation)

      if (!any(compatible_query)) {
        stop(
          "Candidate peaks and gene annotations share no chromosome names."
        )
      }

      query_ranges_compatible <- query_ranges[compatible_query]

      nearest_hits <- GenomicRanges::distanceToNearest(
        x = query_ranges_compatible,
        subject = gene_annotation,
        ignore.strand = TRUE
      )

      nearest <- as.data.frame(
        S4Vectors::mcols(
          gene_annotation[
            S4Vectors::subjectHits(nearest_hits)
          ]
        )
      )

      nearest$feature <- names(query_ranges_compatible)[
        S4Vectors::queryHits(nearest_hits)
      ]

      nearest$distance <- S4Vectors::mcols(
        nearest_hits
      )$distance

      if (!"gene_name" %in% colnames(nearest)) {
        nearest$gene_name <- NA_character_
      }

      nearest |>
        dplyr::select(
          feature,
          dplyr::any_of(c(
            "gene_name",
            "gene_id",
            "distance",
            "gene_biotype"
          ))
        ) |>
        dplyr::distinct(feature, .keep_all = TRUE)
    },
    error = function(error_condition) {
      warning(
        "Nearest-gene annotation was skipped: ",
        conditionMessage(error_condition)
      )

      data.frame(
        feature = unique_peaks,
        gene_name = NA_character_,
        distance = NA_real_,
        stringsAsFactors = FALSE
      )
    }
  )

  peak_table |>
    dplyr::left_join(nearest_result, by = "feature")
}

atac_top_annotated <- annotate_nearest_gene(
  multiome,
  atac_top_candidates
)

readr::write_csv(
  atac_top_annotated,
  file.path(
    STEP7_TABLE_DIR,
    "ATAC_top_candidates_with_nearest_gene.csv"
  )
)
```

# 11. Join directionally concordant RNA and ATAC candidates

```{r rna-atac-concordance}
rna_for_concordance <- rna_effects |>
  dplyr::select(
    analysis_unit,
    gene = feature,
    RNA_log2FC = log2FC_comparison_vs_reference,
    RNA_detection_difference =
      detection_difference_percentage_points,
    RNA_reference_CPM = reference_CPM,
    RNA_comparison_CPM = comparison_CPM
  )

rna_atac_concordance <- atac_top_annotated |>
  dplyr::filter(!is.na(gene_name), nzchar(gene_name)) |>
  dplyr::rename(
    peak = feature,
    nearest_gene = gene_name,
    ATAC_log2FC = log2FC_comparison_vs_reference,
    ATAC_detection_difference =
      detection_difference_percentage_points,
    ATAC_reference_CPM = reference_CPM,
    ATAC_comparison_CPM = comparison_CPM
  ) |>
  dplyr::left_join(
    rna_for_concordance,
    by = c(
      "analysis_unit",
      "nearest_gene" = "gene"
    )
  ) |>
  dplyr::mutate(
    RNA_ATAC_direction = dplyr::case_when(
      is.na(RNA_log2FC) ~ "RNA_not_tested_or_filtered",
      sign(RNA_log2FC) == sign(ATAC_log2FC) ~ "concordant",
      TRUE ~ "discordant"
    ),
    joint_effect_magnitude =
      abs(ATAC_log2FC) + abs(RNA_log2FC)
  ) |>
  dplyr::arrange(
    analysis_unit,
    dplyr::desc(RNA_ATAC_direction == "concordant"),
    dplyr::desc(joint_effect_magnitude)
  )

readr::write_csv(
  rna_atac_concordance,
  file.path(STEP7_TABLE_DIR, "RNA_ATAC_nearest_gene_concordance.csv")
)

concordance_plot_data <- rna_atac_concordance |>
  dplyr::filter(
    !is.na(RNA_log2FC),
    is.finite(RNA_log2FC),
    is.finite(ATAC_log2FC)
  )

if (nrow(concordance_plot_data) > 0L) {
  p_rna_atac_concordance <- ggplot2::ggplot(
    concordance_plot_data,
    ggplot2::aes(
      x = ATAC_log2FC,
      y = RNA_log2FC,
      color = RNA_ATAC_direction
    )
  ) +
    ggplot2::geom_hline(
      yintercept = 0,
      color = "grey70",
      linewidth = 0.3
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "grey70",
      linewidth = 0.3
    ) +
    ggplot2::geom_point(size = 0.7, alpha = 0.55) +
    ggplot2::facet_wrap(~ analysis_unit, scales = "free") +
    ggplot2::scale_color_manual(
      values = c(
        concordant = "#2A9D8F",
        discordant = "#E76F51"
      )
    ) +
    ggplot2::labs(
      title = "Nearest-gene RNA and ATAC effect concordance",
      subtitle = paste0(
        COMPARISON_DIET,
        " versus ",
        REFERENCE_DIET,
        "; nearest-gene context is not a regulatory link"
      ),
      x = "ATAC pooled-library log2 fold change",
      y = "RNA pooled-library log2 fold change",
      color = NULL
    ) +
    theme_multiome(base_size = 9)

  print(p_rna_atac_concordance)

  ggplot2::ggsave(
    file.path(
      STEP7_PLOT_DIR,
      "03_RNA_ATAC_nearest_gene_concordance.pdf"
    ),
    p_rna_atac_concordance,
    width = 13,
    height = 9
  )
}
```

# 12. Optional motif-enrichment gate

Motif enrichment requires a motif matrix compatible with the active GRCm39
peak set. It is off by default so that Step 7 cannot silently substitute an
incompatible genome build or motif annotation.

```{r optional-motif-enrichment}
motif_object_present <- !is.null(
  tryCatch(
    Signac::Motifs(multiome[["ATAC"]]),
    error = function(error_condition) NULL
  )
)

motif_status <- data.frame(
  requested = REQUEST_MOTIF_ENRICHMENT,
  motif_object_present = motif_object_present,
  performed = FALSE,
  note = if (REQUEST_MOTIF_ENRICHMENT && !motif_object_present) {
    paste(
      "Motif enrichment was requested, but the ATAC assay lacks a",
      "compatible motif object. Add GRCm39-compatible motifs first."
    )
  } else if (!REQUEST_MOTIF_ENRICHMENT) {
    "Motif enrichment is intentionally disabled for the first-pass benchmark."
  } else {
    paste(
      "A motif object is present. Perform reviewed enrichment only after",
      "the candidate peak lists have been inspected."
    )
  },
  stringsAsFactors = FALSE
)

motif_status

readr::write_csv(
  motif_status,
  file.path(STEP7_TABLE_DIR, "motif_analysis_status.csv")
)
```

# 13. Save the Step 7 checkpoint

```{r save-step-seven}
step7_summary <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  cohort_group = COHORT_GROUP,
  nuclei = nrow(metadata),
  analysis_unit_source = ANALYSIS_UNIT_COLUMN,
  annotation_units = dplyr::n_distinct(metadata$analysis_unit),
  Tier_A_units = sum(eligibility$eligibility_tier == "Tier_A"),
  Tier_B_units = sum(eligibility$eligibility_tier == "Tier_B"),
  Tier_C_units = sum(eligibility$eligibility_tier == "Tier_C"),
  secondary_compartment_units = sum(eligibility$eligibility_tier == "secondary_compartment"),
  excluded_primary_units = sum(eligibility$eligibility_tier == "exclude_primary"),
  primary_RNA_ATAC_units = length(primary_units),
  RNA_effect_rows = nrow(rna_effects),
  RNA_top_candidate_rows = nrow(rna_top_candidates),
  ATAC_effect_rows = nrow(atac_effects),
  ATAC_top_candidate_rows = nrow(atac_top_candidates),
  replicated_diet_design = replicated_design,
  biological_inference_performed = FALSE,
  stringsAsFactors = FALSE
)

readr::write_csv(
  step7_summary,
  file.path(STEP7_DIR, "step7_summary.csv")
)

writeLines(
  c(
    paste("Step 7 completed:", format(Sys.time())),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste("Cells:", nrow(metadata)),
    paste("Analysis unit source:", ANALYSIS_UNIT_COLUMN),
    paste(
      "Annotation units:",
      dplyr::n_distinct(metadata$analysis_unit)
    ),
    paste(
      "Primary RNA/ATAC analysis units:",
      length(primary_units)
    ),
    paste(
      "Balanced cell downsampling performed:",
      RUN_BALANCED_DOWNSAMPLING
    ),
    paste("Independent libraries per diet:", paste(
      paste0(libraries_per_diet$diet, "=", libraries_per_diet$libraries),
      collapse = ";"
    )),
    paste("Replicated diet design:", replicated_design),
    "Biological-replicate differential inference performed: FALSE",
    paste(
      "Interpretation:",
      "descriptive diet-associated effects between pooled libraries"
    )
  ),
  STEP7_COMPLETE_MARKER
)

step7_summary
```

# Interpretation checkpoint

Before moving to trajectories or cross-genotype comparison, answer:

1. Which composition differences persist after restricting the denominator to
   high- or moderate-confidence epithelial nuclei?
2. Are crypt changes best described as expansion, or as redistribution among
   ISC-like, cycling, and progenitor states?
3. Do candidate regional features differ between diets within the same
   annotation unit?
4. Which RNA effects remain directionally stable after balanced downsampling?
5. Which ATAC effects remain stable and have concordant nearby RNA changes?
6. Are any conclusions driven by low-confidence clusters or rare populations?
7. Which effects should be prioritized for coverage plots, motif analysis, and
   later genotype comparison?

# What comes next?

Step 7B audits and, only after manual approval, models WT epithelial
trajectories. Cross-genotype analysis should compare matched biological states
and directionally stable effect sizes rather than matching cluster numbers.
