---
title: "Step 7B - Audit and model WT epithelial trajectories"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 7B evaluates whether the annotated WT d21 epithelial states support a
defensible RNA trajectory. It does not assume that cluster 8 is an intestinal
stem-cell root simply because it was labeled `ISC-like crypt`.

The d21 epithelium may show greater transcriptional infidelity than later
timepoints. Mixed lineage programs, sparse canonical stem markers, and
cell-cycle structure can therefore create a smooth embedding without defining
a reliable developmental trajectory.

This notebook is intentionally divided by a manual checkpoint:

1. audit candidate root and terminal states in both pooled libraries;
2. quantify stem, cycling, absorptive, and secretory program mixing;
3. visualize RNA topology with and without cell-cycle labels;
4. write a proposed trajectory decision table; and
5. stop until the root, included states, and terminal states are approved.

Only after approval does the notebook fit separate absorptive and secretory
Slingshot trajectories. The separation avoids forcing all epithelial fates
through one highly connected graph.

## Statistical limitation

There is one pooled CON and one pooled HFD library. Pseudotime-distribution and
gene-trend differences are descriptive library comparisons. They do not
estimate mouse-to-mouse variance and must not be presented as replicated diet
effects.

A practical review-stage allocation is **4 CPU cores, 64--96 GiB memory, and
8--12 hours**. The post-approval Slingshot models are smaller than the global
Multiome object but retain RNA data for gene-trend summaries.

# 1. Setup

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
  library(Matrix)
  library(patchwork)
})

ggplot2::theme_set(theme_multiome())

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 7B is the WT d21 trajectory review. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt before running it.",
    call. = FALSE
  )
}

set.seed(20260904)
future::plan(future::sequential)
```

# 2. Define inputs, outputs, and candidate lineages

The proposed root is global WT annotation unit 8, but it is not approved in
code. Cluster 2 is excluded despite crypt-associated markers because its
low-complexity profile and high mitochondrial fraction make its position
difficult to interpret. Cluster 10 and mixed cluster 17 are also excluded.

```{r define-inputs-and-candidates}
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP7_DIR <- file.path(OUTPUT_DIR, "07_diet_response")
TRAJECTORY_DIR <- file.path(OUTPUT_DIR, "07b_trajectories")
TRAJECTORY_PLOT_DIR <- file.path(TRAJECTORY_DIR, "plots")
TRAJECTORY_TABLE_DIR <- file.path(TRAJECTORY_DIR, "tables")

STEP6_OBJECT_PATH <- file.path(
  STEP6_DIR,
  "multiome_annotated.rds"
)

STEP7_COMPLETE_MARKER <- file.path(
  STEP7_DIR,
  "STEP_07_COMPLETE.txt"
)

TRAJECTORY_PROPOSAL_PATH <- file.path(
  TRAJECTORY_DIR,
  "trajectory_decision_proposed.csv"
)

TRAJECTORY_DECISION_PATH <- file.path(
  COHORT_CONFIG_DIR,
  "trajectory_decision.csv"
)

TRAJECTORY_REVIEW_MARKER <- file.path(
  TRAJECTORY_DIR,
  "STEP_07B_READY_FOR_TRAJECTORY_REVIEW.txt"
)

TRAJECTORY_COMPLETE_MARKER <- file.path(
  TRAJECTORY_DIR,
  "STEP_07B_COMPLETE.txt"
)

dir.create(TRAJECTORY_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TRAJECTORY_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TRAJECTORY_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)

assert_file(STEP6_OBJECT_PATH, "Step 6 annotated multimodal object")
assert_file(STEP7_COMPLETE_MARKER, "Step 7 completion marker")

PREFERRED_ANALYSIS_UNIT_COLUMN <- "annotation_unit"
LEGACY_ANALYSIS_UNIT_COLUMN <- "wnn_cluster"
REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"

MINIMUM_NUCLEI_PER_DIET_FOR_ROOT <- 100L
N_VARIABLE_FEATURES <- 3000L
N_RNA_PCS <- 30L
RNA_DIMS_TRAJECTORY <- 1:20
N_PSEUDOTIME_BINS <- 10L
MINIMUM_CURVE_WEIGHT <- 0.50

trajectory_programs <- list(
  stem = c(
    "Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Ephb2"
  ),
  crypt = c(
    "Cftr", "Htr4", "Slc12a2", "Pde3a", "Dmbt1"
  ),
  cycling = c(
    "Mki67", "Top2a", "Cdk1", "Ccnb1", "Ccna2", "Kif11"
  ),
  absorptive = c(
    "Car1", "Car4", "Aqp8", "Slc26a3", "Krt20", "Tgm3"
  ),
  secretory = c(
    "Atoh1", "Spdef", "Muc2", "Agr2", "Kit", "Spink4"
  ),
  mature_goblet = c(
    "Fcgbp", "Clca1", "Zg16", "Tff3", "Bcas1"
  ),
  deep_crypt_secretory = c(
    "Reg4", "Mptx1", "Retnlb", "Ccl6", "Selenom"
  )
)

candidate_trajectories <- data.frame(
  cohort_id = "wt",
  trajectory_id = c(
    "crypt_to_absorptive",
    "crypt_to_secretory"
  ),
  analysis_units = c(
    "8;0;6;1;3",
    "8;0;6;18;4;7;5;9"
  ),
  root_unit = c("8", "8"),
  terminal_units = c("3", "5;9"),
  approved = FALSE,
  reviewer = "",
  review_date = "",
  decision_notes = c(
    paste(
      "Proposed ISC-like to immature absorptive to mature colonocyte path.",
      "Approve only if unit 8 is root-like in both libraries and RNA topology",
      "is not dominated by cell cycle or mixed lineage expression."
    ),
    paste(
      "Proposed ISC-like/crypt to cycling secretory and goblet/DCS branches.",
      "Approve only if secretory commitment is coherent and the Reg4+ DCS",
      "endpoint is supported independently of diet composition."
    )
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  candidate_trajectories,
  TRAJECTORY_PROPOSAL_PATH
)
```

# 3. Load and validate the WT object

```{r load-and-validate-wt-object}
multiome <- readRDS(STEP6_OBJECT_PATH)
metadata <- multiome[[]]

resolve_analysis_unit_column <- function(metadata) {
  if (PREFERRED_ANALYSIS_UNIT_COLUMN %in% colnames(metadata)) {
    return(PREFERRED_ANALYSIS_UNIT_COLUMN)
  }

  if (!LEGACY_ANALYSIS_UNIT_COLUMN %in% colnames(metadata)) {
    stop(
      "The Step 6 object contains neither '",
      PREFERRED_ANALYSIS_UNIT_COLUMN,
      "' nor '",
      LEGACY_ANALYSIS_UNIT_COLUMN,
      "'.",
      call. = FALSE
    )
  }

  focused_decision_path <- file.path(
    COHORT_CONFIG_DIR,
    "subclustering_decision.csv"
  )

  if (file.exists(focused_decision_path)) {
    focused_decision <- suppressMessages(
      readr::read_csv(
        focused_decision_path,
        show_col_types = FALSE,
        progress = FALSE
      )
    )

    if ("approved" %in% colnames(focused_decision)) {
      approved_focused_decision <- tolower(
        trimws(as.character(focused_decision$approved))
      ) %in% c("true", "t", "1", "yes", "y")

      if (any(approved_focused_decision)) {
        stop(
          "This cohort has an approved focused-subclustering decision, ",
          "but the Step 6 object lacks annotation_unit. Rerun Step 6 ",
          "before trajectory analysis.",
          call. = FALSE
        )
      }
    }
  }

  message(
    "Legacy WT Step 6 object detected without annotation_unit; ",
    "using wnn_cluster as the trajectory analysis unit."
  )

  LEGACY_ANALYSIS_UNIT_COLUMN
}

ANALYSIS_UNIT_COLUMN <- resolve_analysis_unit_column(metadata)

trajectory_analysis_unit_resolution <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  preferred_column = PREFERRED_ANALYSIS_UNIT_COLUMN,
  resolved_column = ANALYSIS_UNIT_COLUMN,
  legacy_fallback_used = !identical(
    ANALYSIS_UNIT_COLUMN,
    PREFERRED_ANALYSIS_UNIT_COLUMN
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  trajectory_analysis_unit_resolution,
  file.path(
    TRAJECTORY_DIR,
    "trajectory_analysis_unit_resolution.csv"
  )
)

if (!all(c("RNA", "ATAC") %in% names(multiome))) {
  stop(
    "The WT object must contain RNA and ATAC assays.",
    call. = FALSE
  )
}

rna_layers <- SeuratObject::Layers(multiome[["RNA"]])

if (!all(c("counts", "data") %in% rna_layers)) {
  stop(
    "The RNA assay must contain joined counts and data layers.",
    call. = FALSE
  )
}

cell_cycle_columns <- c(
  "Phase",
  "S.Score",
  "G2M.Score"
)

cell_cycle_columns_present_before <- cell_cycle_columns %in%
  colnames(metadata)

cell_cycle_score_source <- "Step 6 metadata"
matched_s_features <- character()
matched_g2m_features <- character()

if (!all(cell_cycle_columns_present_before)) {
  message(
    "Step 6 metadata lacks one or more cell-cycle columns; ",
    "computing Phase, S.Score, and G2M.Score from normalized RNA data."
  )

  match_features_case_insensitive <- function(
      requested_features,
      available_features
  ) {
    matched_index <- match(
      toupper(requested_features),
      toupper(available_features)
    )

    unique(
      available_features[
        matched_index[!is.na(matched_index)]
      ]
    )
  }

  cell_cycle_reference <- Seurat::cc.genes.updated.2019
  available_rna_features <- rownames(multiome[["RNA"]])

  matched_s_features <- match_features_case_insensitive(
    cell_cycle_reference$s.genes,
    available_rna_features
  )

  matched_g2m_features <- match_features_case_insensitive(
    cell_cycle_reference$g2m.genes,
    available_rna_features
  )

  minimum_cell_cycle_features <- 10L

  if (
    length(matched_s_features) < minimum_cell_cycle_features ||
      length(matched_g2m_features) < minimum_cell_cycle_features
  ) {
    stop(
      "Too few mouse-matched cell-cycle genes were available. ",
      "Matched S genes: ",
      length(matched_s_features),
      "; matched G2M genes: ",
      length(matched_g2m_features),
      ".",
      call. = FALSE
    )
  }

  previous_default_assay <- DefaultAssay(multiome)
  DefaultAssay(multiome) <- "RNA"

  multiome <- CellCycleScoring(
    multiome,
    s.features = matched_s_features,
    g2m.features = matched_g2m_features,
    set.ident = FALSE
  )

  DefaultAssay(multiome) <- previous_default_assay

  # Refresh metadata after CellCycleScoring added the columns.
  metadata <- multiome[[]]
  cell_cycle_score_source <- "computed in Step 7B"
}

missing_cell_cycle_columns <- setdiff(
  cell_cycle_columns,
  colnames(metadata)
)

if (length(missing_cell_cycle_columns) > 0L) {
  stop(
    "Cell-cycle scoring did not create the required column(s): ",
    paste(missing_cell_cycle_columns, collapse = ", "),
    ".",
    call. = FALSE
  )
}

cell_cycle_scoring_audit <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  score_source = cell_cycle_score_source,
  Phase_present_before =
    cell_cycle_columns_present_before[[1]],
  S_Score_present_before =
    cell_cycle_columns_present_before[[2]],
  G2M_Score_present_before =
    cell_cycle_columns_present_before[[3]],
  matched_S_features = if (
    identical(cell_cycle_score_source, "computed in Step 7B")
  ) {
    length(matched_s_features)
  } else {
    NA_integer_
  },
  matched_G2M_features = if (
    identical(cell_cycle_score_source, "computed in Step 7B")
  ) {
    length(matched_g2m_features)
  } else {
    NA_integer_
  },
  stringsAsFactors = FALSE
)

readr::write_csv(
  cell_cycle_scoring_audit,
  file.path(
    TRAJECTORY_DIR,
    "trajectory_cell_cycle_scoring_audit.csv"
  )
)

required_metadata <- c(
  "sample_id",
  "diet",
  "wnn_cluster",
  ANALYSIS_UNIT_COLUMN,
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence",
  "nCount_RNA",
  "percent.mt",
  "TSS.enrichment",
  "frip_cr",
  "Phase",
  "S.Score",
  "G2M.Score"
)

missing_metadata <- setdiff(required_metadata, colnames(metadata))

if (length(missing_metadata) > 0L) {
  stop(
    "The Step 6 object is missing trajectory metadata column(s): ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

metadata$cell_barcode <- rownames(metadata)
metadata$sample_id <- as.character(metadata$sample_id)
metadata$diet <- as.character(metadata$diet)
metadata$analysis_unit <- as.character(
  metadata[[ANALYSIS_UNIT_COLUMN]]
)
metadata$wnn_cluster <- as.character(metadata$wnn_cluster)
metadata$broad_compartment <- as.character(
  metadata$broad_compartment
)
metadata$cell_state <- as.character(metadata$cell_state)
metadata$cell_state_full <- as.character(metadata$cell_state_full)
metadata$annotation_confidence <- tolower(
  as.character(metadata$annotation_confidence)
)

expected_diets <- c(REFERENCE_DIET, COMPARISON_DIET)

if (!setequal(unique(metadata$diet), expected_diets)) {
  stop(
    "Expected exactly CON and HFD in the WT object.",
    call. = FALSE
  )
}

rna_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)

rna_data <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "data"
)

metadata_cells <- rownames(metadata)

align_matrix_to_metadata <- function(
    feature_matrix,
    metadata_cells,
    matrix_label
) {
  matrix_cells <- colnames(feature_matrix)

  if (is.null(matrix_cells)) {
    stop(matrix_label, " has no cell names.", call. = FALSE)
  }

  if (anyDuplicated(matrix_cells) > 0L) {
    stop(matrix_label, " contains duplicate cell names.", call. = FALSE)
  }

  if (anyDuplicated(metadata_cells) > 0L) {
    stop(
      "The Step 6 metadata contains duplicate cell names.",
      call. = FALSE
    )
  }

  missing_from_matrix <- setdiff(metadata_cells, matrix_cells)
  extra_in_matrix <- setdiff(matrix_cells, metadata_cells)

  if (
    length(missing_from_matrix) > 0L ||
      length(extra_in_matrix) > 0L
  ) {
    stop(
      matrix_label,
      " and the Step 6 metadata contain different cell sets. ",
      "Missing from matrix: ",
      length(missing_from_matrix),
      "; extra in matrix: ",
      length(extra_in_matrix),
      ".",
      call. = FALSE
    )
  }

  if (!identical(matrix_cells, metadata_cells)) {
    message(
      matrix_label,
      " contains the correct cells in a different order; ",
      "reordering columns to match the Step 6 metadata."
    )

    feature_matrix <- feature_matrix[
      ,
      metadata_cells,
      drop = FALSE
    ]
  }

  if (!identical(colnames(feature_matrix), metadata_cells)) {
    stop(
      "Failed to align ",
      matrix_label,
      " to the Step 6 metadata.",
      call. = FALSE
    )
  }

  feature_matrix
}

trajectory_matrix_alignment_audit <- data.frame(
  matrix = c("RNA counts", "RNA data"),
  matrix_cells = c(ncol(rna_counts), ncol(rna_data)),
  metadata_cells = length(metadata_cells),
  same_cell_set = c(
    setequal(colnames(rna_counts), metadata_cells),
    setequal(colnames(rna_data), metadata_cells)
  ),
  identical_order_before_alignment = c(
    identical(colnames(rna_counts), metadata_cells),
    identical(colnames(rna_data), metadata_cells)
  ),
  stringsAsFactors = FALSE
)

rna_counts <- align_matrix_to_metadata(
  rna_counts,
  metadata_cells,
  "RNA counts"
)

rna_data <- align_matrix_to_metadata(
  rna_data,
  metadata_cells,
  "RNA data"
)

readr::write_csv(
  trajectory_matrix_alignment_audit,
  file.path(
    TRAJECTORY_DIR,
    "trajectory_matrix_cell_alignment_audit.csv"
  )
)

candidate_units <- unique(
  unlist(
    strsplit(
      candidate_trajectories$analysis_units,
      split = ";",
      fixed = TRUE
    )
  )
)

missing_candidate_units <- setdiff(
  candidate_units,
  unique(metadata$analysis_unit)
)

if (length(missing_candidate_units) > 0L) {
  stop(
    "Candidate trajectory unit(s) are absent from the WT object: ",
    paste(missing_candidate_units, collapse = ", "),
    call. = FALSE
  )
}

candidate_annotation <- metadata |>
  dplyr::filter(analysis_unit %in% candidate_units) |>
  dplyr::distinct(
    analysis_unit,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence
  )

if (
  any(candidate_annotation$broad_compartment != "epithelial") ||
    any(candidate_annotation$annotation_confidence == "low")
) {
  stop(
    "Candidate trajectories must contain only high- or moderate-confidence ",
    "epithelial units.",
    call. = FALSE
  )
}
```

# 4. Audit candidate-state representation in both libraries

```{r audit-candidate-state-representation}
candidate_unit_counts <- metadata |>
  dplyr::filter(analysis_unit %in% candidate_units) |>
  dplyr::count(
    analysis_unit,
    cell_state,
    sample_id,
    diet,
    name = "nuclei"
  ) |>
  tidyr::complete(
    nesting(analysis_unit, cell_state),
    nesting(sample_id, diet),
    fill = list(nuclei = 0L)
  )

readr::write_csv(
  candidate_unit_counts,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "candidate_unit_nuclei_by_library.csv"
  )
)

root_count_audit <- candidate_unit_counts |>
  dplyr::filter(analysis_unit == "8") |>
  dplyr::group_by(diet) |>
  dplyr::summarise(
    nuclei = sum(nuclei),
    .groups = "drop"
  )

if (any(root_count_audit$nuclei < MINIMUM_NUCLEI_PER_DIET_FOR_ROOT)) {
  warning(
    "The proposed root has fewer than ",
    MINIMUM_NUCLEI_PER_DIET_FOR_ROOT,
    " nuclei in at least one diet."
  )
}
```

# 5. Quantify lineage-program co-detection and transcriptional mixing

Program scores are the mean log-normalized expression of available genes in
each program. They are used for within-dataset review, not as universal or
batch-corrected scores.

```{r calculate-program-scores}
program_feature_audit <- dplyr::bind_rows(
  lapply(
    names(trajectory_programs),
    function(program_name) {
      genes <- trajectory_programs[[program_name]]

      data.frame(
        program = program_name,
        gene = genes,
        present_in_RNA = genes %in% rownames(rna_counts),
        stringsAsFactors = FALSE
      )
    }
  )
)

readr::write_csv(
  program_feature_audit,
  file.path(TRAJECTORY_TABLE_DIR, "trajectory_program_feature_audit.csv")
)

essential_program_coverage <- program_feature_audit |>
  dplyr::filter(program %in% c("stem", "absorptive", "secretory")) |>
  dplyr::group_by(program) |>
  dplyr::summarise(
    genes_present = sum(present_in_RNA),
    .groups = "drop"
  )

if (any(essential_program_coverage$genes_present < 2L)) {
  stop(
    "At least one essential trajectory program has fewer than two RNA ",
    "features in the active object.",
    call. = FALSE
  )
}

score_cells_for_program <- function(
  normalized_matrix,
  count_matrix,
  genes,
  cells
) {
  matched_genes <- intersect(genes, rownames(normalized_matrix))

  if (length(matched_genes) == 0L) {
    return(data.frame(
      score = rep(NA_real_, length(cells)),
      genes_detected = rep(NA_integer_, length(cells))
    ))
  }

  data.frame(
    score = Matrix::colMeans(
      normalized_matrix[matched_genes, cells, drop = FALSE]
    ),
    genes_detected = Matrix::colSums(
      count_matrix[matched_genes, cells, drop = FALSE] > 0
    )
  )
}

trajectory_metadata <- metadata |>
  dplyr::filter(analysis_unit %in% candidate_units)

for (program_name in names(trajectory_programs)) {
  program_result <- score_cells_for_program(
    normalized_matrix = rna_data,
    count_matrix = rna_counts,
    genes = trajectory_programs[[program_name]],
    cells = trajectory_metadata$cell_barcode
  )

  trajectory_metadata[[paste0(program_name, "_score")]] <-
    program_result$score

  trajectory_metadata[[paste0(
    program_name,
    "_genes_detected"
  )]] <- program_result$genes_detected
}

trajectory_metadata <- trajectory_metadata |>
  dplyr::mutate(
    stem_program_positive = stem_genes_detected >= 2L,
    absorptive_program_positive =
      absorptive_genes_detected >= 2L,
    secretory_program_positive =
      secretory_genes_detected >= 2L,
    absorptive_and_secretory_codetected =
      absorptive_program_positive &
      secretory_program_positive,
    lineage_programs_positive =
      as.integer(stem_program_positive) +
      as.integer(absorptive_program_positive) +
      as.integer(secretory_program_positive),
    multiple_lineage_programs_detected =
      lineage_programs_positive >= 2L
  )

program_score_columns <- grep(
  "_score$",
  colnames(trajectory_metadata),
  value = TRUE
)

program_detection_columns <- grep(
  "_genes_detected$",
  colnames(trajectory_metadata),
  value = TRUE
)

trajectory_program_summary <- trajectory_metadata |>
  dplyr::group_by(
    analysis_unit,
    cell_state,
    sample_id,
    diet
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    dplyr::across(
      dplyr::all_of(program_score_columns),
      ~ stats::median(.x, na.rm = TRUE),
      .names = "median_{.col}"
    ),
    dplyr::across(
      dplyr::all_of(program_detection_columns),
      ~ stats::median(.x, na.rm = TRUE),
      .names = "median_{.col}"
    ),
    stem_program_positive_percent =
      100 * mean(stem_program_positive),
    absorptive_program_positive_percent =
      100 * mean(absorptive_program_positive),
    secretory_program_positive_percent =
      100 * mean(secretory_program_positive),
    absorptive_secretory_codetection_percent =
      100 * mean(absorptive_and_secretory_codetected),
    multiple_lineage_program_percent =
      100 * mean(multiple_lineage_programs_detected),
    median_nCount_RNA = stats::median(nCount_RNA),
    median_percent_mt = stats::median(percent.mt),
    .groups = "drop"
  )

readr::write_csv(
  trajectory_program_summary,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "trajectory_program_summary_by_unit_and_library.csv"
  )
)

readr::write_csv(
  trajectory_metadata |>
    dplyr::select(
      cell_barcode,
      sample_id,
      diet,
      analysis_unit,
      cell_state,
      Phase,
      dplyr::all_of(program_score_columns),
      dplyr::all_of(program_detection_columns),
      multiple_lineage_programs_detected,
      absorptive_and_secretory_codetected
    ),
  file.path(
    TRAJECTORY_TABLE_DIR,
    "trajectory_program_scores_by_nucleus.csv.gz"
  )
)
```

## Cell-cycle and program-mixing summaries

```{r cell-cycle-and-mixing-audit}
candidate_cell_cycle_summary <- trajectory_metadata |>
  dplyr::count(
    analysis_unit,
    cell_state,
    sample_id,
    diet,
    Phase,
    name = "nuclei"
  ) |>
  dplyr::group_by(analysis_unit, sample_id) |>
  dplyr::mutate(
    percent_of_unit_library = 100 * nuclei / sum(nuclei)
  ) |>
  dplyr::ungroup()

readr::write_csv(
  candidate_cell_cycle_summary,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "candidate_unit_cell_cycle_by_library.csv"
  )
)

p_program_mixing <- ggplot2::ggplot(
  trajectory_metadata,
  ggplot2::aes(
    x = absorptive_score,
    y = secretory_score,
    color = analysis_unit
  )
) +
  ggplot2::geom_point(
    size = 0.35,
    alpha = 0.55
  ) +
  ggplot2::facet_wrap(~ diet) +
  ggplot2::labs(
    title = "Absorptive and secretory program mixing",
    subtitle = paste(
      "A smooth distribution can indicate transitional biology,",
      "transcriptional infidelity, or both."
    ),
    x = "Mean absorptive-program expression",
    y = "Mean secretory-program expression",
    color = "Annotation unit"
  ) +
  theme_multiome(base_size = 9)

print(p_program_mixing)

ggplot2::ggsave(
  file.path(
    TRAJECTORY_PLOT_DIR,
    "01_absorptive_secretory_program_mixing.pdf"
  ),
  p_program_mixing,
  width = 11,
  height = 7
)
```

# 6. Recompute a focused RNA embedding for topology review

Cell-cycle scores are retained as metadata and are not regressed from the
primary audit embedding. Removing them could erase a genuine cycling
transit-amplifying state. A cell-cycle-regressed embedding is generated only
as a sensitivity view; Slingshot does not use it unless a later reviewed
decision explicitly changes the model.

```{r focused-rna-topology-audit}
focused_cells <- trajectory_metadata$cell_barcode

focused_audit <- subset(
  multiome,
  cells = focused_cells
)

DefaultAssay(focused_audit) <- "RNA"

focused_audit <- NormalizeData(
  focused_audit,
  verbose = FALSE
)

focused_audit <- FindVariableFeatures(
  focused_audit,
  selection.method = "vst",
  nfeatures = N_VARIABLE_FEATURES,
  verbose = FALSE
)

trajectory_variable_features <- VariableFeatures(focused_audit)

excluded_variable_features <- grepl(
  "^mt-|^Rpl|^Rps",
  trajectory_variable_features
)

trajectory_variable_features <- trajectory_variable_features[
  !excluded_variable_features
]

VariableFeatures(focused_audit) <- trajectory_variable_features

focused_audit <- ScaleData(
  focused_audit,
  features = trajectory_variable_features,
  verbose = FALSE
)

focused_audit <- RunPCA(
  focused_audit,
  features = trajectory_variable_features,
  npcs = N_RNA_PCS,
  reduction.name = "trajectory.audit.pca",
  reduction.key = "trajAuditPC_",
  verbose = FALSE
)

focused_audit <- RunUMAP(
  focused_audit,
  reduction = "trajectory.audit.pca",
  dims = RNA_DIMS_TRAJECTORY,
  reduction.name = "trajectory.audit.umap",
  reduction.key = "trajAuditUMAP_",
  seed.use = 20260904,
  verbose = FALSE
)

focused_audit <- ScaleData(
  focused_audit,
  features = trajectory_variable_features,
  vars.to.regress = c("S.Score", "G2M.Score"),
  verbose = FALSE
)

focused_audit <- RunPCA(
  focused_audit,
  features = trajectory_variable_features,
  npcs = N_RNA_PCS,
  reduction.name = "trajectory.cycle.sensitivity.pca",
  reduction.key = "trajCyclePC_",
  verbose = FALSE
)

focused_audit <- RunUMAP(
  focused_audit,
  reduction = "trajectory.cycle.sensitivity.pca",
  dims = RNA_DIMS_TRAJECTORY,
  reduction.name = "trajectory.cycle.sensitivity.umap",
  reduction.key = "trajCycleUMAP_",
  seed.use = 20260905,
  verbose = FALSE
)

p_audit_unit <- DimPlot(
  focused_audit,
  reduction = "trajectory.audit.umap",
  group.by = ANALYSIS_UNIT_COLUMN,
  label = TRUE,
  repel = TRUE,
  pt.size = 0.35,
  seed = 20260904,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Candidate WT epithelial trajectory units",
    subtitle = "Focused RNA topology; low-confidence units are excluded.",
    color = "Annotation unit"
  ) +
  theme_multiome(base_size = 9)

p_audit_diet <- DimPlot(
  focused_audit,
  reduction = "trajectory.audit.umap",
  group.by = "diet",
  cols = DIET_COLORS,
  pt.size = 0.35,
  seed = 20260904,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Candidate trajectory topology by pooled library",
    subtitle = "Diet and library remain perfectly confounded.",
    color = NULL
  ) +
  theme_multiome(base_size = 9)

p_audit_phase <- DimPlot(
  focused_audit,
  reduction = "trajectory.audit.umap",
  group.by = "Phase",
  cols = c(
    G1 = "#4C78A8",
    S = "#F2CF5B",
    G2M = "#E45756"
  ),
  pt.size = 0.35,
  seed = 20260904,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Cell-cycle support across candidate topology",
    subtitle = "Review whether the proposed paths are phase-dominated.",
    color = NULL
  ) +
  theme_multiome(base_size = 9)

p_audit_cycle_sensitivity <- DimPlot(
  focused_audit,
  reduction = "trajectory.cycle.sensitivity.umap",
  group.by = ANALYSIS_UNIT_COLUMN,
  label = TRUE,
  repel = TRUE,
  pt.size = 0.35,
  seed = 20260905,
  shuffle = TRUE
) +
  ggplot2::labs(
    title = "Cell-cycle-regressed sensitivity topology",
    subtitle = "Sensitivity view only; not the primary trajectory input.",
    color = "Annotation unit"
  ) +
  theme_multiome(base_size = 9)

p_topology_audit <-
  p_audit_unit + p_audit_diet +
  p_audit_phase + p_audit_cycle_sensitivity +
  patchwork::plot_layout(ncol = 2)

print(p_topology_audit)

ggplot2::ggsave(
  file.path(
    TRAJECTORY_PLOT_DIR,
    "02_candidate_RNA_topology_unit_diet_phase.pdf"
  ),
  p_topology_audit,
  width = 14,
  height = 12
)
```

# 7. Audit the proposed root directly

```{r proposed-root-audit}
root_marker_table <- program_feature_audit |>
  dplyr::filter(program %in% c("stem", "crypt"))

root_marker_summary <- list()
root_summary_index <- 1L

for (gene in root_marker_table$gene[root_marker_table$present_in_RNA]) {
  for (analysis_unit_value in candidate_units) {
    for (diet_value in expected_diets) {
      selected_cells <- trajectory_metadata$cell_barcode[
        trajectory_metadata$analysis_unit == analysis_unit_value &
          trajectory_metadata$diet == diet_value
      ]

      if (length(selected_cells) == 0L) {
        next
      }

      root_marker_summary[[root_summary_index]] <- data.frame(
        analysis_unit = analysis_unit_value,
        diet = diet_value,
        gene = gene,
        nuclei = length(selected_cells),
        percent_detected = 100 * mean(
          rna_counts[gene, selected_cells, drop = TRUE] > 0
        ),
        average_log_normalized_expression = mean(
          rna_data[gene, selected_cells, drop = TRUE]
        ),
        stringsAsFactors = FALSE
      )

      root_summary_index <- root_summary_index + 1L
    }
  }
}

root_marker_summary <- dplyr::bind_rows(root_marker_summary) |>
  dplyr::left_join(
    candidate_annotation |>
      dplyr::select(analysis_unit, cell_state),
    by = "analysis_unit"
  )

readr::write_csv(
  root_marker_summary,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "proposed_root_marker_expression.csv"
  )
)

p_root_markers <- ggplot2::ggplot(
  root_marker_summary,
  ggplot2::aes(
    x = analysis_unit,
    y = gene,
    size = percent_detected,
    color = average_log_normalized_expression
  )
) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~ diet) +
  ggplot2::scale_color_gradientn(
    colors = c("#2C7BB6", "#F7F7F7", "#D7191C")
  ) +
  ggplot2::labs(
    title = "Proposed root marker audit",
    subtitle = "Unit 8 must be root-like in both pooled libraries.",
    x = "Annotation unit",
    y = NULL,
    size = "Percent detected",
    color = "Average log-normalized expression"
  ) +
  theme_multiome(base_size = 9)

print(p_root_markers)

ggplot2::ggsave(
  file.path(
    TRAJECTORY_PLOT_DIR,
    "03_proposed_root_marker_audit.pdf"
  ),
  p_root_markers,
  width = 11,
  height = 7
)
```

# STOP HERE FOR MANUAL ROOT AND TOPOLOGY REVIEW

Before continuing, inspect:

1. `candidate_unit_nuclei_by_library.csv`;
2. `trajectory_program_summary_by_unit_and_library.csv`;
3. the absorptive-versus-secretory program-mixing plot;
4. the focused RNA topology colored by annotation unit, diet, and phase;
5. the proposed root marker audit; and
6. whether unit 8 remains the most defensible root in both libraries.

The proposed trajectory table is written to:

```text
07b_trajectories/trajectory_decision_proposed.csv
```

Copy it to the active WT cohort configuration directory as:

```text
config/datasets/v2_resequenced/cohorts/wt/trajectory_decision.csv
```

Complete the audit fields and set `approved=TRUE` only after review. It is
valid to leave one proposed trajectory unapproved or to change its root,
included units, or terminal units based on the audit.

```{r write-ready-for-review-marker}
writeLines(
  c(
    paste("Trajectory review checkpoint created:", format(Sys.time())),
    paste("Candidate nuclei:", nrow(trajectory_metadata)),
    paste("Candidate annotation units:", paste(candidate_units, collapse = ",")),
    paste("Proposed root:", "8"),
    paste("Decision proposal:", TRAJECTORY_PROPOSAL_PATH),
    "Trajectory root and topology approved: FALSE"
  ),
  TRAJECTORY_REVIEW_MARKER
)
```

# 8. Read and validate the trajectory decision

```{r validate-trajectory-decision}
if (!file.exists(TRAJECTORY_DECISION_PATH)) {
  stop(
    "Manual trajectory checkpoint reached. Review the Step 7B audit, copy:\n",
    TRAJECTORY_PROPOSAL_PATH,
    "\nto:\n",
    TRAJECTORY_DECISION_PATH,
    "\nand approve only the trajectories supported by the data.",
    call. = FALSE
  )
}

trajectory_decision <- suppressMessages(
  readr::read_csv(
    TRAJECTORY_DECISION_PATH,
    show_col_types = FALSE,
    progress = FALSE
  )
)

trajectory_decision <- as.data.frame(
  trajectory_decision,
  stringsAsFactors = FALSE
)

required_decision_columns <- c(
  "cohort_id",
  "trajectory_id",
  "analysis_units",
  "root_unit",
  "terminal_units",
  "approved",
  "reviewer",
  "review_date",
  "decision_notes"
)

missing_decision_columns <- setdiff(
  required_decision_columns,
  colnames(trajectory_decision)
)

if (length(missing_decision_columns) > 0L) {
  stop(
    "Trajectory decision table is missing column(s): ",
    paste(missing_decision_columns, collapse = ", "),
    call. = FALSE
  )
}

trajectory_decision$cohort_id <- tolower(trimws(
  as.character(trajectory_decision$cohort_id)
))

trajectory_decision$trajectory_id <- trimws(
  as.character(trajectory_decision$trajectory_id)
)

trajectory_decision$approved <- tolower(trimws(
  as.character(trajectory_decision$approved)
)) %in% c("true", "t", "1", "yes", "y")

if (any(trajectory_decision$cohort_id != COHORT_ID)) {
  stop(
    "Every trajectory decision row must match the active WT cohort.",
    call. = FALSE
  )
}

if (anyDuplicated(trajectory_decision$trajectory_id) > 0L) {
  stop("Trajectory IDs must be unique.", call. = FALSE)
}

approved_trajectories <- trajectory_decision |>
  dplyr::filter(approved)

if (nrow(approved_trajectories) == 0L) {
  stop(
    "No trajectory has been approved. This is a valid biological decision, ",
    "but there is no model to fit.",
    call. = FALSE
  )
}

split_units <- function(value) {
  if (length(value) != 1L || is.na(value)) {
    return(character())
  }

  units <- trimws(unlist(strsplit(value, ";", fixed = TRUE)))
  unique(units[nzchar(units)])
}

for (decision_index in seq_len(nrow(approved_trajectories))) {
  included_units <- split_units(
    approved_trajectories$analysis_units[[decision_index]]
  )
  root_unit <- trimws(
    approved_trajectories$root_unit[[decision_index]]
  )
  terminal_units <- split_units(
    approved_trajectories$terminal_units[[decision_index]]
  )

  if (
    length(included_units) < 2L ||
      !nzchar(root_unit) ||
      length(terminal_units) == 0L
  ) {
    stop(
      "Approved trajectory ",
      approved_trajectories$trajectory_id[[decision_index]],
      " requires at least two included units, one root, and at least one ",
      "terminal unit.",
      call. = FALSE
    )
  }

  if (root_unit %in% terminal_units) {
    stop(
      "The root cannot also be a terminal unit for ",
      approved_trajectories$trajectory_id[[decision_index]],
      ".",
      call. = FALSE
    )
  }

  if (!root_unit %in% included_units) {
    stop(
      "Trajectory root is not included in analysis_units for ",
      approved_trajectories$trajectory_id[[decision_index]],
      ".",
      call. = FALSE
    )
  }

  if (!all(terminal_units %in% included_units)) {
    stop(
      "At least one terminal unit is absent from analysis_units for ",
      approved_trajectories$trajectory_id[[decision_index]],
      ".",
      call. = FALSE
    )
  }

  if (!all(included_units %in% unique(metadata$analysis_unit))) {
    stop(
      "An approved trajectory references an unavailable annotation unit.",
      call. = FALSE
    )
  }

  included_annotation <- metadata |>
    dplyr::filter(analysis_unit %in% included_units) |>
    dplyr::distinct(
      analysis_unit,
      broad_compartment,
      annotation_confidence
    )

  if (
    any(included_annotation$broad_compartment != "epithelial") ||
      any(included_annotation$annotation_confidence == "low")
  ) {
    stop(
      "Approved trajectories may contain only high- or ",
      "moderate-confidence epithelial annotation units.",
      call. = FALSE
    )
  }

  approved_root_counts <- metadata |>
    dplyr::filter(analysis_unit == root_unit) |>
    dplyr::count(diet, name = "nuclei")

  if (
    nrow(approved_root_counts) != length(expected_diets) ||
      any(
        approved_root_counts$nuclei <
          MINIMUM_NUCLEI_PER_DIET_FOR_ROOT
      )
  ) {
    warning(
      "The approved root for ",
      approved_trajectories$trajectory_id[[decision_index]],
      " has fewer than ",
      MINIMUM_NUCLEI_PER_DIET_FOR_ROOT,
      " nuclei in at least one diet."
    )
  }

  audit_values <- trimws(as.character(unlist(
    approved_trajectories[
      decision_index,
      c("reviewer", "review_date", "decision_notes")
    ],
    use.names = FALSE
  )))

  if (any(is.na(audit_values) | !nzchar(audit_values))) {
    stop(
      "Approved trajectories require reviewer, review_date, and notes.",
      call. = FALSE
    )
  }

  review_date <- trimws(as.character(
    approved_trajectories$review_date[[decision_index]]
  ))

  parsed_review_date <- suppressWarnings(
    as.Date(review_date, format = "%Y-%m-%d")
  )

  if (
    !grepl("^[0-9]{4}-[0-9]{2}-[0-9]{2}$", review_date) ||
      is.na(parsed_review_date)
  ) {
    stop(
      "Approved trajectory review_date must use YYYY-MM-DD format.",
      call. = FALSE
    )
  }
}

approved_trajectories
```

# 9. Fit approved Slingshot trajectories

This section requires `SingleCellExperiment`, `S4Vectors`, and `slingshot`.
Install them with Bioconductor before running beyond the manual checkpoint if
they are not already available.

```{r fit-approved-trajectories}
trajectory_packages <- c(
  "SingleCellExperiment",
  "S4Vectors",
  "slingshot"
)

missing_trajectory_packages <- trajectory_packages[
  !vapply(
    trajectory_packages,
    requireNamespace,
    quietly = TRUE,
    FUN.VALUE = logical(1)
  )
]

if (length(missing_trajectory_packages) > 0L) {
  stop(
    "Install the following trajectory package(s) before continuing: ",
    paste(missing_trajectory_packages, collapse = ", "),
    ". Use BiocManager::install(c(",
    paste0('"', missing_trajectory_packages, '"', collapse = ", "),
    ")).",
    call. = FALSE
  )
}

trajectory_model_paths <- character()
trajectory_assignment_tables <- list()
trajectory_summary_tables <- list()
trajectory_trend_tables <- list()
trajectory_lineage_tables <- list()

approved_trajectory_units <- unique(
  unlist(
    lapply(
      approved_trajectories$analysis_units,
      split_units
    )
  )
)

trajectory_annotation_lookup <- metadata |>
  dplyr::filter(analysis_unit %in% approved_trajectory_units) |>
  dplyr::distinct(
    analysis_unit,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence
  )

duplicate_trajectory_annotations <- trajectory_annotation_lookup |>
  dplyr::count(analysis_unit, name = "annotation_rows") |>
  dplyr::filter(annotation_rows != 1L)

if (nrow(duplicate_trajectory_annotations) > 0L) {
  stop(
    "At least one approved trajectory unit maps to multiple annotations.",
    call. = FALSE
  )
}

trajectory_annotation_key <- metadata |>
  dplyr::filter(analysis_unit %in% approved_trajectory_units) |>
  dplyr::count(
    analysis_unit,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence,
    diet,
    name = "nuclei"
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = nuclei,
    names_prefix = "nuclei_",
    values_fill = 0L
  ) |>
  dplyr::arrange(as.integer(analysis_unit))

lookup_annotation <- function(unit_values, annotation_column) {
  matched_values <- trajectory_annotation_lookup[[annotation_column]][
    match(
      as.character(unit_values),
      trajectory_annotation_lookup$analysis_unit
    )
  ]

  if (anyNA(matched_values)) {
    stop(
      "Could not map every trajectory unit to ",
      annotation_column,
      ".",
      call. = FALSE
    )
  }

  as.character(matched_values)
}

annotate_unit_path <- function(unit_path, annotation_column) {
  path_units <- split_units(unit_path)
  path_labels <- lookup_annotation(path_units, annotation_column)

  paste(
    paste0(path_units, " (", path_labels, ")"),
    collapse = " -> "
  )
}

wnn_context_available <- "wnn.umap" %in% names(multiome@reductions)

if (wnn_context_available) {
  epithelial_wnn_background <- as.data.frame(
    Embeddings(multiome, "wnn.umap")[, 1:2, drop = FALSE]
  )
  colnames(epithelial_wnn_background) <- c("wnn_1", "wnn_2")
  epithelial_wnn_background$cell_barcode <- rownames(
    epithelial_wnn_background
  )

  epithelial_wnn_background <- epithelial_wnn_background |>
    dplyr::inner_join(
      metadata |>
        dplyr::filter(broad_compartment == "epithelial") |>
        dplyr::select(
          cell_barcode,
          analysis_unit,
          cell_state,
          cell_state_full
        ),
      by = "cell_barcode"
    )
} else {
  warning(
    "The Step 6 object lacks wnn.umap; epithelial WNN context plots will ",
    "be skipped. Slingshot fitting and RNA trajectory plots are unaffected."
  )
}

step7_rna_candidates_path <- file.path(
  STEP7_DIR,
  "tables",
  "RNA_top_diet_associated_candidates.csv"
)

step7_rna_candidates <- if (file.exists(step7_rna_candidates_path)) {
  suppressMessages(
    readr::read_csv(
      step7_rna_candidates_path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
} else {
  data.frame()
}

for (decision_index in seq_len(nrow(approved_trajectories))) {
  trajectory_id <- approved_trajectories$trajectory_id[[decision_index]]
  included_units <- split_units(
    approved_trajectories$analysis_units[[decision_index]]
  )
  root_unit <- trimws(
    approved_trajectories$root_unit[[decision_index]]
  )
  terminal_units <- split_units(
    approved_trajectories$terminal_units[[decision_index]]
  )

  trajectory_cells <- metadata$cell_barcode[
    metadata$analysis_unit %in% included_units
  ]

  trajectory_object <- subset(
    multiome,
    cells = trajectory_cells
  )

  DefaultAssay(trajectory_object) <- "RNA"

  trajectory_object <- NormalizeData(
    trajectory_object,
    verbose = FALSE
  )

  trajectory_object <- FindVariableFeatures(
    trajectory_object,
    selection.method = "vst",
    nfeatures = N_VARIABLE_FEATURES,
    verbose = FALSE
  )

  trajectory_features <- VariableFeatures(trajectory_object)
  trajectory_features <- trajectory_features[
    !grepl("^mt-|^Rpl|^Rps", trajectory_features)
  ]

  VariableFeatures(trajectory_object) <- trajectory_features

  trajectory_object <- ScaleData(
    trajectory_object,
    features = trajectory_features,
    verbose = FALSE
  )

  trajectory_object <- RunPCA(
    trajectory_object,
    features = trajectory_features,
    npcs = N_RNA_PCS,
    reduction.name = "trajectory.pca",
    reduction.key = "trajPC_",
    verbose = FALSE
  )

  trajectory_object <- RunUMAP(
    trajectory_object,
    reduction = "trajectory.pca",
    dims = RNA_DIMS_TRAJECTORY,
    reduction.name = "trajectory.umap",
    reduction.key = "trajUMAP_",
    seed.use = 20260904 + decision_index,
    verbose = FALSE
  )

  trajectory_full_log_data <- SeuratObject::LayerData(
    trajectory_object[["RNA"]],
    layer = "data"
  )

  trajectory_log_data <- trajectory_full_log_data[
    trajectory_features,
    ,
    drop = FALSE
  ]

  trajectory_coldata <- trajectory_object[[]]
  trajectory_coldata$analysis_unit <- factor(
    as.character(
      trajectory_coldata[[ANALYSIS_UNIT_COLUMN]]
    ),
    levels = included_units
  )

  trajectory_sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(logcounts = trajectory_log_data),
    colData = S4Vectors::DataFrame(trajectory_coldata),
    reducedDims = list(
      PCA = Embeddings(trajectory_object, "trajectory.pca")
    )
  )

  trajectory_sce <- slingshot::slingshot(
    trajectory_sce,
    clusterLabels = "analysis_unit",
    reducedDim = "PCA",
    start.clus = root_unit,
    end.clus = terminal_units
  )

  pseudotime_matrix <- slingshot::slingPseudotime(
    trajectory_sce,
    na = TRUE
  )

  curve_weight_matrix <- slingshot::slingCurveWeights(
    trajectory_sce
  )

  inferred_lineages <- slingshot::slingLineages(
    trajectory_sce
  )

  if (
    length(inferred_lineages) != ncol(pseudotime_matrix) ||
      ncol(curve_weight_matrix) != ncol(pseudotime_matrix)
  ) {
    stop(
      "Slingshot lineage, pseudotime, and curve-weight counts disagree for ",
      trajectory_id,
      ".",
      call. = FALSE
    )
  }

  lineage_ids <- names(inferred_lineages)

  if (
    is.null(lineage_ids) ||
      any(is.na(lineage_ids) | !nzchar(lineage_ids))
  ) {
    lineage_ids <- paste0(
      "Lineage",
      seq_along(inferred_lineages)
    )
  }

  lineage_mapping <- data.frame(
    trajectory_id = trajectory_id,
    lineage_index = seq_along(inferred_lineages),
    lineage_id = lineage_ids,
    cluster_path = vapply(
      inferred_lineages,
      function(lineage_path) {
        paste(as.character(lineage_path), collapse = ";")
      },
      FUN.VALUE = character(1)
    ),
    terminal_unit = vapply(
      inferred_lineages,
      function(lineage_path) {
        tail(as.character(lineage_path), 1L)
      },
      FUN.VALUE = character(1)
    ),
    stringsAsFactors = FALSE
  )

  lineage_mapping$root_unit <- root_unit
  lineage_mapping$root_cell_state <- lookup_annotation(
    root_unit,
    "cell_state"
  )
  lineage_mapping$terminal_cell_state <- lookup_annotation(
    lineage_mapping$terminal_unit,
    "cell_state"
  )
  lineage_mapping$cell_state_path <- vapply(
    lineage_mapping$cluster_path,
    annotate_unit_path,
    annotation_column = "cell_state",
    FUN.VALUE = character(1)
  )
  lineage_mapping$cell_state_full_path <- vapply(
    lineage_mapping$cluster_path,
    annotate_unit_path,
    annotation_column = "cell_state_full",
    FUN.VALUE = character(1)
  )

  trajectory_lineage_tables[[trajectory_id]] <- lineage_mapping

  inferred_terminal_units <- unique(
    lineage_mapping$terminal_unit
  )

  if (!setequal(inferred_terminal_units, terminal_units)) {
    stop(
      "Slingshot inferred terminal unit(s) ",
      paste(inferred_terminal_units, collapse = ", "),
      " for ",
      trajectory_id,
      ", but the approved terminal unit(s) are ",
      paste(terminal_units, collapse = ", "),
      ". Revise the included units or decision table before interpreting ",
      "pseudotime.",
      call. = FALSE
    )
  }

  assignment_table <- data.frame(
    cell_barcode = colnames(trajectory_sce),
    trajectory_id = trajectory_id,
    sample_id = as.character(trajectory_coldata$sample_id),
    diet = as.character(trajectory_coldata$diet),
    analysis_unit = as.character(
      trajectory_coldata$analysis_unit
    ),
    broad_compartment = as.character(
      trajectory_coldata$broad_compartment
    ),
    cell_state = as.character(trajectory_coldata$cell_state),
    cell_state_full = as.character(
      trajectory_coldata$cell_state_full
    ),
    annotation_confidence = as.character(
      trajectory_coldata$annotation_confidence
    ),
    Phase = as.character(trajectory_coldata$Phase),
    stringsAsFactors = FALSE
  )

  pseudotime_names <- paste0(
    "pseudotime_",
    seq_len(ncol(pseudotime_matrix))
  )

  weight_names <- paste0(
    "curve_weight_",
    seq_len(ncol(curve_weight_matrix))
  )

  colnames(pseudotime_matrix) <- pseudotime_names
  colnames(curve_weight_matrix) <- weight_names

  assignment_table <- cbind(
    assignment_table,
    as.data.frame(pseudotime_matrix),
    as.data.frame(curve_weight_matrix)
  )

  trajectory_assignment_tables[[trajectory_id]] <- assignment_table

  pseudotime_long <- assignment_table |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(pseudotime_names),
      names_to = "pseudotime_column",
      values_to = "pseudotime"
    ) |>
    dplyr::mutate(
      lineage_index = as.integer(sub(
        "pseudotime_",
        "",
        pseudotime_column
      ))
    )

  curve_weight_long <- assignment_table |>
    dplyr::select(
      cell_barcode,
      dplyr::all_of(weight_names)
    ) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(weight_names),
      names_to = "weight_column",
      values_to = "curve_weight"
    ) |>
    dplyr::mutate(
      lineage_index = as.integer(sub(
        "curve_weight_",
        "",
        weight_column
      ))
    ) |>
    dplyr::select(
      cell_barcode,
      lineage_index,
      curve_weight
    )

  assignment_long <- pseudotime_long |>
    dplyr::left_join(
      curve_weight_long,
      by = c("cell_barcode", "lineage_index")
    ) |>
    dplyr::left_join(
      lineage_mapping,
      by = c("trajectory_id", "lineage_index")
    )

  trajectory_summary_tables[[trajectory_id]] <- assignment_long |>
    dplyr::filter(
      !is.na(pseudotime),
      curve_weight >= MINIMUM_CURVE_WEIGHT
    ) |>
    dplyr::group_by(
      trajectory_id,
      lineage_id,
      terminal_unit,
      analysis_unit,
      cell_state,
      cell_state_full,
      diet
    ) |>
    dplyr::summarise(
      nuclei = dplyr::n(),
      median_pseudotime = stats::median(pseudotime),
      pseudotime_q25 = stats::quantile(pseudotime, 0.25),
      pseudotime_q75 = stats::quantile(pseudotime, 0.75),
      .groups = "drop"
    )

  trajectory_umap <- as.data.frame(
    Embeddings(trajectory_object, "trajectory.umap")
  )
  trajectory_umap$cell_barcode <- rownames(trajectory_umap)

  trajectory_state_plot_data <- assignment_table |>
    dplyr::select(
      cell_barcode,
      analysis_unit,
      cell_state,
      cell_state_full
    ) |>
    dplyr::distinct() |>
    dplyr::left_join(
      trajectory_umap,
      by = "cell_barcode"
    ) |>
    dplyr::mutate(
      state_display = paste0(analysis_unit, ": ", cell_state)
    )

  trajectory_state_centroids <- trajectory_state_plot_data |>
    dplyr::group_by(analysis_unit, cell_state, state_display) |>
    dplyr::summarise(
      trajUMAP_1 = stats::median(trajUMAP_1),
      trajUMAP_2 = stats::median(trajUMAP_2),
      .groups = "drop"
    )

  p_trajectory_states <- ggplot2::ggplot(
    trajectory_state_plot_data,
    ggplot2::aes(
      x = trajUMAP_1,
      y = trajUMAP_2,
      color = state_display
    )
  ) +
    ggplot2::geom_point(size = 0.35, alpha = 0.65) +
    ggplot2::geom_label(
      data = trajectory_state_centroids,
      ggplot2::aes(
        x = trajUMAP_1,
        y = trajUMAP_2,
        label = state_display
      ),
      inherit.aes = FALSE,
      size = 2.5,
      label.size = 0.15,
      fill = "white",
      alpha = 0.90
    ) +
    ggplot2::labs(
      title = paste("Annotated RNA topology:", trajectory_id),
      subtitle = paste(
        "State labels come from Step 6; this UMAP is a display of the",
        "trajectory-specific RNA PCA neighborhood."
      ),
      x = "Trajectory UMAP 1",
      y = "Trajectory UMAP 2",
      color = "Annotation unit"
    ) +
    theme_multiome(base_size = 9) +
    ggplot2::theme(legend.position = "none")

  pseudotime_plot_data <- assignment_long |>
    dplyr::left_join(
      trajectory_umap,
      by = "cell_barcode"
    )

  p_trajectory_pseudotime <- ggplot2::ggplot(
    pseudotime_plot_data,
    ggplot2::aes(
      x = trajUMAP_1,
      y = trajUMAP_2,
      color = pseudotime
    )
  ) +
    ggplot2::geom_point(size = 0.35) +
    ggplot2::facet_wrap(~ lineage_id) +
    ggplot2::scale_color_viridis_c(na.value = "grey85") +
    ggplot2::labs(
      title = paste("Descriptive RNA pseudotime:", trajectory_id),
      subtitle = paste(
        "Slingshot fitted in RNA PCA space; root and endpoints",
        "were manually approved."
      ),
      x = "Trajectory UMAP 1",
      y = "Trajectory UMAP 2",
      color = "Pseudotime"
    ) +
    theme_multiome(base_size = 9)

  print(p_trajectory_pseudotime)

  ggplot2::ggsave(
    file.path(
      TRAJECTORY_PLOT_DIR,
      paste0("04_", trajectory_id, "_pseudotime.pdf")
    ),
    p_trajectory_pseudotime,
    width = 11,
    height = 7
  )

  p_annotated_trajectory_overview <-
    p_trajectory_states /
    p_trajectory_pseudotime +
    patchwork::plot_layout(heights = c(0.9, 1.1))

  print(p_annotated_trajectory_overview)

  ggplot2::ggsave(
    file.path(
      TRAJECTORY_PLOT_DIR,
      paste0("05_", trajectory_id, "_annotated_trajectory_overview.pdf")
    ),
    p_annotated_trajectory_overview,
    width = 12,
    height = 12
  )

  trajectory_pca <- as.data.frame(
    Embeddings(trajectory_object, "trajectory.pca")[, 1:2, drop = FALSE]
  )
  colnames(trajectory_pca) <- c("trajectory_PC1", "trajectory_PC2")
  trajectory_pca$cell_barcode <- rownames(trajectory_pca)

  trajectory_pca_plot_data <- assignment_table |>
    dplyr::select(
      cell_barcode,
      analysis_unit,
      cell_state,
      cell_state_full
    ) |>
    dplyr::distinct() |>
    dplyr::left_join(
      trajectory_pca,
      by = "cell_barcode"
    ) |>
    dplyr::mutate(
      state_display = paste0(analysis_unit, ": ", cell_state)
    )

  trajectory_pca_centroids <- trajectory_pca_plot_data |>
    dplyr::group_by(analysis_unit, cell_state, state_display) |>
    dplyr::summarise(
      trajectory_PC1 = stats::median(trajectory_PC1),
      trajectory_PC2 = stats::median(trajectory_PC2),
      .groups = "drop"
    )

  fitted_curves <- slingshot::slingCurves(trajectory_sce)

  curve_plot_data <- dplyr::bind_rows(
    lapply(
      seq_along(fitted_curves),
      function(lineage_index) {
        fitted_curve <- fitted_curves[[lineage_index]]
        curve_coordinates <- fitted_curve$s[
          fitted_curve$ord,
          1:2,
          drop = FALSE
        ]

        data.frame(
          lineage_id = lineage_ids[[lineage_index]],
          curve_order = seq_len(nrow(curve_coordinates)),
          trajectory_PC1 = curve_coordinates[, 1],
          trajectory_PC2 = curve_coordinates[, 2],
          stringsAsFactors = FALSE
        )
      }
    )
  )

  p_slingshot_curves <- ggplot2::ggplot(
    trajectory_pca_plot_data,
    ggplot2::aes(
      x = trajectory_PC1,
      y = trajectory_PC2,
      color = state_display
    )
  ) +
    ggplot2::geom_point(size = 0.30, alpha = 0.45) +
    ggplot2::geom_path(
      data = curve_plot_data,
      ggplot2::aes(
        x = trajectory_PC1,
        y = trajectory_PC2,
        group = lineage_id
      ),
      inherit.aes = FALSE,
      color = "black",
      linewidth = 0.8,
      arrow = grid::arrow(
        length = grid::unit(0.10, "inches"),
        type = "closed"
      )
    ) +
    ggplot2::geom_label(
      data = trajectory_pca_centroids,
      ggplot2::aes(
        x = trajectory_PC1,
        y = trajectory_PC2,
        label = state_display
      ),
      inherit.aes = FALSE,
      size = 2.5,
      label.size = 0.15,
      fill = "white",
      alpha = 0.90
    ) +
    ggplot2::labs(
      title = paste("Slingshot RNA-PCA curves:", trajectory_id),
      subtitle = paste(
        "Arrows project the fitted 20-dimensional RNA-PCA curves onto",
        "PC1/PC2 from the approved root toward the terminal state(s)."
      ),
      x = "Trajectory PC 1",
      y = "Trajectory PC 2",
      color = "Annotation unit"
    ) +
    theme_multiome(base_size = 9) +
    ggplot2::theme(legend.position = "none")

  print(p_slingshot_curves)

  ggplot2::ggsave(
    file.path(
      TRAJECTORY_PLOT_DIR,
      paste0("06_", trajectory_id, "_RNA_PCA_slingshot_curves.pdf")
    ),
    p_slingshot_curves,
    width = 10,
    height = 8
  )

  if (wnn_context_available) {
    trajectory_wnn_coordinates <- epithelial_wnn_background |>
      dplyr::filter(cell_barcode %in% trajectory_cells) |>
      dplyr::select(cell_barcode, wnn_1, wnn_2)

    wnn_pseudotime_plot_data <- assignment_long |>
      dplyr::filter(
        !is.na(pseudotime),
        curve_weight >= MINIMUM_CURVE_WEIGHT
      ) |>
      dplyr::left_join(
        trajectory_wnn_coordinates,
        by = "cell_barcode"
      )

    trajectory_wnn_centroids <- trajectory_state_plot_data |>
      dplyr::select(
        cell_barcode,
        analysis_unit,
        cell_state,
        state_display
      ) |>
      dplyr::left_join(
        trajectory_wnn_coordinates,
        by = "cell_barcode"
      ) |>
      dplyr::group_by(analysis_unit, cell_state, state_display) |>
      dplyr::summarise(
        wnn_1 = stats::median(wnn_1),
        wnn_2 = stats::median(wnn_2),
        .groups = "drop"
      )

    p_wnn_pseudotime_context <- ggplot2::ggplot() +
      ggplot2::geom_point(
        data = epithelial_wnn_background,
        ggplot2::aes(x = wnn_1, y = wnn_2),
        color = "grey88",
        size = 0.20
      ) +
      ggplot2::geom_point(
        data = wnn_pseudotime_plot_data,
        ggplot2::aes(
          x = wnn_1,
          y = wnn_2,
          color = pseudotime
        ),
        size = 0.35
      ) +
      ggplot2::geom_label(
        data = trajectory_wnn_centroids,
        ggplot2::aes(
          x = wnn_1,
          y = wnn_2,
          label = state_display
        ),
        inherit.aes = FALSE,
        size = 2.2,
        label.size = 0.12,
        fill = "white",
        alpha = 0.88
      ) +
      ggplot2::facet_wrap(~ lineage_id) +
      ggplot2::scale_color_viridis_c() +
      ggplot2::labs(
        title = paste("WT epithelial WNN context:", trajectory_id),
        subtitle = paste(
          "Pseudotime was inferred in trajectory-specific RNA PCA space;",
          "WNN coordinates are shown only for multimodal anatomical context."
        ),
        x = "WNN UMAP 1",
        y = "WNN UMAP 2",
        color = "Pseudotime"
      ) +
      theme_multiome(base_size = 9)

    print(p_wnn_pseudotime_context)

    ggplot2::ggsave(
      file.path(
        TRAJECTORY_PLOT_DIR,
        paste0("07_", trajectory_id, "_epithelial_WNN_context.pdf")
      ),
      p_wnn_pseudotime_context,
      width = 12,
      height = 8
    )
  }

  pseudotime_density_data <- assignment_long |>
    dplyr::filter(
      !is.na(pseudotime),
      curve_weight >= MINIMUM_CURVE_WEIGHT
    )

  p_pseudotime_by_diet <- ggplot2::ggplot(
    pseudotime_density_data,
    ggplot2::aes(
      x = pseudotime,
      color = diet,
      fill = diet
    )
  ) +
    ggplot2::geom_density(alpha = 0.18, linewidth = 0.7) +
    ggplot2::facet_wrap(~ lineage_id, scales = "free_x") +
    ggplot2::scale_color_manual(values = DIET_COLORS) +
    ggplot2::scale_fill_manual(values = DIET_COLORS) +
    ggplot2::labs(
      title = paste("Descriptive pseudotime occupancy:", trajectory_id),
      subtitle = paste(
        "One pooled library per diet; density differences describe library",
        "occupancy and are not replicated diet-effect estimates."
      ),
      x = "Pseudotime",
      y = "Density",
      color = NULL,
      fill = NULL
    ) +
    theme_multiome(base_size = 9)

  print(p_pseudotime_by_diet)

  ggplot2::ggsave(
    file.path(
      TRAJECTORY_PLOT_DIR,
      paste0("08_", trajectory_id, "_pseudotime_by_diet.pdf")
    ),
    p_pseudotime_by_diet,
    width = 10,
    height = 7
  )

  trend_genes <- unique(unlist(trajectory_programs))

  if (nrow(step7_rna_candidates) > 0L) {
    diet_candidate_genes <- step7_rna_candidates |>
      dplyr::filter(analysis_unit %in% included_units) |>
      dplyr::group_by(analysis_unit) |>
      dplyr::slice_max(
        order_by = abs(log2FC_comparison_vs_reference),
        n = 25L,
        with_ties = FALSE
      ) |>
      dplyr::ungroup() |>
      dplyr::pull(feature)

    trend_genes <- unique(c(trend_genes, diet_candidate_genes))
  }

  trend_genes <- intersect(
    trend_genes,
    rownames(trajectory_full_log_data)
  )

  trend_results <- list()
  trend_index <- 1L

  for (lineage_name in unique(assignment_long$lineage_id)) {
    lineage_cells <- assignment_long |>
      dplyr::filter(
        lineage_id == lineage_name,
        !is.na(pseudotime),
        curve_weight >= MINIMUM_CURVE_WEIGHT
      ) |>
      dplyr::group_by(diet) |>
      dplyr::mutate(
        pseudotime_bin = dplyr::ntile(
          pseudotime,
          N_PSEUDOTIME_BINS
        )
      ) |>
      dplyr::ungroup()

    for (diet_value in expected_diets) {
      for (bin_value in seq_len(N_PSEUDOTIME_BINS)) {
        selected_cells <- lineage_cells$cell_barcode[
          lineage_cells$diet == diet_value &
            lineage_cells$pseudotime_bin == bin_value
        ]

        if (length(selected_cells) == 0L) {
          next
        }

        mean_expression <- Matrix::rowMeans(
          trajectory_full_log_data[
            trend_genes,
            selected_cells,
            drop = FALSE
          ]
        )

        trend_results[[trend_index]] <- data.frame(
          trajectory_id = trajectory_id,
          lineage_id = lineage_name,
          terminal_unit = lineage_cells$terminal_unit[[1]],
          diet = diet_value,
          pseudotime_bin = bin_value,
          nuclei = length(selected_cells),
          median_pseudotime = stats::median(
            lineage_cells$pseudotime[
              lineage_cells$diet == diet_value &
                lineage_cells$pseudotime_bin == bin_value
            ]
          ),
          gene = trend_genes,
          mean_log_normalized_expression = mean_expression,
          stringsAsFactors = FALSE
        )

        trend_index <- trend_index + 1L
      }
    }
  }

  trajectory_trend_tables[[trajectory_id]] <- dplyr::bind_rows(
    trend_results
  )

  model_path <- file.path(
    TRAJECTORY_DIR,
    paste0(trajectory_id, "_slingshot_model.rds")
  )

  saveRDS(
    trajectory_sce,
    model_path,
    compress = FALSE
  )

  trajectory_model_paths <- c(
    trajectory_model_paths,
    model_path
  )
}
```

# 10. Save trajectory assignments and summaries

```{r save-trajectory-outputs}
trajectory_assignments <- dplyr::bind_rows(
  trajectory_assignment_tables
)

trajectory_pseudotime_summary <- dplyr::bind_rows(
  trajectory_summary_tables
)

trajectory_gene_trends <- dplyr::bind_rows(
  trajectory_trend_tables
)

trajectory_lineage_definitions <- dplyr::bind_rows(
  trajectory_lineage_tables
)

readr::write_csv(
  trajectory_assignments,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "approved_trajectory_pseudotime_assignments.csv.gz"
  )
)

readr::write_csv(
  trajectory_pseudotime_summary,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "pseudotime_summary_by_unit_and_diet.csv"
  )
)

readr::write_csv(
  trajectory_gene_trends,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "descriptive_gene_trends_by_pseudotime_and_diet.csv.gz"
  )
)

readr::write_csv(
  trajectory_lineage_definitions,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "trajectory_lineage_definitions.csv"
  )
)

readr::write_csv(
  trajectory_annotation_key,
  file.path(
    TRAJECTORY_TABLE_DIR,
    "trajectory_annotation_key.csv"
  )
)

readr::write_csv(
  approved_trajectories,
  file.path(
    TRAJECTORY_DIR,
    "trajectory_decision_applied.csv"
  )
)

writeLines(
  c(
    paste("Step 7B completed:", format(Sys.time())),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste(
      "Approved trajectories:",
      paste(approved_trajectories$trajectory_id, collapse = ",")
    ),
    paste(
      "Proposed root annotation unit:",
      paste(unique(approved_trajectories$root_unit), collapse = ",")
    ),
    paste(
      "Slingshot model files:",
      paste(basename(trajectory_model_paths), collapse = ",")
    ),
    paste(
      "Annotated lineage definitions:",
      file.path(
        TRAJECTORY_TABLE_DIR,
        "trajectory_lineage_definitions.csv"
      )
    ),
    paste(
      "Trajectory annotation key:",
      file.path(
        TRAJECTORY_TABLE_DIR,
        "trajectory_annotation_key.csv"
      )
    ),
    "Cell-cycle scores regressed from primary topology: FALSE",
    "Biological-replicate diet inference performed: FALSE",
    paste(
      "Interpretation:",
      "exploratory RNA trajectories and pooled-library trends"
    )
  ),
  TRAJECTORY_COMPLETE_MARKER
)
```

# Interpretation checkpoint

For each approved trajectory, answer:

1. Is the proposed root supported by stem/crypt markers in both libraries?
2. Does the inferred ordering agree with the approved biological labels?
3. Is topology dominated by S/G2M separation rather than lineage identity?
4. Where are absorptive and secretory programs co-detected, and is that pattern
   compatible with a transition or more consistent with d21 transcriptional
   infidelity?
5. Do CON and HFD occupy different pseudotime ranges because of composition,
   or do expression trends differ within overlapping pseudotime?
6. Which trajectory-associated RNA genes also have stable Step 7 ATAC effects?
7. Would a later timepoint or lineage-tracing experiment be required before
   assigning developmental direction?

The fitted arrows are displayed only in trajectory-specific RNA PCA space,
where Slingshot was fit. The epithelial WNN panels are contextual overlays and
must not be interpreted as independently fitted WNN trajectories.

# What comes next?

The WT Step 7 and approved Step 7B results provide a descriptive baseline for
matching biological states across WT, PPAR, and IL17. Cross-genotype work
should compare marker-supported states and effect directions, not cluster
numbers or unqualified pseudotime values.
