---
title: "Step 7C - Prioritize WT diet-associated RNA and ATAC candidates"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 7 produced descriptive HFD-versus-CON RNA and ATAC effect sizes within
reviewed WT annotation units. Step 7B added exploratory epithelial RNA
trajectories. Step 7C integrates those outputs into transparent review lists
for experimental follow-up and later state-matched comparison with PPAR and
IL17.

The notebook creates four distinct products:

1. RNA candidates ranked by effect size, detection change, and balanced
   cell-sampling stability;
2. ATAC candidates ranked by effect size, stability, and nearest-gene context;
3. directionally concordant RNA/ATAC candidate pairs; and
4. trajectory-aware RNA candidates whose annotation unit lies on an approved
   epithelial lineage.

These tables are intentionally called *candidate* tables rather than formal
DEG or DAR tables. There is one pooled CON library and one pooled HFD library,
so diet and library are perfectly confounded. Balanced downsampling measures
cell-sampling stability; it does not create biological replication.

# 1. Setup

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
  library(patchwork)
})

ggplot2::theme_set(theme_multiome())

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 7C currently defines the WT d21 benchmark. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}
```

# 2. Define inputs, outputs, and support rules

The support labels below are review aids, not statistical significance tiers.
An interval that excludes zero is the interquartile range across repeated
balanced cell samples, not a biological confidence interval.

```{r define-inputs-and-parameters}
STEP7_DIR <- file.path(OUTPUT_DIR, "07_diet_response")
STEP7_TABLE_DIR <- file.path(STEP7_DIR, "tables")
STEP7B_DIR <- file.path(OUTPUT_DIR, "07b_trajectories")
STEP7B_TABLE_DIR <- file.path(STEP7B_DIR, "tables")

STEP7C_DIR <- file.path(OUTPUT_DIR, "07c_candidate_prioritization")
STEP7C_TABLE_DIR <- file.path(STEP7C_DIR, "tables")
STEP7C_PLOT_DIR <- file.path(STEP7C_DIR, "plots")

dir.create(STEP7C_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP7C_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP7C_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

STEP7_COMPLETE_MARKER <- file.path(STEP7_DIR, "STEP_07_COMPLETE.txt")
STEP7B_COMPLETE_MARKER <- file.path(STEP7B_DIR, "STEP_07B_COMPLETE.txt")
STEP7C_COMPLETE_MARKER <- file.path(STEP7C_DIR, "STEP_07C_COMPLETE.txt")

input_paths <- c(
  eligibility = file.path(
    STEP7_TABLE_DIR,
    "analysis_population_eligibility.csv"
  ),
  RNA_candidates = file.path(
    STEP7_TABLE_DIR,
    "RNA_top_diet_associated_candidates.csv"
  ),
  RNA_stability = file.path(
    STEP7_TABLE_DIR,
    "RNA_balanced_downsampling_stability.csv"
  ),
  ATAC_candidates = file.path(
    STEP7_TABLE_DIR,
    "ATAC_top_candidates_with_nearest_gene.csv"
  ),
  ATAC_stability = file.path(
    STEP7_TABLE_DIR,
    "ATAC_balanced_downsampling_stability.csv"
  ),
  RNA_ATAC_concordance = file.path(
    STEP7_TABLE_DIR,
    "RNA_ATAC_nearest_gene_concordance.csv"
  ),
  lineage_definitions = file.path(
    STEP7B_TABLE_DIR,
    "trajectory_lineage_definitions.csv"
  ),
  pseudotime_summary = file.path(
    STEP7B_TABLE_DIR,
    "pseudotime_summary_by_unit_and_diet.csv"
  ),
  trajectory_gene_trends = file.path(
    STEP7B_TABLE_DIR,
    "descriptive_gene_trends_by_pseudotime_and_diet.csv.gz"
  )
)

assert_file(STEP7_COMPLETE_MARKER, "Step 7 completion marker")
assert_file(STEP7B_COMPLETE_MARKER, "Step 7B completion marker")

for (input_label in names(input_paths)) {
  assert_file(input_paths[[input_label]], input_label)
}

REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"

MINIMUM_SIGN_CONCORDANCE_PERCENT <- 80
NEAR_GENE_DISTANCE_BP <- 10000
MINIMUM_PAIRED_PSEUDOTIME_BINS <- 5L
TOP_RNA_CANDIDATES_PER_UNIT_AND_DIRECTION <- 25L
TOP_ATAC_CANDIDATES_PER_UNIT_AND_DIRECTION <- 50L
TOP_MULTIMODAL_PAIRS_PER_UNIT_AND_DIRECTION <- 25L
TOP_TRAJECTORY_CANDIDATES_PER_LINEAGE <- 30L

prioritization_parameters <- data.frame(
  parameter = c(
    "reference diet",
    "comparison diet",
    "minimum balanced-sampling sign concordance percent",
    "near-gene distance threshold in bp",
    "minimum paired pseudotime bins",
    "top RNA candidates per unit and direction",
    "top ATAC candidates per unit and direction",
    "top multimodal pairs per unit and direction",
    "top trajectory candidates per lineage"
  ),
  value = c(
    REFERENCE_DIET,
    COMPARISON_DIET,
    MINIMUM_SIGN_CONCORDANCE_PERCENT,
    NEAR_GENE_DISTANCE_BP,
    MINIMUM_PAIRED_PSEUDOTIME_BINS,
    TOP_RNA_CANDIDATES_PER_UNIT_AND_DIRECTION,
    TOP_ATAC_CANDIDATES_PER_UNIT_AND_DIRECTION,
    TOP_MULTIMODAL_PAIRS_PER_UNIT_AND_DIRECTION,
    TOP_TRAJECTORY_CANDIDATES_PER_LINEAGE
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  prioritization_parameters,
  file.path(STEP7C_DIR, "candidate_prioritization_parameters.csv")
)
```

# 3. Read and validate upstream tables

```{r read-and-validate-upstream-tables}
read_table_quietly <- function(path) {
  suppressMessages(
    readr::read_csv(
      path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
}

assert_columns <- function(data, required_columns, table_label) {
  missing_columns <- setdiff(required_columns, colnames(data))

  if (length(missing_columns) > 0L) {
    stop(
      table_label,
      " is missing required column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
}

eligibility <- read_table_quietly(input_paths[["eligibility"]])
rna_candidates <- read_table_quietly(input_paths[["RNA_candidates"]])
rna_stability <- read_table_quietly(input_paths[["RNA_stability"]])
atac_candidates <- read_table_quietly(input_paths[["ATAC_candidates"]])
atac_stability <- read_table_quietly(input_paths[["ATAC_stability"]])
rna_atac_concordance <- read_table_quietly(
  input_paths[["RNA_ATAC_concordance"]]
)
lineage_definitions <- read_table_quietly(
  input_paths[["lineage_definitions"]]
)
pseudotime_summary <- read_table_quietly(
  input_paths[["pseudotime_summary"]]
)
trajectory_gene_trends <- read_table_quietly(
  input_paths[["trajectory_gene_trends"]]
)

eligibility$analysis_unit <- as.character(eligibility$analysis_unit)
rna_candidates$analysis_unit <- as.character(rna_candidates$analysis_unit)
rna_stability$analysis_unit <- as.character(rna_stability$analysis_unit)
atac_candidates$analysis_unit <- as.character(atac_candidates$analysis_unit)
atac_stability$analysis_unit <- as.character(atac_stability$analysis_unit)
rna_atac_concordance$analysis_unit <- as.character(
  rna_atac_concordance$analysis_unit
)
lineage_definitions$terminal_unit <- as.character(
  lineage_definitions$terminal_unit
)
pseudotime_summary$analysis_unit <- as.character(
  pseudotime_summary$analysis_unit
)
pseudotime_summary$terminal_unit <- as.character(
  pseudotime_summary$terminal_unit
)
trajectory_gene_trends$terminal_unit <- as.character(
  trajectory_gene_trends$terminal_unit
)

assert_columns(
  eligibility,
  c(
    "analysis_unit", "broad_compartment", "analysis_lineage",
    "cell_state", "cell_state_full", "annotation_confidence",
    "nuclei_CON", "nuclei_HFD", "eligibility_tier",
    "primary_RNA_ATAC"
  ),
  "analysis population eligibility"
)

effect_columns <- c(
  "analysis_unit", "feature", "reference_diet", "comparison_diet",
  "reference_nuclei", "comparison_nuclei", "total_counts",
  "reference_CPM", "comparison_CPM",
  "log2FC_comparison_vs_reference",
  "reference_percent_detected", "comparison_percent_detected",
  "detection_difference_percentage_points", "cell_state",
  "cell_state_full", "eligibility_tier", "direction"
)

assert_columns(rna_candidates, effect_columns, "RNA candidate table")
assert_columns(
  atac_candidates,
  c(effect_columns, "gene_name", "distance"),
  "ATAC candidate table"
)

stability_columns <- c(
  "analysis_unit", "feature", "sampled_nuclei_per_diet",
  "full_log2FC", "downsampled_log2FC_median",
  "downsampled_log2FC_q25", "downsampled_log2FC_q75",
  "sign_concordance_percent"
)

assert_columns(rna_stability, stability_columns, "RNA stability table")
assert_columns(atac_stability, stability_columns, "ATAC stability table")

assert_columns(
  lineage_definitions,
  c(
    "trajectory_id", "lineage_index", "lineage_id",
    "cluster_path", "terminal_unit"
  ),
  "trajectory lineage definitions"
)

assert_columns(
  pseudotime_summary,
  c(
    "trajectory_id", "lineage_id", "terminal_unit", "analysis_unit",
    "cell_state", "diet", "nuclei", "median_pseudotime",
    "pseudotime_q25", "pseudotime_q75"
  ),
  "pseudotime summary"
)

assert_columns(
  trajectory_gene_trends,
  c(
    "trajectory_id", "lineage_id", "terminal_unit", "diet",
    "pseudotime_bin", "nuclei", "median_pseudotime", "gene",
    "mean_log_normalized_expression"
  ),
  "trajectory gene-trend table"
)

if (
  any(rna_candidates$reference_diet != REFERENCE_DIET) ||
    any(rna_candidates$comparison_diet != COMPARISON_DIET) ||
    any(atac_candidates$reference_diet != REFERENCE_DIET) ||
    any(atac_candidates$comparison_diet != COMPARISON_DIET)
) {
  stop("Upstream candidate tables do not encode CON as reference and HFD as comparison.")
}

input_inventory <- data.frame(
  input = names(input_paths),
  path = unname(input_paths),
  rows = c(
    nrow(eligibility),
    nrow(rna_candidates),
    nrow(rna_stability),
    nrow(atac_candidates),
    nrow(atac_stability),
    nrow(rna_atac_concordance),
    nrow(lineage_definitions),
    nrow(pseudotime_summary),
    nrow(trajectory_gene_trends)
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  input_inventory,
  file.path(STEP7C_DIR, "candidate_prioritization_input_inventory.csv")
)
```

# 4. Create the WT state reference key

This key is the starting scaffold for later cross-genotype matching. It does
not claim that a PPAR or IL17 state matches a WT state merely because their
short labels are similar. Matching must later incorporate de novo markers,
full labels, broad lineage, and multimodal neighborhood evidence.

```{r create-wt-state-reference-key}
wt_state_reference_key <- eligibility |>
  dplyr::mutate(
    state_matching_priority = dplyr::case_when(
      annotation_confidence == "high" &
        eligibility_tier %in% c("Tier_A", "Tier_B") ~
        "preferred WT reference state",
      annotation_confidence == "moderate" &
        eligibility_tier %in% c("Tier_A", "Tier_B") ~
        "usable WT reference with marker revalidation",
      eligibility_tier == "Tier_C" ~
        "exploratory low-nucleus reference",
      eligibility_tier == "secondary_compartment" ~
        "match only within the same non-epithelial compartment",
      TRUE ~
        "exclude from primary state matching"
    )
  ) |>
  dplyr::arrange(
    broad_compartment,
    analysis_lineage,
    as.integer(analysis_unit)
  )

readr::write_csv(
  wt_state_reference_key,
  file.path(STEP7C_TABLE_DIR, "WT_state_reference_key.csv")
)

wt_state_reference_key |>
  dplyr::count(
    broad_compartment,
    state_matching_priority,
    name = "annotation_units"
  )
```

# 5. Prioritize RNA candidates

```{r prioritize-rna-candidates}
prepare_stability <- function(stability_table, prefix) {
  stability_table |>
    dplyr::mutate(
      direction_stable =
        sign(full_log2FC) == sign(downsampled_log2FC_median) &
        sign_concordance_percent >= MINIMUM_SIGN_CONCORDANCE_PERCENT,
      downsampling_iqr_excludes_zero = dplyr::case_when(
        full_log2FC > 0 ~ downsampled_log2FC_q25 > 0,
        full_log2FC < 0 ~ downsampled_log2FC_q75 < 0,
        TRUE ~ FALSE
      )
    ) |>
    dplyr::rename_with(
      ~ paste0(prefix, .x),
      -dplyr::all_of(c("analysis_unit", "feature"))
    )
}

rna_stability_prepared <- prepare_stability(rna_stability, "RNA_")

rna_prioritized <- rna_candidates |>
  dplyr::left_join(
    rna_stability_prepared,
    by = c("analysis_unit", "feature")
  ) |>
  dplyr::mutate(
    RNA_stability_available = !is.na(RNA_sign_concordance_percent),
    RNA_support_class = dplyr::case_when(
      RNA_direction_stable & RNA_downsampling_iqr_excludes_zero ~
        "Tier 1: stable direction and IQR excludes zero",
      RNA_direction_stable ~
        "Tier 2: stable direction",
      RNA_stability_available ~
        "Tier 3: unstable across balanced cell samples",
      TRUE ~
        "Tier 4: stability not evaluated"
    ),
    absolute_log2FC = abs(log2FC_comparison_vs_reference),
    absolute_detection_difference = abs(
      detection_difference_percentage_points
    ),
    support_order = dplyr::case_when(
      grepl("^Tier 1", RNA_support_class) ~ 1L,
      grepl("^Tier 2", RNA_support_class) ~ 2L,
      grepl("^Tier 3", RNA_support_class) ~ 3L,
      TRUE ~ 4L
    )
  ) |>
  dplyr::group_by(analysis_unit, direction) |>
  dplyr::arrange(
    support_order,
    dplyr::desc(absolute_log2FC),
    dplyr::desc(absolute_detection_difference),
    dplyr::desc(total_counts),
    .by_group = TRUE
  ) |>
  dplyr::mutate(within_unit_direction_rank = dplyr::row_number()) |>
  dplyr::ungroup()

rna_review_shortlist <- rna_prioritized |>
  dplyr::filter(
    within_unit_direction_rank <=
      TOP_RNA_CANDIDATES_PER_UNIT_AND_DIRECTION
  )

readr::write_csv(
  rna_prioritized,
  file.path(STEP7C_TABLE_DIR, "RNA_prioritized_diet_candidates.csv")
)

readr::write_csv(
  rna_review_shortlist,
  file.path(STEP7C_TABLE_DIR, "RNA_review_shortlist.csv")
)
```

# 6. Prioritize ATAC candidates

Nearest-gene assignments are retained as genomic context. Distance zero means
the peak overlaps a gene annotation; it does not establish the target gene or
whether accessibility is activating, repressive, or unrelated to that gene.

```{r prioritize-atac-candidates}
atac_stability_prepared <- prepare_stability(atac_stability, "ATAC_")

atac_prioritized <- atac_candidates |>
  dplyr::left_join(
    atac_stability_prepared,
    by = c("analysis_unit", "feature")
  ) |>
  dplyr::mutate(
    ATAC_stability_available = !is.na(ATAC_sign_concordance_percent),
    ATAC_support_class = dplyr::case_when(
      ATAC_direction_stable & ATAC_downsampling_iqr_excludes_zero ~
        "Tier 1: stable direction and IQR excludes zero",
      ATAC_direction_stable ~
        "Tier 2: stable direction",
      ATAC_stability_available ~
        "Tier 3: unstable across balanced cell samples",
      TRUE ~
        "Tier 4: stability not evaluated"
    ),
    nearest_gene_context = dplyr::case_when(
      is.na(gene_name) | !nzchar(gene_name) ~ "nearest gene unavailable",
      is.na(distance) ~ "distance unavailable",
      distance == 0 ~ "overlaps gene annotation",
      distance <= NEAR_GENE_DISTANCE_BP ~ "within 10 kb of nearest gene",
      TRUE ~ "more than 10 kb from nearest gene"
    ),
    absolute_log2FC = abs(log2FC_comparison_vs_reference),
    absolute_detection_difference = abs(
      detection_difference_percentage_points
    ),
    support_order = dplyr::case_when(
      grepl("^Tier 1", ATAC_support_class) ~ 1L,
      grepl("^Tier 2", ATAC_support_class) ~ 2L,
      grepl("^Tier 3", ATAC_support_class) ~ 3L,
      TRUE ~ 4L
    )
  ) |>
  dplyr::group_by(analysis_unit, direction) |>
  dplyr::arrange(
    support_order,
    dplyr::desc(absolute_log2FC),
    dplyr::desc(absolute_detection_difference),
    dplyr::desc(total_counts),
    .by_group = TRUE
  ) |>
  dplyr::mutate(within_unit_direction_rank = dplyr::row_number()) |>
  dplyr::ungroup()

atac_review_shortlist <- atac_prioritized |>
  dplyr::filter(
    within_unit_direction_rank <=
      TOP_ATAC_CANDIDATES_PER_UNIT_AND_DIRECTION
  )

readr::write_csv(
  atac_prioritized,
  file.path(STEP7C_TABLE_DIR, "ATAC_prioritized_diet_candidates.csv")
)

readr::write_csv(
  atac_review_shortlist,
  file.path(STEP7C_TABLE_DIR, "ATAC_review_shortlist.csv")
)
```

# 7. Prioritize directionally concordant RNA/ATAC pairs

The concordance table asks whether an RNA effect and the accessibility effect
of a peak assigned to that gene have the same sign. A concordant sign is
supportive context, not proof of a peak-to-gene regulatory link.

```{r prioritize-rna-atac-pairs}
assert_columns(
  rna_atac_concordance,
  c(
    "analysis_unit", "peak", "nearest_gene", "distance",
    "ATAC_log2FC", "RNA_log2FC", "RNA_ATAC_direction",
    "joint_effect_magnitude", "cell_state", "cell_state_full"
  ),
  "RNA/ATAC concordance table"
)

rna_top_key <- rna_candidates |>
  dplyr::transmute(
    analysis_unit,
    nearest_gene = feature,
    RNA_top_candidate = TRUE
  ) |>
  dplyr::distinct()

rna_stability_for_pairs <- rna_stability_prepared |>
  dplyr::rename(nearest_gene = feature)

atac_stability_for_pairs <- atac_stability_prepared |>
  dplyr::rename(peak = feature)

multimodal_candidates <- rna_atac_concordance |>
  dplyr::filter(
    RNA_ATAC_direction == "concordant",
    !is.na(RNA_log2FC),
    is.finite(RNA_log2FC),
    is.finite(ATAC_log2FC)
  ) |>
  dplyr::left_join(
    rna_top_key,
    by = c("analysis_unit", "nearest_gene")
  ) |>
  dplyr::left_join(
    rna_stability_for_pairs,
    by = c("analysis_unit", "nearest_gene")
  ) |>
  dplyr::left_join(
    atac_stability_for_pairs,
    by = c("analysis_unit", "peak")
  ) |>
  dplyr::mutate(
    RNA_top_candidate = dplyr::coalesce(RNA_top_candidate, FALSE),
    RNA_direction_stable = dplyr::coalesce(
      RNA_direction_stable,
      FALSE
    ),
    ATAC_direction_stable = dplyr::coalesce(
      ATAC_direction_stable,
      FALSE
    ),
    RNA_downsampling_iqr_excludes_zero = dplyr::coalesce(
      RNA_downsampling_iqr_excludes_zero,
      FALSE
    ),
    ATAC_downsampling_iqr_excludes_zero = dplyr::coalesce(
      ATAC_downsampling_iqr_excludes_zero,
      FALSE
    ),
    both_modalities_direction_stable =
      RNA_direction_stable & ATAC_direction_stable,
    both_downsampling_iqrs_exclude_zero =
      RNA_downsampling_iqr_excludes_zero &
      ATAC_downsampling_iqr_excludes_zero,
    nearest_gene_context = dplyr::case_when(
      is.na(distance) ~ "distance unavailable",
      distance == 0 ~ "overlaps gene annotation",
      distance <= NEAR_GENE_DISTANCE_BP ~ "within 10 kb of nearest gene",
      TRUE ~ "more than 10 kb from nearest gene"
    ),
    multimodal_support_class = dplyr::case_when(
      RNA_top_candidate &
        both_modalities_direction_stable &
        both_downsampling_iqrs_exclude_zero ~
        "Tier 1: RNA candidate plus stable RNA and ATAC",
      both_modalities_direction_stable ~
        "Tier 2: concordant and direction-stable in both modalities",
      RNA_direction_stable | ATAC_direction_stable ~
        "Tier 3: concordant with one stable modality",
      TRUE ~
        "Tier 4: directional concordance only"
    ),
    effect_direction = dplyr::if_else(
      RNA_log2FC > 0,
      "HFD_higher",
      "CON_higher"
    ),
    support_order = dplyr::case_when(
      grepl("^Tier 1", multimodal_support_class) ~ 1L,
      grepl("^Tier 2", multimodal_support_class) ~ 2L,
      grepl("^Tier 3", multimodal_support_class) ~ 3L,
      TRUE ~ 4L
    )
  ) |>
  dplyr::group_by(analysis_unit, effect_direction) |>
  dplyr::arrange(
    support_order,
    dplyr::desc(joint_effect_magnitude),
    .by_group = TRUE
  ) |>
  dplyr::mutate(within_unit_direction_rank = dplyr::row_number()) |>
  dplyr::ungroup()

multimodal_review_shortlist <- multimodal_candidates |>
  dplyr::filter(
    within_unit_direction_rank <=
      TOP_MULTIMODAL_PAIRS_PER_UNIT_AND_DIRECTION
  )

readr::write_csv(
  multimodal_candidates,
  file.path(
    STEP7C_TABLE_DIR,
    "RNA_ATAC_concordant_prioritized_candidates.csv"
  )
)

readr::write_csv(
  multimodal_review_shortlist,
  file.path(STEP7C_TABLE_DIR, "RNA_ATAC_concordant_review_shortlist.csv")
)
```

# 8. Add trajectory context to RNA candidates

The trajectory layer answers whether the state containing an RNA candidate is
on an approved lineage and whether that gene has a reproducible descriptive
shape across relative pseudotime bins. It does not turn a pooled-library RNA
effect into evidence of developmental regulation.

```{r create-trajectory-context}
split_units <- function(unit_string) {
  units <- trimws(
    unlist(strsplit(as.character(unit_string), ";", fixed = TRUE))
  )
  units[nzchar(units)]
}

lineage_unit_membership <- dplyr::bind_rows(
  lapply(
    seq_len(nrow(lineage_definitions)),
    function(lineage_row) {
      unit_path <- split_units(
        lineage_definitions$cluster_path[[lineage_row]]
      )

      data.frame(
        trajectory_id = lineage_definitions$trajectory_id[[lineage_row]],
        lineage_id = lineage_definitions$lineage_id[[lineage_row]],
        terminal_unit = as.character(
          lineage_definitions$terminal_unit[[lineage_row]]
        ),
        analysis_unit = unit_path,
        path_position = seq_along(unit_path),
        trajectory_role = dplyr::case_when(
          seq_along(unit_path) == 1L ~ "root",
          seq_along(unit_path) == length(unit_path) ~ "terminal",
          TRUE ~ "intermediate"
        ),
        stringsAsFactors = FALSE
      )
    }
  )
) |>
  dplyr::left_join(
    eligibility |>
      dplyr::select(
        analysis_unit,
        analysis_lineage,
        cell_state,
        cell_state_full,
        annotation_confidence
      ),
    by = "analysis_unit"
  ) |>
  dplyr::mutate(
    annotated_unit = paste0(analysis_unit, ": ", cell_state),
    model_role = dplyr::if_else(
      grepl("sensitivity", trajectory_id, ignore.case = TRUE),
      "root sensitivity model",
      "primary descriptive model"
    )
  )

readr::write_csv(
  lineage_unit_membership,
  file.path(STEP7C_TABLE_DIR, "trajectory_unit_membership.csv")
)

safe_spearman <- function(x, y) {
  complete_rows <- is.finite(x) & is.finite(y)

  if (
    sum(complete_rows) < 3L ||
      length(unique(x[complete_rows])) < 2L ||
      length(unique(y[complete_rows])) < 2L
  ) {
    return(NA_real_)
  }

  suppressWarnings(
    stats::cor(
      x[complete_rows],
      y[complete_rows],
      method = "spearman"
    )
  )
}

trajectory_shape_by_diet <- trajectory_gene_trends |>
  dplyr::group_by(
    trajectory_id,
    lineage_id,
    terminal_unit,
    diet,
    gene
  ) |>
  dplyr::summarise(
    pseudotime_bins = dplyr::n_distinct(pseudotime_bin),
    mean_expression_across_bins = mean(
      mean_log_normalized_expression,
      na.rm = TRUE
    ),
    pseudotime_spearman_rho = safe_spearman(
      median_pseudotime,
      mean_log_normalized_expression
    ),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = c(
      pseudotime_bins,
      mean_expression_across_bins,
      pseudotime_spearman_rho
    ),
    names_glue = "{.value}_{diet}"
  )

trajectory_bin_pairs <- trajectory_gene_trends |>
  dplyr::select(
    trajectory_id,
    lineage_id,
    terminal_unit,
    gene,
    pseudotime_bin,
    diet,
    mean_log_normalized_expression
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = mean_log_normalized_expression,
    names_prefix = "mean_expression_"
  )

trajectory_diet_profile_summary <- trajectory_bin_pairs |>
  dplyr::group_by(
    trajectory_id,
    lineage_id,
    terminal_unit,
    gene
  ) |>
  dplyr::summarise(
    paired_pseudotime_bins = sum(
      is.finite(mean_expression_CON) &
        is.finite(mean_expression_HFD)
    ),
    mean_HFD_minus_CON_across_relative_bins = {
      paired_rows <- is.finite(mean_expression_CON) &
        is.finite(mean_expression_HFD)

      if (any(paired_rows)) {
        mean(
          mean_expression_HFD[paired_rows] -
            mean_expression_CON[paired_rows]
        )
      } else {
        NA_real_
      }
    },
    diet_profile_spearman_rho = safe_spearman(
      mean_expression_CON,
      mean_expression_HFD
    ),
    .groups = "drop"
  )

trajectory_gene_context <- trajectory_shape_by_diet |>
  dplyr::left_join(
    trajectory_diet_profile_summary,
    by = c(
      "trajectory_id", "lineage_id", "terminal_unit", "gene"
    )
  )

trajectory_rna_candidates <- rna_prioritized |>
  dplyr::rename(gene = feature) |>
  dplyr::inner_join(
    lineage_unit_membership |>
      dplyr::select(
        trajectory_id,
        lineage_id,
        terminal_unit,
        analysis_unit,
        path_position,
        trajectory_role,
        annotation_confidence,
        annotated_unit,
        model_role
      ),
    by = "analysis_unit"
  ) |>
  dplyr::left_join(
    trajectory_gene_context,
    by = c(
      "trajectory_id", "lineage_id", "terminal_unit", "gene"
    )
  ) |>
  dplyr::mutate(
    sufficient_trajectory_bins =
      dplyr::coalesce(pseudotime_bins_CON, 0L) >=
        MINIMUM_PAIRED_PSEUDOTIME_BINS &
      dplyr::coalesce(pseudotime_bins_HFD, 0L) >=
        MINIMUM_PAIRED_PSEUDOTIME_BINS &
      dplyr::coalesce(paired_pseudotime_bins, 0L) >=
        MINIMUM_PAIRED_PSEUDOTIME_BINS,
    state_effect_agrees_with_mean_lineage_profile = dplyr::case_when(
      is.na(mean_HFD_minus_CON_across_relative_bins) ~ NA,
      mean_HFD_minus_CON_across_relative_bins == 0 ~ NA,
      TRUE ~ sign(log2FC_comparison_vs_reference) ==
        sign(mean_HFD_minus_CON_across_relative_bins)
    ),
    trajectory_context_class = dplyr::case_when(
      RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero &
        sufficient_trajectory_bins ~
        "Tier 1: stable state effect with trajectory profile available",
      RNA_direction_stable & sufficient_trajectory_bins ~
        "Tier 2: stable state effect with trajectory profile available",
      sufficient_trajectory_bins ~
        "Tier 3: trajectory profile available; state effect less stable",
      TRUE ~
        "Tier 4: limited trajectory-bin context"
    ),
    trajectory_context_order = dplyr::case_when(
      grepl("^Tier 1", trajectory_context_class) ~ 1L,
      grepl("^Tier 2", trajectory_context_class) ~ 2L,
      grepl("^Tier 3", trajectory_context_class) ~ 3L,
      TRUE ~ 4L
    )
  ) |>
  dplyr::group_by(trajectory_id, lineage_id) |>
  dplyr::arrange(
    trajectory_context_order,
    support_order,
    dplyr::desc(absolute_log2FC),
    .by_group = TRUE
  ) |>
  dplyr::mutate(within_lineage_rank = dplyr::row_number()) |>
  dplyr::ungroup()

trajectory_rna_shortlist <- trajectory_rna_candidates |>
  dplyr::filter(
    within_lineage_rank <= TOP_TRAJECTORY_CANDIDATES_PER_LINEAGE
  )

readr::write_csv(
  trajectory_gene_context,
  file.path(STEP7C_TABLE_DIR, "trajectory_gene_profile_context.csv")
)

readr::write_csv(
  trajectory_rna_candidates,
  file.path(STEP7C_TABLE_DIR, "trajectory_aware_RNA_candidates.csv")
)

readr::write_csv(
  trajectory_rna_shortlist,
  file.path(STEP7C_TABLE_DIR, "trajectory_aware_RNA_shortlist.csv")
)
```

## Absorptive root-sensitivity summary

```{r summarize-absorptive-root-sensitivity}
absorptive_root_sensitivity <- pseudotime_summary |>
  dplyr::filter(
    trajectory_id %in% c(
      "crypt_to_absorptive",
      "crypt_to_absorptive_root0_sensitivity"
    )
  ) |>
  dplyr::group_by(
    trajectory_id,
    lineage_id,
    analysis_unit,
    cell_state
  ) |>
  dplyr::summarise(
    nuclei_across_diet_rows = sum(nuclei),
    median_of_diet_medians = stats::median(median_pseudotime),
    .groups = "drop"
  ) |>
  dplyr::group_by(trajectory_id, lineage_id) |>
  dplyr::arrange(median_of_diet_medians, .by_group = TRUE) |>
  dplyr::mutate(pseudotime_order = dplyr::row_number()) |>
  dplyr::ungroup()

readr::write_csv(
  absorptive_root_sensitivity,
  file.path(STEP7C_TABLE_DIR, "absorptive_root_sensitivity_ordering.csv")
)
```

# 9. Visual review

```{r plot-rna-support, fig.width=14, fig.height=10}
rna_plot_data <- rna_review_shortlist |>
  dplyr::mutate(
    state_display = paste0(analysis_unit, ": ", cell_state),
    RNA_support_short = sub(":.*$", "", RNA_support_class)
  )

p_rna_support <- ggplot2::ggplot(
  rna_plot_data,
  ggplot2::aes(
    x = log2FC_comparison_vs_reference,
    y = detection_difference_percentage_points,
    color = RNA_support_short,
    size = log10(total_counts + 1)
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_point(alpha = 0.70) +
  ggplot2::facet_wrap(~ state_display, scales = "free", ncol = 4) +
  ggplot2::labs(
    title = "WT RNA candidate support landscape",
    subtitle = paste(
      "HFD versus CON pooled-library effects; support reflects balanced",
      "cell-sampling stability, not biological replication."
    ),
    x = "RNA pooled-library log2 fold change",
    y = "Detection difference (percentage points)",
    color = "Support",
    size = "log10 total counts"
  ) +
  theme_multiome(base_size = 9)

print(p_rna_support)

ggplot2::ggsave(
  file.path(STEP7C_PLOT_DIR, "01_RNA_candidate_support_landscape.pdf"),
  p_rna_support,
  width = 14,
  height = 10
)
```

```{r plot-atac-support, fig.width=14, fig.height=10}
atac_plot_data <- atac_review_shortlist |>
  dplyr::mutate(
    state_display = paste0(analysis_unit, ": ", cell_state),
    ATAC_support_short = sub(":.*$", "", ATAC_support_class)
  )

p_atac_support <- ggplot2::ggplot(
  atac_plot_data,
  ggplot2::aes(
    x = log2FC_comparison_vs_reference,
    y = detection_difference_percentage_points,
    color = ATAC_support_short,
    shape = nearest_gene_context
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_point(alpha = 0.65, size = 1.1) +
  ggplot2::facet_wrap(~ state_display, scales = "free", ncol = 4) +
  ggplot2::labs(
    title = "WT ATAC candidate support landscape",
    subtitle = paste(
      "HFD versus CON pooled-library effects; nearest-gene assignment is",
      "genomic context and not a regulatory link."
    ),
    x = "ATAC pooled-library log2 fold change",
    y = "Detection difference (percentage points)",
    color = "Support",
    shape = "Nearest-gene context"
  ) +
  theme_multiome(base_size = 9)

print(p_atac_support)

ggplot2::ggsave(
  file.path(STEP7C_PLOT_DIR, "02_ATAC_candidate_support_landscape.pdf"),
  p_atac_support,
  width = 14,
  height = 10
)
```

```{r plot-multimodal-candidates, fig.width=14, fig.height=10}
multimodal_plot_data <- multimodal_review_shortlist |>
  dplyr::mutate(
    state_display = paste0(analysis_unit, ": ", cell_state),
    multimodal_support_short = sub(
      ":.*$",
      "",
      multimodal_support_class
    )
  )

p_multimodal <- ggplot2::ggplot(
  multimodal_plot_data,
  ggplot2::aes(
    x = ATAC_log2FC,
    y = RNA_log2FC,
    color = multimodal_support_short,
    shape = nearest_gene_context
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, color = "grey75", linewidth = 0.3) +
  ggplot2::geom_point(alpha = 0.75, size = 1.2) +
  ggplot2::facet_wrap(~ state_display, scales = "free", ncol = 4) +
  ggplot2::labs(
    title = "Directionally concordant WT RNA/ATAC candidates",
    subtitle = paste(
      "Each point is an RNA gene and nearest-gene ATAC peak pair;",
      "concordance does not establish peak-to-gene regulation."
    ),
    x = "ATAC pooled-library log2 fold change",
    y = "RNA pooled-library log2 fold change",
    color = "Support",
    shape = "Nearest-gene context"
  ) +
  theme_multiome(base_size = 9)

print(p_multimodal)

ggplot2::ggsave(
  file.path(STEP7C_PLOT_DIR, "03_RNA_ATAC_concordant_candidates.pdf"),
  p_multimodal,
  width = 14,
  height = 10
)
```

```{r plot-root-sensitivity, fig.width=11, fig.height=7}
p_root_sensitivity <- ggplot2::ggplot(
  absorptive_root_sensitivity,
  ggplot2::aes(
    x = pseudotime_order,
    y = median_of_diet_medians,
    color = trajectory_id,
    group = trajectory_id
  )
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_text(
    ggplot2::aes(label = paste0(analysis_unit, ": ", cell_state)),
    check_overlap = TRUE,
    nudge_y = 0.8,
    size = 3
  ) +
  ggplot2::labs(
    title = "Absorptive ordering under the approved root sensitivity model",
    subtitle = paste(
      "Medians summarize pooled-library pseudotime and compare ordering,",
      "not absolute pseudotime values between models."
    ),
    x = "Within-model state order",
    y = "Median of CON and HFD median pseudotime",
    color = "Model"
  ) +
  theme_multiome(base_size = 9)

print(p_root_sensitivity)

ggplot2::ggsave(
  file.path(STEP7C_PLOT_DIR, "04_absorptive_root_sensitivity_ordering.pdf"),
  p_root_sensitivity,
  width = 11,
  height = 7
)
```

# 10. Save a compact review summary

```{r save-step-seven-c}
candidate_support_summary <- dplyr::bind_rows(
  rna_prioritized |>
    dplyr::count(
      modality = "RNA",
      support_class = RNA_support_class,
      name = "candidates"
    ),
  atac_prioritized |>
    dplyr::count(
      modality = "ATAC",
      support_class = ATAC_support_class,
      name = "candidates"
    ),
  multimodal_candidates |>
    dplyr::count(
      modality = "RNA_ATAC_pair",
      support_class = multimodal_support_class,
      name = "candidates"
    ),
  trajectory_rna_candidates |>
    dplyr::count(
      modality = "trajectory_RNA",
      support_class = trajectory_context_class,
      name = "candidates"
    )
)

readr::write_csv(
  candidate_support_summary,
  file.path(STEP7C_DIR, "candidate_support_summary.csv")
)

step7c_summary <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  RNA_candidates = nrow(rna_prioritized),
  RNA_shortlist = nrow(rna_review_shortlist),
  ATAC_candidates = nrow(atac_prioritized),
  ATAC_shortlist = nrow(atac_review_shortlist),
  concordant_RNA_ATAC_pairs = nrow(multimodal_candidates),
  concordant_RNA_ATAC_shortlist = nrow(multimodal_review_shortlist),
  trajectory_RNA_candidates = nrow(trajectory_rna_candidates),
  trajectory_RNA_shortlist = nrow(trajectory_rna_shortlist),
  replicated_diet_design = FALSE,
  biological_inference_performed = FALSE,
  stringsAsFactors = FALSE
)

readr::write_csv(
  step7c_summary,
  file.path(STEP7C_DIR, "step7c_summary.csv")
)

writeLines(
  c(
    paste("Step 7C completed:", format(Sys.time())),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste("RNA candidate rows:", nrow(rna_prioritized)),
    paste("ATAC candidate rows:", nrow(atac_prioritized)),
    paste(
      "Directionally concordant RNA/ATAC pairs:",
      nrow(multimodal_candidates)
    ),
    paste(
      "Trajectory-aware RNA candidate rows:",
      nrow(trajectory_rna_candidates)
    ),
    "Biological-replicate diet inference performed: FALSE",
    paste(
      "Interpretation:",
      "descriptive WT pooled-library candidates for later validation"
    )
  ),
  STEP7C_COMPLETE_MARKER
)

step7c_summary
```

# Interpretation checkpoint

Before promoting a candidate for cross-genotype comparison, ask:

1. Is the candidate found in a high- or moderate-confidence matched state?
2. Does its direction remain stable under balanced cell sampling?
3. Is an RNA/ATAC pair supported by more than nearest-gene proximity?
4. Does coverage inspection support the candidate peak in both libraries?
5. Does the RNA candidate occur in a biologically coherent part of an
   approved trajectory, and is that conclusion stable to the alternative
   absorptive root?
6. Does the same marker-supported state exist in PPAR and IL17?
7. Is the direction reproduced across genotypes, or is it genotype-specific?

# What comes next?

Step 8 should construct a cross-genotype state-matching table before joining
diet-associated effects. Comparisons should use biological state labels and
marker evidence rather than raw cluster numbers. A candidate becomes suitable
for inferential differential analysis only when independent biological
replicates are available.
