---
title: "Step 07D v3 | Diet response in whole annotated WNN clusters"
author: "Mana Lab multiome teaching pipeline"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# Aim and scope

This companion to Step 07 calculates descriptive HFD-versus-CON RNA and ATAC
changes **within each whole Step 06 WNN cluster**. Step 07 can use refined
`annotation_unit` children; this notebook deliberately uses `wnn_cluster`.
It looks up approved *parent-cluster* labels in the cohort's Step 06 decision
file and audits any difference from nucleus-level refined labels. A parent
with an approved child is explicitly labeled as a **mixed whole cluster**;
the residual parent's label is not asserted for every nucleus in that WNN
cluster.

Run the same notebook separately for WT (`wt`), PPAR (`ppar`), and IL17
(`il17`). Numeric cluster IDs belong to their own cohort; do not join a WT
cluster to a knockout cluster by number. The full ATAC peak sets may also
differ between cohorts. The frozen-interval recount in Step 08J addresses
exact-locus comparison separately.

There is **one pooled library per diet per cohort**. The reported nucleus
counts, CPM, detection differences, fold changes, and candidate counts are
descriptive statistics. They do not supply animal-level P values, significant
DARs, or a genotype-by-diet interaction test. A clear-effect threshold is an
inspectable screen, not a significance threshold.

The notebook requires the standard `config/project_config.R` and the saved
Step 06 object/annotation CSV already used by the teaching pipeline. All
calculation helpers are defined below; the dated `deliverables/` helper is
**not** required. Allow substantial RAM (the Step 06 object plus one assay
matrix) and time for the ATAC tables.

## 1. Select one cohort and load the standard configuration

From the repository root on SOL, set `MULTIOME_COHORT_ID` to `wt`, `ppar`, or
`il17` before running the chunks or rendering. If unset, this notebook starts
with `wt`. The project root is explicit so it also works when RStudio's working
directory is different. Results are separated by cohort.

For a noninteractive run after downloading this Rmd to `analysis/`:

```bash
cd /scratch/dsaiz/multiome_scRNA-ATAC
MULTIOME_COHORT_ID=wt Rscript -e 'rmarkdown::render("analysis/07d_whole_wnn_cluster_diet_response.Rmd", output_file="07d_wt_whole_WNN_v3.html")'
MULTIOME_COHORT_ID=ppar Rscript -e 'rmarkdown::render("analysis/07d_whole_wnn_cluster_diet_response.Rmd", output_file="07d_ppar_whole_WNN_v3.html")'
MULTIOME_COHORT_ID=il17 Rscript -e 'rmarkdown::render("analysis/07d_whole_wnn_cluster_diet_response.Rmd", output_file="07d_il17_whole_WNN_v3.html")'
```

Run one cohort at a time. The HTML notebook is a run record; detailed results
are written beneath each cohort's `07d_whole_wnn_cluster_diet_response`
directory in the configured results tree.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
  fig.width = 11, fig.height = 7)
REPO_ROOT <- Sys.getenv("MULTIOME_REPO_ROOT",
  unset = getwd())
if (!file.exists(file.path(REPO_ROOT, "config", "project_config.R"))) {
  stop("Cannot find the teaching pipeline at REPO_ROOT: ", REPO_ROOT)
}
REPO_ROOT <- normalizePath(REPO_ROOT)
COHORT_TO_RUN <- tolower(Sys.getenv("MULTIOME_COHORT_ID", unset = "wt"))
if (!COHORT_TO_RUN %in% c("wt", "ppar", "il17")) {
  stop("MULTIOME_COHORT_ID must be wt, ppar, or il17.")
}
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = COHORT_TO_RUN)
source(file.path(REPO_ROOT, "config", "project_config.R"))
suppressPackageStartupMessages({
  library(SeuratObject)
  library(Signac)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(tibble)
  library(ggplot2)
})
stopifnot(DATASET_VERSION == "v2_resequenced", COHORT_ID == COHORT_TO_RUN)

STEP6_OBJECT <- file.path(OUTPUT_DIR, "06_annotation", "multiome_annotated.rds")
DECISIONS <- file.path(DATASET_CONFIG_DIR, "cohorts", COHORT_ID,
  "cell_state_annotations.csv")
OUT <- file.path(OUTPUT_DIR, "07d_whole_wnn_cluster_diet_response")
TABLES <- file.path(OUT, "tables")
PLOTS <- file.path(OUT, "plots")
for (p in c(OUT, TABLES, PLOTS))
  dir.create(p, recursive = TRUE, showWarnings = FALSE)

# Match the effect and feature-evaluability thresholds in teaching Step 07.
MIN_N_EXPLORATORY <- 30L
MIN_N_TIER_B <- 100L
MIN_N_TIER_A <- 200L
CPM_PSEUDOCOUNT <- 0.10
MIN_TOTAL_COUNTS <- c(RNA = 20L, ATAC = 30L)
MIN_DETECTION_PERCENT <- c(RNA = 5, ATAC = 1)
CLEAR_ABS_LOG2FC <- 0.50
RNA_MIN_DETECTION_DELTA_PP <- 5
TOP_N_PER_DIRECTION <- c(RNA = 250L, ATAC = 500L)

need <- function(ok, message) {
  if (length(ok) != 1L || is.na(ok) || !ok) stop(message, call. = FALSE)
}
need_columns <- function(x, columns, label) {
  missing <- setdiff(columns, names(x))
  need(!length(missing), paste(label, "missing columns:",
    paste(missing, collapse = ", ")))
}
write_table <- function(x, directory, filename) {
  readr::write_csv(x, file.path(directory, filename), na = "NA")
}
align_counts <- function(x, cells, label) {
  need(!is.null(colnames(x)) && !anyDuplicated(colnames(x)) &&
    !anyDuplicated(cells), paste(label, "has missing/duplicated cell IDs."))
  need(setequal(colnames(x), cells),
    paste(label, "does not contain exactly the Step 06 nuclei."))
  x <- x[, cells, drop = FALSE]
  need(identical(colnames(x), cells) && !anyDuplicated(rownames(x)),
    paste(label, "could not be aligned or has duplicate features."))
  x
}
```

## 2. Verify the saved object, pooled libraries, and approved parent labels

The annotation CSV contains one row for each global WNN parent and may also
contain approved child `annotation_unit` rows. IL17's `11_tuft` is a child of
WNN 11, which also contains the residual annotation unit `11`. Their distinct
labels remain in the audit; the whole WNN 11 label is marked mixed.

```{r load-and-audit}
need(file.exists(STEP6_OBJECT), paste("Missing Step 06 object:", STEP6_OBJECT))
need(file.exists(DECISIONS), paste("Missing approved annotation CSV:", DECISIONS))
manifest_paths <- c(STEP6_OBJECT, DECISIONS)
manifest_info <- file.info(manifest_paths)
write_table(tibble::tibble(path = normalizePath(manifest_paths),
  bytes = manifest_info$size, modified = as.character(manifest_info$mtime)),
  OUT, "00_input_manifest.csv")

decisions <- readr::read_csv(DECISIONS, show_col_types = FALSE)
need_columns(decisions, c("cluster", "cell_state", "approved",
  "annotation_confidence"), "Step 06 decisions")
decisions$cluster <- as.character(decisions$cluster)
need(!anyDuplicated(decisions$cluster) && all(decisions$approved %in% TRUE),
  "An annotation unit is duplicated or lacks approval in Step 06 decisions.")
if (!"parent_wnn_cluster" %in% names(decisions)) {
  # WT and PPAR approvals have only global parent rows.
  decisions$parent_wnn_cluster <- decisions$cluster
}
decisions$parent_wnn_cluster <- as.character(decisions$parent_wnn_cluster)
need(!anyNA(decisions$parent_wnn_cluster) &&
  all(nzchar(decisions$parent_wnn_cluster)),
  "An approved annotation unit lacks its global WNN parent.")
is_child <- decisions$cluster != decisions$parent_wnn_cluster
parent_decisions <- decisions[!is_child, , drop = FALSE]
child_decisions <- decisions[is_child, , drop = FALSE]
need(!anyDuplicated(parent_decisions$cluster) &&
  all(child_decisions$parent_wnn_cluster %in% parent_decisions$cluster),
  "An approved child lacks an approved global WNN parent.")

obj <- readRDS(STEP6_OBJECT)
need(all(c("RNA", "ATAC") %in% names(obj)),
  "The Step 06 object must contain RNA and ATAC assays.")
meta <- obj[[]]
meta$cell_barcode <- rownames(meta)
need_columns(meta, c("cell_barcode", "wnn_cluster", "cell_state",
  "sample_id", "diet"), "Step 06 metadata")
meta$wnn_cluster <- as.character(meta$wnn_cluster)
meta$diet <- as.character(meta$diet)
meta$sample_id <- as.character(meta$sample_id)
if ("annotation_unit" %in% names(meta)) {
  meta$annotation_unit <- as.character(meta$annotation_unit)
} else {
  need(!nrow(child_decisions),
    "Approved focused children require annotation_unit in the Step 06 object.")
  meta$annotation_unit <- meta$wnn_cluster
}
need(!anyDuplicated(meta$cell_barcode) &&
  setequal(unique(meta$diet), c("CON", "HFD")) &&
  setequal(unique(meta$wnn_cluster), parent_decisions$cluster) &&
  setequal(unique(meta$annotation_unit), decisions$cluster),
  paste("Step 06 global WNN/annotation unit identities differ from the",
    "approved parent/child decision rows. Review the Step 06 object."))
unit_match <- match(meta$annotation_unit, decisions$cluster)
need(!anyNA(unit_match) &&
  identical(meta$wnn_cluster,
    decisions$parent_wnn_cluster[unit_match]) &&
  identical(as.character(meta$cell_state),
    as.character(decisions$cell_state[unit_match])),
  "A Step 06 annotation unit has the wrong global WNN parent or approved label.")
observed_unit_n <- table(meta$annotation_unit)
write_table(tibble::tibble(annotation_unit = decisions$cluster,
  parent_wnn_cluster = decisions$parent_wnn_cluster,
  decision_n = if ("nuclei" %in% names(decisions))
    as.integer(decisions$nuclei) else NA_integer_,
  observed_n = as.integer(observed_unit_n[decisions$cluster])),
  TABLES, "01_approved_annotation_unit_count_audit.csv")

library_map <- dplyr::distinct(meta, sample_id, diet)
need(nrow(library_map) == 2L && !anyDuplicated(library_map$sample_id) &&
  setequal(library_map$diet, c("CON", "HFD")),
  "Expected one pooled sample_id for CON and one for HFD in this cohort.")
write_table(library_map, TABLES, "01_pooled_sample_id_and_diet.csv")

k <- match(meta$wnn_cluster, parent_decisions$cluster)
need(!anyNA(k), "Some Step 06 nuclei lack an approved WNN parent row.")
parent_decisions$whole_cluster_label <- parent_decisions$cell_state
parent_decisions$whole_label_basis <- "approved_parent_label"
parent_decisions$whole_annotation_confidence <-
  parent_decisions$annotation_confidence
for (parent_id in unique(child_decisions$parent_wnn_cluster)) {
  child_labels <- unique(child_decisions$cell_state[
    child_decisions$parent_wnn_cluster == parent_id])
  pk <- match(parent_id, parent_decisions$cluster)
  parent_decisions$whole_cluster_label[pk] <- paste0("Mixed: ",
    parent_decisions$cell_state[pk], " + ",
    paste(child_labels, collapse = " + "))
  parent_decisions$whole_label_basis[pk] <- "composite_parent_plus_child"
  parent_decisions$whole_annotation_confidence[pk] <- "mixed"
}
write_table(dplyr::select(decisions, cluster, parent_wnn_cluster,
  cell_state, annotation_confidence, approved), TABLES,
  "01_approved_parent_and_child_annotation_units.csv")
write_table(dplyr::select(parent_decisions, cluster, cell_state,
  annotation_confidence, whole_cluster_label, whole_label_basis,
  whole_annotation_confidence), TABLES,
  "01_whole_parent_label_resolution.csv")
meta$whole_cluster_label <- parent_decisions$whole_cluster_label[k]
meta$parent_decision_label <- parent_decisions$cell_state[k]
meta$annotation_confidence <-
  parent_decisions$whole_annotation_confidence[k]
meta$whole_label_basis <- parent_decisions$whole_label_basis[k]
meta$refined_label_differs <- !is.na(meta$cell_state) &
  as.character(meta$cell_state) != as.character(meta$parent_decision_label)
label_audit <- meta |>
  dplyr::count(wnn_cluster, annotation_unit, parent_decision_label,
    whole_cluster_label, cell_state, refined_label_differs,
    name = "n_nuclei") |>
  dplyr::arrange(wnn_cluster, dplyr::desc(n_nuclei))
write_table(label_audit, TABLES, "01_whole_vs_nucleus_label_audit.csv")

units <- meta |>
  dplyr::count(wnn_cluster, whole_cluster_label,
    annotation_confidence, diet, name = "n") |>
  tidyr::complete(tidyr::nesting(wnn_cluster, whole_cluster_label,
    annotation_confidence), diet = c("CON", "HFD"), fill = list(n = 0L)) |>
  tidyr::pivot_wider(names_from = diet, values_from = n, names_prefix = "n_") |>
  dplyr::mutate(n_total = n_CON + n_HFD,
    tier = dplyr::case_when(
      pmin(n_CON, n_HFD) >= MIN_N_TIER_A ~ "A_200_per_diet",
      pmin(n_CON, n_HFD) >= MIN_N_TIER_B ~ "B_100_per_diet",
      pmin(n_CON, n_HFD) >= MIN_N_EXPLORATORY ~ "C_exploratory_30_per_diet",
      TRUE ~ "composition_only"),
    effect_evaluable = tier != "composition_only",
    primary = tier %in% c("A_200_per_diet", "B_100_per_diet")) |>
  dplyr::arrange(as.integer(wnn_cluster))
need(sum(units$n_total) == nrow(meta),
  "Whole-cluster count table did not include every Step 06 nucleus.")
write_table(units, TABLES, "01_whole_cluster_diet_counts_and_tiers.csv")
print(units)
```

## 3. Show population composition and eligibility

The denominator is all retained Step 06 nuclei in that diet's pooled library.
This is a descriptive difference in representation, which may also reflect
capture and library composition.

```{r population-composition}
totals <- meta |> dplyr::count(diet, name = "library_n")
composition <- units |>
  dplyr::mutate(CON_percent = 100 * n_CON /
      totals$library_n[match("CON", totals$diet)],
    HFD_percent = 100 * n_HFD /
      totals$library_n[match("HFD", totals$diet)],
    difference_pp = HFD_percent - CON_percent)
write_table(composition, TABLES, "02_whole_cluster_composition.csv")
p_composition <- ggplot2::ggplot(composition,
  ggplot2::aes(x = difference_pp,
    y = reorder(paste0("W", wnn_cluster, " ", whole_cluster_label),
      difference_pp), color = tier)) +
  ggplot2::geom_vline(xintercept = 0, color = "grey60") +
  ggplot2::geom_segment(ggplot2::aes(x = 0, xend = difference_pp,
    yend = reorder(paste0("W", wnn_cluster, " ", whole_cluster_label),
      difference_pp)), linewidth = .5) +
  ggplot2::geom_point(size = 2) +
  ggplot2::labs(title = paste(toupper(COHORT_ID), "whole WNN clusters"),
    subtitle = "Percent of pooled HFD nuclei minus percent of pooled CON nuclei",
    x = "HFD minus CON (percentage points)", y = NULL,
    caption = "Descriptive library composition, not a replicated abundance test.") +
  ggplot2::theme_bw(base_size = 10)
print(p_composition)
ggplot2::ggsave(file.path(PLOTS, "01_whole_cluster_composition.pdf"),
  p_composition, width = 11, height = 8, device = grDevices::pdf,
  useDingbats = FALSE)
```

## 4. Calculate whole-cluster pooled RNA and ATAC effects

For each eligible whole cluster, pool raw counts within each diet and divide
by the **total assay counts in its selected nuclei** to calculate CPM. Use
identical pseudocounts and feature thresholds across cohorts. The RNA clear
screen also requires at least a five-percentage-point detection change.
ATAC clear effects require the count/detection minimum and absolute pooled
log2FC >= 0.5. Neither rule is a hypothesis test.

A separate compressed full feature table is written for each cluster. That
keeps memory bounded and preserves counts, CPM, detection, eligibility, and
effect size for later re-screening. The compact top table and whole-cluster
summary are also exported for each modality.

```{r pooled-effect-function}
pooled_effect <- function(counts, meta, cluster_id, modality, depth) {
  ix <- which(meta$wnn_cluster == cluster_id)
  con <- ix[meta$diet[ix] == "CON"]
  hfd <- ix[meta$diet[ix] == "HFD"]
  need(length(con) >= MIN_N_EXPLORATORY &&
    length(hfd) >= MIN_N_EXPLORATORY,
    paste("Insufficient nuclei for WNN", cluster_id))
  a <- Matrix::rowSums(counts[, con, drop = FALSE])
  b <- Matrix::rowSums(counts[, hfd, drop = FALSE])
  total_con <- sum(depth[con]); total_hfd <- sum(depth[hfd])
  need(total_con > 0 && total_hfd > 0,
    paste("Zero assay depth in WNN", cluster_id, modality))
  con_detection <- 100 * Matrix::rowMeans(counts[, con, drop = FALSE] > 0)
  hfd_detection <- 100 * Matrix::rowMeans(counts[, hfd, drop = FALSE] > 0)
  con_cpm <- 1e6 * a / total_con
  hfd_cpm <- 1e6 * b / total_hfd
  effect <- log2((hfd_cpm + CPM_PSEUDOCOUNT) /
    (con_cpm + CPM_PSEUDOCOUNT))
  detected_delta <- hfd_detection - con_detection
  evaluable <- (a + b >= MIN_TOTAL_COUNTS[[modality]]) &
    pmax(con_detection, hfd_detection) >=
      MIN_DETECTION_PERCENT[[modality]]
  clear <- evaluable & abs(effect) >= CLEAR_ABS_LOG2FC &
    (modality == "ATAC" | abs(detected_delta) >=
      RNA_MIN_DETECTION_DELTA_PP)
  tibble::tibble(feature = rownames(counts),
    CON_n = length(con), HFD_n = length(hfd),
    CON_counts = as.numeric(a), HFD_counts = as.numeric(b),
    CON_CPM = as.numeric(con_cpm), HFD_CPM = as.numeric(hfd_cpm),
    CON_percent_detected = as.numeric(con_detection),
    HFD_percent_detected = as.numeric(hfd_detection),
    detection_delta_pp = as.numeric(detected_delta),
    log2FC_HFD_vs_CON = as.numeric(effect),
    evaluable = as.logical(evaluable), clear_effect = as.logical(clear),
    direction = dplyr::case_when(effect > 0 ~ "HFD_higher",
      effect < 0 ~ "HFD_lower", TRUE ~ "unchanged"))
}

# The number to retain is evaluated outside dplyr's data mask. Use ordinary
# indices to avoid evaluating .env or a similarly named feature-table column
# inside slice_max()/slice_head(). Break equal-effect ties by feature name.
select_top_clear_effects <- function(eff, n_per_direction) {
  need(length(n_per_direction) == 1L && is.finite(n_per_direction) &&
    n_per_direction >= 1L, "Invalid number of top features.")
  clear_rows <- which(eff$clear_effect %in% TRUE)
  if (!length(clear_rows)) return(eff[0, , drop = FALSE])
  by_direction <- split(clear_rows, eff$direction[clear_rows])
  top_rows <- unlist(lapply(by_direction, function(ix) {
    ix <- ix[order(-abs(eff$log2FC_HFD_vs_CON[ix]), eff$feature[ix])]
    utils::head(ix, as.integer(n_per_direction))
  }), use.names = FALSE)
  eff[top_rows, , drop = FALSE]
}
```

```{r whole-cluster-effects}
for (modality in c("RNA", "ATAC")) {
  message("Calculating whole-cluster ", modality, " effects: ", COHORT_ID)
  need("counts" %in% SeuratObject::Layers(obj[[modality]]),
    paste("Joined counts layer missing from", modality))
  counts <- align_counts(SeuratObject::LayerData(
    obj[[modality]], layer = "counts"), meta$cell_barcode,
    paste(COHORT_ID, modality))
  depth <- Matrix::colSums(counts)
  need(all(is.finite(depth)) && all(depth >= 0),
    paste("Nonfinite or negative depth in", modality))
  top_n <- unname(TOP_N_PER_DIRECTION[[modality]])
  detail_dir <- file.path(TABLES, paste0(modality, "_full_by_WNN"))
  dir.create(detail_dir, recursive = TRUE, showWarnings = FALSE)
  summary_rows <- list(); top_rows <- list()
  eligible <- units[units$effect_evaluable, , drop = FALSE]
  for (j in seq_len(nrow(eligible))) {
    row <- eligible[j, ]
    cluster_id <- row$wnn_cluster[[1]]
    message("  WNN ", cluster_id, ": ", row$n_CON,
      " CON / ", row$n_HFD, " HFD")
    eff <- pooled_effect(counts, meta, cluster_id, modality, depth) |>
      dplyr::mutate(cohort = COHORT_ID, modality = .env$modality,
        wnn_cluster = .env$cluster_id,
        whole_cluster_label = row$whole_cluster_label[[1]],
        tier = row$tier[[1]], .before = 1)
    write_table(eff, detail_dir,
      paste0("WNN_", cluster_id, "_pooled_effects.csv.gz"))
    ev <- eff$evaluable
    summary_rows[[j]] <- tibble::tibble(
      cohort = COHORT_ID, modality = modality,
      wnn_cluster = cluster_id,
      whole_cluster_label = row$whole_cluster_label[[1]],
      tier = row$tier[[1]], n_CON = row$n_CON[[1]],
      n_HFD = row$n_HFD[[1]], total_features = nrow(eff),
      evaluable_features = sum(ev),
      clear_HFD_higher = sum(eff$clear_effect &
        eff$log2FC_HFD_vs_CON > 0),
      clear_HFD_lower = sum(eff$clear_effect &
        eff$log2FC_HFD_vs_CON < 0),
      median_abs_log2FC_evaluable = if (any(ev))
        median(abs(eff$log2FC_HFD_vs_CON[ev])) else NA_real_)
    top_rows[[j]] <- select_top_clear_effects(eff, top_n)
    rm(eff); invisible(gc())
  }
  modality_summary <- dplyr::bind_rows(summary_rows)
  write_table(modality_summary, TABLES,
    paste0("03_", modality, "_whole_cluster_effect_summary.csv"))
  write_table(dplyr::bind_rows(top_rows), TABLES,
    paste0("04_", modality, "_top_descriptive_effects.csv.gz"))
  rm(counts, depth, summary_rows, top_rows); invisible(gc())
}
```

## 5. Review the effect counts and complete the cohort run

Counts of screened features depend on population size and assay depth. The
full per-cluster tables provide the denominators and continuous effect sizes;
compare those before describing differences between clusters or cohorts.

```{r effect-summary-figure}
rna_summary <- readr::read_csv(file.path(TABLES,
  "03_RNA_whole_cluster_effect_summary.csv"), show_col_types = FALSE)
atac_summary <- readr::read_csv(file.path(TABLES,
  "03_ATAC_whole_cluster_effect_summary.csv"), show_col_types = FALSE)
combined <- dplyr::bind_rows(rna_summary, atac_summary)
plot_table <- combined |>
  tidyr::pivot_longer(c(clear_HFD_higher, clear_HFD_lower),
    names_to = "direction", values_to = "n_features") |>
  dplyr::mutate(display_label = paste0("W", wnn_cluster,
    " ", whole_cluster_label))
p_effects <- ggplot2::ggplot(plot_table,
  ggplot2::aes(x = display_label, y = n_features, fill = direction)) +
  ggplot2::geom_col() +
  ggplot2::coord_flip() +
  ggplot2::facet_wrap(~modality, scales = "free_x") +
  ggplot2::scale_fill_manual(values = c(clear_HFD_higher = "#56B4E9",
    clear_HFD_lower = "#E69F00"),
    labels = c("HFD higher", "HFD lower")) +
  ggplot2::labs(title = paste(toupper(COHORT_ID),
      "whole WNN cluster RNA and ATAC screens"),
    subtitle = "One pooled CON and one pooled HFD library",
    x = "Approved whole WNN cluster", y = "Features passing descriptive clear-effect rule",
    caption = "Counts depend on evaluable features, cell numbers and depth; no biological-replicate P values.") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(legend.title = ggplot2::element_blank(),
    plot.caption = ggplot2::element_text(hjust = 0))
print(p_effects)
ggplot2::ggsave(file.path(PLOTS, "02_whole_cluster_RNA_ATAC_effect_counts.pdf"),
  p_effects, width = 13, height = 9, device = grDevices::pdf,
  useDingbats = FALSE)

write_table(tibble::tibble(cohort = COHORT_ID, dataset = DATASET_VERSION,
  analysis_unit = "whole approved WNN cluster", all_nuclei = nrow(meta),
  all_clusters = nrow(units), primary_clusters = sum(units$primary),
  exploratory_clusters = sum(units$tier == "C_exploratory_30_per_diet"),
  library_design = "one pooled CON and one pooled HFD library",
  rna_full_tables = sum(units$effect_evaluable),
  atac_full_tables = sum(units$effect_evaluable)),
  OUT, "05_run_summary.csv")
writeLines(c(paste("Step 07D complete:", COHORT_ID),
  "Unit: whole approved Step 06 WNN clusters.",
  "One pooled library per diet; no mouse-level diet test.",
  paste("Outputs:", OUT)), file.path(OUT, "STEP_07D_COMPLETE.txt"))
message("Step 07D output: ", OUT)
```

## Questions for review

1. Do WNN 0 in WT, annotated cycling ISC parents in PPAR/IL17, and their
   context populations retain the expected marker identities?
2. Are RNA and ATAC effect sizes stable enough to describe at the whole-state
   level, and which changes merit exact-coordinate genotype comparison?
3. Which small or diet-imbalanced states should remain exploratory?
