---
title: "Step 8 - Validate persistent HFD-associated genes in WT"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Gourab's reference lists contain genes that changed in the same direction in
an annotated stem-cell population at day 21 and day 50. Step 8 asks whether
those persistent directions are also present in the WT d21 Multiome benchmark.

The analysis proceeds in two linked stages:

1. test every persistent reference gene against the complete Step 7 RNA effect
   table, with the annotated WT `ISC-like crypt` state as the primary
   comparison and other eligible states as context; and
2. identify ATAC peaks that physically overlap a strand-aware plus-or-minus
   2-kb transcription-start-site window for each persistent gene, then compare
   their pooled HFD-versus-CON accessibility direction with the RNA direction.

Step 7C rankings are retained as supporting context. They are not used to
remove reference genes before testing, and their nearest-gene assignments are
kept separate from the exact promoter-overlap analysis.

The legacy HyperMet and HypoImmune gene sets are deliberately excluded from
primary selection here. They can be overlaid after the complete persistent
gene lists have been evaluated.

## Statistical limitation

The WT cohort contains one pooled CON library and one pooled HFD library.
Mouse identities cannot be recovered after pooling, so diet and library are
perfectly confounded. The reported RNA and ATAC changes are descriptive
pooled-library effect sizes. Balanced cell downsampling measures cell-sampling
stability and does not create biological replication.

A promoter-overlapping ATAC peak is genomic context. Directional agreement
between promoter accessibility and RNA expression does not prove that the
peak regulates the gene or that increased accessibility is activating.

# 1. Setup

Run this notebook in a fresh R session configured for the WT v2 cohort:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight()
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(GenomicRanges)
  library(IRanges)
  library(GenomeInfoDb)
  library(S4Vectors)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
})

ggplot2::theme_set(theme_multiome())

set.seed(20260911)
future::plan(future::sequential)

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 8 currently validates the WT v2 benchmark. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}
```

# 2. Define inputs, outputs, and evidence rules

The existing Step 7 thresholds define whether an RNA gene or ATAC peak enters
the complete effect tables. The thresholds below distinguish a clear effect
from a directionally concordant but smaller effect. They do not define
statistical significance.

```{r define-inputs-and-parameters}
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP7_DIR <- file.path(OUTPUT_DIR, "07_diet_response")
STEP7_TABLE_DIR <- file.path(STEP7_DIR, "tables")
STEP7C_DIR <- file.path(OUTPUT_DIR, "07c_candidate_prioritization")
STEP7C_TABLE_DIR <- file.path(STEP7C_DIR, "tables")

STEP8_DIR <- file.path(OUTPUT_DIR, "08_persistent_hfd_programs")
STEP8_TABLE_DIR <- file.path(STEP8_DIR, "tables")
STEP8_PLOT_DIR <- file.path(STEP8_DIR, "plots")

dir.create(STEP8_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

STEP6_OBJECT_PATH <- file.path(
  STEP6_DIR,
  "multiome_annotated.rds"
)

STEP6_COMPLETE_MARKER <- file.path(
  STEP6_DIR,
  "STEP_06_COMPLETE.txt"
)

STEP7_COMPLETE_MARKER <- file.path(
  STEP7_DIR,
  "STEP_07_COMPLETE.txt"
)

STEP7C_COMPLETE_MARKER <- file.path(
  STEP7C_DIR,
  "STEP_07C_COMPLETE.txt"
)

STEP8_COMPLETE_MARKER <- file.path(
  STEP8_DIR,
  "STEP_08_COMPLETE.txt"
)

REFERENCE_GENE_PATH <- file.path(
  DATASET_CONFIG_DIR,
  "reference_programs",
  "persistent_hfd_genes.csv"
)

input_paths <- c(
  eligibility = file.path(
    STEP7_TABLE_DIR,
    "analysis_population_eligibility.csv"
  ),
  RNA_effects = file.path(
    STEP7_TABLE_DIR,
    "RNA_pooled_library_effects_by_analysis_unit.csv.gz"
  ),
  RNA_stability = file.path(
    STEP7_TABLE_DIR,
    "RNA_balanced_downsampling_stability.csv"
  ),
  ATAC_effects = file.path(
    STEP7_TABLE_DIR,
    "ATAC_pooled_library_effects_by_analysis_unit.csv.gz"
  ),
  ATAC_stability = file.path(
    STEP7_TABLE_DIR,
    "ATAC_balanced_downsampling_stability.csv"
  ),
  Step7C_RNA = file.path(
    STEP7C_TABLE_DIR,
    "RNA_prioritized_diet_candidates.csv"
  ),
  Step7C_ATAC = file.path(
    STEP7C_TABLE_DIR,
    "ATAC_prioritized_diet_candidates.csv"
  ),
  Step7C_multimodal = file.path(
    STEP7C_TABLE_DIR,
    "RNA_ATAC_concordant_prioritized_candidates.csv"
  )
)

assert_file(STEP6_OBJECT_PATH, "Step 6 annotated WT object")
assert_file(STEP6_COMPLETE_MARKER, "Step 6 completion marker")
assert_file(STEP7_COMPLETE_MARKER, "Step 7 completion marker")
assert_file(STEP7C_COMPLETE_MARKER, "Step 7C completion marker")
assert_file(REFERENCE_GENE_PATH, "persistent HFD reference gene table")
assert_file(GRCM39_GTF, "GRCm39 GENCODE vM33 GTF")

for (input_label in names(input_paths)) {
  assert_file(input_paths[[input_label]], input_label)
}

REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"

WT_STEM_CELL_STATE <- "ISC-like crypt"
TSS_WINDOW_BP <- 2000L

MINIMUM_SIGN_CONCORDANCE_PERCENT <- 80
RNA_CLEAR_LOG2FC <- 0.50
RNA_CLEAR_DETECTION_DELTA <- 5
ATAC_CLEAR_LOG2FC <- 0.50

TOP_GENES_PER_REFERENCE_DIRECTION <- 30L
TOP_LABELS_PER_REFERENCE_DIRECTION <- 12L

step8_parameters <- data.frame(
  parameter = c(
    "reference diet",
    "comparison diet",
    "primary WT stem-cell annotation",
    "TSS window upstream bp",
    "TSS window downstream bp",
    "minimum balanced-sampling sign concordance percent",
    "RNA clear-effect absolute log2 fold-change threshold",
    "RNA clear-effect absolute detection-difference threshold",
    "ATAC clear-effect absolute log2 fold-change threshold",
    "top genes per reference direction"
  ),
  value = c(
    REFERENCE_DIET,
    COMPARISON_DIET,
    WT_STEM_CELL_STATE,
    TSS_WINDOW_BP,
    TSS_WINDOW_BP,
    MINIMUM_SIGN_CONCORDANCE_PERCENT,
    RNA_CLEAR_LOG2FC,
    RNA_CLEAR_DETECTION_DELTA,
    ATAC_CLEAR_LOG2FC,
    TOP_GENES_PER_REFERENCE_DIRECTION
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  step8_parameters,
  file.path(STEP8_DIR, "persistent_hfd_validation_parameters.csv")
)
```

# 3. Read and validate the reference and upstream tables

```{r read-and-validate-inputs}
read_table_quietly <- function(path) {
  suppressMessages(
    readr::read_csv(
      path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
}

assert_columns <- function(data, required_columns, table_label) {
  missing_columns <- setdiff(required_columns, colnames(data))

  if (length(missing_columns) > 0L) {
    stop(
      table_label,
      " is missing required column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
}

as_logical_flag <- function(values, label) {
  normalized <- tolower(trimws(as.character(values)))
  result <- dplyr::case_when(
    normalized %in% c("true", "t", "1", "yes", "y") ~ TRUE,
    normalized %in% c("false", "f", "0", "no", "n") ~ FALSE,
    TRUE ~ NA
  )

  if (anyNA(result)) {
    stop(
      "Could not parse logical values in ",
      label,
      ": ",
      paste(unique(values[is.na(result)]), collapse = ", "),
      call. = FALSE
    )
  }

  result
}

reference_genes <- read_table_quietly(REFERENCE_GENE_PATH)
eligibility <- read_table_quietly(input_paths[["eligibility"]])
rna_effects <- read_table_quietly(input_paths[["RNA_effects"]])
rna_stability <- read_table_quietly(input_paths[["RNA_stability"]])
atac_effects <- read_table_quietly(input_paths[["ATAC_effects"]])
atac_stability <- read_table_quietly(input_paths[["ATAC_stability"]])
step7c_rna <- read_table_quietly(input_paths[["Step7C_RNA"]])
step7c_atac <- read_table_quietly(input_paths[["Step7C_ATAC"]])
step7c_multimodal <- read_table_quietly(
  input_paths[["Step7C_multimodal"]]
)

assert_columns(
  reference_genes,
  c(
    "gene", "reference_direction", "reference_program",
    "source_rank", "avg_log2FC_day21", "avg_log2FC_day50",
    "source_file"
  ),
  "persistent HFD reference gene table"
)

assert_columns(
  eligibility,
  c(
    "analysis_unit", "broad_compartment", "analysis_lineage",
    "cell_state", "cell_state_full", "annotation_confidence",
    "nuclei_CON", "nuclei_HFD", "minimum_nuclei_per_diet",
    "eligibility_tier",
    "primary_RNA_ATAC"
  ),
  "analysis population eligibility table"
)

effect_columns <- c(
  "analysis_unit", "feature", "feature_type",
  "reference_diet", "comparison_diet",
  "reference_nuclei", "comparison_nuclei",
  "reference_counts", "comparison_counts", "total_counts",
  "reference_CPM", "comparison_CPM",
  "log2FC_comparison_vs_reference",
  "reference_percent_detected", "comparison_percent_detected",
  "detection_difference_percentage_points",
  "cell_state", "cell_state_full", "eligibility_tier"
)

assert_columns(rna_effects, effect_columns, "complete RNA effect table")
assert_columns(atac_effects, effect_columns, "complete ATAC effect table")

stability_columns <- c(
  "analysis_unit", "feature", "sampled_nuclei_per_diet",
  "full_log2FC", "downsampled_log2FC_median",
  "downsampled_log2FC_q25", "downsampled_log2FC_q75",
  "sign_concordance_percent"
)

assert_columns(rna_stability, stability_columns, "RNA stability table")
assert_columns(atac_stability, stability_columns, "ATAC stability table")

assert_columns(
  step7c_rna,
  c(
    "analysis_unit", "feature", "RNA_support_class",
    "within_unit_direction_rank"
  ),
  "Step 7C RNA prioritization table"
)

assert_columns(
  step7c_atac,
  c(
    "analysis_unit", "feature", "ATAC_support_class",
    "within_unit_direction_rank"
  ),
  "Step 7C ATAC prioritization table"
)

assert_columns(
  step7c_multimodal,
  c(
    "analysis_unit", "nearest_gene", "peak",
    "multimodal_support_class", "support_order",
    "within_unit_direction_rank"
  ),
  "Step 7C multimodal prioritization table"
)

reference_genes <- reference_genes |>
  dplyr::mutate(
    source_gene = trimws(as.character(gene)),
    gene = dplyr::recode(
      source_gene,
      "Atpif1" = "Atp5if1"
    ),
    symbol_update = dplyr::if_else(
      source_gene == gene,
      NA_character_,
      paste0(source_gene, " renamed to ", gene)
    ),
    reference_direction = trimws(as.character(reference_direction)),
    reference_program = trimws(as.character(reference_program)),
    source_rank = as.integer(source_rank),
    avg_log2FC_day21 = as.numeric(avg_log2FC_day21),
    avg_log2FC_day50 = as.numeric(avg_log2FC_day50)
  )

if (
  anyNA(reference_genes$gene) ||
    any(!nzchar(reference_genes$gene))
) {
  stop("The persistent reference contains a missing gene symbol.")
}

if (
  anyNA(reference_genes$source_rank) ||
    anyNA(reference_genes$avg_log2FC_day21) ||
    anyNA(reference_genes$avg_log2FC_day50)
) {
  stop(
    "The persistent reference contains a missing rank or source effect.",
    call. = FALSE
  )
}

if (anyDuplicated(reference_genes$gene) > 0L) {
  duplicated_genes <- unique(
    reference_genes$gene[duplicated(reference_genes$gene)]
  )

  stop(
    "The persistent reference contains duplicate gene symbols: ",
    paste(duplicated_genes, collapse = ", "),
    call. = FALSE
  )
}

readr::write_csv(
  reference_genes |>
    dplyr::filter(!is.na(symbol_update)),
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_symbol_update_audit.csv"
  )
)

valid_reference_directions <- c("HFD_higher", "CON_higher")
invalid_reference_directions <- setdiff(
  unique(reference_genes$reference_direction),
  valid_reference_directions
)

if (length(invalid_reference_directions) > 0L) {
  stop(
    "Invalid persistent reference direction(s): ",
    paste(invalid_reference_directions, collapse = ", "),
    call. = FALSE
  )
}

invalid_program_mapping <- with(
  reference_genes,
  (
    reference_direction == "HFD_higher" &
      reference_program != "persistent_HFD_up"
  ) |
    (
      reference_direction == "CON_higher" &
        reference_program != "persistent_HFD_down"
    )
)

if (any(invalid_program_mapping)) {
  stop(
    "reference_program and reference_direction disagree for: ",
    paste(reference_genes$gene[invalid_program_mapping], collapse = ", "),
    call. = FALSE
  )
}

invalid_reference_sign <- with(
  reference_genes,
  (
    reference_direction == "HFD_higher" &
      (avg_log2FC_day21 <= 0 | avg_log2FC_day50 <= 0)
  ) |
    (
      reference_direction == "CON_higher" &
        (avg_log2FC_day21 >= 0 | avg_log2FC_day50 >= 0)
    )
)

if (any(invalid_reference_sign)) {
  stop(
    "Reference effect signs do not agree with reference_direction for: ",
    paste(reference_genes$gene[invalid_reference_sign], collapse = ", "),
    call. = FALSE
  )
}

eligibility$analysis_unit <- as.character(eligibility$analysis_unit)
eligibility$primary_RNA_ATAC <- as_logical_flag(
  eligibility$primary_RNA_ATAC,
  "eligibility$primary_RNA_ATAC"
)

rna_effects$analysis_unit <- as.character(rna_effects$analysis_unit)
rna_stability$analysis_unit <- as.character(rna_stability$analysis_unit)
atac_effects$analysis_unit <- as.character(atac_effects$analysis_unit)
atac_stability$analysis_unit <- as.character(atac_stability$analysis_unit)
step7c_rna$analysis_unit <- as.character(step7c_rna$analysis_unit)
step7c_atac$analysis_unit <- as.character(step7c_atac$analysis_unit)
step7c_multimodal$analysis_unit <- as.character(
  step7c_multimodal$analysis_unit
)

if (
  any(rna_effects$reference_diet != REFERENCE_DIET) ||
    any(rna_effects$comparison_diet != COMPARISON_DIET) ||
    any(atac_effects$reference_diet != REFERENCE_DIET) ||
    any(atac_effects$comparison_diet != COMPARISON_DIET)
) {
  stop(
    "The Step 7 effect tables do not encode CON as reference and HFD ",
    "as comparison.",
    call. = FALSE
  )
}

if (
  any(rna_effects$feature_type != "gene") ||
    any(atac_effects$feature_type != "peak")
) {
  stop(
    "The complete Step 7 effect tables contain an unexpected feature type.",
    call. = FALSE
  )
}

if (
  anyDuplicated(rna_effects[c("analysis_unit", "feature")]) > 0L ||
    anyDuplicated(atac_effects[c("analysis_unit", "feature")]) > 0L
) {
  stop(
    "A complete Step 7 effect table contains duplicate unit-feature keys.",
    call. = FALSE
  )
}

input_inventory <- data.frame(
  input = c(
    "persistent reference genes",
    names(input_paths)
  ),
  path = c(
    REFERENCE_GENE_PATH,
    unname(input_paths)
  ),
  rows = c(
    nrow(reference_genes),
    nrow(eligibility),
    nrow(rna_effects),
    nrow(rna_stability),
    nrow(atac_effects),
    nrow(atac_stability),
    nrow(step7c_rna),
    nrow(step7c_atac),
    nrow(step7c_multimodal)
  ),
  stringsAsFactors = FALSE
)

readr::write_csv(
  input_inventory,
  file.path(STEP8_DIR, "persistent_hfd_validation_input_inventory.csv")
)
```

# 4. Resolve the WT stem anchor and contextual units

The reference was generated from an annotated stem-cell population. The
primary WT validation target is therefore the single eligible annotation unit
named `ISC-like crypt`. Other crypt units test state breadth, the Reg4-positive
unit provides niche context, and the remaining eligible epithelial units show
whether a signal is broadly distributed.

```{r resolve-analysis-units}
unit_key <- eligibility |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::mutate(
    analysis_unit = as.character(analysis_unit),
    analysis_scope = dplyr::case_when(
      cell_state == WT_STEM_CELL_STATE ~ "primary_WT_ISC_like",
      analysis_lineage == "crypt" ~ "crypt_context",
      grepl("Reg4", cell_state, ignore.case = TRUE) |
        grepl("Reg4", cell_state_full, ignore.case = TRUE) ~
        "Reg4_secretory_context",
      TRUE ~ "other_primary_epithelial_context"
    ),
    analysis_scope_order = dplyr::case_when(
      analysis_scope == "primary_WT_ISC_like" ~ 1L,
      analysis_scope == "crypt_context" ~ 2L,
      analysis_scope == "Reg4_secretory_context" ~ 3L,
      TRUE ~ 4L
    ),
    analysis_unit_display = paste0(
      analysis_unit,
      ": ",
      cell_state
    )
  ) |>
  dplyr::arrange(
    analysis_scope_order,
    dplyr::desc(minimum_nuclei_per_diet),
    analysis_unit
  )

if (nrow(unit_key) == 0L) {
  stop("No primary RNA/ATAC units were found in Step 7.")
}

missing_rna_effect_units <- setdiff(
  unit_key$analysis_unit,
  unique(rna_effects$analysis_unit)
)

missing_atac_effect_units <- setdiff(
  unit_key$analysis_unit,
  unique(atac_effects$analysis_unit)
)

if (
  length(missing_rna_effect_units) > 0L ||
    length(missing_atac_effect_units) > 0L
) {
  stop(
    "A primary Step 7 unit is missing from a complete effect table. ",
    "RNA: ",
    paste(missing_rna_effect_units, collapse = ", "),
    "; ATAC: ",
    paste(missing_atac_effect_units, collapse = ", "),
    call. = FALSE
  )
}

if (any(unit_key$broad_compartment != "epithelial")) {
  stop(
    "Step 8 expected every primary RNA/ATAC unit to be epithelial.",
    call. = FALSE
  )
}

stem_unit_key <- unit_key |>
  dplyr::filter(cell_state == WT_STEM_CELL_STATE)

if (nrow(stem_unit_key) != 1L) {
  stop(
    "Expected exactly one eligible WT unit annotated as '",
    WT_STEM_CELL_STATE,
    "'; found ",
    nrow(stem_unit_key),
    ".",
    call. = FALSE
  )
}

WT_STEM_UNIT <- stem_unit_key$analysis_unit[[1]]

readr::write_csv(
  unit_key,
  file.path(STEP8_TABLE_DIR, "WT_persistent_gene_analysis_unit_key.csv")
)

unit_key |>
  dplyr::select(
    analysis_unit,
    cell_state,
    cell_state_full,
    analysis_scope,
    nuclei_CON,
    nuclei_HFD,
    eligibility_tier
  )
```

# 5. Load feature inventories from the Step 6 object

The large object is loaded only long enough to retain RNA feature names and
ATAC peak coordinates. Expression and accessibility effects are read from the
completed Step 7 tables.

```{r load-feature-inventories}
multiome <- readRDS(STEP6_OBJECT_PATH)

required_assays <- c("RNA", "ATAC")
missing_assays <- setdiff(required_assays, names(multiome))

if (length(missing_assays) > 0L) {
  stop(
    "The Step 6 object is missing assay(s): ",
    paste(missing_assays, collapse = ", "),
    call. = FALSE
  )
}

rna_feature_inventory <- rownames(multiome[["RNA"]])
atac_feature_inventory <- rownames(multiome[["ATAC"]])
atac_ranges <- GenomicRanges::granges(multiome[["ATAC"]])

if (
  length(atac_feature_inventory) != length(atac_ranges) ||
    anyDuplicated(atac_feature_inventory) > 0L
) {
  stop(
    "ATAC feature names and genomic ranges are not one-to-one.",
    call. = FALSE
  )
}

names(atac_ranges) <- atac_feature_inventory

missing_effect_peaks <- setdiff(
  unique(atac_effects$feature),
  atac_feature_inventory
)

if (length(missing_effect_peaks) > 0L) {
  stop(
    "The complete ATAC effect table contains peaks absent from the ",
    "Step 6 ATAC assay.",
    call. = FALSE
  )
}

reference_feature_audit <- reference_genes |>
  dplyr::mutate(
    gene_in_RNA_assay = gene %in% rna_feature_inventory
  )

rm(multiome)
invisible(gc())
```

# 6. Test every persistent gene against WT RNA effects

`RNA_reference_aligned_log2FC` is positive when the WT direction agrees with
the persistent reference and negative when it opposes the reference,
regardless of whether the source gene was HFD-up or HFD-down.

```{r prepare-stability-and-priority-keys}
prepare_stability <- function(stability_table, prefix) {
  stability_table |>
    dplyr::mutate(
      direction_stable =
        sign(full_log2FC) == sign(downsampled_log2FC_median) &
        sign_concordance_percent >=
          MINIMUM_SIGN_CONCORDANCE_PERCENT,
      downsampling_iqr_excludes_zero = dplyr::case_when(
        full_log2FC > 0 ~ downsampled_log2FC_q25 > 0,
        full_log2FC < 0 ~ downsampled_log2FC_q75 < 0,
        TRUE ~ FALSE
      )
    ) |>
    dplyr::rename_with(
      ~ paste0(prefix, .x),
      -dplyr::all_of(c("analysis_unit", "feature"))
    )
}

rna_stability_key <- prepare_stability(
  rna_stability,
  "RNA_"
) |>
  dplyr::rename(gene = feature)

atac_stability_key <- prepare_stability(
  atac_stability,
  "ATAC_"
) |>
  dplyr::rename(peak = feature)

step7c_rna_key <- step7c_rna |>
  dplyr::transmute(
    analysis_unit,
    gene = feature,
    Step7C_RNA_support_class = RNA_support_class,
    Step7C_RNA_rank = within_unit_direction_rank
  )

step7c_atac_key <- step7c_atac |>
  dplyr::transmute(
    analysis_unit,
    peak = feature,
    Step7C_ATAC_support_class = ATAC_support_class,
    Step7C_ATAC_rank = within_unit_direction_rank
  )

step7c_multimodal_key <- step7c_multimodal |>
  dplyr::filter(nearest_gene %in% reference_genes$gene) |>
  dplyr::arrange(
    analysis_unit,
    nearest_gene,
    support_order,
    within_unit_direction_rank
  ) |>
  dplyr::group_by(analysis_unit, nearest_gene) |>
  dplyr::summarise(
    Step7C_nearest_gene_concordant_pairs = dplyr::n(),
    Step7C_best_multimodal_support_class =
      dplyr::first(multimodal_support_class),
    Step7C_best_multimodal_rank =
      dplyr::first(within_unit_direction_rank),
    .groups = "drop"
  ) |>
  dplyr::rename(gene = nearest_gene)
```

```{r build-persistent-gene-rna-evidence}
rna_effect_key <- rna_effects |>
  dplyr::select(
    analysis_unit,
    feature,
    reference_diet,
    comparison_diet,
    reference_nuclei,
    comparison_nuclei,
    reference_counts,
    comparison_counts,
    total_counts,
    reference_CPM,
    comparison_CPM,
    log2FC_comparison_vs_reference,
    reference_percent_detected,
    comparison_percent_detected,
    detection_difference_percentage_points
  ) |>
  dplyr::rename(gene = feature) |>
  dplyr::mutate(RNA_effect_available = TRUE)

rna_evidence <- tidyr::crossing(
  reference_genes,
  unit_key |>
    dplyr::select(
      analysis_unit,
      cell_state,
      cell_state_full,
      analysis_lineage,
      eligibility_tier,
      analysis_scope,
      analysis_scope_order,
      analysis_unit_display
    )
) |>
  dplyr::left_join(
    rna_effect_key,
    by = c("analysis_unit", "gene")
  ) |>
  dplyr::left_join(
    rna_stability_key,
    by = c("analysis_unit", "gene")
  ) |>
  dplyr::left_join(
    step7c_rna_key,
    by = c("analysis_unit", "gene")
  ) |>
  dplyr::mutate(
    gene_in_RNA_assay = gene %in% rna_feature_inventory,
    RNA_effect_available = dplyr::coalesce(
      RNA_effect_available,
      FALSE
    ),
    RNA_observed_direction = dplyr::case_when(
      !RNA_effect_available ~ NA_character_,
      log2FC_comparison_vs_reference > 0 ~ "HFD_higher",
      log2FC_comparison_vs_reference < 0 ~ "CON_higher",
      TRUE ~ "no_change"
    ),
    RNA_reference_aligned_log2FC = dplyr::case_when(
      !RNA_effect_available ~ NA_real_,
      reference_direction == "HFD_higher" ~
        log2FC_comparison_vs_reference,
      TRUE ~ -log2FC_comparison_vs_reference
    ),
    RNA_reference_direction_concordant = dplyr::case_when(
      !RNA_effect_available ~ NA,
      RNA_reference_aligned_log2FC > 0 ~ TRUE,
      RNA_reference_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    RNA_clear_effect =
      RNA_effect_available &
      abs(log2FC_comparison_vs_reference) >= RNA_CLEAR_LOG2FC &
      abs(detection_difference_percentage_points) >=
        RNA_CLEAR_DETECTION_DELTA,
    RNA_stability_available = !is.na(RNA_sign_concordance_percent),
    RNA_direction_stable = dplyr::coalesce(
      RNA_direction_stable,
      FALSE
    ),
    RNA_downsampling_iqr_excludes_zero = dplyr::coalesce(
      RNA_downsampling_iqr_excludes_zero,
      FALSE
    ),
    Step7C_RNA_candidate = !is.na(Step7C_RNA_rank),
    RNA_validation_class = dplyr::case_when(
      !gene_in_RNA_assay ~
        "absent from WT RNA assay",
      !RNA_effect_available ~
        "present but below Step 7 RNA filters",
      RNA_reference_aligned_log2FC == 0 ~
        "no directional RNA change",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero ~
        "reference-concordant clear RNA effect; stable with IQR excluding zero",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        RNA_direction_stable ~
        "reference-concordant clear RNA effect; stable direction",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        RNA_stability_available ~
        "reference-concordant clear RNA effect; unstable in downsampling",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect ~
        "reference-concordant clear RNA effect; stability not evaluated",
      RNA_reference_aligned_log2FC > 0 ~
        "reference-concordant subthreshold RNA effect",
      RNA_reference_aligned_log2FC < 0 & RNA_clear_effect ~
        "reference-discordant clear RNA effect",
      RNA_reference_aligned_log2FC < 0 ~
        "reference-discordant subthreshold RNA effect",
      TRUE ~ "unclassified"
    )
  ) |>
  dplyr::arrange(
    analysis_scope_order,
    analysis_unit,
    reference_direction,
    source_rank
  )

stability_effect_mismatch <- rna_evidence |>
  dplyr::filter(RNA_stability_available) |>
  dplyr::filter(
    abs(
      RNA_full_log2FC -
        log2FC_comparison_vs_reference
    ) > 1e-8
  )

if (nrow(stability_effect_mismatch) > 0L) {
  stop(
    "RNA stability and complete effect tables disagree for at least one ",
    "unit-gene pair.",
    call. = FALSE
  )
}

readr::write_csv(
  rna_evidence,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_WT_RNA_evidence_by_unit.csv"
  )
)
```

## Primary WT stem-state summary

```{r summarize-primary-stem-rna-evidence}
stem_rna_evidence <- rna_evidence |>
  dplyr::filter(analysis_unit == WT_STEM_UNIT) |>
  dplyr::mutate(
    primary_WT_stem_status = dplyr::case_when(
      !gene_in_RNA_assay ~ "absent from WT RNA assay",
      !RNA_effect_available ~ "below Step 7 RNA filters",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        RNA_direction_stable ~
        "reference-concordant clear and stable",
      RNA_reference_aligned_log2FC > 0 & RNA_clear_effect ~
        "reference-concordant clear",
      RNA_reference_aligned_log2FC > 0 ~
        "reference-concordant subthreshold",
      RNA_reference_aligned_log2FC < 0 & RNA_clear_effect ~
        "reference-discordant clear",
      RNA_reference_aligned_log2FC < 0 ~
        "reference-discordant subthreshold",
      TRUE ~ "no directional RNA change"
    )
  )

crypt_context_summary <- rna_evidence |>
  dplyr::filter(
    analysis_scope %in% c(
      "primary_WT_ISC_like",
      "crypt_context"
    )
  ) |>
  dplyr::group_by(gene) |>
  dplyr::summarise(
    crypt_units_evaluable = sum(RNA_effect_available),
    crypt_units_reference_concordant = sum(
      RNA_reference_direction_concordant %in% TRUE
    ),
    crypt_units_reference_discordant = sum(
      RNA_reference_direction_concordant %in% FALSE
    ),
    crypt_units_clear_concordant = sum(
      RNA_reference_direction_concordant %in% TRUE &
        RNA_clear_effect
    ),
    crypt_units_clear_discordant = sum(
      RNA_reference_direction_concordant %in% FALSE &
        RNA_clear_effect
    ),
    crypt_units_stable_concordant = sum(
      RNA_reference_direction_concordant %in% TRUE &
        RNA_direction_stable
    ),
    .groups = "drop"
  )

persistent_gene_rna_summary <- stem_rna_evidence |>
  dplyr::select(
    gene,
    reference_direction,
    reference_program,
    source_rank,
    avg_log2FC_day21,
    avg_log2FC_day50,
    source_file,
    analysis_unit,
    cell_state,
    cell_state_full,
    gene_in_RNA_assay,
    RNA_effect_available,
    RNA_observed_direction,
    RNA_reference_aligned_log2FC,
    log2FC_comparison_vs_reference,
    detection_difference_percentage_points,
    reference_percent_detected,
    comparison_percent_detected,
    RNA_clear_effect,
    RNA_stability_available,
    RNA_direction_stable,
    RNA_downsampling_iqr_excludes_zero,
    RNA_sign_concordance_percent,
    Step7C_RNA_candidate,
    Step7C_RNA_support_class,
    Step7C_RNA_rank,
    primary_WT_stem_status
  ) |>
  dplyr::left_join(crypt_context_summary, by = "gene") |>
  dplyr::arrange(reference_direction, source_rank)

readr::write_csv(
  persistent_gene_rna_summary,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_primary_WT_stem_RNA_summary.csv"
  )
)

persistent_gene_rna_summary |>
  dplyr::count(
    reference_program,
    primary_WT_stem_status,
    name = "genes"
  )
```

# 7. Build exact strand-aware plus-or-minus 2-kb TSS windows

Each GENCODE gene record contributes one strand-aware TSS. The reported
window spans 2,000 bp upstream through 2,000 bp downstream in genomic
coordinates. `transcriptional_offset_bp` uses the gene strand, so positive
values are downstream of the TSS.

```{r build-persistent-gene-promoter-windows}
annotation <- build_grcm39_annotation(GRCM39_GTF)

gene_annotation <- annotation[
  as.character(annotation$type) == "gene"
]

gene_annotation$gene_name <- as.character(
  gene_annotation$gene_name
)
gene_annotation$gene_id <- as.character(
  gene_annotation$gene_id
)

gene_annotation <- gene_annotation[
  gene_annotation$gene_name %in% reference_genes$gene
]

if (length(gene_annotation) == 0L) {
  stop(
    "None of the persistent reference genes were found in the GRCm39 GTF.",
    call. = FALSE
  )
}

gene_strand <- as.character(GenomicRanges::strand(gene_annotation))

if (any(!gene_strand %in% c("+", "-"))) {
  stop(
    "At least one persistent reference gene has an unresolved strand.",
    call. = FALSE
  )
}

tss_position <- ifelse(
  gene_strand == "-",
  GenomicRanges::end(gene_annotation),
  GenomicRanges::start(gene_annotation)
)

promoter_ranges <- GenomicRanges::GRanges(
  seqnames = GenomeInfoDb::seqnames(gene_annotation),
  ranges = IRanges::IRanges(
    start = pmax(1L, tss_position - TSS_WINDOW_BP),
    end = tss_position + TSS_WINDOW_BP
  ),
  strand = GenomicRanges::strand(gene_annotation),
  gene = gene_annotation$gene_name,
  gene_id = gene_annotation$gene_id,
  tss = as.integer(tss_position)
)

names(promoter_ranges) <- make.unique(
  paste0(
    promoter_ranges$gene,
    "|",
    promoter_ranges$gene_id
  )
)

promoter_window_table <- data.frame(
  gene = as.character(promoter_ranges$gene),
  gene_id = as.character(promoter_ranges$gene_id),
  chromosome = as.character(
    GenomeInfoDb::seqnames(promoter_ranges)
  ),
  strand = as.character(GenomicRanges::strand(promoter_ranges)),
  tss = as.integer(promoter_ranges$tss),
  promoter_start = GenomicRanges::start(promoter_ranges),
  promoter_end = GenomicRanges::end(promoter_ranges),
  stringsAsFactors = FALSE
) |>
  dplyr::left_join(
    reference_genes,
    by = "gene"
  )

readr::write_csv(
  promoter_window_table,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_GRCm39_TSS_windows.csv"
  )
)
```

```{r find-exact-promoter-peak-overlaps}
compatible_promoter <- as.character(
  GenomeInfoDb::seqnames(promoter_ranges)
) %in% GenomeInfoDb::seqlevels(atac_ranges)

promoter_ranges_compatible <- promoter_ranges[compatible_promoter]

promoter_hits <- GenomicRanges::findOverlaps(
  query = promoter_ranges_compatible,
  subject = atac_ranges,
  ignore.strand = TRUE
)

if (length(promoter_hits) == 0L) {
  promoter_peak_overlaps <- tibble::tibble(
    gene = character(),
    gene_id = character(),
    chromosome = character(),
    strand = character(),
    tss = integer(),
    promoter_start = integer(),
    promoter_end = integer(),
    peak = character(),
    peak_start = integer(),
    peak_end = integer(),
    peak_center = integer(),
    transcriptional_offset_bp = integer(),
    peak_overlaps_tss = logical()
  )
} else {
  hit_promoters <- promoter_ranges_compatible[
    S4Vectors::queryHits(promoter_hits)
  ]

  hit_peaks <- atac_ranges[
    S4Vectors::subjectHits(promoter_hits)
  ]

  peak_center <- floor(
    (
      GenomicRanges::start(hit_peaks) +
        GenomicRanges::end(hit_peaks)
    ) / 2
  )

  transcriptional_offset <- ifelse(
    as.character(GenomicRanges::strand(hit_promoters)) == "-",
    as.integer(hit_promoters$tss) - peak_center,
    peak_center - as.integer(hit_promoters$tss)
  )

  promoter_peak_overlaps <- tibble::tibble(
    gene = as.character(hit_promoters$gene),
    gene_id = as.character(hit_promoters$gene_id),
    chromosome = as.character(
      GenomeInfoDb::seqnames(hit_promoters)
    ),
    strand = as.character(GenomicRanges::strand(hit_promoters)),
    tss = as.integer(hit_promoters$tss),
    promoter_start = GenomicRanges::start(hit_promoters),
    promoter_end = GenomicRanges::end(hit_promoters),
    peak = names(hit_peaks),
    peak_start = GenomicRanges::start(hit_peaks),
    peak_end = GenomicRanges::end(hit_peaks),
    peak_center = as.integer(peak_center),
    transcriptional_offset_bp = as.integer(
      transcriptional_offset
    ),
    peak_overlaps_tss =
      GenomicRanges::start(hit_peaks) <=
        as.integer(hit_promoters$tss) &
      GenomicRanges::end(hit_peaks) >=
        as.integer(hit_promoters$tss)
  ) |>
    dplyr::distinct(gene_id, peak, .keep_all = TRUE)
}

promoter_peak_overlaps <- promoter_peak_overlaps |>
  dplyr::left_join(
    reference_genes,
    by = "gene"
  )

readr::write_csv(
  promoter_peak_overlaps,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_exact_promoter_peak_overlaps.csv"
  )
)

promoter_peak_counts <- promoter_peak_overlaps |>
  dplyr::count(gene, name = "promoter_peaks_in_ATAC_assay")

gtf_gene_counts <- promoter_window_table |>
  dplyr::count(gene, name = "GRCm39_gene_loci")

reference_feature_audit <- reference_feature_audit |>
  dplyr::left_join(gtf_gene_counts, by = "gene") |>
  dplyr::left_join(promoter_peak_counts, by = "gene") |>
  dplyr::mutate(
    gene_in_GRCm39_GTF = !is.na(GRCm39_gene_loci),
    GRCm39_gene_loci = dplyr::coalesce(
      GRCm39_gene_loci,
      0L
    ),
    promoter_peaks_in_ATAC_assay = dplyr::coalesce(
      promoter_peaks_in_ATAC_assay,
      0L
    )
  )

readr::write_csv(
  reference_feature_audit,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_feature_inventory_audit.csv"
  )
)
```

# 8. Quantify promoter-overlapping ATAC effects

Every exact promoter-overlapping peak is retained. The representative peak is
the evaluable promoter peak with the largest absolute ATAC effect. It is a
summary for plotting and does not replace the complete peak-level table.

```{r build-promoter-atac-evidence}
atac_effect_key <- atac_effects |>
  dplyr::select(
    analysis_unit,
    feature,
    reference_diet,
    comparison_diet,
    reference_nuclei,
    comparison_nuclei,
    reference_counts,
    comparison_counts,
    total_counts,
    reference_CPM,
    comparison_CPM,
    log2FC_comparison_vs_reference,
    reference_percent_detected,
    comparison_percent_detected,
    detection_difference_percentage_points
  ) |>
  dplyr::rename(peak = feature) |>
  dplyr::mutate(ATAC_effect_available = TRUE)

if (nrow(promoter_peak_overlaps) == 0L) {
  promoter_atac_evidence <- tibble::tibble()
} else {
  promoter_atac_evidence <- tidyr::crossing(
    promoter_peak_overlaps,
    unit_key |>
      dplyr::select(
        analysis_unit,
        cell_state,
        cell_state_full,
        analysis_lineage,
        eligibility_tier,
        analysis_scope,
        analysis_scope_order,
        analysis_unit_display
      )
  ) |>
    dplyr::left_join(
      atac_effect_key,
      by = c("analysis_unit", "peak")
    ) |>
    dplyr::left_join(
      atac_stability_key,
      by = c("analysis_unit", "peak")
    ) |>
    dplyr::left_join(
      step7c_atac_key,
      by = c("analysis_unit", "peak")
    ) |>
    dplyr::mutate(
      ATAC_effect_available = dplyr::coalesce(
        ATAC_effect_available,
        FALSE
      ),
      ATAC_observed_direction = dplyr::case_when(
        !ATAC_effect_available ~ NA_character_,
        log2FC_comparison_vs_reference > 0 ~ "HFD_higher",
        log2FC_comparison_vs_reference < 0 ~ "CON_higher",
        TRUE ~ "no_change"
      ),
      ATAC_reference_aligned_log2FC = dplyr::case_when(
        !ATAC_effect_available ~ NA_real_,
        reference_direction == "HFD_higher" ~
          log2FC_comparison_vs_reference,
        TRUE ~ -log2FC_comparison_vs_reference
      ),
      ATAC_reference_direction_concordant = dplyr::case_when(
        !ATAC_effect_available ~ NA,
        ATAC_reference_aligned_log2FC > 0 ~ TRUE,
        ATAC_reference_aligned_log2FC < 0 ~ FALSE,
        TRUE ~ NA
      ),
      ATAC_clear_effect =
        ATAC_effect_available &
        abs(log2FC_comparison_vs_reference) >=
          ATAC_CLEAR_LOG2FC,
      ATAC_stability_available =
        !is.na(ATAC_sign_concordance_percent),
      ATAC_direction_stable = dplyr::coalesce(
        ATAC_direction_stable,
        FALSE
      ),
      ATAC_downsampling_iqr_excludes_zero = dplyr::coalesce(
        ATAC_downsampling_iqr_excludes_zero,
        FALSE
      ),
      Step7C_ATAC_candidate = !is.na(Step7C_ATAC_rank),
      promoter_ATAC_validation_class = dplyr::case_when(
        !ATAC_effect_available ~
          "promoter peak below Step 7 ATAC filters",
        ATAC_reference_aligned_log2FC == 0 ~
          "no directional promoter ATAC change",
        ATAC_reference_aligned_log2FC > 0 &
          ATAC_clear_effect &
          ATAC_direction_stable &
          ATAC_downsampling_iqr_excludes_zero ~
          "reference-concordant clear promoter ATAC effect; stable with IQR excluding zero",
        ATAC_reference_aligned_log2FC > 0 &
          ATAC_clear_effect &
          ATAC_direction_stable ~
          "reference-concordant clear promoter ATAC effect; stable direction",
        ATAC_reference_aligned_log2FC > 0 &
          ATAC_clear_effect ~
          "reference-concordant clear promoter ATAC effect",
        ATAC_reference_aligned_log2FC > 0 ~
          "reference-concordant subthreshold promoter ATAC effect",
        ATAC_reference_aligned_log2FC < 0 &
          ATAC_clear_effect ~
          "reference-discordant clear promoter ATAC effect",
        ATAC_reference_aligned_log2FC < 0 ~
          "reference-discordant subthreshold promoter ATAC effect",
        TRUE ~ "unclassified"
      )
    ) |>
    dplyr::arrange(
      analysis_scope_order,
      analysis_unit,
      reference_direction,
      source_rank,
      transcriptional_offset_bp
    )
}

readr::write_csv(
  promoter_atac_evidence,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_promoter_ATAC_evidence_by_unit_and_peak.csv.gz"
  )
)
```

```{r summarize-promoter-atac-evidence}
if (nrow(promoter_atac_evidence) == 0L) {
  promoter_atac_summary <- tibble::tibble(
    gene = character(),
    analysis_unit = character(),
    promoter_peaks_in_assay = integer(),
    promoter_peaks_evaluable = integer(),
    promoter_peaks_reference_concordant = integer(),
    promoter_peaks_reference_discordant = integer(),
    promoter_peaks_clear_concordant = integer(),
    promoter_peaks_clear_discordant = integer(),
    promoter_peaks_stable_concordant = integer()
  )

  representative_promoter_peak <- tibble::tibble(
    gene = character(),
    analysis_unit = character(),
    representative_promoter_peak = character(),
    representative_peak_offset_bp = integer(),
    representative_peak_overlaps_tss = logical(),
    representative_ATAC_log2FC = double(),
    representative_ATAC_reference_aligned_log2FC = double(),
    representative_ATAC_detection_difference = double(),
    representative_ATAC_direction_stable = logical(),
    representative_ATAC_IQR_excludes_zero = logical(),
    representative_Step7C_ATAC_support_class = character(),
    representative_Step7C_ATAC_rank = integer()
  )
} else {
  promoter_atac_summary <- promoter_atac_evidence |>
    dplyr::group_by(gene, analysis_unit) |>
    dplyr::summarise(
      promoter_peaks_in_assay = dplyr::n_distinct(peak),
      promoter_peaks_evaluable = dplyr::n_distinct(
        peak[ATAC_effect_available]
      ),
      promoter_peaks_reference_concordant = dplyr::n_distinct(
        peak[
          ATAC_effect_available &
            ATAC_reference_direction_concordant %in% TRUE
        ]
      ),
      promoter_peaks_reference_discordant = dplyr::n_distinct(
        peak[
          ATAC_effect_available &
            ATAC_reference_direction_concordant %in% FALSE
        ]
      ),
      promoter_peaks_clear_concordant = dplyr::n_distinct(
        peak[
          ATAC_effect_available &
            ATAC_reference_direction_concordant %in% TRUE &
            ATAC_clear_effect
        ]
      ),
      promoter_peaks_clear_discordant = dplyr::n_distinct(
        peak[
          ATAC_effect_available &
            ATAC_reference_direction_concordant %in% FALSE &
            ATAC_clear_effect
        ]
      ),
      promoter_peaks_stable_concordant = dplyr::n_distinct(
        peak[
          ATAC_effect_available &
            ATAC_reference_direction_concordant %in% TRUE &
            ATAC_direction_stable
        ]
      ),
      .groups = "drop"
    )

  representative_promoter_peak <- promoter_atac_evidence |>
    dplyr::filter(ATAC_effect_available) |>
    dplyr::mutate(
      absolute_ATAC_log2FC =
        abs(log2FC_comparison_vs_reference)
    ) |>
    dplyr::group_by(gene, analysis_unit) |>
    dplyr::arrange(
      dplyr::desc(absolute_ATAC_log2FC),
      abs(transcriptional_offset_bp),
      .by_group = TRUE
    ) |>
    dplyr::slice_head(n = 1L) |>
    dplyr::ungroup() |>
    dplyr::transmute(
      gene,
      analysis_unit,
      representative_promoter_peak = peak,
      representative_peak_offset_bp =
        transcriptional_offset_bp,
      representative_peak_overlaps_tss = peak_overlaps_tss,
      representative_ATAC_log2FC =
        log2FC_comparison_vs_reference,
      representative_ATAC_reference_aligned_log2FC =
        ATAC_reference_aligned_log2FC,
      representative_ATAC_detection_difference =
        detection_difference_percentage_points,
      representative_ATAC_direction_stable =
        ATAC_direction_stable,
      representative_ATAC_IQR_excludes_zero =
        ATAC_downsampling_iqr_excludes_zero,
      representative_Step7C_ATAC_support_class =
        Step7C_ATAC_support_class,
      representative_Step7C_ATAC_rank =
        Step7C_ATAC_rank
    )
}
```

# 9. Integrate persistent RNA and exact promoter ATAC evidence

The primary shortlist favors agreement with the source direction in the WT
ISC-like state. Exact promoter ATAC support raises priority but is not
required, because RNA and chromatin accessibility need not change uniformly
or simultaneously.

```{r integrate-rna-and-promoter-atac-evidence}
integrated_evidence <- rna_evidence |>
  dplyr::left_join(
    promoter_atac_summary,
    by = c("gene", "analysis_unit")
  ) |>
  dplyr::left_join(
    representative_promoter_peak,
    by = c("gene", "analysis_unit")
  ) |>
  dplyr::left_join(
    step7c_multimodal_key,
    by = c("gene", "analysis_unit")
  ) |>
  dplyr::mutate(
    promoter_peaks_in_assay = dplyr::coalesce(
      promoter_peaks_in_assay,
      0L
    ),
    promoter_peaks_evaluable = dplyr::coalesce(
      promoter_peaks_evaluable,
      0L
    ),
    promoter_peaks_reference_concordant = dplyr::coalesce(
      promoter_peaks_reference_concordant,
      0L
    ),
    promoter_peaks_reference_discordant = dplyr::coalesce(
      promoter_peaks_reference_discordant,
      0L
    ),
    promoter_peaks_clear_concordant = dplyr::coalesce(
      promoter_peaks_clear_concordant,
      0L
    ),
    promoter_peaks_clear_discordant = dplyr::coalesce(
      promoter_peaks_clear_discordant,
      0L
    ),
    promoter_peaks_stable_concordant = dplyr::coalesce(
      promoter_peaks_stable_concordant,
      0L
    ),
    Step7C_nearest_gene_concordant_pairs = dplyr::coalesce(
      Step7C_nearest_gene_concordant_pairs,
      0L
    ),
    promoter_ATAC_summary_class = dplyr::case_when(
      promoter_peaks_in_assay == 0L ~
        "no assayed ATAC peak overlaps the TSS window",
      promoter_peaks_evaluable == 0L ~
        "promoter peaks present but below Step 7 ATAC filters",
      promoter_peaks_clear_concordant > 0L &
        promoter_peaks_clear_discordant > 0L ~
        "mixed clear promoter ATAC directions",
      promoter_peaks_clear_concordant > 0L ~
        "reference-concordant clear promoter ATAC support",
      promoter_peaks_clear_discordant > 0L ~
        "reference-discordant clear promoter ATAC signal",
      promoter_peaks_reference_concordant > 0L &
        promoter_peaks_reference_discordant > 0L ~
        "mixed subthreshold promoter ATAC directions",
      promoter_peaks_reference_concordant > 0L ~
        "reference-concordant subthreshold promoter ATAC support",
      promoter_peaks_reference_discordant > 0L ~
        "reference-discordant subthreshold promoter ATAC signal",
      TRUE ~ "no directional promoter ATAC change"
    ),
    multimodal_validation_class = dplyr::case_when(
      !gene_in_RNA_assay ~
        "reference gene absent from WT RNA assay",
      !RNA_effect_available ~
        "RNA below Step 7 filters; promoter context not primary",
      RNA_reference_aligned_log2FC < 0 & RNA_clear_effect ~
        "clear WT RNA effect opposes persistent reference",
      RNA_reference_aligned_log2FC < 0 ~
        "subthreshold WT RNA effect opposes persistent reference",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        promoter_peaks_clear_concordant > 0L &
        promoter_peaks_clear_discordant == 0L ~
        "WT RNA and exact promoter ATAC support persistent reference",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        promoter_peaks_clear_concordant > 0L &
        promoter_peaks_clear_discordant > 0L ~
        "WT RNA supports reference; exact promoter ATAC is mixed",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        promoter_peaks_clear_discordant > 0L ~
        "WT RNA supports reference; exact promoter ATAC opposes",
      RNA_reference_aligned_log2FC > 0 & RNA_clear_effect ~
        "clear WT RNA support; exact promoter ATAC unresolved",
      RNA_reference_aligned_log2FC > 0 ~
        "subthreshold WT RNA support; promoter context secondary",
      TRUE ~ "no directional WT RNA support"
    ),
    priority_order = dplyr::case_when(
      multimodal_validation_class ==
        "WT RNA and exact promoter ATAC support persistent reference" &
        RNA_direction_stable &
        promoter_peaks_stable_concordant > 0L ~ 1L,
      multimodal_validation_class ==
        "WT RNA and exact promoter ATAC support persistent reference" ~ 2L,
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect &
        RNA_direction_stable ~ 3L,
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect ~ 4L,
      RNA_reference_aligned_log2FC > 0 ~ 5L,
      !RNA_effect_available ~ 6L,
      RNA_reference_aligned_log2FC < 0 ~ 7L,
      TRUE ~ 8L
    )
  ) |>
  dplyr::arrange(
    analysis_scope_order,
    priority_order,
    dplyr::desc(RNA_reference_aligned_log2FC),
    reference_direction,
    source_rank
  )

readr::write_csv(
  integrated_evidence,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_WT_RNA_promoter_ATAC_evidence_by_unit.csv"
  )
)

primary_stem_integrated <- integrated_evidence |>
  dplyr::filter(analysis_unit == WT_STEM_UNIT) |>
  dplyr::arrange(
    priority_order,
    dplyr::desc(RNA_reference_aligned_log2FC),
    reference_direction,
    source_rank
  )

readr::write_csv(
  primary_stem_integrated,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_primary_WT_stem_integrated_evidence.csv"
  )
)
```

## Create the focused review shortlist

```{r create-persistent-gene-review-shortlist}
persistent_gene_review_shortlist <- primary_stem_integrated |>
  dplyr::filter(
    RNA_effect_available,
    RNA_reference_aligned_log2FC > 0
  ) |>
  dplyr::group_by(reference_direction) |>
  dplyr::arrange(
    priority_order,
    dplyr::desc(RNA_reference_aligned_log2FC),
    dplyr::desc(
      abs(detection_difference_percentage_points)
    ),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = TOP_GENES_PER_REFERENCE_DIRECTION) |>
  dplyr::mutate(
    within_reference_direction_rank = dplyr::row_number()
  ) |>
  dplyr::ungroup()

readr::write_csv(
  persistent_gene_review_shortlist,
  file.path(
    STEP8_TABLE_DIR,
    "persistent_gene_primary_WT_stem_review_shortlist.csv"
  )
)

persistent_gene_review_shortlist |>
  dplyr::select(
    reference_direction,
    within_reference_direction_rank,
    gene,
    RNA_reference_aligned_log2FC,
    RNA_direction_stable,
    promoter_ATAC_summary_class,
    multimodal_validation_class
  )
```

# 10. Visual review

## Reference-direction recovery in the WT ISC-like state

```{r plot-primary-stem-rna-classification, fig.width=12, fig.height=8}
primary_status_order <- c(
  "reference-concordant clear and stable",
  "reference-concordant clear",
  "reference-concordant subthreshold",
  "reference-discordant subthreshold",
  "reference-discordant clear",
  "below Step 7 RNA filters",
  "absent from WT RNA assay",
  "no directional RNA change"
)

primary_status_plot_data <- stem_rna_evidence |>
  dplyr::mutate(
    primary_WT_stem_status = factor(
      primary_WT_stem_status,
      levels = rev(primary_status_order)
    )
  ) |>
  dplyr::count(
    reference_program,
    primary_WT_stem_status,
    name = "genes"
  )

p_primary_status <- ggplot2::ggplot(
  primary_status_plot_data,
  ggplot2::aes(
    x = primary_WT_stem_status,
    y = genes,
    fill = reference_program
  )
) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::coord_flip() +
  ggplot2::labs(
    title = "Persistent reference directions in the WT ISC-like state",
    subtitle = paste(
      "Clear effects require the Step 7 RNA log2FC and detection-difference",
      "thresholds; stability is balanced cell-sampling support."
    ),
    x = NULL,
    y = "Reference genes",
    fill = "Reference program"
  ) +
  theme_multiome(base_size = 10)

print(p_primary_status)

ggplot2::ggsave(
  file.path(
    STEP8_PLOT_DIR,
    "01_primary_WT_stem_RNA_validation_classes.pdf"
  ),
  p_primary_status,
  width = 12,
  height = 8
)
```

## Compare source and WT RNA effect sizes

```{r plot-source-versus-wt-rna, fig.width=12, fig.height=8}
source_wt_plot_data <- primary_stem_integrated |>
  dplyr::filter(RNA_effect_available) |>
  dplyr::mutate(
    agreement = dplyr::case_when(
      RNA_reference_aligned_log2FC > 0 ~ "reference-concordant",
      RNA_reference_aligned_log2FC < 0 ~ "reference-discordant",
      TRUE ~ "no directional change"
    )
  )

source_wt_labels <- source_wt_plot_data |>
  dplyr::group_by(reference_direction) |>
  dplyr::slice_max(
    order_by = abs(RNA_reference_aligned_log2FC),
    n = TOP_LABELS_PER_REFERENCE_DIRECTION,
    with_ties = FALSE
  ) |>
  dplyr::ungroup()

p_source_wt <- ggplot2::ggplot(
  source_wt_plot_data,
  ggplot2::aes(
    x = avg_log2FC_day50,
    y = log2FC_comparison_vs_reference,
    color = agreement,
    shape = RNA_clear_effect
  )
) +
  ggplot2::geom_hline(
    yintercept = 0,
    color = "grey70",
    linewidth = 0.4
  ) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey70",
    linewidth = 0.4
  ) +
  ggplot2::geom_point(alpha = 0.72, size = 1.8) +
  ggplot2::geom_text(
    data = source_wt_labels,
    ggplot2::aes(label = gene),
    size = 2.5,
    vjust = -0.6,
    check_overlap = TRUE,
    show.legend = FALSE
  ) +
  ggplot2::facet_wrap(
    ~ reference_program,
    scales = "free"
  ) +
  ggplot2::labs(
    title = "Persistent d50 source effects versus WT d21 RNA effects",
    subtitle = paste0(
      "WT unit ",
      WT_STEM_UNIT,
      ": ",
      WT_STEM_CELL_STATE,
      "; positive WT log2FC means HFD higher."
    ),
    x = "Source average log2FC at day 50",
    y = "WT pooled-library RNA log2FC (HFD versus CON)",
    color = "Direction",
    shape = "Clear WT effect"
  ) +
  theme_multiome(base_size = 10)

print(p_source_wt)

ggplot2::ggsave(
  file.path(
    STEP8_PLOT_DIR,
    "02_source_day50_vs_primary_WT_stem_RNA_effect.pdf"
  ),
  p_source_wt,
  width = 12,
  height = 8
)
```

## Exact promoter ATAC context

```{r plot-primary-stem-multimodal-support, fig.width=12, fig.height=8}
primary_multimodal_plot_data <- primary_stem_integrated |>
  dplyr::filter(
    RNA_effect_available,
    !is.na(representative_ATAC_log2FC)
  )

if (nrow(primary_multimodal_plot_data) > 0L) {
  p_primary_multimodal <- ggplot2::ggplot(
    primary_multimodal_plot_data,
    ggplot2::aes(
      x = representative_ATAC_log2FC,
      y = log2FC_comparison_vs_reference,
      color = reference_program,
      shape = RNA_clear_effect
    )
  ) +
    ggplot2::geom_hline(
      yintercept = 0,
      color = "grey70",
      linewidth = 0.4
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "grey70",
      linewidth = 0.4
    ) +
    ggplot2::geom_point(alpha = 0.75, size = 1.9) +
    ggplot2::labs(
      title = "WT RNA and exact promoter ATAC effects",
      subtitle = paste(
        "Each gene is represented by its largest absolute effect among",
        "peaks overlapping the strand-aware plus-or-minus 2-kb TSS window."
      ),
      x = "Representative promoter ATAC log2FC (HFD versus CON)",
      y = "RNA log2FC (HFD versus CON)",
      color = "Reference program",
      shape = "Clear RNA effect"
    ) +
    theme_multiome(base_size = 10)

  print(p_primary_multimodal)

  ggplot2::ggsave(
    file.path(
      STEP8_PLOT_DIR,
      "03_primary_WT_stem_RNA_exact_promoter_ATAC.pdf"
    ),
    p_primary_multimodal,
    width = 12,
    height = 8
  )
} else {
  message(
    "No persistent gene has an evaluable exact promoter-overlapping ATAC ",
    "peak in the primary WT stem unit; plot 03 was not created."
  )
}
```

## State breadth of the primary RNA candidates

```{r plot-rna-state-breadth, fig.width=14, fig.height=10}
shortlist_gene_order <- persistent_gene_review_shortlist |>
  dplyr::arrange(
    reference_direction,
    within_reference_direction_rank
  ) |>
  dplyr::pull(gene) |>
  unique()

state_breadth_plot_data <- rna_evidence |>
  dplyr::filter(gene %in% shortlist_gene_order) |>
  dplyr::mutate(
    gene = factor(gene, levels = rev(shortlist_gene_order)),
    analysis_unit_display = factor(
      analysis_unit_display,
      levels = unit_key$analysis_unit_display
    )
  )

p_state_breadth <- ggplot2::ggplot(
  state_breadth_plot_data,
  ggplot2::aes(
    x = analysis_unit_display,
    y = gene,
    fill = RNA_reference_aligned_log2FC
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.15) +
  ggplot2::scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    na.value = "grey88"
  ) +
  ggplot2::labs(
    title = "State breadth of persistent genes recovered in WT",
    subtitle = paste(
      "Positive values agree with the persistent reference direction;",
      "grey cells were absent from the filtered Step 7 RNA effect table."
    ),
    x = "WT analysis unit",
    y = NULL,
    fill = "Reference-aligned\nRNA log2FC"
  ) +
  theme_multiome(base_size = 9) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(
      angle = 45,
      hjust = 1
    )
  )

print(p_state_breadth)

ggplot2::ggsave(
  file.path(
    STEP8_PLOT_DIR,
    "04_persistent_gene_reference_aligned_RNA_across_WT_states.pdf"
  ),
  p_state_breadth,
  width = 14,
  height = 10
)
```

# 11. Write the completion summary

```{r write-completion-summary}
primary_summary_counts <- primary_stem_integrated |>
  dplyr::summarise(
    genes_in_reference = dplyr::n(),
    genes_in_WT_RNA_assay = sum(gene_in_RNA_assay),
    genes_with_WT_stem_RNA_effect = sum(RNA_effect_available),
    genes_with_clear_reference_concordant_RNA = sum(
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect,
      na.rm = TRUE
    ),
    genes_with_clear_reference_discordant_RNA = sum(
      RNA_reference_aligned_log2FC < 0 &
        RNA_clear_effect,
      na.rm = TRUE
    ),
    genes_with_stable_reference_concordant_RNA = sum(
      RNA_reference_aligned_log2FC > 0 &
        RNA_direction_stable,
      na.rm = TRUE
    ),
    genes_with_assayed_promoter_peak = sum(
      promoter_peaks_in_assay > 0L
    ),
    genes_with_evaluable_promoter_peak = sum(
      promoter_peaks_evaluable > 0L
    ),
    genes_with_clear_reference_concordant_promoter_ATAC = sum(
      promoter_peaks_clear_concordant > 0L
    ),
    genes_with_RNA_and_promoter_ATAC_support = sum(
      multimodal_validation_class ==
        "WT RNA and exact promoter ATAC support persistent reference"
    )
  )

step8_summary <- data.frame(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  cohort_group = COHORT_GROUP,
  primary_WT_stem_unit = WT_STEM_UNIT,
  primary_WT_stem_state = WT_STEM_CELL_STATE,
  persistent_genes = nrow(reference_genes),
  persistent_HFD_up_genes = sum(
    reference_genes$reference_direction == "HFD_higher"
  ),
  persistent_HFD_down_genes = sum(
    reference_genes$reference_direction == "CON_higher"
  ),
  genes_in_WT_RNA_assay =
    primary_summary_counts$genes_in_WT_RNA_assay,
  genes_with_WT_stem_RNA_effect =
    primary_summary_counts$genes_with_WT_stem_RNA_effect,
  clear_reference_concordant_RNA =
    primary_summary_counts$
      genes_with_clear_reference_concordant_RNA,
  clear_reference_discordant_RNA =
    primary_summary_counts$
      genes_with_clear_reference_discordant_RNA,
  stable_reference_concordant_RNA =
    primary_summary_counts$
      genes_with_stable_reference_concordant_RNA,
  genes_with_assayed_promoter_peak =
    primary_summary_counts$genes_with_assayed_promoter_peak,
  genes_with_evaluable_promoter_peak =
    primary_summary_counts$genes_with_evaluable_promoter_peak,
  clear_reference_concordant_promoter_ATAC =
    primary_summary_counts$
      genes_with_clear_reference_concordant_promoter_ATAC,
  RNA_and_exact_promoter_ATAC_support =
    primary_summary_counts$
      genes_with_RNA_and_promoter_ATAC_support,
  replicated_diet_design = FALSE,
  biological_inference_performed = FALSE,
  stringsAsFactors = FALSE
)

readr::write_csv(
  step8_summary,
  file.path(STEP8_DIR, "step8_summary.csv")
)

writeLines(
  c(
    paste("Step 8 completed:", Sys.time()),
    paste("Persistent reference genes:", nrow(reference_genes)),
    paste(
      "Persistent HFD-up genes:",
      sum(reference_genes$reference_direction == "HFD_higher")
    ),
    paste(
      "Persistent HFD-down genes:",
      sum(reference_genes$reference_direction == "CON_higher")
    ),
    paste("Primary WT stem unit:", WT_STEM_UNIT),
    paste("Primary WT stem annotation:", WT_STEM_CELL_STATE),
    paste(
      "Genes with clear reference-concordant WT stem RNA:",
      primary_summary_counts$
        genes_with_clear_reference_concordant_RNA
    ),
    paste(
      "Genes with clear reference-discordant WT stem RNA:",
      primary_summary_counts$
        genes_with_clear_reference_discordant_RNA
    ),
    paste(
      "Genes with RNA and exact promoter ATAC support:",
      primary_summary_counts$
        genes_with_RNA_and_promoter_ATAC_support
    ),
    "HyperMet overlay performed: FALSE",
    "HypoImmune overlay performed: FALSE",
    "Replicated diet inference performed: FALSE"
  ),
  STEP8_COMPLETE_MARKER
)

step8_summary
```

# Interpretation boundary

The primary output is
`persistent_gene_primary_WT_stem_integrated_evidence.csv`. It separates:

- persistent genes whose WT stem-state RNA direction agrees with the source;
- genes whose WT direction opposes the source;
- genes present in the assay but below Step 7 detection and count filters;
- genes absent from the WT RNA assay; and
- genes with concordant, discordant, mixed, or unresolved exact promoter ATAC
  context.

The full by-unit table should be used to decide whether a recovered program is
specific to the ISC-like state, shared across crypt states, associated with
the Reg4-positive niche, or broadly epithelial. These patterns remain
descriptive until independent biological replicates are available.
