---
title: "Step 8B - Refine persistent HFD programs in source-matched WT populations"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 8 is the immutable, strict audit of Gourab's persistent HFD-associated
gene lists in WT annotation unit 8 (`ISC-like crypt`). Step 8B asks the next,
different question: do the same source directions emerge when the WT cells
are regrouped into populations that better match the source annotations?

The source-aligned primary populations are:

1. `broad_stem`: WT units 8 and 6 (`ISC-like crypt` plus
   `Crypt progenitor`);
2. `DCS`: WT unit 9 (`Reg4+ deep crypt secretory`); and
3. `broad_goblet_all`: WT units 4, 5, 7, and 13 (all approved immature and
   mature goblet populations).

Strict and component populations are retained as sensitivity analyses. In
particular, unit 13 remains in the biologically complete goblet composite,
while a `core_goblet` analysis using units 4, 5, and 7 tests how strongly its
CON-dominant composition influences the pooled effect.

For each population, this notebook:

- recomputes descriptive HFD-versus-CON RNA and ATAC pooled-library effects;
- evaluates cell-sampling stability by balanced repeated downsampling;
- tests the persistent HFD-up and persistent HFD-down lists using ranked RNA
  enrichment;
- produces explicit concordant and discordant persistent-gene tables before
  performing targeted MSigDB over-representation analysis;
- characterizes ATAC changes independently; and
- integrates RNA with exact plus-or-minus 2-kb TSS accessibility evidence.

Distal peak-to-gene linkage, HyperMet/HypoImmune scoring, and cross-genotype
comparisons are deliberately deferred. They should use the benchmark and
candidate tables produced here rather than changing the definition of this
analysis after results are known.

## Statistical limitation

WT contains one pooled CON library and one pooled HFD library. Diet and
library are perfectly confounded, and individual mice cannot be recovered
after pooling. All effect sizes, ranked enrichment results, and enrichment
P values in this notebook are descriptive at the pooled-library level.
Balanced cell downsampling tests sensitivity to unequal cell numbers and a
small number of nuclei; it does not create biological replication.

Composite-population sensitivity analyses also are not independent
replication: the primary composite and its components contain overlapping
cells. They measure dependence on the chosen annotation boundary. A composite
pseudobulk effect can reflect both expression changes within a component and
diet-associated changes in the relative abundance of its component states;
the component tables are therefore essential to interpretation.

# 1. Setup

Run in a fresh R session configured for WT v2:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

This notebook additionally requires `fgsea` and `msigdbr`. Installation is
documented in `analysis/README_STEP08_PERSISTENT_HFD.md`.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight(
  packages = unique(c(required_packages, "fgsea", "msigdbr"))
)
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Signac)
  library(Matrix)
  library(GenomicRanges)
  library(IRanges)
  library(GenomeInfoDb)
  library(S4Vectors)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
  library(fgsea)
  library(msigdbr)
})

ggplot2::theme_set(theme_multiome())
future::plan(future::sequential)
set.seed(20260915)

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 8B is defined for the WT v2 benchmark. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}
```

# 2. Define inputs, outputs, and evidence rules

The pooled-effect thresholds mirror Step 7. `source_matched_primary`,
`sensitivity`, and `component` describe scientific scope. `Tier_A`, `Tier_B`,
and `Tier_C` describe only cell-number support. Neither is an annotation
confidence score.

```{r define-inputs-and-parameters}
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP7_DIR <- file.path(OUTPUT_DIR, "07_diet_response")
STEP8_DIR <- file.path(OUTPUT_DIR, "08_persistent_hfd_programs")
STEP8_TABLE_DIR <- file.path(STEP8_DIR, "tables")

STEP8B_DIR <- file.path(
  OUTPUT_DIR,
  "08b_source_matched_persistent_hfd"
)
STEP8B_TABLE_DIR <- file.path(STEP8B_DIR, "tables")
STEP8B_PLOT_DIR <- file.path(STEP8B_DIR, "plots")

dir.create(STEP8B_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8B_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8B_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

STEP6_OBJECT_PATH <- file.path(
  STEP6_DIR,
  "multiome_annotated.rds"
)
STEP6_COMPLETE_MARKER <- file.path(
  STEP6_DIR,
  "STEP_06_COMPLETE.txt"
)
STEP7_COMPLETE_MARKER <- file.path(
  STEP7_DIR,
  "STEP_07_COMPLETE.txt"
)
STEP8_COMPLETE_MARKER <- file.path(
  STEP8_DIR,
  "STEP_08_COMPLETE.txt"
)

REFERENCE_GENE_PATH <- file.path(
  DATASET_CONFIG_DIR,
  "reference_programs",
  "persistent_hfd_genes.csv"
)
STRICT_STEP8_RNA_PATH <- file.path(
  STEP8_TABLE_DIR,
  "persistent_gene_primary_WT_stem_RNA_summary.csv"
)
PROMOTER_OVERLAP_PATH <- file.path(
  STEP8_TABLE_DIR,
  "persistent_gene_exact_promoter_peak_overlaps.csv"
)

required_inputs <- c(
  Step6_object = STEP6_OBJECT_PATH,
  Step6_marker = STEP6_COMPLETE_MARKER,
  Step7_marker = STEP7_COMPLETE_MARKER,
  Step8_marker = STEP8_COMPLETE_MARKER,
  persistent_reference = REFERENCE_GENE_PATH,
  strict_Step8_RNA = STRICT_STEP8_RNA_PATH,
  persistent_promoter_overlaps = PROMOTER_OVERLAP_PATH,
  GRCm39_GTF = GRCM39_GTF
)

for (input_label in names(required_inputs)) {
  assert_file(required_inputs[[input_label]], input_label)
}

REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"

TIER_A_MIN_NUCLEI_PER_DIET <- 200L
TIER_B_MIN_NUCLEI_PER_DIET <- 100L

RNA_MIN_TOTAL_COUNTS <- 20L
RNA_MIN_DETECTION_PERCENT <- 5
RNA_CLEAR_LOG2FC <- 0.50
RNA_CLEAR_DETECTION_DELTA <- 5
RNA_TOP_FEATURES_PER_DIRECTION <- 250L

ATAC_MIN_TOTAL_COUNTS <- 30L
ATAC_MIN_DETECTION_PERCENT <- 1
ATAC_CLEAR_LOG2FC <- 0.50
ATAC_TOP_FEATURES_PER_DIRECTION <- 500L

CPM_PSEUDOCOUNT <- 0.10

DOWNSAMPLING_ITERATIONS <- 25L
DOWNSAMPLING_MAX_CELLS_PER_DIET <- 500L
DOWNSAMPLING_MAX_FEATURES_PER_POPULATION <- 1500L
MINIMUM_SIGN_CONCORDANCE_PERCENT <- 80

FGSEA_MIN_SIZE <- 10L
FGSEA_MAX_SIZE <- 500L

ORA_MIN_QUERY_GENES <- 5L
ORA_MIN_PATHWAY_GENES_IN_UNIVERSE <- 5L
ORA_MAX_PATHWAY_GENES_IN_UNIVERSE <- 500L

TSS_WINDOW_BP <- 2000L
PREVIEW_PLOTS_IN_RSTUDIO <- FALSE

analysis_parameters <- tibble::tibble(
  parameter = c(
    "reference diet",
    "comparison diet",
    "Tier A minimum nuclei per diet",
    "Tier B minimum nuclei per diet",
    "RNA minimum total counts",
    "RNA minimum detection percent",
    "RNA clear-effect absolute log2FC",
    "RNA clear-effect absolute detection delta",
    "ATAC minimum total counts",
    "ATAC minimum detection percent",
    "ATAC clear-effect absolute log2FC",
    "balanced downsampling iterations",
    "maximum sampled nuclei per diet",
    "minimum downsampling sign concordance percent",
    "exact promoter window bp each side of TSS",
    "replicated diet design",
    "biological inference performed"
  ),
  value = c(
    REFERENCE_DIET,
    COMPARISON_DIET,
    TIER_A_MIN_NUCLEI_PER_DIET,
    TIER_B_MIN_NUCLEI_PER_DIET,
    RNA_MIN_TOTAL_COUNTS,
    RNA_MIN_DETECTION_PERCENT,
    RNA_CLEAR_LOG2FC,
    RNA_CLEAR_DETECTION_DELTA,
    ATAC_MIN_TOTAL_COUNTS,
    ATAC_MIN_DETECTION_PERCENT,
    ATAC_CLEAR_LOG2FC,
    DOWNSAMPLING_ITERATIONS,
    DOWNSAMPLING_MAX_CELLS_PER_DIET,
    MINIMUM_SIGN_CONCORDANCE_PERCENT,
    TSS_WINDOW_BP,
    FALSE,
    FALSE
  )
)

readr::write_csv(
  analysis_parameters,
  file.path(STEP8B_DIR, "step8b_parameters.csv")
)
```

# 3. General helper functions

PDFs use the base PDF device because Cairo is not reliable in the SOL
interactive environment.

```{r helper-functions}
read_table_quietly <- function(path) {
  suppressMessages(
    readr::read_csv(
      path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
}

assert_columns <- function(data, required_columns, table_label) {
  missing_columns <- setdiff(required_columns, colnames(data))

  if (length(missing_columns) > 0L) {
    stop(
      table_label,
      " is missing required column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
}

align_matrix_to_metadata <- function(
  feature_matrix,
  metadata_cells,
  matrix_label
) {
  matrix_cells <- colnames(feature_matrix)

  if (!setequal(matrix_cells, metadata_cells)) {
    stop(
      matrix_label,
      " and the Step 6 metadata do not contain the same cells.",
      call. = FALSE
    )
  }

  if (!identical(matrix_cells, metadata_cells)) {
    message(
      matrix_label,
      " has the correct cells in a different order; reordering columns."
    )
    feature_matrix <- feature_matrix[
      ,
      metadata_cells,
      drop = FALSE
    ]
  }

  if (!identical(colnames(feature_matrix), metadata_cells)) {
    stop("Failed to align ", matrix_label, ".", call. = FALSE)
  }

  feature_matrix
}

preview_plot <- function(plot_object) {
  if (isTRUE(PREVIEW_PLOTS_IN_RSTUDIO)) {
    print(plot_object)
  }
  invisible(plot_object)
}

save_native_pdf <- function(
  filename,
  plot_object,
  width = 11,
  height = 7
) {
  output_path <- file.path(STEP8B_PLOT_DIR, filename)
  grDevices::pdf(
    file = output_path,
    width = width,
    height = height,
    family = "Helvetica",
    useDingbats = FALSE
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot_object)
  invisible(output_path)
}

observed_direction <- function(effect) {
  dplyr::case_when(
    is.na(effect) ~ NA_character_,
    effect > 0 ~ "HFD_higher",
    effect < 0 ~ "CON_higher",
    TRUE ~ "no_change"
  )
}

reference_aligned_effect <- function(effect, reference_direction) {
  dplyr::case_when(
    is.na(effect) ~ NA_real_,
    reference_direction == "HFD_higher" ~ effect,
    reference_direction == "CON_higher" ~ -effect,
    TRUE ~ NA_real_
  )
}

safe_spearman <- function(x, y) {
  complete <- is.finite(x) & is.finite(y)

  if (sum(complete) < 3L) {
    return(NA_real_)
  }

  suppressWarnings(
    stats::cor(
      x[complete],
      y[complete],
      method = "spearman"
    )
  )
}

prepare_stability <- function(stability_table, prefix) {
  stability_table |>
    dplyr::mutate(
      direction_stable =
        sign(full_log2FC) == sign(downsampled_log2FC_median) &
        sign_concordance_percent >=
          MINIMUM_SIGN_CONCORDANCE_PERCENT,
      downsampling_iqr_excludes_zero = dplyr::case_when(
        full_log2FC > 0 ~ downsampled_log2FC_q25 > 0,
        full_log2FC < 0 ~ downsampled_log2FC_q75 < 0,
        TRUE ~ FALSE
      )
    ) |>
    dplyr::rename_with(
      ~ paste0(prefix, .x),
      -dplyr::all_of(c("comparison_population", "feature"))
    )
}
```

# 4. Load and validate the WT object

The legacy WT Step 6 object does not contain `annotation_unit`. In that
expected case, `wnn_cluster` is used. This fallback is recorded, not inferred
silently.

```{r load-and-validate-wt-object}
multiome <- readRDS(STEP6_OBJECT_PATH)
metadata <- multiome[[]]

analysis_unit_column <- if (
  "annotation_unit" %in% colnames(metadata)
) {
  "annotation_unit"
} else if ("wnn_cluster" %in% colnames(metadata)) {
  "wnn_cluster"
} else {
  stop(
    "The Step 6 object contains neither annotation_unit nor wnn_cluster.",
    call. = FALSE
  )
}

required_metadata <- c(
  "sample_id",
  "diet",
  analysis_unit_column,
  "broad_compartment",
  "cell_state",
  "cell_state_full",
  "annotation_confidence"
)

missing_metadata <- setdiff(required_metadata, colnames(metadata))

if (length(missing_metadata) > 0L) {
  stop(
    "The Step 6 object is missing metadata column(s): ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

metadata$cell_barcode <- rownames(metadata)
metadata$sample_id <- as.character(metadata$sample_id)
metadata$diet <- as.character(metadata$diet)
metadata$analysis_unit <- as.character(
  metadata[[analysis_unit_column]]
)
metadata$broad_compartment <- as.character(
  metadata$broad_compartment
)
metadata$cell_state <- as.character(metadata$cell_state)
metadata$cell_state_full <- as.character(metadata$cell_state_full)
metadata$annotation_confidence <- tolower(
  as.character(metadata$annotation_confidence)
)

expected_diets <- c(REFERENCE_DIET, COMPARISON_DIET)

if (!setequal(unique(metadata$diet), expected_diets)) {
  stop("Expected exactly CON and HFD in the WT object.", call. = FALSE)
}

if (!all(c("RNA", "ATAC") %in% names(multiome))) {
  stop("The WT object must contain RNA and ATAC assays.", call. = FALSE)
}

rna_layers <- SeuratObject::Layers(multiome[["RNA"]])
atac_layers <- SeuratObject::Layers(multiome[["ATAC"]])

if (!"counts" %in% rna_layers || !"counts" %in% atac_layers) {
  stop("RNA and ATAC must each contain a joined counts layer.", call. = FALSE)
}

rna_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)
atac_counts <- SeuratObject::LayerData(
  multiome[["ATAC"]],
  layer = "counts"
)

metadata_cells <- rownames(metadata)

matrix_alignment_audit <- tibble::tibble(
  matrix = c("RNA counts", "ATAC counts"),
  matrix_cells = c(ncol(rna_counts), ncol(atac_counts)),
  metadata_cells = length(metadata_cells),
  same_cell_set = c(
    setequal(colnames(rna_counts), metadata_cells),
    setequal(colnames(atac_counts), metadata_cells)
  ),
  identical_order_before_alignment = c(
    identical(colnames(rna_counts), metadata_cells),
    identical(colnames(atac_counts), metadata_cells)
  )
)

rna_counts <- align_matrix_to_metadata(
  rna_counts,
  metadata_cells,
  "RNA counts"
)
atac_counts <- align_matrix_to_metadata(
  atac_counts,
  metadata_cells,
  "ATAC counts"
)

atac_ranges <- GenomicRanges::granges(multiome[["ATAC"]])

if (length(atac_ranges) != nrow(atac_counts)) {
  stop(
    "ATAC genomic ranges and count features are not one-to-one.",
    call. = FALSE
  )
}

names(atac_ranges) <- rownames(atac_counts)

rm(multiome)
invisible(gc())

readr::write_csv(
  matrix_alignment_audit,
  file.path(STEP8B_DIR, "matrix_cell_alignment_audit.csv")
)

run_inventory <- tibble::tibble(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  cohort_group = COHORT_GROUP,
  nuclei = nrow(metadata),
  RNA_features = nrow(rna_counts),
  ATAC_features = nrow(atac_counts),
  analysis_unit_column = analysis_unit_column,
  independent_libraries = dplyr::n_distinct(metadata$sample_id),
  replicated_diet_design = FALSE,
  biological_inference_performed = FALSE
)

readr::write_csv(
  run_inventory,
  file.path(STEP8B_DIR, "step8b_run_inventory.csv")
)

run_inventory
```

# 5. Define source-matched and sensitivity populations

The mapping is explicit and is written to disk before any diet effect is
calculated. Units 0 and 18 are biologically informative cycling populations,
but are excluded from these source-matched composites to avoid making cell
cycle the definition of the source population.

```{r define-source-matched-populations}
population_definitions <- tibble::tibble(
  comparison_population = c(
    "broad_stem",
    "strict_ISC",
    "crypt_progenitor",
    "DCS",
    "broad_goblet_all",
    "core_goblet",
    "immature_goblet_4",
    "mature_goblet_5",
    "immature_goblet_7",
    "immature_goblet_13"
  ),
  display_label = c(
    "Broad stem (8 + 6)",
    "Strict ISC-like (8)",
    "Crypt progenitor (6)",
    "Reg4+ DCS (9)",
    "All goblet (4 + 5 + 7 + 13)",
    "Core goblet (4 + 5 + 7)",
    "Immature goblet (4)",
    "Mature goblet (5)",
    "Immature goblet (7)",
    "Immature goblet, CON-dominant (13)"
  ),
  source_population = c(
    "Gourab stem",
    "Gourab stem",
    "Gourab stem",
    "Gourab DCS",
    "Gourab goblet",
    "Gourab goblet",
    "Gourab goblet",
    "Gourab goblet",
    "Gourab goblet",
    "Gourab goblet"
  ),
  analysis_role = c(
    "source_matched_primary",
    "sensitivity",
    "component",
    "source_matched_primary",
    "source_matched_primary",
    "sensitivity",
    "component",
    "component",
    "component",
    "component"
  ),
  member_units = list(
    c("8", "6"),
    "8",
    "6",
    "9",
    c("4", "5", "7", "13"),
    c("4", "5", "7"),
    "4",
    "5",
    "7",
    "13"
  ),
  rationale = c(
    "Source-aligned broad stem population spanning ISC-like and noncycling crypt progenitor states.",
    "Immutable Step 8 anchor retained as a strict sensitivity analysis.",
    "Tests the noncycling crypt-progenitor contribution to the broad-stem result.",
    "Approved Reg4-positive deep crypt secretory population.",
    "Biologically complete approved immature-plus-mature goblet composite, including unit 13.",
    "Tests sensitivity to removal of the strongly CON-dominant unit 13.",
    "Component audit for immature goblet unit 4.",
    "Component audit for mature goblet unit 5.",
    "Component audit for immature goblet unit 7.",
    "Component audit for approved but strongly CON-dominant immature goblet unit 13."
  )
)

expected_unit_annotations <- tibble::tribble(
  ~analysis_unit, ~expected_cell_state, ~expected_confidence,
  "4", "Immature goblet", "moderate",
  "5", "Mature goblet", "high",
  "6", "Crypt progenitor", "moderate",
  "7", "Immature goblet", "moderate",
  "8", "ISC-like crypt", "moderate",
  "9", "Reg4+ DCS", "high",
  "13", "Immature goblet", "moderate"
)

unit_annotation <- metadata |>
  dplyr::distinct(
    analysis_unit,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence
  )

duplicate_unit_annotation <- unit_annotation |>
  dplyr::count(analysis_unit) |>
  dplyr::filter(n != 1L)

if (nrow(duplicate_unit_annotation) > 0L) {
  stop(
    "At least one analysis unit has inconsistent annotation metadata.",
    call. = FALSE
  )
}

unit_validation <- expected_unit_annotations |>
  dplyr::left_join(unit_annotation, by = "analysis_unit") |>
  dplyr::mutate(
    unit_present = !is.na(cell_state),
    epithelial = broad_compartment == "epithelial",
    cell_state_matches = cell_state == expected_cell_state,
    confidence_matches = annotation_confidence == expected_confidence,
    validation_pass = unit_present & epithelial &
      cell_state_matches & confidence_matches
  )

if (any(!unit_validation$validation_pass)) {
  stop(
    "The WT annotations no longer match the approved Step 8B population ",
    "mapping. Review step8b_unit_annotation_validation.csv.",
    call. = FALSE
  )
}

readr::write_csv(
  unit_validation,
  file.path(STEP8B_DIR, "step8b_unit_annotation_validation.csv")
)

population_definition_export <- population_definitions |>
  dplyr::mutate(
    member_units = vapply(
      member_units,
      paste,
      collapse = ";",
      FUN.VALUE = character(1)
    )
  )

readr::write_csv(
  population_definition_export,
  file.path(STEP8B_DIR, "matched_population_definitions.csv")
)
```

```{r quantify-population-support}
population_cell_counts <- dplyr::bind_rows(
  lapply(
    seq_len(nrow(population_definitions)),
    function(population_index) {
      population_id <- population_definitions$comparison_population[[
        population_index
      ]]
      member_units <- population_definitions$member_units[[
        population_index
      ]]

      metadata |>
        dplyr::filter(analysis_unit %in% member_units) |>
        dplyr::count(diet, name = "nuclei") |>
        tidyr::complete(
          diet = expected_diets,
          fill = list(nuclei = 0L)
        ) |>
        dplyr::mutate(comparison_population = population_id)
    }
  )
)

population_support <- population_cell_counts |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = nuclei,
    names_prefix = "nuclei_"
  ) |>
  dplyr::left_join(
    population_definition_export,
    by = "comparison_population"
  ) |>
  dplyr::mutate(
    minimum_nuclei_per_diet = pmin(nuclei_CON, nuclei_HFD),
    cell_support_tier = dplyr::case_when(
      minimum_nuclei_per_diet >= TIER_A_MIN_NUCLEI_PER_DIET ~
        "Tier_A",
      minimum_nuclei_per_diet >= TIER_B_MIN_NUCLEI_PER_DIET ~
        "Tier_B",
      TRUE ~ "Tier_C"
    ),
    primary_RNA_ATAC =
      analysis_role == "source_matched_primary" &
      cell_support_tier %in% c("Tier_A", "Tier_B"),
    interpretation_scope = dplyr::case_when(
      analysis_role == "source_matched_primary" &
        cell_support_tier == "Tier_A" ~
        "Primary source-matched descriptive RNA and ATAC benchmark",
      analysis_role == "source_matched_primary" &
        cell_support_tier == "Tier_B" ~
        "Primary source-matched benchmark with reduced cell support",
      analysis_role == "sensitivity" ~
        "Annotation-boundary sensitivity analysis",
      TRUE ~ "Internal component consistency analysis"
    )
  ) |>
  dplyr::arrange(
    factor(
      analysis_role,
      levels = c(
        "source_matched_primary",
        "sensitivity",
        "component"
      )
    ),
    comparison_population
  )

primary_populations <- population_support |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::pull(comparison_population)

expected_primary_populations <- c(
  "broad_stem",
  "DCS",
  "broad_goblet_all"
)

if (!setequal(primary_populations, expected_primary_populations)) {
  stop(
    "The source-matched primary population set is not the expected ",
    "broad_stem, DCS, and broad_goblet_all set.",
    call. = FALSE
  )
}

readr::write_csv(
  population_cell_counts,
  file.path(STEP8B_TABLE_DIR, "matched_population_counts_by_diet.csv")
)
readr::write_csv(
  population_support,
  file.path(STEP8B_TABLE_DIR, "matched_population_support.csv")
)

population_support
```

# 6. Calculate descriptive RNA effects

Each population and diet is collapsed into one count vector. These are pooled
library contrasts, not replicate-aware pseudobulk models.

```{r pooled-effect-functions}
summarize_pooled_effect <- function(
  count_matrix,
  feature_label,
  metadata,
  comparison_population,
  member_units,
  minimum_total_counts,
  minimum_detection_percent,
  cpm_pseudocount
) {
  reference_cells <- metadata$cell_barcode[
    metadata$analysis_unit %in% member_units &
      metadata$diet == REFERENCE_DIET
  ]
  comparison_cells <- metadata$cell_barcode[
    metadata$analysis_unit %in% member_units &
      metadata$diet == COMPARISON_DIET
  ]

  if (length(reference_cells) == 0L || length(comparison_cells) == 0L) {
    return(NULL)
  }

  reference_counts <- Matrix::rowSums(
    count_matrix[, reference_cells, drop = FALSE]
  )
  comparison_counts <- Matrix::rowSums(
    count_matrix[, comparison_cells, drop = FALSE]
  )
  reference_detected <- Matrix::rowSums(
    count_matrix[, reference_cells, drop = FALSE] > 0
  )
  comparison_detected <- Matrix::rowSums(
    count_matrix[, comparison_cells, drop = FALSE] > 0
  )

  reference_library_size <- sum(reference_counts)
  comparison_library_size <- sum(comparison_counts)

  if (reference_library_size <= 0 || comparison_library_size <= 0) {
    stop(
      "A pooled library has zero counts for ",
      comparison_population,
      ".",
      call. = FALSE
    )
  }

  reference_cpm <- 1e6 * reference_counts / reference_library_size
  comparison_cpm <- 1e6 * comparison_counts / comparison_library_size
  reference_percent <-
    100 * reference_detected / length(reference_cells)
  comparison_percent <-
    100 * comparison_detected / length(comparison_cells)

  tibble::tibble(
    comparison_population = comparison_population,
    feature = rownames(count_matrix),
    feature_type = feature_label,
    reference_diet = REFERENCE_DIET,
    comparison_diet = COMPARISON_DIET,
    reference_nuclei = length(reference_cells),
    comparison_nuclei = length(comparison_cells),
    reference_counts = reference_counts,
    comparison_counts = comparison_counts,
    total_counts = reference_counts + comparison_counts,
    reference_CPM = reference_cpm,
    comparison_CPM = comparison_cpm,
    log2FC_comparison_vs_reference = log2(
      (comparison_cpm + cpm_pseudocount) /
        (reference_cpm + cpm_pseudocount)
    ),
    reference_percent_detected = reference_percent,
    comparison_percent_detected = comparison_percent,
    detection_difference_percentage_points =
      comparison_percent - reference_percent
  ) |>
    dplyr::filter(
      total_counts >= minimum_total_counts,
      pmax(
        reference_percent_detected,
        comparison_percent_detected
      ) >= minimum_detection_percent
    )
}

calculate_population_effects <- function(
  count_matrix,
  feature_label,
  minimum_total_counts,
  minimum_detection_percent
) {
  dplyr::bind_rows(
    lapply(
      seq_len(nrow(population_definitions)),
      function(population_index) {
        summarize_pooled_effect(
          count_matrix = count_matrix,
          feature_label = feature_label,
          metadata = metadata,
          comparison_population =
            population_definitions$comparison_population[[
              population_index
            ]],
          member_units = population_definitions$member_units[[
            population_index
          ]],
          minimum_total_counts = minimum_total_counts,
          minimum_detection_percent = minimum_detection_percent,
          cpm_pseudocount = CPM_PSEUDOCOUNT
        )
      }
    )
  ) |>
    dplyr::left_join(
      population_support |>
        dplyr::select(
          comparison_population,
          display_label,
          source_population,
          analysis_role,
          cell_support_tier
        ),
      by = "comparison_population"
    )
}

select_top_effects <- function(
  effects,
  absolute_log2fc_threshold,
  top_features_per_direction,
  detection_delta_threshold = NULL
) {
  selected <- effects |>
    dplyr::filter(
      abs(log2FC_comparison_vs_reference) >=
        absolute_log2fc_threshold
    )

  if (!is.null(detection_delta_threshold)) {
    selected <- selected |>
      dplyr::filter(
        abs(detection_difference_percentage_points) >=
          detection_delta_threshold
      )
  }

  selected |>
    dplyr::mutate(
      direction = observed_direction(
        log2FC_comparison_vs_reference
      )
    ) |>
    dplyr::group_by(comparison_population, direction) |>
    dplyr::slice_max(
      order_by = abs(log2FC_comparison_vs_reference),
      n = top_features_per_direction,
      with_ties = FALSE
    ) |>
    dplyr::ungroup()
}
```

```{r calculate-rna-effects}
rna_effects <- calculate_population_effects(
  count_matrix = rna_counts,
  feature_label = "gene",
  minimum_total_counts = RNA_MIN_TOTAL_COUNTS,
  minimum_detection_percent = RNA_MIN_DETECTION_PERCENT
)

rna_top_candidates <- select_top_effects(
  effects = rna_effects,
  absolute_log2fc_threshold = RNA_CLEAR_LOG2FC,
  top_features_per_direction = RNA_TOP_FEATURES_PER_DIRECTION,
  detection_delta_threshold = RNA_CLEAR_DETECTION_DELTA
)

readr::write_csv(
  rna_effects,
  file.path(
    STEP8B_TABLE_DIR,
    "RNA_pooled_library_effects_by_matched_population.csv.gz"
  )
)
readr::write_csv(
  rna_top_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "RNA_top_diet_candidates_by_matched_population.csv"
  )
)

rna_top_candidates |>
  dplyr::count(comparison_population, direction)
```

# 7. Evaluate balanced cell-sampling stability

All evaluable persistent genes are included in RNA downsampling. ATAC
downsampling includes the top clear-effect peaks plus every evaluable exact
promoter peak for a persistent gene.

```{r balanced-downsampling-functions}
downsample_effect_stability <- function(
  count_matrix,
  full_effects,
  required_features,
  maximum_features_per_population,
  random_seed
) {
  set.seed(random_seed)
  cell_library_sizes <- Matrix::colSums(count_matrix)
  iteration_results <- list()
  result_index <- 1L

  for (population_index in seq_len(nrow(population_definitions))) {
    population_id <- population_definitions$comparison_population[[
      population_index
    ]]
    member_units <- population_definitions$member_units[[
      population_index
    ]]

    population_effects <- full_effects |>
      dplyr::filter(comparison_population == population_id)

    required_rows <- population_effects |>
      dplyr::filter(feature %in% required_features)

    top_rows <- population_effects |>
      dplyr::slice_max(
        order_by = abs(log2FC_comparison_vs_reference),
        n = maximum_features_per_population,
        with_ties = FALSE
      )

    selected_effects <- dplyr::bind_rows(
      required_rows,
      top_rows
    ) |>
      dplyr::distinct(feature, .keep_all = TRUE)

    if (nrow(selected_effects) == 0L) {
      next
    }

    feature_index <- match(
      selected_effects$feature,
      rownames(count_matrix)
    )
    valid_features <- !is.na(feature_index)
    selected_effects <- selected_effects[valid_features, , drop = FALSE]
    feature_index <- feature_index[valid_features]

    reference_cells <- metadata$cell_barcode[
      metadata$analysis_unit %in% member_units &
        metadata$diet == REFERENCE_DIET
    ]
    comparison_cells <- metadata$cell_barcode[
      metadata$analysis_unit %in% member_units &
        metadata$diet == COMPARISON_DIET
    ]

    sampled_nuclei <- min(
      length(reference_cells),
      length(comparison_cells),
      DOWNSAMPLING_MAX_CELLS_PER_DIET
    )

    if (sampled_nuclei < 1L) {
      next
    }

    for (iteration in seq_len(DOWNSAMPLING_ITERATIONS)) {
      sampled_reference <- sample(
        reference_cells,
        sampled_nuclei,
        replace = FALSE
      )
      sampled_comparison <- sample(
        comparison_cells,
        sampled_nuclei,
        replace = FALSE
      )

      sampled_reference_counts <- Matrix::rowSums(
        count_matrix[
          feature_index,
          sampled_reference,
          drop = FALSE
        ]
      )
      sampled_comparison_counts <- Matrix::rowSums(
        count_matrix[
          feature_index,
          sampled_comparison,
          drop = FALSE
        ]
      )

      sampled_reference_cpm <-
        1e6 * sampled_reference_counts /
        sum(cell_library_sizes[sampled_reference])
      sampled_comparison_cpm <-
        1e6 * sampled_comparison_counts /
        sum(cell_library_sizes[sampled_comparison])

      iteration_results[[result_index]] <- tibble::tibble(
        comparison_population = population_id,
        feature = selected_effects$feature,
        iteration = iteration,
        sampled_nuclei_per_diet = sampled_nuclei,
        downsampled_log2FC = log2(
          (sampled_comparison_cpm + CPM_PSEUDOCOUNT) /
            (sampled_reference_cpm + CPM_PSEUDOCOUNT)
        ),
        full_log2FC =
          selected_effects$log2FC_comparison_vs_reference
      )

      result_index <- result_index + 1L
    }
  }

  stability_iterations <- dplyr::bind_rows(iteration_results)

  if (nrow(stability_iterations) == 0L) {
    stop("Balanced downsampling produced no results.", call. = FALSE)
  }

  stability_iterations |>
    dplyr::group_by(comparison_population, feature) |>
    dplyr::summarise(
      sampled_nuclei_per_diet = dplyr::first(
        sampled_nuclei_per_diet
      ),
      full_log2FC = dplyr::first(full_log2FC),
      downsampled_log2FC_median = stats::median(
        downsampled_log2FC
      ),
      downsampled_log2FC_q25 = stats::quantile(
        downsampled_log2FC,
        0.25
      ),
      downsampled_log2FC_q75 = stats::quantile(
        downsampled_log2FC,
        0.75
      ),
      sign_concordance_percent = 100 * mean(
        sign(downsampled_log2FC) == sign(full_log2FC)
      ),
      .groups = "drop"
    )
}
```

```{r run-rna-downsampling}
reference_genes <- read_table_quietly(REFERENCE_GENE_PATH)

assert_columns(
  reference_genes,
  c(
    "gene",
    "reference_direction",
    "reference_program",
    "source_rank",
    "avg_log2FC_day21",
    "avg_log2FC_day50",
    "source_file"
  ),
  "persistent HFD reference table"
)

reference_genes <- reference_genes |>
  dplyr::mutate(
    gene = dplyr::recode(gene, Atpif1 = "Atp5if1")
  ) |>
  dplyr::distinct(gene, .keep_all = TRUE)

if (
  any(!reference_genes$reference_direction %in%
    c("HFD_higher", "CON_higher")) ||
    any(!reference_genes$reference_program %in%
      c("persistent_HFD_up", "persistent_HFD_down"))
) {
  stop("The persistent reference contains invalid directions.", call. = FALSE)
}

source_direction_mismatch <- reference_genes |>
  dplyr::filter(
    (reference_direction == "HFD_higher" &
      avg_log2FC_day21 <= 0) |
      (reference_direction == "CON_higher" &
        avg_log2FC_day21 >= 0)
  )

if (nrow(source_direction_mismatch) > 0L) {
  stop(
    "At least one day-21 source effect disagrees with its stored direction.",
    call. = FALSE
  )
}

rna_stability <- downsample_effect_stability(
  count_matrix = rna_counts,
  full_effects = rna_effects,
  required_features = reference_genes$gene,
  maximum_features_per_population =
    DOWNSAMPLING_MAX_FEATURES_PER_POPULATION,
  random_seed = 20260916
)

readr::write_csv(
  rna_stability,
  file.path(
    STEP8B_TABLE_DIR,
    "RNA_balanced_downsampling_stability.csv"
  )
)
```

# 8. Build persistent-gene RNA evidence

Positive `RNA_reference_aligned_log2FC` means that WT agrees with the source
direction, regardless of whether the source gene was HFD-up or HFD-down.

```{r build-persistent-rna-evidence}
rna_stability_key <- prepare_stability(
  rna_stability,
  "RNA_"
) |>
  dplyr::rename(gene = feature)

population_key <- population_support |>
  dplyr::select(
    comparison_population,
    display_label,
    source_population,
    analysis_role,
    member_units,
    cell_support_tier,
    primary_RNA_ATAC
  )

rna_effect_key <- rna_effects |>
  dplyr::select(
    comparison_population,
    feature,
    reference_nuclei,
    comparison_nuclei,
    reference_counts,
    comparison_counts,
    total_counts,
    reference_CPM,
    comparison_CPM,
    log2FC_comparison_vs_reference,
    reference_percent_detected,
    comparison_percent_detected,
    detection_difference_percentage_points
  ) |>
  dplyr::rename(gene = feature) |>
  dplyr::mutate(RNA_effect_available = TRUE)

rna_persistent_evidence <- tidyr::crossing(
  reference_genes,
  population_key
) |>
  dplyr::left_join(
    rna_effect_key,
    by = c("comparison_population", "gene")
  ) |>
  dplyr::left_join(
    rna_stability_key,
    by = c("comparison_population", "gene")
  ) |>
  dplyr::mutate(
    gene_in_RNA_assay = gene %in% rownames(rna_counts),
    RNA_effect_available = dplyr::coalesce(
      RNA_effect_available,
      FALSE
    ),
    RNA_observed_direction = observed_direction(
      log2FC_comparison_vs_reference
    ),
    RNA_reference_aligned_log2FC = reference_aligned_effect(
      log2FC_comparison_vs_reference,
      reference_direction
    ),
    RNA_reference_direction_concordant = dplyr::case_when(
      !RNA_effect_available ~ NA,
      RNA_reference_aligned_log2FC > 0 ~ TRUE,
      RNA_reference_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    RNA_clear_effect =
      RNA_effect_available &
      abs(log2FC_comparison_vs_reference) >= RNA_CLEAR_LOG2FC &
      abs(detection_difference_percentage_points) >=
        RNA_CLEAR_DETECTION_DELTA,
    RNA_stability_available = !is.na(RNA_sign_concordance_percent),
    RNA_direction_stable = dplyr::coalesce(
      RNA_direction_stable,
      FALSE
    ),
    RNA_downsampling_iqr_excludes_zero = dplyr::coalesce(
      RNA_downsampling_iqr_excludes_zero,
      FALSE
    ),
    RNA_evidence_class = dplyr::case_when(
      !gene_in_RNA_assay ~ "absent_from_RNA_assay",
      !RNA_effect_available ~ "below_RNA_effect_filters",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect & RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero ~
        "concordant_clear_stable_IQR",
      RNA_reference_aligned_log2FC > 0 &
        RNA_clear_effect & RNA_direction_stable ~
        "concordant_clear_stable",
      RNA_reference_aligned_log2FC > 0 & RNA_clear_effect ~
        "concordant_clear_unstable",
      RNA_reference_aligned_log2FC > 0 ~
        "concordant_subthreshold",
      RNA_reference_aligned_log2FC < 0 &
        RNA_clear_effect & RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero ~
        "discordant_clear_stable_IQR",
      RNA_reference_aligned_log2FC < 0 &
        RNA_clear_effect & RNA_direction_stable ~
        "discordant_clear_stable",
      RNA_reference_aligned_log2FC < 0 & RNA_clear_effect ~
        "discordant_clear_unstable",
      RNA_reference_aligned_log2FC < 0 ~
        "discordant_subthreshold",
      TRUE ~ "no_directional_change"
    )
  ) |>
  dplyr::arrange(
    factor(
      analysis_role,
      levels = c(
        "source_matched_primary",
        "sensitivity",
        "component"
      )
    ),
    comparison_population,
    reference_program,
    source_rank
  )

readr::write_csv(
  rna_persistent_evidence,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_RNA_evidence_by_matched_population.csv.gz"
  )
)

rna_persistent_evidence |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::count(
    comparison_population,
    reference_program,
    RNA_evidence_class
  )
```

# 9. Confirm strict unit-8 provenance

The recomputed `strict_ISC` pooled effect must reproduce Step 8. Step 8 is not
rewritten or reinterpreted by this notebook.

```{r audit-strict-step8-reproduction}
strict_step8 <- read_table_quietly(STRICT_STEP8_RNA_PATH)

assert_columns(
  strict_step8,
  c(
    "gene",
    "RNA_effect_available",
    "log2FC_comparison_vs_reference",
    "detection_difference_percentage_points"
  ),
  "strict Step 8 RNA summary"
)

strict_recomputed <- rna_persistent_evidence |>
  dplyr::filter(comparison_population == "strict_ISC") |>
  dplyr::select(
    gene,
    recomputed_effect_available = RNA_effect_available,
    recomputed_log2FC = log2FC_comparison_vs_reference,
    recomputed_detection_delta =
      detection_difference_percentage_points
  )

strict_reproduction_audit <- strict_step8 |>
  dplyr::select(
    gene,
    upstream_effect_available = RNA_effect_available,
    upstream_log2FC = log2FC_comparison_vs_reference,
    upstream_detection_delta =
      detection_difference_percentage_points
  ) |>
  dplyr::full_join(strict_recomputed, by = "gene") |>
  dplyr::mutate(
    gene_present_in_both =
      !is.na(upstream_effect_available) &
      !is.na(recomputed_effect_available),
    availability_matches =
      upstream_effect_available == recomputed_effect_available,
    log2FC_absolute_difference = abs(
      upstream_log2FC - recomputed_log2FC
    ),
    detection_delta_absolute_difference = abs(
      upstream_detection_delta - recomputed_detection_delta
    ),
    reproduction_pass =
      gene_present_in_both &
      availability_matches &
      (
        !upstream_effect_available |
          (
            log2FC_absolute_difference <= 1e-8 &
              detection_delta_absolute_difference <= 1e-8
          )
      )
  )

readr::write_csv(
  strict_reproduction_audit,
  file.path(
    STEP8B_TABLE_DIR,
    "strict_Step8_unit8_reproduction_audit.csv"
  )
)

if (any(!strict_reproduction_audit$reproduction_pass)) {
  stop(
    "The Step 8B strict_ISC recomputation does not reproduce Step 8. ",
    "Do not interpret regrouped results until the audit is resolved.",
    call. = FALSE
  )
}
```

# 10. Test persistent lists by ranked enrichment

Ranked enrichment uses every gene that passes the population-specific RNA
count and detection filters. A positive NES for `persistent_HFD_up` and a
negative NES for `persistent_HFD_down` agree with the source. Because the two
diets are single pooled libraries, enrichment P values do not quantify
mouse-to-mouse reproducibility.

```{r run-ranked-persistent-enrichment}
persistent_pathways <- split(
  reference_genes$gene,
  reference_genes$reference_program
)

run_one_ranked_enrichment <- function(population_id) {
  population_effects <- rna_effects |>
    dplyr::filter(comparison_population == population_id) |>
    dplyr::filter(is.finite(log2FC_comparison_vs_reference)) |>
    dplyr::arrange(dplyr::desc(log2FC_comparison_vs_reference))

  ranked_statistics <- population_effects$log2FC_comparison_vs_reference
  names(ranked_statistics) <- population_effects$feature
  ranked_statistics <- sort(ranked_statistics, decreasing = TRUE)

  enrichment <- fgsea::fgseaMultilevel(
    pathways = persistent_pathways,
    stats = ranked_statistics,
    minSize = FGSEA_MIN_SIZE,
    maxSize = FGSEA_MAX_SIZE,
    eps = 0,
    scoreType = "std"
  ) |>
    as.data.frame()

  if (nrow(enrichment) == 0L) {
    return(tibble::tibble())
  }

  enrichment |>
    tibble::as_tibble() |>
    dplyr::mutate(
      comparison_population = population_id,
      leadingEdge = vapply(
        leadingEdge,
        paste,
        collapse = ";",
        FUN.VALUE = character(1)
      ),
      expected_NES_direction = dplyr::if_else(
        pathway == "persistent_HFD_up",
        "positive",
        "negative"
      ),
      expected_direction_recovered = dplyr::if_else(
        pathway == "persistent_HFD_up",
        NES > 0,
        NES < 0
      )
    )
}

ranked_enrichment <- dplyr::bind_rows(
  lapply(
    population_definitions$comparison_population,
    run_one_ranked_enrichment
  )
) |>
  dplyr::left_join(
    population_support |>
      dplyr::select(
        comparison_population,
        display_label,
        analysis_role,
        cell_support_tier
      ),
    by = "comparison_population"
  ) |>
  dplyr::arrange(analysis_role, comparison_population, pathway)

readr::write_csv(
  ranked_enrichment,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_program_ranked_RNA_enrichment.csv"
  )
)

ranked_enrichment |>
  dplyr::filter(comparison_population %in% primary_populations) |>
  dplyr::select(
    comparison_population,
    pathway,
    size,
    NES,
    pval,
    padj,
    expected_direction_recovered
  )
```

# 11. Quantify annotation-boundary sensitivity

These comparisons test whether a composite result is carried by one of its
components. They do not represent independent samples.

```{r calculate-population-sensitivity}
sensitivity_pairs <- tibble::tribble(
  ~primary_population, ~comparator_population, ~comparison_type,
  "broad_stem", "strict_ISC", "strict_anchor",
  "broad_stem", "crypt_progenitor", "component",
  "broad_goblet_all", "core_goblet", "unit13_exclusion",
  "broad_goblet_all", "immature_goblet_4", "component",
  "broad_goblet_all", "mature_goblet_5", "component",
  "broad_goblet_all", "immature_goblet_7", "component",
  "broad_goblet_all", "immature_goblet_13", "component"
)

primary_sensitivity_values <- rna_persistent_evidence |>
  dplyr::select(
    gene,
    comparison_population,
    primary_effect_available = RNA_effect_available,
    primary_log2FC = log2FC_comparison_vs_reference,
    primary_clear_effect = RNA_clear_effect,
    primary_direction_stable = RNA_direction_stable
  )

comparator_sensitivity_values <- rna_persistent_evidence |>
  dplyr::select(
    gene,
    comparison_population,
    comparator_effect_available = RNA_effect_available,
    comparator_log2FC = log2FC_comparison_vs_reference,
    comparator_clear_effect = RNA_clear_effect,
    comparator_direction_stable = RNA_direction_stable
  )

sensitivity_gene_evidence <- tidyr::crossing(
  sensitivity_pairs,
  gene = reference_genes$gene
) |>
  dplyr::left_join(
    primary_sensitivity_values,
    by = c(
      "primary_population" = "comparison_population",
      "gene" = "gene"
    )
  ) |>
  dplyr::left_join(
    comparator_sensitivity_values,
    by = c(
      "comparator_population" = "comparison_population",
      "gene" = "gene"
    )
  ) |>
  dplyr::mutate(
    both_evaluable =
      primary_effect_available & comparator_effect_available,
    direction_agrees = dplyr::case_when(
      !both_evaluable ~ NA,
      sign(primary_log2FC) == sign(comparator_log2FC) ~ TRUE,
      TRUE ~ FALSE
    ),
    both_clear = primary_clear_effect & comparator_clear_effect,
    both_stable =
      primary_direction_stable & comparator_direction_stable
  )

sensitivity_summary <- sensitivity_gene_evidence |>
  dplyr::group_by(
    primary_population,
    comparator_population,
    comparison_type
  ) |>
  dplyr::summarise(
    persistent_genes = dplyr::n(),
    genes_evaluable_in_both = sum(both_evaluable),
    genes_same_direction = sum(direction_agrees %in% TRUE),
    genes_opposite_direction = sum(direction_agrees %in% FALSE),
    same_direction_percent = 100 * genes_same_direction /
      pmax(1L, genes_evaluable_in_both),
    clear_in_both = sum(both_clear, na.rm = TRUE),
    stable_in_both = sum(both_stable, na.rm = TRUE),
    spearman_effect_correlation = safe_spearman(
      primary_log2FC[both_evaluable],
      comparator_log2FC[both_evaluable]
    ),
    .groups = "drop"
  )

candidate_sensitivity_summary <- sensitivity_gene_evidence |>
  dplyr::filter(primary_effect_available) |>
  dplyr::group_by(primary_population, gene) |>
  dplyr::summarise(
    sensitivity_comparators = dplyr::n(),
    sensitivity_comparators_evaluable = sum(comparator_effect_available),
    sensitivity_comparators_same_direction = sum(
      direction_agrees %in% TRUE
    ),
    sensitivity_comparators_opposite_direction = sum(
      direction_agrees %in% FALSE
    ),
    all_evaluable_sensitivities_agree =
      sensitivity_comparators_evaluable > 0L &
      sensitivity_comparators_opposite_direction == 0L,
    .groups = "drop"
  )

readr::write_csv(
  sensitivity_gene_evidence,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_population_sensitivity.csv.gz"
  )
)
readr::write_csv(
  sensitivity_summary,
  file.path(
    STEP8B_TABLE_DIR,
    "population_sensitivity_summary.csv"
  )
)

sensitivity_summary
```

# 12. Create concordant and discordant RNA candidate tables

Candidate tables are created before ORA. `direction_stable` refers only to
cell-sampling stability, not mouse-level replication.

```{r create-rna-candidate-tables}
primary_rna_candidates <- rna_persistent_evidence |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::left_join(
    candidate_sensitivity_summary,
    by = c(
      "comparison_population" = "primary_population",
      "gene"
    )
  ) |>
  dplyr::mutate(
    candidate_direction = dplyr::case_when(
      RNA_reference_direction_concordant %in% TRUE ~ "concordant",
      RNA_reference_direction_concordant %in% FALSE ~ "discordant",
      TRUE ~ "not_evaluable"
    ),
    candidate_priority = dplyr::case_when(
      !RNA_effect_available ~ "not_evaluable",
      candidate_direction == "concordant" &
        RNA_clear_effect & RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero ~
        "concordant_clear_stable_IQR",
      candidate_direction == "concordant" &
        RNA_clear_effect & RNA_direction_stable ~
        "concordant_clear_stable",
      candidate_direction == "concordant" & RNA_clear_effect ~
        "concordant_clear",
      candidate_direction == "concordant" ~
        "concordant_subthreshold",
      candidate_direction == "discordant" &
        RNA_clear_effect & RNA_direction_stable &
        RNA_downsampling_iqr_excludes_zero ~
        "discordant_clear_stable_IQR",
      candidate_direction == "discordant" &
        RNA_clear_effect & RNA_direction_stable ~
        "discordant_clear_stable",
      candidate_direction == "discordant" & RNA_clear_effect ~
        "discordant_clear",
      candidate_direction == "discordant" ~
        "discordant_subthreshold",
      TRUE ~ "not_evaluable"
    )
  ) |>
  dplyr::group_by(comparison_population, candidate_priority) |>
  dplyr::arrange(
    dplyr::desc(abs(RNA_reference_aligned_log2FC)),
    source_rank,
    .by_group = TRUE
  ) |>
  dplyr::mutate(within_priority_rank = dplyr::row_number()) |>
  dplyr::ungroup()

concordant_candidates <- primary_rna_candidates |>
  dplyr::filter(candidate_direction == "concordant") |>
  dplyr::arrange(
    comparison_population,
    dplyr::desc(RNA_clear_effect),
    dplyr::desc(RNA_direction_stable),
    dplyr::desc(abs(RNA_reference_aligned_log2FC))
  )

discordant_candidates <- primary_rna_candidates |>
  dplyr::filter(candidate_direction == "discordant") |>
  dplyr::arrange(
    comparison_population,
    dplyr::desc(RNA_clear_effect),
    dplyr::desc(RNA_direction_stable),
    dplyr::desc(abs(RNA_reference_aligned_log2FC))
  )

readr::write_csv(
  primary_rna_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_primary_RNA_candidate_inventory.csv"
  )
)
readr::write_csv(
  concordant_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_concordant_RNA_candidates.csv"
  )
)
readr::write_csv(
  discordant_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_discordant_RNA_candidates.csv"
  )
)

primary_rna_candidates |>
  dplyr::count(
    comparison_population,
    reference_program,
    candidate_priority
  )
```

# 13. Perform targeted MSigDB ORA

ORA asks whether stable, clear concordant or discordant candidates are
enriched for Hallmark or GO Biological Process terms relative to persistent
genes evaluable in the same WT population. This deliberately narrow universe
tests structure within the supplied persistent lists; it is not a genome-wide
discovery analysis.

```{r prepare-msigdb-collections}
get_msigdb_collection <- function(
    collection,
    subcollection = NULL
) {
  msigdbr_arguments <- list(
    db_species = "MM",
    species = "Mus musculus",
    collection = collection
  )

  if (!is.null(subcollection)) {
    msigdbr_arguments$subcollection <- subcollection
  }

  do.call(
    msigdbr::msigdbr,
    msigdbr_arguments
  )
}

hallmark_msigdb <- get_msigdb_collection(
  collection = "MH"
) |>
  dplyr::transmute(
    collection = "Hallmark",
    gene_set = gs_name,
    gene = gene_symbol
  ) |>
  dplyr::distinct()

go_bp_msigdb <- get_msigdb_collection(
  collection = "M5",
  subcollection = "GO:BP"
) |>
  dplyr::transmute(
    collection = "GO Biological Process",
    gene_set = gs_name,
    gene = gene_symbol
  ) |>
  dplyr::distinct()

msigdb_pathways <- dplyr::bind_rows(
  hallmark_msigdb,
  go_bp_msigdb
) |>
  dplyr::distinct(collection, gene_set, gene)

empty_ora_result <- tibble::tibble(
  comparison_population = character(),
  ora_query = character(),
  collection = character(),
  gene_set = character(),
  pathway_size_in_universe = integer(),
  overlap_genes = character(),
  overlap_n = integer(),
  universe_n = integer(),
  query_n = integer(),
  p_value = double(),
  fold_enrichment = double(),
  p_adj = double()
)

msigdb_audit <- tibble::tibble(
  msigdbr_version = as.character(
    utils::packageVersion("msigdbr")
  ),
  species = "Mus musculus",
  collection = c("Hallmark", "GO_Biological_Process"),
  pathway_gene_rows = c(
    nrow(hallmark_msigdb),
    nrow(go_bp_msigdb)
  ),
  note = paste(
    "MSigDB terms are used for targeted interpretation after candidate",
    "definition; pooled-library design remains descriptive."
  )
)

readr::write_csv(
  msigdb_audit,
  file.path(STEP8B_DIR, "msigdb_input_audit.csv")
)
```

```{r run-targeted-ora}
ora_query_grid <- tidyr::crossing(
  comparison_population = primary_populations,
  candidate_direction = c("concordant", "discordant"),
  reference_program = c(
    "persistent_HFD_up",
    "persistent_HFD_down"
  )
) |>
  dplyr::mutate(
    ora_query = paste(
      candidate_direction,
      reference_program,
      sep = "__"
    )
  )

ora_candidate_sets <- dplyr::bind_rows(
  lapply(
    seq_len(nrow(ora_query_grid)),
    function(query_index) {
      query_row <- ora_query_grid[query_index, , drop = FALSE]
      query_genes <- primary_rna_candidates |>
        dplyr::filter(
          comparison_population ==
            query_row$comparison_population[[1]],
          candidate_direction == query_row$candidate_direction[[1]],
          reference_program == query_row$reference_program[[1]],
          RNA_effect_available,
          RNA_clear_effect,
          RNA_direction_stable
        ) |>
        dplyr::pull(gene) |>
        unique()

      query_row |>
        dplyr::transmute(
          comparison_population,
          ora_query,
          query_genes = list(query_genes)
        )
    }
  )
)

ora_universes <- primary_rna_candidates |>
  dplyr::filter(RNA_effect_available) |>
  dplyr::group_by(comparison_population) |>
  dplyr::summarise(
    universe_genes = list(unique(gene)),
    .groups = "drop"
  )

run_one_ora <- function(
  population_id,
  query_label,
  query_genes,
  universe_genes
) {
  query_genes <- intersect(unique(query_genes), universe_genes)
  universe_genes <- unique(universe_genes)

  if (length(query_genes) < ORA_MIN_QUERY_GENES) {
    return(empty_ora_result)
  }

  pathway_counts <- msigdb_pathways |>
    dplyr::filter(gene %in% universe_genes) |>
    dplyr::group_by(collection, gene_set) |>
    dplyr::summarise(
      pathway_genes = list(unique(gene)),
      pathway_size_in_universe = dplyr::n_distinct(gene),
      .groups = "drop"
    ) |>
    dplyr::filter(
      pathway_size_in_universe >=
        ORA_MIN_PATHWAY_GENES_IN_UNIVERSE,
      pathway_size_in_universe <=
        ORA_MAX_PATHWAY_GENES_IN_UNIVERSE
    )

  if (nrow(pathway_counts) == 0L) {
    return(empty_ora_result)
  }

  pathway_counts |>
    dplyr::rowwise() |>
    dplyr::mutate(
      overlap_genes = list(
        intersect(query_genes, pathway_genes)
      ),
      overlap_n = length(overlap_genes),
      universe_n = length(universe_genes),
      query_n = length(query_genes),
      p_value = stats::phyper(
        overlap_n - 1L,
        pathway_size_in_universe,
        universe_n - pathway_size_in_universe,
        query_n,
        lower.tail = FALSE
      ),
      fold_enrichment =
        (overlap_n / query_n) /
        (pathway_size_in_universe / universe_n),
      overlap_genes = paste(
        sort(overlap_genes),
        collapse = ";"
      )
    ) |>
    dplyr::ungroup() |>
    dplyr::filter(overlap_n > 0L) |>
    dplyr::group_by(collection) |>
    dplyr::mutate(p_adj = stats::p.adjust(p_value, method = "BH")) |>
    dplyr::ungroup() |>
    dplyr::mutate(
      comparison_population = population_id,
      ora_query = query_label,
      .before = 1
    )
}

ora_input_audit <- ora_candidate_sets |>
  dplyr::left_join(ora_universes, by = "comparison_population") |>
  dplyr::mutate(
    query_n = lengths(query_genes),
    universe_n = lengths(universe_genes),
    ORA_run = query_n >= ORA_MIN_QUERY_GENES,
    exclusion_reason = dplyr::if_else(
      ORA_run,
      NA_character_,
      paste0("fewer than ", ORA_MIN_QUERY_GENES, " query genes")
    )
  )

targeted_ora <- dplyr::bind_rows(
  empty_ora_result,
  lapply(
    seq_len(nrow(ora_input_audit)),
    function(query_index) {
      run_one_ora(
        population_id =
          ora_input_audit$comparison_population[[query_index]],
        query_label = ora_input_audit$ora_query[[query_index]],
        query_genes = ora_input_audit$query_genes[[query_index]],
        universe_genes =
          ora_input_audit$universe_genes[[query_index]]
      )
    }
  )
)

readr::write_csv(
  ora_input_audit |>
    dplyr::select(-query_genes, -universe_genes),
  file.path(STEP8B_TABLE_DIR, "targeted_ORA_input_audit.csv")
)
readr::write_csv(
  targeted_ora,
  file.path(STEP8B_TABLE_DIR, "targeted_MSigDB_ORA_results.csv")
)

targeted_ora |>
  dplyr::arrange(p_adj, dplyr::desc(fold_enrichment)) |>
  dplyr::select(
    comparison_population,
    ora_query,
    collection,
    gene_set,
    overlap_n,
    fold_enrichment,
    p_value,
    p_adj,
    overlap_genes
  ) |>
  utils::head(30)
```

# 14. Characterize ATAC effects independently

ATAC effects are calculated without requiring RNA concordance. Peak context
uses this priority: exact plus-or-minus 2-kb TSS overlap, gene-body overlap
outside that promoter window, then intergenic. `nearest_gene` is descriptive
context and is not a regulatory assignment.

```{r annotate-atac-feature-context}
annotation <- build_grcm39_annotation(GRCM39_GTF)
gene_annotation <- annotation[
  as.character(annotation$type) == "gene"
]

gene_annotation$gene_name <- as.character(
  gene_annotation$gene_name
)
gene_annotation$gene_id <- as.character(gene_annotation$gene_id)

compatible_genes <- as.character(
  GenomeInfoDb::seqnames(gene_annotation)
) %in% GenomeInfoDb::seqlevels(atac_ranges)
gene_ranges <- gene_annotation[compatible_genes]

gene_strand <- as.character(GenomicRanges::strand(gene_ranges))
tss_position <- ifelse(
  gene_strand == "-",
  GenomicRanges::end(gene_ranges),
  GenomicRanges::start(gene_ranges)
)

all_gene_promoters <- GenomicRanges::GRanges(
  seqnames = GenomeInfoDb::seqnames(gene_ranges),
  ranges = IRanges::IRanges(
    start = pmax(1L, tss_position - TSS_WINDOW_BP),
    end = tss_position + TSS_WINDOW_BP
  ),
  strand = "*"
)

body_hits <- GenomicRanges::findOverlaps(
  atac_ranges,
  gene_ranges,
  ignore.strand = TRUE
)
promoter_hits <- GenomicRanges::findOverlaps(
  atac_ranges,
  all_gene_promoters,
  ignore.strand = TRUE
)

peak_context <- rep("intergenic", length(atac_ranges))
peak_context[unique(S4Vectors::queryHits(body_hits))] <-
  "gene_body_nonpromoter"
peak_context[unique(S4Vectors::queryHits(promoter_hits))] <-
  "promoter_proximal"

nearest_hits <- GenomicRanges::distanceToNearest(
  atac_ranges,
  gene_ranges,
  ignore.strand = TRUE
)

nearest_gene <- rep(NA_character_, length(atac_ranges))
nearest_gene_id <- rep(NA_character_, length(atac_ranges))
nearest_gene_distance <- rep(NA_integer_, length(atac_ranges))

nearest_gene[S4Vectors::queryHits(nearest_hits)] <-
  gene_ranges$gene_name[S4Vectors::subjectHits(nearest_hits)]
nearest_gene_id[S4Vectors::queryHits(nearest_hits)] <-
  gene_ranges$gene_id[S4Vectors::subjectHits(nearest_hits)]
nearest_gene_distance[S4Vectors::queryHits(nearest_hits)] <-
  S4Vectors::mcols(nearest_hits)$distance

atac_feature_context <- tibble::tibble(
  peak = names(atac_ranges),
  chromosome = as.character(GenomeInfoDb::seqnames(atac_ranges)),
  peak_start = GenomicRanges::start(atac_ranges),
  peak_end = GenomicRanges::end(atac_ranges),
  genomic_context = peak_context,
  nearest_gene = nearest_gene,
  nearest_gene_id = nearest_gene_id,
  nearest_gene_distance_bp = nearest_gene_distance
)

readr::write_csv(
  atac_feature_context,
  file.path(STEP8B_TABLE_DIR, "ATAC_feature_context_GRCm39.csv.gz")
)
```

```{r calculate-atac-effects}
atac_effects <- calculate_population_effects(
  count_matrix = atac_counts,
  feature_label = "peak",
  minimum_total_counts = ATAC_MIN_TOTAL_COUNTS,
  minimum_detection_percent = ATAC_MIN_DETECTION_PERCENT
)

atac_top_candidates <- select_top_effects(
  effects = atac_effects,
  absolute_log2fc_threshold = ATAC_CLEAR_LOG2FC,
  top_features_per_direction = ATAC_TOP_FEATURES_PER_DIRECTION
) |>
  dplyr::left_join(
    atac_feature_context,
    by = c("feature" = "peak")
  )

readr::write_csv(
  atac_effects,
  file.path(
    STEP8B_TABLE_DIR,
    "ATAC_pooled_library_effects_by_matched_population.csv.gz"
  )
)
readr::write_csv(
  atac_top_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "ATAC_top_diet_candidates_by_matched_population.csv"
  )
)
```

```{r run-atac-downsampling}
promoter_peak_overlaps <- read_table_quietly(PROMOTER_OVERLAP_PATH)

assert_columns(
  promoter_peak_overlaps,
  c(
    "gene",
    "gene_id",
    "peak",
    "transcriptional_offset_bp",
    "peak_overlaps_tss"
  ),
  "persistent-gene promoter-overlap table"
)

promoter_geometry_columns <- c(
  "gene",
  "gene_id",
  "chromosome",
  "strand",
  "tss",
  "promoter_start",
  "promoter_end",
  "peak",
  "peak_start",
  "peak_end",
  "peak_center",
  "transcriptional_offset_bp",
  "peak_overlaps_tss"
)

promoter_peak_overlaps <- promoter_peak_overlaps |>
  dplyr::select(dplyr::all_of(promoter_geometry_columns)) |>
  dplyr::distinct(gene_id, peak, .keep_all = TRUE)

missing_promoter_peaks <- setdiff(
  promoter_peak_overlaps$peak,
  rownames(atac_counts)
)

if (length(missing_promoter_peaks) > 0L) {
  stop(
    "Step 8 promoter overlaps contain peaks absent from the WT ATAC assay.",
    call. = FALSE
  )
}

atac_stability <- downsample_effect_stability(
  count_matrix = atac_counts,
  full_effects = atac_effects,
  required_features = unique(promoter_peak_overlaps$peak),
  maximum_features_per_population =
    DOWNSAMPLING_MAX_FEATURES_PER_POPULATION,
  random_seed = 20260917
)

atac_stability_key <- prepare_stability(
  atac_stability,
  "ATAC_"
) |>
  dplyr::rename(peak = feature)

readr::write_csv(
  atac_stability,
  file.path(
    STEP8B_TABLE_DIR,
    "ATAC_balanced_downsampling_stability.csv"
  )
)

atac_effect_summary <- atac_effects |>
  dplyr::mutate(
    direction = observed_direction(log2FC_comparison_vs_reference),
    ATAC_clear_effect =
      abs(log2FC_comparison_vs_reference) >= ATAC_CLEAR_LOG2FC
  ) |>
  dplyr::left_join(
    atac_feature_context,
    by = c("feature" = "peak")
  ) |>
  dplyr::left_join(
    atac_stability_key,
    by = c(
      "comparison_population",
      "feature" = "peak"
    )
  ) |>
  dplyr::mutate(
    ATAC_stability_evaluated = !is.na(
      ATAC_sign_concordance_percent
    ),
    ATAC_direction_stable = dplyr::coalesce(
      ATAC_direction_stable,
      FALSE
    )
  ) |>
  dplyr::group_by(
    comparison_population,
    display_label,
    analysis_role,
    cell_support_tier,
    direction,
    genomic_context
  ) |>
  dplyr::summarise(
    filtered_peaks = dplyr::n(),
    clear_effect_peaks = sum(ATAC_clear_effect),
    stability_evaluated_peaks = sum(ATAC_stability_evaluated),
    clear_stable_peaks = sum(
      ATAC_clear_effect & ATAC_direction_stable
    ),
    .groups = "drop"
  )

readr::write_csv(
  atac_effect_summary,
  file.path(
    STEP8B_TABLE_DIR,
    "ATAC_effect_summary_by_population_direction_context.csv"
  )
)

atac_effect_summary |>
  dplyr::filter(comparison_population %in% primary_populations)
```

# 15. Integrate RNA with exact promoter accessibility

All exact promoter-overlapping peaks are retained in the peak-level table.
The representative peak is the evaluable overlapping peak with the largest
absolute ATAC effect and is used only for compact plotting.

```{r build-exact-promoter-evidence}
promoter_atac_effect_key <- atac_effects |>
  dplyr::select(
    comparison_population,
    feature,
    reference_nuclei,
    comparison_nuclei,
    reference_counts,
    comparison_counts,
    total_counts,
    reference_CPM,
    comparison_CPM,
    log2FC_comparison_vs_reference,
    reference_percent_detected,
    comparison_percent_detected,
    detection_difference_percentage_points
  ) |>
  dplyr::rename(peak = feature) |>
  dplyr::mutate(ATAC_effect_available = TRUE)

promoter_peak_evidence <- tidyr::crossing(
  promoter_peak_overlaps,
  population_key
) |>
  dplyr::left_join(
    promoter_atac_effect_key,
    by = c("comparison_population", "peak")
  ) |>
  dplyr::left_join(
    atac_stability_key,
    by = c("comparison_population", "peak")
  ) |>
  dplyr::left_join(
    rna_persistent_evidence |>
      dplyr::select(
        comparison_population,
        gene,
        reference_direction,
        reference_program,
        source_rank,
        avg_log2FC_day21,
        avg_log2FC_day50,
        RNA_effect_available,
        RNA_observed_direction,
        RNA_reference_aligned_log2FC,
        RNA_clear_effect,
        RNA_direction_stable,
        RNA_downsampling_iqr_excludes_zero,
        RNA_evidence_class
      ),
    by = c("comparison_population", "gene")
  ) |>
  dplyr::mutate(
    ATAC_effect_available = dplyr::coalesce(
      ATAC_effect_available,
      FALSE
    ),
    ATAC_observed_direction = observed_direction(
      log2FC_comparison_vs_reference
    ),
    ATAC_reference_aligned_log2FC = reference_aligned_effect(
      log2FC_comparison_vs_reference,
      reference_direction
    ),
    ATAC_source_direction_concordant = dplyr::case_when(
      !ATAC_effect_available ~ NA,
      ATAC_reference_aligned_log2FC > 0 ~ TRUE,
      ATAC_reference_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    ATAC_clear_effect =
      ATAC_effect_available &
      abs(log2FC_comparison_vs_reference) >= ATAC_CLEAR_LOG2FC,
    ATAC_stability_available = !is.na(
      ATAC_sign_concordance_percent
    ),
    ATAC_direction_stable = dplyr::coalesce(
      ATAC_direction_stable,
      FALSE
    ),
    ATAC_downsampling_iqr_excludes_zero = dplyr::coalesce(
      ATAC_downsampling_iqr_excludes_zero,
      FALSE
    ),
    RNA_ATAC_direction_match = dplyr::case_when(
      !RNA_effect_available | !ATAC_effect_available ~ NA,
      sign(RNA_reference_aligned_log2FC) ==
        sign(ATAC_reference_aligned_log2FC) ~ TRUE,
      TRUE ~ FALSE
    ),
    promoter_support_class = dplyr::case_when(
      !ATAC_effect_available ~ "promoter_peak_below_ATAC_filters",
      RNA_ATAC_direction_match %in% TRUE &
        ATAC_clear_effect & ATAC_direction_stable ~
        "RNA_matched_clear_stable_promoter_ATAC",
      RNA_ATAC_direction_match %in% TRUE & ATAC_clear_effect ~
        "RNA_matched_clear_promoter_ATAC",
      RNA_ATAC_direction_match %in% TRUE ~
        "RNA_matched_subthreshold_promoter_ATAC",
      RNA_ATAC_direction_match %in% FALSE & ATAC_clear_effect ~
        "RNA_opposed_clear_promoter_ATAC",
      RNA_ATAC_direction_match %in% FALSE ~
        "RNA_opposed_subthreshold_promoter_ATAC",
      TRUE ~ "not_classified"
    )
  ) |>
  dplyr::arrange(
    analysis_role,
    comparison_population,
    reference_program,
    source_rank,
    transcriptional_offset_bp
  )

readr::write_csv(
  promoter_peak_evidence,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_exact_promoter_evidence_by_population_peak.csv.gz"
  )
)
```

```{r summarize-exact-promoter-evidence}
promoter_gene_summary_observed <- promoter_peak_evidence |>
  dplyr::filter(primary_RNA_ATAC) |>
  dplyr::group_by(comparison_population, gene) |>
  dplyr::summarise(
    promoter_peaks_in_assay = dplyr::n_distinct(peak),
    promoter_peaks_evaluable = dplyr::n_distinct(
      peak[ATAC_effect_available]
    ),
    promoter_peaks_RNA_matched = dplyr::n_distinct(
      peak[RNA_ATAC_direction_match %in% TRUE]
    ),
    promoter_peaks_RNA_opposed = dplyr::n_distinct(
      peak[RNA_ATAC_direction_match %in% FALSE]
    ),
    promoter_peaks_RNA_matched_clear = dplyr::n_distinct(
      peak[
        RNA_ATAC_direction_match %in% TRUE &
          ATAC_clear_effect
      ]
    ),
    promoter_peaks_RNA_matched_clear_stable = dplyr::n_distinct(
      peak[
        RNA_ATAC_direction_match %in% TRUE &
          ATAC_clear_effect & ATAC_direction_stable
      ]
    ),
    promoter_peak_directions_mixed =
      promoter_peaks_RNA_matched > 0L &
      promoter_peaks_RNA_opposed > 0L,
    .groups = "drop"
  )

representative_promoter_peak <- promoter_peak_evidence |>
  dplyr::filter(primary_RNA_ATAC, ATAC_effect_available) |>
  dplyr::mutate(
    absolute_ATAC_log2FC = abs(log2FC_comparison_vs_reference)
  ) |>
  dplyr::group_by(comparison_population, gene) |>
  dplyr::arrange(
    dplyr::desc(absolute_ATAC_log2FC),
    abs(transcriptional_offset_bp),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = 1L) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    comparison_population,
    gene,
    representative_promoter_peak = peak,
    representative_peak_offset_bp = transcriptional_offset_bp,
    representative_peak_overlaps_TSS = peak_overlaps_tss,
    representative_ATAC_log2FC =
      log2FC_comparison_vs_reference,
    representative_ATAC_reference_aligned_log2FC =
      ATAC_reference_aligned_log2FC,
    representative_ATAC_clear_effect = ATAC_clear_effect,
    representative_ATAC_direction_stable =
      ATAC_direction_stable,
    representative_RNA_ATAC_direction_match =
      RNA_ATAC_direction_match,
    representative_promoter_support_class =
      promoter_support_class
  )

integrated_primary_candidates <- primary_rna_candidates |>
  dplyr::left_join(
    promoter_gene_summary_observed,
    by = c("comparison_population", "gene")
  ) |>
  dplyr::left_join(
    representative_promoter_peak,
    by = c("comparison_population", "gene")
  ) |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("promoter_peaks_"),
      ~ dplyr::coalesce(.x, 0L)
    ),
    promoter_peak_directions_mixed = dplyr::coalesce(
      promoter_peak_directions_mixed,
      FALSE
    ),
    integrated_priority = dplyr::case_when(
      candidate_direction == "concordant" &
        RNA_clear_effect & RNA_direction_stable &
        promoter_peaks_RNA_matched_clear_stable > 0L ~
        "Tier_1_stable_source_concordant_RNA_plus_matched_promoter_ATAC",
      candidate_direction == "concordant" &
        RNA_clear_effect & RNA_direction_stable ~
        "Tier_2_stable_source_concordant_RNA",
      candidate_direction == "concordant" & RNA_clear_effect ~
        "Tier_3_source_concordant_clear_RNA",
      candidate_direction == "discordant" &
        RNA_clear_effect & RNA_direction_stable ~
        "discordance_priority_stable_clear_RNA",
      candidate_direction == "discordant" & RNA_clear_effect ~
        "discordance_priority_clear_RNA",
      candidate_direction == "concordant" ~
        "concordant_subthreshold_RNA",
      candidate_direction == "discordant" ~
        "discordant_subthreshold_RNA",
      TRUE ~ "not_evaluable"
    )
  ) |>
  dplyr::arrange(
    comparison_population,
    factor(
      integrated_priority,
      levels = c(
        "Tier_1_stable_source_concordant_RNA_plus_matched_promoter_ATAC",
        "Tier_2_stable_source_concordant_RNA",
        "Tier_3_source_concordant_clear_RNA",
        "discordance_priority_stable_clear_RNA",
        "discordance_priority_clear_RNA",
        "concordant_subthreshold_RNA",
        "discordant_subthreshold_RNA",
        "not_evaluable"
      )
    ),
    dplyr::desc(abs(RNA_reference_aligned_log2FC))
  )

readr::write_csv(
  promoter_gene_summary_observed,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_exact_promoter_summary.csv"
  )
)
readr::write_csv(
  integrated_primary_candidates,
  file.path(
    STEP8B_TABLE_DIR,
    "persistent_gene_integrated_primary_candidates.csv"
  )
)

integrated_primary_candidates |>
  dplyr::count(comparison_population, integrated_priority)
```

# 16. Generate review figures

The primary source plots combine HFD-up and HFD-down genes on one signed
axis. Green indicates directional agreement; red indicates disagreement.
Point shape distinguishes clear and stable effects from weaker observations.

```{r plot-source-versus-wt-rna}
build_source_scatter <- function(source_column, source_time_label) {
  plot_data <- rna_persistent_evidence |>
    dplyr::filter(comparison_population == "broad_stem") |>
    dplyr::mutate(
      source_effect = .data[[source_column]],
      direction_class = dplyr::case_when(
        !RNA_effect_available ~ "not evaluable",
        RNA_reference_direction_concordant %in% TRUE ~ "concordant",
        RNA_reference_direction_concordant %in% FALSE ~ "discordant",
        TRUE ~ "no directional change"
      ),
      evidence_shape = dplyr::case_when(
        RNA_clear_effect & RNA_direction_stable ~ "clear + stable",
        RNA_clear_effect ~ "clear; unstable",
        RNA_effect_available ~ "subthreshold",
        TRUE ~ "not evaluable"
      )
    )

  label_data <- plot_data |>
    dplyr::filter(RNA_clear_effect) |>
    dplyr::slice_max(
      order_by = abs(source_effect) +
        abs(log2FC_comparison_vs_reference),
      n = 18L,
      with_ties = FALSE
    )

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = source_effect,
      y = log2FC_comparison_vs_reference,
      color = direction_class,
      shape = evidence_shape
    )
  ) +
    ggplot2::geom_hline(yintercept = 0, color = "grey65") +
    ggplot2::geom_vline(xintercept = 0, color = "grey65") +
    ggplot2::geom_point(size = 2.2, alpha = 0.80) +
    ggplot2::geom_text(
      data = label_data,
      ggplot2::aes(label = gene),
      color = "black",
      size = 3,
      check_overlap = TRUE,
      nudge_y = 0.08,
      show.legend = FALSE
    ) +
    ggplot2::scale_color_manual(
      values = c(
        concordant = "#009E73",
        discordant = "#D55E00",
        `no directional change` = "grey55",
        `not evaluable` = "grey82"
      )
    ) +
    ggplot2::labs(
      title = paste0(
        "Persistent genes: source ",
        source_time_label,
        " versus WT d21 broad stem"
      ),
      subtitle = paste(
        "WT broad stem combines annotation units 8 and 6;",
        "green agrees with the source direction and red opposes it."
      ),
      x = paste0("Source ", source_time_label, " HFD versus CON log2FC"),
      y = "WT d21 broad-stem HFD versus CON log2FC",
      color = "Direction",
      shape = "WT evidence",
      caption = paste(
        "Descriptive pooled-library effects; diet and library are confounded."
      )
    ) +
    theme_multiome(base_size = 12)
}

p_source_day21 <- build_source_scatter(
  "avg_log2FC_day21",
  "d21"
)
p_source_day50 <- build_source_scatter(
  "avg_log2FC_day50",
  "d50"
)

print(p_source_day21)
save_native_pdf(
  "01_source_d21_vs_WT_d21_broad_stem_RNA.pdf",
  p_source_day21,
  width = 11,
  height = 8
)
print(p_source_day50)
save_native_pdf(
  "S01_source_d50_vs_WT_d21_broad_stem_RNA.pdf",
  p_source_day50,
  width = 11,
  height = 8
)
```

```{r plot-ranked-enrichment}
ranked_plot_data <- ranked_enrichment |>
  dplyr::mutate(
    display_label = factor(
      display_label,
      levels = rev(population_support$display_label)
    ),
    expected_result = dplyr::if_else(
      expected_direction_recovered,
      "expected direction",
      "opposite direction"
    )
  )

p_ranked_enrichment <- ggplot2::ggplot(
  ranked_plot_data,
  ggplot2::aes(
    x = NES,
    y = display_label,
    color = expected_result,
    shape = pathway
  )
) +
  ggplot2::geom_vline(xintercept = 0, color = "grey70") +
  ggplot2::geom_point(size = 3) +
  ggplot2::scale_color_manual(
    values = c(
      `expected direction` = "#009E73",
      `opposite direction` = "#D55E00"
    )
  ) +
  ggplot2::labs(
    title = "Ranked recovery of persistent HFD programs",
    subtitle = paste(
      "Expected: positive NES for persistent HFD-up and negative NES",
      "for persistent HFD-down."
    ),
    x = "Normalized enrichment score",
    y = NULL,
    color = NULL,
    shape = "Reference list",
    caption = paste(
      "Gene-rank enrichment is descriptive and is not evidence of",
      "mouse-level reproducibility."
    )
  ) +
  theme_multiome(base_size = 11)

print(p_ranked_enrichment)
save_native_pdf(
  "02_persistent_program_ranked_RNA_enrichment.pdf",
  p_ranked_enrichment,
  width = 12,
  height = 8
)
```

```{r plot-population-sensitivity}
sensitivity_plot_data <- sensitivity_gene_evidence |>
  dplyr::filter(both_evaluable) |>
  dplyr::mutate(
    direction_class = dplyr::if_else(
      direction_agrees,
      "same direction",
      "opposite direction"
    ),
    pair_label = paste0(
      primary_population,
      " vs ",
      comparator_population
    )
  )

p_sensitivity <- ggplot2::ggplot(
  sensitivity_plot_data,
  ggplot2::aes(
    x = primary_log2FC,
    y = comparator_log2FC,
    color = direction_class
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75") +
  ggplot2::geom_vline(xintercept = 0, color = "grey75") +
  ggplot2::geom_point(size = 1.3, alpha = 0.65) +
  ggplot2::facet_wrap(~ pair_label, scales = "free", ncol = 2) +
  ggplot2::scale_color_manual(
    values = c(
      `same direction` = "#0072B2",
      `opposite direction` = "#D55E00"
    )
  ) +
  ggplot2::labs(
    title = "Sensitivity of persistent-gene RNA effects to regrouping",
    subtitle = paste(
      "Each comparator overlaps its primary composite; this tests",
      "annotation-boundary dependence, not replication."
    ),
    x = "Primary composite HFD versus CON log2FC",
    y = "Sensitivity/component HFD versus CON log2FC",
    color = NULL
  ) +
  theme_multiome(base_size = 10)

print(p_sensitivity)
save_native_pdf(
  "03_persistent_gene_population_sensitivity.pdf",
  p_sensitivity,
  width = 13,
  height = 12
)
```

```{r plot-atac-effect-summary}
atac_plot_data <- atac_effect_summary |>
  dplyr::filter(
    comparison_population %in% primary_populations,
    clear_effect_peaks > 0L
  ) |>
  dplyr::mutate(
    display_label = factor(
      display_label,
      levels = population_support$display_label[
        population_support$comparison_population %in%
          primary_populations
      ]
    )
  )

p_atac_summary <- ggplot2::ggplot(
  atac_plot_data,
  ggplot2::aes(
    x = display_label,
    y = clear_effect_peaks,
    fill = genomic_context
  )
) +
  ggplot2::geom_col() +
  ggplot2::facet_wrap(~ direction, scales = "free_y") +
  ggplot2::scale_fill_manual(
    values = c(
      promoter_proximal = "#CC79A7",
      gene_body_nonpromoter = "#56B4E9",
      intergenic = "#999999"
    )
  ) +
  ggplot2::labs(
    title = "Independent ATAC diet-effect inventory",
    subtitle = paste(
      "Clear effect denotes absolute pooled-library log2FC >=",
      ATAC_CLEAR_LOG2FC
    ),
    x = NULL,
    y = "Clear-effect peaks",
    fill = "Genomic context",
    caption = paste(
      "Nearest-gene and genomic-context labels are descriptive;",
      "they do not establish regulation."
    )
  ) +
  theme_multiome(base_size = 11) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1)
  )

print(p_atac_summary)
save_native_pdf(
  "04_primary_population_ATAC_effect_inventory.pdf",
  p_atac_summary,
  width = 12,
  height = 8
)
```

```{r plot-rna-promoter-atac}
promoter_plot_data <- integrated_primary_candidates |>
  dplyr::filter(
    RNA_effect_available,
    !is.na(representative_ATAC_log2FC)
  ) |>
  dplyr::mutate(
    RNA_ATAC_class = dplyr::if_else(
      representative_RNA_ATAC_direction_match %in% TRUE,
      "RNA and promoter ATAC match",
      "RNA and promoter ATAC oppose"
    )
  )

p_rna_promoter <- ggplot2::ggplot(
  promoter_plot_data,
  ggplot2::aes(
    x = log2FC_comparison_vs_reference,
    y = representative_ATAC_log2FC,
    color = RNA_ATAC_class
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey70") +
  ggplot2::geom_vline(xintercept = 0, color = "grey70") +
  ggplot2::geom_point(size = 2, alpha = 0.75) +
  ggplot2::facet_wrap(~ display_label, scales = "free") +
  ggplot2::scale_color_manual(
    values = c(
      `RNA and promoter ATAC match` = "#009E73",
      `RNA and promoter ATAC oppose` = "#D55E00"
    )
  ) +
  ggplot2::labs(
    title = "Persistent-gene RNA and exact-promoter ATAC effects",
    subtitle = paste(
      "One representative exact promoter peak per gene is shown;",
      "all peaks remain in the exported evidence table."
    ),
    x = "RNA HFD versus CON log2FC",
    y = "Representative promoter ATAC HFD versus CON log2FC",
    color = NULL,
    caption = paste(
      "Directional agreement is regulatory context, not proof of a",
      "causal peak-to-gene relationship."
    )
  ) +
  theme_multiome(base_size = 10)

print(p_rna_promoter)
save_native_pdf(
  "05_persistent_gene_RNA_exact_promoter_ATAC.pdf",
  p_rna_promoter,
  width = 13,
  height = 8
)
```

# 17. Write completion summaries

```{r write-completion-summary}
primary_summary <- integrated_primary_candidates |>
  dplyr::group_by(
    comparison_population,
    display_label,
    source_population,
    cell_support_tier
  ) |>
  dplyr::summarise(
    persistent_genes = dplyr::n(),
    RNA_effect_evaluable = sum(RNA_effect_available),
    clear_stable_concordant_RNA = sum(
      candidate_direction == "concordant" &
        RNA_clear_effect & RNA_direction_stable
    ),
    clear_stable_discordant_RNA = sum(
      candidate_direction == "discordant" &
        RNA_clear_effect & RNA_direction_stable
    ),
    Tier_1_integrated_candidates = sum(
      integrated_priority ==
        "Tier_1_stable_source_concordant_RNA_plus_matched_promoter_ATAC"
    ),
    genes_with_evaluable_exact_promoter_ATAC = sum(
      promoter_peaks_evaluable > 0L
    ),
    .groups = "drop"
  )

ranked_primary_summary <- ranked_enrichment |>
  dplyr::filter(comparison_population %in% primary_populations) |>
  dplyr::select(
    comparison_population,
    pathway,
    NES,
    pval,
    padj,
    expected_direction_recovered
  ) |>
  tidyr::pivot_wider(
    names_from = pathway,
    values_from = c(
      NES,
      pval,
      padj,
      expected_direction_recovered
    ),
    names_glue = "{pathway}_{.value}"
  )

step8b_primary_summary <- primary_summary |>
  dplyr::left_join(
    ranked_primary_summary,
    by = "comparison_population"
  ) |>
  dplyr::mutate(
    replicated_diet_design = FALSE,
    biological_inference_performed = FALSE,
    exact_promoter_integration_performed = TRUE,
    distal_regulatory_integration_performed = FALSE,
    HyperMet_overlay_performed = FALSE,
    HypoImmune_overlay_performed = FALSE
  )

readr::write_csv(
  step8b_primary_summary,
  file.path(STEP8B_DIR, "step8b_primary_population_summary.csv")
)

writeLines(
  c(
    paste("Step 8B completed:", Sys.time()),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste(
      "Analysis unit source:",
      analysis_unit_column
    ),
    paste(
      "Source-matched primary populations:",
      paste(primary_populations, collapse = ",")
    ),
    paste(
      "Persistent reference genes:",
      nrow(reference_genes)
    ),
    "Strict Step 8 unit-8 reproduction passed: TRUE",
    "Rank-based persistent-list enrichment performed: TRUE",
    "Targeted MSigDB ORA performed after candidate definition: TRUE",
    "Independent ATAC characterization performed: TRUE",
    "Exact plus-or-minus 2-kb TSS integration performed: TRUE",
    "Distal regulatory integration performed: FALSE",
    "HyperMet overlay performed: FALSE",
    "HypoImmune overlay performed: FALSE",
    "Cross-genotype comparison performed: FALSE",
    "Replicated diet inference performed: FALSE"
  ),
  con = file.path(STEP8B_DIR, "STEP_08B_COMPLETE.txt")
)

step8b_primary_summary
```

# Interpretation checkpoint

Before carrying these benchmarks into PPAR or IL17, review:

1. whether the broad-stem and all-goblet ranked enrichments recover the
   expected directions;
2. which clear and stable concordant or discordant genes depend on a single
   component population;
3. whether including unit 13 materially changes the all-goblet result;
4. whether exact-promoter ATAC evidence supports, opposes, or is absent for
   the leading RNA candidates; and
5. whether any ORA term is driven by only one or two genes despite meeting the
   numerical query threshold.

None of these checks removes the pooled-library limitation. Final biological
claims require replicated material or orthogonal validation.
