---
title: "Step 8C - Decompose the WT broad-stem diet response and prioritize regulatory loci"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 8B established that the source-matched WT broad-stem population (`8 + 6`) recovers both persistent HFD-associated RNA programs better than strict ISC-like unit 8 alone. It also showed a large diet-associated shift in the composition of that population: HFD contains many more crypt-progenitor unit-6 nuclei and fewer ISC-like unit-8 nuclei.

Step 8C separates those two phenomena. It asks whether a broad-stem diet effect is attributable primarily to:

1.  a response occurring in both component states;
2.  a response concentrated in crypt-progenitor unit 6;
3.  a response concentrated in ISC-like unit 8;
4.  opposing effects in the two states; or
5.  the changed unit-8/unit-6 mixture itself.

The decomposition is performed on linear quantities so that the terms add exactly. RNA uses the mean per-cell library-normalized abundance (counts per 10,000); ATAC uses the fraction of nuclei with at least one fragment in each peak. Existing Step 8B pooled-library log2 fold changes remain beside these quantities as the familiar effect-size summaries.

The notebook then creates stable, component-resolved ATAC peak sets for the next motif-enrichment checkpoint and reprioritizes persistent-gene exact promoter candidates. It does not perform motif enrichment itself. This keeps the peak-set definition inspectable before sequence motifs are tested.

## Statistical limitation

WT contains one pooled CON library and one pooled HFD library. Diet and library are perfectly confounded. The decomposition is an exact descriptive partition of these two pooled samples; it is not a replicate-aware estimate and does not recover mouse-level variance. Repeated cell downsampling checks sampling stability only.

# 1. Setup

Run in a fresh WT v2 R session:

``` r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

check_r_version()
package_preflight(packages = required_packages)
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Signac)
  library(Matrix)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
})

ggplot2::theme_set(theme_multiome())
future::plan(future::sequential)
set.seed(20260918)

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 8C is defined for the WT v2 benchmark. Set ",
    "MULTIOME_DATASET_VERSION=v2_resequenced and ",
    "MULTIOME_COHORT_ID=wt.",
    call. = FALSE
  )
}
```

# 2. Define inputs, outputs, and evidence rules

```{r define-inputs-and-parameters}
STEP6_DIR <- file.path(OUTPUT_DIR, "06_annotation")
STEP8B_DIR <- file.path(
  OUTPUT_DIR,
  "08b_source_matched_persistent_hfd"
)
STEP8B_TABLE_DIR <- file.path(STEP8B_DIR, "tables")

STEP8C_DIR <- file.path(
  OUTPUT_DIR,
  "08c_broad_stem_decomposition"
)
STEP8C_TABLE_DIR <- file.path(STEP8C_DIR, "tables")
STEP8C_PLOT_DIR <- file.path(STEP8C_DIR, "plots")

dir.create(STEP8C_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8C_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8C_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

STEP6_OBJECT_PATH <- file.path(
  STEP6_DIR,
  "multiome_annotated.rds"
)
STEP8B_COMPLETE_MARKER <- file.path(
  STEP8B_DIR,
  "STEP_08B_COMPLETE.txt"
)
REFERENCE_GENE_PATH <- file.path(
  DATASET_CONFIG_DIR,
  "reference_programs",
  "persistent_hfd_genes.csv"
)
STEP8B_RNA_EFFECT_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "RNA_pooled_library_effects_by_matched_population.csv.gz"
)
STEP8B_ATAC_EFFECT_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "ATAC_pooled_library_effects_by_matched_population.csv.gz"
)
STEP8B_RNA_STABILITY_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "RNA_balanced_downsampling_stability.csv"
)
STEP8B_ATAC_CONTEXT_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "ATAC_feature_context_GRCm39.csv.gz"
)
STEP8B_PROMOTER_EVIDENCE_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "persistent_gene_exact_promoter_evidence_by_population_peak.csv.gz"
)

required_inputs <- c(
  Step6_object = STEP6_OBJECT_PATH,
  Step8B_marker = STEP8B_COMPLETE_MARKER,
  persistent_reference = REFERENCE_GENE_PATH,
  Step8B_RNA_effects = STEP8B_RNA_EFFECT_PATH,
  Step8B_ATAC_effects = STEP8B_ATAC_EFFECT_PATH,
  Step8B_RNA_stability = STEP8B_RNA_STABILITY_PATH,
  Step8B_ATAC_context = STEP8B_ATAC_CONTEXT_PATH,
  Step8B_promoter_evidence = STEP8B_PROMOTER_EVIDENCE_PATH
)

for (input_label in names(required_inputs)) {
  assert_file(required_inputs[[input_label]], input_label)
}

REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"
BROAD_STEM_UNITS <- c("8", "6")
UNIT8 <- "8"
UNIT6 <- "6"

RNA_SCALE_FACTOR <- 10000
RNA_LINEAR_PSEUDOCOUNT <- 0.01
RNA_CLEAR_LOG2FC <- 0.50
RNA_CLEAR_DETECTION_DELTA <- 5

ATAC_CLEAR_LOG2FC <- 0.50
ATAC_CLEAR_DETECTION_DELTA <- 2
ATAC_DOWNSAMPLING_ITERATIONS <- 25L
MINIMUM_SIGN_CONCORDANCE_PERCENT <- 80

MINIMUM_MOTIF_SET_SIZE <- 25L
PREVIEW_PLOTS_IN_RSTUDIO <- FALSE

analysis_parameters <- tibble::tibble(
  parameter = c(
    "reference diet",
    "comparison diet",
    "broad-stem component units",
    "RNA linear scale factor",
    "RNA clear-effect absolute pooled log2FC",
    "RNA clear-effect absolute detection delta",
    "ATAC clear-effect absolute pooled log2FC",
    "ATAC clear-effect absolute detection delta",
    "ATAC downsampling iterations",
    "minimum sign-concordance percent",
    "minimum motif-set size",
    "replicated diet design",
    "biological inference performed"
  ),
  value = c(
    REFERENCE_DIET,
    COMPARISON_DIET,
    paste(BROAD_STEM_UNITS, collapse = ";"),
    RNA_SCALE_FACTOR,
    RNA_CLEAR_LOG2FC,
    RNA_CLEAR_DETECTION_DELTA,
    ATAC_CLEAR_LOG2FC,
    ATAC_CLEAR_DETECTION_DELTA,
    ATAC_DOWNSAMPLING_ITERATIONS,
    MINIMUM_SIGN_CONCORDANCE_PERCENT,
    MINIMUM_MOTIF_SET_SIZE,
    FALSE,
    FALSE
  )
)

readr::write_csv(
  analysis_parameters,
  file.path(STEP8C_DIR, "step8c_parameters.csv")
)
```

# 3. Helper functions

```{r helper-functions}
read_table_quietly <- function(path) {
  suppressMessages(
    readr::read_csv(
      path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
}

assert_columns <- function(data, required_columns, table_label) {
  missing_columns <- setdiff(required_columns, colnames(data))

  if (length(missing_columns) > 0L) {
    stop(
      table_label,
      " is missing required column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
}

align_matrix_to_metadata <- function(
  feature_matrix,
  metadata_cells,
  matrix_label
) {
  if (!setequal(colnames(feature_matrix), metadata_cells)) {
    stop(
      matrix_label,
      " and Step 6 metadata do not contain the same cells.",
      call. = FALSE
    )
  }

  if (!identical(colnames(feature_matrix), metadata_cells)) {
    message(matrix_label, " columns are being reordered to metadata.")
    feature_matrix <- feature_matrix[
      ,
      metadata_cells,
      drop = FALSE
    ]
  }

  feature_matrix
}

# preview_plot <- function(plot_object) {
#   if (isTRUE(PREVIEW_PLOTS_IN_RSTUDIO)) print(plot_object)
#   invisible(plot_object)
# }

save_native_pdf <- function(
  filename,
  plot_object,
  width = 11,
  height = 7
) {
  output_path <- file.path(STEP8C_PLOT_DIR, filename)
  grDevices::pdf(
    file = output_path,
    width = width,
    height = height,
    family = "Helvetica",
    useDingbats = FALSE
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot_object)
  invisible(output_path)
}

effect_direction <- function(effect) {
  dplyr::case_when(
    is.na(effect) ~ NA_character_,
    effect > 0 ~ "HFD_higher",
    effect < 0 ~ "CON_higher",
    TRUE ~ "no_change"
  )
}

source_aligned_effect <- function(effect, reference_direction) {
  dplyr::case_when(
    is.na(effect) ~ NA_real_,
    reference_direction == "HFD_higher" ~ effect,
    reference_direction == "CON_higher" ~ -effect,
    TRUE ~ NA_real_
  )
}

summarize_matrix_by_component <- function(
  feature_matrix,
  metadata,
  value_label,
  binary = FALSE
) {
  dplyr::bind_rows(
    lapply(
      BROAD_STEM_UNITS,
      function(component_unit) {
        dplyr::bind_rows(
          lapply(
            c(REFERENCE_DIET, COMPARISON_DIET),
            function(diet_value) {
              selected_cells <- metadata$cell_barcode[
                metadata$analysis_unit == component_unit &
                  metadata$diet == diet_value
              ]

              if (length(selected_cells) == 0L) {
                stop(
                  "No cells for unit ", component_unit,
                  " and diet ", diet_value, ".",
                  call. = FALSE
                )
              }

              feature_values <- Matrix::rowMeans(
                feature_matrix[, selected_cells, drop = FALSE]
              )

              tibble::tibble(
                feature = rownames(feature_matrix),
                component_unit = component_unit,
                component = dplyr::recode(
                  component_unit,
                  `8` = "unit8",
                  `6` = "unit6"
                ),
                diet = diet_value,
                nuclei = length(selected_cells),
                value = as.numeric(feature_values),
                value_type = value_label,
                binary = binary
              )
            }
          )
        )
      }
    )
  )
}

midpoint_decomposition <- function(
  component_means,
  component_proportions,
  pseudocount,
  scale_to_percentage_points = FALSE
) {
  means_wide <- component_means |>
    dplyr::select(feature, component, diet, value) |>
    tidyr::pivot_wider(
      names_from = c(component, diet),
      values_from = value,
      names_glue = "{component}_{diet}"
    )

  required_mean_columns <- c(
    "unit8_CON",
    "unit8_HFD",
    "unit6_CON",
    "unit6_HFD"
  )
  assert_columns(
    means_wide,
    c("feature", required_mean_columns),
    "component-mean table"
  )

  proportion_value <- function(component_value, diet_value) {
    selected <- component_proportions |>
      dplyr::filter(
        component == component_value,
        diet == diet_value
      ) |>
      dplyr::pull(proportion)

    if (length(selected) != 1L) {
      stop("Component proportions are not one-to-one.", call. = FALSE)
    }
    selected[[1]]
  }

  p8_con <- proportion_value("unit8", REFERENCE_DIET)
  p8_hfd <- proportion_value("unit8", COMPARISON_DIET)
  p6_con <- proportion_value("unit6", REFERENCE_DIET)
  p6_hfd <- proportion_value("unit6", COMPARISON_DIET)

  output <- means_wide |>
    dplyr::mutate(
      unit8_proportion_CON = p8_con,
      unit8_proportion_HFD = p8_hfd,
      unit6_proportion_CON = p6_con,
      unit6_proportion_HFD = p6_hfd,
      unit8_midpoint_proportion = (p8_con + p8_hfd) / 2,
      unit6_midpoint_proportion = (p6_con + p6_hfd) / 2,
      observed_CON =
        p8_con * unit8_CON + p6_con * unit6_CON,
      observed_HFD =
        p8_hfd * unit8_HFD + p6_hfd * unit6_HFD,
      observed_difference = observed_HFD - observed_CON,
      observed_log2FC = log2(
        (observed_HFD + pseudocount) /
          (observed_CON + pseudocount)
      ),
      standardized_CON =
        unit8_midpoint_proportion * unit8_CON +
        unit6_midpoint_proportion * unit6_CON,
      standardized_HFD =
        unit8_midpoint_proportion * unit8_HFD +
        unit6_midpoint_proportion * unit6_HFD,
      standardized_difference =
        standardized_HFD - standardized_CON,
      standardized_log2FC = log2(
        (standardized_HFD + pseudocount) /
          (standardized_CON + pseudocount)
      ),
      within_state_component =
        unit8_midpoint_proportion * (unit8_HFD - unit8_CON) +
        unit6_midpoint_proportion * (unit6_HFD - unit6_CON),
      composition_component =
        ((unit8_HFD + unit8_CON) / 2) * (p8_hfd - p8_con) +
        ((unit6_HFD + unit6_CON) / 2) * (p6_hfd - p6_con),
      decomposition_residual =
        observed_difference -
        within_state_component -
        composition_component,
      absolute_within_fraction = dplyr::if_else(
        abs(within_state_component) + abs(composition_component) > 0,
        abs(within_state_component) /
          (abs(within_state_component) + abs(composition_component)),
        NA_real_
      ),
      unit8_log2FC = log2(
        (unit8_HFD + pseudocount) /
          (unit8_CON + pseudocount)
      ),
      unit6_log2FC = log2(
        (unit6_HFD + pseudocount) /
          (unit6_CON + pseudocount)
      )
    )

  if (isTRUE(scale_to_percentage_points)) {
    output <- output |>
      dplyr::mutate(
        dplyr::across(
          c(
            observed_CON,
            observed_HFD,
            observed_difference,
            standardized_CON,
            standardized_HFD,
            standardized_difference,
            within_state_component,
            composition_component,
            decomposition_residual,
            unit8_CON,
            unit8_HFD,
            unit6_CON,
            unit6_HFD
          ),
          ~ 100 * .x
        )
      )
  }

  output
}

prepare_effect_wide <- function(effect_table, feature_name = "feature") {
  effect_table |>
    dplyr::filter(
      comparison_population %in%
        c("broad_stem", "strict_ISC", "crypt_progenitor")
    ) |>
    dplyr::mutate(
      effect_population = dplyr::recode(
        comparison_population,
        broad_stem = "broad",
        strict_ISC = "unit8",
        crypt_progenitor = "unit6"
      )
    ) |>
    dplyr::select(
      dplyr::all_of(feature_name),
      effect_population,
      log2FC_comparison_vs_reference,
      detection_difference_percentage_points,
      reference_percent_detected,
      comparison_percent_detected
    ) |>
    tidyr::pivot_wider(
      names_from = effect_population,
      values_from = c(
        log2FC_comparison_vs_reference,
        detection_difference_percentage_points,
        reference_percent_detected,
        comparison_percent_detected
      ),
      names_glue = "{effect_population}_{.value}"
    )
}

prepare_rna_stability_wide <- function(stability_table) {
  stability_table |>
    dplyr::filter(
      comparison_population %in%
        c("broad_stem", "strict_ISC", "crypt_progenitor")
    ) |>
    dplyr::mutate(
      effect_population = dplyr::recode(
        comparison_population,
        broad_stem = "broad",
        strict_ISC = "unit8",
        crypt_progenitor = "unit6"
      ),
      direction_stable =
        sign(full_log2FC) == sign(downsampled_log2FC_median) &
        sign_concordance_percent >=
          MINIMUM_SIGN_CONCORDANCE_PERCENT,
      downsampling_iqr_excludes_zero = dplyr::case_when(
        full_log2FC > 0 ~ downsampled_log2FC_q25 > 0,
        full_log2FC < 0 ~ downsampled_log2FC_q75 < 0,
        TRUE ~ FALSE
      )
    ) |>
    dplyr::select(
      feature,
      effect_population,
      sign_concordance_percent,
      direction_stable,
      downsampling_iqr_excludes_zero
    ) |>
    tidyr::pivot_wider(
      names_from = effect_population,
      values_from = c(
        sign_concordance_percent,
        direction_stable,
        downsampling_iqr_excludes_zero
      ),
      names_glue = "{effect_population}_{.value}"
    )
}

classify_component_response <- function(
  broad_clear,
  unit8_clear,
  unit6_clear,
  broad_effect,
  unit8_effect,
  unit6_effect,
  within_component,
  composition_component
) {
  dplyr::case_when(
    unit8_clear & unit6_clear &
      sign(unit8_effect) == sign(unit6_effect) ~
      "shared_within_state",
    unit8_clear & unit6_clear &
      sign(unit8_effect) != sign(unit6_effect) ~
      "opposing_state_effects",
    unit6_clear & !unit8_clear ~ "unit6_specific",
    unit8_clear & !unit6_clear ~ "unit8_specific",
    broad_clear &
      abs(composition_component) > abs(within_component) ~
      "composition_dominant",
    broad_clear ~ "broad_mixed_within_state",
    TRUE ~ "weak_or_unresolved"
  )
}

run_binary_detection_stability <- function(
  binary_matrix,
  metadata,
  component_unit,
  features,
  iterations,
  random_seed
) {
  features <- intersect(features, rownames(binary_matrix))

  if (length(features) == 0L) {
    return(tibble::tibble())
  }

  reference_cells <- metadata$cell_barcode[
    metadata$analysis_unit == component_unit &
      metadata$diet == REFERENCE_DIET
  ]
  comparison_cells <- metadata$cell_barcode[
    metadata$analysis_unit == component_unit &
      metadata$diet == COMPARISON_DIET
  ]
  sampled_nuclei <- min(
    length(reference_cells),
    length(comparison_cells)
  )

  selected_matrix <- binary_matrix[features, , drop = FALSE]
  full_difference <- 100 * (
    Matrix::rowMeans(
      selected_matrix[, comparison_cells, drop = FALSE]
    ) -
      Matrix::rowMeans(
        selected_matrix[, reference_cells, drop = FALSE]
      )
  )

  set.seed(random_seed)
  sampled_differences <- matrix(
    NA_real_,
    nrow = length(features),
    ncol = iterations,
    dimnames = list(features, NULL)
  )

  for (iteration_index in seq_len(iterations)) {
    sampled_reference <- sample(
      reference_cells,
      size = sampled_nuclei,
      replace = FALSE
    )
    sampled_comparison <- sample(
      comparison_cells,
      size = sampled_nuclei,
      replace = FALSE
    )

    sampled_differences[, iteration_index] <- 100 * (
      Matrix::rowMeans(
        selected_matrix[, sampled_comparison, drop = FALSE]
      ) -
        Matrix::rowMeans(
          selected_matrix[, sampled_reference, drop = FALSE]
        )
    )
  }

  quantiles <- t(apply(
    sampled_differences,
    1,
    stats::quantile,
    probs = c(0.25, 0.50, 0.75),
    na.rm = TRUE,
    names = FALSE
  ))

  sign_concordance <- vapply(
    seq_along(features),
    function(feature_index) {
      if (full_difference[[feature_index]] == 0) return(NA_real_)
      100 * mean(
        sign(sampled_differences[feature_index, ]) ==
          sign(full_difference[[feature_index]]),
        na.rm = TRUE
      )
    },
    numeric(1)
  )

  tibble::tibble(
    feature = features,
    component_unit = component_unit,
    sampled_nuclei_per_diet = sampled_nuclei,
    full_detection_difference_percentage_points =
      as.numeric(full_difference),
    downsampled_detection_difference_q25 = quantiles[, 1],
    downsampled_detection_difference_median = quantiles[, 2],
    downsampled_detection_difference_q75 = quantiles[, 3],
    sign_concordance_percent = sign_concordance,
    direction_stable =
      sign_concordance >= MINIMUM_SIGN_CONCORDANCE_PERCENT &
      sign(full_difference) == sign(quantiles[, 2]),
    downsampling_iqr_excludes_zero = dplyr::case_when(
      full_difference > 0 ~ quantiles[, 1] > 0,
      full_difference < 0 ~ quantiles[, 3] < 0,
      TRUE ~ FALSE
    )
  )
}
```

# 4. Load and validate the WT object

```{r load-and-validate-wt-object}
multiome <- readRDS(STEP6_OBJECT_PATH)
metadata <- multiome[[]]

analysis_unit_column <- if (
  "annotation_unit" %in% colnames(metadata)
) {
  "annotation_unit"
} else if ("wnn_cluster" %in% colnames(metadata)) {
  "wnn_cluster"
} else {
  stop(
    "The Step 6 object contains neither annotation_unit nor wnn_cluster.",
    call. = FALSE
  )
}

required_metadata <- c(
  "sample_id",
  "diet",
  analysis_unit_column,
  "broad_compartment",
  "cell_state",
  "annotation_confidence"
)
missing_metadata <- setdiff(required_metadata, colnames(metadata))

if (length(missing_metadata) > 0L) {
  stop(
    "The Step 6 object is missing metadata column(s): ",
    paste(missing_metadata, collapse = ", "),
    call. = FALSE
  )
}

metadata$cell_barcode <- rownames(metadata)
metadata$sample_id <- as.character(metadata$sample_id)
metadata$diet <- as.character(metadata$diet)
metadata$analysis_unit <- as.character(
  metadata[[analysis_unit_column]]
)
metadata$broad_compartment <- as.character(
  metadata$broad_compartment
)
metadata$cell_state <- as.character(metadata$cell_state)
metadata$annotation_confidence <- tolower(
  as.character(metadata$annotation_confidence)
)

if (!setequal(unique(metadata$diet), c(REFERENCE_DIET, COMPARISON_DIET))) {
  stop("Expected exactly CON and HFD in the WT object.", call. = FALSE)
}

unit_annotation_validation <- metadata |>
  dplyr::filter(analysis_unit %in% BROAD_STEM_UNITS) |>
  dplyr::distinct(
    analysis_unit,
    broad_compartment,
    cell_state,
    annotation_confidence
  ) |>
  dplyr::mutate(
    expected_cell_state = dplyr::recode(
      analysis_unit,
      `8` = "ISC-like crypt",
      `6` = "Crypt progenitor"
    ),
    validation_pass =
      broad_compartment == "epithelial" &
      cell_state == expected_cell_state &
      annotation_confidence %in% c("high", "moderate")
  )

if (
  nrow(unit_annotation_validation) != length(BROAD_STEM_UNITS) ||
    any(!unit_annotation_validation$validation_pass)
) {
  stop(
    "The approved WT unit-8/unit-6 annotations no longer match Step 8C.",
    call. = FALSE
  )
}

if (!all(c("RNA", "ATAC") %in% names(multiome))) {
  stop("The WT object must contain RNA and ATAC assays.", call. = FALSE)
}

rna_counts <- SeuratObject::LayerData(
  multiome[["RNA"]],
  layer = "counts"
)
atac_counts <- SeuratObject::LayerData(
  multiome[["ATAC"]],
  layer = "counts"
)

metadata_cells <- rownames(metadata)
matrix_alignment_audit <- tibble::tibble(
  matrix = c("RNA counts", "ATAC counts"),
  matrix_cells = c(ncol(rna_counts), ncol(atac_counts)),
  metadata_cells = length(metadata_cells),
  same_cell_set = c(
    setequal(colnames(rna_counts), metadata_cells),
    setequal(colnames(atac_counts), metadata_cells)
  ),
  identical_order_before_alignment = c(
    identical(colnames(rna_counts), metadata_cells),
    identical(colnames(atac_counts), metadata_cells)
  )
)

rna_counts <- align_matrix_to_metadata(
  rna_counts,
  metadata_cells,
  "RNA counts"
)
atac_counts <- align_matrix_to_metadata(
  atac_counts,
  metadata_cells,
  "ATAC counts"
)

readr::write_csv(
  matrix_alignment_audit,
  file.path(STEP8C_DIR, "matrix_cell_alignment_audit.csv")
)
readr::write_csv(
  unit_annotation_validation,
  file.path(STEP8C_DIR, "step8c_unit_annotation_validation.csv")
)

rm(multiome)
invisible(gc())
```

# 5. Quantify the broad-stem mixture

```{r quantify-component-composition}
component_composition <- metadata |>
  dplyr::filter(analysis_unit %in% BROAD_STEM_UNITS) |>
  dplyr::count(diet, analysis_unit, name = "nuclei") |>
  dplyr::group_by(diet) |>
  dplyr::mutate(
    broad_stem_nuclei = sum(nuclei),
    proportion = nuclei / broad_stem_nuclei,
    percent = 100 * proportion,
    component = dplyr::recode(
      analysis_unit,
      `8` = "unit8",
      `6` = "unit6"
    ),
    component_label = dplyr::recode(
      analysis_unit,
      `8` = "Unit 8: ISC-like crypt",
      `6` = "Unit 6: Crypt progenitor"
    )
  ) |>
  dplyr::ungroup()

composition_totals <- component_composition |>
  dplyr::group_by(diet) |>
  dplyr::summarise(total = sum(proportion), .groups = "drop") |>
  dplyr::pull(total)

if (
  nrow(component_composition) != 4L ||
    any(abs(composition_totals - 1) > 1e-12)
) {
  stop("Broad-stem component proportions failed validation.", call. = FALSE)
}

readr::write_csv(
  component_composition,
  file.path(
    STEP8C_TABLE_DIR,
    "broad_stem_component_counts_and_proportions.csv"
  )
)

p_composition <- ggplot2::ggplot(
  component_composition,
  ggplot2::aes(
    x = diet,
    y = percent,
    fill = component_label
  )
) +
  ggplot2::geom_col(width = 0.72) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = paste0(nuclei, "\n", sprintf("%.1f%%", percent))
    ),
    position = ggplot2::position_stack(vjust = 0.5),
    size = 4
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "Unit 8: ISC-like crypt" = "green3",
      "Unit 6: Crypt progenitor" = "yellow3"
    )
  ) +
  ggplot2::labs(
    title = "WT broad-stem composition differs strongly by pooled diet library",
    subtitle = "This redistribution can contribute to an observed composite diet effect",
    x = NULL,
    y = "Percent of broad-stem nuclei",
    fill = NULL,
    caption = "Descriptive: one pooled CON and one pooled HFD library."
  ) +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.major.x = ggplot2::element_blank()
  )

save_native_pdf(
  "01_broad_stem_component_composition.pdf",
  p_composition,
  width = 9,
  height = 7
)
print(p_composition)
component_composition
```

# 6. Decompose RNA effects on a linear scale

Per-cell RNA counts are divided by each nucleus's total RNA count and scaled to 10,000 before component means are calculated. This is a linear abundance scale; log-normalized Seurat values are deliberately not averaged for the additive decomposition.

```{r calculate-rna-decomposition}
rna_cell_totals <- Matrix::colSums(rna_counts)

if (any(!is.finite(rna_cell_totals)) || any(rna_cell_totals <= 0)) {
  stop("RNA counts contain a cell with a non-positive library size.")
}

rna_linear <- rna_counts %*% Matrix::Diagonal(
  x = RNA_SCALE_FACTOR / rna_cell_totals
)
colnames(rna_linear) <- colnames(rna_counts)
rownames(rna_linear) <- rownames(rna_counts)

rna_component_means <- summarize_matrix_by_component(
  feature_matrix = rna_linear,
  metadata = metadata,
  value_label = "mean_per_cell_counts_per_10000",
  binary = FALSE
)

rna_decomposition <- midpoint_decomposition(
  component_means = rna_component_means,
  component_proportions = component_composition,
  pseudocount = RNA_LINEAR_PSEUDOCOUNT,
  scale_to_percentage_points = FALSE
)

maximum_rna_residual <- max(
  abs(rna_decomposition$decomposition_residual),
  na.rm = TRUE
)

if (maximum_rna_residual > 1e-8) {
  stop(
    "RNA midpoint decomposition residual exceeded tolerance: ",
    signif(maximum_rna_residual, 4),
    call. = FALSE
  )
}

step8b_rna_effects <- read_table_quietly(STEP8B_RNA_EFFECT_PATH)
step8b_rna_stability <- read_table_quietly(
  STEP8B_RNA_STABILITY_PATH
)

rna_effect_wide <- prepare_effect_wide(step8b_rna_effects)
rna_stability_wide <- prepare_rna_stability_wide(
  step8b_rna_stability
)

rna_decomposition <- rna_decomposition |>
  dplyr::left_join(rna_effect_wide, by = "feature") |>
  dplyr::left_join(rna_stability_wide, by = "feature") |>
  dplyr::mutate(
    broad_effect_available = !is.na(
      broad_log2FC_comparison_vs_reference
    ),
    unit8_effect_available = !is.na(
      unit8_log2FC_comparison_vs_reference
    ),
    unit6_effect_available = !is.na(
      unit6_log2FC_comparison_vs_reference
    ),
    broad_clear = broad_effect_available &
      abs(broad_log2FC_comparison_vs_reference) >=
        RNA_CLEAR_LOG2FC &
      abs(broad_detection_difference_percentage_points) >=
        RNA_CLEAR_DETECTION_DELTA,
    unit8_clear = unit8_effect_available &
      abs(unit8_log2FC_comparison_vs_reference) >=
        RNA_CLEAR_LOG2FC &
      abs(unit8_detection_difference_percentage_points) >=
        RNA_CLEAR_DETECTION_DELTA,
    unit6_clear = unit6_effect_available &
      abs(unit6_log2FC_comparison_vs_reference) >=
        RNA_CLEAR_LOG2FC &
      abs(unit6_detection_difference_percentage_points) >=
        RNA_CLEAR_DETECTION_DELTA,
    RNA_response_class = classify_component_response(
      broad_clear = broad_clear,
      unit8_clear = unit8_clear,
      unit6_clear = unit6_clear,
      broad_effect = broad_log2FC_comparison_vs_reference,
      unit8_effect = unit8_log2FC_comparison_vs_reference,
      unit6_effect = unit6_log2FC_comparison_vs_reference,
      within_component = within_state_component,
      composition_component = composition_component
    ),
    pooled_pseudobulk_direction = effect_direction(
      broad_log2FC_comparison_vs_reference
    ),
    observed_linear_direction = effect_direction(observed_log2FC),
    standardized_direction = effect_direction(
      standardized_log2FC
    ),
    direction_changed_after_composition_standardization =
      dplyr::case_when(
        is.na(observed_log2FC) | is.na(standardized_log2FC) ~ NA,
        sign(observed_log2FC) !=
          sign(standardized_log2FC) ~ TRUE,
        TRUE ~ FALSE
      ),
    pooled_and_linear_observed_direction_agree = dplyr::case_when(
      !broad_effect_available | is.na(observed_log2FC) ~ NA,
      sign(broad_log2FC_comparison_vs_reference) ==
        sign(observed_log2FC) ~ TRUE,
      TRUE ~ FALSE
    )
  )

readr::write_csv(
  rna_component_means,
  file.path(
    STEP8C_TABLE_DIR,
    "RNA_linear_component_means.csv.gz"
  )
)
readr::write_csv(
  rna_decomposition,
  file.path(
    STEP8C_TABLE_DIR,
    "RNA_broad_stem_composition_decomposition.csv.gz"
  )
)

rm(rna_linear, rna_counts, rna_cell_totals)
invisible(gc())
```

# 7. Reprioritize persistent RNA genes

```{r prioritize-persistent-rna-genes}
reference_genes <- read_table_quietly(REFERENCE_GENE_PATH)
assert_columns(
  reference_genes,
  c(
    "gene",
    "reference_direction",
    "reference_program",
    "source_rank",
    "avg_log2FC_day21",
    "avg_log2FC_day50"
  ),
  "persistent-HFD reference table"
)

persistent_rna_decomposition <- reference_genes |>
  dplyr::left_join(
    rna_decomposition,
    by = c("gene" = "feature")
  ) |>
  dplyr::mutate(
    pooled_source_aligned_log2FC = source_aligned_effect(
      broad_log2FC_comparison_vs_reference,
      reference_direction
    ),
    observed_linear_source_aligned_log2FC = source_aligned_effect(
      observed_log2FC,
      reference_direction
    ),
    standardized_source_aligned_log2FC = source_aligned_effect(
      standardized_log2FC,
      reference_direction
    ),
    unit8_source_aligned_log2FC = source_aligned_effect(
      unit8_log2FC_comparison_vs_reference,
      reference_direction
    ),
    unit6_source_aligned_log2FC = source_aligned_effect(
      unit6_log2FC_comparison_vs_reference,
      reference_direction
    ),
    pooled_source_concordant = dplyr::case_when(
      !broad_effect_available ~ NA,
      pooled_source_aligned_log2FC > 0 ~ TRUE,
      pooled_source_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    observed_linear_source_concordant = dplyr::case_when(
      is.na(observed_log2FC) ~ NA,
      observed_linear_source_aligned_log2FC > 0 ~ TRUE,
      observed_linear_source_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    standardized_source_concordant = dplyr::case_when(
      is.na(standardized_log2FC) ~ NA,
      standardized_source_aligned_log2FC > 0 ~ TRUE,
      standardized_source_aligned_log2FC < 0 ~ FALSE,
      TRUE ~ NA
    ),
    RNA_candidate_priority = dplyr::case_when(
      standardized_source_concordant %in% TRUE &
        RNA_response_class == "shared_within_state" ~
        "A_source_concordant_shared_within_state",
      standardized_source_concordant %in% TRUE &
        RNA_response_class == "unit6_specific" ~
        "B_source_concordant_unit6_specific",
      standardized_source_concordant %in% TRUE &
        RNA_response_class == "unit8_specific" ~
        "C_source_concordant_unit8_specific",
      observed_linear_source_concordant %in% TRUE &
        RNA_response_class == "composition_dominant" ~
        "D_observed_concordance_composition_dominant",
      standardized_source_concordant %in% TRUE ~
        "E_source_concordant_mixed_or_subthreshold",
      observed_linear_source_concordant %in% FALSE ~
        "F_source_discordant",
      TRUE ~ "G_not_evaluable"
    )
  ) |>
  dplyr::arrange(
    RNA_candidate_priority,
    dplyr::desc(abs(standardized_source_aligned_log2FC)),
    reference_program,
    source_rank
  )

readr::write_csv(
  persistent_rna_decomposition,
  file.path(
    STEP8C_TABLE_DIR,
    "persistent_gene_RNA_composition_decomposition.csv"
  )
)

rna_mechanism_summary <- persistent_rna_decomposition |>
  dplyr::count(
    reference_program,
    RNA_response_class,
    standardized_source_concordant,
    name = "genes"
  ) |>
  dplyr::arrange(
    reference_program,
    dplyr::desc(genes),
    RNA_response_class
  )

readr::write_csv(
  rna_mechanism_summary,
  file.path(
    STEP8C_TABLE_DIR,
    "persistent_gene_RNA_mechanism_summary.csv"
  )
)

rna_mechanism_summary
```

```{r plot-rna-decomposition}
rna_class_colors <- c(
  shared_within_state = "#009E73",
  unit6_specific = "#E69F00",
  unit8_specific = "#0072B2",
  opposing_state_effects = "#CC79A7",
  composition_dominant = "#D55E00",
  broad_mixed_within_state = "#56B4E9",
  weak_or_unresolved = "#9CA3AF"
)

plot_persistent <- persistent_rna_decomposition |>
  dplyr::filter(
    is.finite(observed_log2FC),
    is.finite(standardized_log2FC)
  )

label_genes <- plot_persistent |>
  dplyr::filter(
    RNA_candidate_priority %in% c(
      "A_source_concordant_shared_within_state",
      "B_source_concordant_unit6_specific",
      "C_source_concordant_unit8_specific",
      "D_observed_concordance_composition_dominant"
    )
  ) |>
  dplyr::slice_max(
    order_by = abs(
      observed_log2FC - standardized_log2FC
    ) + abs(standardized_log2FC),
    n = 14,
    with_ties = FALSE
  )

p_observed_standardized <- ggplot2::ggplot(
  plot_persistent,
  ggplot2::aes(
    x = observed_log2FC,
    y = standardized_log2FC,
    color = RNA_response_class
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75") +
  ggplot2::geom_vline(xintercept = 0, color = "grey75") +
  ggplot2::geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey45"
  ) +
  ggplot2::geom_point(alpha = 0.78, size = 2.1) +
  ggplot2::geom_text(
    data = label_genes,
    ggplot2::aes(label = gene),
    size = 3,
    check_overlap = TRUE,
    vjust = -0.5,
    show.legend = FALSE
  ) +
  ggplot2::scale_color_manual(values = rna_class_colors) +
  ggplot2::facet_wrap(~ reference_program) +
  ggplot2::labs(
    title = "Observed and composition-standardized persistent-gene RNA effects",
    subtitle = "Departure from the diagonal indicates sensitivity to the unit-8/unit-6 mixture",
    x = "Observed cell-weighted broad-stem log2FC (HFD / CON)",
    y = "Midpoint-composition-standardized log2FC",
    color = "Response class",
    caption = "Standardization uses fixed midpoint component weights; effects remain pooled-library descriptive."
  ) +
  ggplot2::theme(legend.position = "bottom")

print(p_observed_standardized)
save_native_pdf(
  "02_persistent_RNA_observed_vs_standardized.pdf",
  p_observed_standardized,
  width = 13,
  height = 8
)

p_component_rna <- ggplot2::ggplot(
  plot_persistent |>
    dplyr::filter(
      is.finite(unit8_log2FC_comparison_vs_reference),
      is.finite(unit6_log2FC_comparison_vs_reference)
    ),
  ggplot2::aes(
    x = unit8_log2FC_comparison_vs_reference,
    y = unit6_log2FC_comparison_vs_reference,
    color = RNA_response_class
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75") +
  ggplot2::geom_vline(xintercept = 0, color = "grey75") +
  ggplot2::geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey45"
  ) +
  ggplot2::geom_point(alpha = 0.8, size = 2.1) +
  ggplot2::scale_color_manual(values = rna_class_colors) +
  ggplot2::facet_wrap(~ reference_program) +
  ggplot2::labs(
    title = "Persistent-gene diet effects differ between stem components",
    x = "Unit 8 pooled RNA log2FC",
    y = "Unit 6 pooled RNA log2FC",
    color = "Response class",
    caption = "Axes are HFD versus CON within each annotated component."
  ) +
  ggplot2::theme(legend.position = "bottom")

print(p_component_rna)
save_native_pdf(
  "03_persistent_RNA_unit8_vs_unit6.pdf",
  p_component_rna,
  width = 13,
  height = 8
)
```

# 8. Decompose ATAC accessibility and create stable peak sets

ATAC is decomposed using peak detection probability: the fraction of nuclei with at least one fragment in a peak. This does not imply that every detected fragment is activating. `HFD_opening` and `CON_opening` below describe only the direction of accessibility.

```{r calculate-atac-decomposition}
atac_binary <- atac_counts
atac_binary@x[] <- 1

atac_component_means <- summarize_matrix_by_component(
  feature_matrix = atac_binary,
  metadata = metadata,
  value_label = "fraction_nuclei_peak_detected",
  binary = TRUE
)

atac_decomposition <- midpoint_decomposition(
  component_means = atac_component_means,
  component_proportions = component_composition,
  pseudocount = 1e-4,
  scale_to_percentage_points = TRUE
)

maximum_atac_residual <- max(
  abs(atac_decomposition$decomposition_residual),
  na.rm = TRUE
)

if (maximum_atac_residual > 1e-8) {
  stop(
    "ATAC midpoint decomposition residual exceeded tolerance: ",
    signif(maximum_atac_residual, 4),
    call. = FALSE
  )
}

step8b_atac_effects <- read_table_quietly(STEP8B_ATAC_EFFECT_PATH)
atac_effect_wide <- prepare_effect_wide(step8b_atac_effects)
atac_context <- read_table_quietly(STEP8B_ATAC_CONTEXT_PATH)

atac_decomposition <- atac_decomposition |>
  dplyr::left_join(atac_effect_wide, by = "feature") |>
  dplyr::left_join(
    atac_context,
    by = c("feature" = "peak")
  ) |>
  dplyr::mutate(
    peak_width = peak_end - peak_start + 1L,
    broad_effect_available = !is.na(
      broad_log2FC_comparison_vs_reference
    ),
    unit8_effect_available = !is.na(
      unit8_log2FC_comparison_vs_reference
    ),
    unit6_effect_available = !is.na(
      unit6_log2FC_comparison_vs_reference
    ),
    broad_clear = broad_effect_available &
      abs(broad_log2FC_comparison_vs_reference) >=
        ATAC_CLEAR_LOG2FC &
      abs(broad_detection_difference_percentage_points) >=
        ATAC_CLEAR_DETECTION_DELTA,
    unit8_clear = unit8_effect_available &
      abs(unit8_log2FC_comparison_vs_reference) >=
        ATAC_CLEAR_LOG2FC &
      abs(unit8_detection_difference_percentage_points) >=
        ATAC_CLEAR_DETECTION_DELTA,
    unit6_clear = unit6_effect_available &
      abs(unit6_log2FC_comparison_vs_reference) >=
        ATAC_CLEAR_LOG2FC &
      abs(unit6_detection_difference_percentage_points) >=
        ATAC_CLEAR_DETECTION_DELTA,
    ATAC_response_class = classify_component_response(
      broad_clear = broad_clear,
      unit8_clear = unit8_clear,
      unit6_clear = unit6_clear,
      broad_effect = broad_log2FC_comparison_vs_reference,
      unit8_effect = unit8_log2FC_comparison_vs_reference,
      unit6_effect = unit6_log2FC_comparison_vs_reference,
      within_component = within_state_component,
      composition_component = composition_component
    )
  )

candidate_atac_features <- atac_decomposition |>
  dplyr::filter(
    unit8_clear | unit6_clear | broad_clear |
      abs(standardized_difference) >=
        ATAC_CLEAR_DETECTION_DELTA
  ) |>
  dplyr::pull(feature) |>
  unique()

unit8_atac_stability <- run_binary_detection_stability(
  binary_matrix = atac_binary,
  metadata = metadata,
  component_unit = UNIT8,
  features = candidate_atac_features,
  iterations = ATAC_DOWNSAMPLING_ITERATIONS,
  random_seed = 20260919
) |>
  dplyr::rename_with(
    ~ paste0("unit8_", .x),
    -feature
  )

unit6_atac_stability <- run_binary_detection_stability(
  binary_matrix = atac_binary,
  metadata = metadata,
  component_unit = UNIT6,
  features = candidate_atac_features,
  iterations = ATAC_DOWNSAMPLING_ITERATIONS,
  random_seed = 20260920
) |>
  dplyr::rename_with(
    ~ paste0("unit6_", .x),
    -feature
  )

atac_stability <- dplyr::full_join(
  unit8_atac_stability,
  unit6_atac_stability,
  by = "feature"
)

atac_decomposition <- atac_decomposition |>
  dplyr::left_join(atac_stability, by = "feature") |>
  dplyr::mutate(
    unit8_direction_stable = dplyr::coalesce(
      unit8_direction_stable,
      FALSE
    ),
    unit6_direction_stable = dplyr::coalesce(
      unit6_direction_stable,
      FALSE
    ),
    unit8_downsampling_iqr_excludes_zero = dplyr::coalesce(
      unit8_downsampling_iqr_excludes_zero,
      FALSE
    ),
    unit6_downsampling_iqr_excludes_zero = dplyr::coalesce(
      unit6_downsampling_iqr_excludes_zero,
      FALSE
    ),
    motif_peak_set = dplyr::case_when(
      ATAC_response_class == "shared_within_state" &
        unit8_direction_stable & unit6_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit6_downsampling_iqr_excludes_zero &
        unit8_log2FC_comparison_vs_reference > 0 ~
        "shared_HFD_opening",
      ATAC_response_class == "shared_within_state" &
        unit8_direction_stable & unit6_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit6_downsampling_iqr_excludes_zero &
        unit8_log2FC_comparison_vs_reference < 0 ~
        "shared_CON_opening",
      ATAC_response_class == "unit6_specific" &
        unit6_direction_stable &
        unit6_downsampling_iqr_excludes_zero &
        unit6_log2FC_comparison_vs_reference > 0 ~
        "unit6_HFD_opening",
      ATAC_response_class == "unit6_specific" &
        unit6_direction_stable &
        unit6_downsampling_iqr_excludes_zero &
        unit6_log2FC_comparison_vs_reference < 0 ~
        "unit6_CON_opening",
      ATAC_response_class == "unit8_specific" &
        unit8_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit8_log2FC_comparison_vs_reference > 0 ~
        "unit8_HFD_opening",
      ATAC_response_class == "unit8_specific" &
        unit8_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit8_log2FC_comparison_vs_reference < 0 ~
        "unit8_CON_opening",
      ATAC_response_class == "opposing_state_effects" &
        unit8_direction_stable & unit6_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit6_downsampling_iqr_excludes_zero &
        unit6_log2FC_comparison_vs_reference > 0 ~
        "opposing_unit6_HFD_unit8_CON",
      ATAC_response_class == "opposing_state_effects" &
        unit8_direction_stable & unit6_direction_stable &
        unit8_downsampling_iqr_excludes_zero &
        unit6_downsampling_iqr_excludes_zero &
        unit6_log2FC_comparison_vs_reference < 0 ~
        "opposing_unit6_CON_unit8_HFD",
      TRUE ~ NA_character_
    ),
    motif_ready = !is.na(motif_peak_set)
  )

readr::write_csv(
  atac_component_means,
  file.path(
    STEP8C_TABLE_DIR,
    "ATAC_detection_component_means.csv.gz"
  )
)
readr::write_csv(
  atac_stability,
  file.path(
    STEP8C_TABLE_DIR,
    "ATAC_binary_detection_downsampling_stability.csv.gz"
  )
)
readr::write_csv(
  atac_decomposition,
  file.path(
    STEP8C_TABLE_DIR,
    "ATAC_broad_stem_composition_decomposition.csv.gz"
  )
)

motif_ready_peak_sets <- atac_decomposition |>
  dplyr::filter(motif_ready) |>
  dplyr::select(
    peak = feature,
    motif_peak_set,
    ATAC_response_class,
    chromosome,
    peak_start,
    peak_end,
    peak_width,
    genomic_context,
    nearest_gene,
    nearest_gene_distance_bp,
    dplyr::everything()
  )

motif_peak_set_summary <- motif_ready_peak_sets |>
  dplyr::count(
    motif_peak_set,
    ATAC_response_class,
    genomic_context,
    name = "peaks"
  ) |>
  dplyr::group_by(motif_peak_set) |>
  dplyr::mutate(
    total_peak_set_size = sum(peaks),
    meets_minimum_motif_set_size =
      total_peak_set_size >= MINIMUM_MOTIF_SET_SIZE
  ) |>
  dplyr::ungroup()

motif_background_universe <- atac_decomposition |>
  dplyr::filter(unit8_effect_available | unit6_effect_available) |>
  dplyr::transmute(
    peak = feature,
    chromosome,
    peak_start,
    peak_end,
    peak_width,
    genomic_context,
    nearest_gene,
    nearest_gene_distance_bp,
    testable_unit8 = unit8_effect_available,
    testable_unit6 = unit6_effect_available,
    mean_detection_percent = rowMeans(
      cbind(unit8_CON, unit8_HFD, unit6_CON, unit6_HFD),
      na.rm = TRUE
    ),
    broad_log2FC = broad_log2FC_comparison_vs_reference,
    unit8_log2FC = unit8_log2FC_comparison_vs_reference,
    unit6_log2FC = unit6_log2FC_comparison_vs_reference
  )

readr::write_csv(
  motif_ready_peak_sets,
  file.path(STEP8C_TABLE_DIR, "motif_ready_peak_sets.csv.gz")
)
readr::write_csv(
  motif_peak_set_summary,
  file.path(STEP8C_TABLE_DIR, "motif_peak_set_summary.csv")
)
readr::write_csv(
  motif_background_universe,
  file.path(STEP8C_TABLE_DIR, "motif_background_universe.csv.gz")
)

rm(atac_binary, atac_counts)
invisible(gc())

motif_peak_set_summary
```

```{r plot-atac-component-effects}
atac_plot_data <- atac_decomposition |>
  dplyr::filter(
    unit8_effect_available,
    unit6_effect_available,
    unit8_clear | unit6_clear
  )

p_atac_components <- ggplot2::ggplot(
  atac_plot_data,
  ggplot2::aes(
    x = unit8_log2FC_comparison_vs_reference,
    y = unit6_log2FC_comparison_vs_reference,
    color = ATAC_response_class
  )
) +
  ggplot2::geom_hline(yintercept = 0, color = "grey75") +
  ggplot2::geom_vline(xintercept = 0, color = "grey75") +
  ggplot2::geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey45"
  ) +
  ggplot2::geom_point(size = 0.55, alpha = 0.45) +
  ggplot2::scale_color_manual(values = rna_class_colors) +
  ggplot2::coord_cartesian(xlim = c(-4, 4), ylim = c(-4, 4)) +
  ggplot2::labs(
    title = "WT broad-stem ATAC effects are strongly component resolved",
    subtitle = "Only peaks clear in unit 8 or unit 6 are shown",
    x = "Unit 8 pooled ATAC log2FC",
    y = "Unit 6 pooled ATAC log2FC",
    color = "Response class",
    caption = "Motif-ready sets additionally require stable binary-detection direction under cell downsampling."
  ) +
  ggplot2::theme(legend.position = "bottom")

print(p_atac_components)
save_native_pdf(
  "04_ATAC_unit8_vs_unit6_response_classes.pdf",
  p_atac_components,
  width = 12,
  height = 8
)

p_motif_inventory <- motif_peak_set_summary |>
  dplyr::group_by(motif_peak_set) |>
  dplyr::summarise(
    peaks = max(total_peak_set_size),
    meets_minimum_motif_set_size =
      all(meets_minimum_motif_set_size),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = reorder(motif_peak_set, peaks),
      y = peaks,
      fill = meets_minimum_motif_set_size
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::coord_flip() +
  ggplot2::scale_fill_manual(
    values = c(`TRUE` = "#009E73", `FALSE` = "#999999")
  ) +
  ggplot2::labs(
    title = "Stable component-resolved ATAC peak sets prepared for motif review",
    x = NULL,
    y = "Peaks",
    fill = paste0("At least ", MINIMUM_MOTIF_SET_SIZE, " peaks")
  ) +
  ggplot2::theme(legend.position = "bottom")

print(p_motif_inventory)
save_native_pdf(
  "05_motif_ready_peak_set_inventory.pdf",
  p_motif_inventory,
  width = 11,
  height = 7
)
```

# 9. Integrate persistent RNA with exact promoter accessibility

```{r integrate-exact-promoter-evidence}
step8b_promoter_evidence <- read_table_quietly(
  STEP8B_PROMOTER_EVIDENCE_PATH
)

promoter_geometry_columns <- c(
  "gene",
  "gene_id",
  "chromosome",
  "strand",
  "tss",
  "promoter_start",
  "promoter_end",
  "peak",
  "peak_start",
  "peak_end",
  "peak_center",
  "transcriptional_offset_bp",
  "peak_overlaps_tss"
)
assert_columns(
  step8b_promoter_evidence,
  promoter_geometry_columns,
  "Step 8B promoter evidence"
)

promoter_geometry <- step8b_promoter_evidence |>
  dplyr::select(dplyr::all_of(promoter_geometry_columns)) |>
  dplyr::distinct(gene, peak, .keep_all = TRUE)

promoter_decomposition <- promoter_geometry |>
  dplyr::left_join(
    persistent_rna_decomposition,
    by = "gene"
  ) |>
  dplyr::left_join(
    atac_decomposition |>
      dplyr::rename(peak = feature),
    by = "peak",
    suffix = c("_gene", "_peak")
  ) |>
  dplyr::mutate(
    standardized_RNA_ATAC_direction_match = dplyr::case_when(
      is.na(standardized_log2FC_gene) |
        is.na(within_state_component_peak) ~ NA,
      sign(standardized_log2FC_gene) ==
        sign(within_state_component_peak) ~ TRUE,
      TRUE ~ FALSE
    ),
    source_aligned_promoter_effect = source_aligned_effect(
      within_state_component_peak,
      reference_direction
    ),
    promoter_candidate_priority = dplyr::case_when(
      standardized_source_concordant %in% TRUE &
        standardized_RNA_ATAC_direction_match %in% TRUE &
        motif_ready &
        ATAC_response_class %in% c(
          "shared_within_state",
          "unit6_specific",
          "unit8_specific"
        ) ~
        "Tier_1_source_concordant_RNA_plus_stable_component_promoter_ATAC",
      standardized_source_concordant %in% TRUE &
        standardized_RNA_ATAC_direction_match %in% TRUE &
        ATAC_response_class %in% c(
          "shared_within_state",
          "unit6_specific",
          "unit8_specific"
        ) ~
        "Tier_2_source_concordant_RNA_plus_component_promoter_ATAC",
      standardized_source_concordant %in% TRUE &
        standardized_RNA_ATAC_direction_match %in% FALSE ~
        "Tier_3_source_concordant_RNA_opposed_promoter_ATAC",
      observed_linear_source_concordant %in% TRUE &
        RNA_response_class == "composition_dominant" ~
        "Tier_4_composition_sensitive_RNA",
      TRUE ~ "Not_prioritized"
    )
  )

representative_promoter_candidates <- promoter_decomposition |>
  dplyr::filter(promoter_candidate_priority != "Not_prioritized") |>
  dplyr::group_by(gene) |>
  dplyr::arrange(
    promoter_candidate_priority,
    dplyr::desc(abs(within_state_component_peak)),
    abs(transcriptional_offset_bp),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = 1L) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    promoter_candidate_priority,
    RNA_candidate_priority,
    dplyr::desc(abs(standardized_source_aligned_log2FC))
  )

readr::write_csv(
  promoter_decomposition,
  file.path(
    STEP8C_TABLE_DIR,
    "persistent_gene_exact_promoter_decomposition.csv.gz"
  )
)
readr::write_csv(
  representative_promoter_candidates,
  file.path(
    STEP8C_TABLE_DIR,
    "persistent_gene_prioritized_promoter_candidates.csv"
  )
)

representative_promoter_candidates |>
  dplyr::select(
    gene,
    reference_program,
    RNA_candidate_priority,
    RNA_response_class,
    standardized_log2FC_gene,
    peak,
    ATAC_response_class,
    motif_peak_set,
    within_state_component_peak,
    standardized_RNA_ATAC_direction_match,
    promoter_candidate_priority
  )
```

```{r plot-prioritized-promoter-candidates}
promoter_plot_data <- representative_promoter_candidates |>
  dplyr::slice_head(n = 30L) |>
  dplyr::mutate(
    gene = factor(gene, levels = rev(unique(gene)))
  )

if (nrow(promoter_plot_data) > 0L) {
  p_promoter <- ggplot2::ggplot(
    promoter_plot_data,
    ggplot2::aes(
      x = standardized_log2FC_gene,
      y = gene,
      color = promoter_candidate_priority,
      size = abs(within_state_component_peak)
    )
  ) +
    ggplot2::geom_vline(xintercept = 0, color = "grey75") +
    ggplot2::geom_point(alpha = 0.85) +
    ggplot2::facet_wrap(~ reference_program, scales = "free_y") +
    ggplot2::labs(
      title = "Persistent-gene candidates after composition standardization",
      subtitle = "Point size is the exact-promoter within-state ATAC detection change",
      x = "Composition-standardized RNA log2FC",
      y = NULL,
      color = "Integrated priority",
      size = "Absolute promoter\nchange (percentage points)",
      caption = "Exact promoter overlap is regulatory context, not proof of causality."
    ) +
    ggplot2::theme(legend.position = "bottom")

  print(p_promoter)
  save_native_pdf(
    "06_prioritized_persistent_gene_promoter_candidates.pdf",
    p_promoter,
    width = 13,
    height = 9
  )
}
```

# 10. Write completion summaries

```{r write-completion-summary}
classification_summary <- dplyr::bind_rows(
  rna_mechanism_summary |>
    dplyr::transmute(
      feature_type = "RNA_persistent_gene",
      group = reference_program,
      classification = RNA_response_class,
      direction_or_concordance = as.character(
        standardized_source_concordant
      ),
      features = genes
    ),
  motif_peak_set_summary |>
    dplyr::transmute(
      feature_type = "ATAC_peak",
      group = motif_peak_set,
      classification = ATAC_response_class,
      direction_or_concordance = genomic_context,
      features = peaks
    )
)

readr::write_csv(
  classification_summary,
  file.path(STEP8C_DIR, "step8c_classification_summary.csv")
)

step8c_summary <- tibble::tibble(
  dataset_version = DATASET_VERSION,
  cohort_id = COHORT_ID,
  broad_stem_units = paste(BROAD_STEM_UNITS, collapse = ";"),
  broad_stem_CON_nuclei = sum(
    component_composition$nuclei[
      component_composition$diet == REFERENCE_DIET
    ]
  ),
  broad_stem_HFD_nuclei = sum(
    component_composition$nuclei[
      component_composition$diet == COMPARISON_DIET
    ]
  ),
  persistent_genes = nrow(reference_genes),
  persistent_genes_source_concordant_after_standardization = sum(
    persistent_rna_decomposition$standardized_source_concordant %in% TRUE,
    na.rm = TRUE
  ),
  motif_ready_peak_sets = dplyr::n_distinct(
    motif_ready_peak_sets$motif_peak_set
  ),
  motif_ready_peaks = nrow(motif_ready_peak_sets),
  prioritized_exact_promoter_genes = dplyr::n_distinct(
    representative_promoter_candidates$gene
  ),
  motif_enrichment_performed = FALSE,
  distal_linking_performed = FALSE,
  cross_genotype_comparison_performed = FALSE,
  replicated_diet_design = FALSE,
  biological_inference_performed = FALSE
)

readr::write_csv(
  step8c_summary,
  file.path(STEP8C_DIR, "step8c_summary.csv")
)

writeLines(
  c(
    paste("Step 8C completed:", Sys.time()),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste("Broad-stem units:", paste(BROAD_STEM_UNITS, collapse = ",")),
    paste(
      "Maximum RNA decomposition residual:",
      format(maximum_rna_residual, scientific = TRUE)
    ),
    paste(
      "Maximum ATAC decomposition residual:",
      format(maximum_atac_residual, scientific = TRUE)
    ),
    paste(
      "Motif-ready peak sets:",
      dplyr::n_distinct(motif_ready_peak_sets$motif_peak_set)
    ),
    paste("Motif-ready peaks:", nrow(motif_ready_peak_sets)),
    paste(
      "Prioritized exact-promoter genes:",
      dplyr::n_distinct(representative_promoter_candidates$gene)
    ),
    "Midpoint composition decomposition performed: TRUE",
    "Component-resolved ATAC downsampling performed: TRUE",
    "Motif enrichment performed: FALSE",
    "Distal peak-to-gene linking performed: FALSE",
    "Cross-genotype comparison performed: FALSE",
    "Replicated diet inference performed: FALSE"
  ),
  con = file.path(STEP8C_DIR, "STEP_08C_COMPLETE.txt")
)

step8c_summary
```

# Interpretation checkpoint

Before motif enrichment:

1.  confirm the unit-8 and unit-6 mixture and the exact decomposition residuals;
2.  inspect genes whose observed and standardized directions differ;
3.  distinguish shared, unit-6-specific, unit-8-specific, opposing, and composition-dominant RNA candidates;
4.  confirm that each proposed motif peak set has adequate size and stable binary-detection direction;
5.  inspect exact-promoter candidates, especially cases in which the RNA and promoter directions oppose; and
6.  approve only biologically interpretable peak sets in the Step 8D decision table.

The observed broad-stem result remains useful: it captures the combined effect of state redistribution and molecular change. The standardized and component-resolved results answer the narrower regulatory-mechanism question.
