---
title: "Step 8D - Enrich motifs in reviewed WT broad-stem regulatory peak sets"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

# The goal of this notebook

Step 8C separates the broad-stem diet response into shared, unit-6-specific,
unit-8-specific, opposing, and composition-dominant components. It also
creates stable ATAC peak sets without looking at DNA motifs. Step 8D tests
JASPAR 2024 CORE vertebrate motifs only in peak sets that have been manually
approved after Step 8C review.

For each approved target set, background peaks are drawn from the same
testable component universe and matched on:

- baseline accessibility;
- GC content;
- peak width; and
- promoter, gene-body, or intergenic context.

Motif enrichment is a sequence-composition result, not proof that the
corresponding transcription factor is expressed, active, or bound. A compact
TF-expression column is included as secondary context only.

## Statistical limitation

The target peak sets originate from one pooled CON and one pooled HFD WT
library. Motif P values quantify over-representation among sequences; they do
not create biological replication or test a mouse-level diet effect.

# 1. Setup

Run after Step 8C in a fresh WT v2 session:

```r
Sys.setenv(
  MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt"
)
```

Additional Bioconductor packages are required:

```r
BiocManager::install(
  c(
    "JASPAR2024",
    "TFBSTools",
    "motifmatchr",
    "Biostrings",
    "Rsamtools",
    "DBI",
    "RSQLite"
  ),
  ask = FALSE,
  update = FALSE
)
```

JASPAR2024 may populate its versioned cache the first time it is used.

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 11,
  fig.height = 7
)

REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else {
  stop("Open multiome_scRNA-ATAC.Rproj before running this notebook.")
}

source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))

motif_packages <- c(
  "JASPAR2024",
  "TFBSTools",
  "motifmatchr",
  "Biostrings",
  "Rsamtools",
  "DBI",
  "RSQLite"
)

check_r_version()
package_preflight(
  packages = unique(c(required_packages, motif_packages))
)
initialize_output_dirs()

suppressPackageStartupMessages({
  library(Matrix)
  library(GenomicRanges)
  library(IRanges)
  library(ggplot2)
  library(dplyr)
  library(readr)
  library(tidyr)
  library(tibble)
  library(Biostrings)
  library(Rsamtools)
  library(TFBSTools)
  library(motifmatchr)
  library(DBI)
  library(RSQLite)
  library(JASPAR2024)
})

ggplot2::theme_set(theme_multiome())
future::plan(future::sequential)
set.seed(20260921)

if (
  !identical(DATASET_VERSION, "v2_resequenced") ||
    !identical(COHORT_ID, "wt")
) {
  stop(
    "Step 8D is defined for the WT v2 benchmark.",
    call. = FALSE
  )
}
```

# 2. Define inputs, outputs, and matching rules

```{r define-inputs-and-parameters}
STEP8B_DIR <- file.path(
  OUTPUT_DIR,
  "08b_source_matched_persistent_hfd"
)
STEP8B_TABLE_DIR <- file.path(STEP8B_DIR, "tables")
STEP8C_DIR <- file.path(
  OUTPUT_DIR,
  "08c_broad_stem_decomposition"
)
STEP8C_TABLE_DIR <- file.path(STEP8C_DIR, "tables")

STEP8D_DIR <- file.path(
  OUTPUT_DIR,
  "08d_broad_stem_motif_enrichment"
)
STEP8D_TABLE_DIR <- file.path(STEP8D_DIR, "tables")
STEP8D_PLOT_DIR <- file.path(STEP8D_DIR, "plots")

dir.create(STEP8D_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8D_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(STEP8D_PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

STEP8C_COMPLETE_MARKER <- file.path(
  STEP8C_DIR,
  "STEP_08C_COMPLETE.txt"
)
MOTIF_PEAK_SET_PATH <- file.path(
  STEP8C_TABLE_DIR,
  "motif_ready_peak_sets.csv.gz"
)
MOTIF_BACKGROUND_PATH <- file.path(
  STEP8C_TABLE_DIR,
  "motif_background_universe.csv.gz"
)
MOTIF_PEAK_SET_SUMMARY_PATH <- file.path(
  STEP8C_TABLE_DIR,
  "motif_peak_set_summary.csv"
)
MOTIF_DECISION_PATH <- file.path(
  COHORT_CONFIG_DIR,
  "motif_peak_set_decision.csv"
)
STEP8B_RNA_EFFECT_PATH <- file.path(
  STEP8B_TABLE_DIR,
  "RNA_pooled_library_effects_by_matched_population.csv.gz"
)

GRCM39_FASTA <- Sys.getenv(
  "MULTIOME_GRCM39_FASTA",
  unset = file.path(
    PROJECT_DIR,
    "reference",
    "GRCm39_2024-A",
    "fasta",
    "genome.fa"
  )
)

required_inputs <- c(
  Step8C_marker = STEP8C_COMPLETE_MARKER,
  motif_peak_sets = MOTIF_PEAK_SET_PATH,
  motif_background = MOTIF_BACKGROUND_PATH,
  motif_peak_set_summary = MOTIF_PEAK_SET_SUMMARY_PATH,
  motif_decision = MOTIF_DECISION_PATH,
  Step8B_RNA_effects = STEP8B_RNA_EFFECT_PATH,
  GRCm39_FASTA = GRCM39_FASTA
)

for (input_label in names(required_inputs)) {
  assert_file(required_inputs[[input_label]], input_label)
}

if (!file.exists(paste0(GRCM39_FASTA, ".fai"))) {
  stop(
    "The GRCm39 FASTA index is missing: ",
    paste0(GRCM39_FASTA, ".fai"),
    "\nIndex the exact FASTA once with Rsamtools::indexFa() or samtools faidx.",
    call. = FALSE
  )
}

MINIMUM_TARGET_PEAKS <- 25L
BACKGROUND_RATIO <- 10L
MINIMUM_BACKGROUND_RATIO <- 3
GC_BINS <- 4L
WIDTH_BINS <- 3L
ACCESSIBILITY_BINS <- 4L
MOTIF_PVALUE_CUTOFF <- 5e-5
TOP_MOTIFS_PER_SET <- 15L
PREVIEW_PLOTS_IN_RSTUDIO <- FALSE

analysis_parameters <- tibble::tibble(
  parameter = c(
    "minimum target peaks",
    "target background ratio",
    "minimum accepted background ratio",
    "GC matching bins",
    "width matching bins",
    "accessibility matching bins",
    "motif-match p-value cutoff",
    "JASPAR release",
    "JASPAR collection",
    "JASPAR taxonomic group",
    "replicated diet design",
    "biological inference performed"
  ),
  value = c(
    MINIMUM_TARGET_PEAKS,
    BACKGROUND_RATIO,
    MINIMUM_BACKGROUND_RATIO,
    GC_BINS,
    WIDTH_BINS,
    ACCESSIBILITY_BINS,
    MOTIF_PVALUE_CUTOFF,
    "JASPAR2024",
    "CORE",
    "vertebrates",
    FALSE,
    FALSE
  )
)

readr::write_csv(
  analysis_parameters,
  file.path(STEP8D_DIR, "step8d_parameters.csv")
)
```

# 3. Helper functions

```{r helper-functions}
read_table_quietly <- function(path) {
  suppressMessages(
    readr::read_csv(
      path,
      show_col_types = FALSE,
      progress = FALSE
    )
  )
}

assert_columns <- function(data, required_columns, table_label) {
  missing_columns <- setdiff(required_columns, colnames(data))

  if (length(missing_columns) > 0L) {
    stop(
      table_label,
      " is missing required column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
}

preview_plot <- function(plot_object) {
  if (isTRUE(PREVIEW_PLOTS_IN_RSTUDIO)) print(plot_object)
  invisible(plot_object)
}

save_native_pdf <- function(
  filename,
  plot_object,
  width = 11,
  height = 7
) {
  output_path <- file.path(STEP8D_PLOT_DIR, filename)
  grDevices::pdf(
    file = output_path,
    width = width,
    height = height,
    family = "Helvetica",
    useDingbats = FALSE
  )
  on.exit(grDevices::dev.off(), add = TRUE)
  print(plot_object)
  invisible(output_path)
}

parse_approved <- function(values) {
  tolower(trimws(as.character(values))) %in%
    c("true", "t", "1", "yes", "y")
}

matching_scope <- function(peak_set_id) {
  dplyr::case_when(
    grepl("^unit6_", peak_set_id) ~ "unit6",
    grepl("^unit8_", peak_set_id) ~ "unit8",
    grepl("^shared_", peak_set_id) ~ "both",
    grepl("^opposing_", peak_set_id) ~ "both",
    TRUE ~ "unknown"
  )
}

matching_population <- function(peak_set_id) {
  dplyr::case_when(
    grepl("^unit6_", peak_set_id) ~ "crypt_progenitor",
    grepl("^unit8_", peak_set_id) ~ "strict_ISC",
    TRUE ~ "broad_stem"
  )
}

accessibility_direction <- function(peak_set_id) {
  dplyr::case_when(
    grepl("HFD_opening", peak_set_id) ~ "HFD_opening",
    grepl("CON_opening", peak_set_id) ~ "CON_opening",
    grepl("unit6_HFD_unit8_CON", peak_set_id) ~
      "opposing_unit6_HFD_unit8_CON",
    grepl("unit6_CON_unit8_HFD", peak_set_id) ~
      "opposing_unit6_CON_unit8_HFD",
    TRUE ~ "not_classified"
  )
}

safe_ntile <- function(x, bins) {
  if (length(unique(x[is.finite(x)])) < 2L) {
    return(rep.int(1L, length(x)))
  }
  dplyr::ntile(x, bins)
}

standardize_numeric <- function(x) {
  x_center <- mean(x, na.rm = TRUE)
  x_scale <- stats::sd(x, na.rm = TRUE)
  if (!is.finite(x_scale) || x_scale == 0) x_scale <- 1
  (x - x_center) / x_scale
}

match_background_for_set <- function(
  peak_set_id,
  target_peaks,
  universe,
  excluded_target_peaks,
  background_ratio,
  random_seed
) {
  scope <- matching_scope(peak_set_id)

  eligible_universe <- universe |>
    dplyr::filter(
      dplyr::case_when(
        scope == "unit6" ~ testable_unit6,
        scope == "unit8" ~ testable_unit8,
        scope == "both" ~ testable_unit6 & testable_unit8,
        TRUE ~ FALSE
      )
    ) |>
    dplyr::filter(!peak %in% excluded_target_peaks)

  if (nrow(eligible_universe) == 0L) {
    stop("No eligible background peaks for ", peak_set_id, ".")
  }

  combined <- dplyr::bind_rows(
    target_peaks |>
      dplyr::select(
        peak,
        gc_percent,
        peak_width,
        mean_detection_percent,
        genomic_context
      ) |>
      dplyr::mutate(record_role = "target"),
    eligible_universe |>
      dplyr::select(
        peak,
        gc_percent,
        peak_width,
        mean_detection_percent,
        genomic_context
      ) |>
      dplyr::mutate(record_role = "background_candidate")
  ) |>
    dplyr::mutate(
      gc_bin = safe_ntile(gc_percent, GC_BINS),
      width_bin = safe_ntile(peak_width, WIDTH_BINS),
      accessibility_bin = safe_ntile(
        mean_detection_percent,
        ACCESSIBILITY_BINS
      ),
      matching_stratum = paste(
        genomic_context,
        gc_bin,
        width_bin,
        accessibility_bin,
        sep = "|"
      )
    )

  target_binned <- combined |>
    dplyr::filter(record_role == "target")
  background_binned <- combined |>
    dplyr::filter(record_role == "background_candidate")

  requested_by_stratum <- target_binned |>
    dplyr::count(matching_stratum, name = "target_n") |>
    dplyr::mutate(requested_background_n = target_n * background_ratio)

  set.seed(random_seed)
  initially_matched <- dplyr::bind_rows(
    lapply(
      seq_len(nrow(requested_by_stratum)),
      function(stratum_index) {
        stratum_value <- requested_by_stratum$matching_stratum[[
          stratum_index
        ]]
        requested_n <- requested_by_stratum$requested_background_n[[
          stratum_index
        ]]
        available <- background_binned |>
          dplyr::filter(matching_stratum == stratum_value)

        if (nrow(available) == 0L) return(NULL)
        available |>
          dplyr::slice_sample(n = min(requested_n, nrow(available)))
      }
    )
  ) |>
    dplyr::distinct(peak, .keep_all = TRUE)

  desired_background_n <- min(
    background_ratio * nrow(target_binned),
    nrow(background_binned)
  )

  if (nrow(initially_matched) < desired_background_n) {
    target_centers <- target_binned |>
      dplyr::summarise(
        gc = mean(gc_percent, na.rm = TRUE),
        width = mean(log1p(peak_width), na.rm = TRUE),
        accessibility = mean(
          mean_detection_percent,
          na.rm = TRUE
        )
      )

    supplemental <- background_binned |>
      dplyr::filter(!peak %in% initially_matched$peak)

    gc_scaled <- standardize_numeric(
      c(supplemental$gc_percent, target_centers$gc)
    )
    width_scaled <- standardize_numeric(
      c(log1p(supplemental$peak_width), target_centers$width)
    )
    accessibility_scaled <- standardize_numeric(
      c(
        supplemental$mean_detection_percent,
        target_centers$accessibility
      )
    )

    target_index <- nrow(supplemental) + 1L
    supplemental$gc_z <- gc_scaled[-target_index]
    supplemental$width_z <- width_scaled[-target_index]
    supplemental$accessibility_z <- accessibility_scaled[-target_index]

    target_gc_z <- gc_scaled[[target_index]]
    target_width_z <- width_scaled[[target_index]]
    target_accessibility_z <- accessibility_scaled[[target_index]]

    target_contexts <- unique(target_binned$genomic_context)
    supplemental <- supplemental |>
      dplyr::mutate(
        matching_distance =
          (gc_z - target_gc_z)^2 +
          (width_z - target_width_z)^2 +
          (accessibility_z - target_accessibility_z)^2 +
          dplyr::if_else(
            genomic_context %in% target_contexts,
            0,
            4
          )
      ) |>
      dplyr::arrange(matching_distance) |>
      dplyr::slice_head(
        n = desired_background_n - nrow(initially_matched)
      )

    initially_matched <- dplyr::bind_rows(
      initially_matched,
      supplemental
    ) |>
      dplyr::distinct(peak, .keep_all = TRUE)
  }

  background_output <- initially_matched |>
    dplyr::slice_head(n = desired_background_n) |>
    dplyr::mutate(
      motif_peak_set = peak_set_id,
      membership = "background"
    )

  target_output <- target_binned |>
    dplyr::mutate(
      motif_peak_set = peak_set_id,
      membership = "target"
    )

  membership <- dplyr::bind_rows(
    target_output,
    background_output
  ) |>
    dplyr::select(
      motif_peak_set,
      peak,
      membership,
      gc_percent,
      peak_width,
      mean_detection_percent,
      genomic_context,
      matching_stratum
    )

  achieved_ratio <- nrow(background_output) / nrow(target_output)

  audit <- membership |>
    dplyr::group_by(motif_peak_set, membership) |>
    dplyr::summarise(
      peaks = dplyr::n(),
      GC_median = stats::median(gc_percent, na.rm = TRUE),
      width_median = stats::median(peak_width, na.rm = TRUE),
      accessibility_median = stats::median(
        mean_detection_percent,
        na.rm = TRUE
      ),
      promoter_percent = 100 * mean(
        genomic_context == "promoter_proximal"
      ),
      gene_body_percent = 100 * mean(
        genomic_context == "gene_body_nonpromoter"
      ),
      intergenic_percent = 100 * mean(
        genomic_context == "intergenic"
      ),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      requested_background_ratio = background_ratio,
      achieved_background_ratio = achieved_ratio,
      minimum_ratio_pass =
        achieved_ratio >= MINIMUM_BACKGROUND_RATIO
    )

  if (!all(audit$minimum_ratio_pass)) {
    stop(
      "Background matching for ", peak_set_id,
      " achieved only ", signif(achieved_ratio, 3),
      " background peaks per target peak.",
      call. = FALSE
    )
  }

  list(membership = membership, audit = audit)
}

mouse_symbol_from_motif_name <- function(motif_name) {
  simple_name <- grepl("^[A-Za-z][A-Za-z0-9]+$", motif_name)
  output <- rep(NA_character_, length(motif_name))
  output[simple_name] <- paste0(
    toupper(substr(motif_name[simple_name], 1, 1)),
    tolower(substr(motif_name[simple_name], 2, nchar(motif_name[simple_name])))
  )
  output
}
```

# 4. Validate the manual motif decision

Every approved row must have a reviewer and review date. The decision file is
committed with all rows unapproved; approval should occur only after review of
`motif_peak_set_summary.csv` and the Step 8C plots.

```{r validate-motif-decision}
motif_peak_sets <- read_table_quietly(MOTIF_PEAK_SET_PATH)
motif_background <- read_table_quietly(MOTIF_BACKGROUND_PATH)
motif_peak_set_summary <- read_table_quietly(
  MOTIF_PEAK_SET_SUMMARY_PATH
)
motif_decision <- read_table_quietly(MOTIF_DECISION_PATH)

assert_columns(
  motif_peak_sets,
  c(
    "peak",
    "motif_peak_set",
    "chromosome",
    "peak_start",
    "peak_end",
    "peak_width",
    "genomic_context"
  ),
  "Step 8C motif peak sets"
)
assert_columns(
  motif_background,
  c(
    "peak",
    "chromosome",
    "peak_start",
    "peak_end",
    "peak_width",
    "genomic_context",
    "testable_unit8",
    "testable_unit6",
    "mean_detection_percent"
  ),
  "Step 8C motif background"
)
assert_columns(
  motif_decision,
  c(
    "cohort_id",
    "peak_set_id",
    "approved",
    "reviewer",
    "review_date",
    "decision_notes"
  ),
  "motif decision table"
)

if (anyDuplicated(motif_decision$peak_set_id) > 0L) {
  stop("Motif decision table contains duplicate peak_set_id values.")
}

motif_decision <- motif_decision |>
  dplyr::mutate(
    cohort_id = tolower(trimws(as.character(cohort_id))),
    peak_set_id = trimws(as.character(peak_set_id)),
    approved_logical = parse_approved(approved),
    reviewer = trimws(as.character(reviewer)),
    review_date = trimws(as.character(review_date)),
    decision_notes = trimws(as.character(decision_notes))
  )

if (any(motif_decision$cohort_id != COHORT_ID)) {
  stop("Motif decision table contains a non-WT cohort row.")
}

approved_decisions <- motif_decision |>
  dplyr::filter(approved_logical)

if (nrow(approved_decisions) == 0L) {
  stop(
    "No motif peak set has approved=TRUE. Review Step 8C, complete ",
    MOTIF_DECISION_PATH,
    ", and rerun Step 8D.",
    call. = FALSE
  )
}

if (
  anyNA(approved_decisions$reviewer) ||
    any(!nzchar(approved_decisions$reviewer)) ||
    anyNA(approved_decisions$review_date) ||
    any(!nzchar(approved_decisions$review_date))
) {
  stop(
    "Every approved motif peak set requires reviewer and review_date.",
    call. = FALSE
  )
}

available_peak_sets <- unique(motif_peak_sets$motif_peak_set)
missing_approved_sets <- setdiff(
  approved_decisions$peak_set_id,
  available_peak_sets
)

if (length(missing_approved_sets) > 0L) {
  stop(
    "Approved motif peak set(s) are absent from Step 8C: ",
    paste(missing_approved_sets, collapse = ", "),
    call. = FALSE
  )
}

approved_set_sizes <- motif_peak_sets |>
  dplyr::filter(
    motif_peak_set %in% approved_decisions$peak_set_id
  ) |>
  dplyr::count(motif_peak_set, name = "target_peaks")

if (any(approved_set_sizes$target_peaks < MINIMUM_TARGET_PEAKS)) {
  undersized <- approved_set_sizes$motif_peak_set[
    approved_set_sizes$target_peaks < MINIMUM_TARGET_PEAKS
  ]
  stop(
    "Approved motif set(s) contain fewer than ",
    MINIMUM_TARGET_PEAKS,
    " peaks: ",
    paste(undersized, collapse = ", "),
    call. = FALSE
  )
}

readr::write_csv(
  approved_decisions,
  file.path(STEP8D_DIR, "approved_motif_peak_sets_applied.csv")
)

approved_set_sizes
```

# 5. Retrieve exact GRCm39 sequences and calculate GC content

```{r retrieve-peak-sequences}
approved_target_peaks <- motif_peak_sets |>
  dplyr::filter(
    motif_peak_set %in% approved_decisions$peak_set_id
  ) |>
  dplyr::select(
    peak,
    motif_peak_set,
    chromosome,
    peak_start,
    peak_end,
    peak_width,
    genomic_context
  ) |>
  dplyr::distinct()

all_sequence_candidates <- dplyr::bind_rows(
  approved_target_peaks |>
    dplyr::select(
      peak,
      chromosome,
      peak_start,
      peak_end,
      peak_width,
      genomic_context
    ),
  motif_background |>
    dplyr::select(
      peak,
      chromosome,
      peak_start,
      peak_end,
      peak_width,
      genomic_context
    )
) |>
  dplyr::distinct(peak, .keep_all = TRUE)

fasta_file <- Rsamtools::FaFile(GRCM39_FASTA)
fasta_index <- Rsamtools::scanFaIndex(fasta_file)
fasta_seqnames <- as.character(GenomeInfoDb::seqnames(fasta_index))
fasta_lengths <- GenomicRanges::width(fasta_index)
names(fasta_lengths) <- fasta_seqnames

sequence_qc <- all_sequence_candidates |>
  dplyr::mutate(
    chromosome_present = chromosome %in% fasta_seqnames,
    coordinate_valid = chromosome_present &
      peak_start >= 1L &
      peak_end <= fasta_lengths[chromosome] &
      peak_end >= peak_start,
    exclusion_reason = dplyr::case_when(
      !chromosome_present ~ "chromosome_absent_from_FASTA",
      !coordinate_valid ~ "invalid_or_out_of_bounds_coordinate",
      TRUE ~ NA_character_
    )
  )

readr::write_csv(
  sequence_qc,
  file.path(STEP8D_TABLE_DIR, "motif_sequence_qc.csv.gz")
)

invalid_target_peaks <- approved_target_peaks |>
  dplyr::anti_join(
    sequence_qc |>
      dplyr::filter(coordinate_valid) |>
      dplyr::select(peak),
    by = "peak"
  )

if (nrow(invalid_target_peaks) > 0L) {
  stop(
    "At least one approved target peak cannot be retrieved from the exact ",
    "GRCm39 FASTA. Review motif_sequence_qc.csv before proceeding.",
    call. = FALSE
  )
}

valid_sequence_candidates <- sequence_qc |>
  dplyr::filter(coordinate_valid)

peak_ranges <- GenomicRanges::GRanges(
  seqnames = valid_sequence_candidates$chromosome,
  ranges = IRanges::IRanges(
    start = valid_sequence_candidates$peak_start,
    end = valid_sequence_candidates$peak_end
  )
)
names(peak_ranges) <- valid_sequence_candidates$peak

peak_sequences <- Rsamtools::scanFa(
  fasta_file,
  param = peak_ranges
)
names(peak_sequences) <- names(peak_ranges)
names(peak_sequences) <- names(peak_ranges)

gc_percent <- 100 * rowSums(
  Biostrings::letterFrequency(
    peak_sequences,
    letters = c("G", "C"),
    as.prob = TRUE
  )
)

sequence_metrics <- valid_sequence_candidates |>
  dplyr::mutate(
    sequence_width = width(peak_sequences),
    gc_percent = as.numeric(gc_percent),
    sequence_width_matches_peak = sequence_width == peak_width
  )

if (any(!sequence_metrics$sequence_width_matches_peak)) {
  stop("Retrieved FASTA sequence width does not match a peak width.")
}

readr::write_csv(
  sequence_metrics,
  file.path(STEP8D_TABLE_DIR, "motif_sequence_metrics.csv.gz")
)
```

# 6. Match background peaks

```{r match-background-peaks}
background_with_gc <- motif_background |>
  dplyr::inner_join(
    sequence_metrics |>
      dplyr::select(peak, gc_percent),
    by = "peak"
  )

targets_with_gc <- approved_target_peaks |>
  dplyr::inner_join(
    motif_background |>
      dplyr::select(peak, mean_detection_percent),
    by = "peak"
  ) |>
  dplyr::inner_join(
    sequence_metrics |>
      dplyr::select(peak, gc_percent),
    by = "peak"
  )

missing_target_accessibility <- setdiff(
  approved_target_peaks$peak,
  targets_with_gc$peak
)

if (length(missing_target_accessibility) > 0L) {
  stop(
    "Approved target peaks are missing from the background universe: ",
    paste(head(missing_target_accessibility, 10), collapse = ", "),
    call. = FALSE
  )
}

matched_results <- lapply(
  seq_len(nrow(approved_decisions)),
  function(decision_index) {
    peak_set_id <- approved_decisions$peak_set_id[[decision_index]]
    target_set <- targets_with_gc |>
      dplyr::filter(motif_peak_set == peak_set_id)

    match_background_for_set(
      peak_set_id = peak_set_id,
      target_peaks = target_set,
      universe = background_with_gc,
      excluded_target_peaks = unique(targets_with_gc$peak),
      background_ratio = BACKGROUND_RATIO,
      random_seed = 20260921 + decision_index
    )
  }
)

motif_peak_membership <- dplyr::bind_rows(
  lapply(matched_results, `[[`, "membership")
)
background_matching_audit <- dplyr::bind_rows(
  lapply(matched_results, `[[`, "audit")
)

readr::write_csv(
  motif_peak_membership,
  file.path(STEP8D_TABLE_DIR, "motif_peak_membership.csv.gz")
)
readr::write_csv(
  background_matching_audit,
  file.path(STEP8D_TABLE_DIR, "motif_background_matching_audit.csv")
)

background_matching_audit
```

# 7. Retrieve JASPAR 2024 motifs and scan sequences

The CORE vertebrate collection is queried with `all_versions = FALSE` to
avoid treating multiple versions of the same JASPAR profile as independent
discoveries.

```{r retrieve-and-scan-jaspar-motifs}
jaspar_options <- list(
  collection = "CORE",
  tax_group = "vertebrates",
  all_versions = FALSE
)

load_jaspar_pfms <- function(options) {
  jaspar_database <- JASPAR2024::JASPAR2024()
  jaspar_connection <- DBI::dbConnect(
    RSQLite::SQLite(),
    db(jaspar_database)
  )
  on.exit(DBI::dbDisconnect(jaspar_connection), add = TRUE)

  TFBSTools::getMatrixSet(
    x = jaspar_connection,
    opts = options
  )
}

jaspar_pfms <- load_jaspar_pfms(jaspar_options)

if (length(jaspar_pfms) == 0L) {
  stop("JASPAR2024 returned no CORE vertebrate motifs.")
}

motif_metadata_base <- tibble::tibble(
  motif_id = vapply(
    jaspar_pfms,
    TFBSTools::ID,
    character(1)
  ),
  motif_name = vapply(
    jaspar_pfms,
    TFBSTools::name,
    character(1)
  ),
  motif_class = vapply(
    jaspar_pfms,
    function(x) paste(TFBSTools::tags(x)$class, collapse = ";"),
    character(1)
  ),
  motif_family = vapply(
    jaspar_pfms,
    function(x) paste(TFBSTools::tags(x)$family, collapse = ";"),
    character(1)
  )
) |>
  dplyr::mutate(
    candidate_mouse_gene = mouse_symbol_from_motif_name(motif_name)
  )

scan_peaks <- unique(motif_peak_membership$peak)
scan_sequences <- peak_sequences[scan_peaks]

if (anyNA(match(scan_peaks, names(peak_sequences)))) {
  stop("A matched target or background peak lacks a retrieved sequence.")
}

motif_match_object <- motifmatchr::matchMotifs(
  pwms = jaspar_pfms,
  subject = scan_sequences,
  bg = "subject",
  out = "matches",
  p.cutoff = MOTIF_PVALUE_CUTOFF
)

motif_match_matrix <- motifmatchr::motifMatches(
  motif_match_object
)
rownames(motif_match_matrix) <- scan_peaks

if (nrow(motif_match_matrix) != length(scan_peaks)) {
  stop("Motif match matrix does not align with scanned peaks.")
}

motif_matrix_columns <- colnames(motif_match_matrix)
motif_matrix_ids <- vapply(
  motif_matrix_columns,
  function(matrix_column) {
    matching_ids <- motif_metadata_base$motif_id[
      startsWith(matrix_column, motif_metadata_base$motif_id)
    ]
    if (length(matching_ids) == 0L) return(NA_character_)
    matching_ids[[which.max(nchar(matching_ids))]]
  },
  character(1)
)

motif_metadata <- tibble::tibble(
  motif_match_column = motif_matrix_columns,
  motif_id = motif_matrix_ids
) |>
  dplyr::left_join(motif_metadata_base, by = "motif_id")

if (anyNA(motif_metadata$motif_name)) {
  stop("JASPAR motif metadata and match matrix IDs do not agree.")
}

readr::write_csv(
  motif_metadata,
  file.path(STEP8D_TABLE_DIR, "JASPAR2024_CORE_motif_metadata.csv")
)
```

# 8. Test motif enrichment

```{r test-motif-enrichment}
test_one_peak_set <- function(peak_set_id) {
  membership <- motif_peak_membership |>
    dplyr::filter(motif_peak_set == peak_set_id)

  target_peaks <- membership$peak[membership$membership == "target"]
  background_peaks <- membership$peak[
    membership$membership == "background"
  ]

  target_matches <- motif_match_matrix[target_peaks, , drop = FALSE]
  background_matches <- motif_match_matrix[
    background_peaks,
    ,
    drop = FALSE
  ]

  target_with_motif <- Matrix::colSums(target_matches)
  background_with_motif <- Matrix::colSums(background_matches)
  target_without_motif <- length(target_peaks) - target_with_motif
  background_without_motif <-
    length(background_peaks) - background_with_motif

  p_values <- vapply(
    seq_along(target_with_motif),
    function(motif_index) {
      stats::fisher.test(
        matrix(
          c(
            target_with_motif[[motif_index]],
            target_without_motif[[motif_index]],
            background_with_motif[[motif_index]],
            background_without_motif[[motif_index]]
          ),
          nrow = 2,
          byrow = TRUE
        ),
        alternative = "greater"
      )$p.value
    },
    numeric(1)
  )

  tibble::tibble(
    motif_peak_set = peak_set_id,
    motif_match_column = colnames(motif_match_matrix),
    target_peaks = length(target_peaks),
    background_peaks = length(background_peaks),
    target_with_motif = as.integer(target_with_motif),
    background_with_motif = as.integer(background_with_motif),
    target_motif_percent =
      100 * target_with_motif / length(target_peaks),
    background_motif_percent =
      100 * background_with_motif / length(background_peaks),
    motif_percentage_point_difference =
      target_motif_percent - background_motif_percent,
    odds_ratio_Haldane =
      ((target_with_motif + 0.5) *
        (background_without_motif + 0.5)) /
      ((target_without_motif + 0.5) *
        (background_with_motif + 0.5)),
    p_value = p_values,
    p_adjust_BH = stats::p.adjust(p_values, method = "BH")
  ) |>
    dplyr::left_join(motif_metadata, by = "motif_match_column") |>
    dplyr::mutate(
      matching_scope = matching_scope(motif_peak_set),
      accessibility_direction = accessibility_direction(
        motif_peak_set
      )
    ) |>
    dplyr::arrange(p_adjust_BH, p_value, dplyr::desc(odds_ratio_Haldane))
}

motif_enrichment <- dplyr::bind_rows(
  lapply(approved_decisions$peak_set_id, test_one_peak_set)
)

step8b_rna_effects <- read_table_quietly(STEP8B_RNA_EFFECT_PATH) |>
  dplyr::mutate(feature_upper = toupper(feature))

motif_enrichment <- motif_enrichment |>
  dplyr::mutate(
    comparison_population = matching_population(motif_peak_set),
    candidate_mouse_gene_upper = toupper(candidate_mouse_gene)
  ) |>
  dplyr::left_join(
    step8b_rna_effects |>
      dplyr::select(
        comparison_population,
        feature_upper,
        TF_RNA_log2FC = log2FC_comparison_vs_reference,
        TF_RNA_detection_delta =
          detection_difference_percentage_points
      ),
    by = c(
      "comparison_population",
      "candidate_mouse_gene_upper" = "feature_upper"
    )
  ) |>
  dplyr::mutate(
    TF_RNA_context_available = !is.na(TF_RNA_log2FC),
    TF_RNA_direction_matches_peak_opening = dplyr::case_when(
      !TF_RNA_context_available ~ NA,
      accessibility_direction == "HFD_opening" &
        TF_RNA_log2FC > 0 ~ TRUE,
      accessibility_direction == "CON_opening" &
        TF_RNA_log2FC < 0 ~ TRUE,
      accessibility_direction %in% c(
        "HFD_opening",
        "CON_opening"
      ) ~ FALSE,
      TRUE ~ NA
    )
  )

top_motif_candidates <- motif_enrichment |>
  dplyr::filter(
    is.finite(p_adjust_BH),
    p_adjust_BH <= 0.10,
    odds_ratio_Haldane > 1
  ) |>
  dplyr::group_by(motif_peak_set) |>
  dplyr::slice_min(
    order_by = p_adjust_BH,
    n = TOP_MOTIFS_PER_SET,
    with_ties = FALSE
  ) |>
  dplyr::ungroup()

readr::write_csv(
  motif_enrichment,
  file.path(STEP8D_TABLE_DIR, "motif_enrichment_results.csv.gz")
)
readr::write_csv(
  top_motif_candidates,
  file.path(STEP8D_TABLE_DIR, "motif_top_candidates.csv")
)

top_motif_candidates |>
  dplyr::select(
    motif_peak_set,
    motif_id,
    motif_name,
    odds_ratio_Haldane,
    p_adjust_BH,
    target_motif_percent,
    background_motif_percent,
    candidate_mouse_gene,
    TF_RNA_log2FC,
    TF_RNA_direction_matches_peak_opening
  )
```

# 9. Generate review figures

```{r plot-background-audit}
p_background <- background_matching_audit |>
  dplyr::select(
    motif_peak_set,
    membership,
    GC_median,
    width_median,
    accessibility_median,
    promoter_percent,
    gene_body_percent,
    intergenic_percent
  ) |>
  tidyr::pivot_longer(
    cols = c(
      GC_median,
      width_median,
      accessibility_median,
      promoter_percent,
      gene_body_percent,
      intergenic_percent
    ),
    names_to = "matching_metric",
    values_to = "value"
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = membership,
      y = value,
      color = membership,
      group = motif_peak_set
    )
  ) +
  ggplot2::geom_line(alpha = 0.55) +
  ggplot2::geom_point(size = 2) +
  ggplot2::facet_grid(
    matching_metric ~ motif_peak_set,
    scales = "free_y"
  ) +
  ggplot2::scale_color_manual(
    values = c(target = "#D55E00", background = "#0072B2")
  ) +
  ggplot2::labs(
    title = "Target and matched-background peak characteristics",
    x = NULL,
    y = NULL,
    color = NULL
  ) +
  ggplot2::theme(
    legend.position = "bottom",
    axis.text.x = ggplot2::element_text(angle = 30, hjust = 1)
  )

preview_plot(p_background)
save_native_pdf(
  "01_motif_background_matching_audit.pdf",
  p_background,
  width = max(11, 3 * nrow(approved_decisions)),
  height = 12
)
```

```{r plot-top-motifs}
if (nrow(top_motif_candidates) > 0L) {
  plot_motifs <- top_motif_candidates |>
    dplyr::mutate(
      motif_label = paste0(motif_name, " [", motif_id, "]"),
      minus_log10_q = -log10(pmax(p_adjust_BH, 1e-300))
    )

  p_motifs <- ggplot2::ggplot(
    plot_motifs,
    ggplot2::aes(
      x = odds_ratio_Haldane,
      y = reorder(motif_label, odds_ratio_Haldane),
      size = minus_log10_q,
      color = motif_percentage_point_difference
    )
  ) +
    ggplot2::geom_vline(xintercept = 1, color = "grey70") +
    ggplot2::geom_point(alpha = 0.85) +
    ggplot2::facet_wrap(
      ~ motif_peak_set,
      scales = "free_y"
    ) +
    ggplot2::scale_color_gradient2(
      low = "#0072B2",
      mid = "grey85",
      high = "#D55E00",
      midpoint = 0
    ) +
    ggplot2::labs(
      title = "JASPAR 2024 motifs enriched in reviewed WT regulatory peak sets",
      subtitle = "Fisher tests use GC-, width-, accessibility-, and context-matched background peaks",
      x = "Motif odds ratio (Haldane corrected)",
      y = NULL,
      size = expression(-log[10](BH~q)),
      color = "Target minus background\npercentage points",
      caption = "Motif enrichment is sequence evidence; TF RNA is secondary context and binding is not established."
    ) +
    ggplot2::theme(legend.position = "bottom")

  preview_plot(p_motifs)
  save_native_pdf(
    "02_reviewed_peak_set_motif_enrichment.pdf",
    p_motifs,
    width = 14,
    height = max(8, 4 * ceiling(nrow(approved_decisions) / 2))
  )
}
```

# 10. Write completion summaries

```{r write-completion-summary}
motif_summary <- motif_enrichment |>
  dplyr::group_by(motif_peak_set) |>
  dplyr::summarise(
    target_peaks = dplyr::first(target_peaks),
    background_peaks = dplyr::first(background_peaks),
    motifs_tested = dplyr::n(),
    motifs_BH_q_0_05 = sum(p_adjust_BH <= 0.05, na.rm = TRUE),
    motifs_BH_q_0_10 = sum(p_adjust_BH <= 0.10, na.rm = TRUE),
    top_motif = motif_name[which.min(p_adjust_BH)],
    top_motif_q = min(p_adjust_BH, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    approved_decisions |>
      dplyr::select(
        peak_set_id,
        reviewer,
        review_date,
        decision_notes
      ),
    by = c("motif_peak_set" = "peak_set_id")
  ) |>
  dplyr::mutate(
    replicated_diet_design = FALSE,
    biological_inference_performed = FALSE
  )

readr::write_csv(
  motif_summary,
  file.path(STEP8D_DIR, "step8d_motif_summary.csv")
)

writeLines(
  c(
    paste("Step 8D completed:", Sys.time()),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohort:", COHORT_ID),
    paste(
      "Approved motif peak sets:",
      paste(approved_decisions$peak_set_id, collapse = ",")
    ),
    paste("JASPAR motifs tested:", length(jaspar_pfms)),
    paste("GRCm39 FASTA:", GRCM39_FASTA),
    "Matched-background motif enrichment performed: TRUE",
    "GC matching performed: TRUE",
    "Accessibility matching performed: TRUE",
    "Peak-width matching performed: TRUE",
    "Genomic-context matching performed: TRUE",
    "TF expression used as required evidence: FALSE",
    "Motif enrichment interpreted as TF binding: FALSE",
    "Replicated diet inference performed: FALSE"
  ),
  con = file.path(STEP8D_DIR, "STEP_08D_COMPLETE.txt")
)

motif_summary
```

# Interpretation checkpoint

Prioritize motifs that:

1. remain enriched against a well-matched background;
2. recur in biologically related peak sets or distinguish unit 6 from unit 8;
3. have a plausible TF expressed in the relevant component population; and
4. connect to source-concordant RNA plus promoter candidates without being
   treated as proof of TF binding.

Do not select a TF solely because its motif has the smallest P value. Similar
DNA-binding families often share motifs, and motif presence cannot identify
which family member occupies a locus. Candidate mechanisms should ultimately
be supported by TF expression, motif deviation/activity, footprinting where
coverage permits, distal peak-to-gene evidence, or orthogonal experiments.
