---
title: "Step 08E | Refined WT crypt diet benchmarks"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This release compares CON and HFD **within the refined WT focused r0.3
populations** identified in Step 05E and reviewed in 05F. C0 is the
ISC-enriched primary population and C2 the Dmbt1/Hmgcs2 progenitor-like
primary population. C1 and C4 are exploratory comparators; C3 and C5, plus
the historical global 8+6 grouping, are sensitivity contexts. Step 05E,
Step 06 and previous Steps 08–08D are unchanged.

One pooled CON library and one pooled HFD library means fold changes,
classification and resampling are **descriptive**. Repeated draws of nuclei
measure dependence on the included nuclei and depth; they are not biological
replicates and yield no sample-level P value or FDR. No claim of differential
expression, differential accessibility, fetal conversion, enhancer targeting
or causal genotype effect follows from this notebook.

Run after the completed 05F notebook from the project root in the WT R
session. Recommended SOL allocation: 8 CPUs, 96 GB, 8 hours.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
  fig.width = 11, fig.height = 7)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) {
  normalizePath(".")
} else if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj"))) {
  normalizePath("..")
} else stop("Open the project root before running 08E.", call. = FALSE)
source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
source(file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R"))
suppressPackageStartupMessages({
  library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges); library(IRanges)
})
need(identical(DATASET_VERSION, "v2_resequenced") && identical(COHORT_ID, "wt"),
  "08E requires the v2_resequenced WT run.")
set.seed(12345)
```

```{r inputs-and-working-map}
STEP5E <- file.path(OUTPUT_DIR, "05_wnn", "focused_wt_crypt_stem_progenitor")
ROOT_FOLLOWUP <- file.path(OUTPUT_DIR, "05_wnn", FOLLOWUP_RELEASE)
STEP5F <- file.path(ROOT_FOLLOWUP, "05f_review")
OBJECT <- file.path(STEP5E, "objects", "WT_crypt_core_0_6_8_focused.rds")
ANNOTATED <- file.path(OUTPUT_DIR, "06_annotation", "multiome_annotated.rds")
DECISION <- file.path(STEP5F, "refined_working_labels.csv")
COMPLETE <- file.path(STEP5F, "STEP_05F_COMPLETE.txt")
REFERENCE <- file.path(DATASET_CONFIG_DIR, "reference_programs",
  "persistent_hfd_genes.csv")
OUT <- file.path(OUTPUT_DIR, "08e_refined_wt_crypt_diet")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
for (path in c(OBJECT, ANNOTATED, DECISION, COMPLETE, REFERENCE))
  need(file.exists(path), paste("Required input absent:", path))

map <- read_csv_checked(DECISION) |>
  dplyr::mutate(
    focused_cluster = as.character(focused_cluster)
  )

require_columns(map, c("focused_cluster", "analysis_unit", "working_label",
  "analysis_role", "label_status", "approved_for_global_annotation",
  "focused_annotation_confidence"),
  "working decision")

need(setequal(as.character(map$focused_cluster), as.character(0:5)) &&
  !anyDuplicated(map$focused_cluster), "Expected exactly one working row for C0-C5.")
need(all(!map$approved_for_global_annotation) &&
  all(map$label_status == "provisional_working"),
  "The working map was modified into an annotation approval; review it separately.")
manifest <- capture_manifest(c(OBJECT, ANNOTATED, DECISION, COMPLETE, REFERENCE), OUT)
```

```{r align-focused-matrices}
focused <- load_focused(OBJECT)
obj <- focused$object
meta <- focused$metadata
rna <- focused$matrices$RNA_counts
atac <- focused$matrices$ATAC_counts
write_table(focused$alignment, OUT, "matrix_cell_alignment_audit.csv")
require_columns(meta, "annotation_confidence", "Step 05E metadata")
meta <- dplyr::left_join(
  dplyr::rename(meta, parent_annotation_confidence = annotation_confidence),
  map, by = "focused_cluster",
  relationship = "many-to-one")
need(!anyNA(meta$analysis_unit), "Some nuclei have no provisional working mapping.")
require_columns(meta,
  c("parent_annotation_confidence", "focused_annotation_confidence"),
  "joined 08E metadata")
reference <- read_csv_checked(REFERENCE) |>
  dplyr::mutate(gene_original = gene,
    gene = dplyr::if_else(gene == "Atpif1", "Atp5if1", gene))
require_columns(reference, c("gene", "reference_direction", "reference_program",
  "avg_log2FC_day21"), "persistent source")
need(!anyDuplicated(reference$gene), "Duplicate persistent genes after canonicalization.")
need(all(reference$reference_direction %in% c("HFD_higher", "CON_higher")),
  "Unrecognized source direction.")
write_table(reference, OUT, "persistent_reference_gene_canonicalization.csv")
```

```{r define-contrasts}
# Separate within-state effects from changes in cluster abundance. Preserve
# previous global 8+6 definition as a sensitivity, not as the new ISC label.
contrast <- tibble::tribble(
  ~contrast, ~population, ~member, ~role,
  "C0_ISC_enriched", "C0", "0", "primary",
  "C2_progenitor_like", "C2", "2", "primary",
  "C1_Htr4_Fut9", "C1", "1", "exploratory",
  "C3_unresolved", "C3", "3", "sensitivity",
  "C4_secretory", "C4", "4", "secretory_comparator",
  "C5_stem_associated", "C5", "5", "sensitivity",
  "C0_C3_C5_broad_crypt_sensitivity", "C0+C3+C5", "0;3;5", "sensitivity",
  "legacy_global_8_6", "global_8+6", "8;6", "historical_sensitivity"
)
contrast$n_CON <- integer(nrow(contrast)); contrast$n_HFD <- integer(nrow(contrast))
contrast$member_definition <- ifelse(contrast$contrast == "legacy_global_8_6",
  "parent_wnn_cluster", "focused_cluster")
contrast_indices <- vector("list", nrow(contrast))
for (i in seq_len(nrow(contrast))) {
  members <- strsplit(contrast$member[i], ";", fixed = TRUE)[[1]]
  key <- contrast$member_definition[i]
  ix <- which(as.character(meta[[key]]) %in% members)
  contrast_indices[[i]] <- ix
  contrast$n_CON[i] <- sum(meta$diet[ix] == "CON")
  contrast$n_HFD[i] <- sum(meta$diet[ix] == "HFD")
}
need(all(contrast$n_CON >= 30L & contrast$n_HFD >= 30L),
  "A requested contrast has fewer than 30 nuclei per diet.")
write_table(contrast, OUT, "refined_diet_contrast_definition_and_counts.csv")
```

```{r pooled-rna-atac-effects}
# Every effect is HFD minus CON. Evaluable filters follow Step 08B, but no
# individual-nucleus inferential P values are reported as biological evidence.
all_effects <- list()
for (modality in c("RNA", "ATAC")) {
  counts <- if (modality == "RNA") rna else atac
  effects <- dplyr::bind_rows(lapply(seq_len(nrow(contrast)), function(i) {
    contrast_id <- contrast$contrast[[i]]
contrast_role <- contrast$role[[i]]
message("Pooled ", modality, " effects: ", contrast_id)

effect <- pooled_effect(counts, meta, contrast_indices[[i]], modality)
effect$contrast <- contrast_id
effect$role <- contrast_role
dplyr::relocate(effect, contrast, role)
  }))
  filename <- paste0(modality, "_all_refined_pooled_effects.csv.gz")
  write_table(effects, OUT, filename)
  all_effects[[modality]] <- effects
  top <- effects |>
    dplyr::filter(evaluable) |>
    dplyr::group_by(contrast, direction) |>
    dplyr::arrange(dplyr::desc(abs(log2FC_HFD_vs_CON)), .by_group = TRUE) |>
    dplyr::slice_head(n = if (modality == "RNA") 250L else 500L) |>
    dplyr::ungroup()
  write_table(top, OUT, paste0(modality, "_top_descriptive_effects.csv"))
  rm(effects, top); invisible(gc())
}
```

```{r source-persistent-concordance}
persistent <- tidyr::crossing(contrast = contrast$contrast,
    feature = reference$gene) |>
  dplyr::left_join(reference, by = c("feature" = "gene"),
    relationship = "many-to-one") |>
  dplyr::left_join(all_effects$RNA |>
    dplyr::select(contrast, feature, evaluable, clear_effect,
      CON_counts, HFD_counts, CON_CPM, HFD_CPM,
      CON_percent, HFD_percent, detection_delta_pp,
      log2FC_HFD_vs_CON), by = c("contrast", "feature"),
    relationship = "many-to-one") |>
  dplyr::mutate(observation = dplyr::case_when(
    is.na(evaluable) ~ "gene_not_in_RNA_assay",
    !evaluable ~ "too_sparse_for_effect_threshold",
    !clear_effect ~ "direction_weak_or_detection_change_small",
    reference_direction == "HFD_higher" & log2FC_HFD_vs_CON > 0 ~ "concordant",
    reference_direction == "CON_higher" & log2FC_HFD_vs_CON < 0 ~ "concordant",
    TRUE ~ "discordant"),
    source_aligned_log2FC = dplyr::if_else(reference_direction == "HFD_higher",
      log2FC_HFD_vs_CON, -log2FC_HFD_vs_CON))
write_table(persistent, OUT, "persistent_genes_refined_WT_concordance.csv")
counts <- persistent |>
  dplyr::count(contrast, reference_program, observation, name = "genes")
write_table(counts, OUT, "persistent_gene_concordance_counts.csv")
```

```{r balanced-sensitivity}
# Select a bounded, fixed set before resampling; never select separately for
# each draw. This avoids full-genome recomputation across repeated subsets.
set.seed(12345)
anchors <- c("C0_ISC_enriched", "C2_progenitor_like", "C1_Htr4_Fut9",
  "C0_C3_C5_broad_crypt_sensitivity")
stability <- list()
for (modality in c("RNA", "ATAC")) {
  full <- all_effects[[modality]]
  matrix_data <- if (modality == "RNA") rna else atac
  bound <- if (modality == "RNA") 150L else 250L
  for (key in anchors) {
    i <- match(key, contrast$contrast)
    t <- full[full$contrast == key & full$evaluable, , drop = FALSE]
    t <- t[order(-abs(t$log2FC_HFD_vs_CON)), , drop = FALSE]
    top <- head(t$feature, bound)
    if (modality == "RNA") top <- union(top, reference$gene)
    message("Resampling ", modality, ": ", key, " (", length(top), " features)")
    x <- resampling_effects(matrix_data, meta, contrast_indices[[i]], t,
      top, modality, iterations = 20L)
    if (!nrow(x)) next
    stability[[length(stability) + 1L]] <- x |>
      dplyr::mutate(contrast = key, modality = modality,
        .before = 1)
  }
}
stability <- dplyr::bind_rows(stability)
write_table(stability, OUT, "equal_cell_and_depth_matched_resampling.csv.gz")
```

```{r phase-composition-sensitivity}
# Phase from Step 05E is RNA-derived. Common phase strata with >=10 nuclei
# per diet are assigned equal weights; no causal or biological variance claim.
phase_effect <- list(); phase_audit <- list()
for (key in c("C0_ISC_enriched", "C2_progenitor_like", "C1_Htr4_Fut9")) {
  i <- match(key, contrast$contrast)
  ranked <- all_effects$RNA |>
    dplyr::filter(contrast == key, evaluable) |>
    dplyr::arrange(dplyr::desc(abs(log2FC_HFD_vs_CON)))
  selected <- intersect(union(head(ranked$feature, 150L), reference$gene),
    rownames(rna))
  result <- standardize_strata(rna, meta, contrast_indices[[i]],
    meta$Phase, "RNA", selected, min_per_diet = 10L)
  phase_audit[[key]] <- result$audit |>
    dplyr::mutate(contrast = key, .before = 1)
  if (nrow(result$effects)) phase_effect[[key]] <- result$effects |>
    dplyr::mutate(contrast = key, .before = 1)
}
write_table(dplyr::bind_rows(phase_audit), OUT, "phase_stratum_eligibility.csv")
write_table(dplyr::bind_rows(phase_effect), OUT,
  "RNA_equal_phase_weight_effect_sensitivity.csv")
```

```{r promoter-and-rna-integration}
# One consensus peak can overlap several genes/TSSs, and one gene can have
# several TSSs or peaks. Each row below is a gene–peak hypothesis, not proof of
# regulation. Whole consensus peak counts, not clipped ±2 kb counts, enter
# ATAC effects. Missing overlapping peak is not evidence of closed chromatin.
peak_map <- promoter_peak_map(obj, window = 2000L)
map_all <- peak_map$map
interesting <- union(c(reference$gene,
    "Olfm4", "Clu", "Ly6a", "Anxa1", "Tacstd2"), #just checking for fetal signature, tom asked to see if there was any signature for it
  all_effects$RNA |>
    dplyr::filter(contrast %in% c("C0_ISC_enriched", "C2_progenitor_like"),
      clear_effect) |>
    dplyr::group_by(contrast) |>
    dplyr::slice_max(abs(log2FC_HFD_vs_CON), n = 250,
      with_ties = FALSE) |>
    dplyr::pull(feature))
map_target <- map_all[map_all$gene %in% interesting, , drop = FALSE]
write_table(map_target, OUT, "target_gene_promoter_overlapping_peak_map.csv.gz")
availability <- tibble::tibble(gene = interesting,
  RNA_feature_present = interesting %in% rownames(rna),
  gene_annotated = interesting %in% peak_map$annotated_genes,
  promoter_overlapping_peak_count = vapply(interesting,
    function(g) length(unique(map_target$peak[map_target$gene == g])), integer(1))) |>
  dplyr::mutate(ATAC_assessment = dplyr::case_when(
    !gene_annotated ~ "gene_not_in_annotation",
    promoter_overlapping_peak_count == 0 ~ "not_assayed_no_overlapping_peak",
    TRUE ~ "promoter_overlapping_peak_present"))
write_table(availability, OUT, "target_gene_promoter_assay_availability.csv")

rna_primary <- all_effects$RNA |>
  dplyr::filter(contrast %in% c("C0_ISC_enriched", "C2_progenitor_like"),
    feature %in% interesting) |>
  dplyr::select(contrast, feature, RNA_evaluable = evaluable,
    RNA_clear_effect = clear_effect,
    RNA_log2FC_HFD_vs_CON = log2FC_HFD_vs_CON,
    RNA_detection_delta_pp = detection_delta_pp,
    RNA_CON_percent = CON_percent, RNA_HFD_percent = HFD_percent)
atac_primary <- all_effects$ATAC |>
  dplyr::filter(contrast %in% c("C0_ISC_enriched", "C2_progenitor_like")) |>
  dplyr::select(contrast, peak = feature, ATAC_evaluable = evaluable,
    ATAC_clear_effect = clear_effect,
    ATAC_log2FC_HFD_vs_CON = log2FC_HFD_vs_CON,
    ATAC_CON_percent = CON_percent, ATAC_HFD_percent = HFD_percent)
paired <- dplyr::left_join(rna_primary, map_target,
  by = c("feature" = "gene"), relationship = "many-to-many") |>
  dplyr::left_join(atac_primary, by = c("contrast", "peak"),
    relationship = "many-to-one") |>
  dplyr::left_join(availability,
    by = c("feature" = "gene"), relationship = "many-to-one") |>
  dplyr::mutate(evidence_class = dplyr::case_when(
    ATAC_assessment == "not_assayed_no_overlapping_peak" ~ "promoter_peak_not_assayed",
    ATAC_assessment == "gene_not_in_annotation" ~ "no_gene_TSS_annotation",
    !RNA_evaluable | is.na(RNA_evaluable) ~ "RNA_not_evaluable",
    !ATAC_evaluable | is.na(ATAC_evaluable) ~ "ATAC_not_evaluable",
    !RNA_clear_effect | !ATAC_clear_effect ~ "one_or_both_effects_weak",
    sign(RNA_log2FC_HFD_vs_CON) == sign(ATAC_log2FC_HFD_vs_CON) ~
      "RNA_and_promoter_peak_same_direction",
    TRUE ~ "RNA_and_promoter_peak_opposite_direction"))
write_table(paired, OUT, "refined_C0_C2_RNA_promoter_peak_evidence.csv.gz")
```

```{r presentation-plots-and-finish}
plot_ref <- persistent |>
  dplyr::filter(contrast %in% c("C0_ISC_enriched", "C2_progenitor_like"),
    is.finite(avg_log2FC_day21), is.finite(log2FC_HFD_vs_CON)) |>
  ggplot2::ggplot(ggplot2::aes(x = avg_log2FC_day21,
    y = log2FC_HFD_vs_CON, color = observation)) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 0.3) +
  ggplot2::geom_point(size = 1.7, alpha = .8) +
  ggplot2::facet_wrap(~ contrast, nrow = 1) +
  ggplot2::scale_color_manual(values = c(concordant = "#009E73",
    discordant = "#D55E00", gene_not_in_RNA_assay = "grey60",
    too_sparse_for_effect_threshold = "grey70",
    direction_weak_or_detection_change_small = "#7570B3")) +
  ggplot2::labs(title = "Persistent source d21 versus refined WT d21 crypt effects",
    subtitle = "Pooled-library descriptive HFD/CON log2 effects; gene availability is retained",
    x = "Gourab d21 source log2 HFD/CON",
    y = "Refined WT d21 pooled log2 HFD/CON", color = "Evidence") +
  ggplot2::theme_bw(base_size = 12)
save_vector_plot(plot_ref, PLOTS, "01_source_d21_vs_refined_WT_d21_RNA.pdf", 12, 6)

plot_depth <- stability |>
  dplyr::filter(method == "depth_matched", is.finite(median_log2FC)) |>
  dplyr::left_join(all_effects$RNA |>
    dplyr::transmute(contrast, modality,
      feature, full_effect = log2FC_HFD_vs_CON),
    by = c("contrast", "modality", "feature")) |>
  dplyr::filter(modality == "RNA", is.finite(full_effect)) |>
  ggplot2::ggplot(ggplot2::aes(full_effect, median_log2FC)) +
  ggplot2::geom_hline(yintercept = 0, linewidth = .2) +
  ggplot2::geom_vline(xintercept = 0, linewidth = .2) +
  ggplot2::geom_point(ggplot2::aes(color = sign_agreement_percent), size = .8) +
  ggplot2::facet_wrap(~ contrast, scales = "free") +
  ggplot2::scale_color_viridis_c(limits = c(0, 100)) +
  ggplot2::labs(title = "RNA effect stability under ATAC/RNA depth matching",
    subtitle = "Depth matched using RNA counts for RNA effects; draws are not mouse replicates",
    x = "Full pooled RNA log2FC", y = "Median resampled RNA log2FC",
    color = "Sign match %") +
  ggplot2::theme_bw(base_size = 11)
save_vector_plot(plot_depth, PLOTS, "02_refined_RNA_depth_sensitivity.pdf", 11, 7)

finish_run(OUT, "STEP_08E_COMPLETE.txt", c(
  paste("Focused nuclei:", ncol(obj)),
  paste("Contrasts:", paste(contrast$contrast, collapse = ";")),
  "Global Step 06 annotation altered: FALSE",
  "Biological diet replication or inferential FDR: FALSE"))
```

Review `persistent_genes_refined_WT_concordance.csv` beside the older Step 08
strict unit-8 and Step 08B broad 8+6 results. Before motif enrichment or
cross-genotype projection, inspect gene expression, gene–peak overlap,
minimum detection and the matching sensitivities together.
