---
title: "Step 08F | ISC UCell gates and WT crypt accessibility"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This notebook continues the completed WT Step 08E analysis without changing
Step 05E/05F labels or the Step 06 object. Its **two prespecified arms** are:

1. Focused **C0+C2**, a crypt-core proxy for the G6/G9 part of Gourab's broad
   source label. It is not a reconstruction of every source "Stem" cell.
2. Within each parent **W0, W6, W8**, the top 15%, 10%, and 5% by one fixed
   six-gene ISC UCell signature. Ranking is across CON and HFD *together within
   each parent*, never separately by diet. The three thresholds are nested.

We calculate each parent separately and a union of the three parent gates for
each threshold. The union is an **exploratory sensitivity** until its parent
states, diet balance, and effect directions have been reviewed. W8 is
intentionally retained: it expresses ISC-associated markers but its reciprocal
source match favors the source TA clusters, and its library composition is
strongly imbalanced. Fixed C0–C5 labels are carried through every output.

The ISC core is `Lgr5, Ascl2, Smoc2, Axin2, Lrig1, Rnf43`. These are the
noncycling, source-independent ISC markers in the Step 05E targeted panel. We
keep `Olfm4` as a seventh-gene sensitivity because its colonic detectability
needs inspection. We audit cell-cycle and secretory programs separately rather
than excluding cycling ISCs or defining cells using Gourab's diet-responsive
genes. All signatures, ranks, and thresholds are fixed before ATAC effects.

**Interpretation:** one pooled CON library and one pooled HFD library. Nuclei
are not mouse replicates. The output contains *descriptive diet-associated
accessibility candidates* and no mouse-level differential-accessibility P
value, FDR, or claim of regulatory causation. A threshold with fewer than 30
nuclei from either diet is recorded but its RNA/ATAC effect calculation is
skipped. In particular, W8 top 5% may be too small in CON. A promoter overlap
is a locus annotation, not evidence that a peak regulates that gene.

Run from the project root on SOL after Step 08E finishes. Use the WT
`v2_resequenced` cohort, R 4.4.2, and the same 8 CPU / 96 GiB interactive
session recommended for 08E. `UCell` must be installed in that R 4.4 library;
the notebook stops with a readable preflight message if it is missing.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE,
                      fig.width = 12, fig.height = 7)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else
      stop("Open the project root before running 08F.", call. = FALSE)
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced", MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) stop("Missing 05F/08E shared helper on this checkout.")
source(FOLLOWUP_HELPER)
if (!requireNamespace("UCell", quietly = TRUE))
  stop("Install UCell for the active R 4.4 Bioconductor library before 08F.")
suppressPackageStartupMessages({
  library(Seurat); library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges); library(IRanges)
  library(patchwork)
})
need(identical(DATASET_VERSION, "v2_resequenced") && identical(COHORT_ID, "wt"),
     "08F requires WT v2_resequenced.")
set.seed(8128)
MIN_PER_DIET <- 30L
GATE_FRACTIONS <- c(top15 = .15, top10 = .10, top05 = .05)
UCELL_MAX_RANK <- 400L   # Focused median nFeature_RNA is ~365; compare 600 below.
UCELL_SENS_RANK <- 600L
N_DRAWS <- 20L
RUN_COVERAGE <- TRUE
MAX_COVERAGE_LOCI <- 4L
```

## 1. Input contract and preserved annotations

```{r inputs}
STEP5E <- file.path(OUTPUT_DIR, "05_wnn", "focused_wt_crypt_stem_progenitor")
ROOT_FOLLOWUP <- file.path(OUTPUT_DIR, "05_wnn", FOLLOWUP_RELEASE)
STEP5F <- file.path(ROOT_FOLLOWUP, "05f_review")
STEP8E <- file.path(OUTPUT_DIR, "08e_refined_wt_crypt_diet")
OBJECT <- file.path(STEP5E, "objects", "WT_crypt_core_0_6_8_focused.rds")
ANNOTATED <- file.path(OUTPUT_DIR, "06_annotation", "multiome_annotated.rds")
DECISION <- file.path(STEP5F, "refined_working_labels.csv")
REFERENCE <- file.path(DATASET_CONFIG_DIR, "reference_programs",
                       "persistent_hfd_genes.csv")
OUT <- file.path(OUTPUT_DIR, "08f_isc_ucell_wt_crypt")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
inputs <- c(OBJECT, ANNOTATED, DECISION, REFERENCE,
  file.path(STEP5F, "STEP_05F_COMPLETE.txt"),
  file.path(STEP8E, "STEP_08E_COMPLETE.txt"),
  file.path(STEP8E, "RNA_all_refined_pooled_effects.csv.gz"),
  file.path(STEP8E, "ATAC_all_refined_pooled_effects.csv.gz"))
for (path in inputs) need(file.exists(path), paste("Missing required input:", path))
capture_manifest(inputs, OUT)
map <- read_csv_checked(DECISION) |>
  mutate(focused_cluster = as.character(focused_cluster))
require_columns(map, c("focused_cluster", "analysis_unit", "working_label",
  "analysis_role", "focused_annotation_confidence", "label_status",
  "approved_for_global_annotation"), "05F working map")
need(setequal(map$focused_cluster, as.character(0:5)) &&
  !anyDuplicated(map$focused_cluster), "Expected one 05F C0-C5 row each.")
need(all(map$label_status == "provisional_working") &&
  !any(map$approved_for_global_annotation), "05F provisional labels changed.")
focused <- load_focused(OBJECT)
obj <- focused$object
meta <- focused$metadata |>
  dplyr::rename(parent_annotation_confidence = annotation_confidence) |>
  dplyr::left_join(map, by = "focused_cluster", relationship = "many-to-one")
need(!anyNA(meta$working_label) && !anyNA(meta$diet),
     "Focused barcode/working-label alignment failed.")
rna <- focused$matrices$RNA_counts
atac <- focused$matrices$ATAC_counts
need(identical(meta$cell_barcode, colnames(rna)) &&
     identical(meta$cell_barcode, colnames(atac)),
     "Focused RNA/ATAC matrices are not aligned with metadata.")
need(setequal(unique(meta$parent_wnn_cluster), c("0", "6", "8")),
     "Expected Step 05E parent W0/W6/W8 only.")
write_table(focused$alignment, OUT, "00_matrix_alignment.csv")
```

## 2. Gene audit and rank-based ISC score

`UCell` ranks RNA genes within each nucleus. Use raw, joined `RNA.crypt`
counts, not ATAC gene activities or smoothed values. A common `maxRank` is
held across all three parents. The seventh-gene score and a second rank cap
are sensitivity checks; neither changes the primary gate.

```{r score-ucell}
ISC_CORE <- c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Rnf43")
ISC_EXTENDED <- c(ISC_CORE, "Olfm4") #use this as a negative control
CYCLE_AUDIT <- c("Mki67", "Top2a", "Cenpf", "Ube2c", "Cdk1", "Ccnb1")
SECRETORY_AUDIT <- c("Atoh1", "Muc2", "Reg4", "Spink4", "Kit")
signatures <- list(ISC_core6 = ISC_CORE, ISC_plus_Olfm4 = ISC_EXTENDED,
  Cell_cycle = CYCLE_AUDIT, Secretory_DSC = SECRETORY_AUDIT)
signature_audit <- dplyr::bind_rows(lapply(names(signatures), function(s) {
  tibble(signature = s, gene = signatures[[s]],
         in_RNA_assay = signatures[[s]] %in% rownames(rna))
}))
write_table(signature_audit, OUT, "01_signature_assay_availability.csv")
need(sum(ISC_CORE %in% rownames(rna)) >= 4L,
  "Fewer than four of the six prespecified ISC genes are assayed; review signature before running 08F.")
available_signatures <- lapply(signatures, intersect, rownames(rna))
need(all(lengths(available_signatures) >= 3L),
     "A signature has fewer than three assayed genes; review signature availability.")
score <- UCell::ScoreSignatures_UCell(rna, features = available_signatures,
  maxRank = UCELL_MAX_RANK)
score_sens <- UCell::ScoreSignatures_UCell(rna,
  features = list(ISC_core6 = available_signatures$ISC_core6),
  maxRank = UCELL_SENS_RANK)
need(setequal(rownames(score), meta$cell_barcode) &&
     setequal(rownames(score_sens), meta$cell_barcode),
     "UCell result barcode set differs from focused object.")
score <- score[meta$cell_barcode, , drop = FALSE]
score_sens <- score_sens[meta$cell_barcode, , drop = FALSE]
need(all(paste0(names(signatures), "_UCell") %in% colnames(score)) &&
     "ISC_core6_UCell" %in% colnames(score_sens),
     "Unexpected UCell output column names.")
for (s in names(signatures))
  meta[[paste0(s, "_UCell")]] <- score[, paste0(s, "_UCell")]
meta$ISC_core6_maxRank600 <- score_sens[, "ISC_core6_UCell"]
need(!anyNA(meta$ISC_core6_UCell) && all(is.finite(meta$ISC_core6_UCell)),
     "Missing or nonfinite core ISC UCell scores.")
expressed_genes <- available_signatures$ISC_core6
meta$ISC_core_genes_detected <- Matrix::colSums(
  rna[expressed_genes, , drop = FALSE] > 0)
marker_detection <- dplyr::bind_rows(lapply(ISC_EXTENDED, function(g) {
  if (!g %in% rownames(rna)) return(tibble(gene = g, parent = NA_character_,
    diet = NA_character_, n = 0L, percent_detected = NA_real_))
  meta |>
    dplyr::mutate(detected = as.numeric(rna[g, meta$cell_barcode]) > 0) |>
    dplyr::group_by(parent = parent_wnn_cluster, diet) |>
    dplyr::summarise(gene = g, n = dplyr::n(),
      percent_detected = 100 * mean(detected), .groups = "drop")
}))
write_table(marker_detection, OUT, "01_marker_detection_by_parent_diet.csv")
write_table(meta |> dplyr::select(cell_barcode, sample_id, diet,
  parent_wnn_cluster, focused_cluster, working_label,
  parent_annotation_confidence, focused_annotation_confidence,
  Phase, nFeature_RNA, nCount_RNA, nCount_ATAC, ISC_core_genes_detected,
  dplyr::ends_with("_UCell"), ISC_core6_maxRank600), OUT,
  "01_scores_and_preserved_annotations.csv.gz")
```

## 3. Three nested, diet-blind gates within each parent

The requested percentages are **within W0, W6, and W8 separately**. Ranking
uses the combined CON+HFD parent, with barcode as a reproducible tie break;
no diet or ATAC information enters the gate. A zero score cannot be called a
high-ISC nucleus. We record actual fractions, boundary ties, and mouse-pool
composition so the labels can be interpreted at the end.

```{r define-gates}
meta <- meta |>
  dplyr::group_by(parent_wnn_cluster) |>
  dplyr::arrange(dplyr::desc(ISC_core6_UCell), cell_barcode,
                 .by_group = TRUE) |>
  dplyr::mutate(rank_in_parent = dplyr::row_number(),
    parent_n = dplyr::n(),
    top15 = rank_in_parent <= ceiling(.15 * parent_n) & ISC_core6_UCell > 0,
    top10 = rank_in_parent <= ceiling(.10 * parent_n) & ISC_core6_UCell > 0,
    top05 = rank_in_parent <= ceiling(.05 * parent_n) & ISC_core6_UCell > 0) |>
  dplyr::ungroup()
need(all(!meta$top05 | meta$top10) && all(!meta$top10 | meta$top15),
     "UCell gates are not nested.")
need(setequal(meta$cell_barcode, colnames(rna)), "Gate lost a barcode.")
# Restore the matrix/metadata order before every pooled-effect call.
meta <- meta[match(colnames(rna), meta$cell_barcode), , drop = FALSE]
need(identical(meta$cell_barcode, colnames(rna)), "Cannot restore matrix order.")
gate_cells <- dplyr::bind_rows(lapply(names(GATE_FRACTIONS), function(g) {
  meta |>
    dplyr::transmute(cell_barcode, sample_id, diet, parent_wnn_cluster,
      focused_cluster, analysis_unit, working_label,
      parent_annotation_confidence, focused_annotation_confidence,
      Phase, ISC_core6_UCell, ISC_plus_Olfm4_UCell,
      ISC_core6_maxRank600, Cell_cycle_UCell, Secretory_DSC_UCell,
      ISC_core_genes_detected, nFeature_RNA, nCount_RNA, nCount_ATAC,
      rank_in_parent, parent_n, gate = g, selected = .data[[g]])
}))
write_table(gate_cells, OUT, "02_all_gate_memberships_with_05f_labels.csv.gz")
gate_audit <- gate_cells |>
  dplyr::group_by(gate, parent_wnn_cluster, diet) |>
  dplyr::summarise(n_parent = dplyr::n(), n_selected = sum(selected),
    selected_percent = 100 * mean(selected),
    median_ISC_score_selected = if (sum(selected)) median(ISC_core6_UCell[selected]) else NA_real_,
    median_marker_genes_selected = if (sum(selected)) median(ISC_core_genes_detected[selected]) else NA_real_,
    median_RNA_features_selected = if (sum(selected)) median(nFeature_RNA[selected]) else NA_real_,
    median_ATAC_counts_selected = if (sum(selected)) median(nCount_ATAC[selected]) else NA_real_,
    .groups = "drop")
write_table(gate_audit, OUT, "02_gate_counts_and_depth_by_parent_diet.csv")
gate_boundary <- dplyr::bind_rows(lapply(names(GATE_FRACTIONS), function(g) {
  meta |>
    dplyr::group_by(parent_wnn_cluster) |>
    dplyr::summarise(gate = g, target_n = ceiling(GATE_FRACTIONS[[g]] * dplyr::n()),
      selected_n = sum(.data[[g]]),
      minimum_selected_score = if (sum(.data[[g]])) min(ISC_core6_UCell[.data[[g]]]) else NA_real_,
      boundary_tied_n = if (sum(.data[[g]]))
        sum(ISC_core6_UCell == min(ISC_core6_UCell[.data[[g]]])) else NA_integer_,
      .groups = "drop")
}))
write_table(gate_boundary, OUT, "02_gate_boundary_and_ties.csv")

# Compare primary ranks with a 600-rank sensitivity without changing selection.
rank600 <- meta |>
  dplyr::group_by(parent_wnn_cluster) |>
  dplyr::arrange(dplyr::desc(ISC_core6_maxRank600), cell_barcode,
                 .by_group = TRUE) |>
  dplyr::mutate(rank600 = dplyr::row_number()) |>
  dplyr::ungroup()
rank_audit <- dplyr::bind_rows(lapply(names(GATE_FRACTIONS), function(g) {
  rank600 |>
    dplyr::group_by(parent_wnn_cluster) |>
    dplyr::summarise(gate = g,
      n_primary = sum(.data[[g]]),
      n_600 = sum(rank600 <= ceiling(GATE_FRACTIONS[[g]] * dplyr::n()) &
                    ISC_core6_maxRank600 > 0),
      overlap = sum(.data[[g]] & rank600 <= ceiling(GATE_FRACTIONS[[g]] * dplyr::n()) &
                      ISC_core6_maxRank600 > 0),
      .groups = "drop") |>
    dplyr::mutate(jaccard = ifelse(n_primary + n_600 - overlap > 0,
      overlap / (n_primary + n_600 - overlap), NA_real_))
}))
write_table(rank_audit, OUT, "02_maxRank400_vs_600_gate_overlap.csv")

spec <- list()
add_contrast <- function(key, role, gate, parent, ix) {
  tibble(contrast = key, role = role, gate = gate, parent = parent,
         indices = list(as.integer(ix)))
}
spec[[1]] <- add_contrast("proxy_C0_C2", "crypt_core_proxy", "fixed", "C0+C2",
                          which(meta$focused_cluster %in% c("0", "2")))
for (g in names(GATE_FRACTIONS)) {
  for (p in c("0", "6", "8")) {
    spec[[length(spec) + 1L]] <- add_contrast(paste0("W", p, "_", g),
      "parent_gate", g, paste0("W", p),
      which(meta$parent_wnn_cluster == p & meta[[g]]))
  }
  spec[[length(spec) + 1L]] <- add_contrast(paste0("ISC_union_", g),
    "union_sensitivity", g, "W0+W6+W8", which(meta[[g]]))
}
contrast <- dplyr::bind_rows(spec) |>
  dplyr::mutate(n_CON = vapply(indices, function(ix) sum(meta$diet[ix] == "CON"), integer(1)),
                n_HFD = vapply(indices, function(ix) sum(meta$diet[ix] == "HFD"), integer(1)),
                eligible = n_CON >= MIN_PER_DIET & n_HFD >= MIN_PER_DIET,
                status = ifelse(eligible, "descriptive_effect_eligible",
                                "insufficient_nuclei_in_one_diet"))
need(!anyDuplicated(contrast$contrast) && contrast$eligible[1],
     "Duplicate contrast or ineligible C0+C2 proxy.")
write_table(dplyr::select(contrast, -indices), OUT,
            "02_contrast_definitions_and_eligibility.csv")
```

## 4. Show UCell membership and source subclusters

```{r membership-figures}
global <- readRDS(ANNOTATED)
need("wnn.umap" %in% names(global@reductions), "Step 06 lacks global WNN UMAP.")
coords <- as.data.frame(SeuratObject::Embeddings(global, "wnn.umap"))
names(coords)[1:2] <- c("x", "y")
coords$cell_barcode <- rownames(coords)
coords$selected_focus <- coords$cell_barcode %in% meta$cell_barcode
fg <- dplyr::left_join(coords, dplyr::select(meta, cell_barcode,
  parent_wnn_cluster, focused_cluster, working_label, diet, ISC_core6_UCell,
  top15, top10, top05), by = "cell_barcode")
need(nrow(fg) == nrow(coords) &&
     !anyNA(fg$ISC_core6_UCell[fg$selected_focus]), "UMAP cell alignment failed.")
base_points <- fg |>
  ggplot2::ggplot(ggplot2::aes(x, y)) +
  ggplot2::geom_point(color = "grey85", size = .10, alpha = .5) +
  ggplot2::coord_equal() + ggplot2::theme_classic(base_size = 12) +
  ggplot2::labs(x = "Global WNN UMAP 1", y = "Global WNN UMAP 2")
p_score <- base_points +
  ggplot2::geom_point(data = dplyr::filter(fg, selected_focus),
    ggplot2::aes(color = ISC_core6_UCell), size = .25) +
  ggplot2::scale_color_viridis_c(option = "magma") +
  ggplot2::labs(title = "Six-gene ISC score on the global WT WNN UMAP",
    subtitle = "Same score definition in W0, W6 and W8; all other cells grey",
    color = "UCell")
save_vector_plot(p_score, PLOTS, "01_ISC_UCell_on_global_WNN.pdf", 12, 7)
selected_plot <- dplyr::bind_rows(lapply(names(GATE_FRACTIONS), function(g) {
  fg |>
    dplyr::filter(selected_focus, .data[[g]]) |>
    dplyr::mutate(gate = g)
}))
background_plot <- dplyr::bind_rows(lapply(names(GATE_FRACTIONS),
  function(g) dplyr::mutate(dplyr::select(coords, x, y), gate = g)))
p_gates <- ggplot2::ggplot() +
  ggplot2::geom_point(data = background_plot, ggplot2::aes(x, y),
    color = "grey87", size = .08, alpha = .35) +
  ggplot2::geom_point(data = selected_plot,
    ggplot2::aes(x, y, color = paste0("W", parent_wnn_cluster)),
    size = .35, alpha = .85) +
  ggplot2::facet_wrap(~gate, nrow = 1) +
  ggplot2::scale_color_manual(values = c(W0 = "#007F86", W6 = "#B34A6A",
                                         W8 = "#3454A5"), name = "Parent") +
  ggplot2::coord_equal() + ggplot2::theme_classic(base_size = 11) +
  ggplot2::labs(title = "Nested ISC score gates on the original WNN embedding",
    subtitle = "15%, 10%, and 5% are ranked within each WNN parent, pooled across diets",
    x = "WNN UMAP 1", y = "WNN UMAP 2")
save_vector_plot(p_gates, PLOTS, "02_nested_gate_locations.pdf", 16, 6)
p_scores <- ggplot2::ggplot(meta,
  ggplot2::aes(x = diet, y = ISC_core6_UCell, fill = diet)) +
  ggplot2::geom_violin(scale = "width", trim = TRUE, color = "grey25") +
  ggplot2::geom_boxplot(width = .14, outlier.shape = NA, fill = "white") +
  ggplot2::facet_wrap(~parent_wnn_cluster, nrow = 1) +
  ggplot2::scale_fill_manual(values = c(CON = "#E69F00", HFD = "#56B4E9"),
                             guide = "none") +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::labs(title = "ISC scores and diet balance in each parent",
    subtitle = "Score distributions are descriptive and can depend on RNA detection",
    x = NULL, y = "ISC core UCell score")
save_vector_plot(p_scores, PLOTS, "03_parent_score_distributions.pdf", 11, 6)
subcluster_counts <- gate_cells |>
  dplyr::filter(selected) |>
  dplyr::count(gate, parent_wnn_cluster, focused_cluster,
               working_label, diet, name = "nuclei")
write_table(subcluster_counts, OUT, "03_selected_05f_subclusters_by_parent_diet.csv")
p_subcluster <- ggplot2::ggplot(subcluster_counts,
  ggplot2::aes(x = paste0("W", parent_wnn_cluster, " | ", diet), y = nuclei,
    fill = paste0("C", focused_cluster, " ", working_label))) +
  ggplot2::geom_col() + ggplot2::facet_wrap(~gate, nrow = 1, scales = "free_y") +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 legend.text = ggplot2::element_text(size = 8)) +
  ggplot2::labs(title = "Which fixed C0-C5 states enter each ISC score gate?",
    subtitle = "No new subclustering or relabeling; one bar per parent and diet",
    x = NULL, y = "Selected nuclei", fill = "05F working annotation")
save_vector_plot(p_subcluster, PLOTS, "04_preserved_subcluster_composition.pdf", 16, 8)
```

## 5. Descriptive RNA and ATAC effects on the same fixed consensus peaks

No significance tests use nuclei as replicates. Run the original Step 08E
`pooled_effect()` with the same CPM pseudocount and evaluable/clear filters.
The newly calculated proxy and UCell results use **exactly the same** focused
RNA genes and ATAC consensus peak rows. The eligible contrasts alone receive
fold changes. Sample-depth and selected-cell counts remain in each record.

```{r pooled-effects}
all_effects <- list()
for (modality in c("RNA", "ATAC")) {
  counts <- if (modality == "RNA") rna else atac
  depth <- Matrix::colSums(counts)
  out <- dplyr::bind_rows(lapply(which(contrast$eligible), function(i) {
    contrast_key <- contrast$contrast[[i]]
    gate_key <- contrast$gate[[i]]
    role_key <- contrast$role[[i]]
    parent_key <- contrast$parent[[i]]
    message("08F ", modality, " ", contrast_key, ": ",
            contrast$n_CON[i], " CON / ", contrast$n_HFD[i], " HFD")
    pooled_effect(counts, meta, contrast$indices[[i]], modality,
                  total_depth = depth) |>
      dplyr::mutate(contrast = .env$contrast_key, gate = .env$gate_key,
        role = .env$role_key, parent = .env$parent_key, .before = 1)
  }))
  need(nrow(out) > 0L && !anyDuplicated(out[c("contrast", "feature")]),
       paste("Bad", modality, "effect table or duplicate feature keys."))
  write_table(out, OUT, paste0("05_", modality, "_all_eligible_pooled_effects.csv.gz"))
  all_effects[[modality]] <- out
  rm(out); invisible(gc())
}
```

## 6. Parent consistency, proxy overlap, and depth stability

"Shared" below means the same consensus peak has a descriptive *clear*
effect in the same direction in **at least two eligible parents**. It does
not mean a replicated diet effect; if W8 was ineligible, it is explicitly
absent from that comparison. The C0+C2 and UCell union nuclei may overlap, so
their peak agreement is a context comparison, not independent validation.

```{r compare-peaks}
atac_effects <- all_effects$ATAC
parent_peaks <- atac_effects |>
  dplyr::filter(role == "parent_gate") |>
  dplyr::group_by(gate, feature) |>
  dplyr::summarise(n_parent_effects = dplyr::n(),
    n_parent_evaluable = sum(evaluable, na.rm = TRUE),
    n_HFD_open = sum(clear_effect & log2FC_HFD_vs_CON > 0, na.rm = TRUE),
    n_CON_open = sum(clear_effect & log2FC_HFD_vs_CON < 0, na.rm = TRUE),
    HFD_open_parents = paste(parent[clear_effect &
      log2FC_HFD_vs_CON > 0], collapse = ";"),
    CON_open_parents = paste(parent[clear_effect &
      log2FC_HFD_vs_CON < 0], collapse = ";"), .groups = "drop") |>
  dplyr::mutate(parent_pattern = dplyr::case_when(
    n_HFD_open > 0 & n_CON_open > 0 ~ "opposing_parent_directions",
    n_HFD_open >= 2 ~ "shared_HFD_open",
    n_CON_open >= 2 ~ "shared_CON_open",
    n_HFD_open + n_CON_open == 1 ~ "one_parent_clear",
    TRUE ~ "no_clear_parent_effect"))
write_table(parent_peaks, OUT, "06_parent_peak_direction_comparison.csv.gz")
parent_review <- contrast |>
  dplyr::filter(gate %in% names(GATE_FRACTIONS), role == "parent_gate") |>
  dplyr::select(gate, parent, n_CON, n_HFD, eligible, status)
write_table(parent_review, OUT, "06_which_parent_effects_are_available.csv")

proxy_peak <- atac_effects |>
  dplyr::filter(contrast == "proxy_C0_C2") |>
  dplyr::select(feature, proxy_evaluable = evaluable,
    proxy_clear = clear_effect, proxy_log2FC = log2FC_HFD_vs_CON,
    proxy_CON_percent = CON_percent, proxy_HFD_percent = HFD_percent)
union_peaks <- atac_effects |>
  dplyr::filter(role == "union_sensitivity") |>
  dplyr::select(gate, feature, union_evaluable = evaluable,
    union_clear = clear_effect, union_log2FC = log2FC_HFD_vs_CON,
    union_CON_percent = CON_percent, union_HFD_percent = HFD_percent) |>
  dplyr::left_join(proxy_peak, by = "feature", relationship = "many-to-one") |>
  dplyr::left_join(parent_peaks, by = c("gate", "feature"),
                   relationship = "one-to-one") |>
  dplyr::mutate(proxy_union_class = dplyr::case_when(
    !union_evaluable | !proxy_evaluable ~ "one_or_both_sparse",
    union_clear & proxy_clear &
      sign(union_log2FC) == sign(proxy_log2FC) ~ "both_clear_same_direction",
    union_clear & proxy_clear &
      sign(union_log2FC) != sign(proxy_log2FC) ~ "both_clear_opposite_direction",
    union_clear & !proxy_clear ~ "union_only_clear",
    !union_clear & proxy_clear ~ "proxy_only_clear",
    TRUE ~ "neither_clear"))
write_table(union_peaks, OUT, "06_proxy_vs_UCell_union_all_peaks.csv.gz")
comparison_counts <- union_peaks |>
  dplyr::count(gate, parent_pattern, proxy_union_class, name = "peaks")
write_table(comparison_counts, OUT, "06_peak_comparison_counts.csv")

p_peak_scatter <- ggplot2::ggplot(dplyr::filter(union_peaks,
  union_evaluable, proxy_evaluable),
  ggplot2::aes(proxy_log2FC, union_log2FC, color = proxy_union_class)) +
  ggplot2::geom_hline(yintercept = 0, linewidth = .2) +
  ggplot2::geom_vline(xintercept = 0, linewidth = .2) +
  ggplot2::geom_point(size = .28, alpha = .25) +
  ggplot2::facet_wrap(~gate, nrow = 1) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Proxy versus UCell-selected accessibility effects",
    subtitle = "Same consensus peaks, partly overlapping nuclei; no independent replication",
    x = "C0+C2 HFD/CON pooled ATAC log2FC",
    y = "Selected W0+W6+W8 pooled ATAC log2FC",
    color = "Description")
save_vector_plot(p_peak_scatter, PLOTS, "05_proxy_vs_union_peak_effects.pdf", 14, 6)

# Inspect group identity and depth before interpreting a union.
identity_review <- gate_cells |>
  dplyr::filter(selected) |>
  dplyr::group_by(gate, parent_wnn_cluster, diet) |>
  dplyr::summarise(n = dplyr::n(),
    median_ISC = median(ISC_core6_UCell),
    median_ISC_plus_Olfm4 = median(ISC_plus_Olfm4_UCell),
    median_cycle = median(Cell_cycle_UCell),
    median_secretory = median(Secretory_DSC_UCell),
    median_ISC_genes_detected = median(ISC_core_genes_detected),
    median_RNA_features = median(nFeature_RNA),
    median_ATAC_counts = median(nCount_ATAC), .groups = "drop")
write_table(identity_review, OUT, "06_parent_identity_and_depth_review.csv")
```

Use the same Step 08E equal-cell and RNA/ATAC-depth-matched draws for a fixed,
bounded set of peaks, selected once per contrast from the full effect. These
draws test dependence on cell/depth composition, **not** biological variance.

```{r resampling}
resample_keys <- c("proxy_C0_C2", "ISC_union_top15", "ISC_union_top10",
  "ISC_union_top05", "W0_top15", "W6_top15", "W8_top15")
resample_keys <- intersect(resample_keys, contrast$contrast[contrast$eligible])
stability <- dplyr::bind_rows(lapply(resample_keys, function(key) {
  i <- match(key, contrast$contrast)
  t <- atac_effects |>
    dplyr::filter(contrast == key, evaluable) |>
    dplyr::arrange(dplyr::desc(abs(log2FC_HFD_vs_CON)))
  picked <- head(t$feature, 200L)
  if (!length(picked)) return(tibble())
  message("Depth and cell resampling: ", key, " (", length(picked), " peaks)")
  resampling_effects(atac, meta, contrast$indices[[i]], t, picked,
                     "ATAC", iterations = N_DRAWS) |>
    dplyr::mutate(contrast = key, .before = 1)
}))
write_table(stability, OUT, "07_ATAC_equal_cell_and_depth_sensitivity.csv.gz")
depth_stability <- stability |>
  dplyr::filter(method == "depth_matched") |>
  dplyr::select(contrast, feature, valid_iterations,
    depth_median_log2FC = median_log2FC,
    depth_sign_agreement_percent = sign_agreement_percent,
    sampled_n_per_diet_min)
```

## 7. Prioritize loci and overlay Gourab's Figure 2 genes afterward

The 37 HyperMet and 91 HypoImmune names supplied by Dom are used **only now**.
They do not enter clustering, the UCell score, thresholding, or peak-effect
selection. A source gene is displayed beside a peak only for an overlap with
its ±2 kb annotated promoter, with RNA shown as a separate observation. This
does not infer an enhancer target. `Atpif1` is mapped to `Atp5if1` in the
current assay; absent genes remain in the gene audit.

```{r posthoc-source-overlay}
GOURAB_PROGRAMS <- list(
 HyperMet = strsplit(paste(
  "Mgst1 Gstm1 Idh1 Gss Gsta4 Gclm Sod1 Sult1a1 Hmgcs2 Acsm3 Csad",
  "Acss2 Shmt1 Impdh2 Akr7a5 Adk Aprt Hao2 Hadh Mecr Scd2 Mif Fads2",
  "Acads Sox9 Aldob Atpif1 Uqcrc1 Slc4a4 Hspd1 Eef1a1 Nox1 Cyba",
  "Tkt Gpx2 Gsto1 Adh1"), " ")[[1]],
 HypoImmune = strsplit(paste(
  "Nupr1 Crip1 Atp2a3 Cdkn1a Cd74 Herpud1 Dnajc10 Ern2 Itpr1 Xbp1",
  "P4hb Selenok Selenos Creb3l1 Serinc3 Sh3glb1 H2-K1 H2-D1 B2m",
  "H2-T23 Psmb8 Calr Ceacam1 Klf4 Mia3 Cd24a Rap1gap Pla2g2f",
  "Spint2 Trp53inp1 Gabarap Gabarapl2 Camk2n1 S100a13 Oit1 Syt7",
  "Serp1 Rab3d Rab15 Agr2 Hspa5 Manf Dnajc3 Creb3l4 Pdia3 Pdia6",
  "Sdf2l1 Sec61b Ramp1 Galnt7 C1galt1 Gfpt1 Gcnt3 St3gal6 Tmem59",
  "St6galnac6 Galnt10 C1galt1c1 Fut2 Galnt3 Rpn1 Krtcap2 Pmm2",
  "Prdx6 Qsox1 Sh3bgrl3 Txndc5 Pdia5 Dap Foxa3 Clca1 Tspan13",
  "Atp2c2 P2rx4 Stim2 Asph Fxyd3 Lrrc26 Kit Plpp1 Pdxdc1 Rrbp1",
  "Tram1 Sec62 Ssr1 Bag1 Lman1 Actb Fos Il13ra1 Ifngr1"), " ")[[1]]
)
need(length(GOURAB_PROGRAMS$HyperMet) == 37L &&
     length(GOURAB_PROGRAMS$HypoImmune) == 91L,
     "Figure 2 program list length changed.")
source_genes <- dplyr::bind_rows(lapply(names(GOURAB_PROGRAMS), function(p) {
  tibble(program = p, original_symbol = GOURAB_PROGRAMS[[p]],
    gene = ifelse(GOURAB_PROGRAMS[[p]] == "Atpif1", "Atp5if1",
                  GOURAB_PROGRAMS[[p]]),
    expected_HFD_sign = ifelse(p == "HyperMet", 1L, -1L))
})) |>
  dplyr::mutate(in_RNA_assay = gene %in% rownames(rna))
write_table(source_genes, OUT, "08_Gourab_gene_assay_audit.csv")
source_RNA <- all_effects$RNA |>
  dplyr::filter(feature %in% source_genes$gene) |>
  dplyr::left_join(source_genes, by = c("feature" = "gene"),
                   relationship = "many-to-one") |>
  dplyr::mutate(source_direction_match = evaluable &
    sign(log2FC_HFD_vs_CON) == expected_HFD_sign,
    source_clear_same_direction = clear_effect & source_direction_match)
write_table(source_RNA, OUT, "08_posthoc_Gourab_RNA_effects_by_population.csv")

promoter_map <- promoter_peak_map(obj, window = 2000L)$map |>
  dplyr::filter(peak %in% rownames(atac)) |>
  dplyr::select(gene, peak, chromosome, peak_start, peak_end,
    strand, tss, strand_oriented_center_distance) |>
  dplyr::distinct()
source_promoters <- promoter_map |>
  dplyr::inner_join(source_genes, by = "gene", relationship = "many-to-one")
write_table(source_promoters, OUT,
            "08_source_gene_promoter_peak_annotations.csv.gz")

# Pick locus examples using ATAC alone. The rank is set by union/proxy ATAC
# effects and parent consistency, before checking any source-gene overlay.
peak_priority <- union_peaks |>
  dplyr::left_join(depth_stability |>
    dplyr::filter(grepl("^ISC_union_", contrast)) |>
    dplyr::mutate(gate = sub("^ISC_union_", "", contrast)) |>
    dplyr::select(gate, feature, union_depth_sign = depth_sign_agreement_percent,
                  union_valid_draws = valid_iterations),
    by = c("gate", "feature"), relationship = "one-to-one") |>
  dplyr::mutate(direction = ifelse(union_log2FC > 0, "HFD_open", "CON_open"),
    eligible_candidate = union_evaluable & union_clear,
    stable_if_resampled = !is.na(union_depth_sign) &
      union_valid_draws >= 15L & union_depth_sign >= 80,
    proxy_agrees = proxy_union_class == "both_clear_same_direction",
    parent_recurrence = pmax(dplyr::coalesce(n_HFD_open, 0L),
                             dplyr::coalesce(n_CON_open, 0L)),
    rank_score = 100 * as.integer(proxy_agrees) +
      10 * parent_recurrence + 5 * as.integer(stable_if_resampled) +
      abs(union_log2FC))
write_table(peak_priority, OUT, "08_ATAC_first_candidate_peak_priority.csv.gz")
annotated_priority <- peak_priority |>
  dplyr::filter(eligible_candidate) |>
  dplyr::left_join(promoter_map |> dplyr::group_by(peak) |>
    dplyr::summarise(promoter_genes = paste(sort(unique(gene)), collapse = ";"),
                     .groups = "drop"), by = c("feature" = "peak"),
    relationship = "many-to-one") |>
  dplyr::left_join(source_promoters |> dplyr::group_by(peak) |>
    dplyr::summarise(Gourab_promoter_genes =
      paste(sort(unique(original_symbol)), collapse = ";"), .groups = "drop"),
    by = c("feature" = "peak"), relationship = "many-to-one")
write_table(annotated_priority, OUT,
            "08_ATAC_candidates_with_posthoc_promoter_context.csv.gz")
```

## 8. Candidate BED sets and locus coverage

Export the descriptive candidate sets in the mouse GRCm39 reference. BED
starts are zero-based and ends are exclusive. These
files are for *review and later motif analysis*, not approved motif target
sets. If coverage fragments are inaccessible, keep the peak tables and record the failed plot attempts rather than suppressing the other outputs.

```{r candidate-bed-and-coverage}
peak_gr <- GenomicRanges::granges(obj[["ATAC.crypt"]])
names(peak_gr) <- rownames(atac)
need(identical(names(peak_gr), rownames(atac)),
     "ATAC range/feature identity mismatch before BED export.")
for (g in names(GATE_FRACTIONS)) {
  for (dir in c("HFD_open", "CON_open")) {
    x <- annotated_priority |>
      dplyr::filter(gate == .env$g, direction == .env$dir,
        stable_if_resampled | parent_recurrence >= 2L) |>
      dplyr::arrange(dplyr::desc(rank_score)) |>
      dplyr::distinct(feature, .keep_all = TRUE)
    gr <- peak_gr[match(x$feature, names(peak_gr))]
    bed <- tibble(chrom = as.character(GenomicRanges::seqnames(gr)),
      start0 = as.integer(BiocGenerics::start(gr)) - 1L,
      end = as.integer(BiocGenerics::end(gr)), peak = x$feature)
    readr::write_tsv(bed, file.path(OUT, paste0("09_", g, "_", dir,
                     "_review_only_GRCm39.bed")), col_names = FALSE)
  }
}

# Prioritize at most four distinct ATAC loci from the 15% union. Source-gene
# annotations are joined only after this selection and cannot drive it.
coverage_loci <- peak_priority |>
  dplyr::filter(gate == "top15", eligible_candidate) |>
  dplyr::arrange(dplyr::desc(proxy_agrees),
                  dplyr::desc(parent_recurrence),
                  dplyr::desc(stable_if_resampled),
                  dplyr::desc(abs(union_log2FC))) |>
  dplyr::distinct(feature, .keep_all = TRUE) |>
  dplyr::slice_head(n = MAX_COVERAGE_LOCI) |>
  dplyr::select(feature, direction, union_log2FC,
                proxy_log2FC, parent_recurrence,
                union_depth_sign, proxy_agrees)
write_table(coverage_loci, OUT, "09_prespecified_ATAC_locus_examples.csv")
# Match these ATAC-first loci to RNA only after the loci have been selected.
# An overlapping promoter annotates location, not a peak-to-gene link.
example_keys <- c("proxy_C0_C2", "ISC_union_top15")
example_peaks <- atac_effects |>
  dplyr::filter(contrast %in% example_keys,
                feature %in% coverage_loci$feature) |>
  dplyr::transmute(contrast, feature, layer = "ATAC peak",
    label = "peak", log2FC_HFD_vs_CON, evaluable, clear_effect,
    CON_percent, HFD_percent)
example_rna <- promoter_map |>
  dplyr::filter(peak %in% coverage_loci$feature,
                gene %in% rownames(rna)) |>
  dplyr::distinct(feature = peak, gene) |>
  dplyr::inner_join(all_effects$RNA |>
    dplyr::filter(contrast %in% example_keys) |>
    dplyr::select(contrast, gene = feature, log2FC_HFD_vs_CON, evaluable,
                  clear_effect, CON_percent, HFD_percent),
    by = "gene", relationship = "many-to-many") |>
  dplyr::transmute(contrast, feature, layer = "Promoter-overlap RNA",
    label = gene, log2FC_HFD_vs_CON, evaluable, clear_effect,
    CON_percent, HFD_percent)
example_context <- dplyr::bind_rows(example_peaks, example_rna) |>
  dplyr::mutate(feature = factor(feature, levels = rev(coverage_loci$feature)))
write_table(example_context, OUT, "09_ATAC_first_examples_with_RNA_context.csv")
if (nrow(example_context)) {
  p_examples <- ggplot2::ggplot(example_context,
    ggplot2::aes(log2FC_HFD_vs_CON, feature, color = clear_effect,
                 shape = evaluable)) +
    ggplot2::geom_vline(xintercept = 0, linewidth = .25) +
    ggplot2::geom_point(size = 2.5,
      position = ggplot2::position_dodge(width = .5)) +
    ggplot2::geom_text(data = dplyr::filter(example_context,
      layer == "Promoter-overlap RNA"), ggplot2::aes(label = label),
      color = "grey20", size = 2.5, vjust = -.8, check_overlap = TRUE) +
    ggplot2::facet_grid(layer ~ contrast, scales = "free_x") +
    ggplot2::scale_color_manual(values = c(`FALSE` = "grey60",
                                    `TRUE` = "#007F86")) +
    ggplot2::theme_bw(base_size = 10) +
    ggplot2::labs(title = "ATAC-first locus examples and overlapping promoter RNA",
      subtitle = "Same peaks in both arms; RNA rows only when an annotated ±2 kb promoter overlaps; pooled descriptive effects",
      x = "Pooled HFD/CON log2FC", y = "Consensus peak",
      color = "Clear effect", shape = "Evaluable")
  save_vector_plot(p_examples, PLOTS, "08_locus_ATAC_and_RNA_context.pdf", 13, 7)
}
coverage_keys <- c("proxy_C0_C2", "ISC_union_top15")
coverage_status <- tidyr::expand_grid(contrast = coverage_keys,
                                     feature = coverage_loci$feature) |>
  dplyr::mutate(status = "not_attempted", detail = "")
if (RUN_COVERAGE && nrow(coverage_loci)) {
  if (!length(Signac::Fragments(global[["ATAC"]]))) {
    coverage_status$status <- "skipped"
    coverage_status$detail <- "Step 06 ATAC fragment references are unavailable"
  } else {
    coverage_groups <- list(
      proxy_C0_C2 = contrast$indices[[match("proxy_C0_C2", contrast$contrast)]],
      ISC_union_top15 = contrast$indices[[match("ISC_union_top15", contrast$contrast)]])
    for (key in names(coverage_groups)) {
      ix <- coverage_groups[[key]]
      if (!contrast$eligible[match(key, contrast$contrast)]) {
        coverage_status$status[coverage_status$contrast == key] <- "ineligible"
        next
      }
      coverage_obj <- subset(global, cells = meta$cell_barcode[ix])
      SeuratObject::DefaultAssay(coverage_obj) <- "ATAC"
      for (j in seq_len(nrow(coverage_loci))) {
        peak <- coverage_loci$feature[j]
        k <- which(coverage_status$contrast == key &
                   coverage_status$feature == peak)
        gr <- peak_gr[match(peak, names(peak_gr))]
        region <- paste0(as.character(GenomicRanges::seqnames(gr)), ":",
          max(1L, as.integer(BiocGenerics::start(gr)) - 5000L), "-",
          as.integer(BiocGenerics::end(gr)) + 5000L)
        tryCatch({
          p <- Signac::CoveragePlot(coverage_obj, region = region,
  assay = "ATAC", annotation = "gene", peaks = TRUE,
  links = FALSE, tile = FALSE, group.by = "diet",
  ymax = "q95", sep = c(":", "-")) +
  patchwork::plot_annotation(title = paste(key, peak),
    subtitle = "CON and HFD pooled depth-normalized coverage; same GRCm39 locus")
          ggplot2::ggsave(file.path(PLOTS, paste0("09_", key, "_locus_", j,
            ".pdf")), plot = p, device = grDevices::pdf,
            useDingbats = FALSE, width = 13, height = 7)
          coverage_status$status[k] <- "saved"
        }, error = function(e) {
          coverage_status$status[k] <<- "plot_failed"
          coverage_status$detail[k] <<- conditionMessage(e)
          warning("Coverage plot failed for ", key, " / ", peak, ": ",
                  conditionMessage(e))
        })
      }
      rm(coverage_obj); invisible(gc())
    }
  }
}
write_table(coverage_status, OUT, "09_coverage_plot_status.csv")
```

## 9. Review and completion

Inspect `02_contrast_definitions_and_eligibility.csv` first. Look at
`06_parent_identity_and_depth_review.csv` and the fixed C0–C5 assignments
before using the pooled UCell union. Next compare the direction of shared and
parent-specific peaks, the bounded depth draws, and the ATAC-first loci. Keep
the source-gene overlap as post hoc context. The BED sets are not yet motif
results; future enrichment will need a matched background for GC, width,
baseline accessibility, and genomic context, plus a minimum peak-set size.

```{r finish}
finish_run(OUT, "STEP_08F_COMPLETE.txt", c(
  "Input: completed WT Step 08E and unchanged Step 05E/05F and 06 objects",
  paste("Signature:", paste(ISC_CORE, collapse = ",")),
  paste("UCell maxRank:", UCELL_MAX_RANK,
        "; sensitivity maxRank:", UCELL_SENS_RANK),
  paste("Eligibility min nuclei per diet:", MIN_PER_DIET),
  paste("Eligible contrasts:", paste(contrast$contrast[contrast$eligible],
                                    collapse = ";")),
  "05F labels altered: FALSE; biological diet replication/FDR: FALSE",
  "Peak sets and locus choices descriptive; motif enrichment: NOT PERFORMED"))
```
