---
title: "Step 08G | Separate crypt states and score-defined C0/W0 ISC nuclei"
author: "Mana Lab"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This notebook follows the completed WT Step 08F review. The prior
W0+W6+W8 union was compositionally uneven: W8 supplied 100 CON and 19 HFD nuclei at the 15% gate, while the W6 gate usually detected only one of the six ISC markers. It must not be interpreted as one high-confidence stem population. The same consensus ATAC peaks and Step 08E pooled-effect function are retained throughout this follow-up.

**Prespecified groups:** C0 alone, C2 alone, C0+C2 (the broad Gourab proxy),
all C0 nuclei from W0, and score-ranked C0/W0 nuclei at 15% and 10%.
The C0/W0 5%, 20%, and 25% gates are sensitivity views. Ranking is within
C0/W0 over both diets together. W6/C2 and W8/C3+C5 are *separate context
comparators*, never pooled into the selected ISC arm. Fixed 05F labels are
carried forward; no object or global annotation is modified.

**Score diagnostic:** the original Step 08F six-gene UCell score (`Lgr5`,
`Ascl2`, `Smoc2`, `Axin2`, `Lrig1`, `Rnf43`) remains the primary rank.
The observed four-gene set (`Lgr5`, `Smoc2`, `Lrig1`, `Rnf43`) is scored at
maxRank 400 and 600 **only as a sensitivity**. This lets us see whether
undetected/rare members or rank cap change membership, without choosing a
signature after looking at ATAC effects. `Olfm4` is not a distal-colon gate
member and its absent detection in 08F is expected.

**Statistical scope:** one pooled CON and one pooled HFD library. Results are
descriptive diet-associated effects; nuclei are not mouse replicates. Do
not call the candidate peaks significant DARs, report FDR, infer causation
from a promoter overlap, or use the post hoc source-gene program to choose
nuclei. Peak-set files are review inputs for a later, matched-background
motif analysis, not motif enrichment results.

Run interactively from the repository root on SOL in the WT R 4.4.2 session
after Step 08F completed. This creates a new `08g_*` result directory and
leaves Step 08E/08F outputs intact.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      fig.width = 12, fig.height = 7)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Open the project root before running 08G.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced", MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) stop("Missing shared WT crypt helper.")
source(FOLLOWUP_HELPER)
if (!requireNamespace("UCell", quietly = TRUE))
  stop("Install UCell in the active R 4.4 library before 08G.")
suppressPackageStartupMessages({
  library(Seurat); library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges); library(IRanges); library(patchwork)
})
need(identical(DATASET_VERSION, "v2_resequenced") && identical(COHORT_ID, "wt"),
     "08G requires WT v2_resequenced.")
set.seed(8139)
MIN_PER_DIET <- 30L
PRIMARY_GATES <- c(top15 = .15, top10 = .10)
SENSITIVITY_GATES <- c(top05 = .05, top20 = .20, top25 = .25)
GATES <- c(PRIMARY_GATES, SENSITIVITY_GATES)
UCELL_MAX_RANK <- 400L
UCELL_SENS_RANK <- 600L
N_DRAWS <- 20L
RUN_COVERAGE <- TRUE
MAX_COVERAGE_LOCI <- 4L
```

## 1. Inputs, barcode alignment, and fixed labels

```{r inputs}
STEP5E <- file.path(OUTPUT_DIR, "05_wnn", "focused_wt_crypt_stem_progenitor")
STEP5F <- file.path(OUTPUT_DIR, "05_wnn", FOLLOWUP_RELEASE, "05f_review")
STEP8F <- file.path(OUTPUT_DIR, "08f_isc_ucell_wt_crypt")
OBJECT <- file.path(STEP5E, "objects", "WT_crypt_core_0_6_8_focused.rds")
ANNOTATED <- file.path(OUTPUT_DIR, "06_annotation", "multiome_annotated.rds")
DECISION <- file.path(STEP5F, "refined_working_labels.csv")
SCORES_08F <- file.path(STEP8F, "01_scores_and_preserved_annotations.csv.gz")
SOURCE_GENES <- file.path(STEP8F, "08_Gourab_gene_assay_audit.csv")
OUT <- file.path(OUTPUT_DIR, "08g_C0_C2_and_C0W0_ISC")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
inputs <- c(OBJECT, ANNOTATED, DECISION, SCORES_08F, SOURCE_GENES,
  file.path(STEP5F, "STEP_05F_COMPLETE.txt"),
  file.path(STEP8F, "STEP_08F_COMPLETE.txt"))
for (p in inputs) need(file.exists(p), paste("Missing required input:", p))
capture_manifest(inputs, OUT)
map <- read_csv_checked(DECISION) |>
  dplyr::mutate(focused_cluster = as.character(focused_cluster))
require_columns(map, c("focused_cluster", "working_label", "analysis_unit",
  "focused_annotation_confidence", "label_status",
  "approved_for_global_annotation"), "Step 05F map")
need(setequal(map$focused_cluster, as.character(0:5)) &&
     !anyDuplicated(map$focused_cluster) &&
     all(map$label_status == "provisional_working") &&
     !any(map$approved_for_global_annotation), "Step 05F labels changed.")
focused <- load_focused(OBJECT)
obj <- focused$object
rna <- focused$matrices$RNA_counts
rna_data <- focused$matrices$RNA_data
atac <- focused$matrices$ATAC_counts
meta <- focused$metadata |>
  dplyr::rename(parent_annotation_confidence = annotation_confidence) |>
  dplyr::left_join(map, by = "focused_cluster", relationship = "many-to-one")
need(identical(meta$cell_barcode, colnames(rna)) &&
     identical(meta$cell_barcode, colnames(atac)) &&
     identical(meta$cell_barcode, colnames(rna_data)) &&
     !anyNA(meta$working_label), "Focused assays/labels are misaligned.")
score_input <- read_csv_checked(SCORES_08F)
require_columns(score_input, c("cell_barcode", "ISC_core6_UCell",
  "ISC_core6_maxRank600", "ISC_core_genes_detected"), "08F score table")
need(!anyDuplicated(score_input$cell_barcode) &&
     setequal(score_input$cell_barcode, meta$cell_barcode),
     "08F score table and focused assay have different barcodes.")
score_input <- score_input[match(meta$cell_barcode, score_input$cell_barcode), ]
need(identical(meta$cell_barcode, score_input$cell_barcode),
     "Could not order the Step 08F scores.")
meta$ISC_core6_UCell <- score_input$ISC_core6_UCell
meta$ISC_core6_rank600 <- score_input$ISC_core6_maxRank600
meta$ISC_core_genes_detected <- score_input$ISC_core_genes_detected
need(all(is.finite(meta$ISC_core6_UCell)) &&
     setequal(unique(meta$diet), c("CON", "HFD")), "Missing score or diet.")
write_table(focused$alignment, OUT, "00_matrix_alignment.csv")
```

## 2. Explain score scale and audit detected genes

The UCell score is a normalized rank statistic. With six positive genes and
`maxRank=400`, a nucleus with just one gene ranked first and all other genes
beyond the cap has an idealized score of about 0.168; two genes ranked first
and second give about 0.335. Detection of one or two of six genes therefore
leaves little room for a large value. The small list is not itself a defect;
the low *observed* detection and rank cap determine the scale here. These
idealized values are explanatory bounds, not thresholds for calling ISCs.

```{r score-diagnostic}
CORE6 <- c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Rnf43")
DETECTED4 <- c("Lgr5", "Smoc2", "Lrig1", "Rnf43")
need(all(CORE6 %in% rownames(rna)),
     "A prespecified ISC marker is missing from the RNA assay.")
score4 <- UCell::ScoreSignatures_UCell(rna,
  features = list(ISC_detected4 = DETECTED4), maxRank = UCELL_MAX_RANK)
score4_600 <- UCell::ScoreSignatures_UCell(rna,
  features = list(ISC_detected4 = DETECTED4), maxRank = UCELL_SENS_RANK)
need(setequal(rownames(score4), meta$cell_barcode) &&
     setequal(rownames(score4_600), meta$cell_barcode) &&
     "ISC_detected4_UCell" %in% colnames(score4) &&
     "ISC_detected4_UCell" %in% colnames(score4_600),
     "Unexpected four-gene UCell output shape/names.")
meta$ISC_detected4_UCell <- score4[meta$cell_barcode, "ISC_detected4_UCell"]
meta$ISC_detected4_rank600 <- score4_600[meta$cell_barcode, "ISC_detected4_UCell"]
meta$ISC_detected4_genes_detected <- Matrix::colSums(
  rna[DETECTED4, , drop = FALSE] > 0)
meta$ISC_core_genes_detected_recount <- Matrix::colSums(
  rna[CORE6, , drop = FALSE] > 0)
need(identical(as.numeric(meta$ISC_core_genes_detected),
               as.numeric(meta$ISC_core_genes_detected_recount)),
     "08F six-gene detection counts differ from reloaded RNA counts.")
ideal_score <- function(k, n, cap) {
  rank_sum <- k * (k + 1) / 2 + (n - k) * cap
  1 - (rank_sum - n * (n + 1) / 2) / (n * cap - n * (n + 1) / 2)
}
write_table(tibble(detected_genes = 0:6, signature_size = 6L,
  maxRank = UCELL_MAX_RANK,
  idealized_score_if_detected_genes_rank_first =
    ideal_score(0:6, 6, UCELL_MAX_RANK)), OUT,
  "01_UCell_score_scale_explanation.csv")
marker_detection <- dplyr::bind_rows(lapply(CORE6, function(g) {
  detected <- as.numeric(rna[g, meta$cell_barcode]) > 0
  meta |>
    dplyr::mutate(marker_detected = .env$detected) |>
    dplyr::group_by(parent_wnn_cluster, focused_cluster, diet) |>
    dplyr::summarise(gene = g, nuclei = dplyr::n(),
      detected_n = sum(marker_detected), detected_percent = 100 * mean(marker_detected),
      .groups = "drop")
}))
write_table(marker_detection, OUT, "01_core_marker_detection_by_state_diet.csv")
score_audit <- meta |>
  dplyr::group_by(parent_wnn_cluster, focused_cluster, diet,
    ISC_core_genes_detected) |>
  dplyr::summarise(nuclei = dplyr::n(),
    median_core6 = median(ISC_core6_UCell),
    median_detected4 = median(ISC_detected4_UCell),
    median_RNA_features = median(nFeature_RNA), .groups = "drop")
write_table(score_audit, OUT, "01_score_by_detected_genes_and_state.csv")
p_score <- ggplot2::ggplot(meta, ggplot2::aes(
  x = ISC_core_genes_detected, y = ISC_core6_UCell, color = diet)) +
  ggplot2::geom_jitter(width = .15, height = 0, alpha = .22, size = .4) +
  ggplot2::facet_grid(~parent_wnn_cluster) +
  ggplot2::scale_color_manual(values = c(CON = "#E69F00", HFD = "#56B4E9")) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Low UCell scores track sparse ISC marker detection",
    x = "Six-gene signature members detected per nucleus",
    y = "Step 08F six-gene UCell score")
save_vector_plot(p_score, PLOTS, "01_score_vs_detected_ISC_genes.pdf", 12, 6)
```

## 3. Select only within C0/W0; preserve all 05F identities

No diet or ATAC information enters the ranking. Gate percentages are computed
over all C0/W0 nuclei together, not separately for CON and HFD. The four-gene
and rank-600 scores never determine primary group membership.

```{r C0W0-gates}
base_ix <- which(meta$focused_cluster == "0" & meta$parent_wnn_cluster == "0")
need(length(base_ix) > 100L && all(c("CON", "HFD") %in% meta$diet[base_ix]),
     "The expected C0/W0 source group is absent.")
ranked <- base_ix[order(-meta$ISC_core6_UCell[base_ix],
                        meta$cell_barcode[base_ix])]
meta$C0W0_rank <- NA_integer_
meta$C0W0_rank[ranked] <- seq_along(ranked)
ranked4 <- base_ix[order(-meta$ISC_detected4_UCell[base_ix],
                         meta$cell_barcode[base_ix])]
ranked600 <- base_ix[order(-meta$ISC_core6_rank600[base_ix],
                           meta$cell_barcode[base_ix])]
for (g in names(GATES)) {
  target <- ceiling(GATES[[g]] * length(base_ix))
  meta[[paste0("C0W0_", g)]] <- !is.na(meta$C0W0_rank) &
    meta$C0W0_rank <= target & meta$ISC_core6_UCell > 0
}
need(all(!meta$C0W0_top05 | meta$C0W0_top10) &&
     all(!meta$C0W0_top10 | meta$C0W0_top15) &&
     all(!meta$C0W0_top15 | meta$C0W0_top20) &&
     all(!meta$C0W0_top20 | meta$C0W0_top25),
     "C0/W0 gates must be nested.")
gate_audit <- dplyr::bind_rows(lapply(names(GATES), function(g) {
  flag <- meta[[paste0("C0W0_", g)]]
  sensitivity4 <- seq_along(meta$cell_barcode) %in%
    ranked4[seq_len(ceiling(GATES[[g]] * length(base_ix)))]
  sensitivity600 <- seq_along(meta$cell_barcode) %in%
    ranked600[seq_len(ceiling(GATES[[g]] * length(base_ix)))]
  tibble(gate = g, target_n = ceiling(GATES[[g]] * length(base_ix)),
    selected_n = sum(flag), CON_n = sum(flag & meta$diet == "CON"),
    HFD_n = sum(flag & meta$diet == "HFD"),
    minimum_primary_score = if (any(flag)) min(meta$ISC_core6_UCell[flag]) else NA_real_,
    median_detected_six = if (any(flag)) median(meta$ISC_core_genes_detected[flag]) else NA_real_,
    overlap_detected4 = sum(flag & sensitivity4),
    overlap_rank600 = sum(flag & sensitivity600),
    jaccard_detected4 = sum(flag & sensitivity4) /
      sum(flag | sensitivity4),
    jaccard_rank600 = sum(flag & sensitivity600) /
      sum(flag | sensitivity600))
}))
write_table(gate_audit, OUT, "02_C0W0_gate_size_and_score_sensitivity.csv")
write_table(meta |>
  dplyr::select(cell_barcode, sample_id, diet, parent_wnn_cluster,
    focused_cluster, working_label, parent_annotation_confidence,
    focused_annotation_confidence, Phase, nFeature_RNA, nCount_RNA,
    nCount_ATAC, ISC_core6_UCell, ISC_core6_rank600,
    ISC_detected4_UCell, ISC_detected4_rank600,
    ISC_core_genes_detected, ISC_detected4_genes_detected, C0W0_rank,
    dplyr::starts_with("C0W0_top")), OUT,
  "02_all_nuclei_fixed_labels_and_C0W0_gates.csv.gz")
gate_identity <- dplyr::bind_rows(lapply(names(GATES), function(g) {
  meta[meta[[paste0("C0W0_", g)]], , drop = FALSE] |>
    dplyr::group_by(diet) |>
    dplyr::summarise(gate = g, nuclei = dplyr::n(),
      median_six_score = median(ISC_core6_UCell),
      median_detected4_score = median(ISC_detected4_UCell),
      median_six_genes_detected = median(ISC_core_genes_detected),
      median_ATAC_counts = median(nCount_ATAC),
      median_RNA_features = median(nFeature_RNA), .groups = "drop")
}))
write_table(gate_identity, OUT, "02_C0W0_selected_identity_and_depth_by_diet.csv")
```

## 4. Fixed comparisons and pooled descriptive effects

Each contrast needs at least 30 nuclei in *each* diet. The same Step 08E
`pooled_effect()` and consensus peaks are used throughout. An ineligible
comparison has missing effects, not zero changed peaks. The W6/C2 and
W8/C3+C5 context groups remain independent of all C0 analyses.

```{r contrasts-and-effects}
make_contrast <- function(key, role_name, ix) {
  tibble(contrast = key, role = role_name, indices = list(as.integer(ix)))
}
spec <- list(
  make_contrast("C0_all", "ISC_enriched_state", which(meta$focused_cluster == "0")),
  make_contrast("C2_all", "crypt_progenitor_state", which(meta$focused_cluster == "2")),
  make_contrast("C0_C2_proxy", "broad_Gourab_proxy",
    which(meta$focused_cluster %in% c("0", "2"))),
  make_contrast("C0W0_all", "parent_restricted_C0", base_ix))
for (g in names(GATES)) spec[[length(spec) + 1L]] <- make_contrast(
  paste0("C0W0_", g), if (g %in% names(PRIMARY_GATES))
    "primary_C0W0_score_gate" else "sensitivity_C0W0_score_gate",
  which(meta[[paste0("C0W0_", g)]]))
spec[[length(spec) + 1L]] <- make_contrast("W6_C2_context", "separate_W6_context",
  which(meta$parent_wnn_cluster == "6" & meta$focused_cluster == "2"))
spec[[length(spec) + 1L]] <- make_contrast("W8_C3_C5_context", "separate_W8_context",
  which(meta$parent_wnn_cluster == "8" & meta$focused_cluster %in% c("3", "5")))
contrast <- dplyr::bind_rows(spec)
contrast$n_CON <- vapply(contrast$indices, function(ix)
  as.integer(sum(meta$diet[ix] == "CON")), integer(1))
contrast$n_HFD <- vapply(contrast$indices, function(ix)
  as.integer(sum(meta$diet[ix] == "HFD")), integer(1))
contrast$eligible <- contrast$n_CON >= MIN_PER_DIET & contrast$n_HFD >= MIN_PER_DIET
need(!anyDuplicated(contrast$contrast) &&
     all(contrast$eligible[contrast$contrast %in%
       c("C0_all", "C2_all", "C0_C2_proxy", "C0W0_top15", "C0W0_top10")]),
     "A primary contrast is duplicated or has too few nuclei per diet.")
write_table(dplyr::select(contrast, -indices), OUT,
  "03_contrast_definitions_and_diet_balance.csv")
all_effects <- list()
effect_summary <- list()
for (modality in c("RNA", "ATAC")) {
  counts <- if (modality == "RNA") rna else atac
  depth <- Matrix::colSums(counts)
  effects <- dplyr::bind_rows(lapply(which(contrast$eligible), function(i) {
    contrast_key <- contrast$contrast[[i]]
    role_name <- contrast$role[[i]]
    message("08G ", modality, ": ", contrast_key, " (",
      contrast$n_CON[i], " CON / ", contrast$n_HFD[i], " HFD)")
    pooled_effect(counts, meta, contrast$indices[[i]], modality,
      total_depth = depth) |>
      dplyr::mutate(contrast = .env$contrast_key, role = .env$role_name,
                    .before = 1)
  }))
  need(!anyDuplicated(effects[c("contrast", "feature")]),
       paste("Duplicate", modality, "feature per contrast."))
  effect_summary[[modality]] <- effects |>
    dplyr::group_by(contrast, role, modality) |>
    dplyr::summarise(consensus_features = dplyr::n(),
      evaluable_features = sum(evaluable), clear_features = sum(clear_effect),
      HFD_higher_clear = sum(clear_effect & log2FC_HFD_vs_CON > 0),
      CON_higher_clear = sum(clear_effect & log2FC_HFD_vs_CON < 0),
      .groups = "drop")
  write_table(dplyr::filter(effects, evaluable), OUT,
    paste0("04_", modality, "_evaluable_pooled_effects.csv.gz"))
  all_effects[[modality]] <- effects
  rm(effects); invisible(gc())
}
write_table(dplyr::bind_rows(effect_summary) |>
  dplyr::left_join(dplyr::select(contrast, contrast, n_CON, n_HFD, eligible),
    by = "contrast", relationship = "many-to-one"), OUT,
  "04_descriptive_effect_counts.csv")
```

## 5. Compare C0, C2, and the score gates at the same consensus peaks

The 15% C0/W0 peak set is a **descriptive anchor**. Each row retains the
original contrast values, detection, and count support. The tiers summarize
whether the same directional clear effect also appears in C0 overall and/or
the 10% gate; they are not P values or evidence from independent samples.
The C0 and C2 comparison is tabulated separately so that their distinct
responses remain visible even if the C0+C2 proxy averages them together.

```{r compare-peak-effects}
atac_effects <- all_effects$ATAC
rna_effects <- all_effects$RNA
feature_effect <- function(key) {
  atac_effects |>
    dplyr::filter(contrast == .env$key) |>
    dplyr::select(feature, evaluable, clear_effect,
      log2FC_HFD_vs_CON, CON_percent, HFD_percent, CON_counts, HFD_counts)
}
comparators <- c("C0_all", "C2_all", "C0_C2_proxy", "C0W0_top10",
  "C0W0_top05", "C0W0_top20", "C0W0_top25",
  "W6_C2_context", "W8_C3_C5_context")
candidate <- feature_effect("C0W0_top15") |>
  dplyr::filter(clear_effect)
names(candidate)[-1] <- paste0("anchor_", names(candidate)[-1])
for (key in comparators) {
  comparison <- feature_effect(key)
  names(comparison)[-1] <- paste0(key, "_", names(comparison)[-1])
  candidate <- dplyr::left_join(candidate, comparison, by = "feature",
    relationship = "one-to-one")
}
agrees <- function(key) {
  dplyr::coalesce(candidate[[paste0(key, "_clear_effect")]] &
    sign(candidate[[paste0(key, "_log2FC_HFD_vs_CON")]]) ==
      sign(candidate$anchor_log2FC_HFD_vs_CON), FALSE)
}
candidate$C0_same <- agrees("C0_all")
candidate$C2_same <- agrees("C2_all")
candidate$top10_same <- agrees("C0W0_top10")
candidate$proxy_same <- agrees("C0_C2_proxy")
candidate$tier <- dplyr::case_when(
  candidate$C0_same & candidate$top10_same ~ "C0_and_top10",
  candidate$C0_same ~ "C0_only",
  candidate$top10_same ~ "top10_only",
  TRUE ~ "anchor_only")
candidate$direction <- ifelse(candidate$anchor_log2FC_HFD_vs_CON > 0,
  "HFD_higher", "CON_higher")
write_table(candidate, OUT, "05_C0W0_top15_peak_candidates_with_all_contrasts.csv.gz")
candidate_summary <- candidate |>
  dplyr::group_by(tier, direction) |>
  dplyr::summarise(candidate_peaks = dplyr::n(),
    also_C2_same = sum(C2_same),
    also_proxy_same = sum(proxy_same),
    .groups = "drop")
write_table(candidate_summary, OUT, "05_candidate_peak_support_tiers.csv")
c0_vs_c2 <- feature_effect("C0_all") |>
  dplyr::rename_with(~ paste0("C0_", .x), -feature) |>
  dplyr::left_join(feature_effect("C2_all") |>
    dplyr::rename_with(~ paste0("C2_", .x), -feature),
    by = "feature", relationship = "one-to-one") |>
  dplyr::mutate(C0_C2_both_evaluable = C0_evaluable & C2_evaluable,
    both_clear_same_direction = C0_clear_effect & C2_clear_effect &
      sign(C0_log2FC_HFD_vs_CON) == sign(C2_log2FC_HFD_vs_CON),
    both_clear_opposite_direction = C0_clear_effect & C2_clear_effect &
      sign(C0_log2FC_HFD_vs_CON) != sign(C2_log2FC_HFD_vs_CON))
write_table(dplyr::filter(c0_vs_c2, C0_clear_effect | C2_clear_effect),
  OUT, "05_C0_C2_at_least_one_clear_peak.csv.gz")
write_table(tibble(
  consensus_peaks = nrow(c0_vs_c2),
  C0_evaluable = sum(c0_vs_c2$C0_evaluable),
  C2_evaluable = sum(c0_vs_c2$C2_evaluable),
  both_evaluable = sum(c0_vs_c2$C0_C2_both_evaluable),
  C0_clear = sum(c0_vs_c2$C0_clear_effect),
  C2_clear = sum(c0_vs_c2$C2_clear_effect),
  both_clear_same_direction = sum(c0_vs_c2$both_clear_same_direction),
  both_clear_opposite_direction = sum(c0_vs_c2$both_clear_opposite_direction)
), OUT, "05_C0_vs_C2_peak_overlap_summary.csv")
peak_counts <- dplyr::bind_rows(effect_summary) |>
  dplyr::filter(modality == "ATAC") |>
  dplyr::left_join(dplyr::select(contrast, contrast, n_CON, n_HFD),
    by = "contrast", relationship = "many-to-one")
write_table(peak_counts, OUT, "05_ATAC_peak_counts_and_diet_balance.csv")
p_counts <- ggplot2::ggplot(peak_counts, ggplot2::aes(
  x = stats::reorder(contrast, evaluable_features), y = evaluable_features,
  fill = clear_features / pmax(evaluable_features, 1))) +
  ggplot2::geom_col() + ggplot2::coord_flip() +
  ggplot2::scale_fill_viridis_c(labels = scales::percent,
    name = "Clear/evaluable") +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Evaluable consensus peaks by fixed population",
    subtitle = "Color shows descriptive clear effects among evaluable peaks; each group needs diet balance",
    x = NULL, y = "Evaluable ATAC peaks")
save_vector_plot(p_counts, PLOTS, "05_evaluable_peak_counts.pdf", 12, 7)
```

## 6. Cell number and depth sensitivity at selected peak candidates

Repeat the same sampling methods used in 08F for the top 200 evaluable
features of each eligible primary contrast. These draws estimate stability
against cell balance and fragment depth; they do not create mouse replicates.

```{r depth-sensitivity}
resample_keys <- intersect(c("C0_all", "C2_all", "C0_C2_proxy",
  "C0W0_top15", "C0W0_top10", "C0W0_top05"),
  contrast$contrast[contrast$eligible])
resampling <- dplyr::bind_rows(lapply(resample_keys, function(key) {
  i <- match(key, contrast$contrast)
  full <- atac_effects |>
    dplyr::filter(contrast == .env$key, evaluable) |>
    dplyr::arrange(dplyr::desc(abs(log2FC_HFD_vs_CON)))
  picked <- head(full$feature, 200L)
  if (!length(picked)) return(tibble::tibble())
  resampling_effects(atac, meta, contrast$indices[[i]], full, picked,
    "ATAC", iterations = N_DRAWS) |>
    dplyr::mutate(contrast = .env$key, .before = 1)
}))
write_table(resampling, OUT, "06_top200_ATAC_equal_cell_and_depth_sensitivity.csv.gz")
candidate <- candidate |>
  dplyr::left_join(resampling |>
    dplyr::filter(contrast == "C0W0_top15", method == "depth_matched") |>
    dplyr::select(feature, depth_matched_valid_draws = valid_iterations,
      depth_matched_sign_agreement_percent = sign_agreement_percent,
      depth_matched_median_log2FC = median_log2FC),
    by = "feature", relationship = "one-to-one")
write_table(candidate, OUT, "06_C0W0_top15_candidates_with_depth_check.csv.gz")
```

## 7. Promoter and expression context after peak selection

The source program is joined **after** the peak list and groups are fixed.
The promoter map uses annotated gene TSS and a symmetric 2-kb window. A
nearby gene and a concordant RNA effect are hypotheses for follow-up, not
validated peak-to-gene links. The example plots require observed RNA and
prefer candidates retained in C0 and the 10% score gate.

```{r promoter-rna-context}
promoter <- promoter_peak_map(obj, window = 2000L)$map |>
  dplyr::distinct(gene, peak, .keep_all = TRUE)
rna_anchor <- rna_effects |>
  dplyr::filter(contrast == "C0W0_top15") |>
  dplyr::transmute(gene = feature, RNA_evaluable = evaluable,
    RNA_clear = clear_effect, RNA_log2FC = log2FC_HFD_vs_CON,
    RNA_CON_percent = CON_percent, RNA_HFD_percent = HFD_percent,
    RNA_CON_counts = CON_counts, RNA_HFD_counts = HFD_counts)
gene_context <- candidate |>
  dplyr::inner_join(promoter, by = c("feature" = "peak"),
    relationship = "many-to-many") |>
  dplyr::left_join(rna_anchor, by = "gene", relationship = "many-to-one")
source_genes <- read_csv_checked(SOURCE_GENES)
require_columns(source_genes, c("gene", "program"), "08F Gourab audit")
source_genes <- dplyr::distinct(source_genes, gene, .keep_all = TRUE)
gene_context <- dplyr::left_join(gene_context, source_genes,
  by = "gene", relationship = "many-to-one") |>
  dplyr::mutate(RNA_ATAC_same_direction = dplyr::coalesce(
    RNA_evaluable & sign(RNA_log2FC) == sign(anchor_log2FC_HFD_vs_CON), FALSE))
write_table(gene_context, OUT, "07_peak_promoter_RNA_and_Gourab_overlay.csv.gz")
examples <- gene_context |>
  dplyr::filter(RNA_evaluable, gene %in% rownames(rna_data)) |>
  dplyr::mutate(tier_priority = dplyr::case_when(
    tier == "C0_and_top10" ~ 3L, tier == "C0_only" ~ 2L,
    tier == "top10_only" ~ 1L, TRUE ~ 0L)) |>
  dplyr::arrange(dplyr::desc(tier_priority),
    dplyr::desc(RNA_ATAC_same_direction),
    dplyr::desc(abs(anchor_log2FC_HFD_vs_CON))) |>
  dplyr::distinct(feature, .keep_all = TRUE) |>
  dplyr::slice_head(n = MAX_COVERAGE_LOCI)
write_table(examples, OUT, "07_example_loci_selected_without_Gourab_overlay.csv")
if (nrow(examples)) {
  effect_plot <- dplyr::bind_rows(
    dplyr::transmute(examples, gene, feature, modality = "ATAC",
      effect = anchor_log2FC_HFD_vs_CON),
    dplyr::transmute(examples, gene, feature, modality = "RNA",
      effect = RNA_log2FC))
  p_effect <- ggplot2::ggplot(effect_plot, ggplot2::aes(
      x = interaction(gene, feature), y = effect, fill = modality)) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::geom_hline(yintercept = 0, linewidth = .25) +
    ggplot2::coord_flip() + ggplot2::theme_bw(base_size = 11) +
    ggplot2::labs(title = "Observed RNA and nearby promoter ATAC effects",
      subtitle = "Descriptive pooled log2 CPM ratios; annotation implies proximity only",
      x = NULL, y = "HFD versus CON pooled log2 fold change")
  save_vector_plot(p_effect, PLOTS, "07_examples_RNA_and_ATAC_effects.pdf", 12, 6)
  violin_keys <- c("C0_all", "C0W0_top15")
  violin_data <- dplyr::bind_rows(lapply(unique(examples$gene), function(gene_name) {
    dplyr::bind_rows(lapply(violin_keys, function(key) {
      ix <- contrast$indices[[match(key, contrast$contrast)]]
      tibble(gene = gene_name, group = key, diet = meta$diet[ix],
        RNA_normalized = as.numeric(rna_data[gene_name, ix]))
    }))
  }))
  write_table(violin_data, OUT, "07_example_gene_per_nucleus_normalized_RNA.csv.gz")
  p_violin <- ggplot2::ggplot(violin_data,
      ggplot2::aes(x = diet, y = RNA_normalized, fill = diet)) +
    ggplot2::geom_violin(scale = "width", trim = TRUE, na.rm = TRUE) +
    ggplot2::geom_boxplot(width = .12, outlier.shape = NA, alpha = .6) +
    ggplot2::facet_grid(gene ~ group, scales = "free_y") +
    ggplot2::scale_fill_manual(values = c(CON = "#E69F00", HFD = "#56B4E9")) +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::labs(title = "Example genes in C0 overall and the C0/W0 score gate",
      subtitle = "Individual nuclei show the distribution; pooled libraries determine descriptive effects",
      x = NULL, y = "Seurat normalized RNA expression")
  save_vector_plot(p_violin, PLOTS, "07_example_gene_RNA_violins.pdf", 11, 9)
}
```

## 8. Coverage views at mapped genomic coordinates

The global WNN view shows the nested score-defined C0/W0 cells in their
original embedding while leaving all unused nuclei grey.

Pass a genomic range to Signac instead of a gene name. This avoids the
`Gene not found` warnings encountered when a region string was combined
with `annotation="gene"` in 08F. If fragment references are missing or
an individual locus fails, save the status and continue.

```{r mapped-coverage}
global <- readRDS(ANNOTATED)
need("wnn.umap" %in% names(global@reductions),
     "Step 06 global WNN UMAP is absent.")
coords <- as.data.frame(SeuratObject::Embeddings(global, "wnn.umap"))
names(coords)[1:2] <- c("x", "y")
coords$cell_barcode <- rownames(coords)
fg <- dplyr::left_join(coords, dplyr::select(meta, cell_barcode,
  focused_cluster, parent_wnn_cluster, C0W0_top15, C0W0_top10),
  by = "cell_barcode", relationship = "one-to-one")
need(nrow(fg) == nrow(coords) &&
     sum(!is.na(fg$focused_cluster)) == nrow(meta),
     "Global WNN and focused barcode sets are misaligned.")
fg$highlight <- dplyr::case_when(
  fg$C0W0_top10 %in% TRUE ~ "C0/W0 top 10%",
  fg$C0W0_top15 %in% TRUE ~ "C0/W0 10–15%",
  fg$focused_cluster == "0" ~ "Other C0",
  fg$focused_cluster == "2" ~ "C2",
  TRUE ~ "Other nuclei")
fg$highlight <- factor(fg$highlight, levels = c("Other nuclei", "C2",
  "Other C0", "C0/W0 10–15%", "C0/W0 top 10%"))
p_wnn <- ggplot2::ggplot(fg, ggplot2::aes(x, y)) +
  ggplot2::geom_point(data = dplyr::filter(fg,
    highlight == "Other nuclei"), color = "grey87", size = .10) +
  ggplot2::geom_point(data = dplyr::filter(fg,
    highlight != "Other nuclei"), ggplot2::aes(color = highlight),
    size = .35, alpha = .85) +
  ggplot2::scale_color_manual(values = c("C2" = "#D18C3A",
    "Other C0" = "#59B2A0", "C0/W0 10–15%" = "#3278AB",
    "C0/W0 top 10%" = "#143D74"), drop = TRUE) +
  ggplot2::coord_equal() + ggplot2::theme_classic(base_size = 11) +
  ggplot2::labs(title = "C0, C2, and score-defined C0/W0 on the global WNN UMAP",
    subtitle = "Fixed focused subclusters; C0/W0 gates ranked by six-gene UCell across both diets",
    x = "Global WNN UMAP 1", y = "Global WNN UMAP 2", color = NULL)
save_vector_plot(p_wnn, PLOTS, "08_C0_C2_and_C0W0_on_global_WNN.pdf", 12, 7)
coverage_status <- tidyr::expand_grid(
  contrast = c("C0_all", "C0W0_top15"), feature = examples$feature) |>
  dplyr::mutate(status = "not_attempted", detail = "")
if (RUN_COVERAGE && nrow(coverage_status)) {
  if (!length(Signac::Fragments(global[["ATAC"]]))) {
    coverage_status$status <- "skipped_no_fragments"
    coverage_status$detail <- "Step 06 ATAC fragment references are unavailable"
  } else {
    peaks <- GenomicRanges::granges(obj[["ATAC.crypt"]])
    names(peaks) <- rownames(atac)
    for (key in unique(coverage_status$contrast)) {
      ix <- contrast$indices[[match(key, contrast$contrast)]]
      coverage_obj <- subset(global, cells = meta$cell_barcode[ix])
      SeuratObject::DefaultAssay(coverage_obj) <- "ATAC"
      for (j in seq_len(nrow(examples))) {
        peak_name <- examples$feature[[j]]
        k <- which(coverage_status$contrast == key &
          coverage_status$feature == peak_name)
        gr <- peaks[match(peak_name, names(peaks))]
        span <- GenomicRanges::GRanges(
          seqnames = as.character(GenomicRanges::seqnames(gr)),
          ranges = IRanges::IRanges(
            start = max(1L, as.integer(BiocGenerics::start(gr)) - 5000L),
            end = as.integer(BiocGenerics::end(gr)) + 5000L))
        tryCatch({
          p <- Signac::CoveragePlot(coverage_obj, region = span,
            assay = "ATAC", annotation = "gene", peaks = TRUE,
            region.highlight = gr, links = FALSE, tile = FALSE,
            group.by = "diet", ymax = "q95") +
            patchwork::plot_annotation(title = paste(key, peak_name),
              subtitle = paste("Promoter-associated", examples$gene[[j]],
                "; pooled depth-normalized CON/HFD coverage"))
          ggplot2::ggsave(file.path(PLOTS, paste0("08_", key, "_example_", j,
            ".pdf")), plot = p, device = grDevices::pdf,
            useDingbats = FALSE, width = 13, height = 6)
          coverage_status$status[k] <- "saved"
        }, error = function(e) {
          coverage_status$status[k] <<- "plot_failed"
          coverage_status$detail[k] <<- conditionMessage(e)
          warning("Coverage view failed for ", key, " / ", peak_name,
            ": ", conditionMessage(e))
        })
      }
      rm(coverage_obj); invisible(gc())
    }
  }
}
rm(global); invisible(gc())
write_table(coverage_status, OUT, "08_example_coverage_status.csv")
```

## 9. Coordinate checked BED inputs for a later motif analysis

Write only the robust descriptive tier (shared by C0 and C0/W0 10%) in
separate HFD-higher and CON-higher sets. The entire evaluable C0/W0 15%
universe is recorded for background design, but is **not** a matched
background. Motif enrichment needs a length, GC, mappability, and detection
matched background and explicit motif choices in a separate analysis.

```{r motif-inputs-and-finish}
peak_ranges <- GenomicRanges::granges(obj[["ATAC.crypt"]])
need(identical(rownames(atac), rownames(obj[["ATAC.crypt"]])) &&
     length(peak_ranges) == nrow(atac), "ATAC ranges/features changed.")
names(peak_ranges) <- rownames(atac)
bed_for <- function(features, filename) {
  gr <- peak_ranges[match(features, names(peak_ranges))]
  need(!anyNA(match(features, names(peak_ranges))),
       paste("Peak coordinates missing for", filename))
  bed <- tibble(chrom = as.character(GenomicRanges::seqnames(gr)),
    start0 = as.integer(BiocGenerics::start(gr)) - 1L,
    end = as.integer(BiocGenerics::end(gr)), feature = features)
  readr::write_tsv(bed, file.path(OUT, filename), col_names = FALSE)
  nrow(bed)
}
motif_inputs <- tibble(
  set = c("C0_and_top10_HFD_higher", "C0_and_top10_CON_higher",
    "C0W0_top15_evaluable_unmatched_universe"),
  file = c("09_HFD_higher_candidate_regions.bed",
    "09_CON_higher_candidate_regions.bed",
    "09_evaluable_unmatched_universe.bed"))
anchor_universe <- feature_effect("C0W0_top15") |>
  dplyr::filter(evaluable)
motif_inputs$n_regions <- c(
  bed_for(candidate$feature[candidate$tier == "C0_and_top10" &
    candidate$direction == "HFD_higher"], motif_inputs$file[[1]]),
  bed_for(candidate$feature[candidate$tier == "C0_and_top10" &
    candidate$direction == "CON_higher"], motif_inputs$file[[2]]),
  bed_for(anchor_universe$feature, motif_inputs$file[[3]]))
write_table(motif_inputs, OUT, "09_motif_input_manifest_not_enrichment.csv")
finish_run(OUT, "STEP_08G_COMPLETE.txt", c(
  "C0, C2, C0+C2, C0/W0 all, C0/W0 score gates and separate W6/W8 contexts audited.",
  "Primary C0/W0 gates: 15% and 10%; 5%, 20%, 25% are sensitivity gates.",
  "One pooled library per diet: effects and resampling are descriptive, not replicate-level DAR tests.",
  "BED sets are candidate and unmatched-universe inputs; motif enrichment was not performed.",
  paste("Results:", OUT)))
```
