---
title: "Step 08H | WT ISC-enriched candidate regions and sex-marker audit"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This follows the completed Step 08G. C0 is the **broader ISC-enriched crypt
state** defined by the independent focused clustering and 05F marker review.
C0/W0 top 15% is a **prespecified marker-enriched sensitivity population**:
the six-gene UCell score ranks C0 nuclei within W0 over both diets together.
The score was set before evaluating these ATAC peaks. Neither the four
gene sensitivity score nor Gourab's programs selects nuclei or peaks here.
Neither group is a purified ISC population. C2 remains a separate crypt
progenitor comparison; W6 and W8 are not silently pooled with C0.

Step 08G has one pooled WT library per diet. All output is descriptive:
"clear effect" is the prior count/detection and absolute log2FC threshold,
not a statistically significant DAR. Retaining an effect in nested groups
uses overlapping nuclei and does not provide independent validation.

This notebook preserves **every one of the 622** Step 08G top-15 clear
peaks with evidence fields. The 37 with a same-direction clear effect in C0
overall form the focused review list. Six of those also have a same-direction
clear effect at top 10%. A count-support flag (at least 40 pooled fragments
in top 15%, with at least 10 in each diet) is a sensitivity view, not an
optimized discovery cutoff. The 20% and 25% effects are reported as
continuous values and separate threshold flags. No RNA or nearest-gene
behavior decides which peak enters the list.

`Xist` and Y-linked transcript detection are **imperfect sex proxies** in
sparse snRNA. The notebook audits composition and conditionally recalculates
candidate ATAC effects within Xist-detected and Xist-undetected nuclei when
both diets have at least 30 nuclei. An Xist-undetected nucleus is not
necessarily male. These analyses do not recover individual mouse identities
or remove pooled-library sex confounding. Sex and depth findings should be
reviewed before treating the focused peak list as ready for a genotype
comparison.

Run interactively from the repository root in the same R 4.4 environment as
08G. Step 08G outputs are not modified. All coordinates use the Step 03
GRCm39 common peak intervals; compare the **exact intervals and same
directional effect definition** in PPAR and IL17 after independently
adjudicating their ISC-like states. No genotype interaction test is possible
from one pooled library per diet per cohort.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Open the repository root before running 08H.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced", MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(HELPER)) HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(HELPER)) stop("Missing WT crypt follow-up helper.")
source(HELPER)
suppressPackageStartupMessages({
  library(Seurat); library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges)
})
need(identical(DATASET_VERSION, "v2_resequenced") && identical(COHORT_ID, "wt"),
     "08H expects the WT v2_resequenced cohort.")
STEP5E <- file.path(OUTPUT_DIR, "05_wnn", "focused_wt_crypt_stem_progenitor")
STEP8G <- file.path(OUTPUT_DIR, "08g_C0_C2_and_C0W0_ISC")
OBJECT <- file.path(STEP5E, "objects", "WT_crypt_core_0_6_8_focused.rds")
CELLS <- file.path(STEP8G, "02_all_nuclei_fixed_labels_and_C0W0_gates.csv.gz")
PEAKS <- file.path(STEP8G, "06_C0W0_top15_candidates_with_depth_check.csv.gz")
OUT <- file.path(OUTPUT_DIR, "08h_wt_ISC_region_freeze_sex_audit")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
inputs <- c(OBJECT, CELLS, PEAKS, SAMPLE_SHEET_PATH,
  file.path(STEP8G, "STEP_08G_COMPLETE.txt"))
for (path in inputs) need(file.exists(path), paste("Missing input:", path))
capture_manifest(inputs, OUT)
```

## 1. Validate cells, assays, and recorded diet balance

```{r load-and-validate}
focused <- load_focused(OBJECT)
obj <- focused$object
rna <- focused$matrices$RNA_counts
atac <- focused$matrices$ATAC_counts
meta <- focused$metadata
recorded <- read_csv_checked(CELLS)
require_columns(recorded, c("cell_barcode", "diet", "parent_wnn_cluster",
  "focused_cluster", "ISC_core6_UCell", "ISC_core_genes_detected",
  "C0W0_top15", "C0W0_top10", "C0W0_top20", "C0W0_top25"),
  "08G nucleus table")
need(!anyDuplicated(recorded$cell_barcode) &&
     setequal(meta$cell_barcode, recorded$cell_barcode),
     "Step 08G nuclei differ from the saved focused object.")
recorded <- recorded[match(meta$cell_barcode, recorded$cell_barcode), ]
need(identical(meta$cell_barcode, recorded$cell_barcode) &&
     identical(meta$cell_barcode, colnames(rna)) &&
     identical(meta$cell_barcode, colnames(atac)) &&
     identical(meta$diet, recorded$diet) &&
     identical(meta$focused_cluster, as.character(recorded$focused_cluster)) &&
     identical(meta$parent_wnn_cluster,
       as.character(recorded$parent_wnn_cluster)),
     "Focused metadata, 08G gates and matrices are not aligned.")
for (field in c("ISC_core6_UCell", "ISC_core_genes_detected",
                "C0W0_top15", "C0W0_top10", "C0W0_top20", "C0W0_top25"))
  meta[[field]] <- recorded[[field]]
need(all(meta$C0W0_top10 <= meta$C0W0_top15) &&
     all(meta$C0W0_top15 <= meta$C0W0_top20) &&
     all(meta$C0W0_top20 <= meta$C0W0_top25) &&
     all(meta$focused_cluster[meta$C0W0_top15] == "0") &&
     all(meta$parent_wnn_cluster[meta$C0W0_top15] == "0"),
     "Recorded score gates are not nested C0/W0 sets.")
sample_sheet <- read_csv_checked(SAMPLE_SHEET_PATH)
require_columns(sample_sheet, c("cohort_id", "sample_id", "diet",
  "mice_pooled", "female_mice", "male_mice"), "v2 sample sheet")
wt_samples <- dplyr::filter(sample_sheet, cohort_id == "wt") |>
  dplyr::select(sample_id, diet, mice_pooled, female_mice, male_mice)
need(nrow(wt_samples) == 2L &&
     setequal(wt_samples$sample_id, unique(meta$sample_id)) &&
     all(wt_samples$mice_pooled == wt_samples$female_mice +
       wt_samples$male_mice), "WT pooled sample records changed.")
write_table(wt_samples, OUT, "00_WT_pooled_mouse_sex_composition.csv")
write_table(focused$alignment, OUT, "00_matrix_alignment.csv")
```

## 2. Audit sex-linked transcript detection, composition, and depth

Do not classify Xist-negative nuclei as male. A Y-transcript-positive
nucleus is suggestive, but non-detection is common in sparse snRNA.

```{r sex-proxy-audit}
need("Xist" %in% rownames(rna), "Xist absent from focused RNA assay.")
y_genes <- intersect(c("Uty", "Ddx3y", "Eif2s3y", "Kdm5d"), rownames(rna))
need(length(y_genes) > 0L, "None of the four prespecified Y genes is in RNA.")
meta$Xist_counts <- as.numeric(rna["Xist", meta$cell_barcode])
meta$Y_genes_detected <- as.integer(Matrix::colSums(
  rna[y_genes, meta$cell_barcode, drop = FALSE] > 0))
meta$Xist_detected <- meta$Xist_counts > 0
meta$Y_any_detected <- meta$Y_genes_detected > 0
meta$sex_proxy_status <- dplyr::case_when(
  meta$Xist_detected & !meta$Y_any_detected ~ "Xist_only",
  !meta$Xist_detected & meta$Y_any_detected ~ "Y_only",
  meta$Xist_detected & meta$Y_any_detected ~ "both_detected",
  TRUE ~ "neither_detected")
groups <- list(C0_all = meta$focused_cluster == "0",
  C2_all = meta$focused_cluster == "2",
  C0_C2_proxy = meta$focused_cluster %in% c("0", "2"),
  C0W0_all = meta$focused_cluster == "0" & meta$parent_wnn_cluster == "0",
  C0W0_top15 = meta$C0W0_top15, C0W0_top10 = meta$C0W0_top10,
  C0W0_top20 = meta$C0W0_top20, C0W0_top25 = meta$C0W0_top25)
sex_audit <- dplyr::bind_rows(lapply(names(groups), function(key) {
  meta[groups[[key]], , drop = FALSE] |>
    dplyr::group_by(diet) |>
    dplyr::summarise(group = key, nuclei = dplyr::n(),
      Xist_positive = sum(Xist_detected),
      Xist_positive_percent = 100 * mean(Xist_detected),
      Y_positive = sum(Y_any_detected),
      Y_positive_percent = 100 * mean(Y_any_detected),
      both_detected = sum(sex_proxy_status == "both_detected"),
      neither_detected = sum(sex_proxy_status == "neither_detected"),
      median_RNA_features = median(nFeature_RNA),
      median_RNA_counts = median(nCount_RNA),
      median_ATAC_counts = median(nCount_ATAC),
      median_ISC_score = median(ISC_core6_UCell), .groups = "drop")
}))
write_table(sex_audit, OUT, "01_sex_marker_and_depth_balance_by_group_diet.csv")
sex_strata <- dplyr::bind_rows(lapply(names(groups), function(key) {
  meta[groups[[key]], , drop = FALSE] |>
    dplyr::count(diet, sex_proxy_status, name = "nuclei") |>
    dplyr::mutate(group = key, .before = 1)
}))
write_table(sex_strata, OUT, "01_sex_marker_proxy_strata_by_group_diet.csv")
write_table(meta |>
  dplyr::select(cell_barcode, sample_id, diet, focused_cluster,
    parent_wnn_cluster, ISC_core6_UCell, ISC_core_genes_detected,
    Xist_counts, Xist_detected, Y_genes_detected, Y_any_detected,
    sex_proxy_status, nFeature_RNA, nCount_RNA, nCount_ATAC,
    dplyr::starts_with("C0W0_top")), OUT,
  "01_nucleus_level_sex_proxy_and_gate_audit.csv.gz")
p_sex <- ggplot2::ggplot(sex_audit,
  ggplot2::aes(x = group, y = Xist_positive_percent, fill = diet)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = .8)) +
  ggplot2::scale_fill_manual(values = c(CON = "#E69F00", HFD = "#56B4E9")) +
  ggplot2::coord_flip() + ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Xist transcript detection across the fixed WT groups",
    subtitle = "A detection audit, not per-mouse sex assignment or a corrected diet effect",
    x = NULL, y = "Nuclei with Xist detected (%)")
save_vector_plot(p_sex, PLOTS, "01_Xist_detection_by_group_diet.pdf", 11, 7)
```

## 3. Freeze every peak and retain transparent support flags

The complete peak list remains available. `C0_supported` is the descriptive
subset for inspection; `count_supported_40` and `nested_top10` are flags,
not replacements for full count and continuous effect values. All BEDs use
0-based starts and 1-based ends from the canonical peak GRanges.

```{r freeze-candidates}
candidate <- read_csv_checked(PEAKS)
require_columns(candidate, c("feature", "anchor_clear_effect", "anchor_evaluable",
  "anchor_CON_counts", "anchor_HFD_counts", "anchor_CON_percent",
  "anchor_HFD_percent", "anchor_log2FC_HFD_vs_CON", "C0_same",
  "C2_same", "top10_same", "C0_all_log2FC_HFD_vs_CON",
  "C0W0_top20_evaluable", "C0W0_top20_clear_effect",
  "C0W0_top20_log2FC_HFD_vs_CON", "C0W0_top25_evaluable",
  "C0W0_top25_clear_effect", "C0W0_top25_log2FC_HFD_vs_CON"),
  "08G top-15 ATAC candidate table")
need(!anyDuplicated(candidate$feature) &&
     all(candidate$anchor_evaluable & candidate$anchor_clear_effect) &&
     identical(nrow(candidate), 622L),
     "08G top-15 candidate set changed; inspect the new run before reusing this release.")
peak_ranges <- GenomicRanges::granges(obj[["ATAC.crypt"]])
need(identical(rownames(atac), rownames(obj[["ATAC.crypt"]])) &&
     length(peak_ranges) == nrow(atac) &&
     all(candidate$feature %in% rownames(atac)),
     "Candidate peaks do not align to the same GRCm39 consensus ranges.")
names(peak_ranges) <- rownames(atac)
selected_ranges <- peak_ranges[match(candidate$feature, names(peak_ranges))]
candidate$chrom <- as.character(GenomicRanges::seqnames(selected_ranges))
candidate$start_1based <- as.integer(BiocGenerics::start(selected_ranges))
candidate$end_1based <- as.integer(BiocGenerics::end(selected_ranges))
candidate$start_0based_bed <- candidate$start_1based - 1L
candidate$anchor_total_counts <- candidate$anchor_CON_counts +
  candidate$anchor_HFD_counts
candidate$count_supported_40 <- candidate$anchor_total_counts >= 40L &
  pmin(candidate$anchor_CON_counts, candidate$anchor_HFD_counts) >= 10L
candidate$C0_supported <- candidate$C0_same
candidate$nested_top10 <- candidate$C0_supported & candidate$top10_same
same_effect <- function(clear, effect, anchor) {
  dplyr::coalesce(clear & sign(effect) == sign(anchor), FALSE)
}
candidate$top20_clear_same <- same_effect(
  candidate$C0W0_top20_clear_effect,
  candidate$C0W0_top20_log2FC_HFD_vs_CON,
  candidate$anchor_log2FC_HFD_vs_CON)
candidate$top25_clear_same <- same_effect(
  candidate$C0W0_top25_clear_effect,
  candidate$C0W0_top25_log2FC_HFD_vs_CON,
  candidate$anchor_log2FC_HFD_vs_CON)
write_table(candidate, OUT, "02_all_622_top15_ATAC_peaks_with_evidence_flags.csv.gz")
focused_candidates <- candidate |>
  dplyr::filter(C0_supported) |>
  dplyr::arrange(dplyr::desc(nested_top10),
    dplyr::desc(count_supported_40),
    dplyr::desc(anchor_total_counts), feature)
need(nrow(focused_candidates) == 37L &&
     sum(focused_candidates$nested_top10) == 6L,
     "Expected C0-supported and nested-region counts changed.")
write_table(focused_candidates, OUT,
  "02_C0_supported_37_peak_candidates_full_evidence.csv")
write_table(dplyr::filter(focused_candidates, count_supported_40), OUT,
  "02_higher_count_C0_supported_sensitivity.csv")
write_table(dplyr::filter(focused_candidates, nested_top10), OUT,
  "02_six_nested_C0_and_top10_HFD_open_peaks.csv")
write_bed <- function(x, filename) {
  readr::write_tsv(dplyr::transmute(x, chrom, start_0based_bed,
    end_1based, feature), file.path(OUT, filename), col_names = FALSE)
}
write_bed(focused_candidates, "02_C0_supported_37_GRCm39.bed")
write_bed(dplyr::filter(focused_candidates, count_supported_40),
  "02_higher_count_C0_supported_GRCm39.bed")
write_table(tibble(list_name = c("C0_supported", "count_supported_40",
  "nested_top10", "top20_clear_same", "top25_clear_same"),
  n_of_37 = c(nrow(focused_candidates),
    sum(focused_candidates$count_supported_40),
    sum(focused_candidates$nested_top10),
    sum(focused_candidates$top20_clear_same),
    sum(focused_candidates$top25_clear_same))), OUT,
  "02_candidate_support_summary.csv")
```

## 4. Xist detection strata as a bounded sensitivity analysis

Evaluate the 37 regions in the same C0 and top-15 group separately among
Xist-detected versus Xist-undetected nuclei only when both diets have at
least 30 nuclei. The results are a sensitivity diagnostic: Xist-undetected
is a mixture of male and female nuclei, and these groups are neither
independent libraries nor proper biological replicates.

```{r Xist-stratum-peak-audit}
depth <- Matrix::colSums(atac)
sex_effects <- list()
sex_stratum_eligibility <- list()
for (key in c("C0_all", "C0W0_top15")) {
  for (status in c("Xist_detected", "Xist_undetected")) {
    ix <- which(groups[[key]] & meta$Xist_detected ==
      identical(status, "Xist_detected"))
    n_con <- sum(meta$diet[ix] == "CON")
    n_hfd <- sum(meta$diet[ix] == "HFD")
    eligible <- n_con >= 30L && n_hfd >= 30L
    sex_stratum_eligibility[[length(sex_stratum_eligibility) + 1L]] <-
      tibble(group = key, Xist_status = status, n_CON = n_con,
        n_HFD = n_hfd, eligible = eligible)
    if (eligible) {
      sex_effects[[length(sex_effects) + 1L]] <- pooled_effect(
        atac, meta, ix, "ATAC", features = focused_candidates$feature,
        total_depth = depth) |>
        dplyr::mutate(group = key, Xist_status = status, .before = 1)
    }
  }
}
write_table(dplyr::bind_rows(sex_stratum_eligibility), OUT,
  "03_Xist_stratum_eligibility.csv")
sex_effects <- dplyr::bind_rows(sex_effects)
write_table(sex_effects, OUT,
  "03_C0_supported_region_effects_by_Xist_detection.csv")
if (nrow(sex_effects)) {
  sex_summary <- sex_effects |>
    dplyr::left_join(dplyr::select(focused_candidates, feature,
      anchor_log2FC_HFD_vs_CON), by = "feature", relationship = "many-to-one") |>
    dplyr::group_by(group, Xist_status) |>
    dplyr::summarise(region_n = dplyr::n(),
      evaluable_n = sum(evaluable),
      evaluable_same_direction = sum(evaluable &
        sign(log2FC_HFD_vs_CON) == sign(anchor_log2FC_HFD_vs_CON)),
      clear_same_direction = sum(clear_effect &
        sign(log2FC_HFD_vs_CON) == sign(anchor_log2FC_HFD_vs_CON)),
      .groups = "drop")
  write_table(sex_summary, OUT, "03_Xist_stratum_peak_summary.csv")
}
```

## 5. Handoff rules for an exact-interval genotype comparison

Use the same Step 03 GRCm39 common peak coordinates in WT, PPAR, and IL17.
Do not map cluster numbers across cohorts. For each other cohort, first
adjudicate an ISC-like state using its own marker, cycling, and secretory
context; independently score the same ISC signature and audit CON/HFD
membership. Then quantify **all 37 frozen WT intervals**, plus the 622
top-15 peaks as a broader view, on that cohort's unmodified common-peak
counts. Show each cohort's CON/HFD nucleus counts, peak counts, detection,
pooled CPM, and HFD-versus-CON effect, with missing evaluability explicit.
Compare directional effect vectors and exposure of the same exact intervals;
do not test a genotype-by-diet interaction with these pooled libraries.
Check the genotype-specific sex/proxy and depth balance before interpreting
different directions. A peak absent from a filtered assay is *missing*, not
a zero accessibility effect.

```{r finish}
finish_run(OUT, "STEP_08H_COMPLETE.txt", c(
  "C0 overall remains the broader ISC-enriched population; C0/W0 top15 is the marker-enriched sensitivity view.",
  "All 622 top15 clear peaks retained; 37 C0-supported and six nested-top10 flags exported without using nearby RNA.",
  "Xist/Y transcript detection and depth audited; stratum effects are sensitivity only, not mouse-adjusted inference.",
  "Exact canonical GRCm39 peak BEDs exported for later PPAR and IL17 comparison.",
  "One pooled WT library per diet: no significant DARs or genotype-by-diet inference.",
  paste("Results:", OUT)))
```
