---
title: "Step 08I | Frozen WT crypt peaks across PPAR and IL17"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This notebook follows the completed WT 08H audit. It carries **all 622**
top-15 WT candidate peaks forward. The **37** peaks with a same-direction
effect in WT C0 are a labeled subset; the 16 with stronger pooled counts are
another flag, not a new discovery list. None is a significant DAR.

The WT C0/W0 score gate is preserved. In PPAR, approved WNN clusters 3
(cycling ISC/early TA) and 6 (ISC-like) are inspected separately. In IL17,
approved clusters 9 (cycling ISC/TA) and 10 (ISC-like) are inspected
separately. Their unions are shown with parent-by-diet composition. Approved
crypt progenitors PPAR 2 and IL17 0 are context comparisons. Labels are
checked against the approved Step 6 decision files; numeric cluster IDs are
never transferred from WT to another genotype.

All new scores use the **same six genes** (`Lgr5`, `Ascl2`, `Smoc2`, `Axin2`,
`Lrig1`, `Rnf43`) on joined raw RNA counts with UCell `maxRank=400`.
Each parent's top 15% is ranked across both diets together. A second gate
uses the *observed WT top-15 minimum score* as a fixed reference threshold;
it is diagnostic because different cohorts can have different RNA sensitivity.
The combined score gate is the union of separately gated parents. Neither
diet effects nor Gourab's genes determine gate membership.

For WT, the notebook additionally computes the same 622-region effects in
**Xist-only** and **Y-only** nuclei within C0. These transcript categories
are suggestive sex proxies, not mouse sex calls. One pooled library per diet
per cohort means that all effects and cross-genotype comparisons remain
descriptive. Separate nuclei, nested gates, and within-library subsampling
do not create mouse replicates or enable genotype-by-diet inference.

Run chunks in order from the repository root on SOL, with R 4.4 and at least
96 GB RAM available. PPAR and IL17 objects are loaded and released in turn.
Outputs go to the WT `08i_frozen_WT_peaks_genotype_context`
directory. Input objects, annotations, and Step 08G/H outputs are not modified.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
  fig.width = 12, fig.height = 7)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Open the repository root before running 08I.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced", MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
source(file.path(REPO_ROOT, "R", "helpers.R"))
FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) FOLLOWUP_HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(FOLLOWUP_HELPER)) stop("Missing WT follow-up helper.")
source(FOLLOWUP_HELPER)
suppressPackageStartupMessages({
  library(Seurat); library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges)
})
need(requireNamespace("UCell", quietly = TRUE), "Install UCell in the active R library.")
need(identical(DATASET_VERSION, "v2_resequenced") && identical(COHORT_ID, "wt"),
  "08I expects the WT v2_resequenced configuration.")
CORE6 <- c("Lgr5", "Ascl2", "Smoc2", "Axin2", "Lrig1", "Rnf43")
Y_GENES <- c("Uty", "Ddx3y", "Eif2s3y", "Kdm5d")
UCELL_MAX_RANK <- 400L
MIN_PER_DIET <- 30L
COHORT_STATES <- list(
  ppar = list(primary = c("3", "6"), context = "2"),
  il17 = list(primary = c("9", "10"), context = "0"))
WT_ROOT <- OUTPUT_DIR
RESULT_ROOT <- file.path(RESULTS_ROOT, DATASET_VERSION, "independent")
STEP5E <- file.path(WT_ROOT, "05_wnn", "focused_wt_crypt_stem_progenitor",
  "objects", "WT_crypt_core_0_6_8_focused.rds")
STEP8G <- file.path(WT_ROOT, "08g_C0_C2_and_C0W0_ISC")
STEP8H <- file.path(WT_ROOT, "08h_wt_ISC_region_freeze_sex_audit")
OUT <- file.path(WT_ROOT, "08i_frozen_WT_peaks_genotype_context")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
```

## 1. Freeze the inputs and verify the WT reference

All coordinate matching below uses the exact Step 03 GRCm39 peak name and
interval. A peak absent from a cohort's filtered ATAC assay is reported as
missing, never as a zero effect. The WT score threshold comes from the
saved 08G gate rather than a value typed into this notebook.

```{r inputs}
PEAK_PATH <- file.path(STEP8H,
  "02_all_622_top15_ATAC_peaks_with_evidence_flags.csv.gz")
GATE_PATH <- file.path(STEP8G, "02_C0W0_gate_size_and_score_sensitivity.csv")
NUCLEI_PATH <- file.path(STEP8H,
  "01_nucleus_level_sex_proxy_and_gate_audit.csv.gz")
source_paths <- c(STEP5E, PEAK_PATH, GATE_PATH, NUCLEI_PATH,
  file.path(STEP8G, "STEP_08G_COMPLETE.txt"),
  file.path(STEP8H, "STEP_08H_COMPLETE.txt"), SAMPLE_SHEET_PATH)
for (cohort in names(COHORT_STATES)) {
  source_paths <- c(source_paths,
    file.path(RESULT_ROOT, cohort, "06_annotation", "multiome_annotated.rds"),
    file.path(DATASET_CONFIG_DIR, "cohorts", cohort,
      "cell_state_annotations.csv"))
}
need(all(file.exists(source_paths)), paste("Missing input(s):",
  paste(source_paths[!file.exists(source_paths)], collapse = "; ")))
# Hash small tables; hashing several large Seurat RDS files needlessly repeats
# expensive I/O. Record their full paths, sizes, and timestamps instead.
info <- file.info(source_paths)
manifest <- tibble(path = normalizePath(source_paths), bytes = info$size,
  modified = as.character(info$mtime), md5 = NA_character_)
small <- !grepl("\\.rds$", source_paths, ignore.case = TRUE)
manifest$md5[small] <- unname(tools::md5sum(source_paths[small]))
write_table(manifest, OUT, "00_input_manifest.csv")
sample_sheet <- read_csv_checked(SAMPLE_SHEET_PATH)
require_columns(sample_sheet, c("cohort_id", "sample_id", "diet",
  "mice_pooled", "female_mice", "male_mice"), "sample sheet")
write_table(sample_sheet |>
  dplyr::filter(cohort_id %in% c("wt", names(COHORT_STATES))) |>
  dplyr::select(cohort_id, sample_id, diet, mice_pooled,
    female_mice, male_mice), OUT, "00_pooled_mouse_sex_composition.csv")

anchor <- read_csv_checked(PEAK_PATH)
require_columns(anchor, c("feature", "chrom", "start_1based", "end_1based",
  "anchor_log2FC_HFD_vs_CON", "anchor_CON_counts", "anchor_HFD_counts",
  "anchor_CON_percent", "anchor_HFD_percent", "C0_all_log2FC_HFD_vs_CON",
  "C0_supported", "count_supported_40", "nested_top10",
  "anchor_evaluable", "anchor_clear_effect"), "08H peak table")
need(nrow(anchor) == 622L && !anyDuplicated(anchor$feature) &&
  sum(anchor$C0_supported) == 37L &&
  sum(anchor$C0_supported & anchor$count_supported_40) == 16L &&
  sum(anchor$nested_top10) == 6L &&
  all(anchor$anchor_evaluable & anchor$anchor_clear_effect),
  "Frozen WT 622/37/16/6 peak contract changed. Inspect Step 08H before rerunning.")
gate <- read_csv_checked(GATE_PATH)
require_columns(gate, c("gate", "minimum_primary_score", "selected_n",
  "CON_n", "HFD_n"), "08G gate table")
need(sum(gate$gate == "top15") == 1L, "Expected one top15 WT gate row.")
WT_SCORE_CUTOFF <- gate$minimum_primary_score[gate$gate == "top15"]
need(is.finite(WT_SCORE_CUTOFF) && WT_SCORE_CUTOFF > 0 &&
  gate$selected_n[gate$gate == "top15"] == 174L,
  "WT top15 reference gate changed.")
write_table(tibble(signature = paste(CORE6, collapse = ";"),
  maxRank = UCELL_MAX_RANK, WT_top15_minimum_score = WT_SCORE_CUTOFF,
  WT_top15_CON_n = gate$CON_n[gate$gate == "top15"],
  WT_top15_HFD_n = gate$HFD_n[gate$gate == "top15"],
  gate_rule = "Within each approved genotype parent, rank all diets together; positive score; ceiling(0.15*n)."),
  OUT, "00_fixed_signature_and_WT_gate.csv")
write_table(anchor, OUT, "00_frozen_WT_622_peak_reference.csv.gz")
write_table(dplyr::filter(anchor, C0_supported), OUT,
  "00_frozen_WT_37_C0_supported_reference.csv")
```

## 2. WT exclusive sex-proxy sensitivity for all 622 peaks

Xist-only and Y-only are more specific transcript patterns than Xist
non-detection, but neither assigns a mouse's sex. We only calculate a
stratum's effect with at least 30 nuclei in **each** diet. Top-15 Y-only is
expected to be too small; it is marked ineligible rather than treated as
having zero diet-associated peaks. The 37 and 16 are reported as subsets
of the unchanged 622.

```{r wt-exclusive-proxy}
focused <- load_focused(STEP5E)
wt_meta <- focused$metadata
wt_atac <- focused$matrices$ATAC_counts
recorded <- read_csv_checked(NUCLEI_PATH)
require_columns(recorded, c("cell_barcode", "diet", "focused_cluster",
  "parent_wnn_cluster", "C0W0_top15", "sex_proxy_status"),
  "08H nucleus sex/gate table")
need(!anyDuplicated(recorded$cell_barcode) &&
  setequal(recorded$cell_barcode, wt_meta$cell_barcode),
  "08H nuclei do not match the WT focused object.")
recorded <- recorded[match(wt_meta$cell_barcode, recorded$cell_barcode), ]
need(identical(recorded$cell_barcode, wt_meta$cell_barcode) &&
  identical(as.character(recorded$diet), wt_meta$diet) &&
  identical(as.character(recorded$focused_cluster), wt_meta$focused_cluster) &&
  identical(as.character(recorded$parent_wnn_cluster),
    wt_meta$parent_wnn_cluster) &&
  identical(colnames(wt_atac), wt_meta$cell_barcode),
  "WT 08H assignments and ATAC counts are not aligned.")
wt_meta$sex_proxy_status <- recorded$sex_proxy_status
wt_meta$C0W0_top15 <- as.logical(recorded$C0W0_top15)
need(!anyNA(wt_meta$C0W0_top15) &&
  all(anchor$feature %in% rownames(wt_atac)),
  "WT focused assay lacks frozen peaks or recorded score gate.")
wt_ix <- list(C0_all = which(wt_meta$focused_cluster == "0"),
  C0W0_top15 = which(wt_meta$C0W0_top15))
wt_specs <- dplyr::bind_rows(lapply(names(wt_ix), function(key) {
  dplyr::bind_rows(lapply(c("Xist_only", "Y_only"), function(status) {
    ix <- wt_ix[[key]][wt_meta$sex_proxy_status[wt_ix[[key]]] == status]
    tibble(group = key, sex_proxy_status = status, indices = list(ix),
      n_CON = sum(wt_meta$diet[ix] == "CON"),
      n_HFD = sum(wt_meta$diet[ix] == "HFD"))
  }))
})) |>
  dplyr::mutate(eligible = n_CON >= MIN_PER_DIET & n_HFD >= MIN_PER_DIET)
need(all(wt_specs$eligible[wt_specs$group == "C0_all"]),
  "The two C0 exclusive-proxy groups no longer meet the minimum size.")
write_table(dplyr::select(wt_specs, -indices), OUT,
  "01_WT_exclusive_sex_proxy_eligibility.csv")
wt_depth <- Matrix::colSums(wt_atac)
wt_sex_effects <- dplyr::bind_rows(lapply(which(wt_specs$eligible), function(i) {
  key <- wt_specs$group[[i]]
  proxy <- wt_specs$sex_proxy_status[[i]]
  message("WT exclusive proxy: ", key, " / ", proxy)
  pooled_effect(wt_atac, wt_meta, wt_specs$indices[[i]], "ATAC",
    features = anchor$feature, total_depth = wt_depth) |>
    dplyr::mutate(group = .env$key, sex_proxy_status = .env$proxy, .before = 1)
}))
need(!anyDuplicated(wt_sex_effects[c("group", "sex_proxy_status", "feature")]),
  "WT sex-proxy effect rows are duplicated.")
wt_sex_effects <- dplyr::left_join(wt_sex_effects,
  dplyr::select(anchor, feature, C0_supported, count_supported_40,
    nested_top10, anchor_log2FC_HFD_vs_CON), by = "feature",
  relationship = "many-to-one") |>
  dplyr::mutate(same_WT_direction = evaluable &
    sign(log2FC_HFD_vs_CON) == sign(anchor_log2FC_HFD_vs_CON))
write_table(wt_sex_effects, OUT,
  "01_WT_622_peak_effects_by_exclusive_proxy.csv.gz")
wt_sex_summary <- dplyr::bind_rows(lapply(c("all_622", "C0_supported_37",
  "count_supported_16", "nested_top10_6"), function(tier) {
  z <- dplyr::filter(wt_sex_effects,
    if (tier == "all_622") TRUE else if (tier == "C0_supported_37")
      C0_supported else if (tier == "count_supported_16")
      C0_supported & count_supported_40 else nested_top10)
  z |>
    dplyr::group_by(group, sex_proxy_status) |>
    dplyr::summarise(tier = tier, n_regions = dplyr::n(),
      evaluable_n = sum(evaluable), same_direction_n = sum(same_WT_direction),
      clear_same_n = sum(clear_effect & same_WT_direction), .groups = "drop")
}))
write_table(wt_sex_summary, OUT, "01_WT_exclusive_proxy_peak_summary.csv")
wt_pair <- wt_sex_effects |>
  dplyr::filter(group == "C0_all") |>
  dplyr::select(feature, sex_proxy_status, log2FC_HFD_vs_CON,
    CON_counts, HFD_counts, evaluable, clear_effect) |>
  tidyr::pivot_wider(names_from = sex_proxy_status,
    values_from = c(log2FC_HFD_vs_CON, CON_counts, HFD_counts,
      evaluable, clear_effect)) |>
  dplyr::left_join(dplyr::select(anchor, feature, chrom, start_1based,
    end_1based, C0_supported, count_supported_40, nested_top10,
    anchor_log2FC_HFD_vs_CON), by = "feature", relationship = "one-to-one") |>
  dplyr::mutate(both_proxy_evaluable = evaluable_Xist_only & evaluable_Y_only,
    both_proxy_clear_same = both_proxy_evaluable &
      clear_effect_Xist_only & clear_effect_Y_only &
      sign(log2FC_HFD_vs_CON_Xist_only) ==
        sign(anchor_log2FC_HFD_vs_CON) &
      sign(log2FC_HFD_vs_CON_Y_only) ==
        sign(anchor_log2FC_HFD_vs_CON))
write_table(wt_pair, OUT, "01_WT_C0_622_peak_exclusive_proxy_pair.csv.gz")
wt_37_review <- wt_pair |>
  dplyr::filter(C0_supported) |>
  dplyr::arrange(dplyr::desc(both_proxy_clear_same),
    dplyr::desc(count_supported_40), feature)
write_table(wt_37_review, OUT,
  "01_WT_37_C0_supported_exclusive_proxy_review.csv")
print(wt_sex_summary)
cat("WT C0 regions clear and same-direction in both exclusive proxy strata:",
  sum(wt_37_review$both_proxy_clear_same), "of 37; higher-count subset:",
  sum(wt_37_review$both_proxy_clear_same &
    wt_37_review$count_supported_40), "of 16.\n")
rm(focused, wt_meta, wt_atac, wt_depth, recorded, wt_sex_effects)
invisible(gc())
```

## 3. Fixed cohort labels, six-gene score, and comparable peak effects

Each candidate parent has its own top-15 gate. We also export the entire
parent, a WT-calibrated score gate, the union of primary parents, and one
separate crypt progenitor context group. An ineligible group or missing
peak produces `NA` effects and a specific status. The full cohort ATAC
matrix supplies each nucleus's sequencing-depth denominator, even when
only the frozen 622 peaks are evaluated.

```{r genotype-cohorts}
check_interval_identity <- function(object, features, reference) {
  gr <- Signac::granges(object[["ATAC"]])
  names(gr) <- rownames(object[["ATAC"]])
  ix <- match(features, names(gr))
  present <- !is.na(ix)
  if (any(present)) {
    x <- gr[ix[present]]
    ref <- reference[match(features[present], reference$feature), ]
    need(all(as.character(GenomicRanges::seqnames(x)) == ref$chrom) &&
      all(as.integer(BiocGenerics::start(x)) == ref$start_1based) &&
      all(as.integer(BiocGenerics::end(x)) == ref$end_1based),
      "A cohort peak name maps to a different genomic interval.")
  }
  present
}
build_spec <- function(meta, states) {
  out <- list()
  add <- function(key, role, ix) {
    out[[length(out) + 1L]] <<- tibble(contrast = key, role = role,
      indices = list(as.integer(ix)))
  }
  selected <- list()
  for (parent in states$primary) {
    ix <- which(meta$wnn_cluster == parent)
    need(length(ix) > 0L, paste("Missing approved ISC parent:", parent))
    ranked <- ix[order(-meta$ISC_core6_UCell[ix], meta$cell_barcode[ix])]
    top <- ranked[seq_len(ceiling(.15 * length(ranked)))]
    top <- top[meta$ISC_core6_UCell[top] > 0]
    calibrated <- ix[meta$ISC_core6_UCell[ix] >= WT_SCORE_CUTOFF]
    add(paste0("parent_", parent, "_all"), "approved_ISC_parent", ix)
    add(paste0("parent_", parent, "_top15"), "within_parent_top15", top)
    add(paste0("parent_", parent, "_WTcal"), "WT_reference_score_gate", calibrated)
    selected[[parent]] <- list(all = ix, top15 = top, WTcal = calibrated)
  }
  for (gate_name in c("all", "top15", "WTcal")) {
    ix <- sort(unique(unlist(lapply(selected, `[[`, gate_name),
      use.names = FALSE)))
    add(paste0("annotated_union_", gate_name),
      paste0("union_of_parent_", gate_name), ix)
  }
  add("crypt_progenitor_context", "approved_nonprimary_crypt_progenitor",
    which(meta$wnn_cluster %in% states$context))
  dplyr::bind_rows(out) |>
    dplyr::mutate(n_CON = vapply(indices, function(ix)
      as.integer(sum(meta$diet[ix] == "CON")), integer(1)),
      n_HFD = vapply(indices, function(ix)
        as.integer(sum(meta$diet[ix] == "HFD")), integer(1)),
      eligible = n_CON >= MIN_PER_DIET & n_HFD >= MIN_PER_DIET)
}

all_effects <- list()
all_balance <- list()
all_score_audit <- list()
all_presence <- list()
all_marker_audit <- list()
all_marker_detection <- list()
for (cohort in names(COHORT_STATES)) {
  message("08I cohort: ", toupper(cohort))
  obj_path <- file.path(RESULT_ROOT, cohort, "06_annotation",
    "multiome_annotated.rds")
  decision_path <- file.path(DATASET_CONFIG_DIR, "cohorts", cohort,
    "cell_state_annotations.csv")
  decisions <- read_csv_checked(decision_path)
  require_columns(decisions, c("cluster", "approved", "broad_compartment",
    "cell_state", "annotation_confidence"), paste(cohort, "annotation decision"))
  decisions$cluster <- as.character(decisions$cluster)
  chosen <- c(COHORT_STATES[[cohort]]$primary,
    COHORT_STATES[[cohort]]$context)
  map <- decisions[match(chosen, decisions$cluster), ]
  need(!anyNA(map$cluster) && all(map$approved) &&
    all(map$broad_compartment == "epithelial") &&
    all(map$annotation_confidence %in% c("high", "moderate")),
    paste("Approved primary/context annotations changed in", cohort))
  write_table(dplyr::mutate(map, cohort = .env$cohort, .before = 1),
    OUT, paste0("02_", cohort, "_approved_crypt_parent_decisions.csv"))

  obj <- readRDS(obj_path)
  need(all(c("RNA", "ATAC") %in% names(obj)) &&
    "counts" %in% SeuratObject::Layers(obj[["RNA"]]) &&
    "counts" %in% SeuratObject::Layers(obj[["ATAC"]]),
    paste("Missing joined RNA/ATAC counts in", cohort))
  meta <- obj[[]]
  require_columns(meta, c("wnn_cluster", "cell_state", "sample_id", "diet",
    "nCount_RNA", "nCount_ATAC", "nFeature_RNA"),
    paste(cohort, "annotated metadata"))
  meta$cell_barcode <- rownames(meta)
  meta$wnn_cluster <- as.character(meta$wnn_cluster)
  meta$cell_state <- as.character(meta$cell_state)
  meta$diet <- as.character(meta$diet)
  meta$sample_id <- as.character(meta$sample_id)
  cohort_samples <- sample_sheet |>
    dplyr::filter(cohort_id == .env$cohort) |>
    dplyr::select(sample_id, diet)
  need(!anyDuplicated(meta$cell_barcode) && !anyNA(meta$diet) &&
    setequal(unique(meta$diet), c("CON", "HFD")) &&
    setequal(meta$cell_barcode, colnames(obj)) &&
    nrow(cohort_samples) == 2L &&
    setequal(cohort_samples$diet, c("CON", "HFD")) &&
    !anyDuplicated(cohort_samples$sample_id) &&
    setequal(unique(meta$sample_id), cohort_samples$sample_id) &&
    identical(meta$diet, cohort_samples$diet[
      match(meta$sample_id, cohort_samples$sample_id)]),
    paste("Unexpected samples/diets/barcodes in", cohort))
  for (j in seq_len(nrow(map))) {
    ix <- meta$wnn_cluster == map$cluster[[j]]
    need(any(ix) && all(meta$cell_state[ix] == map$cell_state[[j]]),
      paste(cohort, "stored annotation differs from the approved decision:",
        map$cluster[[j]]))
  }
  rna_aligned <- align_columns(SeuratObject::LayerData(obj[["RNA"]],
    layer = "counts"), meta$cell_barcode, paste0(cohort, " RNA"))
  atac_aligned <- align_columns(SeuratObject::LayerData(obj[["ATAC"]],
    layer = "counts"), meta$cell_barcode, paste0(cohort, " ATAC"))
  rna <- rna_aligned$matrix
  atac <- atac_aligned$matrix
  write_table(dplyr::bind_rows(rna_aligned$audit, atac_aligned$audit),
    OUT, paste0("02_", cohort, "_matrix_alignment.csv"))
  availability <- tibble(cohort = cohort, gene = CORE6,
    present_in_RNA = CORE6 %in% rownames(rna))
  all_marker_audit[[cohort]] <- availability
  need(all(availability$present_in_RNA),
    paste("An ISC signature gene is absent in", cohort,
      "- cannot compare a shortened score."))
  scores <- UCell::ScoreSignatures_UCell(rna,
    features = list(ISC_core6 = CORE6), maxRank = UCELL_MAX_RANK)
  need(setequal(rownames(scores), meta$cell_barcode) &&
    "ISC_core6_UCell" %in% colnames(scores),
    paste("Unexpected UCell score format in", cohort))
  meta$ISC_core6_UCell <- as.numeric(scores[meta$cell_barcode,
    "ISC_core6_UCell"])
  meta$ISC_core_genes_detected <- as.integer(Matrix::colSums(
    rna[CORE6, meta$cell_barcode, drop = FALSE] > 0))
  need(all(is.finite(meta$ISC_core6_UCell)),
    paste("Nonfinite UCell scores in", cohort))
  all_marker_detection[[cohort]] <- dplyr::bind_rows(lapply(CORE6, function(gene) {
    meta |>
      dplyr::filter(wnn_cluster %in% chosen) |>
      dplyr::mutate(detected = as.numeric(rna[gene, cell_barcode]) > 0) |>
      dplyr::group_by(wnn_cluster, cell_state, diet) |>
      dplyr::summarise(cohort = cohort, gene = gene, nuclei = dplyr::n(),
        detected_n = sum(detected), detected_percent = 100 * mean(detected),
        .groups = "drop")
  }))
  meta$Xist_detected <- if ("Xist" %in% rownames(rna)) {
    as.numeric(rna["Xist", meta$cell_barcode]) > 0
  } else {
    rep(NA, nrow(meta))
  }
  y_present <- intersect(Y_GENES, rownames(rna))
  meta$Y_any_detected <- if (length(y_present)) {
    Matrix::colSums(rna[y_present, meta$cell_barcode, drop = FALSE] > 0) > 0
  } else {
    rep(NA, nrow(meta))
  }
  spec <- build_spec(meta, COHORT_STATES[[cohort]])
  need(!anyDuplicated(spec$contrast) &&
    all(spec$eligible[spec$contrast %in%
      paste0("parent_", COHORT_STATES[[cohort]]$primary, "_all")]),
    paste("An approved ISC parent has fewer than 30 nuclei per diet in", cohort))

  # Keep the universe as 622 exact WT intervals; never equate a filtered-out
  # peak with a zero effect in the genotype cohort.
  present <- check_interval_identity(obj, anchor$feature, anchor)
  all_presence[[cohort]] <- tibble(cohort = cohort,
    feature = anchor$feature, feature_present = present)
  selected_features <- anchor$feature[present]
  need(length(selected_features) > 0L &&
    all(selected_features %in% rownames(atac)),
    paste("No comparable WT peaks in the", cohort, "ATAC assay."))
  depth <- Matrix::colSums(atac)
  all_balance[[cohort]] <- dplyr::bind_rows(lapply(seq_len(nrow(spec)), function(i) {
    ix <- spec$indices[[i]]
    dplyr::bind_rows(lapply(c("CON", "HFD"), function(diet_value) {
      k <- ix[meta$diet[ix] == diet_value]
      tibble(cohort = cohort, contrast = spec$contrast[[i]],
        role = spec$role[[i]], diet = diet_value, n = length(k),
        eligible = spec$eligible[[i]],
        median_ISC_UCell = if (length(k)) median(meta$ISC_core6_UCell[k]) else NA_real_,
        median_detected_six = if (length(k)) median(meta$ISC_core_genes_detected[k]) else NA_real_,
        median_RNA_features = if (length(k)) median(meta$nFeature_RNA[k]) else NA_real_,
        median_RNA_counts = if (length(k)) median(meta$nCount_RNA[k]) else NA_real_,
        median_ATAC_counts = if (length(k)) median(meta$nCount_ATAC[k]) else NA_real_,
        Xist_detected_percent = if (length(k)) 100 * mean(meta$Xist_detected[k]) else NA_real_,
        Y_detected_percent = if (length(k)) 100 * mean(meta$Y_any_detected[k]) else NA_real_)
    }))
  }))
  write_table(dplyr::select(spec, -indices) |>
    dplyr::mutate(cohort = .env$cohort, .before = 1), OUT,
    paste0("02_", cohort, "_contrast_eligibility.csv"))
  top15_ix <- sort(unique(unlist(spec$indices[
    grepl("^parent_.*_top15$", spec$contrast)], use.names = FALSE)))
  top15_cells <- meta$cell_barcode[top15_ix]
  write_table(meta |>
    dplyr::filter(wnn_cluster %in% c(COHORT_STATES[[cohort]]$primary,
      COHORT_STATES[[cohort]]$context)) |>
    dplyr::transmute(cohort = .env$cohort, cell_barcode, sample_id, diet,
      wnn_cluster, cell_state, ISC_core6_UCell, ISC_core_genes_detected,
      Xist_detected, Y_any_detected, nFeature_RNA, nCount_RNA, nCount_ATAC,
      top15 = cell_barcode %in% .env$top15_cells,
      WTcal = ISC_core6_UCell >= WT_SCORE_CUTOFF),
    OUT, paste0("02_", cohort, "_crypt_nuclei_scores.csv.gz"))

  cohort_effect <- dplyr::bind_rows(lapply(which(spec$eligible), function(i) {
    key <- spec$contrast[[i]]
    message(cohort, " ATAC exact intervals: ", key,
      " (", spec$n_CON[[i]], " CON / ", spec$n_HFD[[i]], " HFD)")
    pooled_effect(atac, meta, spec$indices[[i]], "ATAC",
      features = selected_features, total_depth = depth) |>
      dplyr::mutate(contrast = .env$key, .before = 1)
  }))
  need(!anyDuplicated(cohort_effect[c("contrast", "feature")]),
    paste("Duplicate exact-interval effects in", cohort))
  full <- tidyr::expand_grid(contrast = spec$contrast,
    feature = anchor$feature) |>
    dplyr::left_join(cohort_effect,
      by = c("contrast", "feature"), relationship = "one-to-one") |>
    dplyr::left_join(dplyr::select(spec, contrast, role, n_CON, n_HFD,
      eligible), by = "contrast", relationship = "many-to-one") |>
    dplyr::mutate(cohort = .env$cohort, .before = 1) |>
    dplyr::left_join(all_presence[[cohort]],
      by = c("cohort", "feature"), relationship = "many-to-one") |>
    dplyr::left_join(dplyr::select(anchor, feature, chrom, start_1based,
      end_1based, C0_supported, count_supported_40, nested_top10,
      anchor_log2FC_HFD_vs_CON, C0_all_log2FC_HFD_vs_CON,
      anchor_CON_counts, anchor_HFD_counts), by = "feature",
      relationship = "many-to-one") |>
    dplyr::mutate(status = dplyr::case_when(
      !eligible ~ "gate_ineligible", !feature_present ~ "peak_absent_from_assay",
      !evaluable ~ "peak_below_evaluability", TRUE ~ "evaluable"),
      same_WT_top15_direction = dplyr::if_else(status == "evaluable",
        sign(log2FC_HFD_vs_CON) == sign(anchor_log2FC_HFD_vs_CON), NA))
  need(all(!is.na(full$status)) &&
    sum(full$status == "peak_absent_from_assay") ==
      sum(!present) * sum(spec$eligible),
    paste("Missing/evaluable peak bookkeeping failed in", cohort))
  all_effects[[cohort]] <- full

  all_score_audit[[cohort]] <- meta |>
    dplyr::filter(wnn_cluster %in% COHORT_STATES[[cohort]]$primary) |>
    dplyr::group_by(wnn_cluster, diet, ISC_core_genes_detected) |>
    dplyr::summarise(cohort = cohort, nuclei = dplyr::n(),
      median_score = median(ISC_core6_UCell),
      median_RNA_features = median(nFeature_RNA), .groups = "drop")
  rm(obj, meta, rna, atac, rna_aligned, atac_aligned, scores,
    depth, cohort_effect, full, spec)
  invisible(gc())
}
write_table(dplyr::bind_rows(all_marker_audit), OUT,
  "02_fixed_six_gene_assay_availability.csv")
write_table(dplyr::bind_rows(all_marker_detection), OUT,
  "02_six_gene_detection_by_approved_parent_diet.csv")
write_table(dplyr::bind_rows(all_score_audit), OUT,
  "02_crypt_ISC_score_by_detected_genes_diet.csv")
write_table(dplyr::bind_rows(all_balance), OUT,
  "02_crypt_gate_parent_diet_sex_depth_balance.csv")
write_table(dplyr::bind_rows(all_presence), OUT,
  "02_frozen_peak_presence_by_cohort.csv")
effects <- dplyr::bind_rows(all_effects)
need(!anyDuplicated(effects[c("cohort", "contrast", "feature")]) &&
  all(vapply(all_effects, nrow, integer(1)) ==
    622L * vapply(names(COHORT_STATES), function(cohort)
      length(COHORT_STATES[[cohort]]$primary) * 3L + 4L, integer(1))),
  "Not every frozen peak has a status row for each genotype contrast.")
write_table(effects, OUT, "03_all_622_exact_interval_cohort_effects.csv.gz")
write_table(dplyr::filter(effects, C0_supported), OUT,
  "03_C0_supported_37_exact_interval_cohort_effects.csv.gz")
```

## 4. Summarize effect direction, evaluability, and composition

The descriptive `clear_effect` threshold still means >=30 pooled fragments,
>=1% detection in either diet, and absolute depth-normalized log2FC >=0.5.
No P values are generated. All cohort outcomes report the original WT
reference effect alongside the corresponding genotype effect. Same-direction
counts are descriptive summaries of WT-selected peaks,
not tests of a genotype interaction. Inspect each parent's CON/HFD balance
before interpreting its combined union.

```{r summaries-and-plots}
summary_by_tier <- function(tier, subset) {
  effects |>
    dplyr::filter(.env$subset) |>
    dplyr::group_by(cohort, contrast, role, n_CON, n_HFD, eligible) |>
    dplyr::summarise(tier = tier, frozen_regions = dplyr::n(),
      present_regions = sum(feature_present),
      evaluable_regions = sum(status == "evaluable"),
      clear_regions = sum(clear_effect, na.rm = TRUE),
      evaluable_same_WT_direction = sum(same_WT_top15_direction,
        na.rm = TRUE),
      clear_same_WT_direction = sum(clear_effect &
        same_WT_top15_direction, na.rm = TRUE),
      clear_opposite_WT_direction = sum(clear_effect &
        !same_WT_top15_direction, na.rm = TRUE),
      .groups = "drop") |>
    dplyr::mutate(dplyr::across(c(evaluable_regions, clear_regions,
      evaluable_same_WT_direction, clear_same_WT_direction,
      clear_opposite_WT_direction), ~dplyr::if_else(eligible, .x, NA_integer_)))
}
summary <- dplyr::bind_rows(
  summary_by_tier("all_622", rep(TRUE, nrow(effects))),
  summary_by_tier("C0_supported_37", effects$C0_supported),
  summary_by_tier("count_supported_16", effects$C0_supported &
    effects$count_supported_40),
  summary_by_tier("nested_top10_6", effects$nested_top10))
write_table(summary, OUT, "03_cohort_gate_peak_effect_summary.csv")
source_composition <- dplyr::bind_rows(lapply(names(COHORT_STATES), function(cohort) {
  a <- all_balance[[cohort]]
  parents <- paste0("parent_", COHORT_STATES[[cohort]]$primary, "_")
  dplyr::bind_rows(lapply(c("all", "top15", "WTcal"), function(g) {
    selected <- a |>
      dplyr::filter(contrast %in% paste0(parents, g)) |>
      dplyr::mutate(gate = g,
        parent = sub("^parent_([^_]+)_.*$", "\\1", contrast))
    selected |>
      dplyr::group_by(cohort, gate, diet) |>
      dplyr::mutate(parent_percent_of_union = if (sum(n) > 0L)
        100 * n / sum(n) else NA_real_) |>
      dplyr::ungroup() |>
      dplyr::select(cohort, gate, parent, diet, n,
        parent_percent_of_union, median_ISC_UCell,
        median_detected_six, median_ATAC_counts, Xist_detected_percent)
  }))
}))
write_table(source_composition, OUT,
  "03_parent_contributions_to_annotated_unions.csv")
print(summary |>
  dplyr::filter(tier %in% c("all_622", "C0_supported_37"),
    contrast %in% c("annotated_union_all", "annotated_union_top15",
      paste0("parent_", unlist(lapply(COHORT_STATES, `[[`, "primary")),
        "_all"))))
print(source_composition)

p_scatter <- effects |>
  dplyr::filter(contrast %in% c("annotated_union_all",
    "annotated_union_top15"), status == "evaluable") |>
  ggplot2::ggplot(ggplot2::aes(x = anchor_log2FC_HFD_vs_CON,
    y = log2FC_HFD_vs_CON,
    color = ifelse(C0_supported, "WT C0-supported 37", "Other WT top-15 peaks"))) +
  ggplot2::geom_hline(yintercept = 0, linewidth = .25) +
  ggplot2::geom_vline(xintercept = 0, linewidth = .25) +
  ggplot2::geom_point(alpha = .55, size = 1) +
  ggplot2::facet_grid(cohort ~ contrast) +
  ggplot2::scale_color_manual(values = c("WT C0-supported 37" = "#B24B3A",
    "Other WT top-15 peaks" = "#5C7080")) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "Exact WT candidate intervals in approved genotype crypt states",
    subtitle = "Evaluable peaks only; pooled descriptive log2FC; check parent balance before union interpretation",
    x = "WT C0/W0 top-15 log2FC (HFD vs CON)",
    y = "Genotype crypt-state log2FC (HFD vs CON)", color = NULL)
save_vector_plot(p_scatter, PLOTS,
  "03_WT_vs_genotype_exact_interval_effects.pdf", 13, 8)

ordered37 <- dplyr::filter(anchor, C0_supported) |>
  dplyr::arrange(anchor_log2FC_HFD_vs_CON) |>
  dplyr::pull(feature)
p_heat <- effects |>
  dplyr::filter(C0_supported,
    contrast %in% c("annotated_union_all", "annotated_union_top15",
      paste0("parent_", unlist(lapply(COHORT_STATES, `[[`, "primary")),
        "_all"))) |>
  dplyr::mutate(feature = factor(feature, levels = ordered37),
    contrast = factor(contrast,
      levels = c("annotated_union_all", "annotated_union_top15",
        paste0("parent_", unlist(lapply(COHORT_STATES, `[[`, "primary")),
          "_all"))),
    effect_for_display = ifelse(status == "evaluable",
      log2FC_HFD_vs_CON, NA_real_)) |>
  ggplot2::ggplot(ggplot2::aes(x = feature, y = contrast,
    fill = effect_for_display)) +
  ggplot2::geom_tile(color = "white", linewidth = .1) +
  ggplot2::facet_wrap(~cohort, ncol = 1, scales = "free_y") +
  ggplot2::scale_fill_gradient2(low = "#3279A8", mid = "white",
    high = "#C45F48", midpoint = 0, na.value = "grey85") +
  ggplot2::theme_bw(base_size = 8) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90,
    hjust = 1, vjust = .5, size = 6)) +
  ggplot2::labs(title = "Frozen WT C0-supported regions in each genotype parent",
    subtitle = "Grey: missing or below evaluability; colors show descriptive log2FC only",
    x = "Exact GRCm39 peak", y = NULL, fill = "HFD vs CON\nlog2FC")
save_vector_plot(p_heat, PLOTS,
  "03_frozen_37_peak_genotype_context_heatmap.pdf", 17, 7)
```

## 5. Handoff

Read `01_WT_exclusive_proxy_peak_summary.csv` first. Then review
`02_crypt_gate_parent_diet_sex_depth_balance.csv` and
`03_parent_contributions_to_annotated_unions.csv` before the effect table.
Rows labeled `peak_absent_from_assay`, `peak_below_evaluability`, or
`gate_ineligible` must remain missing in subsequent comparisons. The
full 622-row WT source table and all genotype contrasts are exported to
support a review pack and a later matched-background motif analysis.

```{r finish}
finish_run(OUT, "STEP_08I_COMPLETE.txt", c(
  "The WT 622/37/16/6 candidate lists and exact GRCm39 intervals were kept fixed.",
  "WT C0 Xist-only and Y-only effects are transcript-proxy sensitivity views.",
  "PPAR 3/6 and IL17 9/10 were scored with six fixed ISC genes, maxRank=400.",
  "Parent effects precede any union interpretation; progenitor context remains separate.",
  "Peak absence and insufficient counts remain explicit missing values.",
  "One pooled library per diet per genotype: no significant DARs or genotype interaction test.",
  paste("Results:", OUT)))
```
