---
title: "Step 08K | WT ISC interval attenuation, coverage, and motif context"
output:
  html_notebook:
    toc: true
    toc_float: true
---

Run from the repository root on SOL after Steps 08H, 08I, and 08J. This
notebook **does not change** the six-gene UCell definition, approved parent
clusters, frozen WT GRCm39 intervals, or pooled effect estimates. It compares
the same genomic intervals across the existing cohorts. WT C0 is compared
with all annotated ISC parents; the WT C0/W0 top-15% score gate is compared
with the genotype within-parent top-15% and WT-calibrated gates. These are
related biological populations, not identical cell labels.

Attenuation means a **smaller absolute depth-normalized log2FC**, including
effects whose direction reversed but whose magnitude shrank. Direction and
parent-specific effects are always shown separately. Peak-resampling intervals
below quantify variation across the selected regions conditional on these
three pooled datasets; they are not biological confidence intervals or
genotype-by-diet tests. JASPAR matched-set permutation P values, if
generated, test motif sequence over-representation against matched peaks,
**not diet response**.
Motif presence alone does not establish TF binding or activity.

The motif section requires packages `JASPAR2024`, `TFBSTools`, `motifmatchr`,
`Biostrings`, `Rsamtools`, `DBI`, and `RSQLite` and an indexed GRCm39 FASTA
(as used in Step 08D). Install in the active R 4.4 library if absent:

```r
BiocManager::install(c("JASPAR2024", "TFBSTools", "motifmatchr",
  "Biostrings", "Rsamtools", "DBI", "RSQLite"), ask = FALSE,
  update = FALSE)
```

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
  fig.width = 12, fig.height = 7)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Open the repository root before 08K.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(HELPER)) HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(HELPER)) stop("Missing WT follow-up helper.")
source(HELPER)
suppressPackageStartupMessages({
  library(SeuratObject); library(Signac); library(Matrix)
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges); library(IRanges)
  library(patchwork)
})
need(identical(DATASET_VERSION, "v2_resequenced") &&
  identical(COHORT_ID, "wt"), "08K expects WT v2_resequenced config.")
RUN_COVERAGE <- TRUE
RUN_MOTIFS <- TRUE
BOOT_DRAWS <- 1000L
MOTIF_BACKGROUND_PER_TARGET <- 3L
MOTIF_MATCH_P <- 5e-5
MOTIF_PERM_DRAWS <- 5000L
RESULT_ROOT <- file.path(RESULTS_ROOT, DATASET_VERSION,
  "independent")
STEP8H <- file.path(OUTPUT_DIR,
  "08h_wt_ISC_region_freeze_sex_audit")
STEP8I <- file.path(OUTPUT_DIR,
  "08i_frozen_WT_peaks_genotype_context")
STEP8J <- file.path(OUTPUT_DIR,
  "08j_recount_frozen_WT_intervals")
OUT <- file.path(OUTPUT_DIR,
  "08k_wt_ISC_attenuation_motif_context")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
PARENTS <- list(ppar = c("3", "6"), il17 = c("9", "10"))
EXAMPLES <- tibble::tribble(
  ~feature, ~why,
  "chr3-96471088-96472340", "WT C0 effect in both Xist-detection strata; smaller knockout effects",
  "chr4-40142794-40143817", "smaller knockout effects; WT Xist-proxy sensitivity",
  "chr15-79166451-79167322", "smaller knockout effects; WT Xist-proxy sensitivity",
  "chr17-40158319-40159811", "pooled attenuation with opposing PPAR parent effects",
  "chr2-98497019-98497830", "retained diet response comparator")
```

## 1. Validate fixed inputs and the fragment-recount parity check

```{r inputs}
peak_file <- file.path(STEP8H,
  "02_all_622_top15_ATAC_peaks_with_evidence_flags.csv.gz")
effect_file <- file.path(STEP8J,
  "02_recounted_622_interval_genotype_effects.csv.gz")
parity_file <- file.path(STEP8J,
  "01_both_cohorts_fragment_recount_parity.csv")
wt_cell_file <- file.path(STEP8H,
  "01_nucleus_level_sex_proxy_and_gate_audit.csv.gz")
wt_sex_file <- file.path(STEP8H,
  "03_C0_supported_region_effects_by_Xist_detection.csv")
paths <- c(peak_file, effect_file, parity_file, wt_cell_file,
  wt_sex_file, file.path(STEP8J, "STEP_08J_COMPLETE.txt"),
  file.path(RESULT_ROOT, "wt", "06_annotation", "multiome_annotated.rds"))
for (cohort in names(PARENTS)) paths <- c(paths,
  file.path(STEP8I, paste0("02_", cohort,
    "_crypt_nuclei_scores.csv.gz")),
  file.path(RESULT_ROOT, cohort, "06_annotation",
    "multiome_annotated.rds"))
need(all(file.exists(paths)), paste("Missing 08K input(s):",
  paste(paths[!file.exists(paths)], collapse = "; ")))
if (RUN_MOTIFS) {
  motif_packages <- c("JASPAR2024", "TFBSTools", "motifmatchr",
    "Biostrings", "Rsamtools", "DBI", "RSQLite")
  missing_motif_packages <- motif_packages[!vapply(motif_packages,
    requireNamespace, logical(1), quietly = TRUE)]
  need(!length(missing_motif_packages), paste("Install motif packages:",
    paste(missing_motif_packages, collapse = ", ")))
  genome_path <- Sys.getenv("MULTIOME_GRCM39_FASTA",
    unset = file.path(PROJECT_DIR, "reference", "GRCm39_2024-A",
      "fasta", "genome.fa"))
  need(file.exists(genome_path) && file.exists(paste0(genome_path, ".fai")),
    paste("Indexed GRCm39 FASTA absent:", genome_path))
}
info <- file.info(paths)
manifest <- tibble(path = normalizePath(paths), bytes = info$size,
  modified = as.character(info$mtime))
write_table(manifest, OUT, "00_input_manifest.csv")
parity <- read_csv_checked(parity_file)
require_columns(parity, c("cohort", "requested_WT_intervals",
  "previous_exact_name_overlap", "original_shared_peak_count_mismatches",
  "recount_nonzero_intervals"), "08J parity")
need(setequal(parity$cohort, names(PARENTS)) && nrow(parity) == 2L &&
  all(parity$requested_WT_intervals == 622L) &&
  all(parity$recount_nonzero_intervals == 622L) &&
  all(parity$previous_exact_name_overlap >= 1L) &&
  all(parity$original_shared_peak_count_mismatches == 0L),
  "08J fragment recount parity failed; stop before comparing effects.")
anchor <- read_csv_checked(peak_file)
effects <- read_csv_checked(effect_file)
require_columns(anchor, c("feature", "chrom", "start_1based",
  "end_1based", "C0_supported", "count_supported_40",
  "nested_top10", "anchor_log2FC_HFD_vs_CON",
  "C0_all_log2FC_HFD_vs_CON", "anchor_CON_counts",
  "anchor_HFD_counts", "anchor_CON_percent"), "08H peaks")
require_columns(effects, c("cohort", "contrast", "feature", "eligible",
  "evaluable", "clear_effect", "log2FC_HFD_vs_CON", "CON_counts",
  "HFD_counts", "CON_percent", "HFD_percent"), "08J effects")
need(nrow(anchor) == 622L && !anyDuplicated(anchor$feature) &&
  sum(anchor$C0_supported) == 37L &&
  sum(anchor$C0_supported & anchor$count_supported_40) == 16L &&
  all(EXAMPLES$feature %in% anchor$feature),
  "The frozen WT interval set or example loci changed.")
need(!anyDuplicated(effects[c("cohort", "contrast", "feature")]) &&
  all(effects$feature %in% anchor$feature) &&
  setequal(effects$cohort, names(PARENTS)),
  "08J effect identifiers differ from the frozen WT intervals.")
intervals <- GenomicRanges::GRanges(seqnames = anchor$chrom,
  ranges = IRanges::IRanges(start = anchor$start_1based,
    end = anchor$end_1based))
names(intervals) <- anchor$feature
need(identical(paste0(as.character(GenomicRanges::seqnames(intervals)),
  "-", BiocGenerics::start(intervals), "-",
  BiocGenerics::end(intervals)), anchor$feature),
  "WT interval names and coordinates disagree.")
```

## 2. Region-level attenuation and parent-state checks

The output retains raw pooled counts, detection percentages, signed effects,
both parent effects, and eligibility. Categories are a reading aid; ranking
and plotting always retain continuous values.

```{r attenuation-table}
fields <- c("log2FC_HFD_vs_CON", "evaluable", "clear_effect",
  "CON_counts", "HFD_counts", "CON_percent", "HFD_percent")
add_contrast <- function(data, cohort, contrast, prefix) {
  x <- dplyr::filter(data, .data$cohort == .env$cohort,
    .data$contrast == .env$contrast)
  need(nrow(x) == nrow(anchor) &&
    setequal(x$feature, anchor$feature) && all(x$eligible),
    paste("Missing or ineligible genotype contrast:", cohort, contrast))
  x <- x[match(anchor$feature, x$feature), c("feature", fields)]
  names(x)[-1] <- paste0(prefix, "_", fields)
  x
}
wide <- dplyr::select(anchor, feature, chrom, start_1based, end_1based,
  C0_supported, count_supported_40, nested_top10,
  anchor_log2FC_HFD_vs_CON, C0_all_log2FC_HFD_vs_CON,
  anchor_CON_counts, anchor_HFD_counts, anchor_CON_percent)
for (cohort in names(PARENTS)) {
  keys <- c("annotated_union_all", "annotated_union_top15",
    "annotated_union_WTcal", paste0("parent_", PARENTS[[cohort]], "_all"))
  for (key in keys) {
    prefix <- paste(cohort, key, sep = "_")
    wide <- dplyr::left_join(wide,
      add_contrast(effects, cohort, key, prefix), by = "feature",
      relationship = "one-to-one")
  }
}
need(nrow(wide) == 622L && !anyNA(wide$ppar_annotated_union_all_evaluable),
  "Region-level join lost frozen intervals.")
for (cohort in names(PARENTS)) {
  for (gate in c("all", "top15", "WTcal")) {
    key <- paste(cohort, "annotated_union", gate, sep = "_")
    reference <- if (gate == "all")
      wide$C0_all_log2FC_HFD_vs_CON else
      wide$anchor_log2FC_HFD_vs_CON
    value <- wide[[paste0(key, "_log2FC_HFD_vs_CON")]]
    evaluable <- wide[[paste0(key, "_evaluable")]]
    need(all(is.finite(reference)) && all(abs(reference) > 0),
      "An attenuation reference is missing or zero.")
    wide[[paste0(key, "_magnitude_ratio_to_WT")]] <-
      ifelse(evaluable, abs(value) / abs(reference), NA_real_)
    wide[[paste0(key, "_direction_vs_WT")]] <-
      ifelse(evaluable, ifelse(value * reference > 0,
        "same", ifelse(value * reference < 0, "opposite", "zero")),
        "not_evaluable")
  }
}
for (gate in c("all", "top15", "WTcal")) {
  a <- wide[[paste0("ppar_annotated_union_", gate,
    "_magnitude_ratio_to_WT")]]
  b <- wide[[paste0("il17_annotated_union_", gate,
    "_magnitude_ratio_to_WT")]]
  wide[[paste0("both_", gate, "_below_WT")]] <-
    !is.na(a) & !is.na(b) & a < 1 & b < 1
  wide[[paste0("both_", gate, "_below_half_WT")]] <-
    !is.na(a) & !is.na(b) & a < .5 & b < .5
}
for (cohort in names(PARENTS)) {
  parent_values <- lapply(PARENTS[[cohort]], function(parent)
    wide[[paste0(cohort, "_parent_", parent,
      "_all_log2FC_HFD_vs_CON")]])
  wide[[paste0(cohort, "_parent_opposite_signs")]] <-
    parent_values[[1]] * parent_values[[2]] < 0
}
write_table(wide, OUT, "01_all_622_region_effects_and_attenuation.csv.gz")
write_table(dplyr::filter(wide, C0_supported), OUT,
  "01_C0_supported_37_region_effects_and_attenuation.csv")
write_table(dplyr::filter(wide, C0_supported, both_all_below_half_WT),
  OUT, "01_dual_attenuated_22_candidates.csv")
```

## 3. Descriptive magnitude summaries with region-resampling intervals

These percentile intervals resample **features**, not mice. They only convey
how the paired median across this selected interval set changes when regions
are left in or out; peaks may be correlated. They are intentionally labelled
as descriptive and never used as a diet or genotype interaction P value.

```{r summaries-and-plots}
set.seed(20260930)
tiers <- list(all_622 = rep(TRUE, nrow(wide)),
  C0_supported_37 = wide$C0_supported,
  count_supported_16 = wide$C0_supported & wide$count_supported_40)
summary_rows <- list()
for (tier in names(tiers)) for (gate in c("all", "top15", "WTcal")) {
  if (tier == "all_622" && gate == "all") next
  reference <- if (gate == "all")
    wide$C0_all_log2FC_HFD_vs_CON else
    wide$anchor_log2FC_HFD_vs_CON
  for (cohort in names(PARENTS)) {
    key <- paste(cohort, "annotated_union", gate, sep = "_")
    evaluable <- wide[[paste0(key, "_evaluable")]]
    index <- which(tiers[[tier]] & evaluable)
    effect <- wide[[paste0(key, "_log2FC_HFD_vs_CON")]][index]
    wt <- reference[index]
    need(length(index) >= 4L, paste("Insufficient evaluable regions:",
      tier, gate, cohort))
    delta <- abs(wt) - abs(effect)
    draw <- replicate(BOOT_DRAWS, stats::median(
      delta[sample.int(length(delta), length(delta), replace = TRUE)]))
    summary_rows[[length(summary_rows) + 1L]] <- tibble(
      tier = tier, cohort = cohort, gate = gate,
      reference_group = if (gate == "all") "WT_C0_all" else "WT_C0W0_top15",
      frozen_n = sum(tiers[[tier]]), evaluable_n = length(index),
      median_abs_WT_log2FC = stats::median(abs(wt)),
      median_abs_genotype_log2FC = stats::median(abs(effect)),
      median_paired_abs_reduction = stats::median(delta),
      peak_resampling_lo = unname(stats::quantile(draw, .025)),
      peak_resampling_hi = unname(stats::quantile(draw, .975)),
      smaller_magnitude_n = sum(abs(effect) < abs(wt)),
      below_half_magnitude_n = sum(abs(effect) < .5 * abs(wt)),
      same_direction_n = sum(wt * effect > 0),
      reverse_direction_n = sum(wt * effect < 0),
      clear_genotype_n = sum(wide[[paste0(key, "_clear_effect")]][index]))
  }
}
write_table(dplyr::bind_rows(summary_rows), OUT,
  "02_region_magnitude_descriptive_summary.csv")

primary <- dplyr::filter(wide, C0_supported) |>
  dplyr::select(feature, count_supported_40,
    C0_all_log2FC_HFD_vs_CON,
    ppar_annotated_union_all_log2FC_HFD_vs_CON,
    il17_annotated_union_all_log2FC_HFD_vs_CON) |>
  tidyr::pivot_longer(cols = c(ppar_annotated_union_all_log2FC_HFD_vs_CON,
    il17_annotated_union_all_log2FC_HFD_vs_CON),
    names_to = "cohort", values_to = "genotype_log2FC") |>
  dplyr::mutate(cohort = ifelse(grepl("^ppar", cohort), "PPAR", "IL17"))
p_magnitude <- ggplot2::ggplot(primary,
  ggplot2::aes(x = abs(C0_all_log2FC_HFD_vs_CON),
    y = abs(genotype_log2FC), shape = count_supported_40)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, color = "grey45") +
  ggplot2::geom_abline(slope = .5, intercept = 0, linetype = "dashed",
    color = "#B24B3A") +
  ggplot2::geom_point(size = 2, alpha = .8) +
  ggplot2::facet_wrap(~cohort) + ggplot2::coord_equal() +
  ggplot2::scale_shape_manual(values = c(`TRUE` = 19, `FALSE` = 1),
    labels = c(`TRUE` = "Higher-count 16", `FALSE` = "Other C0-supported")) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::labs(title = "WT C0 region effect magnitude in genotype ISC unions",
    subtitle = "37 fixed WT regions; dashed line is half the WT magnitude",
    x = "|WT C0 HFD-versus-CON log2FC|",
    y = "|Genotype HFD-versus-CON log2FC|", shape = NULL)
save_vector_plot(p_magnitude, PLOTS,
  "02_C0_supported_37_magnitude_by_genotype.pdf", 11, 6)

focus <- wide |> dplyr::filter(C0_supported) |>
  dplyr::select(feature, C0_all_log2FC_HFD_vs_CON,
    ppar_parent_3_all_log2FC_HFD_vs_CON,
    ppar_parent_6_all_log2FC_HFD_vs_CON,
    ppar_annotated_union_all_log2FC_HFD_vs_CON,
    il17_parent_9_all_log2FC_HFD_vs_CON,
    il17_parent_10_all_log2FC_HFD_vs_CON,
    il17_annotated_union_all_log2FC_HFD_vs_CON) |>
  tidyr::pivot_longer(-feature, names_to = "group", values_to = "log2FC")
focus$feature <- factor(focus$feature,
  levels = dplyr::filter(wide, C0_supported) |>
    dplyr::arrange(C0_all_log2FC_HFD_vs_CON) |>
    dplyr::pull(feature))
p_parent <- ggplot2::ggplot(focus,
  ggplot2::aes(x = feature, y = group, fill = log2FC)) +
  ggplot2::geom_tile(color = "white", linewidth = .2) +
  ggplot2::scale_fill_gradient2(low = "#3279A8", mid = "white",
    high = "#C45F48", midpoint = 0) +
  ggplot2::theme_bw(base_size = 9) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90,
    hjust = 1, vjust = .5, size = 6)) +
  ggplot2::labs(title = "Parent-specific and pooled effects at 37 WT regions",
    subtitle = "Opposing parent effects can yield a small union estimate",
    x = "WT interval", y = NULL, fill = "HFD vs CON\nlog2FC")
save_vector_plot(p_parent, PLOTS,
  "02_parent_and_union_effects_37_regions.pdf", 16, 5)
```

## 4. WT sex-proxy context and selected coverage views

Coverage uses the original indexed fragments and highlights the *same WT
interval* in each cohort. The cohorts have distinct peak calls, so their peak
tracks are omitted. Signac normalizes coverage within each plotted group;
compare CON with HFD **within a panel** and use the tabulated interval
effects for cross-cohort comparison. Coverage is visualization, not a
statistical test. The `chr17` example gets parent-specific tracks because
its union effect masks opposing parent responses.

```{r coverage}
sex_wt <- read_csv_checked(wt_sex_file)
require_columns(sex_wt, c("group", "Xist_status", "feature",
  "CON_n", "HFD_n", "evaluable", "clear_effect",
  "log2FC_HFD_vs_CON"), "08H Xist sensitivity")
sex_wt <- dplyr::filter(sex_wt, group == "C0_all")
write_table(dplyr::filter(sex_wt,
  feature %in% EXAMPLES$feature), OUT,
  "03_example_WT_Xist_detection_sensitivity.csv")
write_table(EXAMPLES, OUT, "03_prespecified_coverage_examples.csv")
wt_cells <- read_csv_checked(wt_cell_file)
require_columns(wt_cells, c("cell_barcode", "diet", "focused_cluster",
  "C0W0_top15"), "08H saved WT cell gates")
need(!anyDuplicated(wt_cells$cell_barcode), "Duplicate WT cell barcode.")
coverage_status <- list()
wt_annotation <- NULL
for (cohort in c("wt", names(PARENTS))) {
  obj <- readRDS(file.path(RESULT_ROOT, cohort, "06_annotation",
    "multiome_annotated.rds"))
  need("ATAC" %in% names(obj) &&
    length(Signac::Fragments(obj[["ATAC"]])) > 0L,
    paste(cohort, "has no indexed ATAC fragments for coverage."))
  meta <- obj[[]]
  meta$cell_barcode <- rownames(meta)
  meta$diet <- as.character(meta$diet)
  need(!anyDuplicated(meta$cell_barcode) &&
    setequal(unique(meta$diet), c("CON", "HFD")),
    paste(cohort, "invalid cell barcodes or diet labels."))
  if (cohort == "wt") {
    wt_annotation <- Signac::Annotation(obj[["ATAC"]])
    ix <- match(wt_cells$cell_barcode, meta$cell_barcode)
    need(!anyNA(ix) && identical(wt_cells$diet, meta$diet[ix]),
      "08H WT cells do not match the annotated object.")
    group_cells <- list(
      C0_all = wt_cells$cell_barcode[
        as.character(wt_cells$focused_cluster) == "0"],
      C0W0_top15 = wt_cells$cell_barcode[wt_cells$C0W0_top15])
  } else {
    scores <- read_csv_checked(file.path(STEP8I,
      paste0("02_", cohort, "_crypt_nuclei_scores.csv.gz")))
    require_columns(scores, c("cell_barcode", "diet", "wnn_cluster",
      "top15"), paste(cohort, "08I fixed gate table"))
    ix <- match(scores$cell_barcode, meta$cell_barcode)
    need(!anyNA(ix) && !anyDuplicated(scores$cell_barcode) &&
      identical(scores$diet, meta$diet[ix]) &&
      identical(as.character(scores$wnn_cluster),
        as.character(meta$wnn_cluster[ix])) &&
      !anyNA(scores$top15),
      paste(cohort, "08I scores disagree with annotated cells."))
    in_parent <- as.character(scores$wnn_cluster) %in% PARENTS[[cohort]]
    group_cells <- list(
      annotated_union_all = scores$cell_barcode[in_parent],
      annotated_union_top15 = scores$cell_barcode[
        in_parent & scores$top15])
    for (parent in PARENTS[[cohort]])
      group_cells[[paste0("parent_", parent, "_all")]] <-
        scores$cell_barcode[as.character(scores$wnn_cluster) == parent]
  }
  for (key in names(group_cells)) {
    cells <- group_cells[[key]]
    n_con <- sum(meta$diet[match(cells, meta$cell_barcode)] == "CON")
    n_hfd <- sum(meta$diet[match(cells, meta$cell_barcode)] == "HFD")
    need(length(cells) && !anyDuplicated(cells) &&
      n_con >= 30L && n_hfd >= 30L,
      paste(cohort, key, "has an ineligible coverage group."))
    selected <- if (grepl("^parent_", key))
      "chr17-40158319-40159811" else EXAMPLES$feature
    for (feature in selected) {
      status <- "not_attempted"; detail <- ""
      if (RUN_COVERAGE) {
        gr <- intervals[match(feature, names(intervals))]
        span <- GenomicRanges::GRanges(
          seqnames = as.character(GenomicRanges::seqnames(gr)),
          ranges = IRanges::IRanges(
            start = max(1L, as.integer(BiocGenerics::start(gr)) - 5000L),
            end = as.integer(BiocGenerics::end(gr)) + 5000L))
        tryCatch({
          p <- Signac::CoveragePlot(obj, region = span,
            assay = "ATAC", cells = cells, annotation = "gene",
            peaks = FALSE, region.highlight = gr, links = FALSE,
            tile = FALSE, group.by = "diet", ymax = "q95") +
            patchwork::plot_annotation(title = paste(cohort, key, feature),
              subtitle = paste("CON", n_con, "/ HFD", n_hfd,
                "nuclei; identical WT interval highlighted"))
          ggplot2::ggsave(file.path(PLOTS,
            paste0("04_", cohort, "_", key, "_", feature, ".pdf")),
            plot = p, device = grDevices::pdf, useDingbats = FALSE,
            width = 13, height = 6)
          status <- "saved"
        }, error = function(e) {
          status <<- "plot_failed"
          detail <<- conditionMessage(e)
          warning("Coverage failed for ", cohort, " / ", key,
            " / ", feature, ": ", detail)
        })
      }
      coverage_status[[length(coverage_status) + 1L]] <- tibble(
        cohort = cohort, group = key, feature = feature,
        n_CON = n_con, n_HFD = n_hfd, status = status,
        detail = detail)
    }
  }
  rm(obj, meta, group_cells); invisible(gc())
}
coverage_status <- dplyr::bind_rows(coverage_status)
write_table(coverage_status, OUT, "04_coverage_status.csv")
```

## 5. Explore motif recurrence and enrichment among attenuated intervals

Target: the fixed WT C0-supported regions whose **full parent union** effect
is evaluable and less than half the WT C0 magnitude in *both* genotype
cohorts. Target definition uses no DNA sequence information. Background:
other intervals in the 622 WT top-15 effect universe, matched to target
on WT effect direction, promoter/gene-body context where available, GC,
width, and WT CON detection. Context mismatches are explicitly audited.
A motif comparison is conditional on this selected WT universe;
it cannot identify the causal TF, and the small target set limits power.

```{r motif-context}
if (RUN_MOTIFS) {
  target <- wide$feature[wide$C0_supported & wide$both_all_below_half_WT]
  need(length(target) == 22L,
    "Dual-attenuation target count changed from the reviewed 22; inspect Step 08J before motif analysis.")
  fa <- Rsamtools::FaFile(genome_path)
  index <- Rsamtools::scanFaIndex(fa)
  seq_width <- setNames(as.integer(BiocGenerics::width(index)),
    as.character(GenomicRanges::seqnames(index)))
  need(all(anchor$chrom %in% names(seq_width)) &&
    all(anchor$end_1based <= seq_width[anchor$chrom]),
    "Frozen WT coordinates do not fit the supplied FASTA index.")
  Rsamtools::open(fa)
  seqs <- tryCatch(Rsamtools::scanFa(fa, param = intervals),
    finally = Rsamtools::close(fa))
  need(length(seqs) == nrow(anchor) &&
    all(as.integer(BiocGenerics::width(seqs)) ==
      anchor$end_1based - anchor$start_1based + 1L),
    "FASTA sequence retrieval did not preserve interval widths.")
  names(seqs) <- anchor$feature
  gc <- 100 * rowSums(Biostrings::letterFrequency(seqs,
    letters = c("G", "C"), as.prob = TRUE))
  need(length(wt_annotation) > 0L,
    "WT Step 06 assay has no gene annotation for motif matching.")
  promoters <- GenomicRanges::promoters(wt_annotation,
    upstream = 2000L, downstream = 2000L)
  in_promoter <- GenomicRanges::countOverlaps(intervals,
    promoters, ignore.strand = TRUE) > 0
  in_gene <- GenomicRanges::countOverlaps(intervals, wt_annotation,
    ignore.strand = TRUE) > 0
  covars <- tibble(feature = anchor$feature,
    gc_percent = as.numeric(gc),
    peak_width = as.integer(BiocGenerics::width(intervals)),
    wt_CON_detection_percent = anchor$anchor_CON_percent,
    genomic_context = dplyr::case_when(
      in_promoter ~ "promoter", in_gene ~ "gene_body",
      TRUE ~ "intergenic"),
    WT_direction = ifelse(anchor$anchor_log2FC_HFD_vs_CON > 0,
      "HFD_open", "HFD_closed"))
  need(all(is.finite(covars$gc_percent)) &&
    all(is.finite(covars$wt_CON_detection_percent)) &&
    all(covars$peak_width > 0L), "Invalid motif matching covariate.")
  write_table(covars, OUT, "05_all_622_motif_sequence_covariates.csv")
  # Greedy without-replacement nearest matching, same direction.
  # Prefer the same genomic context, with an audited fallback.
  v <- cbind(gc = covars$gc_percent,
    log_width = log1p(covars$peak_width),
    CON_detection = covars$wt_CON_detection_percent)
  center <- colMeans(v)
  spread <- apply(v, 2, stats::sd)
  spread[!is.finite(spread) | spread == 0] <- 1
  z <- sweep(sweep(v, 2, center, "-"), 2, spread, "/")
  rownames(z) <- covars$feature
  candidates <- covars$feature[!covars$feature %in% target]
  # Match scarce direction/context strata first.
  target_rows <- covars[match(target, covars$feature), ]
  available_by_stratum <- vapply(seq_len(nrow(target_rows)), function(j)
    sum(covars$WT_direction[covars$feature %in% candidates] ==
      target_rows$WT_direction[[j]] &
      covars$genomic_context[covars$feature %in% candidates] ==
      target_rows$genomic_context[[j]]), integer(1))
  target_order <- order(available_by_stratum, target)
  matches <- list()
  for (j in target_order) {
    f <- target[[j]]
    rr <- covars[match(f, covars$feature), ]
    pool <- candidates[covars$WT_direction[
      match(candidates, covars$feature)] == rr$WT_direction]
    same_context <- covars$genomic_context[
      match(pool, covars$feature)] == rr$genomic_context
    if (sum(same_context) >= MOTIF_BACKGROUND_PER_TARGET)
      pool <- pool[same_context]
    need(length(pool) >= MOTIF_BACKGROUND_PER_TARGET,
      paste("Insufficient same-direction background for", f))
    distance <- rowSums((z[pool, , drop = FALSE] -
      matrix(z[f, ], nrow = length(pool), ncol = ncol(z),
        byrow = TRUE))^2) +
      4 * (covars$genomic_context[
        match(pool, covars$feature)] != rr$genomic_context)
    chosen <- pool[order(distance, pool)[seq_len(
      MOTIF_BACKGROUND_PER_TARGET)]]
    matches[[length(matches) + 1L]] <- tibble(
      target_feature = f, background_feature = chosen,
      context_matched = covars$genomic_context[
        match(chosen, covars$feature)] == rr$genomic_context,
      matching_distance = distance[match(chosen, pool)])
    candidates <- setdiff(candidates, chosen)
  }
  match_table <- dplyr::bind_rows(matches)
  background <- match_table$background_feature
  need(!anyDuplicated(background) &&
    length(background) == length(target) * MOTIF_BACKGROUND_PER_TARGET,
    "Motif matched background has duplicates or wrong size.")
  write_table(match_table, OUT, "05_target_background_matches.csv")
  membership <- dplyr::bind_rows(
    tibble(feature = target, membership = "target"),
    tibble(feature = background, membership = "background")) |>
    dplyr::left_join(covars, by = "feature",
      relationship = "many-to-one")
  write_table(membership, OUT, "05_motif_membership_and_covariates.csv")
  balance <- membership |>
    dplyr::group_by(membership, WT_direction, genomic_context) |>
    dplyr::summarise(n = dplyr::n(),
      median_gc = stats::median(gc_percent),
      median_width = stats::median(peak_width),
      median_WT_CON_detection = stats::median(wt_CON_detection_percent),
      .groups = "drop")
  write_table(balance, OUT, "05_motif_background_balance.csv")
  print(balance)

  jaspar_db <- JASPAR2024::JASPAR2024()
  jaspar_conn <- DBI::dbConnect(RSQLite::SQLite(),
    methods::slot(jaspar_db, "db"))
  jaspar_pfms <- tryCatch(TFBSTools::getMatrixSet(jaspar_conn,
    opts = list(collection = "CORE", tax_group = "vertebrates",
      all_versions = FALSE)), finally = DBI::dbDisconnect(jaspar_conn))
  need(length(jaspar_pfms) > 0L, "JASPAR returned no CORE vertebrate motifs.")
  motif_meta <- tibble(motif_id = vapply(jaspar_pfms,
    TFBSTools::ID, character(1)),
    motif_name = vapply(jaspar_pfms, TFBSTools::name,
      character(1)),
    motif_family = vapply(jaspar_pfms, function(m)
      paste(TFBSTools::tags(m)$family, collapse = ";"), character(1)))
  simple_symbol <- grepl("^[A-Za-z][A-Za-z0-9]+$",
    motif_meta$motif_name)
  motif_meta$candidate_mouse_symbol <- NA_character_
  motif_meta$candidate_mouse_symbol[simple_symbol] <- paste0(
    toupper(substr(motif_meta$motif_name[simple_symbol], 1, 1)),
    tolower(substr(motif_meta$motif_name[simple_symbol], 2,
      nchar(motif_meta$motif_name[simple_symbol]))))
  hits <- motifmatchr::motifMatches(motifmatchr::matchMotifs(
    pwms = jaspar_pfms, subject = seqs,
    bg = "subject", out = "matches", p.cutoff = MOTIF_MATCH_P))
  need(nrow(hits) == nrow(anchor), "Motif matrix lost WT intervals.")
  rownames(hits) <- anchor$feature
  motif_id <- vapply(colnames(hits), function(col) {
    ids <- motif_meta$motif_id[startsWith(col, motif_meta$motif_id)]
    if (!length(ids)) return(NA_character_)
    ids[[which.max(nchar(ids))]]
  }, character(1))
  need(!anyNA(motif_id), "Motif hit IDs cannot be reconciled with JASPAR.")
  motif_meta <- dplyr::left_join(tibble(motif_column = colnames(hits),
    motif_id = motif_id), motif_meta, by = "motif_id",
    relationship = "many-to-one")
  target_hits <- as.integer(Matrix::colSums(hits[target, , drop = FALSE]))
  bg_hits <- as.integer(Matrix::colSums(hits[background, , drop = FALSE]))
  # Permute one target label among the four matched sequences per locus.
  # This tests sequence over-representation conditional on matched sets.
  order4 <- unlist(lapply(target, function(f)
    c(f, match_table$background_feature[
      match_table$target_feature == f])), use.names = FALSE)
  need(length(order4) == 4L * length(target) &&
    !anyDuplicated(order4), "Motif matched quartets are invalid.")
  quartet_hits <- hits[order4, , drop = FALSE]
  set.seed(20260930)
  null_counts <- matrix(0L, nrow = MOTIF_PERM_DRAWS,
    ncol = ncol(hits))
  for (b in seq_len(MOTIF_PERM_DRAWS)) {
    selected_rows <- 4L * (seq_along(target) - 1L) +
      sample.int(4L, length(target), replace = TRUE)
    null_counts[b, ] <- as.integer(Matrix::colSums(
      quartet_hits[selected_rows, , drop = FALSE]))
  }
  matched_permutation_p <- (1 + colSums(sweep(null_counts, 2,
    target_hits, FUN = ">="))) / (MOTIF_PERM_DRAWS + 1)
  motif_stats <- motif_meta |>
    dplyr::mutate(target_n = length(target), background_n = length(background),
      target_hits = target_hits, background_hits = bg_hits,
      target_percent = 100 * target_hits / target_n,
      background_percent = 100 * background_hits / background_n,
      odds_ratio_Haldane = ((target_hits + .5) *
        (background_n - background_hits + .5)) /
        ((target_n - target_hits + .5) * (background_hits + .5)),
      matched_permutation_p = matched_permutation_p,
      motif_BH = stats::p.adjust(matched_permutation_p, method = "BH")) |>
    dplyr::arrange(motif_BH, matched_permutation_p,
      dplyr::desc(target_hits))
  write_table(motif_stats, OUT, "05_JASPAR2024_motif_enrichment_exploratory.csv")
  motif_presence <- tibble(feature = target,
    matched_motif_ids = vapply(target, function(f)
      paste(motif_meta$motif_id[as.logical(hits[f, ])],
        collapse = ";"), character(1)))
  write_table(motif_presence, OUT,
    "05_22_attenuated_regions_motif_presence.csv")
  plot_motifs <- motif_stats |>
    dplyr::filter(target_hits >= 4) |>
    dplyr::arrange(dplyr::desc(target_percent - background_percent),
      matched_permutation_p) |>
    dplyr::slice_head(n = 20L)
  if (nrow(plot_motifs)) {
    p_motif <- ggplot2::ggplot(plot_motifs,
      ggplot2::aes(x = stats::reorder(paste(motif_name, motif_id),
        target_percent - background_percent),
        y = target_percent - background_percent,
        size = target_hits, color = motif_BH)) +
      ggplot2::geom_point() + ggplot2::coord_flip() +
      ggplot2::scale_color_gradient(low = "#AE4D43", high = "#3B718F") +
      ggplot2::theme_bw(base_size = 10) +
      ggplot2::labs(title = "Candidate motifs in WT regions with smaller knockout effects",
        subtitle = "22 targets versus 66 matched WT peaks; motif permutation BH, not diet inference",
        x = NULL, y = "Target minus background motif prevalence (percentage points)",
        size = "Targets with motif", color = "Motif BH")
    save_vector_plot(p_motif, PLOTS,
      "05_exploratory_motif_prevalence.pdf", 12, 7)
  }
}
```

## 6. Read the results in order

1. `01_C0_supported_37_region_effects_and_attenuation.csv` gives each
   region's signed effect, detection, count support, magnitude ratio and
   parent-state effects. The 22-row table is an **exploratory subset**.
2. `02_region_magnitude_descriptive_summary.csv` and two PDFs show the
   continuous effect changes. Feature-resampling intervals are descriptive.
3. Check `04_coverage_status.csv`, then view CON/HFD coverage within each
   cohort. `chr17` illustrates why a flat pooled estimate need not mean
   loss in every parent state.
4. Inspect `05_motif_background_balance.csv` before any motif ranking. A
   shared motif suggests a candidate family; check TF expression, motif
   ambiguity, chromatin context, and eventually biological replication.

```{r finish}
finish_run(OUT, "STEP_08K_COMPLETE.txt", c(
  "Frozen 08H WT regions, fixed 08I score gates, and validated 08J fragment recount.",
  "Descriptive effects and feature-resampling intervals; no diet or genotype interaction P values.",
  "Matched-set motif permutation/BH values describe sequence enrichment only, when RUN_MOTIFS=TRUE.",
  paste("Results:", OUT)))
```
