---
title: "Step 08L | WT ISC HFD-opening and HFD-closing regions"
output:
  html_notebook:
    toc: true
    toc_float: true
---

Run from the repository root in the SOL R 4.4 session after Steps 08G, 08H,
08J and 08K. This analysis uses the frozen 622 WT top-15% UCell C0/W0
candidate intervals. HFD-opening and HFD-closing refer to the sign of the
**pooled, depth-normalized WT ATAC effect**, not a replicate-level DAR test.
The independently specified C0-supported 37 and higher-count 16 are
sensitivity subsets; neither subset redefines the 622. Other evaluable WT
top-15% consensus peaks provide a neutral, accessible sequence background.
Motif tests ask whether DNA motifs are more frequent in candidate regions
than in matched neutral regions. They do **not** test TF binding, TF activity,
or a diet-by-genotype interaction. One pooled library per diet prevents
mouse-level statistical inference. Use effect sizes, parent estimates, depth,
sex-proxy sensitivity, and coverage when selecting illustrative loci.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Run Step 08L from the repository root.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
if (!file.exists(HELPER)) HELPER <- file.path(REPO_ROOT, "analysis", "R",
  "wt_crypt_followup_helpers.R")
source(HELPER)
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(tibble); library(readr)
  library(ggplot2); library(GenomicRanges); library(IRanges)
  library(Matrix); library(Signac); library(SeuratObject)
})
motif_pkgs <- c("JASPAR2024", "TFBSTools", "motifmatchr", "Biostrings",
  "Rsamtools", "DBI", "RSQLite")
need(all(vapply(motif_pkgs, requireNamespace, logical(1), quietly = TRUE)),
  paste("Missing packages:", paste(motif_pkgs[!vapply(motif_pkgs,
    requireNamespace, logical(1), quietly = TRUE)], collapse = ", ")))
STEP8G <- file.path(OUTPUT_DIR, "08g_C0_C2_and_C0W0_ISC")
STEP8H <- file.path(OUTPUT_DIR, "08h_wt_ISC_region_freeze_sex_audit")
STEP8J <- file.path(OUTPUT_DIR, "08j_recount_frozen_WT_intervals")
STEP8K <- file.path(OUTPUT_DIR, "08k_wt_ISC_attenuation_motif_context")
OUT <- file.path(OUTPUT_DIR, "08l_wt_ISC_open_close_motifs")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
FASTA <- Sys.getenv("MULTIOME_GRCM39_FASTA", unset = file.path(
  PROJECT_DIR, "reference", "GRCm39_2024-A", "fasta", "genome.fa"))
MOTIF_P <- 5e-5
MATCH_Z_LIMIT <- 1.5  # hard limit for EACH of GC, log width and CON detection
NEUTRAL_ABS_LOG2FC <- .25
```

## 1. Load fixed candidate intervals and the evaluable WT peak universe

```{r inputs}
files <- c(
  frozen = file.path(STEP8H,
    "02_all_622_top15_ATAC_peaks_with_evidence_flags.csv.gz"),
  effects = file.path(STEP8K,
    "01_all_622_region_effects_and_attenuation.csv.gz"),
  universe = file.path(STEP8G,
    "04_ATAC_evaluable_pooled_effects.csv.gz"),
  bed = file.path(STEP8G, "09_evaluable_unmatched_universe.bed"),
  sex = file.path(STEP8H,
    "03_C0_supported_region_effects_by_Xist_detection.csv"),
  fasta = FASTA, fai = paste0(FASTA, ".fai"),
  wt_object = file.path(OUTPUT_DIR, "06_annotation", "multiome_annotated.rds"),
  complete08j = file.path(STEP8J, "STEP_08J_COMPLETE.txt"),
  complete08k = file.path(STEP8K, "STEP_08K_COMPLETE.txt"))
need(all(file.exists(files)), paste("Missing inputs:",
  paste(files[!file.exists(files)], collapse = "; ")))
file_info <- file.info(files)
write_table(tibble(input = names(files), path = normalizePath(files),
  bytes = file_info$size, modified = as.character(file_info$mtime)),
  OUT, "00_input_manifest.csv")
frozen <- read_csv_checked(files[["frozen"]])
wide <- read_csv_checked(files[["effects"]])
need(nrow(frozen) == 622L && !anyDuplicated(frozen$feature) &&
  setequal(frozen$feature, wide$feature) && nrow(wide) == 622L,
  "08H/08K frozen interval identity changed.")
require_columns(wide, c("feature", "chrom", "start_1based", "end_1based",
  "anchor_log2FC_HFD_vs_CON", "C0_all_log2FC_HFD_vs_CON",
  "C0_supported", "count_supported_40", "nested_top10",
  "both_all_below_half_WT", "ppar_parent_opposite_signs",
  "il17_parent_opposite_signs"), "08K region effects")
need(all(wide$feature == frozen$feature) &&
  isTRUE(all.equal(wide$anchor_log2FC_HFD_vs_CON,
    frozen$anchor_log2FC_HFD_vs_CON, tolerance = 1e-9)),
  "08K WT anchor effects differ from 08H.")
wide$WT_direction <- ifelse(wide$anchor_log2FC_HFD_vs_CON > 0,
  "HFD_open", "HFD_closed")
need(!anyNA(wide$WT_direction) &&
  all(wide$anchor_log2FC_HFD_vs_CON != 0),
  "Candidate WT directions are missing.")
print(wide |> dplyr::count(WT_direction, C0_supported,
  count_supported_40, name = "regions"))

# Read only needed columns from the Step 08G pooled-effects table.
all_effects <- readr::read_csv(files[["universe"]],
  col_select = c("contrast", "feature", "evaluable", "clear_effect",
    "log2FC_HFD_vs_CON", "CON_percent", "HFD_percent",
    "CON_counts", "HFD_counts"), show_col_types = FALSE,
  progress = FALSE)
pool <- all_effects |>
  dplyr::filter(.data$contrast == "C0W0_top15", .data$evaluable)
rm(all_effects); invisible(gc())
need(!anyDuplicated(pool$feature) && nrow(pool) > nrow(wide) &&
  all(wide$feature %in% pool$feature),
  "Evaluable WT C0/W0 peak universe is missing frozen candidates.")
verify <- pool[match(wide$feature, pool$feature), ]
need(max(abs(verify$log2FC_HFD_vs_CON -
  wide$anchor_log2FC_HFD_vs_CON)) < 1e-7 &&
  all(verify$clear_effect), "08G effects differ from frozen WT effects.")
bed <- readr::read_tsv(files[["bed"]], col_names = c("chrom", "start0",
  "end", "feature"), col_types = "ciic", show_col_types = FALSE)
need(!anyDuplicated(bed$feature) && setequal(bed$feature, pool$feature),
  "08G evaluable BED and pooled effect universe disagree.")
bed <- bed[match(pool$feature, bed$feature), ]
need(identical(bed$feature, pool$feature), "BED alignment failed.")
pool$chrom <- bed$chrom
pool$start_1based <- bed$start0 + 1L
pool$end_1based <- bed$end
ix <- match(wide$feature, pool$feature)
need(all(pool$chrom[ix] == wide$chrom &
  pool$start_1based[ix] == wide$start_1based &
  pool$end_1based[ix] == wide$end_1based),
  "08G and 08H peak coordinates disagree.")
```

## 2. Direction, C0 support, and genotype behavior

This table is the region-level result. A genotype union can have a small
effect if its two parent clusters move in opposite directions. Keep all
four parent estimates in the CSV. A 'both below half' label compares
pooled magnitudes with WT C0 and is descriptive.

```{r region-table}
region <- dplyr::left_join(wide, pool |>
  dplyr::select(feature, anchor_HFD_percent = HFD_percent),
  by = "feature", relationship = "one-to-one")
region$both_parents_below_half_C0 <- vapply(seq_len(nrow(region)),
  function(i) {
    reference <- abs(region$C0_all_log2FC_HFD_vs_CON[i])
    parents <- c("ppar_parent_3_all", "ppar_parent_6_all",
      "il17_parent_9_all", "il17_parent_10_all")
    all(vapply(parents, function(prefix) {
      isTRUE(region[[paste0(prefix, "_evaluable")]][i]) &&
        abs(region[[paste0(prefix, "_log2FC_HFD_vs_CON")]][i]) <
          .5 * reference
    }, logical(1)))
  }, logical(1))
write_table(region, OUT, "01_all_622_open_close_genotype_and_support.csv.gz")
strata <- list(all_622 = rep(TRUE, nrow(region)),
  C0_supported_37 = region$C0_supported,
  higher_count_C0_16 = region$C0_supported & region$count_supported_40)
direction_summary <- dplyr::bind_rows(lapply(names(strata), function(key) {
  region[strata[[key]], , drop = FALSE] |>
    dplyr::group_by(WT_direction) |>
    dplyr::summarise(set = key, n_regions = dplyr::n(),
      n_both_union_below_half = sum(both_all_below_half_WT),
      n_both_parent_below_half = sum(both_parents_below_half_C0),
      n_ppar_opposing_parents = sum(ppar_parent_opposite_signs),
      n_il17_opposing_parents = sum(il17_parent_opposite_signs),
      median_WT_anchor_abs_log2FC = median(abs(anchor_log2FC_HFD_vs_CON)),
      .groups = "drop")
})) |> dplyr::select(set, dplyr::everything())
write_table(direction_summary, OUT, "01_direction_and_attenuation_summary.csv")
print(direction_summary)
sex <- read_csv_checked(files[["sex"]])
write_table(sex, OUT, "01_C0_regions_Xist_detection_sensitivity.csv")
p <- ggplot2::ggplot(region,
  ggplot2::aes(x = anchor_log2FC_HFD_vs_CON,
    y = C0_all_log2FC_HFD_vs_CON, color = C0_supported)) +
  ggplot2::geom_hline(yintercept = 0, linewidth = .2) +
  ggplot2::geom_vline(xintercept = 0, linewidth = .2) +
  ggplot2::geom_point(alpha = .55, size = 1.5) +
  ggplot2::facet_wrap(~ WT_direction) +
  ggplot2::scale_color_manual(values = c(`TRUE` = "#B45142",
    `FALSE` = "#AAB6B7")) + ggplot2::theme_bw() +
  ggplot2::labs(title = "WT ISC candidate region effects by direction",
    subtitle = "622 top-15% candidates; C0 support is a sensitivity flag",
    x = "WT C0/W0 top-15 pooled ATAC log2FC",
    y = "WT C0 pooled ATAC log2FC")
save_vector_plot(p, PLOTS, "01_WT_open_close_C0_support.pdf", 12, 5)
```

## 3. Define neutral accessible backgrounds and audit matchability

Match a distinct neutral region for each candidate within the same genomic
context, with hard limits on standardized GC, log(width), and WT CON
detection. Neutral means evaluable, `abs(log2FC) <= 0.25`, and not one of
the 622; it does not imply a replicate-verified null. An unmatched target
stays visible in the audit but does not enter a motif test. We compare one
target with one matched neutral region, allowing exact paired discordance
tests. If fewer than 80% of a direction matches, report its motif analysis
as insufficiently matched.

```{r sequence-and-matching}
pool$width <- pool$end_1based - pool$start_1based + 1L
gr <- GenomicRanges::GRanges(seqnames = pool$chrom,
  ranges = IRanges::IRanges(pool$start_1based, pool$end_1based))
names(gr) <- pool$feature
fa <- Rsamtools::FaFile(FASTA)
index <- Rsamtools::scanFaIndex(fa)
seq_width <- setNames(as.integer(BiocGenerics::width(index)),
  as.character(GenomicRanges::seqnames(index)))
valid <- pool$chrom %in% names(seq_width) &
  pool$start_1based >= 1 &
  pool$end_1based <= seq_width[pool$chrom] & pool$width > 0L
need(all(valid[match(wide$feature, pool$feature)]),
  "A frozen candidate lies outside the GRCm39 FASTA.")
pool <- pool[which(valid), ]
gr <- gr[which(valid)]
seqs <- Rsamtools::scanFa(fa, param = gr)
need(length(seqs) == nrow(pool) &&
  all(as.integer(BiocGenerics::width(seqs)) == pool$width),
  "FASTA sequence widths do not match WT peak widths.")
names(seqs) <- pool$feature
pool$gc_percent <- 100 * rowSums(Biostrings::letterFrequency(seqs,
  letters = c("G", "C"), as.prob = TRUE))
need(all(is.finite(pool$gc_percent)), "Invalid GC percentage.")
obj <- readRDS(files[["wt_object"]])
annotation <- Signac::Annotation(obj[["ATAC"]])
need(length(annotation) > 0L, "WT annotation unavailable.")
rm(obj); invisible(gc())
promoters <- GenomicRanges::promoters(annotation,
  upstream = 2000L, downstream = 2000L)
pool$context <- dplyr::case_when(
  GenomicRanges::countOverlaps(gr, promoters, ignore.strand = TRUE) > 0 ~
    "promoter",
  GenomicRanges::countOverlaps(gr, annotation, ignore.strand = TRUE) > 0 ~
    "gene_body", TRUE ~ "intergenic")
target_gr <- gr[match(wide$feature, names(gr))]
target_vicinity <- GenomicRanges::GRanges(
  seqnames = as.character(GenomicRanges::seqnames(target_gr)),
  ranges = IRanges::IRanges(
    start = pmax(1L, BiocGenerics::start(target_gr) - 1000L),
    end = BiocGenerics::end(target_gr) + 1000L))
pool$within_1kb_frozen <- GenomicRanges::countOverlaps(gr,
  target_vicinity, ignore.strand = TRUE) > 0
pool$membership <- dplyr::case_when(
  pool$feature %in% wide$feature[wide$WT_direction == "HFD_open"] ~
    "HFD_open",
  pool$feature %in% wide$feature[wide$WT_direction == "HFD_closed"] ~
    "HFD_closed",
  pool$within_1kb_frozen ~ "near_frozen_excluded",
  !(pool$feature %in% wide$feature) &
    abs(pool$log2FC_HFD_vs_CON) <= NEUTRAL_ABS_LOG2FC ~ "neutral",
  TRUE ~ "other")
write_table(pool |>
  dplyr::select(feature, chrom, start_1based, end_1based,
    log2FC_HFD_vs_CON, CON_percent, HFD_percent, CON_counts,
    HFD_counts, width, gc_percent, context, within_1kb_frozen,
    membership), OUT,
  "02_evaluable_WT_universe_sequence_covariates.csv.gz")
z <- cbind(gc_percent = pool$gc_percent,
  log_width = log1p(pool$width), CON_percent = pool$CON_percent)
center <- colMeans(z)
spread <- apply(z, 2, stats::sd)
need(all(is.finite(spread) & spread > 0), "Invalid matching variances.")
z <- sweep(sweep(z, 2, center, "-"), 2, spread, "/")
rownames(z) <- pool$feature
targets <- wide$feature
remaining <- pool$feature[pool$membership == "neutral"]
# Match the scarcest context first, so common promoter peaks do not consume
# the few available intergenic controls. No context fallback or caliper relaxation.
target_order <- order(vapply(targets, function(f) {
  row <- match(f, pool$feature)
  sum(pool$context[match(remaining, pool$feature)] == pool$context[row])
}, integer(1)), targets)
match_rows <- vector("list", length(target_order))
for (k in seq_along(target_order)) {
  f <- targets[target_order[k]]
  i <- match(f, pool$feature)
  candidates <- remaining[pool$context[match(remaining, pool$feature)] ==
    pool$context[i]]
  if (length(candidates)) {
    delta <- abs(sweep(z[candidates, , drop = FALSE], 2, z[f, ], "-"))
    good <- apply(delta <= MATCH_Z_LIMIT, 1, all)
    candidates <- candidates[good]
    delta <- delta[good, , drop = FALSE]
  }
  if (length(candidates)) {
    dist <- rowSums(delta^2)
    selected <- candidates[order(dist, candidates)][1]
    j <- match(selected, pool$feature)
    match_rows[[k]] <- tibble(target_feature = f,
      background_feature = selected, WT_direction = wide$WT_direction[
        match(f, wide$feature)], target_context = pool$context[i],
      background_context = pool$context[j],
      distance = dist[match(selected, candidates)],
      delta_gc_sd = delta[selected, "gc_percent"],
      delta_log_width_sd = delta[selected, "log_width"],
      delta_CON_detection_sd = delta[selected, "CON_percent"])
    remaining <- setdiff(remaining, selected)
  } else {
    match_rows[[k]] <- tibble(target_feature = f,
      background_feature = NA_character_, WT_direction = wide$WT_direction[
        match(f, wide$feature)], target_context = pool$context[i],
      background_context = NA_character_, distance = NA_real_,
      delta_gc_sd = NA_real_, delta_log_width_sd = NA_real_,
      delta_CON_detection_sd = NA_real_)
  }
}
matching <- dplyr::bind_rows(match_rows) |>
  dplyr::arrange(match(target_feature, targets))
need(nrow(matching) == 622L &&
  !anyDuplicated(stats::na.omit(matching$background_feature)) &&
  all(matching$target_context[!is.na(matching$background_feature)] ==
    matching$background_context[!is.na(matching$background_feature)]),
  "Matched pairs have invalid controls or contexts.")
write_table(matching, OUT, "02_all_622_target_neutral_matches.csv")
matched <- dplyr::filter(matching, !is.na(background_feature))
need(nrow(matched) > 0L, "No neutral WT peaks met the matching criteria.")
balance <- dplyr::bind_rows(lapply(c("HFD_open", "HFD_closed"), function(dir) {
  pairs <- dplyr::filter(matched, WT_direction == dir)
  dplyr::bind_rows(lapply(c("target", "background"), function(which) {
    x <- pool[match(pairs[[paste0(which, "_feature")]], pool$feature), ]
    x |>
      dplyr::group_by(context) |>
      dplyr::summarise(direction = dir, membership = which, n = dplyr::n(),
        median_gc = median(gc_percent), median_width = median(width),
        median_CON_detection = median(CON_percent),
        .groups = "drop")
  }))
})) |> dplyr::select(direction, membership, dplyr::everything())
write_table(balance, OUT, "02_matched_background_balance.csv")
match_audit <- matching |>
  dplyr::group_by(WT_direction) |>
  dplyr::summarise(targets = dplyr::n(),
    matched = sum(!is.na(background_feature)),
    fraction_matched = matched / targets,
    max_gc_sd = if (sum(!is.na(background_feature)) > 0)
      max(delta_gc_sd, na.rm = TRUE)
      else NA_real_,
    max_log_width_sd = if (sum(!is.na(background_feature)) > 0)
      max(delta_log_width_sd, na.rm = TRUE) else NA_real_,
    max_CON_detection_sd = if (sum(!is.na(background_feature)) > 0)
      max(delta_CON_detection_sd, na.rm = TRUE) else NA_real_,
    .groups = "drop")
write_table(match_audit, OUT, "02_matching_coverage_audit.csv")
print(match_audit); print(balance)
```

## 4. JASPAR motifs by direction; C0 and count support as checks

One-sided exact paired discordance P values compare the number of
target-positive/control-negative pairs to the reverse. BH adjustment is
performed **separately within each named region set**, across all tested
JASPAR motif profiles. These P values are sequence-enrichment tests only.
The C0-supported tests reuse the same targets and matching; they are
dependent sensitivity checks, not independent validation. The count-supported
16 are retained in the direction summary, but the three closing regions
are too few for a separate motif test.
Report full-universe direction tests as inconclusive if the match rate is
below 80%, and treat the seven C0-supported closing regions as descriptive.

```{r motif-tests}
db <- JASPAR2024::JASPAR2024()
con <- DBI::dbConnect(RSQLite::SQLite(), methods::slot(db, "db"))
pfms <- tryCatch(TFBSTools::getMatrixSet(con,
  opts = list(collection = "CORE", tax_group = "vertebrates",
    all_versions = FALSE)), finally = DBI::dbDisconnect(con))
need(length(pfms) > 0L, "JASPAR returned no CORE vertebrate motifs.")
meta_motif <- tibble(motif_id = vapply(pfms, TFBSTools::ID,
    character(1)), motif_name = vapply(pfms, TFBSTools::name,
    character(1)), motif_family = vapply(pfms, function(x)
    paste(TFBSTools::tags(x)$family, collapse = ";"), character(1)))
scan_features <- unique(c(matched$target_feature,
  matched$background_feature))
hits <- motifmatchr::motifMatches(motifmatchr::matchMotifs(
  pwms = pfms, subject = seqs[scan_features], bg = "subject",
  out = "matches", p.cutoff = MOTIF_P))
need(nrow(hits) == length(scan_features), "Motif scan rows disagree.")
rownames(hits) <- scan_features
motif_id <- vapply(colnames(hits), function(column) {
  ids <- meta_motif$motif_id[startsWith(column, meta_motif$motif_id)]
  if (!length(ids)) return(NA_character_)
  ids[which.max(nchar(ids))]
}, character(1))
need(!anyNA(motif_id), "JASPAR hit column IDs could not be mapped.")
motif_info <- dplyr::left_join(tibble(motif_column = colnames(hits),
  motif_id = motif_id), meta_motif, by = "motif_id",
  relationship = "many-to-one")
pair_sets <- list(
  HFD_open_622 = dplyr::filter(matched, WT_direction == "HFD_open"),
  HFD_closed_622 = dplyr::filter(matched, WT_direction == "HFD_closed"),
  HFD_open_C0 = dplyr::filter(matched, WT_direction == "HFD_open",
    target_feature %in% region$feature[region$C0_supported]),
  HFD_closed_C0 = dplyr::filter(matched, WT_direction == "HFD_closed",
    target_feature %in% region$feature[region$C0_supported]))
motif_results <- dplyr::bind_rows(lapply(names(pair_sets), function(key) {
  pairs <- pair_sets[[key]]
  if (!nrow(pairs)) return(NULL)
  a <- hits[pairs$target_feature, , drop = FALSE]
  b <- hits[pairs$background_feature, , drop = FALSE]
  target_only <- as.integer(Matrix::colSums(a & !b))
  background_only <- as.integer(Matrix::colSums(!a & b))
  discordant <- target_only + background_only
  p <- stats::pbinom(target_only - 1L, discordant, .5,
    lower.tail = FALSE)
  tibble(set = key, n_pairs = nrow(pairs),
    target_hits = as.integer(Matrix::colSums(a)),
    background_hits = as.integer(Matrix::colSums(b)),
    target_only = target_only, background_only = background_only,
    target_percent = 100 * target_hits / nrow(pairs),
    background_percent = 100 * background_hits / nrow(pairs),
    prevalence_difference_pp = target_percent - background_percent,
    paired_one_sided_p = p,
    paired_BH = stats::p.adjust(p, method = "BH"),
    motif_column = colnames(hits)) |>
    dplyr::left_join(motif_info, by = "motif_column",
      relationship = "many-to-one")
})) |>
  dplyr::arrange(set, paired_BH, paired_one_sided_p,
    dplyr::desc(prevalence_difference_pp))
motif_results$direction <- ifelse(startsWith(motif_results$set, "HFD_open"),
  "HFD_open", "HFD_closed")
motif_results$direction_match_fraction <- match_audit$fraction_matched[
  match(motif_results$direction, match_audit$WT_direction)]
motif_results$interpretation_status <- dplyr::case_when(
  motif_results$direction_match_fraction < .8 ~ "insufficient_matching",
  motif_results$n_pairs < 10L ~ "very_small_descriptive_subset",
  TRUE ~ "conditional_sequence_enrichment_only")
write_table(motif_results, OUT,
  "03_JASPAR2024_open_close_paired_motif_tests.csv.gz")
motif_summary <- motif_results |>
  dplyr::group_by(set) |>
  dplyr::summarise(matched_pairs = dplyr::first(n_pairs),
    n_profiles = dplyr::n(), n_BH_below_005 = sum(paired_BH < .05),
    min_BH = min(paired_BH), .groups = "drop")
write_table(motif_summary, OUT, "03_motif_result_summary.csv")
print(motif_summary)
for (key in c("HFD_open_622", "HFD_closed_622")) {
  x <- motif_results |>
    dplyr::filter(set == key, target_hits >= 5L,
      prevalence_difference_pp > 0) |>
    dplyr::slice_head(n = 20L)
  if (!nrow(x)) next
  x$label <- paste(x$motif_name, x$motif_id)
  p <- ggplot2::ggplot(x, ggplot2::aes(
    x = stats::reorder(label, prevalence_difference_pp),
    y = prevalence_difference_pp, size = target_hits,
    color = paired_BH)) +
    ggplot2::geom_point() + ggplot2::coord_flip() +
    ggplot2::theme_bw(base_size = 10) +
    ggplot2::labs(title = paste(key, "motif sequence enrichment"),
      subtitle = "Candidate vs matched neutral WT accessible regions; paired BH tests DNA sequence only",
      x = NULL, y = "Candidate minus control motif prevalence (pp)")
  save_vector_plot(p, PLOTS, paste0("03_", key, "_motif_context.pdf"),
    12, 8)
}
```

## 5. Review handoff

Read the matching audit before any motif table. If either 622-direction
match fraction is below 0.8, widen the candidate background universe or
revisit matching **before** interpreting motif ranks; do not quietly
relax the matching limits. A significant motif requires follow-up with
TF expression and ideally footprinting or an independent assay. Adjacent
peaks can share a regulatory locus; inspect regional redundancy before
counting them as independent examples. Review the saved Xist sensitivity
and individual genotype parents for each prioritized candidate. This step
does not create replicate-level statistics.

```{r finish}
finish_run(OUT, "STEP_08L_COMPLETE.txt", c(
  "Fixed 622 WT top-15 ISC candidates were split by HFD direction.",
  "Neutral evaluable accessible peaks were matched on context, GC, width and WT CON detection.",
  "Paired motif BH tests are sequence-enrichment tests, not biological diet tests.",
  "No replicate-level DAR or genotype-by-diet inference from pooled libraries.",
  paste("Outputs:", OUT)))
```
