---
title: "Step 08O | Motifs in WT ISC interval loss sets"
output:
  html_notebook:
    toc: true
    toc_float: true
---

Run from the repository root after 08N. Each of six predefined sets uses
the corresponding 08L GC/width/WT CON-detection/context matched neutral
accessible intervals. The main test is a one-sided paired discordance test
for motif occurrence, with BH across all JASPAR profiles separately in each
set. An additional exploratory Fisher comparison asks whether a motif is more
common among lost than retained WT candidates in the same direction and
evaluable knockout group. This second comparison lacks matching and can be
underpowered. For shared loss its comparator comprises intervals evaluable in
both genotypes that are **not** shared-loss; it can include one-genotype loss.
The six sets overlap by design, so tests across sets are not independent.
Motif occurrence neither establishes TF binding nor a genotype-by-diet effect.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Run from the repository root.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
needed <- c("readr", "dplyr", "tibble", "GenomicRanges", "IRanges",
  "Rsamtools", "Biostrings", "TFBSTools", "motifmatchr", "Matrix")
missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Missing R packages: ", paste(missing, collapse = ", "))
BASE08L <- file.path(OUTPUT_DIR,
  "08l_wt_ISC_open_close_motifs")
BASE08N <- file.path(OUTPUT_DIR,
  "08n_wt_ISC_interval_loss_pathways")
OUT <- file.path(OUTPUT_DIR,
  "08o_wt_ISC_loss_group_motifs")
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
read_csv <- function(path) readr::read_csv(path, show_col_types = FALSE,
  progress = FALSE)
write_csv <- function(x, path) readr::write_csv(x, path)
assert <- function(ok, why) if (!isTRUE(ok)) stop(why, call. = FALSE)
FASTA <- Sys.getenv("MULTIOME_GRCM39_FASTA", unset = file.path(
  PROJECT_DIR, "reference", "GRCm39_2024-A", "fasta", "genome.fa"))
PFM <- file.path(PROJECT_DIR, "reference", "JASPAR2024",
  "JASPAR2024_CORE_vertebrates_non-redundant_pfms_jaspar.txt")
files <- c(region = file.path(BASE08N,
    "01_all_622_annotated_loss_audit.csv.gz"),
  pairs = file.path(BASE08L, "02_all_622_target_neutral_matches.csv"),
  universe = file.path(BASE08L,
    "02_evaluable_WT_universe_sequence_covariates.csv.gz"),
  complete08n = file.path(BASE08N, "STEP_08N_COMPLETE.txt"),
  fasta = FASTA, fai = paste0(FASTA, ".fai"), pfm = PFM)
assert(all(file.exists(files)), paste("Missing:",
  paste(files[!file.exists(files)], collapse = "; ")))
assert(file.info(PFM)$size > 100000L,
  "The local JASPAR2024 PFM file is unexpectedly small.")
write_csv(tibble::tibble(source = names(files),
  path = normalizePath(files),
  bytes = file.info(files)$size,
  md5 = unname(tools::md5sum(files))),
  file.path(OUT, "00_input_manifest.csv"))
```

## Verify frozen inputs, match coverage, and GRCm39 sequences

Unmatched 08L targets remain in the denominator of match coverage; they do
not enter paired motif tests. We do not rematch after seeing the loss labels.
Within each set, less than 80% matched is flagged as insufficient matching.
Sets below ten matched pairs are descriptive.

```{r sequences}
region <- read_csv(files[["region"]])
pairs <- read_csv(files[["pairs"]])
universe <- read_csv(files[["universe"]])
assert(nrow(region) == 622L && nrow(pairs) == 622L &&
  !anyDuplicated(region$feature) &&
  !anyDuplicated(pairs$target_feature) &&
  setequal(pairs$target_feature, region$feature),
  "08N/08L frozen target IDs disagree.")
assert(!anyDuplicated(universe$feature) &&
  all(region$feature %in% universe$feature),
  "08L sequence covariates do not cover the candidates.")
ix <- match(pairs$target_feature, region$feature)
assert(all(pairs$WT_direction == region$WT_direction[ix]) &&
  all(is.na(pairs$background_feature) |
    pairs$background_feature %in% universe$feature),
  "Direction or matched background does not agree with 08N/08L.")
matched <- pairs[!is.na(pairs$background_feature), , drop = FALSE]
assert(!anyDuplicated(matched$background_feature) &&
  all(matched$target_context == matched$background_context),
  "08L matched controls were duplicated or mismatched.")
scan_features <- unique(c(region$feature, matched$background_feature))
seq_table <- universe[match(scan_features, universe$feature), , drop = FALSE]
assert(all(!is.na(seq_table$chrom)) &&
  all(seq_table$start_1based >= 1L) &&
  all(seq_table$end_1based >= seq_table$start_1based),
  "Missing or invalid GRCm39 coordinates.")
gr <- GenomicRanges::GRanges(seqnames = seq_table$chrom,
  ranges = IRanges::IRanges(start = seq_table$start_1based,
    end = seq_table$end_1based))
names(gr) <- scan_features
fa <- Rsamtools::FaFile(FASTA)
index <- Rsamtools::scanFaIndex(fa)
width_by_chr <- setNames(as.integer(BiocGenerics::width(index)),
  as.character(GenomicRanges::seqnames(index)))
assert(all(seq_table$chrom %in% names(width_by_chr)) &&
  all(seq_table$end_1based <= width_by_chr[seq_table$chrom]),
  "A motif interval lies outside the GRCm39 FASTA.")
seqs <- Rsamtools::scanFa(fa, param = gr)
assert(length(seqs) == length(scan_features) &&
  all(as.integer(BiocGenerics::width(seqs)) ==
    seq_table$end_1based - seq_table$start_1based + 1L),
  "Retrieved FASTA widths differ from frozen interval coordinates.")
names(seqs) <- scan_features
```

## Scan the offline JASPAR 2024 profiles

The PFM file avoids the JASPAR2024 SQLite download/BiocFileCache problem
observed on SOL. It is the same file and matching method as 08L.

```{r motifs}
raw <- readLines(PFM, warn = FALSE)
raw <- raw[nzchar(trimws(raw))]
assert(length(raw) %% 5L == 0L,
  "Expected five nonblank lines per local JASPAR motif.")
headers <- seq.int(1L, length(raw), by = 5L)
assert(all(grepl("^>MA[0-9]+\\.[0-9]+[[:space:]]+", raw[headers])),
  "Unexpected JASPAR PFM motif ID format.")
raw[headers] <- sub("^>(MA[0-9]+\\.[0-9]+)[[:space:]]+(.+)$",
  ">\\1\t\\2", raw[headers])
normalized <- tempfile("jaspar2024_", fileext = ".txt")
writeLines(raw, normalized)
pfms <- tryCatch(TFBSTools::readJASPARMatrix(normalized,
  matrixClass = "PFM"), finally = unlink(normalized))
ids <- vapply(pfms, TFBSTools::ID, character(1))
motif_names <- vapply(pfms, TFBSTools::name, character(1))
assert(length(ids) == length(headers) && !anyDuplicated(ids),
  "Imported PFM count/IDs disagree with the local file.")
hits <- motifmatchr::motifMatches(motifmatchr::matchMotifs(
  pwms = pfms, subject = seqs, bg = "subject", out = "matches",
  p.cutoff = 5e-5))
expected <- names(pfms)
observed <- colnames(hits)
assert(nrow(hits) == length(scan_features) && ncol(hits) == length(ids) &&
  (identical(observed, expected) ||
   identical(observed, make.unique(expected)) ||
   identical(observed, paste0(ids, "_", motif_names))),
  paste("Unexpected motif hit labels/order:",
    paste(utils::head(observed), collapse = ", ")))
rownames(hits) <- scan_features
colnames(hits) <- ids
motif_info <- tibble::tibble(motif_id = ids, motif_name = motif_names)
```

## Six matched tests and six exploratory within-candidate tests

`loss_vs_nonloss` is not a substitute for matched neutral testing: its
comparison group is conditioned on genotype effects and not sequence matched.
The main matched test assesses the selected regions' motif composition
relative to accessible WT neutral peaks, which can also reflect a generic
WT HFD program. Neither test shows TF binding.

```{r tests}
keys <- as.vector(outer(c("PPAR", "IL17", "shared"),
  c("HFD_open", "HFD_closed"),
  FUN = function(g, d) paste(g, d, "lost", sep = "_")))
assert(length(keys) == 6L && !anyDuplicated(keys),
  "Six motif comparisons were not enumerated.")
set_files <- file.path(BASE08N, "sets", paste0(keys, ".csv.gz"))
assert(all(file.exists(set_files)), paste("Missing 08N set files:",
  paste(set_files[!file.exists(set_files)], collapse = "; ")))
audit <- list()
for (j in seq_along(keys)) {
  key <- keys[[j]]
  genotype <- sub("_HFD_.*$", "", key)
  direction <- if (grepl("_HFD_open_", key)) "HFD_open" else "HFD_closed"
  eligible <- if (genotype == "PPAR") region$ppar_evaluable else
    if (genotype == "IL17") region$il17_evaluable else
      region$both_evaluable
  foreground <- if (genotype == "PPAR") region$ppar_loss else
    if (genotype == "IL17") region$il17_loss else
      region$comparison == "shared_loss"
  candidates <- region$feature[region$WT_direction == direction & eligible]
  target <- region$feature[region$WT_direction == direction & foreground]
  recorded <- read_csv(set_files[[j]])
  assert(!anyDuplicated(recorded$feature) &&
    setequal(recorded$feature, target),
    paste("08N set file and loss audit disagree:", key))
  control <- setdiff(candidates, target)
  assert(all(target %in% candidates),
    paste("Loss contains unevaluable intervals:", key))
  pairs_here <- matched[matched$target_feature %in% target, , drop = FALSE]
  matched_n <- nrow(pairs_here)
  match_fraction <- if (length(target)) matched_n / length(target) else NA_real_
  status <- if (!length(target)) "empty_set" else
    if (match_fraction < .8) "insufficient_matching" else
      if (matched_n < 10L) "small_descriptive_subset" else
        "conditional_sequence_test"
  audit[[key]] <- tibble::tibble(set = key, direction = direction,
    n_evaluable_candidates = length(candidates),
    n_lost = length(target), n_nonloss_candidates = length(control),
    n_matched_pairs = matched_n, match_fraction = match_fraction,
    status = status)

  if (matched_n) {
    a <- hits[pairs_here$target_feature, , drop = FALSE]
    b <- hits[pairs_here$background_feature, , drop = FALSE]
    a_only <- as.integer(Matrix::colSums(a & !b))
    b_only <- as.integer(Matrix::colSums(!a & b))
    p <- stats::pbinom(a_only - 1L, a_only + b_only, .5,
      lower.tail = FALSE)
    target_hits <- as.integer(Matrix::colSums(a))
    background_hits <- as.integer(Matrix::colSums(b))
    matched_result <- motif_info |>
      dplyr::mutate(set = key, n_pairs = matched_n,
        target_hits = target_hits, background_hits = background_hits,
        target_only = a_only, background_only = b_only,
        target_percent = 100 * target_hits / matched_n,
        background_percent = 100 * background_hits / matched_n,
        prevalence_difference_pp = target_percent - background_percent,
        paired_one_sided_p = p, paired_BH = stats::p.adjust(p, "BH"),
        interpretation_status = status) |>
      dplyr::arrange(.data$paired_BH, .data$paired_one_sided_p,
        dplyr::desc(.data$prevalence_difference_pp))
  } else matched_result <- tibble::tibble(motif_id = character(),
    motif_name = character(), set = character(), n_pairs = integer(),
    target_hits = integer(), background_hits = integer(),
    target_only = integer(), background_only = integer(),
    target_percent = numeric(), background_percent = numeric(),
    prevalence_difference_pp = numeric(), paired_one_sided_p = numeric(),
    paired_BH = numeric(), interpretation_status = character())
  write_csv(matched_result, file.path(OUT,
    paste0("01_", key, "_matched_neutral_JASPAR.csv.gz")))

  if (length(target) >= 3L && length(control) >= 3L) {
    target_hits <- as.integer(Matrix::colSums(hits[target, , drop = FALSE]))
    control_hits <- as.integer(Matrix::colSums(hits[control, , drop = FALSE]))
    p <- vapply(seq_along(ids), function(i)
      stats::fisher.test(matrix(c(target_hits[i], length(target) - target_hits[i],
        control_hits[i], length(control) - control_hits[i]), nrow = 2L,
        byrow = TRUE), alternative = "greater")$p.value, numeric(1))
    within_result <- motif_info |>
      dplyr::mutate(set = key, n_lost = length(target),
        n_nonloss = length(control), lost_hits = target_hits,
        nonloss_hits = control_hits,
        lost_percent = 100 * target_hits / length(target),
        nonloss_percent = 100 * control_hits / length(control),
        prevalence_difference_pp = lost_percent - nonloss_percent,
        fisher_one_sided_p = p, fisher_BH = stats::p.adjust(p, "BH")) |>
      dplyr::arrange(.data$fisher_BH, .data$fisher_one_sided_p)
  } else within_result <- tibble::tibble(motif_id = character(),
    motif_name = character(), set = character(), n_lost = integer(),
    n_nonloss = integer(), lost_hits = integer(), nonloss_hits = integer(),
    lost_percent = numeric(), nonloss_percent = numeric(),
    prevalence_difference_pp = numeric(), fisher_one_sided_p = numeric(),
    fisher_BH = numeric())
  write_csv(within_result, file.path(OUT,
    paste0("02_", key, "_loss_vs_nonloss_exploratory.csv.gz")))
}
audit <- dplyr::bind_rows(audit)
write_csv(audit, file.path(OUT, "00_six_set_motif_audit.csv"))
print(audit)
writeLines(c("08O finished; six matched-neutral and six loss-vs-nonloss motif tables.",
  "BH correction is within each motif table across all JASPAR profiles.",
  "Motifs test sequence occurrence, not diet interaction or TF occupancy."),
  file.path(OUT, "STEP_08O_COMPLETE.txt"))
```
