---
title: "Step 08P | WT ISC loss comparison figures"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This standalone notebook recreates the three 08N/08O review figures from
their completed output directories. Run from the repository root, after
Steps 08N and 08O. It does not require a Seurat object or fragments. It
saves PDF, SVG, PNG, and the two small figure audit tables under
`08p_wt_ISC_loss_review_figures/` in the WT output directory.
If needed, override the input/output directories using `STEP08N_DIR`,
`STEP08O_DIR`, and `STEP08P_FIGURE_DIR` environment variables.

The 622 WT candidate intervals were selected on pooled WT accessibility
effects. The knockout gates contain different nuclei; both genotype
comparisons are descriptive. In particular, an interval called "lost" is
still present. It has an effect at most half as large as the WT effect, or
an effect of opposite sign. The overlap comparison is a diagnostic among
selected intervals, not a biological replicate or interaction test.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Run from the repository root.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(tibble)
  library(readr); library(ggplot2)
})
need <- function(ok, why) if (!isTRUE(ok)) stop(why, call. = FALSE)
STEP08N <- Sys.getenv("STEP08N_DIR", unset = file.path(OUTPUT_DIR,
  "08n_wt_ISC_interval_loss_pathways"))
STEP08O <- Sys.getenv("STEP08O_DIR", unset = file.path(OUTPUT_DIR,
  "08o_wt_ISC_loss_group_motifs"))
OUT <- Sys.getenv("STEP08P_FIGURE_DIR", unset = file.path(OUTPUT_DIR,
  "08p_wt_ISC_loss_review_figures"))
dir.create(OUT, recursive = TRUE, showWarnings = FALSE)
input <- c(region = file.path(STEP08N,
    "01_all_622_annotated_loss_audit.csv.gz"),
  sets = file.path(STEP08N, "03_set_manifest.csv"),
  motif_audit = file.path(STEP08O, "00_six_set_motif_audit.csv"),
  complete08n = file.path(STEP08N, "STEP_08N_COMPLETE.txt"),
  complete08o = file.path(STEP08O, "STEP_08O_COMPLETE.txt"))
need(all(file.exists(input)), paste("Missing inputs:",
  paste(input[!file.exists(input)], collapse = "; ")))
region <- readr::read_csv(input[["region"]], show_col_types = FALSE,
  progress = FALSE)
set_manifest <- readr::read_csv(input[["sets"]], show_col_types = FALSE)
motif_audit <- readr::read_csv(input[["motif_audit"]], show_col_types = FALSE)
required <- c("feature", "WT_direction", "anchor_log2FC_HFD_vs_CON",
  "C0_all_log2FC_HFD_vs_CON", "C0_supported", "count_supported_40",
  "ppar_evaluable", "il17_evaluable", "both_evaluable", "comparison",
  "ppar_effect", "il17_effect", "nearest_symbol", "anchor_CON_counts",
  "anchor_HFD_counts", "ppar_parent_effect", "il17_parent_effect")
need(all(required %in% names(region)),
  paste("Missing 08N audit columns:",
    paste(setdiff(required, names(region)), collapse = ", ")))
need(nrow(region) == 622L && !anyDuplicated(region$feature) &&
  all(!is.na(region$both_evaluable)) && nrow(motif_audit) == 6L &&
  all(c("shared_HFD_open_lost", "shared_HFD_closed_lost") %in%
    set_manifest$set), "Step 08N/08O review inputs disagree.")
readr::write_csv(tibble::tibble(source = names(input),
  path = normalizePath(input), bytes = file.info(input)$size,
  md5 = unname(tools::md5sum(input))),
  file.path(OUT, "00_figure_input_manifest.csv"))
save_figure <- function(plot, stem, width, height) {
  ggplot2::ggsave(file.path(OUT, paste0(stem, ".pdf")), plot = plot,
    device = grDevices::pdf, useDingbats = FALSE,
    width = width, height = height)
  ggplot2::ggsave(file.path(OUT, paste0(stem, ".png")), plot = plot,
    device = "png", width = width, height = height, dpi = 220,
    bg = "white")
  # SVG remains editable in Illustrator. The PDF and PNG are saved first.
  tryCatch(ggplot2::ggsave(file.path(OUT, paste0(stem, ".svg")),
    plot = plot, device = grDevices::svg, width = width, height = height),
    error = function(e) warning("SVG unavailable for ", stem, ": ",
      conditionMessage(e)))
}
```

## Figure 1: four-way overlap among doubly evaluable intervals

The PPAR-only and IL17-only counts have the **same denominator** as the
shared and retained counts. Marginal expectation conditions on the separate
PPAR and IL17 loss totals in that WT-direction subset. A large shared count
need not indicate a special common program when each knockout has a high
loss rate.

```{r overlap}
dirs <- c("HFD_open", "HFD_closed")
status <- c("shared_loss", "PPAR_only_loss", "IL17_only_loss",
  "retained_in_both")
status_labels <- c(shared_loss = "Lost in both",
  PPAR_only_loss = "PPAR only", IL17_only_loss = "IL17 only",
  retained_in_both = "Retained in both")
status_colors <- c(shared_loss = "#6B31A7", PPAR_only_loss = "#C26345",
  IL17_only_loss = "#277A9B", retained_in_both = "#758175")
joint <- region |>
  dplyr::filter(.data$both_evaluable)
need(all(joint$comparison %in% status),
  "A doubly evaluable interval lacks a four-way classification.")
overlap <- dplyr::bind_rows(lapply(dirs, function(direction) {
  x <- joint[joint$WT_direction == direction, , drop = FALSE]
  counts <- table(factor(x$comparison, levels = status))
  a <- unname(counts[["shared_loss"]])
  b <- unname(counts[["PPAR_only_loss"]])
  c <- unname(counts[["IL17_only_loss"]])
  d <- unname(counts[["retained_in_both"]])
  need(a + b + c + d == nrow(x), "Four-way counts are incomplete.")
  ft <- stats::fisher.test(matrix(c(a, b, c, d), nrow = 2L,
    byrow = TRUE), alternative = "greater")
  tibble::tibble(WT_direction = direction, n_doubly_evaluable = nrow(x),
    observed_shared = a, expected_shared_under_independent_loss_calls =
      (a+b)*(a+c)/nrow(x), PPAR_only = b, IL17_only = c,
    retained_both = d, conditional_overlap_odds_ratio =
      (a*d)/(b*c), exploratory_one_sided_fisher_p = ft$p.value)
}))
readr::write_csv(overlap, file.path(OUT, "01_overlap_diagnostic.csv"))
print(overlap)
segments <- joint |>
  dplyr::count(.data$WT_direction, .data$comparison, name = "n") |>
  tidyr::complete(WT_direction = dirs, comparison = status,
    fill = list(n = 0L)) |>
  dplyr::mutate(status_index = match(.data$comparison, status),
    y = ifelse(.data$WT_direction == "HFD_open", 2, 1)) |>
  dplyr::group_by(.data$WT_direction) |>
  dplyr::arrange(.data$status_index, .by_group = TRUE) |>
  dplyr::mutate(x0 = dplyr::lag(cumsum(.data$n), default = 0L),
    x1 = cumsum(.data$n), mid = (.data$x0 + .data$x1) / 2) |>
  dplyr::ungroup()
right <- overlap |>
  dplyr::mutate(y = ifelse(.data$WT_direction == "HFD_open", 2, 1),
    note = sprintf("Observed %d; expected %.1f from loss rates\nOverlap OR %.2f",
      .data$observed_shared,
      .data$expected_shared_under_independent_loss_calls,
      .data$conditional_overlap_odds_ratio))
right_x <- max(overlap$n_doubly_evaluable) + 50
p_overlap <- ggplot2::ggplot(segments) +
  ggplot2::geom_rect(ggplot2::aes(xmin = .data$x0, xmax = .data$x1,
    ymin = .data$y - .29, ymax = .data$y + .29,
    fill = .data$comparison)) +
  ggplot2::geom_text(data = dplyr::filter(segments, .data$n >= 9L),
    ggplot2::aes(x = .data$mid, y = .data$y, label = .data$n),
    color = "white", fontface = "bold", size = 4) +
  ggplot2::geom_text(data = right,
    ggplot2::aes(x = .data$n_doubly_evaluable + 14, y = .data$y,
      label = paste0("n=", .data$n_doubly_evaluable)),
    hjust = 0, inherit.aes = FALSE, size = 3.7) +
  ggplot2::geom_text(data = right,
    ggplot2::aes(x = right_x, y = .data$y, label = .data$note),
    hjust = 0, lineheight = 1.4, inherit.aes = FALSE, size = 3.4) +
  ggplot2::scale_fill_manual(values = status_colors,
    breaks = status, labels = status_labels, name = NULL) +
  ggplot2::scale_y_continuous(breaks = c(2, 1),
    labels = c("WT HFD-opening", "WT HFD-closing"),
    limits = c(.55, 2.45), expand = c(0, 0)) +
  ggplot2::scale_x_continuous(limits = c(0, right_x + 205),
    breaks = seq(0, 350, by = 50), expand = c(0, 0)) +
  ggplot2::labs(title = "WT ISC-enriched HFD regions in PPAR and IL17",
    subtitle = "Loss categories among intervals evaluable in both knockouts",
    x = "Exact WT candidate intervals", y = NULL,
    caption = "Shared-loss expectation uses the two marginal loss rates. Overlap OR is a selected-interval diagnostic, not a genotype-by-diet test.") +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(legend.position = "bottom",
    plot.caption = ggplot2::element_text(hjust = 0, size = 8),
    panel.grid.major.y = ggplot2::element_blank()) +
  ggplot2::guides(fill = ggplot2::guide_legend(nrow = 2, byrow = TRUE))
save_figure(p_overlap, "01_directional_loss_and_overlap", 13, 5)
```

## Figure 2: effects at the same consensus intervals

Grey points are all evaluable WT candidates. Purple points were independently
designated C0-supported before the genotype comparison. The diagonal shows
equal effects; the dotted line shows half the WT effect on the same sign.
WT candidate selection at `|log2FC| >= 0.5` means regression to the mean is
an alternative explanation for some of the apparent drop-off.

```{r effect-scatter}
effect_long <- dplyr::bind_rows(lapply(c("ppar", "il17"), function(g) {
  keep <- region[[paste0(g, "_evaluable")]]
  tibble::tibble(feature = region$feature[keep],
    WT_direction = region$WT_direction[keep],
    genotype = toupper(g),
    WT_effect = region$anchor_log2FC_HFD_vs_CON[keep],
    KO_effect = region[[paste0(g, "_effect")]][keep],
    C0_supported = region$C0_supported[keep])
})) |>
  dplyr::mutate(direction = factor(.data$WT_direction,
    levels = dirs, labels = c("WT HFD opening", "WT HFD closing")),
    genotype = factor(.data$genotype, levels = c("PPAR", "IL17")))
need(all(is.finite(effect_long$WT_effect)) &&
  all(is.finite(effect_long$KO_effect)) &&
  all(!is.na(effect_long$C0_supported)),
  "Nonfinite effect or missing C0 status in an evaluable interval.")
limits <- range(c(effect_long$WT_effect, effect_long$KO_effect, 0))
limits <- c(floor(limits[[1]] * 5)/5, ceiling(limits[[2]] * 5)/5)
p_effect <- ggplot2::ggplot(effect_long,
  ggplot2::aes(x = .data$WT_effect, y = .data$KO_effect)) +
  ggplot2::geom_hline(yintercept = 0, color = "grey55", linewidth = .35) +
  ggplot2::geom_abline(intercept = 0, slope = 1, linetype = "dashed",
    color = "grey55", linewidth = .45) +
  ggplot2::geom_abline(intercept = 0, slope = .5, linetype = "dotted",
    color = "grey72", linewidth = .45) +
  ggplot2::geom_point(data = dplyr::filter(effect_long,
    !.data$C0_supported), color = "#B7BBC1", size = 1,
    alpha = .6) +
  ggplot2::geom_point(data = dplyr::filter(effect_long,
    .data$C0_supported), color = "#6933A6", shape = 21,
    fill = "#6933A6", stroke = .2, size = 2.3) +
  ggplot2::facet_grid(direction ~ genotype) +
  ggplot2::coord_equal(xlim = limits, ylim = limits) +
  ggplot2::labs(title = "Effect estimates at the same WT-selected coordinates",
    subtitle = "Purple = previously identified WT C0-supported intervals; dashed = WT magnitude; dotted = half WT magnitude",
    x = "WT top-15% pooled ATAC log2(HFD/CON)",
    y = "Knockout top-15% pooled ATAC log2(HFD/CON)",
    caption = "Each dot is an exact consensus interval. WT was screened on |log2FC| >= 0.5; knockout gates contain distinct nuclei. No mouse-level P values.") +
  ggplot2::theme_classic(base_size = 10) +
  ggplot2::theme(plot.caption = ggplot2::element_text(hjust = 0,
    size = 8))
save_figure(p_effect, "02_same_interval_effects_by_genotype", 11.5, 9)
```

## Figure 3: four C0-supported, higher-count examples

These exact regions were selected from the C0-supported and higher-count
set. All four have the shared-loss label and are evaluable in both knockouts.
The adjacent gene is a locus label **only**. This figure does not demonstrate
RNA change, gene regulation, TF binding, or replicate-level significance.
Fragment coverage with a common scale remains the necessary next view.

```{r loci}
examples <- c("chr4-40142794-40143817",
  "chr13-69759275-69760309", "chr18-65563454-65564359",
  "chr5-97144996-97145914")
need(all(examples %in% region$feature),
  "A preselected illustrative interval is missing.")
chosen <- region[match(examples, region$feature), , drop = FALSE]
need(all(chosen$C0_supported & chosen$count_supported_40 &
  chosen$comparison == "shared_loss" & chosen$both_evaluable),
  "An example no longer meets its prespecified evidence criteria.")
readr::write_csv(chosen |>
  dplyr::select(feature, nearest_symbol, WT_direction,
    anchor_CON_counts, anchor_HFD_counts,
    anchor_log2FC_HFD_vs_CON, C0_all_log2FC_HFD_vs_CON,
    ppar_effect, il17_effect, ppar_parent_effect, il17_parent_effect),
  file.path(OUT, "03_four_loci_audit.csv"))
source_levels <- c("WT selected", "WT C0 all", "PPAR top 15%",
  "IL17 top 15%")
source_colors <- c(`WT selected` = "#E59F26", `WT C0 all` = "#B57817",
  `PPAR top 15%` = "#C26345", `IL17 top 15%` = "#277A9B")
loci <- chosen |>
  dplyr::mutate(locus = paste0(.data$nearest_symbol, "\n",
    ifelse(.data$WT_direction == "HFD_open", "HFD open", "HFD closed"),
    "\n", .data$feature)) |>
  dplyr::select(feature, locus,
    `WT selected` = anchor_log2FC_HFD_vs_CON,
    `WT C0 all` = C0_all_log2FC_HFD_vs_CON,
    `PPAR top 15%` = ppar_effect,
    `IL17 top 15%` = il17_effect) |>
  tidyr::pivot_longer(cols = dplyr::all_of(source_levels),
    names_to = "source", values_to = "effect") |>
  dplyr::mutate(locus = factor(.data$locus,
    levels = unique(.data$locus)),
    source = factor(.data$source, levels = source_levels),
    label = sprintf("%+.2f", ifelse(abs(.data$effect) < .005,
      0, .data$effect)),
    label_vjust = ifelse(.data$effect >= 0, -1, 1.7))
need(all(is.finite(loci$effect)), "Example effects are not finite.")
p_loci <- ggplot2::ggplot(loci,
  ggplot2::aes(x = .data$source, y = .data$effect,
    color = .data$source)) +
  ggplot2::geom_hline(yintercept = 0, color = "grey55",
    linewidth = .35) +
  ggplot2::geom_point(size = 3.1) +
  ggplot2::geom_text(ggplot2::aes(label = .data$label,
    vjust = .data$label_vjust), size = 3, show.legend = FALSE) +
  ggplot2::facet_wrap(~ locus, nrow = 1) +
  ggplot2::scale_color_manual(values = source_colors, guide = "none") +
  ggplot2::scale_y_continuous(limits = c(-1.1, 2.1),
    breaks = seq(-1, 2, by = .5)) +
  ggplot2::labs(title = "Illustrative C0-supported, higher-count intervals",
    x = NULL, y = "Pooled ATAC log2(HFD/CON)",
    caption = "Nearest gene names indicate genomic proximity, not regulatory targets. WT C0 all is a within-cohort sensitivity estimate; inspect matched-scale coverage and RNA separately.") +
  ggplot2::theme_classic(base_size = 10) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 40,
    hjust = 1), plot.caption = ggplot2::element_text(hjust = 0,
    size = 8))
save_figure(p_loci, "03_four_loci_pooled_effects", 14, 5)
writeLines(c("08P figures complete.",
  "Opening and closing status bars use only doubly evaluable exact intervals.",
  "No genotype-by-diet P values or peak-to-gene claims are established."),
  file.path(OUT, "STEP_08P_COMPLETE.txt"))
```
