---
title: "Step 08Q | Functions represented by the WT ISC HFD peak program"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This is a separate biological question from Step 08N. It asks whether genes
nearest to the frozen WT **HFD-opening** or **HFD-closing** peaks are more
frequently assigned GO or KEGG terms than genes near the **2,028 evaluable
WT C0/W0 top-15% ATAC peaks**. The background includes both WT candidate
directions and the other evaluable WT peaks. It does not test PPAR or IL17
dependence. Promoter analyses restrict **both** foreground and universe to
peaks within ±2 kb of the nearest annotated GRCm39 vM33 TSS.

Nearest genes are proximity assignments, not established targets. A gene
with multiple evaluable peaks has more opportunities to enter a peak-selected
foreground; simple gene-level hypergeometric tests do not correct that bias.
These terms, if any, are exploratory functions associated with the WT peak
screen, not replicated differential accessibility or TF binding.

Run from the repository root after Steps 08L and 08N. Step 08L already
performed the complementary WT-direction motif test against matched neutral
evaluable WT peaks; this notebook copies its motif summary for review and
does not rescan motifs.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Run from the repository root.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced",
  MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
required_packages <- c("readr", "dplyr", "tibble", "GenomicRanges",
  "IRanges", "ChIPseeker", "GenomicFeatures", "AnnotationDbi",
  "org.Mm.eg.db", "clusterProfiler")
missing <- required_packages[!vapply(required_packages,
  requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Missing packages: ", paste(missing,
  collapse = ", "))
need <- function(ok, why) if (!isTRUE(ok)) stop(why, call. = FALSE)
STEP08L <- file.path(OUTPUT_DIR,
  "08l_wt_ISC_open_close_motifs")
STEP08N <- file.path(OUTPUT_DIR,
  "08n_wt_ISC_interval_loss_pathways")
OUT <- file.path(OUTPUT_DIR,
  "08q_wt_ISC_program_pathways")
ORA <- file.path(OUT, "ora")
dir.create(ORA, recursive = TRUE, showWarnings = FALSE)
read_csv <- function(f) readr::read_csv(f, show_col_types = FALSE,
  progress = FALSE)
write_csv <- function(x, f) readr::write_csv(x, f)
files <- c(universe = file.path(STEP08L,
    "02_evaluable_WT_universe_sequence_covariates.csv.gz"),
  frozen = file.path(STEP08N,
    "01_all_622_annotated_loss_audit.csv.gz"),
  motif_summary = file.path(STEP08L, "03_motif_result_summary.csv"),
  gtf = GRCM39_GTF)
need(all(file.exists(files)), paste("Missing:",
  paste(files[!file.exists(files)], collapse = "; ")))
write_csv(tibble::tibble(input = names(files),
  path = normalizePath(files), md5 = unname(tools::md5sum(files))),
  file.path(OUT, "00_input_manifest.csv"))
```

## Annotate the evaluable WT opportunity set

The candidate 622 retain their Step 08N annotation exactly. We annotate
only the other 1,406 intervals with the same ChIPseeker/TxDb/TSS window, then
map Ensembl gene IDs to Entrez IDs with the same deterministic rule as 08N.

```{r annotate-universe}
u <- read_csv(files[["universe"]])
frozen <- read_csv(files[["frozen"]])
need(nrow(u) == 2028L && !anyDuplicated(u$feature) &&
  nrow(frozen) == 622L && !anyDuplicated(frozen$feature) &&
  all(frozen$feature %in% u$feature),
  "The 08L WT opportunity set or 08N candidate identity changed.")
need(all(c("chrom", "start_1based", "end_1based", "membership") %in%
  names(u)) && all(c("WT_direction", "nearest_ensembl",
    "nearest_entrez", "nearest_symbol", "annotation", "distanceToTSS",
    "promoter_pm2kb") %in% names(frozen)),
  "Annotation or coordinate columns are missing.")
ix <- match(frozen$feature, u$feature)
need(all(frozen$chrom == u$chrom[ix] &
  frozen$start_1based == u$start_1based[ix] &
  frozen$end_1based == u$end_1based[ix]) &&
  all(u$membership[ix] == frozen$WT_direction),
  "08L/08N exact coordinates or directions differ.")
cols <- c("feature", "chrom", "start_1based", "end_1based",
  "annotation", "distanceToTSS", "nearest_ensembl",
  "nearest_entrez", "nearest_symbol", "promoter_pm2kb")
fixed <- frozen[, cols]
fixed$WT_direction <- frozen$WT_direction
other <- u[!u$feature %in% fixed$feature, , drop = FALSE]
need(nrow(other) == 1406L, "Noncandidate WT opportunity count changed.")
make_txdb <- if (requireNamespace("txdbmaker", quietly = TRUE) &&
  "makeTxDbFromGFF" %in% getNamespaceExports("txdbmaker"))
  getExportedValue("txdbmaker", "makeTxDbFromGFF") else
    getExportedValue("GenomicFeatures", "makeTxDbFromGFF")
txdb <- make_txdb(files[["gtf"]], format = "gtf",
  organism = "Mus musculus")
gr <- GenomicRanges::GRanges(seqnames = other$chrom,
  ranges = IRanges::IRanges(other$start_1based, other$end_1based))
names(gr) <- other$feature
need(all(as.character(GenomicRanges::seqnames(gr)) %in%
  GenomeInfoDb::seqlevels(txdb)), "GTF and peak sequence names differ.")
annotated <- as.data.frame(ChIPseeker::annotatePeak(gr,
  TxDb = txdb, tssRegion = c(-2000, 2000),
  level = "gene", annoDb = NULL, verbose = FALSE))
annotated$feature <- paste0(as.character(annotated$seqnames), "-",
  annotated$start, "-", annotated$end)
need(nrow(annotated) == nrow(other) &&
  !anyDuplicated(annotated$feature) &&
  setequal(annotated$feature, other$feature),
  "Noncandidate ChIPseeker annotations could not be aligned.")
annotated <- annotated[match(other$feature, annotated$feature), ]
other$annotation <- as.character(annotated$annotation)
other$distanceToTSS <- as.numeric(annotated$distanceToTSS)
other$nearest_ensembl <- sub("\\.[0-9]+$", "",
  as.character(annotated$geneId))
other$nearest_ensembl[is.na(annotated$geneId) |
  !grepl("^ENSMUSG", other$nearest_ensembl)] <- NA_character_
other$promoter_pm2kb <- !is.na(other$distanceToTSS) &
  abs(other$distanceToTSS) <= 2000
keys <- sort(unique(stats::na.omit(other$nearest_ensembl)))
mapped <- AnnotationDbi::select(org.Mm.eg.db::org.Mm.eg.db,
  keys = keys, keytype = "ENSEMBL",
  columns = c("ENTREZID", "SYMBOL")) |>
  dplyr::filter(!is.na(.data$ENTREZID)) |>
  dplyr::arrange(.data$ENSEMBL, as.integer(.data$ENTREZID)) |>
  dplyr::distinct(.data$ENSEMBL, .keep_all = TRUE)
other <- dplyr::left_join(other, mapped |>
  dplyr::transmute(nearest_ensembl = .data$ENSEMBL,
    nearest_entrez = .data$ENTREZID,
    nearest_symbol = .data$SYMBOL),
  by = "nearest_ensembl", relationship = "many-to-one")
other$WT_direction <- "other_evaluable_WT"
fixed$nearest_entrez <- as.character(fixed$nearest_entrez)
other$nearest_entrez <- as.character(other$nearest_entrez)
all_peaks <- dplyr::bind_rows(fixed, other[, names(fixed),
  drop = FALSE])
need(nrow(all_peaks) == 2028L &&
  setequal(all_peaks$feature, u$feature),
  "The background does not contain every evaluable WT peak.")
write_csv(all_peaks, file.path(OUT,
  "01_all_2028_WT_evaluable_annotated_peaks.csv.gz"))
write_csv(all_peaks |>
  dplyr::count(.data$WT_direction, .data$promoter_pm2kb,
    name = "intervals"),
  file.path(OUT, "01_opportunity_set_composition.csv"))
write_csv(all_peaks |>
  dplyr::filter(!is.na(.data$nearest_entrez)) |>
  dplyr::count(.data$nearest_entrez, name = "evaluable_peak_opportunities") |>
  dplyr::arrange(dplyr::desc(.data$evaluable_peak_opportunities)),
  file.path(OUT, "01_gene_peak_opportunity_audit.csv"))
```

## WT HFD-opening and HFD-closing GO/KEGG tests

The WT foregrounds are compared with all evaluable WT peaks under the same
gate. Each gene enters the hypergeometric test once regardless of its number
of peaks; the opportunity audit above reports that potential bias. GO terms
are split by BP/MF/CC. Benjamini-Hochberg adjustment is within each
direction/scope/database result. The KEGG status is recorded if the online
resource is unavailable on SOL. No result is interpreted as a genotype
effect or a gene regulatory link.

```{r wt-program-ora}
status <- list()
kegg_unavailable <- FALSE
old_timeout <- getOption("timeout")
options(timeout = min(20L, old_timeout))
for (direction in c("HFD_open", "HFD_closed")) {
  for (scope in c("all", "promoter_pm2kb")) {
    universe <- all_peaks
    if (scope == "promoter_pm2kb")
      universe <- universe[universe$promoter_pm2kb, , drop = FALSE]
    target <- universe[universe$WT_direction == direction, , drop = FALSE]
    bg <- sort(unique(stats::na.omit(as.character(
      universe$nearest_entrez))))
    genes <- intersect(bg, sort(unique(stats::na.omit(as.character(
      target$nearest_entrez)))))
    write_csv(tibble::tibble(entrez_id = bg),
      file.path(ORA, paste0(direction, "_", scope,
        "_background_genes.csv")))
    write_csv(tibble::tibble(entrez_id = genes),
      file.path(ORA, paste0(direction, "_", scope,
        "_target_genes.csv")))
    for (db in c("GO_BP", "GO_MF", "GO_CC", "KEGG")) {
      message("WT ", direction, " / ", scope, " / ", db,
        ": ", length(genes), " of ", length(bg), " genes")
      result <- NULL; issue <- NA_character_
      if (db == "KEGG" && kegg_unavailable) {
        issue <- "skipped_after_earlier_KEGG_error"
      } else if (length(genes) >= 3L &&
          length(setdiff(bg, genes)) >= 3L) {
        result <- tryCatch({
          if (db == "KEGG") clusterProfiler::enrichKEGG(
            gene = genes, universe = bg, organism = "mmu",
            keyType = "ncbi-geneid", minGSSize = 5L,
            maxGSSize = 500L, pvalueCutoff = 1, qvalueCutoff = 1) else
              clusterProfiler::enrichGO(gene = genes,
                OrgDb = org.Mm.eg.db::org.Mm.eg.db,
                keyType = "ENTREZID", ont = sub("GO_", "", db),
                universe = bg, minGSSize = 5L, maxGSSize = 500L,
                pvalueCutoff = 1, qvalueCutoff = 1,
                readable = FALSE)
        }, error = function(e) { issue <<- conditionMessage(e); NULL })
        if (db == "KEGG" && !is.na(issue))
          kegg_unavailable <- TRUE
      } else issue <- "too_few_mapped_target_or_non_target_genes"
      table <- if (is.null(result)) data.frame(
        ID = character(), Description = character(),
        GeneRatio = character(), BgRatio = character(),
        pvalue = numeric(), p.adjust = numeric(),
        qvalue = numeric(), geneID = character(),
        Count = integer()) else as.data.frame(result)
      write_csv(table, file.path(ORA, paste0(direction, "_", scope,
        "_", db, ".csv")))
      status[[length(status) + 1L]] <- tibble::tibble(
        WT_direction = direction, scope = scope, database = db,
        foreground_peaks = nrow(target), background_peaks = nrow(universe),
        foreground_genes = length(genes), background_genes = length(bg),
        pathways_returned = nrow(table),
        min_BH = if (nrow(table) && any(is.finite(table$p.adjust)))
          min(table$p.adjust, na.rm = TRUE) else NA_real_,
        BH_below_005 = if (nrow(table))
          sum(table$p.adjust < .05, na.rm = TRUE) else 0L,
        status = if (!is.na(issue)) issue else if (!nrow(table))
          "no_pathways_after_size_filter" else "complete")
    }
  }
}
options(timeout = old_timeout)
write_csv(dplyr::bind_rows(status),
  file.path(OUT, "02_WT_program_ORA_summary.csv"))
motif <- read_csv(files[["motif_summary"]]) |>
  dplyr::filter(.data$set %in% c("HFD_open_622", "HFD_closed_622"))
need(nrow(motif) == 2L, "The two Step 08L WT motif results are missing.")
write_csv(motif, file.path(OUT,
  "03_existing_08L_WT_matched_neutral_motif_summary.csv"))
print(dplyr::bind_rows(status)); print(motif)
writeLines(c("08Q WT program ORA complete.",
  "This tests WT direction candidate peak genes versus all 2028 evaluable WT gate peaks.",
  "It does not test PPAR/IL17 loss or genotype-by-diet interaction.",
  "Step 08L already tested WT direction peak motifs versus matched WT neutral peaks."),
  file.path(OUT, "STEP_08Q_COMPLETE.txt"))
```
