---
title: "Step 08T v3 | All Gourab program gene-region coverage"
output:
  html_notebook:
    toc: true
    toc_float: true
params:
  run_coverage: true
---

Run from the repository root after Step 08N:

```sh
Rscript -e 'rmarkdown::render("analysis/08t_gourab_gene_region_overlay.Rmd")'
```

For a quick table-only pass, supply `params = list(run_coverage = FALSE)` to
`rmarkdown::render()`. The frozen 622 WT intervals and the Step 08N loss rule
remain unchanged. This is a **post hoc overlay**, not a new differential peak
call or a peak-to-gene regulatory link. A promoter overlap, gene-body overlap,
nearest-gene assignment, and proximity to a TSS are recorded separately. The
panel image's readable qPCR targets are prioritized; several are not in the
original curated lists and are explicitly labeled `qPCR_only` if matched.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
REPO_ROOT <- if (file.exists("multiome_scRNA-ATAC.Rproj")) normalizePath(".") else
  if (file.exists(file.path("..", "multiome_scRNA-ATAC.Rproj")))
    normalizePath("..") else stop("Run from the repository root.")
Sys.setenv(MULTIOME_DATASET_VERSION = "v2_resequenced", MULTIOME_COHORT_ID = "wt")
source(file.path(REPO_ROOT, "config", "project_config.R"))
packages <- c("readr", "dplyr", "tibble", "tidyr", "ggplot2", "IRanges",
  "GenomicRanges", "rtracklayer", "SeuratObject", "Signac", "patchwork")
missing <- packages[!vapply(packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Missing packages: ", paste(missing, collapse = ", "))
need <- function(test, why) if (!isTRUE(test)) stop(why, call. = FALSE)
read <- function(path) {
  need(file.exists(path), paste("Missing:", path))
  readr::read_csv(path, show_col_types = FALSE, progress = FALSE)
}
BASE <- file.path(RESULTS_ROOT, DATASET_VERSION, "independent")
INPUT <- file.path(OUTPUT_DIR, "08n_wt_ISC_interval_loss_pathways",
  "01_all_622_annotated_loss_audit.csv.gz")
OUT <- file.path(OUTPUT_DIR, "08t_gourab_gene_region_overlay")
PLOTS <- file.path(OUT, "plots")
dir.create(PLOTS, recursive = TRUE, showWarnings = FALSE)
need(file.exists(GRCM39_GTF), paste("Missing GRCm39 GTF:", GRCM39_GTF))
```

## Reference genes and matching rule

The curated lists below are the user-supplied sets. The image lists qPCR
targets; the image's Fc receptor label is too small to transcribe confidently
and is omitted pending confirmation. The qPCR comparison is descriptive: the
paper measured crypt RNA, whereas this step measures accessibility in
ISC-enriched nuclei.

```{r reference-genes}
hyper <- scan(text = paste(
  "Mgst1 Gstm1 Idh1 Gss Gsta4 Gclm Sod1 Sult1a1 Hmgcs2 Acsm3 Csad Acss2",
  "Shmt1 Impdh2 Akr7a5 Adk Aprt Hao2 Hadh Mecr Scd2 Mif Fads2 Acads Sox9",
  "Aldob Atpif1 Uqcrc1 Slc4a4 Hspd1 Eef1a1 Nox1 Cyba Tkt Gpx2 Gsto1 Adh1"),
  what = "", quiet = TRUE)
hypo <- scan(text = paste(
  "Nupr1 Crip1 Atp2a3 Cdkn1a Cd74 Herpud1 Dnajc10 Ern2 Itpr1 Xbp1 P4hb",
  "Selenok Selenos Creb3l1 Serinc3 Sh3glb1 H2-K1 H2-D1 B2m H2-T23 Psmb8",
  "Calr Ceacam1 Klf4 Mia3 Cd24a Rap1gap Pla2g2f Spint2 Trp53inp1 Gabarap",
  "Gabarapl2 Camk2n1 S100a13 Oit1 Syt7 Serp1 Rab3d Rab15 Agr2 Hspa5",
  "Manf Dnajc3 Creb3l4 Pdia3 Pdia6 Sdf2l1 Sec61b Ramp1 Galnt7 C1galt1",
  "Gfpt1 Gcnt3 St3gal6 Tmem59 St6galnac6 Galnt10 C1galt1c1 Fut2 Galnt3",
  "Rpn1 Krtcap2 Pmm2 Prdx6 Qsox1 Sh3bgrl3 Txndc5 Pdia5 Dap Foxa3 Clca1",
  "Tspan13 Atp2c2 P2rx4 Stim2 Asph Fxyd3 Lrrc26 Kit Plpp1 Pdxdc1 Rrbp1",
  "Tram1 Sec62 Ssr1 Bag1 Lman1 Actb Fos Il13ra1 Ifngr1"),
  what = "", quiet = TRUE)
qpcr_hyper <- c("Gsta4", "Mgst1", "Hmgcs2", "Idh1", "Aldob", "Shmt1",
  "Hadh", "Hao2")
qpcr_hypo <- c("H2-Eb1", "H2-Ab1", "Ifngr1", "Cd3e", "Tnfaip2",
  "Itgal", "Nox1", "Fos", "Tgfb1")
genes <- dplyr::bind_rows(
  tibble::tibble(gene = hyper, program = "HyperMetab", curated = TRUE,
    qpcr = hyper %in% qpcr_hyper),
  tibble::tibble(gene = hypo, program = "HypoImmune", curated = TRUE,
    qpcr = hypo %in% qpcr_hypo),
  tibble::tibble(gene = setdiff(qpcr_hyper, hyper), program = "HyperMetab",
    curated = FALSE, qpcr = TRUE),
  tibble::tibble(gene = setdiff(qpcr_hypo, hypo), program = "HypoImmune",
    curated = FALSE, qpcr = TRUE)) |>
  dplyr::distinct(.data$gene, .data$program, .keep_all = TRUE)
need(!anyDuplicated(genes[c("gene", "program")]), "Duplicate reference gene")
readr::write_csv(genes, file.path(OUT, "00_reference_gene_sets.csv"))

r <- read(INPUT)
need(nrow(r) == 622L && !anyDuplicated(r$feature),
  "Step 08N frozen 622 candidates changed")
required <- c("feature", "chrom", "start_1based", "end_1based",
  "nearest_symbol", "distanceToTSS", "annotation", "promoter_pm2kb",
  "WT_direction", "comparison", "anchor_log2FC_HFD_vs_CON", "C0_supported",
  "count_supported_40", "ppar_evaluable", "il17_evaluable", "ppar_loss",
  "il17_loss", "ppar_effect", "il17_effect", "ppar_loss_subtype",
  "il17_loss_subtype", "anchor_CON_counts", "anchor_HFD_counts")
need(all(required %in% names(r)), "Step 08N annotated audit schema changed")
need(all(r$ppar_loss <= r$ppar_evaluable) && all(r$il17_loss <= r$il17_evaluable),
  "An unevaluable knockout interval was called lost")
peaks <- GenomicRanges::GRanges(seqnames = r$chrom,
  ranges = IRanges::IRanges(r$start_1based, r$end_1based))

# Retain one GTF gene record per Ensembl ID; multi-transcript entries are
# deliberately ignored. Match overlapping gene bodies and TSS +/-2 kb; allow
# TSS +/-20 kb only as an explicitly weaker proximity match.
gtf <- rtracklayer::import(GRCM39_GTF)
gtf <- gtf[as.character(S4Vectors::mcols(gtf)$type) == "gene"]
need(length(gtf) > 0L && all(c("gene_name", "gene_id") %in%
  names(S4Vectors::mcols(gtf))), "GTF gene_id/gene_name unavailable")
gtf_symbol <- as.character(S4Vectors::mcols(gtf)$gene_name)
gtf <- gtf[!is.na(gtf_symbol) & gtf_symbol %in% genes$gene]
gtf_symbol <- as.character(S4Vectors::mcols(gtf)$gene_name)
gtf_id <- as.character(S4Vectors::mcols(gtf)$gene_id)
gtf <- gtf[!duplicated(gtf_id)]
gtf_symbol <- as.character(S4Vectors::mcols(gtf)$gene_name)
need(length(gtf) > 0L, "None of the curated genes were found in the GTF")
tss_pos <- ifelse(as.character(GenomicRanges::strand(gtf)) == "-",
  BiocGenerics::end(gtf), BiocGenerics::start(gtf))
tss <- GenomicRanges::GRanges(GenomicRanges::seqnames(gtf),
  IRanges::IRanges(tss_pos, tss_pos))
prom <- GenomicRanges::resize(tss, width = 4001L, fix = "center")
IRanges::start(prom) <- pmax(1L, IRanges::start(prom))
near <- GenomicRanges::resize(tss, width = 40001L, fix = "center")
IRanges::start(near) <- pmax(1L, IRanges::start(near))
body_hits <- GenomicRanges::findOverlaps(peaks, gtf, ignore.strand = TRUE)
prom_hits <- GenomicRanges::findOverlaps(peaks, prom, ignore.strand = TRUE)
near_hits <- GenomicRanges::findOverlaps(peaks, near, ignore.strand = TRUE)
hit_keys <- function(h) {
  if (!length(h)) return(tibble::tibble(feature = character(), gene = character()))
  tibble::tibble(feature = r$feature[S4Vectors::queryHits(h)],
    gene = gtf_symbol[S4Vectors::subjectHits(h)]) |> dplyr::distinct()
}
body_keys <- hit_keys(body_hits)
prom_keys <- hit_keys(prom_hits)
near_keys <- hit_keys(near_hits)
nearest_keys <- r |>
  dplyr::filter(!is.na(.data$nearest_symbol),
    .data$nearest_symbol %in% genes$gene) |>
  dplyr::transmute(feature = .data$feature, gene = .data$nearest_symbol) |>
  dplyr::distinct()
matches <- dplyr::bind_rows(body_keys, prom_keys, near_keys, nearest_keys) |>
  dplyr::distinct() |>
  dplyr::left_join(genes, by = "gene", relationship = "many-to-many") |>
  dplyr::left_join(r, by = "feature", relationship = "many-to-one") |>
  dplyr::mutate(
    promoter_overlap = paste(.data$feature, .data$gene) %in%
      paste(prom_keys$feature, prom_keys$gene),
    gene_body_overlap = paste(.data$feature, .data$gene) %in%
      paste(body_keys$feature, body_keys$gene),
    within_20kb_TSS = paste(.data$feature, .data$gene) %in%
      paste(near_keys$feature, near_keys$gene),
    assigned_nearest_gene = paste(.data$feature, .data$gene) %in%
      paste(nearest_keys$feature, nearest_keys$gene),
    evidence = dplyr::case_when(.data$promoter_overlap ~ "promoter_overlap_pm2kb",
      .data$gene_body_overlap ~ "gene_body_overlap",
      .data$assigned_nearest_gene & !is.na(.data$distanceToTSS) &
        abs(.data$distanceToTSS) <= 20000 ~ "nearest_gene_TSS_20kb",
      .data$assigned_nearest_gene ~ "nearest_gene_more_distant",
      TRUE ~ "other_gene_TSS_20kb"),
    evidence_rank = match(.data$evidence,
      c("promoter_overlap_pm2kb", "gene_body_overlap", "nearest_gene_TSS_20kb",
        "nearest_gene_more_distant", "other_gene_TSS_20kb")),
    loss_group = dplyr::case_when(.data$ppar_loss & .data$il17_loss ~ "shared",
      .data$ppar_loss ~ "PPAR_only", .data$il17_loss ~ "IL17_only",
      TRUE ~ "neither"),
    expected_RNA_direction = ifelse(.data$program == "HyperMetab",
      "HFD_higher", "HFD_lower"),
    matching_ATAC_direction = (.data$program == "HyperMetab" &
      .data$WT_direction == "HFD_open") | (.data$program == "HypoImmune" &
      .data$WT_direction == "HFD_closed"),
    list_origin = ifelse(.data$curated, "curated", "qPCR_only")) |>
  dplyr::arrange(.data$program, .data$gene, .data$evidence_rank,
    .data$feature)
need(!anyDuplicated(matches[c("feature", "gene", "program")]),
  "Multiple unresolved associations per peak/gene/program")
readr::write_csv(matches, file.path(OUT, "01_all_candidate_gene_matches.csv.gz"))
attenuated <- matches |>
  dplyr::filter(.data$ppar_loss | .data$il17_loss) |>
  dplyr::arrange(dplyr::desc(.data$qpcr), .data$evidence_rank,
    dplyr::desc(.data$matching_ATAC_direction),
    dplyr::desc(.data$C0_supported), dplyr::desc(.data$count_supported_40),
    dplyr::desc(abs(.data$anchor_log2FC_HFD_vs_CON)), .data$feature)
readr::write_csv(attenuated, file.path(OUT, "02_attenuated_gene_matches.csv"))
coverage <- genes |>
  dplyr::left_join(matches |>
    dplyr::group_by(.data$program, .data$gene) |>
    dplyr::summarise(n_candidate = dplyr::n(),
      n_attenuated = sum(.data$ppar_loss | .data$il17_loss),
      n_shared = sum(.data$loss_group == "shared"),
      best_match = .data$evidence[which.min(.data$evidence_rank)],
      .groups = "drop"), by = c("program", "gene")) |>
  dplyr::mutate(dplyr::across(c("n_candidate", "n_attenuated", "n_shared"),
    ~ tidyr::replace_na(.x, 0L)))
readr::write_csv(coverage, file.path(OUT, "03_all_genes_including_zero_hits.csv"))
print(attenuated |>
  dplyr::count(.data$program, .data$WT_direction, .data$loss_group,
    .data$evidence, name = "n_peak_gene_matches"))
```

## All matched region effects

Include every peak-gene match in the two programs, including opposite-direction
and genotype-specific losses. The match table also preserves `qpcr`, evidence
class, and the expected RNA direction. A peak is plotted once per matched
gene/program, so multiple peaks at the same gene get separate coverage PDFs.
Unevaluable knockout effects remain missing in the effect plot.

```{r select-and-plot}
pick <- matches |>
  dplyr::arrange(.data$program, .data$gene, .data$feature)
readr::write_csv(pick, file.path(OUT, "04_all_coverage_candidates.csv"))
if (nrow(pick)) {
  effect <- dplyr::bind_rows(lapply(seq_len(nrow(pick)), function(j) {
    v <- pick[j, ]
    tibble::tibble(program = v$program, gene = v$gene, feature = v$feature,
      locus = paste(v$gene, v$feature),
      genotype = c("WT", "PPAR", "IL17"),
      log2FC = c(v$anchor_log2FC_HFD_vs_CON,
        if (v$ppar_evaluable) v$ppar_effect else NA_real_,
        if (v$il17_evaluable) v$il17_effect else NA_real_),
      status = c("WT candidate", v$ppar_loss_subtype,
        v$il17_loss_subtype))
  }))
  readr::write_csv(effect, file.path(OUT, "04_all_pooled_ATAC_effects.csv"))
  p <- ggplot2::ggplot(effect, ggplot2::aes(.data$genotype, .data$log2FC,
    color = .data$genotype)) +
    ggplot2::geom_hline(yintercept = 0, color = "grey65") +
    ggplot2::geom_point(size = 2.8, na.rm = TRUE) +
    ggplot2::facet_wrap(~ program + locus, scales = "free_x", ncol = 2) +
    ggplot2::scale_color_manual(values = c(WT = "#E69F00", PPAR = "#5A3E9A",
      IL17 = "#56B4E9")) +
    ggplot2::labs(title = "All Gourab gene overlays: pooled ATAC effects at exact peaks",
      subtitle = "Matched genes identify loci; a nearby peak does not establish regulatory coupling",
      x = NULL, y = "ATAC log2(HFD / CON)",
      caption = "One pooled library per diet and genotype; descriptive effects only. Missing points are unevaluable.") +
    ggplot2::theme_bw(base_size = 10)
  ggplot2::ggsave(file.path(PLOTS, "01_all_locus_effects.pdf"), p,
    width = 11, height = max(5, 3 * ceiling(nrow(pick) / 2)),
    device = grDevices::pdf, useDingbats = FALSE)
}
```

## Coverage on the same coordinates across cohorts

Use the recorded top-15% nuclei that generated the Step 08N effects (WT
C0/W0, PPAR 3/6, IL17 9/10). A common y-axis for the three normalized tracks
within each locus avoids independent `q95` clipping. Signac's group scaling
accounts for cell number and sequencing depth; sparse signals and unequal
fragment quality can still limit interpretation. We record any failure.

```{r coverage}
gate_files <- c(wt = file.path(OUTPUT_DIR,
    "08h_wt_ISC_region_freeze_sex_audit",
    "01_nucleus_level_sex_proxy_and_gate_audit.csv.gz"),
  ppar = file.path(OUTPUT_DIR,
    "08i_frozen_WT_peaks_genotype_context",
    "02_ppar_crypt_nuclei_scores.csv.gz"),
  il17 = file.path(OUTPUT_DIR,
    "08i_frozen_WT_peaks_genotype_context",
    "02_il17_crypt_nuclei_scores.csv.gz"))
get_cells <- function(cohort) {
  gate <- read(gate_files[[cohort]])
  need(!anyDuplicated(gate$cell_barcode), paste("Duplicate gate cells", cohort))
  if (cohort == "wt") {
    need(all(c("C0W0_top15", "cell_barcode") %in% names(gate)),
      "WT gate columns changed")
    selected <- as.logical(gate$C0W0_top15)
  } else {
    need(all(c("top15", "wnn_cluster", "cell_barcode") %in% names(gate)),
      paste("KO gate columns changed:", cohort))
    parents <- if (cohort == "ppar") c("3", "6") else c("9", "10")
    selected <- as.logical(gate$top15) &
      as.character(gate$wnn_cluster) %in% parents
  }
  need(!anyNA(selected), paste("Gate has missing selection:", cohort))
  unique(gate$cell_barcode[selected])
}
status <- list()
if (isTRUE(params$run_coverage) && nrow(pick)) {
  need(all(file.exists(gate_files)), "Recorded Step 08H/08I gate table is missing")
  # Preserve cohort names so every plot uses its recorded top-15% gate.
  cells_by_cohort <- stats::setNames(lapply(names(gate_files), get_cells),
    names(gate_files))
  gate_counts <- dplyr::bind_rows(lapply(names(gate_files), function(cohort) {
    gate <- read(gate_files[[cohort]])
    need("diet" %in% names(gate), paste("Missing diet in gate:", cohort))
    selected_diets <- as.character(gate$diet[match(cells_by_cohort[[cohort]],
      gate$cell_barcode)])
    need(!anyNA(selected_diets) &&
      all(selected_diets %in% c("CON", "HFD")),
      paste("Selected cells absent from gate or have unknown diet:", cohort))
    # Compute before tibble(): its diet column would otherwise shadow the
    # selected_diets vector and produce the misleading counts 1 and 1.
    diet_counts <- table(factor(selected_diets, levels = c("CON", "HFD")))
    need(sum(diet_counts) == length(cells_by_cohort[[cohort]]),
      paste("Incomplete selected-nucleus diet count:", cohort))
    tibble::tibble(cohort = cohort, diet = names(diet_counts),
      n_selected = as.integer(diet_counts))
  }))
  readr::write_csv(gate_counts,
    file.path(OUT, "05_recorded_gate_nuclei_by_diet.csv"))
  print(gate_counts)
  need(all(gate_counts$n_selected >= 30L), paste(
    "Recorded top-15% gate below 30 nuclei per diet:",
    paste(gate_counts$cohort[gate_counts$n_selected < 30L],
      gate_counts$diet[gate_counts$n_selected < 30L],
      gate_counts$n_selected[gate_counts$n_selected < 30L],
      collapse = "; ")))
  for (j in seq_len(nrow(pick))) {
    x <- pick[j, ]
    loss_label <- switch(as.character(x$loss_group),
      shared = "PPAR_and_IL17_loss", PPAR_only = "PPAR_only_loss",
      IL17_only = "IL17_only_loss", neither = "no_recorded_loss")
    need(!is.null(loss_label), paste("Unexpected loss label:", x$loss_group))
    stem <- paste0(x$program, "_", x$gene, "_", x$feature, "_", loss_label)
    loc <- GenomicRanges::GRanges(x$chrom,
      IRanges::IRanges(x$start_1based, x$end_1based))
    span <- GenomicRanges::GRanges(x$chrom,
      IRanges::IRanges(max(1L, x$start_1based - 5000L),
        x$end_1based + 5000L))
    prelim <- list()
    for (cohort in names(gate_files)) {
      tryCatch({
        obj <- readRDS(file.path(BASE, cohort, "06_annotation",
          "multiome_annotated.rds"))
        SeuratObject::DefaultAssay(obj) <- "ATAC"
        cells <- cells_by_cohort[[cohort]]
        m <- obj[[]]
        need(all(cells %in% rownames(m)), paste("Missing selected cells:", cohort))
        n <- table(factor(as.character(m[cells, "diet"]),
          levels = c("CON", "HFD")))
        expected_n <- gate_counts$n_selected[gate_counts$cohort == cohort]
        need(identical(as.integer(n), as.integer(expected_n)),
          paste("Recorded gate and object diet counts disagree:", cohort,
            "object", paste(as.integer(n), collapse = "/"),
            "gate", paste(expected_n, collapse = "/")))
        need(length(Signac::Fragments(obj[["ATAC"]])) > 0L,
          paste("Missing ATAC fragments:", cohort))
        p0 <- Signac::CoveragePlot(obj, region = span, assay = "ATAC",
          cells = cells, group.by = "diet", annotation = FALSE,
          peaks = FALSE, links = FALSE, tile = FALSE,
          region.highlight = loc, window = 100L,
          max.downsample = 3000L, downsample.rate = 1, ymax = NULL)
        yy <- p0[[1]]$data$score
        if (is.null(yy) || !length(yy)) {
          b <- ggplot2::ggplot_build(p0[[1]])$data
          yy <- unlist(lapply(b, function(a)
            if ("y" %in% names(a)) a$y else numeric()), use.names = FALSE)
        }
        yy <- yy[is.finite(yy)]
        need(length(yy) > 0L, paste("Cannot read normalized track:", cohort))
        prelim[[cohort]] <- list(max = max(yy), n = n)
      }, error = function(e) {
        status[[length(status) + 1L]] <<- tibble::tibble(
          feature = x$feature, gene = x$gene, cohort = cohort,
          status = "preflight_failed", detail = conditionMessage(e))
      })
      if (exists("obj", inherits = FALSE)) rm(obj)
      invisible(gc())
    }
    if (!length(prelim)) next
    ymax <- max(1e-3, max(vapply(prelim, `[[`, numeric(1), "max")) * 1.04)
    pages <- list()
    for (cohort in names(prelim)) {
      tryCatch({
        obj <- readRDS(file.path(BASE, cohort, "06_annotation",
          "multiome_annotated.rds"))
        SeuratObject::DefaultAssay(obj) <- "ATAC"
        p1 <- Signac::CoveragePlot(obj, region = span, assay = "ATAC",
          cells = cells_by_cohort[[cohort]], group.by = "diet",
          annotation = FALSE, peaks = FALSE, links = FALSE, tile = FALSE,
          region.highlight = loc, window = 100L, max.downsample = 3000L,
          downsample.rate = 1, ymax = ymax) +
          patchwork::plot_annotation(title = paste(toupper(cohort), "CON",
            prelim[[cohort]]$n[["CON"]], "/ HFD",
            prelim[[cohort]]$n[["HFD"]]),
            subtitle = paste("Common normalized y max", signif(ymax, 3)))
        pages[[cohort]] <- p1
        filename <- paste0("02_", stem, "_", cohort, "_coverage.pdf")
        ggplot2::ggsave(file.path(PLOTS, filename), p1, width = 13,
          height = 6, device = grDevices::pdf, useDingbats = FALSE)
        status[[length(status) + 1L]] <- tibble::tibble(
          feature = x$feature, gene = x$gene, cohort = cohort,
          status = "saved_cohort", detail = filename)
      }, error = function(e) {
        status[[length(status) + 1L]] <<- tibble::tibble(
          feature = x$feature, gene = x$gene, cohort = cohort,
          status = "plot_failed", detail = conditionMessage(e))
      })
      if (exists("obj", inherits = FALSE)) rm(obj)
      invisible(gc())
    }
    if (length(pages) >= 2L) {
      fig <- patchwork::wrap_plots(pages, ncol = 1L) +
        patchwork::plot_annotation(
          title = paste(x$program, x$gene, x$feature),
          subtitle = paste("GTF match:", x$evidence, "|", loss_label,
            "| WT", x$WT_direction),
          caption = "Same genomic interval highlighted in three pooled, depth-normalized cohort tracks. Gene proximity does not imply regulation.")
      filename <- paste0("03_", stem, "_common_scale_coverage.pdf")
      tryCatch({
        ggplot2::ggsave(file.path(PLOTS, filename), fig,
          width = 13, height = 4.7 * length(pages), device = grDevices::pdf,
          useDingbats = FALSE)
        status[[length(status) + 1L]] <- tibble::tibble(
          feature = x$feature, gene = x$gene, cohort = "combined",
          status = "saved_combined", detail = filename)
      }, error = function(e) {
        status[[length(status) + 1L]] <<- tibble::tibble(
          feature = x$feature, gene = x$gene, cohort = "combined",
          status = "save_failed", detail = conditionMessage(e))
      })
    }
    rm(prelim, pages); invisible(gc())
  }
}
status_table <- if (length(status)) dplyr::bind_rows(status) else
  tibble::tibble(feature = character(), gene = character(), cohort = character(),
    status = character(), detail = character())
readr::write_csv(status_table, file.path(OUT, "05_coverage_status.csv"))
writeLines(c("Step 08T v3 complete", paste("matches", nrow(matches)),
  paste("attenuated_matches", nrow(attenuated)),
  paste("all_matched_tracks", nrow(pick)),
  paste("coverage_requested", isTRUE(params$run_coverage)),
  "Review 03_all_genes_including_zero_hits.csv and 05_coverage_status.csv before presenting."),
  file.path(OUT, "STEP_08T_V3_COMPLETE.txt"))
```
