---
title: "Meeting figures: WT diet response and cross-cohort program projection"
author: "Dominic Saiz"
date: "2026-09-15"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# Purpose and interpretation boundary

This standalone notebook prepares editable, presentation-ready PDF figures for
the WT diet-response update and projects the same expression programs into the
PPAR and IL17 cohorts. It does not alter the primary pipeline objects.

Each diet/genotype combination is represented by one pooled library. Therefore:

- every CON-versus-HFD comparison is descriptive;
- nuclei are not biological replicates;
- balanced cell downsampling measures cell-sampling stability, not biological
  uncertainty;
- nearest-gene or nearby-peak concordance does not establish a regulatory
  connection;
- cross-cohort score differences may reflect genotype, pool composition, and
  technical differences in addition to biology.

The notebook writes the exact table underlying every plotted claim. The PDF
pages use a 16:9 canvas and vector graphics for editing in Illustrator.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c(
  "Seurat",
  "SeuratObject",
  "UCell",
  "Matrix",
  "dplyr",
  "tidyr",
  "readr",
  "ggplot2",
  "patchwork",
  "tibble",
  "future",
  "scales",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    "Install the following package(s) before running this notebook: ",
    paste(missing_packages, collapse = ", "),
    "\nFor UCell, use BiocManager::install('UCell').",
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(UCell)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(readr)
  library(ggplot2)
  library(patchwork)
  library(tibble)
})

set.seed(12345)
future::plan(future::sequential)

knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center"
)
```

# 1. Configuration

Change only the values in this chunk when the results root or meeting label
changes.

```{r configure-analysis}
PROJECT_DIR <- "/scratch/dsaiz/Yesenia_scData2026"
DATASET_VERSION <- "v2_resequenced"
PRESENTATION_TAG <- "2026-09-15_diet_program_update"

REFERENCE_DIET <- "CON"
COMPARISON_DIET <- "HFD"
COHORT_IDS <- c("wt", "ppar", "il17")
COHORT_LABELS <- c(
  wt = "WT",
  ppar = "Ppar iKO",
  il17 = "Il17ra iKO"
)

UCELL_MAX_RANK <- 1000L
MIN_NUCLEI_PER_DIET <- 50L
MAX_DOWNSAMPLED_NUCLEI_PER_DIET <- 500L
DOWNSAMPLING_ITERATIONS <- 100L
TOP_RNA_GENES_PER_STATE_AND_DIRECTION <- 3L
TOP_MULTIMODAL_PAIRS_PER_STATE_AND_DIRECTION <- 3L
NEARBY_GENE_DISTANCE_BP <- 2000L
RANDOM_SEED <- 12345L

RESULT_ROOT <- file.path(RESULTS_ROOT,
  DATASET_VERSION,
  "independent"
)

MEETING_OUTPUT_DIR <- file.path(RESULTS_ROOT,
  DATASET_VERSION,
  "presentation",
  PRESENTATION_TAG
)

PLOT_DIR <- file.path(MEETING_OUTPUT_DIR, "plots")
TABLE_DIR <- file.path(MEETING_OUTPUT_DIR, "tables")

dir.create(PLOT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)

STEP6_OBJECT_PATHS <- setNames(
  file.path(
    RESULT_ROOT,
    COHORT_IDS,
    "06_annotation",
    "multiome_annotated.rds"
  ),
  COHORT_IDS
)

WT_STEP7C_TABLE_DIR <- file.path(
  RESULT_ROOT,
  "wt",
  "07c_candidate_prioritization",
  "tables"
)

WT_STEP7_TABLE_DIR <- file.path(
  RESULT_ROOT,
  "wt",
  "07_diet_response",
  "tables"
)

WT_COMPOSITION_EFFECT_PATH <- file.path(
  WT_STEP7_TABLE_DIR,
  "composition_effect_sizes.csv"
)

WT_STEP8_TABLE_DIR <- file.path(
  RESULT_ROOT,
  "wt",
  "08_persistent_hfd_programs",
  "tables"
)

WT_RNA_PRIORITY_PATH <- file.path(
  WT_STEP7C_TABLE_DIR,
  "RNA_prioritized_diet_candidates.csv"
)

WT_MULTIMODAL_PRIORITY_PATH <- file.path(
  WT_STEP7C_TABLE_DIR,
  "RNA_ATAC_concordant_prioritized_candidates.csv"
)

WT_PERSISTENT_EVIDENCE_PATH <- file.path(
  WT_STEP8_TABLE_DIR,
  "persistent_gene_primary_WT_stem_integrated_evidence.csv"
)
```

## Plot styling

```{r define-presentation-style}
DIET_COLORS <- c(
  CON = "#E69F00",
  HFD = "#56B4E9"
)

COHORT_COLORS <- c(
  WT = "#333333",
  `Ppar iKO` = "#CC79A7",
  `Il17ra iKO` = "#009E73"
)

DIRECTION_COLORS <- c(
  CON_higher = DIET_COLORS[["CON"]],
  HFD_higher = DIET_COLORS[["HFD"]]
)

STATE_COLORS <- c(
  "ISC-like crypt" = "#4C78A8",
  "Cycling ISC/TA" = "#72B7B2",
  "Crypt progenitor" = "#54A24B",
  "Deep crypt secretory" = "#F58518",
  "Secretory progenitor" = "#E45756",
  "Immature absorptive" = "#B279A2",
  "Mature absorptive / colonocyte" = "#FF9DA6",
  "Immature goblet" = "#9D755D",
  "Mature goblet" = "#BAB0AC",
  "Tuft" = "#EDC948",
  "Enterochromaffin" = "#8CD17D",
  "Other EEC" = "#499894",
  "Other epithelial" = "#A0A0A0"
)

theme_presentation <- function(base_size = 18) {
  ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold",
        size = base_size * 1.30,
        margin = ggplot2::margin(b = 8)
      ),
      plot.subtitle = ggplot2::element_text(
        size = base_size * 0.80,
        color = "grey25",
        margin = ggplot2::margin(b = 12)
      ),
      plot.caption = ggplot2::element_text(
        size = base_size * 0.62,
        color = "grey35",
        hjust = 0,
        margin = ggplot2::margin(t = 10)
      ),
      axis.title = ggplot2::element_text(face = "bold"),
      axis.text = ggplot2::element_text(color = "black"),
      strip.background = ggplot2::element_rect(
        fill = "grey94",
        color = NA
      ),
      strip.text = ggplot2::element_text(face = "bold"),
      legend.title = ggplot2::element_text(face = "bold"),
      legend.position = "right",
      panel.grid.major.y = ggplot2::element_line(
        color = "grey90",
        linewidth = 0.3
      )
    )
}

PREVIEW_PLOTS_IN_RSTUDIO <- FALSE

preview_plot <- function(plot) {
  if (isTRUE(PREVIEW_PLOTS_IN_RSTUDIO)) {
    print(plot)
  }

  invisible(plot)
}

save_presentation_pdf <- function(filename, plot, width = 13.333,
                                  height = 7.5) {
  output_path <- file.path(PLOT_DIR, filename)

  grDevices::pdf(
    file = output_path,
    width = width,
    height = height,
    onefile = TRUE,
    family = "Helvetica",
    paper = "special",
    useDingbats = FALSE
  )

  opened_device <- grDevices::dev.cur()

  on.exit(
    {
      open_devices <- grDevices::dev.list()
      if (!is.null(open_devices) && opened_device %in% open_devices) {
        grDevices::dev.off(which = opened_device)
      }
    },
    add = TRUE
  )

  print(plot)
  invisible(output_path)
}
```

## Validation helpers

```{r define-validation-helpers}
assert_file <- function(path, label) {
  if (!file.exists(path)) {
    stop(label, " was not found:\n  ", path, call. = FALSE)
  }
  invisible(path)
}

assert_columns <- function(data, columns, label) {
  missing_columns <- setdiff(columns, colnames(data))
  if (length(missing_columns) > 0L) {
    stop(
      label,
      " is missing column(s): ",
      paste(missing_columns, collapse = ", "),
      call. = FALSE
    )
  }
  invisible(TRUE)
}

as_logical_safely <- function(x) {
  if (is.logical(x)) {
    return(x)
  }
  tolower(trimws(as.character(x))) %in% c("true", "t", "1", "yes")
}

safe_median <- function(x) {
  if (length(x) == 0L || all(is.na(x))) {
    return(NA_real_)
  }
  stats::median(x, na.rm = TRUE)
}

for (cohort_id in COHORT_IDS) {
  assert_file(
    STEP6_OBJECT_PATHS[[cohort_id]],
    paste0(COHORT_LABELS[[cohort_id]], " Step 6 object")
  )
}

assert_file(WT_RNA_PRIORITY_PATH, "WT Step 7c RNA priority table")
assert_file(WT_COMPOSITION_EFFECT_PATH, "WT Step 7 composition effect table")
assert_file(
  WT_MULTIMODAL_PRIORITY_PATH,
  "WT Step 7c RNA/ATAC priority table"
)
assert_file(WT_PERSISTENT_EVIDENCE_PATH, "WT Step 8 evidence table")
```

# 2. Define the expression programs

The working descriptions are used in figures. The legacy labels remain in the
program key for continuity with Gourab's terminology. `Atpif1` is represented
by the current mouse symbol `Atp5if1`.

```{r define-programs}
HYPERMET_GENES <- c(
  "Mgst1", "Gstm1", "Idh1", "Gss", "Gsta4", "Gclm", "Sod1",
  "Sult1a1", "Hmgcs2", "Acsm3", "Csad", "Acss2", "Shmt1",
  "Impdh2", "Akr7a5", "Adk", "Aprt", "Hao2", "Hadh", "Mecr",
  "Scd2", "Mif", "Fads2", "Acads", "Sox9", "Aldob", "Atp5if1",
  "Uqcrc1", "Slc4a4", "Hspd1", "Eef1a1", "Nox1", "Cyba", "Tkt",
  "Gpx2", "Gsto1", "Adh1"
)

HYPOIMMUNE_GENES <- c(
  "Nupr1", "Crip1", "Atp2a3", "Cdkn1a", "Cd74", "Herpud1",
  "Dnajc10", "Ern2", "Itpr1", "Xbp1", "P4hb", "Selenok", "Selenos",
  "Creb3l1", "Serinc3", "Sh3glb1", "H2-K1", "H2-D1", "B2m",
  "H2-T23", "Psmb8", "Calr", "Ceacam1", "Klf4", "Mia3", "Cd24a",
  "Rap1gap", "Pla2g2f", "Spint2", "Trp53inp1", "Gabarap",
  "Gabarapl2", "Camk2n1", "S100a13", "Oit1", "Syt7", "Serp1",
  "Rab3d", "Rab15", "Agr2", "Hspa5", "Manf", "Dnajc3", "Creb3l4",
  "Pdia3", "Pdia6", "Sdf2l1", "Sec61b", "Ramp1", "Galnt7",
  "C1galt1", "Gfpt1", "Gcnt3", "St3gal6", "Tmem59", "St6galnac6",
  "Galnt10", "C1galt1c1", "Fut2", "Galnt3", "Rpn1", "Krtcap2",
  "Pmm2", "Prdx6", "Qsox1", "Sh3bgrl3", "Txndc5", "Pdia5", "Dap",
  "Foxa3", "Clca1", "Tspan13", "Atp2c2", "P2rx4", "Stim2", "Asph",
  "Fxyd3", "Lrrc26", "Kit", "Plpp1", "Pdxdc1", "Rrbp1", "Tram1",
  "Sec62", "Ssr1", "Bag1", "Lman1", "Actb", "Fos", "Il13ra1",
  "Ifngr1"
)

persistent_evidence <- suppressMessages(
  readr::read_csv(
    WT_PERSISTENT_EVIDENCE_PATH,
    show_col_types = FALSE
  )
)

assert_columns(
  persistent_evidence,
  c(
    "gene", "reference_direction", "reference_program",
    "gene_in_RNA_assay", "RNA_reference_direction_concordant",
    "RNA_clear_effect", "RNA_direction_stable",
    "RNA_downsampling_iqr_excludes_zero", "RNA_validation_class",
    "RNA_reference_aligned_log2FC",
    "representative_ATAC_reference_aligned_log2FC",
    "representative_promoter_peak",
    "representative_peak_offset_bp",
    "representative_peak_overlaps_tss",
    "promoter_ATAC_summary_class",
    "multimodal_validation_class"
  ),
  "WT Step 8 evidence table"
)

persistent_evidence <- persistent_evidence |>
  dplyr::mutate(
    dplyr::across(
      c(
        gene_in_RNA_assay,
        RNA_reference_direction_concordant,
        RNA_clear_effect,
        RNA_direction_stable,
        RNA_downsampling_iqr_excludes_zero
      ),
      as_logical_safely
    )
  )

persistent_stringent <- persistent_evidence |>
  dplyr::filter(
    gene_in_RNA_assay,
    RNA_reference_direction_concordant,
    RNA_clear_effect,
    RNA_direction_stable,
    RNA_downsampling_iqr_excludes_zero
  )

persistent_extended <- persistent_evidence |>
  dplyr::filter(
    gene_in_RNA_assay,
    RNA_reference_direction_concordant,
    RNA_direction_stable
  )

genes_by_reference_direction <- function(data, direction) {
  data |>
    dplyr::filter(reference_direction == direction) |>
    dplyr::pull(gene) |>
    unique() |>
    sort()
}

requested_signatures <- list(
  HyperMet = unique(HYPERMET_GENES),
  HypoImmune = unique(HYPOIMMUNE_GENES),
  Persistent_up_stringent = genes_by_reference_direction(
    persistent_stringent,
    "HFD_higher"
  ),
  Persistent_down_stringent = genes_by_reference_direction(
    persistent_stringent,
    "CON_higher"
  ),
  Persistent_up_extended = genes_by_reference_direction(
    persistent_extended,
    "HFD_higher"
  ),
  Persistent_down_extended = genes_by_reference_direction(
    persistent_extended,
    "CON_higher"
  )
)

empty_signatures <- names(requested_signatures)[
  lengths(requested_signatures) == 0L
]

if (length(empty_signatures) > 0L) {
  stop(
    "Step 8 produced an empty signature for: ",
    paste(empty_signatures, collapse = ", "),
    call. = FALSE
  )
}

program_key <- tibble::tribble(
  ~program, ~working_descriptor, ~legacy_label, ~expected_HFD_direction,
  "HyperMet", "Persistent HFD-associated metabolic/redox module",
  "HyperMet", "higher",
  "HypoImmune", "Persistent HFD-depleted immune/ER-secretory module",
  "HypoImmune", "lower",
  "Persistent_up_stringent", "Stringent WT-recovered persistent HFD-up arm",
  NA_character_, "higher",
  "Persistent_down_stringent", "Stringent WT-recovered persistent HFD-down arm",
  NA_character_, "lower",
  "Persistent_up_extended", "Extended stable WT-recovered persistent HFD-up arm",
  NA_character_, "higher",
  "Persistent_down_extended", "Extended stable WT-recovered persistent HFD-down arm",
  NA_character_, "lower"
)

program_definition_requested <- dplyr::bind_rows(
  lapply(
    names(requested_signatures),
    function(program_name) {
      tibble::tibble(
        program = program_name,
        gene = requested_signatures[[program_name]]
      )
    }
  )
) |>
  dplyr::left_join(program_key, by = "program")

readr::write_csv(
  program_definition_requested,
  file.path(TABLE_DIR, "program_definition_requested.csv")
)

readr::write_csv(
  tibble::tibble(
    source_symbol = "Atpif1",
    scoring_symbol = "Atp5if1",
    program = "HyperMet",
    rationale = "Use the current mouse gene symbol for assay matching."
  ),
  file.path(TABLE_DIR, "program_symbol_updates.csv")
)

program_key
```

# 3. Harmonize epithelial state labels across cohorts

The mapping preserves each original annotation and only creates a shared label
for descriptive cross-cohort displays. It does not claim that similarly named
states are perfectly equivalent.

```{r define-state-harmonization}
harmonize_epithelial_state <- function(cell_state, cell_state_full) {
  label <- paste(
    tolower(as.character(cell_state)),
    tolower(as.character(cell_state_full))
  )

  dplyr::case_when(
    grepl("unresolved|low-complexity|low information|low-information", label) ~
      "Unresolved / low-information",
    grepl("cycling isc|early ta|transit-ampl", label) ~ "Cycling ISC/TA",
    grepl("isc-like", label) ~ "ISC-like crypt",
    grepl("crypt progenitor", label) ~ "Crypt progenitor",
    grepl("reg4|deep crypt secretory|dcs-like|dcs cell|dcs$", label) ~
      "Deep crypt secretory",
    grepl("cycling secretory|secretory progenitor", label) ~
      "Secretory progenitor",
    grepl("immature absorptive", label) ~ "Immature absorptive",
    grepl(
      "mature absorptive|mature colonocyte|surface colonocyte|car4|aqp8|slc40a1|hmgcs2[+] absorptive",
      label
    ) ~ "Mature absorptive / colonocyte",
    grepl("immature goblet", label) ~ "Immature goblet",
    grepl("mature goblet|goblet cell", label) ~ "Mature goblet",
    grepl("tuft", label) ~ "Tuft",
    grepl("enterochromaffin|tph1", label) ~ "Enterochromaffin",
    grepl("l cell|l-cell|enteroendocrine|eec", label) ~ "Other EEC",
    TRUE ~ "Other epithelial"
  )
}

STATE_ORDER <- c(
  "ISC-like crypt",
  "Cycling ISC/TA",
  "Crypt progenitor",
  "Secretory progenitor",
  "Deep crypt secretory",
  "Immature absorptive",
  "Mature absorptive / colonocyte",
  "Immature goblet",
  "Mature goblet",
  "Tuft",
  "Enterochromaffin",
  "Other EEC",
  "Other epithelial",
  "Unresolved / low-information"
)

PRIMARY_CRYPT_STATES <- c(
  "ISC-like crypt",
  "Cycling ISC/TA",
  "Crypt progenitor",
  "Deep crypt secretory"
)

PRESENTATION_EPITHELIAL_STATES <- setdiff(
  STATE_ORDER,
  c(
    "Other epithelial",
    "Unresolved / low-information"
  )
)
```

# 4. Load and score the three cohorts

The WT feature set defines the common signature genes. The PPAR and IL17 RNA
assays must contain those same genes, ensuring that cohort scores use identical
signature membership.

```{r define-cohort-loader}
resolve_analysis_unit <- function(metadata) {
  if ("annotation_unit" %in% colnames(metadata)) {
    return(as.character(metadata$annotation_unit))
  }

  if ("wnn_cluster" %in% colnames(metadata)) {
    return(as.character(metadata$wnn_cluster))
  }

  stop(
    "Metadata contains neither annotation_unit nor wnn_cluster.",
    call. = FALSE
  )
}

extract_joined_rna_data <- function(object, metadata) {
  if (!"RNA" %in% names(object)) {
    stop("The object has no RNA assay.", call. = FALSE)
  }

  rna_layers <- SeuratObject::Layers(object[["RNA"]])

  if (!"data" %in% rna_layers) {
    stop(
      "The RNA assay must contain a joined data layer. Available layers: ",
      paste(rna_layers, collapse = ", "),
      call. = FALSE
    )
  }

  rna_data <- SeuratObject::LayerData(
    object[["RNA"]],
    layer = "data"
  )

  if (!setequal(colnames(rna_data), rownames(metadata))) {
    stop("RNA data and metadata do not contain the same cells.", call. = FALSE)
  }

  if (!identical(colnames(rna_data), rownames(metadata))) {
    rna_data <- rna_data[, rownames(metadata), drop = FALSE]
  }

  rna_data
}

score_one_cohort <- function(cohort_id, signatures_to_use = NULL) {
  cohort_id_value <- as.character(cohort_id)[[1]]
  cohort_label_value <- unname(COHORT_LABELS[[cohort_id_value]])

  message("Loading and scoring ", cohort_label_value, " ...")

  object <- readRDS(STEP6_OBJECT_PATHS[[cohort_id]])
  metadata <- object[[]]

  required_metadata <- c(
    "sample_id", "diet", "broad_compartment", "cell_state",
    "cell_state_full", "annotation_confidence", "nCount_RNA"
  )

  assert_columns(
    metadata,
    required_metadata,
    paste0(COHORT_LABELS[[cohort_id]], " metadata")
  )

  observed_diets <- unique(as.character(metadata$diet))
  if (!setequal(observed_diets, c(REFERENCE_DIET, COMPARISON_DIET))) {
    stop(
      COHORT_LABELS[[cohort_id]],
      " must contain exactly ",
      REFERENCE_DIET,
      " and ",
      COMPARISON_DIET,
      "; observed: ",
      paste(sort(observed_diets), collapse = ", "),
      call. = FALSE
    )
  }

  metadata$analysis_unit <- resolve_analysis_unit(metadata)
  metadata$cell_barcode <- rownames(metadata)

  rna_data <- extract_joined_rna_data(object, metadata)
  available_genes <- rownames(rna_data)

  if (is.null(signatures_to_use)) {
    signatures_to_use <- lapply(
      requested_signatures,
      intersect,
      y = available_genes
    )
  } else {
    missing_used_genes <- unique(
      unlist(signatures_to_use, use.names = FALSE)
    )
    missing_used_genes <- setdiff(missing_used_genes, available_genes)

    if (length(missing_used_genes) > 0L) {
      stop(
        COHORT_LABELS[[cohort_id]],
        " is missing WT-defined signature gene(s): ",
        paste(missing_used_genes, collapse = ", "),
        call. = FALSE
      )
    }
  }

  too_small <- names(signatures_to_use)[lengths(signatures_to_use) < 2L]
  if (length(too_small) > 0L) {
    stop(
      "Fewer than two evaluable genes remain for: ",
      paste(too_small, collapse = ", "),
      call. = FALSE
    )
  }

  ucell_scores <- UCell::ScoreSignatures_UCell(
    rna_data,
    features = signatures_to_use,
    maxRank = UCELL_MAX_RANK
  ) |>
    as.data.frame() |>
    tibble::rownames_to_column("cell_barcode")

  expected_score_columns <- paste0(names(signatures_to_use), "_UCell")
  assert_columns(
    ucell_scores,
    c("cell_barcode", expected_score_columns),
    paste0(COHORT_LABELS[[cohort_id]], " UCell scores")
  )

  metadata_key <- metadata |>
    tibble::as_tibble(rownames = NULL) |>
    dplyr::transmute(
      cell_barcode,
      cohort_id = cohort_id_value,
      cohort = cohort_label_value,
      sample_id = as.character(sample_id),
      diet = as.character(diet),
      analysis_unit = as.character(analysis_unit),
      broad_compartment = tolower(as.character(broad_compartment)),
      cell_state = as.character(cell_state),
      cell_state_full = as.character(cell_state_full),
      annotation_confidence = tolower(
        as.character(annotation_confidence)
      ),
      nCount_RNA = as.numeric(nCount_RNA),
      nCount_ATAC = if ("nCount_ATAC" %in% colnames(metadata)) {
        as.numeric(metadata$nCount_ATAC)
      } else {
        NA_real_
      }
    )

  if ("Lcn2" %in% rownames(rna_data)) {
    metadata_key$Lcn2_log_normalized <- as.numeric(
      rna_data["Lcn2", metadata_key$cell_barcode]
    )
    metadata_key$Lcn2_detected <- metadata_key$Lcn2_log_normalized > 0
  } else {
    metadata_key$Lcn2_log_normalized <- NA_real_
    metadata_key$Lcn2_detected <- FALSE
  }

  reduction_names <- names(object@reductions)
  wnn_reduction <- intersect(
    c("wnn.umap", "wnn_umap", "umap.wnn"),
    reduction_names
  )

  if (length(wnn_reduction) > 0L) {
    wnn_coordinates <- Seurat::Embeddings(
      object,
      reduction = wnn_reduction[[1]]
    )

    if (!setequal(rownames(wnn_coordinates), metadata_key$cell_barcode)) {
      stop("WNN coordinates and metadata have different cells.", call. = FALSE)
    }

    wnn_coordinates <- wnn_coordinates[
      metadata_key$cell_barcode,
      1:2,
      drop = FALSE
    ]

    colnames(wnn_coordinates) <- c("wnn_1", "wnn_2")

    metadata_key <- dplyr::bind_cols(
      metadata_key,
      as.data.frame(wnn_coordinates)
    )
  } else {
    warning(
      "No WNN UMAP was found for ",
      COHORT_LABELS[[cohort_id]],
      "; WNN score maps will be omitted for this cohort."
    )
    metadata_key$wnn_1 <- NA_real_
    metadata_key$wnn_2 <- NA_real_
  }

  output <- metadata_key |>
    dplyr::left_join(ucell_scores, by = "cell_barcode") |>
    dplyr::mutate(
      harmonized_state = harmonize_epithelial_state(
        cell_state,
        cell_state_full
      )
    )

  coverage <- dplyr::bind_rows(
    lapply(
      names(requested_signatures),
      function(program_name) {
        requested <- requested_signatures[[program_name]]
        used <- signatures_to_use[[program_name]]
        tibble::tibble(
          cohort_id = cohort_id,
          cohort = unname(COHORT_LABELS[[cohort_id]]),
          program = program_name,
          requested_genes = length(requested),
          genes_present_in_assay = sum(requested %in% available_genes),
          genes_used_for_common_score = length(used),
          missing_requested_genes = paste(
            setdiff(requested, available_genes),
            collapse = ";"
          )
        )
      }
    )
  )

  rm(object, metadata, metadata_key, rna_data, ucell_scores)
  invisible(gc())

  list(
    cells = output,
    signatures = signatures_to_use,
    coverage = coverage
  )
}
```

```{r score-cohorts, cache=FALSE}
wt_scored <- score_one_cohort("wt")
common_signatures <- wt_scored$signatures

ppar_scored <- score_one_cohort(
  "ppar",
  signatures_to_use = common_signatures
)

il17_scored <- score_one_cohort(
  "il17",
  signatures_to_use = common_signatures
)

cell_scores <- dplyr::bind_rows(
  wt_scored$cells,
  ppar_scored$cells,
  il17_scored$cells
) |>
  dplyr::mutate(
    cohort = factor(
      cohort,
      levels = unname(COHORT_LABELS[COHORT_IDS])
    ),
    diet = factor(diet, levels = c(REFERENCE_DIET, COMPARISON_DIET)),
    harmonized_state = factor(harmonized_state, levels = STATE_ORDER)
  )

program_gene_coverage <- dplyr::bind_rows(
  wt_scored$coverage,
  ppar_scored$coverage,
  il17_scored$coverage
)

input_inventory <- tibble::tibble(
  input = c(
    paste0(COHORT_IDS, " Step 6 object"),
    "WT Step 7 composition effect table",
    "WT Step 7c RNA priority table",
    "WT Step 7c RNA/ATAC priority table",
    "WT Step 8 persistent evidence table"
  ),
  path = c(
    unname(STEP6_OBJECT_PATHS[COHORT_IDS]),
    WT_COMPOSITION_EFFECT_PATH,
    WT_RNA_PRIORITY_PATH,
    WT_MULTIMODAL_PRIORITY_PATH,
    WT_PERSISTENT_EVIDENCE_PATH
  )
) |>
  dplyr::mutate(
    exists = file.exists(path),
    size_bytes = file.info(path)$size,
    modified = as.character(file.info(path)$mtime)
  )

readr::write_csv(
  input_inventory,
  file.path(MEETING_OUTPUT_DIR, "meeting_figure_input_inventory.csv")
)

common_program_definition <- dplyr::bind_rows(
  lapply(
    names(common_signatures),
    function(program_name) {
      tibble::tibble(
        program = program_name,
        gene = common_signatures[[program_name]]
      )
    }
  )
) |>
  dplyr::left_join(program_key, by = "program")

readr::write_csv(
  program_gene_coverage,
  file.path(TABLE_DIR, "program_gene_coverage_by_cohort.csv")
)

readr::write_csv(
  common_program_definition,
  file.path(TABLE_DIR, "program_definition_used_for_scoring.csv")
)

state_harmonization_audit <- cell_scores |>
  dplyr::count(
    cohort_id,
    cohort,
    broad_compartment,
    analysis_unit,
    cell_state,
    cell_state_full,
    annotation_confidence,
    harmonized_state,
    diet,
    name = "nuclei"
  )

readr::write_csv(
  state_harmonization_audit,
  file.path(TABLE_DIR, "cell_state_harmonization_audit.csv")
)

rm(wt_scored, ppar_scored, il17_scored)
invisible(gc())

program_gene_coverage
```

# 5. Construct signed program contrasts and WT anchors

The raw signed contrasts subtract the down-associated arm from the up-associated
arm. WT anchoring is performed only in the harmonized ISC-like crypt state:

\[
\mathrm{WT\ anchored\ score} =
\frac{x - \mathrm{median}(WT\ CON)}
{\mathrm{median}(WT\ HFD) - \mathrm{median}(WT\ CON)}.
\]

Thus, the WT CON median is 0 and the WT HFD median is 1. Values outside 0--1
are valid. Anchoring is a descriptive coordinate transformation, not a
classifier and not an estimate of genotype rescue.

```{r create-program-contrasts}
cell_scores <- cell_scores |>
  dplyr::mutate(
    Persistent_stringent_raw =
      Persistent_up_stringent_UCell -
      Persistent_down_stringent_UCell,
    Persistent_extended_raw =
      Persistent_up_extended_UCell -
      Persistent_down_extended_UCell,
    HyperMet_minus_HypoImmune_raw =
      HyperMet_UCell - HypoImmune_UCell
  )

axis_specification <- tibble::tribble(
  ~raw_score, ~axis_score, ~axis_label,
  "Persistent_stringent_raw", "Persistent_stringent_WT_axis",
  "Stringent WT-recovered persistent axis",
  "Persistent_extended_raw", "Persistent_extended_WT_axis",
  "Extended WT-recovered persistent axis",
  "HyperMet_minus_HypoImmune_raw", "HyperMet_minus_HypoImmune_WT_axis",
  "Metabolic/redox minus immune/ER-secretory contrast"
)

axis_anchor_parameters <- dplyr::bind_rows(
  lapply(
    seq_len(nrow(axis_specification)),
    function(axis_index) {
      raw_score <- axis_specification$raw_score[[axis_index]]

      anchor_cells <- cell_scores |>
        dplyr::filter(
          cohort_id == "wt",
          harmonized_state == "ISC-like crypt",
          broad_compartment == "epithelial",
          annotation_confidence %in% c("high", "moderate")
        )

      con_anchor <- safe_median(
        anchor_cells[[raw_score]][
          anchor_cells$diet == REFERENCE_DIET
        ]
      )

      hfd_anchor <- safe_median(
        anchor_cells[[raw_score]][
          anchor_cells$diet == COMPARISON_DIET
        ]
      )

      anchor_delta <- hfd_anchor - con_anchor

      if (!is.finite(anchor_delta) || abs(anchor_delta) < 1e-8) {
        stop(
          "WT ISC-like anchors do not separate for ",
          raw_score,
          "; do not construct this axis.",
          call. = FALSE
        )
      }

      tibble::tibble(
        raw_score = raw_score,
        axis_score = axis_specification$axis_score[[axis_index]],
        axis_label = axis_specification$axis_label[[axis_index]],
        WT_CON_median = con_anchor,
        WT_HFD_median = hfd_anchor,
        WT_HFD_minus_CON = anchor_delta,
        WT_HFD_is_higher = anchor_delta > 0
      )
    }
  )
)

for (axis_index in seq_len(nrow(axis_anchor_parameters))) {
  raw_score <- axis_anchor_parameters$raw_score[[axis_index]]
  axis_score <- axis_anchor_parameters$axis_score[[axis_index]]
  con_anchor <- axis_anchor_parameters$WT_CON_median[[axis_index]]
  anchor_delta <- axis_anchor_parameters$WT_HFD_minus_CON[[axis_index]]

  cell_scores[[axis_score]] <-
    (cell_scores[[raw_score]] - con_anchor) / anchor_delta
}

readr::write_csv(
  axis_anchor_parameters,
  file.path(TABLE_DIR, "WT_ISC_axis_anchor_parameters.csv")
)

axis_anchor_parameters
```

# 6. Define the curated population and summarize program scores

```{r summarize-program-scores}
curated_epithelial <- cell_scores |>
  dplyr::filter(
    broad_compartment == "epithelial",
    annotation_confidence %in% c("high", "moderate"),
    as.character(harmonized_state) %in% PRESENTATION_EPITHELIAL_STATES
  )

lcn2_detection_audit <- curated_epithelial |>
  dplyr::group_by(cohort_id, cohort, diet) |>
  dplyr::summarise(
    epithelial_nuclei = dplyr::n(),
    Lcn2_positive_nuclei = sum(Lcn2_detected, na.rm = TRUE),
    Lcn2_positive_percent = 100 * mean(Lcn2_detected, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_csv(
  lcn2_detection_audit,
  file.path(TABLE_DIR, "epithelial_Lcn2_detection_audit.csv")
)

PLOTTED_SCORE_KEY <- tibble::tribble(
  ~score_column, ~program_display, ~expected_sign,
  "HyperMet_UCell",
  "Metabolic/redox module\n(HyperMet)", 1,
  "HypoImmune_UCell",
  "Immune/ER-secretory module\n(HypoImmune)", -1,
  "Persistent_extended_raw",
  "WT-recovered persistent\nexpression contrast", 1
)

program_scores_long <- curated_epithelial |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(PLOTTED_SCORE_KEY$score_column),
    names_to = "score_column",
    values_to = "score"
  ) |>
  dplyr::left_join(PLOTTED_SCORE_KEY, by = "score_column")

program_score_summary <- program_scores_long |>
  dplyr::group_by(
    cohort_id,
    cohort,
    diet,
    harmonized_state,
    score_column,
    program_display,
    expected_sign
  ) |>
  dplyr::summarise(
    nuclei = dplyr::n(),
    median_score = stats::median(score, na.rm = TRUE),
    q25_score = stats::quantile(score, 0.25, na.rm = TRUE),
    q75_score = stats::quantile(score, 0.75, na.rm = TRUE),
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

readr::write_csv(
  program_score_summary,
  file.path(TABLE_DIR, "program_score_summary_by_cohort_state_and_diet.csv")
)
```

## Balanced cell-sampling stability

```{r calculate-balanced-score-effects}
balanced_score_effects <- function(score_data) {
  grouping_columns <- c(
    "cohort_id", "cohort", "harmonized_state",
    "score_column", "program_display", "expected_sign"
  )

  grouped_data <- score_data |>
    dplyr::group_by(
      dplyr::across(dplyr::all_of(grouping_columns))
    ) |>
    dplyr::group_split(.keep = TRUE)

  dplyr::bind_rows(
    lapply(
      seq_along(grouped_data),
      function(group_index) {
        group_data <- grouped_data[[group_index]]
        con_scores <- group_data$score[
          group_data$diet == REFERENCE_DIET
        ]
        hfd_scores <- group_data$score[
          group_data$diet == COMPARISON_DIET
        ]

        con_scores <- con_scores[is.finite(con_scores)]
        hfd_scores <- hfd_scores[is.finite(hfd_scores)]

        con_n <- length(con_scores)
        hfd_n <- length(hfd_scores)
        sampled_n <- min(
          con_n,
          hfd_n,
          MAX_DOWNSAMPLED_NUCLEI_PER_DIET
        )

        eligible <- sampled_n >= MIN_NUCLEI_PER_DIET
        full_delta <- safe_median(hfd_scores) - safe_median(con_scores)

        if (eligible) {
          set.seed(RANDOM_SEED + group_index)

          sampled_deltas <- replicate(
            DOWNSAMPLING_ITERATIONS,
            {
              safe_median(sample(hfd_scores, sampled_n)) -
                safe_median(sample(con_scores, sampled_n))
            }
          )

          downsampled_median <- safe_median(sampled_deltas)
          downsampled_q25 <- stats::quantile(sampled_deltas, 0.25)
          downsampled_q75 <- stats::quantile(sampled_deltas, 0.75)
          sign_concordance <- 100 * mean(
            sign(sampled_deltas) == sign(full_delta)
          )
        } else {
          downsampled_median <- NA_real_
          downsampled_q25 <- NA_real_
          downsampled_q75 <- NA_real_
          sign_concordance <- NA_real_
        }

        tibble::tibble(
          cohort_id = group_data$cohort_id[[1]],
          cohort = group_data$cohort[[1]],
          harmonized_state = group_data$harmonized_state[[1]],
          score_column = group_data$score_column[[1]],
          program_display = group_data$program_display[[1]],
          expected_sign = group_data$expected_sign[[1]],
          CON_nuclei = con_n,
          HFD_nuclei = hfd_n,
          sampled_nuclei_per_diet = sampled_n,
          eligible = eligible,
          full_HFD_minus_CON = full_delta,
          balanced_HFD_minus_CON_median = downsampled_median,
          balanced_HFD_minus_CON_q25 = downsampled_q25,
          balanced_HFD_minus_CON_q75 = downsampled_q75,
          full_effect_expected_direction =
            full_delta * group_data$expected_sign[[1]],
          balanced_effect_expected_direction =
            downsampled_median * group_data$expected_sign[[1]],
          balanced_expected_direction_q25 = if (
            group_data$expected_sign[[1]] > 0
          ) {
            downsampled_q25
          } else {
            -downsampled_q75
          },
          balanced_expected_direction_q75 = if (
            group_data$expected_sign[[1]] > 0
          ) {
            downsampled_q75
          } else {
            -downsampled_q25
          },
          sign_concordance_percent = sign_concordance
        )
      }
    )
  )
}

program_score_effects <- balanced_score_effects(program_scores_long)

readr::write_csv(
  program_score_effects,
  file.path(TABLE_DIR, "program_score_balanced_cell_sampling_effects.csv")
)

program_score_effects |>
  dplyr::filter(
    harmonized_state == "ISC-like crypt",
    eligible
  )
```

# 7. Figure 1A: audited WT cell-state composition effects

This first composition panel preserves the denominator and eligibility calls
from Step 7 exactly. It is the cleanest opening figure for the WT diet effect.

```{r figure-01a-wt-composition}
wt_step7_composition <- suppressMessages(
  readr::read_csv(
    WT_COMPOSITION_EFFECT_PATH,
    show_col_types = FALSE
  )
)

assert_columns(
  wt_step7_composition,
  c(
    "analysis_unit", "cell_state", "CON", "HFD",
    "difference_percentage_points", "abundance_ratio",
    "display_label", "eligibility_tier"
  ),
  "WT Step 7 composition effects"
)

wt_primary_composition <- wt_step7_composition |>
  dplyr::filter(eligibility_tier %in% c("Tier_A", "Tier_B", "Tier_C")) |>
  dplyr::arrange(difference_percentage_points) |>
  dplyr::mutate(
    display_label = factor(display_label, levels = display_label),
    direction = dplyr::if_else(
      difference_percentage_points > 0,
      "HFD_higher",
      "CON_higher"
    )
  )

readr::write_csv(
  wt_step7_composition,
  file.path(TABLE_DIR, "WT_Step7_composition_effect_sizes.csv")
)

p_wt_composition <- ggplot2::ggplot(
  wt_primary_composition,
  ggplot2::aes(
    x = difference_percentage_points,
    y = display_label,
    color = direction
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey55",
    linewidth = 0.6
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = difference_percentage_points,
      yend = display_label
    ),
    linewidth = 1.0
  ) +
  ggplot2::geom_point(size = 3.8) +
  ggplot2::scale_color_manual(
    values = DIRECTION_COLORS,
    labels = c(CON_higher = "CON enriched", HFD_higher = "HFD enriched")
  ) +
  ggplot2::labs(
    title = "WT HFD shifts the balance of annotated cell states",
    subtitle = paste0(
      "Step 7 primary analysis units; positive values indicate a larger ",
      "fraction of all retained nuclei in HFD"
    ),
    x = "HFD minus CON composition (percentage points)",
    y = NULL,
    color = "Direction",
    caption = paste0(
      "One pooled library per diet. Composition changes are descriptive and ",
      "may include biological and pool-specific effects."
    )
  ) +
  theme_presentation(base_size = 16) +
  ggplot2::theme(legend.position = "bottom")

preview_plot(p_wt_composition)
save_presentation_pdf(
  "01A_WT_cell_state_composition_HFD_minus_CON.pdf",
  p_wt_composition
)
```

# 8. Figure 1B: harmonized epithelial composition across cohorts

```{r figure-01-composition}
epithelial_composition <- curated_epithelial |>
  dplyr::count(
    cohort_id,
    cohort,
    diet,
    harmonized_state,
    name = "nuclei"
  ) |>
  dplyr::group_by(cohort_id, cohort, diet) |>
  dplyr::mutate(
    curated_epithelial_nuclei = sum(nuclei),
    percent = 100 * nuclei / curated_epithelial_nuclei
  ) |>
  dplyr::ungroup()

epithelial_composition_effects <- epithelial_composition |>
  dplyr::select(
    cohort_id,
    cohort,
    harmonized_state,
    diet,
    nuclei,
    curated_epithelial_nuclei,
    percent
  ) |>
  tidyr::pivot_wider(
    names_from = diet,
    values_from = c(nuclei, curated_epithelial_nuclei, percent),
    values_fill = 0
  ) |>
  dplyr::mutate(
    HFD_minus_CON_percentage_points =
      percent_HFD - percent_CON,
    HFD_to_CON_abundance_ratio =
      (percent_HFD + 0.01) / (percent_CON + 0.01)
  )

readr::write_csv(
  epithelial_composition,
  file.path(TABLE_DIR, "epithelial_composition_by_cohort_and_diet.csv")
)

readr::write_csv(
  epithelial_composition_effects,
  file.path(TABLE_DIR, "epithelial_composition_HFD_minus_CON.csv")
)

composition_plot_data <- epithelial_composition_effects |>
  dplyr::filter(
    as.character(harmonized_state) %in% PRESENTATION_EPITHELIAL_STATES
  ) |>
  dplyr::mutate(
    harmonized_state = factor(
      harmonized_state,
      levels = rev(PRESENTATION_EPITHELIAL_STATES)
    )
  )

p_composition <- ggplot2::ggplot(
  composition_plot_data,
  ggplot2::aes(
    x = HFD_minus_CON_percentage_points,
    y = harmonized_state,
    color = cohort
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey55",
    linewidth = 0.6
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = HFD_minus_CON_percentage_points,
      yend = harmonized_state
    ),
    linewidth = 1.0,
    alpha = 0.75
  ) +
  ggplot2::geom_point(size = 3.4) +
  ggplot2::facet_wrap(~ cohort, nrow = 1) +
  ggplot2::scale_color_manual(values = COHORT_COLORS, guide = "none") +
  ggplot2::labs(
    title = "HFD-associated epithelial composition shifts",
    subtitle = paste0(
      "HFD minus CON within curated high/moderate-confidence epithelial nuclei; ",
      "positive values indicate HFD enrichment"
    ),
    x = "Difference in epithelial composition (percentage points)",
    y = NULL,
    caption = paste0(
      "Descriptive pooled-library comparison: one CON and one HFD library per cohort. ",
      "Shared state labels are a harmonized display layer."
    )
  ) +
  theme_presentation(base_size = 16)

preview_plot(p_composition)
save_presentation_pdf(
  "01B_cross_cohort_epithelial_composition_HFD_minus_CON.pdf",
  p_composition
)
```

# 9. Figure 2: strongest stable WT RNA effects in key epithelial states

```{r prepare-wt-rna-candidates}
wt_rna_prioritized <- suppressMessages(
  readr::read_csv(WT_RNA_PRIORITY_PATH, show_col_types = FALSE)
)

assert_columns(
  wt_rna_prioritized,
  c(
    "analysis_unit", "feature", "log2FC_comparison_vs_reference",
    "detection_difference_percentage_points", "cell_state",
    "cell_state_full", "direction", "RNA_support_class",
    "support_order", "within_unit_direction_rank"
  ),
  "WT prioritized RNA candidates"
)

wt_rna_prioritized <- wt_rna_prioritized |>
  dplyr::mutate(
    harmonized_state = harmonize_epithelial_state(
      cell_state,
      cell_state_full
    ),
    RNA_direction_stable = if (
      "RNA_direction_stable" %in% colnames(wt_rna_prioritized)
    ) {
      as_logical_safely(RNA_direction_stable)
    } else {
      support_order <= 2L
    },
    RNA_downsampling_iqr_excludes_zero = if (
      "RNA_downsampling_iqr_excludes_zero" %in%
        colnames(wt_rna_prioritized)
    ) {
      as_logical_safely(RNA_downsampling_iqr_excludes_zero)
    } else {
      support_order == 1L
    }
  )

WT_RNA_FOCUS_STATES <- c(
  PRIMARY_CRYPT_STATES,
  "Mature absorptive / colonocyte",
  "Mature goblet"
)

wt_rna_meeting_candidates <- wt_rna_prioritized |>
  dplyr::filter(
    harmonized_state %in% WT_RNA_FOCUS_STATES,
    RNA_direction_stable,
    support_order <= 2L
  ) |>
  dplyr::group_by(harmonized_state, direction) |>
  dplyr::arrange(
    support_order,
    within_unit_direction_rank,
    dplyr::desc(abs(log2FC_comparison_vs_reference)),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = TOP_RNA_GENES_PER_STATE_AND_DIRECTION) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    harmonized_state = factor(
      harmonized_state,
      levels = WT_RNA_FOCUS_STATES
    ),
    gene_label = feature
  )

if (nrow(wt_rna_meeting_candidates) == 0L) {
  stop("No stable WT RNA meeting candidates were retained.", call. = FALSE)
}

readr::write_csv(
  wt_rna_meeting_candidates,
  file.path(TABLE_DIR, "WT_RNA_meeting_candidates.csv")
)
```

```{r figure-02-wt-rna-effects}
p_wt_rna <- ggplot2::ggplot(
  wt_rna_meeting_candidates,
  ggplot2::aes(
    x = log2FC_comparison_vs_reference,
    y = stats::reorder(gene_label, log2FC_comparison_vs_reference),
    color = direction
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey55",
    linewidth = 0.5
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = 0,
      xend = log2FC_comparison_vs_reference,
      yend = stats::reorder(
        gene_label,
        log2FC_comparison_vs_reference
      )
    ),
    linewidth = 0.9
  ) +
  ggplot2::geom_point(
    ggplot2::aes(
      size = abs(detection_difference_percentage_points),
      shape = RNA_downsampling_iqr_excludes_zero
    ),
    stroke = 1.0
  ) +
  ggplot2::facet_wrap(
    ~ harmonized_state,
    scales = "free_y",
    ncol = 3
  ) +
  ggplot2::scale_color_manual(
    values = DIRECTION_COLORS,
    labels = c(CON_higher = "CON higher", HFD_higher = "HFD higher")
  ) +
  ggplot2::scale_shape_manual(
    values = c(`FALSE` = 1, `TRUE` = 16),
    labels = c(`FALSE` = "Stable sign", `TRUE` = "IQR excludes zero")
  ) +
  ggplot2::scale_size_continuous(range = c(2.5, 6.5)) +
  ggplot2::labs(
    title = "Strongest cell-sampling-stable WT RNA diet effects",
    subtitle = paste0(
      "Top candidates per state and direction; point size is the absolute ",
      "difference in detection frequency"
    ),
    x = "RNA log2 fold-change (HFD versus CON)",
    y = NULL,
    color = "Direction",
    shape = "Balanced sampling",
    size = "|Δ detected|\n(percentage points)",
    caption = paste0(
      "Descriptive pooled-library effects. Balanced cell sampling tests ",
      "robustness to unequal nucleus counts, not biological replication."
    )
  ) +
  theme_presentation(base_size = 14) +
  ggplot2::theme(legend.position = "bottom")

preview_plot(p_wt_rna)
save_presentation_pdf("02_WT_stable_RNA_diet_effects.pdf", p_wt_rna)
```

# 10. Figure 3: WT RNA/ATAC concordance near the assigned gene

The distance filter below uses Step 7c nearest-gene distance. It is not an exact
TSS window and must not be called promoter evidence. Exact +/-2 kb TSS evidence
for the persistent-gene set is shown in Figure 5.

```{r prepare-wt-multimodal-candidates}
wt_multimodal_prioritized <- suppressMessages(
  readr::read_csv(
    WT_MULTIMODAL_PRIORITY_PATH,
    show_col_types = FALSE
  )
)

assert_columns(
  wt_multimodal_prioritized,
  c(
    "analysis_unit", "peak", "nearest_gene", "distance",
    "ATAC_log2FC", "RNA_log2FC", "cell_state", "cell_state_full",
    "effect_direction", "multimodal_support_class", "support_order",
    "within_unit_direction_rank", "joint_effect_magnitude"
  ),
  "WT prioritized RNA/ATAC candidates"
)

wt_multimodal_meeting_candidates <- wt_multimodal_prioritized |>
  dplyr::mutate(
    harmonized_state = harmonize_epithelial_state(
      cell_state,
      cell_state_full
    ),
    nearby_gene_context = dplyr::case_when(
      distance == 0 ~ "Overlaps gene annotation",
      is.finite(distance) & abs(distance) <= NEARBY_GENE_DISTANCE_BP ~
        "Within 2 kb of assigned gene",
      TRUE ~ "More than 2 kb away"
    )
  ) |>
  dplyr::filter(
    harmonized_state %in% WT_RNA_FOCUS_STATES,
    support_order <= 2L,
    is.finite(distance),
    abs(distance) <= NEARBY_GENE_DISTANCE_BP
  ) |>
  dplyr::group_by(harmonized_state, effect_direction) |>
  dplyr::arrange(
    support_order,
    within_unit_direction_rank,
    dplyr::desc(joint_effect_magnitude),
    .by_group = TRUE
  ) |>
  dplyr::slice_head(n = TOP_MULTIMODAL_PAIRS_PER_STATE_AND_DIRECTION) |>
  dplyr::ungroup()

if (nrow(wt_multimodal_meeting_candidates) == 0L) {
  warning(
    "No Tier 1/2 WT concordant pairs were found within the 2-kb ",
    "nearest-gene filter; Figure 3 will be omitted."
  )
} else {
  readr::write_csv(
    wt_multimodal_meeting_candidates,
    file.path(TABLE_DIR, "WT_nearby_RNA_ATAC_meeting_candidates.csv")
  )
}
```

```{r figure-03-wt-multimodal-effects}
if (nrow(wt_multimodal_meeting_candidates) > 0L) {
  label_candidates <- wt_multimodal_meeting_candidates |>
    dplyr::arrange(support_order, dplyr::desc(joint_effect_magnitude)) |>
    dplyr::distinct(nearest_gene, .keep_all = TRUE)

  p_wt_multimodal <- ggplot2::ggplot(
    wt_multimodal_meeting_candidates,
    ggplot2::aes(
      y = RNA_log2FC,
      x = ATAC_log2FC,
      color = harmonized_state,
      shape = nearby_gene_context
    )
  ) +
    ggplot2::geom_hline(
      yintercept = 0,
      color = "grey65",
      linewidth = 0.5
    ) +
    ggplot2::geom_vline(
      xintercept = 0,
      color = "grey65",
      linewidth = 0.5
    ) +
    ggplot2::geom_abline(
      slope = 1,
      intercept = 0,
      linetype = "dotted",
      color = "grey55"
    ) +
    ggplot2::geom_point(size = 4.2, alpha = 0.85, stroke = 1.1) +
    ggplot2::geom_text(
      data = label_candidates,
      ggplot2::aes(label = nearest_gene),
      color = "black",
      size = 4.0,
      check_overlap = TRUE,
      nudge_y = 0.10,
      show.legend = FALSE
    ) +
    ggplot2::scale_color_manual(values = STATE_COLORS) +
    ggplot2::scale_shape_manual(
      values = c(
        "Overlaps gene annotation" = 16,
        "Within 2 kb of assigned gene" = 17
      )
    ) +
    ggplot2::labs(
      title = "WT genes with directionally concordant nearby accessibility",
      subtitle = paste0(
        "Tier 1/2 RNA-ATAC pairs assigned within 2 kb of the nearest gene; ",
        "upper-right is HFD-higher in both modalities"
      ),
      y = "RNA log2 fold-change (HFD versus CON)",
      x = "ATAC log2 fold-change (HFD versus CON)",
      color = "Harmonized state",
      shape = "Genomic context",
      caption = paste0(
        "Nearest-gene assignment is contextual and does not establish a ",
        "peak-to-gene regulatory relationship."
      )
    ) +
    theme_presentation(base_size = 15) +
    ggplot2::theme(legend.position = "bottom")

  preview_plot(p_wt_multimodal)
  save_presentation_pdf(
    "03_WT_nearby_RNA_ATAC_concordance.pdf",
    p_wt_multimodal
  )
}
```

# 11. Figure 4: how much of the persistent program reappears in WT

```{r figure-04-persistent-summary}
persistent_class_summary <- persistent_evidence |>
  dplyr::mutate(
    validation_group = dplyr::case_when(
      RNA_reference_direction_concordant &
        RNA_clear_effect &
        RNA_direction_stable ~ "Clear, stable concordance",
      RNA_reference_direction_concordant &
        RNA_direction_stable ~ "Stable, subthreshold concordance",
      RNA_reference_direction_concordant ~ "Other directional concordance",
      !RNA_reference_direction_concordant &
        RNA_clear_effect &
        RNA_direction_stable ~ "Clear, stable discordance",
      !RNA_reference_direction_concordant ~ "Other discordance",
      TRUE ~ "Not evaluable"
    ),
    reference_program = dplyr::recode(
      reference_program,
      persistent_HFD_up = "Reference HFD-up",
      persistent_HFD_down = "Reference HFD-down"
    )
  ) |>
  dplyr::count(reference_program, validation_group, name = "genes") |>
  dplyr::group_by(reference_program) |>
  dplyr::mutate(percent = 100 * genes / sum(genes)) |>
  dplyr::ungroup()

readr::write_csv(
  persistent_class_summary,
  file.path(TABLE_DIR, "persistent_program_WT_validation_summary.csv")
)

PERSISTENT_CLASS_COLORS <- c(
  "Clear, stable concordance" = "#0072B2",
  "Stable, subthreshold concordance" = "#56B4E9",
  "Other directional concordance" = "#A6CEE3",
  "Clear, stable discordance" = "#D55E00",
  "Other discordance" = "#E69F00",
  "Not evaluable" = "#BDBDBD"
)

p_persistent_summary <- ggplot2::ggplot(
  persistent_class_summary,
  ggplot2::aes(
    x = reference_program,
    y = percent,
    fill = validation_group
  )
) +
  ggplot2::geom_col(width = 0.68, color = "white", linewidth = 0.4) +
  ggplot2::geom_text(
    ggplot2::aes(
      label = ifelse(percent >= 5, paste0(genes, " (", round(percent), "%)"), "")
    ),
    position = ggplot2::position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 4.4
  ) +
  ggplot2::scale_fill_manual(values = PERSISTENT_CLASS_COLORS) +
  ggplot2::scale_y_continuous(
    limits = c(0, 100),
    expand = ggplot2::expansion(mult = c(0, 0.02))
  ) +
  ggplot2::labs(
    title = "The persistent d21-to-d50 program is only partially recovered in WT",
    subtitle = paste0(
      "Primary WT ISC-like state; the HFD-up arm is more directionally ",
      "coherent than the HFD-down arm"
    ),
    x = NULL,
    y = "Genes (%)",
    fill = "WT RNA validation",
    caption = paste0(
      "Direction is compared with the supplied persistent reference list. ",
      "This is a targeted validation, not discovery of a globally preserved module."
    )
  ) +
  theme_presentation(base_size = 17) +
  ggplot2::theme(legend.position = "right")

preview_plot(p_persistent_summary)
save_presentation_pdf(
  "04_persistent_program_WT_validation_summary.pdf",
  p_persistent_summary
)
```

# 12. Figure 5: stringent persistent genes and exact TSS-window ATAC context

Positive values in this plot agree with the supplied persistent reference
direction. The ATAC layer is exact +/-2 kb TSS-window context from Step 8.

```{r figure-05-persistent-gene-evidence}
persistent_gene_meeting_table <- persistent_stringent |>
  dplyr::select(
    gene,
    reference_direction,
    reference_program,
    RNA_reference_aligned_log2FC,
    representative_promoter_peak,
    representative_peak_offset_bp,
    representative_peak_overlaps_tss,
    representative_ATAC_reference_aligned_log2FC,
    promoter_ATAC_summary_class,
    multimodal_validation_class
  ) |>
  dplyr::arrange(
    reference_direction,
    dplyr::desc(RNA_reference_aligned_log2FC)
  )

readr::write_csv(
  persistent_gene_meeting_table,
  file.path(TABLE_DIR, "persistent_stringent_gene_RNA_TSS_ATAC_evidence.csv")
)

persistent_gene_long <- persistent_gene_meeting_table |>
  dplyr::select(
    gene,
    reference_direction,
    RNA_reference_aligned_log2FC,
    representative_ATAC_reference_aligned_log2FC
  ) |>
  tidyr::pivot_longer(
    cols = c(
      RNA_reference_aligned_log2FC,
      representative_ATAC_reference_aligned_log2FC
    ),
    names_to = "modality",
    values_to = "reference_aligned_log2FC"
  ) |>
  dplyr::mutate(
    modality = dplyr::recode(
      modality,
      RNA_reference_aligned_log2FC = "RNA",
      representative_ATAC_reference_aligned_log2FC =
        "Representative exact TSS-window ATAC"
    ),
    reference_direction = dplyr::recode(
      reference_direction,
      HFD_higher = "Reference HFD-up",
      CON_higher = "Reference HFD-down"
    )
  )

gene_order <- persistent_gene_meeting_table |>
  dplyr::arrange(reference_direction, RNA_reference_aligned_log2FC) |>
  dplyr::pull(gene)

persistent_gene_long$gene <- factor(
  persistent_gene_long$gene,
  levels = unique(gene_order)
)

p_persistent_genes <- ggplot2::ggplot(
  persistent_gene_long,
  ggplot2::aes(
    x = reference_aligned_log2FC,
    y = gene,
    color = reference_direction,
    shape = modality
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey55",
    linewidth = 0.6
  ) +
  ggplot2::geom_segment(
    data = persistent_gene_meeting_table,
    ggplot2::aes(
      x = RNA_reference_aligned_log2FC,
      xend = representative_ATAC_reference_aligned_log2FC,
      y = factor(gene, levels = unique(gene_order)),
      yend = factor(gene, levels = unique(gene_order))
    ),
    inherit.aes = FALSE,
    color = "grey75",
    linewidth = 0.8,
    na.rm = TRUE
  ) +
  ggplot2::geom_point(size = 4.2, stroke = 1.1, na.rm = TRUE) +
  ggplot2::scale_color_manual(
    values = c(
      "Reference HFD-up" = DIET_COLORS[["HFD"]],
      "Reference HFD-down" = DIET_COLORS[["CON"]]
    )
  ) +
  ggplot2::scale_shape_manual(
    values = c(
      RNA = 16,
      `Representative exact TSS-window ATAC` = 17
    )
  ) +
  ggplot2::labs(
    title = "Nine stringent WT-recovered persistent genes",
    subtitle = paste0(
      "Positive values agree with the supplied persistent direction; ",
      "RNA and exact +/-2 kb TSS-window ATAC are shown separately"
    ),
    x = "Reference-aligned log2 fold-change",
    y = NULL,
    color = "Reference arm",
    shape = "Evidence layer",
    caption = paste0(
      "Exact TSS-window accessibility is regulatory context, not proof of ",
      "activation or direct peak-to-gene control."
    )
  ) +
  theme_presentation(base_size = 17) +
  ggplot2::theme(legend.position = "bottom")

preview_plot(p_persistent_genes)
save_presentation_pdf(
  "05_persistent_stringent_RNA_and_TSS_ATAC.pdf",
  p_persistent_genes
)
```

# 13. Figure 6: WT ISC-like program scores

```{r figure-06-wt-isc-scores}
wt_isc_scores <- program_scores_long |>
  dplyr::filter(
    cohort_id == "wt",
    harmonized_state == "ISC-like crypt"
  )

readr::write_csv(
  wt_isc_scores,
  file.path(TABLE_DIR, "WT_ISC_cell_level_program_scores.csv.gz")
)

p_wt_isc_scores <- ggplot2::ggplot(
  wt_isc_scores,
  ggplot2::aes(x = diet, y = score, fill = diet)
) +
  ggplot2::geom_violin(
    width = 0.82,
    scale = "width",
    trim = TRUE,
    color = NA,
    alpha = 0.60
  ) +
  ggplot2::geom_boxplot(
    width = 0.20,
    outlier.shape = NA,
    color = "black",
    fill = "white",
    linewidth = 0.7
  ) +
  ggplot2::stat_summary(
    fun = median,
    geom = "point",
    shape = 23,
    size = 3.2,
    fill = "black"
  ) +
  ggplot2::facet_wrap(
    ~ program_display,
    scales = "free_y",
    nrow = 1
  ) +
  ggplot2::scale_fill_manual(values = DIET_COLORS) +
  ggplot2::labs(
    title = "WT ISC-like cells: targeted program scores",
    subtitle = paste0(
      "UCell scores for HyperMet and HypoImmune plus the signed extended ",
      "WT-recovered persistent contrast"
    ),
    x = NULL,
    y = "Per-nucleus score",
    fill = "Diet",
    caption = paste0(
      "Distributions describe nuclei from one pooled CON and one pooled HFD ",
      "library; no cell-level P values are reported."
    )
  ) +
  theme_presentation(base_size = 16) +
  ggplot2::theme(legend.position = "bottom")

preview_plot(p_wt_isc_scores)
save_presentation_pdf("06_WT_ISC_program_scores.pdf", p_wt_isc_scores)
```

# 14. Figure 7: the same ISC-like scores across cohorts

```{r figure-07-cross-cohort-isc-scores}
cross_cohort_isc_scores <- program_scores_long |>
  dplyr::filter(harmonized_state == "ISC-like crypt")

readr::write_csv(
  cross_cohort_isc_scores,
  file.path(TABLE_DIR, "cross_cohort_ISC_cell_level_program_scores.csv.gz")
)

p_cross_cohort_isc_scores <- ggplot2::ggplot(
  cross_cohort_isc_scores,
  ggplot2::aes(
    x = diet,
    y = score,
    fill = diet
  )
) +
  ggplot2::geom_violin(
    width = 0.84,
    scale = "width",
    trim = TRUE,
    color = NA,
    alpha = 0.60
  ) +
  ggplot2::geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    color = "black",
    fill = "white",
    linewidth = 0.65
  ) +
  ggplot2::facet_grid(
    rows = ggplot2::vars(program_display),
    cols = ggplot2::vars(cohort),
    scales = "free_y"
  ) +
  ggplot2::scale_fill_manual(values = DIET_COLORS) +
  ggplot2::labs(
    title = "Matched ISC-like states scored with identical gene sets",
    subtitle = paste0(
      "The same WT-defined signatures are projected without retraining into ",
      "Ppar iKO and Il17ra iKO"
    ),
    x = NULL,
    y = "Per-nucleus score",
    fill = "Diet",
    caption = paste0(
      "Cross-cohort shifts are descriptive. Pool composition and technical ",
      "differences remain inseparable from genotype effects."
    )
  ) +
  theme_presentation(base_size = 14) +
  ggplot2::theme(legend.position = "bottom")

preview_plot(p_cross_cohort_isc_scores)
save_presentation_pdf(
  "07_cross_cohort_ISC_program_scores.pdf",
  p_cross_cohort_isc_scores
)
```

# 15. Figure 8: direction and stability of ISC-like score changes

All effects are oriented so positive means movement in the expected direction:
HyperMet higher in HFD, HypoImmune lower in HFD, or a more HFD-like persistent
contrast. Error bars are the interquartile range across balanced cell samples,
not confidence intervals.

```{r figure-08-cross-cohort-isc-effects}
isc_score_effects <- program_score_effects |>
  dplyr::filter(
    harmonized_state == "ISC-like crypt",
    eligible
  ) |>
  dplyr::mutate(
    cohort = factor(cohort, levels = unname(COHORT_LABELS[COHORT_IDS])),
    program_display = factor(
      program_display,
      levels = PLOTTED_SCORE_KEY$program_display
    )
  )

readr::write_csv(
  isc_score_effects,
  file.path(TABLE_DIR, "cross_cohort_ISC_program_effects.csv")
)

p_isc_effects <- ggplot2::ggplot(
  isc_score_effects,
  ggplot2::aes(
    x = balanced_effect_expected_direction,
    y = cohort,
    color = cohort
  )
) +
  ggplot2::geom_vline(
    xintercept = 0,
    color = "grey50",
    linewidth = 0.6
  ) +
  ggplot2::geom_errorbarh(
    ggplot2::aes(
      xmin = balanced_expected_direction_q25,
      xmax = balanced_expected_direction_q75
    ),
    height = 0.16,
    linewidth = 1.0
  ) +
  ggplot2::geom_point(size = 4.4) +
  ggplot2::facet_wrap(
    ~ program_display,
    scales = "free_x",
    nrow = 1
  ) +
  ggplot2::scale_color_manual(values = COHORT_COLORS, guide = "none") +
  ggplot2::labs(
    title = "Do matched ISC-like cells move in the expected HFD direction?",
    subtitle = paste0(
      "Point = median HFD-minus-CON effect across 100 balanced cell samples; ",
      "positive values support the expected program direction"
    ),
    x = "Expected-direction score difference",
    y = NULL,
    caption = paste0(
      "Bars are cell-sampling IQRs, not biological confidence intervals. ",
      "One pooled library represents each diet within each cohort."
    )
  ) +
  theme_presentation(base_size = 15)

preview_plot(p_isc_effects)
save_presentation_pdf(
  "08_cross_cohort_ISC_expected_direction_effects.pdf",
  p_isc_effects
)
```

# 16. Figure 9: program effects across matched epithelial states

Color is rescaled separately within each program to emphasize direction and
relative pattern. It must not be used to compare absolute effect magnitude
between programs; the raw effect table contains the exact values.

```{r figure-09-state-program-heatmap}
state_program_effects <- program_score_effects |>
  dplyr::filter(
    eligible,
    as.character(harmonized_state) %in% PRESENTATION_EPITHELIAL_STATES
  ) |>
  dplyr::group_by(score_column) |>
  dplyr::mutate(
    program_scale = max(
      abs(balanced_effect_expected_direction),
      na.rm = TRUE
    ),
    relative_expected_direction_effect = dplyr::if_else(
      is.finite(program_scale) & program_scale > 0,
      balanced_effect_expected_direction / program_scale,
      NA_real_
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    harmonized_state = factor(
      harmonized_state,
      levels = rev(PRESENTATION_EPITHELIAL_STATES)
    ),
    cohort = factor(cohort, levels = unname(COHORT_LABELS[COHORT_IDS])),
    display_value = sprintf("%+.3f", balanced_effect_expected_direction)
  )

readr::write_csv(
  state_program_effects,
  file.path(TABLE_DIR, "cross_cohort_state_program_effects.csv")
)

p_state_program <- ggplot2::ggplot(
  state_program_effects,
  ggplot2::aes(
    x = cohort,
    y = harmonized_state,
    fill = relative_expected_direction_effect
  )
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.7) +
  ggplot2::geom_text(
    ggplot2::aes(label = display_value),
    size = 3.3,
    color = "black"
  ) +
  ggplot2::facet_wrap(
    ~ program_display,
    nrow = 1
  ) +
  ggplot2::scale_fill_gradient2(
    low = DIET_COLORS[["CON"]],
    mid = "white",
    high = DIET_COLORS[["HFD"]],
    midpoint = 0,
    limits = c(-1, 1),
    oob = scales::squish
  ) +
  ggplot2::labs(
    title = "Program direction across matched epithelial states",
    subtitle = paste0(
      "Cell labels show raw expected-direction HFD-minus-CON score differences; ",
      "tile color is scaled within each program"
    ),
    x = NULL,
    y = NULL,
    fill = "Relative effect\nwithin program",
    caption = paste0(
      "Blank state/cohort cells failed the minimum of ",
      MIN_NUCLEI_PER_DIET,
      " nuclei per diet. Color magnitude is not comparable between programs."
    )
  ) +
  theme_presentation(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 25, hjust = 1),
    legend.position = "bottom"
  )

preview_plot(p_state_program)
save_presentation_pdf(
  "09_cross_cohort_state_program_effects.pdf",
  p_state_program
)
```

# 17. Supporting figures: WT epithelial WNN program maps

These maps show where a score resides in the existing multimodal embedding.
They do not imply trajectory direction or RNA velocity.

```{r define-wnn-score-map}
plot_wnn_score <- function(data, score_column, display_label, filename) {
  plot_data <- data |>
    dplyr::filter(
      cohort_id == "wt",
      broad_compartment == "epithelial",
      annotation_confidence %in% c("high", "moderate"),
      is.finite(wnn_1),
      is.finite(wnn_2),
      is.finite(.data[[score_column]])
    )

  if (nrow(plot_data) == 0L) {
    warning("No WT WNN coordinates available for ", display_label)
    return(invisible(NULL))
  }

  score_limits <- stats::quantile(
    plot_data[[score_column]],
    probs = c(0.02, 0.98),
    na.rm = TRUE,
    names = FALSE
  )

  if (
    !all(is.finite(score_limits)) ||
      score_limits[[1]] == score_limits[[2]]
  ) {
    score_center <- stats::median(
      plot_data[[score_column]],
      na.rm = TRUE
    )

    score_padding <- max(
      abs(score_center) * 0.01,
      1e-6
    )

    score_limits <- score_center + c(
      -score_padding,
      score_padding
    )
  }

  # Draw low-scoring cells first so high-scoring cells remain visible.
  plot_data <- plot_data |>
    dplyr::arrange(.data[[score_column]])

  score_plot <- ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = wnn_1,
      y = wnn_2,
      color = .data[[score_column]]
    )
  ) +
    ggplot2::geom_point(
      size = 0.35,
      alpha = 0.90
    ) +
    ggplot2::facet_wrap(
      ~ diet,
      nrow = 1
    ) +
    ggplot2::scale_color_gradient(
      low = "lightgrey",
      high = "blue",
      limits = score_limits,
      oob = scales::squish,
      na.value = "grey90"
    ) +
    ggplot2::coord_equal() +
    ggplot2::labs(
      title = paste0(
        "WT epithelial WNN: ",
        display_label
      ),
      subtitle = paste0(
        "Same embedding and UCell score scale in CON and HFD"
      ),
      x = "wnnUMAP_1",
      y = "wnnUMAP_2",
      color = "UCell score\n(2nd–98th pct.)",
      caption = paste0(
        "The WNN embedding represents multimodal neighborhoods; ",
        "score gradients are descriptive."
      )
    ) +
    theme_presentation(base_size = 16) +
    ggplot2::theme(
      panel.grid = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      axis.text = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank()
    )

  preview_plot(score_plot)
  save_presentation_pdf(filename, score_plot)

  invisible(score_plot)
}
```

```{r figure-10-11-wnn-score-maps}
plot_wnn_score(
  cell_scores,
  "HyperMet_UCell",
  "Persistent HFD-associated metabolic/redox module (HyperMet)",
  "10_supporting_WT_epithelial_WNN_HyperMet.pdf"
)

plot_wnn_score(
  cell_scores,
  "HypoImmune_UCell",
  "Persistent HFD-depleted immune/ER-secretory module (HypoImmune)",
  "11_supporting_WT_epithelial_WNN_HypoImmune.pdf"
)
```

# 18. Meeting-ready result inventory

This section writes a compact numerical scorecard and a manifest that gives the
recommended figure order, the question answered by each PDF, and the principal
interpretive limitation.

```{r create-meeting-scorecard}
top_wt_composition_changes <- epithelial_composition_effects |>
  dplyr::filter(cohort_id == "wt") |>
  dplyr::arrange(dplyr::desc(abs(HFD_minus_CON_percentage_points))) |>
  dplyr::slice_head(n = 8L) |>
  dplyr::transmute(
    result_family = "WT epithelial composition",
    item = as.character(harmonized_state),
    estimate = HFD_minus_CON_percentage_points,
    estimate_definition = "HFD minus CON percentage points",
    direction = dplyr::if_else(estimate > 0, "HFD enriched", "CON enriched")
  )

top_wt_rna_changes <- wt_rna_meeting_candidates |>
  dplyr::arrange(dplyr::desc(abs(log2FC_comparison_vs_reference))) |>
  dplyr::slice_head(n = 12L) |>
  dplyr::transmute(
    result_family = "WT stable RNA effect",
    item = paste0(feature, " | ", harmonized_state),
    estimate = log2FC_comparison_vs_reference,
    estimate_definition = "RNA log2FC HFD versus CON",
    direction = dplyr::recode(
      direction,
      HFD_higher = "HFD higher",
      CON_higher = "CON higher"
    )
  )

persistent_scorecard <- persistent_class_summary |>
  dplyr::transmute(
    result_family = "Persistent program validation",
    item = paste0(reference_program, " | ", validation_group),
    estimate = genes,
    estimate_definition = "genes",
    direction = NA_character_
  )

isc_program_scorecard <- isc_score_effects |>
  dplyr::transmute(
    result_family = "Matched ISC-like program effect",
    item = paste0(as.character(cohort), " | ", program_display),
    estimate = balanced_effect_expected_direction,
    estimate_definition =
      "expected-direction median HFD-CON score difference",
    direction = dplyr::if_else(
      balanced_effect_expected_direction > 0,
      "expected direction",
      "opposite direction"
    )
  )

lcn2_scorecard <- lcn2_detection_audit |>
  dplyr::transmute(
    result_family = "Epithelial Lcn2 detection",
    item = paste0(as.character(cohort), " | ", as.character(diet)),
    estimate = Lcn2_positive_percent,
    estimate_definition = "percent of curated epithelial nuclei with Lcn2 detected",
    direction = NA_character_
  )

meeting_key_results <- dplyr::bind_rows(
  top_wt_composition_changes,
  top_wt_rna_changes,
  persistent_scorecard,
  isc_program_scorecard,
  lcn2_scorecard
)

readr::write_csv(
  meeting_key_results,
  file.path(TABLE_DIR, "meeting_key_results.csv")
)

figure_manifest <- tibble::tribble(
  ~order, ~priority, ~pdf, ~question, ~safe_talk_track, ~backing_table,
  1L, "core", "01A_WT_cell_state_composition_HFD_minus_CON.pdf",
  "Which annotated WT states change in relative abundance with HFD?",
  "Lead with the increase in cycling ISC/TA, mature colonocyte, and Reg4-positive DCS states, alongside depletion of ISC-like and immature absorptive states.",
  "WT_Step7_composition_effect_sizes.csv",
  2L, "core", "01B_cross_cohort_epithelial_composition_HFD_minus_CON.pdf",
  "Do harmonized epithelial composition patterns stack similarly across cohorts?",
  "Describe direction and size in WT first, then show whether Ppar iKO or Il17ra iKO displays the same pattern.",
  "epithelial_composition_HFD_minus_CON.csv",
  3L, "core", "02_WT_stable_RNA_diet_effects.pdf",
  "Which WT genes show the clearest state-matched diet effects?",
  "Emphasize effect size, detection shift, and cell-sampling stability; do not describe significance.",
  "WT_RNA_meeting_candidates.csv",
  4L, "core", "03_WT_nearby_RNA_ATAC_concordance.pdf",
  "Which WT RNA effects have concordant nearby accessibility?",
  "Call these prioritized multimodal candidates, not validated regulatory links.",
  "WT_nearby_RNA_ATAC_meeting_candidates.csv",
  5L, "core", "04_persistent_program_WT_validation_summary.pdf",
  "Does the supplied persistent d21-to-d50 program reproduce in WT?",
  "The recovery is partial and asymmetric, with better support for the HFD-up arm.",
  "persistent_program_WT_validation_summary.csv",
  6L, "core", "05_persistent_stringent_RNA_and_TSS_ATAC.pdf",
  "Which persistent genes have the strongest WT RNA evidence and exact TSS-window context?",
  "Separate RNA agreement from chromatin context and point out discordant loci explicitly.",
  "persistent_stringent_gene_RNA_TSS_ATAC_evidence.csv",
  7L, "core", "06_WT_ISC_program_scores.pdf",
  "Do the targeted modules move within WT ISC-like cells?",
  "Use raw per-nucleus score distributions descriptively; the pool is the experimental unit.",
  "WT_ISC_cell_level_program_scores.csv.gz",
  8L, "core", "07_cross_cohort_ISC_program_scores.pdf",
  "Where do Ppar iKO and Il17ra iKO sit under the same scoring definitions?",
  "The signatures are fixed from WT, but cross-cohort offsets are not pure genotype effects.",
  "cross_cohort_ISC_cell_level_program_scores.csv.gz",
  9L, "core", "08_cross_cohort_ISC_expected_direction_effects.pdf",
  "Does each cohort move in the expected program direction under HFD?",
  "Compare signs and descriptive magnitudes; the bars are cell-sampling IQRs, not confidence intervals.",
  "cross_cohort_ISC_program_effects.csv",
  10L, "supporting", "09_cross_cohort_state_program_effects.pdf",
  "Are program shifts restricted to ISC-like cells or shared across epithelial states?",
  "Use color for relative pattern within a program, and the printed values for raw effects.",
  "cross_cohort_state_program_effects.csv",
  11L, "supporting", "10_supporting_WT_epithelial_WNN_HyperMet.pdf",
  "Where is the metabolic/redox program located in WT epithelial WNN space?",
  "Treat the map as spatial context, not trajectory or RNA velocity.",
  "program_score_summary_by_cohort_state_and_diet.csv",
  12L, "supporting", "11_supporting_WT_epithelial_WNN_HypoImmune.pdf",
  "Where is the immune/ER-secretory program located in WT epithelial WNN space?",
  "Treat the map as spatial context, not trajectory or RNA velocity.",
  "program_score_summary_by_cohort_state_and_diet.csv"
)

readr::write_csv(
  figure_manifest,
  file.path(MEETING_OUTPUT_DIR, "meeting_figure_manifest.csv")
)

figure_manifest
```

# 19. Run summary and completion marker

```{r save-final-audits}
cell_score_export <- cell_scores |>
  dplyr::select(
    cell_barcode,
    cohort_id,
    cohort,
    sample_id,
    diet,
    analysis_unit,
    broad_compartment,
    cell_state,
    cell_state_full,
    annotation_confidence,
    harmonized_state,
    Lcn2_log_normalized,
    Lcn2_detected,
    dplyr::ends_with("_UCell"),
    dplyr::ends_with("_raw"),
    dplyr::ends_with("_WT_axis")
  )

readr::write_csv(
  cell_score_export,
  file.path(TABLE_DIR, "all_cohort_cell_level_program_scores.csv.gz")
)

run_summary <- tibble::tibble(
  dataset_version = DATASET_VERSION,
  presentation_tag = PRESENTATION_TAG,
  cohorts = paste(COHORT_IDS, collapse = ";"),
  total_nuclei = nrow(cell_scores),
  curated_epithelial_nuclei = nrow(curated_epithelial),
  harmonized_epithelial_states = dplyr::n_distinct(
    curated_epithelial$harmonized_state
  ),
  UCell_max_rank = UCELL_MAX_RANK,
  minimum_nuclei_per_diet = MIN_NUCLEI_PER_DIET,
  balanced_sampling_iterations = DOWNSAMPLING_ITERATIONS,
  replicated_diet_design = FALSE,
  biological_inference_performed = FALSE,
  PDFs_requested = nrow(figure_manifest),
  PDFs_created = sum(
    file.exists(file.path(PLOT_DIR, figure_manifest$pdf))
  )
)

readr::write_csv(
  run_summary,
  file.path(MEETING_OUTPUT_DIR, "meeting_figure_run_summary.csv")
)

writeLines(
  capture.output(utils::sessionInfo()),
  file.path(MEETING_OUTPUT_DIR, "sessionInfo.txt")
)

missing_figures <- figure_manifest$pdf[
  !file.exists(file.path(PLOT_DIR, figure_manifest$pdf))
]

writeLines(
  c(
    paste("Meeting figure preparation completed:", Sys.time()),
    paste("Dataset version:", DATASET_VERSION),
    paste("Cohorts:", paste(COHORT_IDS, collapse = ", ")),
    paste("Total nuclei scored:", nrow(cell_scores)),
    paste("Curated epithelial nuclei:", nrow(curated_epithelial)),
    paste("PDFs created:", run_summary$PDFs_created),
    paste(
      "Missing/omitted PDFs:",
      if (length(missing_figures) == 0L) {
        "none"
      } else {
        paste(missing_figures, collapse = "; ")
      }
    ),
    "Replicated diet design: FALSE",
    "Biological inference performed: FALSE"
  ),
  file.path(MEETING_OUTPUT_DIR, "STEP_MEETING_FIGURES_COMPLETE.txt")
)

run_summary
```

# 20. Console debrief

```{r print-debrief}
cat("\nOutput directory:\n", MEETING_OUTPUT_DIR, "\n\n", sep = "")

cat("WT composition changes with largest absolute values:\n")
print(
  epithelial_composition_effects |>
    dplyr::filter(cohort_id == "wt") |>
    dplyr::arrange(dplyr::desc(abs(HFD_minus_CON_percentage_points))) |>
    dplyr::select(
      harmonized_state,
      percent_CON,
      percent_HFD,
      HFD_minus_CON_percentage_points
    ) |>
    dplyr::slice_head(n = 8L),
  n = Inf,
  width = Inf
)

cat("\nMatched ISC-like expected-direction program effects:\n")
print(
  isc_score_effects |>
    dplyr::select(
      cohort,
      program_display,
      CON_nuclei,
      HFD_nuclei,
      balanced_effect_expected_direction,
      balanced_expected_direction_q25,
      balanced_expected_direction_q75,
      sign_concordance_percent
    ),
  n = Inf,
  width = Inf
)

cat("\nPersistent stringent genes:\n")
print(
  persistent_gene_meeting_table |>
    dplyr::select(
      gene,
      reference_direction,
      RNA_reference_aligned_log2FC,
      representative_ATAC_reference_aligned_log2FC,
      multimodal_validation_class
    ),
  n = Inf,
  width = Inf
)
```
